\section*{Chapter \ref{ch:g6:wholes} --- Whole Numbers}

\begin{solution}{exo:g6:wholes:1}
$3\,047$; \quad $20\,500$; \quad $104$. \quad
$60\,013$: ``sixty thousand and thirteen''.
\end{solution}

\begin{solution}{exo:g6:wholes:2}
Hundreds digit: $2$; units digit: $9$; the $4$ counts the
ten-thousands. Decomposition:
\[
47\,209 = 4 \times 10\,000 + 7 \times 1000 + 2 \times 100 + 0 \times 10
+ 9 .
\]
\end{solution}

\begin{solution}{exo:g6:wholes:3}
$507 < 570$ (tens: $0 < 7$); \quad
$1\,099 > 989$ (four digits beat three); \quad
$23\,456 < 23\,465$ (tens: $5 < 6$); \quad
$9\,999 < 10\,000$ (four digits against five).
\end{solution}

\begin{solution}{exo:g6:wholes:4}
By digit count, $978$ and $1\,001$ come before the three
four-digit numbers starting with $2$; among those, compare the tens and
units: $2\,034 < 2\,043 < 2\,304$. Final order:
\[
978 < 1\,001 < 2\,034 < 2\,043 < 2\,304 .
\]
\end{solution}

\begin{solution}{exo:g6:wholes:5}
$348 + 576 = 924$; \quad $805 - 268 = 537$ (check:
$537 + 268 = 805$); \quad $307 \times 26 = 307 \times 20 + 307 \times 6
= 6\,140 + 1\,842 = 7\,982$.
\end{solution}

\begin{solution}{exo:g6:wholes:6}
$38 + 45 + 12 + 5 = (38 + 12) + (45 + 5) = 50 + 50 = 100$.

$2 \times 83 \times 50 = (2 \times 50) \times 83 = 100 \times 83 =
8\,300$.

$4 \times 78 \times 25 = (4 \times 25) \times 78 = 100 \times 78 =
7\,800$.
\end{solution}

\begin{solution}{exo:g6:wholes:7}
$95 = 7 \times 13 + 4$ (\omterm{def:g3:division:remainder}{quotient} $13$, \omterm{def:g3:division:remainder}{remainder} $4$).

$230 = 6 \times 38 + 2$ (\omterm{def:g3:division:remainder}{quotient} $38$, \omterm{def:g3:division:remainder}{remainder} $2$).

$504 = 8 \times 63 + 0$ (\omterm{def:g3:division:remainder}{quotient} $63$, \omterm{def:g3:division:remainder}{remainder} $0$: $504$ is
\omterm{def:g6:wholes:divisible}{divisible} by $8$).
\end{solution}

\begin{solution}{exo:g6:wholes:8}
The equality is true, but the \omterm{def:g3:division:remainder}{remainder} $13$ is \emph{not} smaller than
the \omterm{def:g5:division:multiple}{divisor} $8$, so it is not the Euclidean division. Take one more
$8$ out of the \omterm{def:g3:division:remainder}{remainder}: $173 = 8 \times 21 + 5$, with $0 \leq 5 < 8$:
\omterm{def:g3:division:remainder}{quotient} $21$, \omterm{def:g3:division:remainder}{remainder} $5$.
\end{solution}

\begin{solution}{exo:g6:wholes:9}
\emph{1.} $2\,000 = 12 \times 166 + 8$: the farm fills $166$ full
boxes and $8$ eggs are left over.

\emph{2.} $150 = 12 \times 12 + 6$: twelve boxes contain only $144$
eggs, not enough. The bakery must buy $13$ boxes.
\end{solution}

\begin{solution}{exo:g6:wholes:10}
$100 = 7 \times 14 + 2$: one hundred days are $14$ full weeks and $2$
days more. Fourteen weeks later it is again a Tuesday; two days after a
Tuesday is a \emph{Thursday}.
\end{solution}

\begin{solution}{exo:g6:wholes:11}
The largest possible \omterm{def:g3:division:remainder}{remainder} in a division by $9$ is $8$ (the
\omterm{def:g3:division:remainder}{remainder} must be smaller than the \omterm{def:g5:division:multiple}{divisor}). So the dividend is
\[
9 \times 56 + 8 = 504 + 8 = 512 .
\]
\end{solution}

\begin{solution}{exo:g6:wholes:12}
Let the digits be $h$ (hundreds), $0$ (tens), $u$ (units). We know
$h = 2 \times u$ and $h + 0 + u = 9$, so $2u + u = 9$, giving
$3u = 9$, $u = 3$, and $h = 6$. The number is $603$. Check: $6$ is
twice $3$, the tens digit is $0$, and $6 + 0 + 3 = 9$.
\end{solution}

\begin{solution}{pb:g6:wholes:1}
\textbf{1.} For instance $(1 + 4) + (2 + 3) + 5 = 5 + 5 + 5 =
15$.

\textbf{2.} Each pair is worth $11$: $1 + 10$, $2 + 9$, $3 + 8$,
$4 + 7$, $5 + 6$. That is $5$ pairs (ten numbers, two per pair),
so the \omterm{def:g1:addition:def}{sum} is $5 \times 11 = 55$.

\textbf{3.} Pairs worth $21$ ($1 + 20$, $2 + 19, \dots$), and
there are $10$ of them: $10 \times 21 = 210$.

\textbf{4.} In each pair, when the first number grows by $1$ the
second shrinks by $1$, so the total never changes:
$1 + 100 = 2 + 99 = 3 + 98 = \dots = 101$. The hundred numbers
make $50$ pairs (two numbers each), the innermost being
$50 + 51 = 101$.

\textbf{5.} $50$ pairs of $101$:
\[
1 + 2 + \dots + 100 = 50 \times 101 = 5\,050 .
\]

\textbf{6.} With nine numbers, the middle one, $5$, has no
partner. Around it: $1 + 9 = 2 + 8 = 3 + 7 = 4 + 6 = 10$, four
pairs. Total: $4 \times 10 + 5 = 45$.

\textbf{7.} Each column is worth $1 + 9 = 2 + 8 = \dots =
9 + 1 = 10$, and there are $9$ columns: the two lines together
are worth $9 \times 10 = 90$. But the two lines are the
\emph{same} \omterm{def:g1:addition:def}{sum} written twice, so one copy is worth \omterm{def:g3:division:half}{half}:
$1 + 2 + \dots + 9 = 90 \div 2 = 45$. No number is ever left
without a partner --- its partner is just underneath it.

\textbf{8.} For $1$ to $50$: $50$ columns worth $51$, double \omterm{def:g1:addition:def}{sum}
$50 \times 51 = 2\,550$, so the \omterm{def:g1:addition:def}{sum} is $2\,550 \div 2 = 1\,275$.
For $1$ to $200$: $200$ columns worth $201$, double \omterm{def:g1:addition:def}{sum}
$40\,200$, \omterm{def:g1:addition:def}{sum} $20\,100$.

\textbf{9.} Every \omterm{def:g3:numbers:evenodd}{even number} is the double of a number of
Gauss's \omterm{def:g1:addition:def}{sum}: $2 = 2 \times 1$, $4 = 2 \times 2$, \dots,
$200 = 2 \times 100$. So the whole \omterm{def:g1:addition:def}{sum} is the double of Gauss's:
\[
2 + 4 + \dots + 200 = 2 \times 5\,050 = 10\,100 .
\]

\textbf{10.} The numbers from $101$ to $200$ are the numbers from
$1$ to $200$ with the numbers from $1$ to $100$ removed:
\[
101 + 102 + \dots + 200 = 20\,100 - 5\,050 = 15\,050 .
\]

\textbf{11.} Line up the friends. The first shakes $7$ hands (one
with each of the others). The second has already greeted the
first, so he shakes $6$ \emph{new} hands; the third, $5$; and so
on, the seventh shaking $1$ and the eighth none that is new.
Every handshake is counted exactly once:
\[
7 + 6 + 5 + 4 + 3 + 2 + 1 = 28 \text{ handshakes}
\]
(pair around the middle: $(7+1) + (6+2) + (5+3) + 4 =
8 + 8 + 8 + 4 = 28$).

\textbf{12.} $1 + 2 + \dots + 12$: twelve columns worth $13$,
double \omterm{def:g1:addition:def}{sum} $12 \times 13 = 156$, so
$156 \div 2 = 78$ cans.

\textbf{13.} The staircase of $1 + 2 + 3 + 4$ squares, plus the
same staircase upside down, fill a rectangle of $4$ rows and $5$
columns exactly: each row of the rectangle is one column of the
two-line trick (a step of the rising staircase completed by a
step of the falling one, $1 + 4$, $2 + 3$, $3 + 2$, $4 + 1$ ---
always $5$). So twice the \omterm{def:g1:addition:def}{sum} is $4 \times 5 = 20$ squares, and
$1 + 2 + 3 + 4 = 10$: the picture \emph{is} question 7.

\textbf{14.} $1 + 2 + \dots + 12 = 78$ strikes in twelve hours
(question 12), hence $78 \times 2 = 156$ strikes in a whole day.

\textbf{15.} Divide by $60$: $5\,050 = 60 \times 84 + 10$, so
$5\,050$ seconds is $84$ minutes and $10$ seconds. Divide the
minutes by $60$: $84 = 60 \times 1 + 24$, so $84$ minutes is
$1$ hour $24$ minutes. In all: $1$ h $24$ min $10$ s --- a little
\emph{less} than an hour and a \omterm{def:g3:division:half}{half} ($1$ h $30$ min).
\end{solution}
