\section*{Chapter \ref{ch:g6:decimals} --- Decimal Numbers}

\begin{solution}{exo:g6:decimals:1}
$5.37$; \quad $0.9$; \quad $12.004$; \quad $8.05$.
\end{solution}

\begin{solution}{exo:g6:decimals:2}
Tenths digit: $3$; hundredths digit: $5$; the $8$ counts the tens.
\[
86.354 = 8 \times 10 + 6 + 3 \times \frac{1}{10} + 5 \times
\frac{1}{100} + 4 \times \frac{1}{1000}.
\]
\end{solution}

\begin{solution}{exo:g6:decimals:3}
$5.3 > 5.29$ (compare $5.30$ with $5.29$); \quad
$0.7 = 0.70$ (final \omterm{def:g1:counting:zero}{zeros} change nothing); \quad
$2.09 < 2.9$ (tenths: $0 < 9$); \quad
$14.5 > 14.49$.
\end{solution}

\begin{solution}{exo:g6:decimals:4}
Pad to two decimals: $4.20$; $4.05$; $4.51$; $4.15$; $4.50$. Order:
\[
4.05 < 4.15 < 4.2 < 4.5 < 4.51 .
\]
\end{solution}

\begin{solution}{exo:g6:decimals:5}
Each tick is one tenth. $A$: $6.3$; \quad $B$: $6.7$; \quad
$C$: $7.2$.
\end{solution}

\begin{solution}{exo:g6:decimals:6}
$45.80 + 7.65 = 53.45$.

$23.40 - 8.72 = 14.68$ (check: $14.68 + 8.72 = 23.4$).

$56 \times 24 = 1344$, and two decimal digits in the factors:
$5.6 \times 2.4 = 13.44$.
\end{solution}

\begin{solution}{exo:g6:decimals:7}
$6.42 \times 10 = 64.2$; \quad
$0.35 \times 1000 = 350$; \quad
$78.1 \div 100 = 0.781$; \quad
$5 \div 1000 = 0.005$.
\end{solution}

\begin{solution}{exo:g6:decimals:8}
$2.4$ m $= 240$ cm; \quad $370$ g $= 0.37$ kg (divide by $1000$);
\quad $0.85$ km $= 850$ m; \quad $56$ mm $= 0.056$ m.
\end{solution}

\begin{solution}{exo:g6:decimals:9}
To the unit: $24$ (next digit $8$: \omterm{def:g4:numbers:round}{round} up). To the tenth: $23.9$.
To the hundredth: $23.87$. And $9.97$ to the tenth: the next digit is
$7$, so \omterm{def:g4:numbers:round}{round} up the $9$ tenths --- $10.0$.
\end{solution}

\begin{solution}{exo:g6:decimals:10}
Price of the baguettes: $3 \times 1.15 = 3.45$. Change:
$5 - 3.45 = 1.55$.
\end{solution}

\begin{solution}{exo:g6:decimals:11}
Between $7.4$ and $7.5$: for instance $7.45$ (or $7.41$, $7.499$,
\dots). Between $3.99$ and $4$: for instance $3.995$. In both cases
there are \emph{infinitely many} such numbers: one can always add more
decimal places.
\end{solution}

\begin{solution}{exo:g6:decimals:12}
Largest: put the biggest digits first and the point as late as
possible: $85.2$. Smallest: smallest digits first and the point as
early as possible: $2.58$.
\end{solution}

\begin{solution}{pb:g6:decimals:1}
\textbf{1.} $3$ is larger: padding, $3.000 > 2.999$. The
\omterm{ex:g1:subtraction:difference}{difference} is $3.000 - 2.999 = 0.001$, one thousandth.

\textbf{2.} Strictly between $7.4 = 7.40$ and $7.5 = 7.50$ lie
\[
7.41,\ 7.42,\ 7.43,\ 7.44,\ 7.45,\ 7.46,\ 7.47,\ 7.48,\ 7.49 :
\]
nine numbers, one per new graduation mark.

\textbf{3.} The same picture, ten times smaller: between $7.440$
and $7.450$ lie $7.441, 7.442, \dots, 7.449$ --- nine numbers
again.

\textbf{4.} Between $7.400$ and $7.500$, the numbers with three
decimal digits are $7.401, 7.402, \dots, 7.499$: all the
thousandths from $401$ to $499$, that is $99$ numbers.

\textbf{5.} The zoom recipe: write both numbers with the same
number of decimal places, then \emph{add one more decimal place}
--- the smaller number followed by a digit from $1$ to $9$ lands
strictly between the two. Indeed $5.67 = 5.670$ and
$5.68 = 5.680$, and
\[
5.670 < 5.675 < 5.680 ;
\]
likewise $0.1999 = 0.19990 < 0.19995 < 0.20000 = 0.2$. Between
two neighbouring graduations there is always a whole new level of
nine finer graduations.

\textbf{6.} $3 < 3.05 < 3.1$; then $3 < 3.005 < 3.01$; then
$3 < 3.0005 < 3.001$. Each candidate is beaten by one more zoom.

\textbf{7.} Whatever number Tom proposes --- call it his
candidate, strictly greater than $3$ --- question 5 produces a
number strictly between $3$ and the candidate. So the candidate
was \emph{not} the closest number to $3$: something \omterm{def:g3:numbers:evenodd}{even} closer
exists. Since this happens to every candidate without exception,
no number can be ``the number just after $3$'': it simply does
not exist.

\textbf{8.} $2.9 < 2.95 < 3$, \ $2.99 < 2.995 < 3$, \
$2.999 < 2.9995 < 3$. And $3 - 2.999 = 0.001$. With twenty
nines, the \omterm{ex:g1:subtraction:difference}{difference} is a \omterm{def:g4:decimals:point}{decimal point} followed by nineteen
\omterm{def:g1:counting:zero}{zeros} and a $1$ --- one unit in the twentieth decimal place:
tiny, but not \omterm{def:g1:counting:zero}{zero}. However many nines Tom writes, he never
reaches $3$.

\textbf{9.} Whole numbers climb in steps of $1$: more than $7$
but less than $8$ would mean a whole number of units strictly
between $7$ and $8$ units, and there is none. The zoom escapes
this only by writing digits \emph{after the point} --- exactly
what whole numbers do not have.

\textbf{10.} The smallest length rounding to $3.7$ is $3.65$ cm:
its next digit is $5$, which \omterm{def:g4:numbers:round}{rounds} \emph{up}
(\cref{def:g6:decimals:rounding}). And $3.75$ \omterm{def:g4:numbers:round}{rounds} up to $3.8$
for the same reason. So ``$3.7$ cm to the nearest tenth'' means:
at least $3.65$ cm, and strictly less than $3.75$ cm --- a whole
zoomed-in segment of possible true lengths hides behind one
rounded value.

\textbf{11.} $0.0999 < 0.123 < 0.58 < 0.6$. One sentence:
comparing digit by digit from the left, $0.0999$ has $0$ tenths
while $0.6$ has $6$, and the comparison is settled there --- the
number of digits written says nothing about size
(\cref{met:g6:decimals:compare}).

\textbf{12.} No: in cents, $4.99$ euros is $499$ cents and $5$
euros is $500$ cents --- consecutive \emph{whole} numbers, with
nothing between them. Prices, having exactly two decimal places,
are really whole numbers of cents in disguise: like the whole
numbers of question 9, they do have next-door neighbours. The
endless zoom needs the right to write ever more decimal places.

\textbf{13.} The candidates near $6$ are $5.82$ and $8.25$ (a
number starting with $2$, or with $58$, $25$, $82$, $85$, is far
from $6$). Distances: $6 - 5.82 = 0.18$ and $8.25 - 6 = 2.25$.
The closest is $5.82$.

\textbf{14.} For instance $3.1415$, since
$3.1410 < 3.1415 < 3.1420$; then $3.14159$, since
$3.14150 < 3.14159 < 3.14160$. (Both are real steps in the
actual hunt for $\pi = 3.14159\dots$)

\textbf{15.} Suppose someone names a number, as large as they
like. One zoom turns every pair of neighbouring graduations
between $0$ and $1$ into nine new numbers, and the zoom can be
repeated forever --- questions 2, 3 and 4 were only the first
two levels. Repeating it enough times produces more \omterm{def:g6:decimals:places}{decimal
numbers} between $0$ and $1$ than the named number. So no number
is large enough to count them: there are infinitely many.
\end{solution}
