\section*{Chapter \ref{ch:g6:lines} --- Lines, Circles and Angles}

\begin{solution}{exo:g6:lines:1}
Free construction. The \omterm{def:g6:lines:objects}{segment} stops at $A$ and $B$; the ray starts at
$C$ and continues past $A$; the line continues on both sides of $B$
and $C$.
\end{solution}

\begin{solution}{exo:g6:lines:2}
``$[AB] = [BA]$'': \emph{true} --- the part between the two points does
not depend on the order.

``$[AB) = [BA)$'': \emph{false} --- $[AB)$ starts at $A$, $[BA)$ starts
at $B$; they point in opposite directions.

``$(AB) = (BA)$'': \emph{true} --- both names describe the same
unlimited line.
\end{solution}

\begin{solution}{exo:g6:lines:3}
Steps: slide the set square along $d$ until its \omterm{def:g6:lines:perp}{perpendicular} edge
passes through $P$; draw that \omterm{def:g6:lines:perp}{perpendicular}, call it $p$. Then draw
the \omterm{def:g6:lines:perp}{perpendicular} to $p$ through $P$ the same way: by
\cref{prop:g6:lines:facts} it is \omterm{def:g6:lines:perp}{parallel} to $d$.
\end{solution}

\begin{solution}{exo:g6:lines:4}
$a$ and $b$ are both \omterm{def:g6:lines:perp}{perpendicular} to the same line $c$, so
$a \parallel b$ (fact 1). For $e$ and $c$: both are \omterm{def:g6:lines:perp}{perpendicular} to
the same line $a$ (we are told $e \perp a$, and $a \perp c$ means
$c \perp a$ too), so fact 1 applies again: $e \parallel c$.
\end{solution}

\begin{solution}{exo:g6:lines:5}
$OM = 3$ cm exactly ($M$ is on the circle); $ON < 3$ cm ($N$ inside);
$OP > 3$ cm ($P$ outside).
\end{solution}

\begin{solution}{exo:g6:lines:6}
\omterm{def:g3:shapes:shapes}{Radius} $= 9 \div 2 = 4.5$ cm. \quad \omterm{def:g4:geometry:circle}{Diameter} $= 2 \times 2.6 = 5.2$ cm.
\end{solution}

\begin{solution}{exo:g6:lines:7}
$\widehat{SRT} = 72^\circ$: acute. \quad
$\widehat{ABC} = 148^\circ$: obtuse. \quad
$\widehat{MON} = 90^\circ$: right. (The \omterm{def:g5:solids:def}{vertex} is always the middle
letter.)
\end{solution}

\begin{solution}{exo:g6:lines:8}
Measures depend on the triangle drawn, but the \omterm{def:g1:addition:def}{sum} of the three angles
is always (very close to) $180^\circ$ --- small \omterm{ex:g1:subtraction:difference}{differences} come from
measuring imprecision. \Cref{ch:g7:triangles} proves that the \omterm{def:g1:addition:def}{sum} is
exactly $180^\circ$.
\end{solution}

\begin{solution}{exo:g6:lines:9}
Free construction. To get $130^\circ$ from $65^\circ$ without the
protractor: copy the $65^\circ$ angle twice side by side ($65 + 65 =
130$), for instance with tracing paper or a compass-and-ruler angle
copy.
\end{solution}

\begin{solution}{exo:g6:lines:10}
The points at $3$ cm from $A$ form the circle of center $A$ and \omterm{def:g3:shapes:shapes}{radius}
$3$ cm; those at $2$ cm from $B$ form the circle of center $B$ and
\omterm{def:g3:shapes:shapes}{radius} $2$ cm. The required points are the intersections of the two
circles: two points if the circles cross ($AB$ strictly between $1$ and
$5$ cm), one if they touch ($AB = 5$ cm or $AB = 1$ cm), none if they
are too far apart or one inside the other.
\end{solution}

\begin{solution}{exo:g6:lines:11}
The $12$ hour marks split the full turn into $12$ angles of
$360 \div 12 = 30^\circ$. At 3 o'clock the hands span $3$ marks:
$90^\circ$ (a \omterm{def:g3:shapes:rightangle}{right angle}). At 5 o'clock: $5 \times 30 = 150^\circ$.
At 6 o'clock: $180^\circ$ (a straight angle).
\end{solution}

\begin{solution}{exo:g6:lines:12}
Draw two arcs of \omterm{def:g3:shapes:shapes}{radius} $6$ cm centered at $A$ and at $B$; their
crossing point is $C$. All three sides measure $6$ cm: the triangle is
equilateral, and each angle measures $60^\circ$ (the measure confirms
it). Conjecture: an equilateral triangle has three equal angles of
$60^\circ$.
\end{solution}

\begin{solution}{pb:g6:lines:1}
\textbf{1.} \omterm{def:g3:division:half}{Half} a turn: $360 \div 2 = 180^\circ$. A \omterm{def:g3:division:half}{quarter}:
$360 \div 4 = 90^\circ$. A twelfth: $360 \div 12 = 30^\circ$.

\textbf{2.} Twelve equal angles filling $360^\circ$:
$30^\circ$ between neighbouring marks. At 3 o'clock the hands
span $3$ marks: $3 \times 30 = 90^\circ$ --- a \omterm{def:g3:shapes:rightangle}{right angle}, as
found in \cref{exo:g6:lines:11}.

\textbf{3.} At 1 o'clock: $1 \times 30 = 30^\circ$. At
4 o'clock: $4 \times 30 = 120^\circ$. At 7 o'clock the hands
span $7$ marks on one side, $7 \times 30 = 210^\circ$, so the
angle between the hands is the other side:
$360 - 210 = 150^\circ$.

\textbf{4.} $360^\circ$ in $60$ minutes: $360 \div 60 =
6^\circ$ per minute.

\textbf{5.} $30^\circ$ in $60$ minutes: $30 \div 60 = 0.5^\circ$
per minute --- \omterm{def:g3:division:half}{half} a degree. The hour hand creeps, the minute
hand strides.

\textbf{6.} In the $30$ minutes since 3:00, the hour hand has
moved $30 \times 0.5 = 15^\circ$ past the $3$: it sits exactly
halfway between the $3$ and the $4$, at $90 + 15 = 105^\circ$
from the $12$. The minute hand points at the $6$:
$180^\circ$. Angle between the hands:
$180 - 105 = 75^\circ$.

\textbf{7.} Computation beats the guess: the minute hand is at
$180^\circ$, but the hour hand has left the $6$ --- it is at
$6 \times 30 + 30 \times 0.5 = 180 + 15 = 195^\circ$. The angle
between the hands is $195 - 180 = 15^\circ$: close, but not on
top of each other.

\textbf{8.} Minute hand at the $3$: $90^\circ$. Hour hand:
$9 \times 30 + 15 \times 0.5 = 270 + 7.5 = 277.5^\circ$.
\omterm{ex:g1:subtraction:difference}{Difference}: $277.5 - 90 = 187.5^\circ$, so the angle between the
hands is $360 - 187.5 = 172.5^\circ$ --- almost a straight
angle.

\textbf{9.} At 1:00 the minute hand ($0^\circ$) is \emph{behind}
the hour hand ($30^\circ$). At 1:10 the minute hand
($60^\circ$) is \emph{ahead} of it
($30 + 10 \times 0.5 = 35^\circ$). The minute hand runs faster
($6^\circ$ per minute against $0.5^\circ$), so between 1:00 and
1:10 it catches up and passes the hour hand --- at the moment of
passing, the two hands are exactly on top of each other.

\textbf{10.} The same catching-up happens over and over: the
hands meet at 12:00 exactly, then at about 1:05, 2:11, 3:16,
4:22, 5:27, 6:33, 7:38, 8:44, 9:49 and 10:55. The meeting one
might expect between 11 and 12 falls exactly \emph{at} 12:00,
which already begins the next \omterm{def:g4:numbers:round}{round} --- so in twelve hours the
hands coincide $11$ times, not $12$. In a whole day: $22$ times.

\textbf{11.} Angles sharing a \omterm{def:g5:solids:def}{vertex}, with no overlap and no
gap, tile the full turn around that \omterm{def:g5:solids:def}{vertex}: sweeping once around
the point passes through each angle exactly once, and a full
sweep is $360^\circ$. So the measures add up to $360^\circ$ ---
whatever the number of angles and their sizes.

\textbf{12.} Each hour of the day is worth
$360 \div 24 = 15^\circ$. So:
\[
\text{sleep } 9 \times 15 = 135^\circ, \quad
\text{school } 90^\circ, \quad
\text{play } 45^\circ, \quad
\text{meals } 30^\circ, \quad
\text{other } 60^\circ,
\]
and $135 + 90 + 45 + 30 + 60 = 360^\circ$: the pie is full, with
no gap and no overlap (question 11).

\textbf{13.} Draw a circle with the compass and one \omterm{def:g3:shapes:shapes}{radius} (the
starting line). Place the protractor's center on the center of
the circle, its $0^\circ$ line on the \omterm{def:g3:shapes:shapes}{radius}, and mark
$135^\circ$; draw the new \omterm{def:g3:shapes:shapes}{radius}: the sleep sector is done. Then
place the $0^\circ$ line on \emph{that} \omterm{def:g3:shapes:shapes}{radius} and mark
$90^\circ$ for school, and so on --- each sector starts where
the previous one ends. After the last sector the drawing closes
up exactly on the starting \omterm{def:g3:shapes:shapes}{radius}.

\textbf{14.} $90^\circ$ is $\frac{90}{360} = \frac14$ of the
disk (\cref{met:g6:fractions:simplify}), so it represents a
\omterm{def:g3:division:half}{quarter} of the day: $24 \div 4 = 6$ hours.

\textbf{15.} The hour hand makes one turn in $12$ hours: $2$
turns per day. The minute hand, one turn per hour: $24$ turns.
The second hand, one turn per minute: $24 \times 60 = 1\,440$
turns per day.
\end{solution}
