\section*{Chapter \ref{ch:g6:symmetry} --- Axial Symmetry}

\begin{solution}{exo:g6:symmetry:1}
$A'$ is two squares to the \emph{right} of $d$, on the same row as
$A$; $B'$ is five squares to the right, one square higher --- mirror
positions.
\end{solution}

\begin{solution}{exo:g6:symmetry:2}
\omterm{def:g6:lines:perp}{Perpendicular} to $d$ through $M$, foot $H$; then $M'$ on the other
side with $HM' = MH = 3$ cm. The distance $MM'$ is
$3 + 3 = 6$ cm.
\end{solution}

\begin{solution}{exo:g6:symmetry:3}
Vertical axis (like A): A, H, I, M, O, T, U, V, W, X, Y.
Horizontal axis (like B): B, C, D, E, H, I, K, O, X.
Both: H, I, O, X.
(Exact answers depend slightly on how the letters are drawn.)
\end{solution}

\begin{solution}{exo:g6:symmetry:4}
\omterm{def:g6:shapes:triangles}{Equilateral triangle}: $3$ axes. Rectangle: $2$ (the \omterm{def:g6:symmetry:bisector}{perpendicular
bisectors} of its sides). \omterm{def:g6:shapes:quadrilaterals}{Rhombus}: $2$ (its diagonals). Square: $4$.
Circle: infinitely many (every line through the center).
\end{solution}

\begin{solution}{exo:g6:symmetry:5}
\emph{False.} Folding a (non-square) rectangle along a diagonal does
not make the two halves coincide: the two triangles have the same
shape but sit differently --- try it with a sheet of paper. Only for a
square do the diagonals work as axes.
\end{solution}

\begin{solution}{exo:g6:symmetry:6}
Reflections preserve lengths, angles and \omterm{def:g4:measure:perimeter}{perimeters}
(\cref{prop:g6:symmetry:preserved}): $A'B' = 4$ cm,
$\widehat{B'A'C'} = 70^\circ$, \omterm{def:g4:measure:perimeter}{perimeter} of $A'B'C' = 12$ cm.
\end{solution}

\begin{solution}{exo:g6:symmetry:7}
Compass construction as in \cref{met:g6:symmetry:bisector}. For any
$M$ on the line, the ruler confirms $MA = MB$
(\cref{prop:g6:symmetry:bisector}).
\end{solution}

\begin{solution}{exo:g6:symmetry:8}
All the points at equal distance from $A$ and $B$ form the
\omterm{def:g6:symmetry:bisector}{perpendicular bisector} of $[AB]$ (\cref{prop:g6:symmetry:bisector}).
The fountain can go anywhere on that line (within the garden).
\end{solution}

\begin{solution}{exo:g6:symmetry:9}
Each \omterm{def:g5:solids:def}{vertex} is reflected \omterm{def:g6:lines:perp}{perpendicular} to the axis: for the
$45^\circ$ axis, a point $k$ squares to the right of the axis lands
$k$ squares above it (and vice versa) --- horizontal and vertical
displacements are exchanged. Reflect the \omterm{def:g5:solids:def}{vertices} one by one, then
join them in order.
\end{solution}

\begin{solution}{exo:g6:symmetry:10}
The \omterm{def:g6:symmetry:def}{symmetric} of $M$ is the second intersection of the circle with the
\omterm{def:g6:lines:perp}{perpendicular} to $d$ through $M$ --- it lies on the circle. Reason:
the reflection preserves distances and fixes $O$ (which is on $d$), so
the image of any point at distance $r$ from $O$ is again at distance
$r$ from $O$: the circle maps onto itself.
\end{solution}

\begin{solution}{exo:g6:symmetry:11}
Each crossing point $P$ was drawn with the same compass opening $\rho$
from $A$ and from $B$: so $PA = \rho = PB$. By
\cref{prop:g6:symmetry:bisector}, a point equidistant from $A$ and $B$
lies on the \omterm{def:g6:symmetry:bisector}{perpendicular bisector} of $[AB]$; two such points determine
the line.
\end{solution}

\begin{solution}{exo:g6:symmetry:12}
For any bounce point $P$ on the wall, $PB = PB'$ (reflection across
$d$ preserves distances), so the path length
$AP + PB = AP + PB'$. The broken line $A \to P \to B'$ is shortest
when it is straight, i.e.\ when $P$ is on the \omterm{def:g6:lines:objects}{segment} $[AB']$. So the
best bounce point is the intersection of $[AB']$ with the wall.
\end{solution}

\begin{solution}{pb:g6:symmetry:1}
\textbf{1.} Construction as in \cref{met:g6:symmetry:bisector};
for any $M$ on the line, the ruler confirms $MA = MB$
(\cref{prop:g6:symmetry:bisector}).

\textbf{2.} The fountain question needs the second direction:
\emph{all} the equidistant points are on the bisector, so the
possible spots are exactly that line and nothing else. And a
point not on the bisector is therefore \emph{not} equidistant:
it is strictly closer to one of the two trees.

\textbf{3.} The two paths share the leg $[MP]$, and $PB = PA$
since $P$ is on the bisector. So the path $M \to P \to B$ has
the same length as the path $M \to P \to A$, namely $MP + PA$.
Now $MB$ is the straight route to $B$ --- here it \emph{is} the
path through $P$, so $MB = MP + PA$; and the straight route to
$A$ is shorter than the detour through $P$: $MA < MP + PA = MB$.
Every point on $A$'s side of the bisector is closer to $A$.

\textbf{4.} The border is the \omterm{def:g6:symmetry:bisector}{perpendicular bisector} of $[AB]$:
school $A$ serves exactly the \omterm{def:g3:division:half}{half} of the town on its side of
that line (question 3), school $B$ the other \omterm{def:g3:division:half}{half}, and children
living on the line itself may pick either.

\textbf{5.} $O$ lies on the bisector of $[AB]$, so $OA = OB$;
and on the bisector of $[BC]$, so $OB = OC$
(\cref{prop:g6:symmetry:bisector}).

\textbf{6.} From $OA = OB$ and $OB = OC$: $OA = OC$. By the
second direction of \cref{prop:g6:symmetry:bisector}, a point
equidistant from $A$ and $C$ lies on the \omterm{def:g6:symmetry:bisector}{perpendicular bisector}
of $[AC]$: the third bisector passes through $O$ without being
asked. The three \omterm{def:g6:symmetry:bisector}{perpendicular bisectors} of any triangle are
\emph{concurrent}.

\textbf{7.} The circle of center $O$ and \omterm{def:g3:shapes:shapes}{radius} $OA$ consists of
all points at distance $OA$ from $O$
(\cref{def:g6:lines:circle}); since $OB = OA$ and $OC = OA$, the
points $B$ and $C$ are on it. One center, one \omterm{def:g3:shapes:shapes}{radius}, three
points served: it is the circle through $A$, $B$ and $C$.

\textbf{8.} Two bisectors suffice because the third is automatic
(question 6): their crossing point already satisfies all three
equalities of distance. The circle is drawn with center $O$ and
the compass opened from $O$ to any one of the three \omterm{def:g5:solids:def}{vertices}.

\textbf{9.} Recipe: mark three points $A$, $B$, $C$ on the
surviving arc; draw the \omterm{def:g6:symmetry:bisector}{perpendicular bisectors} of the chords
$[AB]$ and $[BC]$; their crossing point is the center of the
plate, and its distance to $A$ is the \omterm{def:g3:shapes:shapes}{radius}. It works because
the plate's rim is a circle through the three marked points ---
and question 8 finds \emph{the} circle through them. The test
arc rebuilds exactly.

\textbf{10.} If $A$, $B$, $C$ are aligned, the bisector of
$[AB]$ and the bisector of $[BC]$ are both \omterm{def:g6:lines:perp}{perpendicular} to the
same line $(AB) = (BC)$ --- so they are \omterm{def:g6:lines:perp}{parallel} to each other
(\cref{prop:g6:lines:facts}) and, being distinct, never cross.
No point is equidistant from all three, and no circle passes
through three aligned points.

\textbf{11.} Regular \omterm{def:g4:geometry:polygon}{pentagon}: $5$ axes, each through one \omterm{def:g5:solids:def}{vertex}
and the midpoint of the opposite side (\omterm{def:g3:numbers:evenodd}{odd} count: every axis is
of this one kind). Regular \omterm{def:g4:geometry:polygon}{hexagon}: $6$ axes --- three through
opposite \omterm{def:g5:solids:def}{vertices} and three through midpoints of opposite sides
(\omterm{def:g3:numbers:evenodd}{even} count: two kinds, \omterm{def:g3:division:half}{half} and \omterm{def:g3:division:half}{half}). The pattern: a regular
\omterm{def:g4:geometry:polygon}{polygon} with $n$ sides has $n$ axes.

\textbf{12.} The cut goes through both layers at once, so the
two layers lose exactly matching pieces; unfolding is the
reflection across the fold line, which sends each layer's edge
onto the other's. The hole is its own mirror image: the fold
line is its \omterm{def:g6:symmetry:axis}{axis of symmetry} --- whatever was cut.

\textbf{13.} The unfolded hole has at least $2$ axes of
symmetry, the two fold lines, and the cut appears in $4$ copies
(each unfolding doubles the picture: $1 \to 2 \to 4$). Folding
more times in the snowflake style multiplies the copies and the
axes further.

\textbf{14.} $F''$ has the \emph{same} orientation as $F$ (two
mirror flips cancel the mirroring), the same size and the same
height --- it is simply $F$ slid to the right. The slide
measures $8$ squares: exactly \emph{twice} the distance between
the two axes. Two \omterm{def:g6:lines:perp}{parallel} mirrors together do not mirror at
all: they \emph{translate}.

\textbf{15.} $F''$ is again the same size, but now it appears
\emph{upside down}, turned by \omterm{def:g3:division:half}{half} a turn about the crossing
point of the two axes: each point of $F''$ is diametrically
opposite the corresponding point of $F$ through that center.
Two \omterm{def:g6:lines:perp}{perpendicular} mirrors compose into the \emph{half-turn} ---
the central symmetry of \cref{ch:g7:central}.
\end{solution}
