\section*{Chapter \ref{ch:g2:shapes} --- Shapes and Solids}

\begin{solution}{exo:g2:shapes:1}
$5$ sides, $5$ corners: a pentagon. $3$ corners (and $3$ sides): a
triangle. Round: a circle.
\end{solution}

\begin{solution}{exo:g2:shapes:2}
Polygons: square, triangle, hexagon (straight sides). Not a
polygon: the circle (curved border).
\end{solution}

\begin{solution}{exo:g2:shapes:3}
Right angles at, for instance, the corners of the board, a book, a
window; a not-right angle at the open lid of a box or the hands of
a clock at $2$ o'clock.
\end{solution}

\begin{solution}{exo:g2:shapes:4}
Square: $4$ right angles. Rectangle: $4$. A triangle can have at
most $1$ right angle --- two would make the other two sides never
meet.
\end{solution}

\begin{solution}{exo:g2:shapes:5}
Grid copies made by counting squares between corners; comparing
with the model checks the work.
\end{solution}

\begin{solution}{exo:g2:shapes:6}
Die: cube. Can of soup: cylinder. Football: ball. Shoebox: box.
\end{solution}

\begin{solution}{exo:g2:shapes:7}
A cube has $6$ faces, each a square. A box also has $6$ faces
(rectangles).
\end{solution}

\begin{solution}{exo:g2:shapes:8}
The ball rolls (round all over) and the cylinder rolls (on its
curved side). The cube and the box do not: they have only flat
faces and stop on them.
\end{solution}

\begin{solution}{exo:g2:shapes:9}
$6$ sides: a hexagon, with $6$ corners.
\end{solution}

\begin{solution}{exo:g2:shapes:10}
The house outline (square wall with a triangle roof, the roof base
sitting on the square's top side) has $5$ sides and $5$ corners ---
it is a pentagon (bottom, two walls, two roof slopes; the top of
the wall is hidden under the roof).
\end{solution}
