\section*{Chapter \ref{ch:b2:structures} --- Sets and Structures}

\begin{solution}{exo:b2:structures:1}
\emph{Finite subsets of $\N$:} countable --- the set of subsets of
$\intint{0}{n}$ is finite, and the finite subsets form the countable
union over $n$ of these
(\cref{prop:b2:structures:countablestable}~(3)); infinite since it
contains all singletons.

\emph{All subsets of $\N$:} not countable, by Cantor's theorem
(\cref{thm:b2:structures:cantor}~(1) with $E = \N$).

\emph{$\R \setminus \Q$:} not countable --- otherwise $\R = \Q \cup
(\R\setminus\Q)$ would be a union of two countable sets,
contradicting \cref{thm:b2:structures:cantor}~(2).

\emph{Polynomials over $\Q$:} countable --- the polynomials of degree
$\leq n$ inject into $\Q^{n+1}$ (finite products of countable sets),
and take the union over $n$.

\emph{Eventually-zero binary sequences:} countable --- they biject with
finite subsets of $\N$ (the support).
\end{solution}

\begin{solution}{exo:b2:structures:2}
Compute element by element, applying the right factor first.
$\sigma\tau$ sends $1 \mapsto \sigma(1) = 4$, $\;2 \mapsto \sigma(3)
= 5$, $\;3 \mapsto \sigma(7) = 7$, $\;4 \mapsto \sigma(4) = 2$, $\;5
\mapsto \sigma(5) = 3$, $\;6 \mapsto \sigma(6) = 1$, $\;7 \mapsto
\sigma(2) = 6$:
\[
\sigma\tau = (1\,4\,2\,5\,3\,7\,6),
\]
a $7$-cycle. Likewise $\tau\sigma$ sends $1 \mapsto \tau(4) = 4$,
$\;2 \mapsto \tau(6) = 6$, $\;3 \mapsto \tau(5) = 5$, $\;4 \mapsto
\tau(2) = 3$, $\;5 \mapsto \tau(3) = 7$, $\;6 \mapsto \tau(1) = 1$,
$\;7 \mapsto \tau(7) = 2$:
\[
\tau\sigma = (1\,4\,3\,5\,7\,2\,6),
\]
also a $7$-cycle (as expected: $\sigma\tau$ and $\tau\sigma$ are
conjugate, hence share their cycle type).

Orders and signatures: $\sigma$ has cycle type $(4,2)$: order
$\operatorname{lcm}(4,2) = 4$, signature $(-1)^3(-1)^1 = +1$; $\tau$
is a $3$-cycle: order $3$, signature $+1$; both products are
$7$-cycles: order $7$, signature $(-1)^6 = +1$.

$\sigma^{2026}$: $2026 = 4 \times 506 + 2$, so $\sigma^{2026} =
\sigma^2 = (1\,2)(4\,6)$ (square the $4$-cycle; the transposition
squares away).
\end{solution}

\begin{solution}{exo:b2:structures:3}
$\{0,1\}^{\N} \to \mathcal{P}(\N)$: a sequence maps to its support
--- a bijection (indicator functions), no theorem needed.

$\{0,1\}^{\N} \to \intcc{0}{1}$: the base-$3$ map $(a_n) \mapsto
\sum 2a_n 3^{-n-1}$ is injective (two distinct sequences differ
first at rank $N$; the tails cannot compensate a gap of $2\cdot
3^{-N-1}$, since $\sum_{n > N} 2\cdot 3^{-n-1} = 3^{-N-1} <
2\cdot3^{-N-1}$).

$\intcc{0}{1} \to \{0,1\}^{\N}$: binary expansion, choosing (say)
the expansion not ending in all $1$'s: injective.

By Cantor--Bernstein (\cref{thm:b2:structures:cantorbernstein})
applied to the last two injections, $\intcc{0}{1}$ and
$\{0,1\}^{\N}$ are equipotent, hence all three sets are.
\end{solution}

\begin{solution}{exo:b2:structures:4}
Let $c = ab = ba$ and $d = \operatorname{ord}(c)$. First $c^{mn} =
a^{mn} b^{mn} = e$ (commutation allows splitting the power), so $d
\mid mn$. Conversely $c^d = e$ gives $a^d = b^{-d}$; this element
lies in $\langle a\rangle \cap \langle b\rangle$, a subgroup whose
order divides both $m$ and $n$ (Lagrange in each cyclic group),
hence is trivial: $a^d = b^d = e$, so $m \mid d$ and $n \mid d$, and
by coprimality $mn \mid d$. Hence $d = mn$.

In $\mathfrak{S}_3$: take $a = (1\,2)$ (order $2$) and $b =
(1\,2\,3)$ (order $3$), coprime orders, which do not commute: $ab =
(2\,3)$ has order $2 \neq 6$ --- indeed $\mathfrak{S}_3$ has no
element of order $6$. Commutation is essential.
\end{solution}

\begin{solution}{exo:b2:structures:5}
Pair every $x \in G$ with $x^{-1}$. The pairs $\{x, x^{-1}\}$ with
$x \neq x^{-1}$ have two elements and partition their union; the
remaining elements are exactly those with $x = x^{-1}$, i.e.\ $x^2 =
e$. Since $\abs G$ is even and the two-element pairs cover an even
number of elements, the set $\{x : x^2 = e\}$ has even cardinality;
it contains $e$, so it contains at least one other element $x \neq
e$ --- an element of order $2$.
\end{solution}

\begin{solution}{exo:b2:structures:6}
Every element of $A_n$ is a product of an even number of
transpositions (\cref{thm:b2:structures:signature}: decompose into
transpositions; the count is even since the signature is $+1$). It
suffices to write each product of two transpositions with
$3$-cycles:
\[
(a\,b)(a\,c) = (a\,c\,b),
\qquad
(a\,b)(c\,d) = (a\,c\,b)(a\,c\,d) \quad (\text{distinct } a,b,c,d),
\]
(check by evaluation), and $(a\,b)(a\,b) = \mathrm{id}$. So the
$3$-cycles generate $A_n$.
\end{solution}

\begin{solution}{exo:b2:structures:7}
\emph{$(\Q,+) \to (\Z,+)$:} only the zero morphism. For any $x$ and
every $n \geq 1$, $f(x) = n f\bigl(\frac xn\bigr)$ is divisible by
$n$ in $\Z$; the only integer divisible by every $n$ is $0$, so
$f(x) = 0$ for all $x$.

\emph{$(\Z/n\Z, +) \to (\Z/m\Z, +)$:} a morphism is determined by
$c = f(\overline 1)$, which must satisfy $n c \equiv 0 \pmod m$,
i.e.\ $c$ is a multiple of $\frac{m}{\gcd(m,n)}$; there are
$\gcd(m,n)$ such classes, and each choice does define a morphism
(factor $k \mapsto kc$ through $\Z/n\Z$ by the universal property).

\emph{$(\Q, +) \to (\Q_+^*, \times)$:} only the trivial one. If
$f(x) = y$, then for every $n$, $y = f(n \cdot \frac xn) =
f(\frac xn)^n$ is an $n$-th power in $\Q_+^*$. But a rational
$y \neq 1$ cannot be an $n$-th power for all $n$: some prime appears
in $y$ with a nonzero exponent $v$, and $n \nmid v$ for $n > \abs
v$ (exponents of $n$-th powers are multiples of $n$, by unique
factorization). Hence $f \equiv 1$.
\end{solution}

\begin{solution}{exo:b2:structures:8}
$360 = 2^3 \cdot 3^2 \cdot 5$:
$\varphi(360) = 360\bigl(1 - \tfrac12\bigr)\bigl(1 -
\tfrac13\bigr)\bigl(1 - \tfrac15\bigr) = 360 \cdot \tfrac12 \cdot
\tfrac23 \cdot \tfrac45 = 96$.

System: moduli $8, 9, 5$ pairwise coprime, total $360$. From $x
\equiv 3 \pmod 8$ and $x \equiv 5 \pmod 9$: $x = 3 + 8k$ with $3 +
8k \equiv 5 \pmod 9$, i.e.\ $-k \equiv 2$, $k \equiv -2 \equiv 7
\pmod 9$: $x \equiv 3 + 56 = 59 \pmod{72}$. Then $59 + 72\ell \equiv
2 \pmod 5$: $4 + 2\ell \equiv 2$, $2\ell \equiv 3 \equiv 8$, $\ell
\equiv 4 \pmod 5$: $x \equiv 59 + 288 = 347 \pmod{360}$.

Last two digits of $3^{2026}$: mod $4$, $3^{2026} = 9^{1013} \equiv
1$. Mod $25$: $\varphi(25) = 20$ and $2026 = 20\cdot101 + 6$, so
$3^{2026} \equiv 3^6 = 729 \equiv 4 \pmod{25}$. Solve $x \equiv 1
\pmod 4$, $x \equiv 4 \pmod{25}$: $x = 4 + 25k \equiv 1 \pmod 4$
gives $k \equiv 1 \pmod 4$: $x \equiv 29 \pmod{100}$. The last two
digits are $29$.
\end{solution}

\begin{solution}{exo:b2:structures:9}
Let $A$ be a finite integral domain and $a \in A$, $a \neq 0$. The
map $x \mapsto ax$ is injective ($ax = ay \implies a(x - y) = 0
\implies x = y$, no zero divisors); an injective map of a finite set
to itself is surjective (Year 1 volume, the pigeonhole equivalence).
So $1 = ab$ for some $b$: every nonzero element is invertible, $A$
is a field.

$\Z/n\Z$: if $n$ is prime it is an integral domain ($n \mid ab
\implies n \mid a$ or $n \mid b$, Euclid's lemma), finite, hence a
field; if $n = rs$ is composite, $\overline r\,\overline s =
\overline 0$ exhibits zero divisors.
\end{solution}

\begin{solution}{exo:b2:structures:10}
Let $m = \max\{\operatorname{ord}(x) : x \in G\}$, attained at $a$.

\emph{Claim: every $x \in G$ has order dividing $m$.} Suppose some
$x$ has order $q$ with $q \nmid m$: then some prime power $p^k$
divides $q$ but not $m$. Write $m = p^j m'$ with $p \nmid m'$ and $j
< k$. The element $a^{p^j}$ has order $m'$; the element
$x^{q/p^k}$ has order $p^k$; these orders are coprime and the two
elements commute ($G \subseteq K^*$ is abelian), so by
\cref{exo:b2:structures:4} their product has order $p^k m' > p^j m'
= m$: contradicting maximality.

So all $x \in G$ satisfy $x^m = 1$: the polynomial $X^m - 1$ has at
least $\abs G$ roots in the field $K$, whence $\abs G \leq m$
(a nonzero polynomial of degree $m$ has at most $m$ roots, Year 1
volume). But $m = \operatorname{ord}(a) \leq \abs G$ by Lagrange.
Hence $m = \abs G$ and $\langle a \rangle$, of cardinality $m =
\abs G$, is all of $G$: cyclic.

For $K = \Z/p\Z$: $(\Z/p\Z)^*$ is a finite subgroup of $K^*$, hence
cyclic (of order $p - 1$).
\end{solution}

\begin{solution}{exo:b2:structures:11}
\emph{Not cyclic:} the subgroup $\langle \frac pq\rangle$ consists
of the integer multiples of $\frac pq$, all of which have
denominator dividing $q$ (in lowest terms); it therefore misses
$\frac{1}{2q}$. No single generator can reach the unbounded
denominators of $\Q$.

\emph{Not finitely generated:} the subgroup generated by
$\frac{p_1}{q_1}, \dots, \frac{p_k}{q_k}$ consists of rationals
whose denominators divide $Q = q_1 \cdots q_k$ (integer combinations
have denominator dividing $Q$): it misses $\frac{1}{2Q}$.

\emph{Finitely generated subgroups are cyclic:} with $Q$ as above,
the subgroup $H = \langle \frac{p_1}{q_1}, \dots,
\frac{p_k}{q_k}\rangle$ is contained in $\frac{1}{Q}\Z$. The map $x
\mapsto Qx$ is an isomorphism from $\frac1Q\Z$ onto $\Z$ carrying
$H$ to a subgroup of $\Z$, which is $n\Z$ for some $n$ (Year 1
volume): so $H = \frac{n}{Q}\Z$ is cyclic, generated by
$\frac nQ$.
\end{solution}

\begin{solution}{exo:b2:structures:12}
\emph{A countable subset.} Let $E$ be infinite. Construct $a_0,
a_1, a_2, \dots$ inductively: $E$ is nonempty, pick $a_0 \in E$;
if $a_0, \dots, a_n$ are chosen, $E \setminus \{a_0, \dots,
a_n\}$ is nonempty ($E$ is not finite), pick $a_{n+1}$ there. The
$a_n$ are pairwise distinct by construction, so $A = \{a_n : n
\in \N\}$ is a countable subset of $E$.

\emph{Infinite $\implies$ equipotent to a proper subset.} Define
$f \colon E \to E \setminus \{a_0\}$ by $f(a_n) = a_{n+1}$ and
$f(x) = x$ for $x \notin A$. It is injective (the two pieces are
injective with disjoint images) and surjective onto $E \setminus
\{a_0\}$: every $a_{n+1}$ is hit, every $x \notin A$ is hit. So
$E$ is equipotent to the proper subset $E \setminus \{a_0\}$.

\emph{Converse.} If $E$ is finite and $g \colon E \to F$ is a
bijection onto $F \subseteq E$ with $F \neq E$, then $g$ is an
injection of $E$ into itself that is not surjective,
contradicting the pigeonhole principle (Year 1 volume: an
injective self-map of a finite set is bijective). So a set
equipotent to a proper subset is infinite.
\end{solution}

\begin{solution}{pb:b2:structures:1}
\textbf{1.} A configuration assigns to each of the $16$ cells one
of the $16$ contents (tiles $1$--$15$ or the blank $b = 16$),
each exactly once: precisely a bijection $\intint1{16} \to
\intint1{16}$, an element of $\mathfrak{S}_{16}$; there are $16!
= 20\,922\,789\,888\,000$ of them. A legal move slides one tile
adjacent to the blank, so the number of moves is the number of
neighbours of the blank's cell: $2$ for the four corner cells,
$3$ for the eight edge cells, $4$ for the four interior cells.

\textbf{2.} After the slide, cell $p$ holds the former content of
$c$ and cell $c$ holds the blank; all other cells are untouched:
$\sigma'(p) = \sigma(c)$, $\sigma'(c) = \sigma(p) = 16$,
$\sigma' = \sigma$ elsewhere. That is exactly $\sigma' = \sigma
\circ (p\ c)$. Since $\varepsilon$ is a morphism and
$\varepsilon\bigl((p\ c)\bigr) = -1$:
$\varepsilon(\sigma') = -\varepsilon(\sigma)$.

\textbf{3.} Adjacent cells differ by one step in exactly one of
the two coordinates, so $i + j$ changes parity: $\chi$ takes
opposite values on adjacent cells. A move transfers the blank
from $p$ to the adjacent $c$, flipping $\chi(\text{blank cell})$.
Along a closed walk of the blank, $\chi$ is flipped once per move
and returns to its initial value: the number of moves is even.

\textbf{4.} By questions 2 and 3, one move flips both factors of
$I(\sigma) = \varepsilon(\sigma)\chi(\sigma^{-1}(16))$; their
product is unchanged. For the solved configuration:
$\varepsilon(\mathrm{id}) = +1$ and the blank is at cell $16$,
row $4$, column $4$: $\chi(16) = (-1)^{8} = +1$, so
$I(\mathrm{id}) = +1$.

\textbf{5.} $\sigma_L$ is the transposition $(14\ 15)$ of cells:
$\varepsilon(\sigma_L) = -1$; its blank is home, $\chi(16) = +1$:
$I(\sigma_L) = -1 \neq +1 = I(\mathrm{id})$. Since $I$ is
preserved by every move, no sequence of moves joins $\sigma_L$
and $\mathrm{id}$. The prize was structurally safe.

\textbf{6.} Fix a cell $p$ and two other cells $c \neq d$
distinct from $p$, and set $\tau_0 = (c\ d)$. On the set of
configurations with blank at $p$, the map $\sigma \mapsto \sigma
\circ \tau_0$ is an involution (it preserves $\sigma(p) = 16$
since $\tau_0$ fixes $p$) and flips $\varepsilon$, hence flips
$I$: it pairs the configurations with $I = +1$ bijectively with
those with $I = -1$. So each of the $16$ blank positions
contributes $15!/2$ configurations with $I = +1$, and
\[
\abs{\{I = +1\}} = 16 \cdot \frac{15!}{2} = \frac{16!}{2}.
\]

\textbf{7.} The move sliding the tile of $c$ into $p$ is undone
by sliding that same tile (now in $p$) back into $c$: composing
with $(p\ c)$ twice is the identity. Hence: reflexivity (empty
sequence), symmetry (reverse the sequence, undoing each move),
transitivity (concatenate): an equivalence relation. Every
$\sigma \in R$ has $I(\sigma) = I(\mathrm{id}) = +1$ by question
4, so $R \subseteq \{I = +1\}$; and $\sigma_L \notin R$ gives a
second class.

\textbf{8.} If $\sigma(16) = 16$, then $\sigma$ permutes the
cells $1, \dots, 15$; call $\rho$ this restriction. Appending a
fixed point changes neither the cycle type nor the signature
(decompose $\rho$ into transpositions; the same product works in
$\mathfrak{S}_{16}$), so $\varepsilon(\sigma) =
\varepsilon(\rho)$, and $\chi(16) = +1$ gives $I(\sigma) =
\varepsilon(\rho)$. Any configuration can be carried to a
blank-home one: the grid is connected, so walk the blank along a
path of adjacent cells to cell $16$ (each step is a legal move).
Now suppose every even $\rho \in \mathfrak{S}_{15}$ is realized
by a program. Given $\sigma$ with $I(\sigma) = +1$: walk the
blank home to reach $\widetilde\sigma$ (equivalent to $\sigma$),
with $I(\widetilde\sigma) = +1$, i.e.\ its restriction $\rho$ is
even; the program realizing $\rho$ carries $\widetilde\sigma$ to
$\widetilde\sigma \circ \rho^{-1} = \mathrm{id}$ (see question
9). By transitivity $\sigma \in R$, whence $\{I = +1\} \subseteq
R$ and equality.

\textbf{9.} \emph{Single move:} the content of $c$ ends in $p$
and the blank in $c$: the effect is $\pi = (p\ c)$, and indeed
$\sigma' = \sigma \circ (p\ c) = \sigma \circ \pi^{-1}$.
\emph{Induction:} if a sequence has effect $\pi_1$ and carries
$\sigma$ to $\sigma \circ \pi_1^{-1}$, following it with a move
of effect $\pi_2 = (p'\ c')$ yields $(\sigma \circ \pi_1^{-1})
\circ \pi_2^{-1} = \sigma \circ (\pi_2\pi_1)^{-1}$, and contents
move by $\pi_2 \circ \pi_1$ (first $\pi_1$, then $\pi_2$). So
effects compose, and a program run from $\sigma$ ends at $\sigma
\circ \pi^{-1}$. \emph{Subgroup:} the empty program has effect
$\mathrm{id}$; concatenation gives products; reversing a program
(question 7) gives inverses. A program's effect fixes cell $16$
(blank starts and ends home), so $H \leq \mathfrak{S}_{15}$.
\emph{Evenness:} a program of $k$ moves has $k$ even (question
3), and $\varepsilon(\sigma \circ \pi^{-1}) =
(-1)^k\varepsilon(\sigma)$ forces $\varepsilon(\pi) = +1$: $H
\subseteq A_{15}$.

\textbf{10.} Track the four slides from the blank at $16$: move
$16 \to 12$ sends the content of $12$ to $16$; move $12 \to 11$
sends the content of $11$ to $12$; move $11 \to 15$ sends the
content of $15$ to $11$; move $15 \to 16$ sends the content
parked in $16$ (originally in $12$) to $15$. Net: $11 \mapsto
12$, $12 \mapsto 15$, $15 \mapsto 11$, blank home: the effect is
$(11\ 12\ 15)$. The reverse tour undoes it: effect $(11\ 12\
15)^{-1} = (11\ 15\ 12)$. Both are effects of programs, hence in
$H$.

\textbf{11.} Adjacency of consecutive cells: within each listed
pair the cells differ by $1$ in the same row ($16{-}15$,
$15{-}14$, $14{-}13$; $1{-}2$, $2{-}3$, $3{-}4$; $8{-}7$,
$7{-}6$; $10{-}11$, $11{-}12$) or by $4$ within a column
($13{-}9$, $9{-}5$, $5{-}1$; $4{-}8$; $6{-}10$; $12{-}16$): a
closed walk through all $16$ cells, of length $16$. Effect: as
in question 10, writing the visited cells $c_0 = 16, c_1 = 15,
\dots, c_{15} = 12$: the content of $c_i$ moves to $c_{i-1}$ for
$i = 2, \dots, 15$, and the content of $c_1$, parked in $16$
after the first move, is carried to $c_{15}$ by the last move.
So the effect maps $15 \mapsto 12$, and $14 \mapsto 15$, $13
\mapsto 14$, $9 \mapsto 13$, $5 \mapsto 9$, $1 \mapsto 5$, $2
\mapsto 1$, $3 \mapsto 2$, $4 \mapsto 3$, $8 \mapsto 4$, $7
\mapsto 8$, $6 \mapsto 7$, $10 \mapsto 6$, $11 \mapsto 10$, $12
\mapsto 11$: exactly the $15$-cycle $\zeta$. Its cycle order
starts $x_0 = 15$, $x_1 = 12$, $x_2 = 11$, and $(x_0\ x_1\ x_2)
= (15\ 12\ 11)$ maps $15 \mapsto 12 \mapsto 11 \mapsto 15$ ---
which is precisely $(11\ 15\ 12)$, the reverse elementary tour.

\textbf{12.} Let $\gamma = (a\ b\ c)$ and $x \in \intint1n$. If
$x = g(a)$: $g\gamma g^{-1}(x) = g(\gamma(a)) = g(b)$; likewise
$g(b) \mapsto g(c)$ and $g(c) \mapsto g(a)$. If $x \notin
\{g(a), g(b), g(c)\}$, then $g^{-1}(x) \notin \{a,b,c\}$ is
fixed by $\gamma$, so $x$ is fixed. Hence $g\gamma g^{-1} =
(g(a)\ g(b)\ g(c))$. And for $g, h \in H$, $ghg^{-1} \in H$ by
the subgroup axioms.

\textbf{13.} $\zeta \in H$ (question 11) and $s_0 = (x_0\ x_1\
x_2) \in H$ (questions 10--11). Since $\zeta(x_i) = x_{i+1}$
(indices mod $15$), question 12 gives
\[
\zeta^{t}\,s_0\,\zeta^{-t}
= \bigl(\zeta^t(x_0)\ \zeta^t(x_1)\ \zeta^t(x_2)\bigr)
= (x_t\ x_{t+1}\ x_{t+2}) = s_t \in H
\qquad (t = 0, 1, \dots, 14).
\]

\textbf{14.} Up to inverting, assume $s = (a\ b\ c)$ and $t =
(b\ c\ d)$ (a $3$-cycle on $\{a,b,c\}$ is $(a\ b\ c)$ or its
inverse; likewise on $\{b,c,d\}$; replacing a generator by its
inverse leaves $\langle s, t\rangle$ unchanged). Then, applying
$t$ first,
\[
st \colon a \mapsto b,\quad b \mapsto a,\quad c \mapsto d,\quad
d \mapsto c, \qquad\text{i.e.}\quad st = (a\ b)(c\ d),
\]
a double transposition. The subgroup $G = \langle s, t\rangle$
consists of even permutations of the four letters, so $G \leq
A_4$ and $\abs G \mid 12$; it contains an element of order $3$
and one of order $2$, so $6 \mid \abs G$ (Lagrange,
\cref{thm:b2:structures:lagrange}, applied to the two cyclic
subgroups). If $A_4$ had a subgroup $K$ of order $6$, it would
have index $2$, and then $g^2 \in K$ for every $g \in A_4$: for
$g \in K$ this is clear; for $g \notin K$ the only cosets are
$K$ and $gK$, so the coset $g^2K$ is $K$ or $gK$, and $g^2K =
gK$ would force $g \in K$. So every square lies in $K$. But every
$3$-cycle $\gamma$ is a square, $\gamma = (\gamma^2)^2$, and
$A_4$ contains eight $3$-cycles: $8 > 6$, contradiction. Hence
$\abs G = 12$: $G = A_4$.

\textbf{15.} Extend $u \mapsto a$, $v \mapsto b$ to a bijection
$g_0$ of $X$ (send the remaining $k - 2$ letters bijectively
anywhere onto the complement of $\{a, b\}$). If $g_0$ is odd,
pick two distinct letters $s_1, t_1 \in X \setminus \{u, v\}$
(possible: $k \geq 4$) and replace $g_0$ by $g_0 \circ (s_1\
t_1)$, which is even and still maps $u \mapsto a$, $v \mapsto
b$. Extend by the identity off $X$: an even permutation $g \in
G$ (it is an even permutation of $X$). Then question 12:
\[
g\,(u\ v\ w)\,g^{-1} = (g(u)\ g(v)\ g(w)) = (a\ b\ w) \in G,
\]
using $g(w) = w$.

\textbf{16.} Every $3$-cycle of $X \cup \{w\}$ lies in $G$:
those supported in $X$ are even permutations of $X$; one with
support $\{a, b, w\}$ is $(a\ b\ w)$ or $(b\ a\ w)$, both
delivered by question 15. By \cref{exo:b2:structures:6}, the
$3$-cycles of the $(k+1)$-element set $X \cup \{w\}$ generate
its alternating group, so $G$ contains every even permutation of
$X \cup \{w\}$. \emph{Chaining:} let $G = \langle s_0, \dots,
s_{12}\rangle$. Lemma A applied to $s_0 = (x_0\ x_1\ x_2)$ and
$s_1 = (x_1\ x_2\ x_3)$ (supports share $\{x_1, x_2\}$) gives
all even permutations of $X_4 = \{x_0, x_1, x_2, x_3\}$. If $G$
contains all even permutations of $X_m = \{x_0, \dots,
x_{m-1}\}$ ($4 \leq m \leq 14$), then $s_{m-2} = (x_{m-2}\
x_{m-1}\ x_m)$ has $u = x_{m-2}, v = x_{m-1} \in X_m$ and new
letter $w = x_m$: Lemma B and the first part give all even
permutations of $X_{m+1}$. Induction up to $m = 14$: $G \supseteq
A_{15}$ (even permutations of all fifteen cells), and $G
\subseteq A_{15}$ since each $s_t$ is even: $\langle s_0, \dots,
s_{12}\rangle = A_{15}$.

\textbf{17.} Questions 13 and 16: $A_{15} = \langle s_0, \dots,
s_{12}\rangle \subseteq H$; question 9: $H \subseteq A_{15}$. So
$H = A_{15}$, of order $15!/2 = 653\,837\,184\,000$: every even
rearrangement of the fifteen tiles is the effect of a program.

\textbf{18.} Question 8 reduced $R = \{I = +1\}$ to realizing
every even $\rho \in \mathfrak{S}_{15}$ by a program: done by
question 17. With question 6, $\abs R = 16!/2 =
10\,461\,394\,944\,000$. \emph{Two classes:} let $t_0 = (14\
15)$ act on \emph{contents}: $\varphi(\sigma) = t_0 \circ
\sigma$. A legal move from $\sigma$ is a legal move from
$\varphi(\sigma)$ (the blank cell is unchanged:
$(t_0\sigma)^{-1}(16) = \sigma^{-1}(t_0(16)) = \sigma^{-1}(16)$,
and the moved cell is the same), and $\varphi(\sigma \circ \tau)
= \varphi(\sigma) \circ \tau$: $\varphi$ maps move sequences to
move sequences, bijectively (it is an involution). It flips $I$:
$\varepsilon(t_0\sigma) = -\varepsilon(\sigma)$, same blank
cell. Hence $\varphi$ maps the class $R = \{I = +1\}$ of
$\mathrm{id}$ bijectively onto the class of $\varphi(\mathrm{id})
= \sigma_L$, which is therefore all of $\{I = -1\}$: exactly two
classes. This is the Johnson--Story theorem.

\textbf{19.} Index the cells in reading order and let $k = 4(i -
1) + j$ be the blank's cell. Count the inversions of $\sigma$
(pairs of cells $x < y$ with $\sigma(x) > \sigma(y)$): pairs of
two tile cells contribute $N$; pairs involving the blank: cells
after the blank all hold tiles $< 16$, each inverted ($16 - k$
pairs), cells before it are never inverted. So
$\varepsilon(\sigma) = (-1)^{N + 16 - k} = (-1)^{N + k}$. Since
$k = 4(i-1) + j \equiv j \pmod 2$,
\[
I(\sigma) = (-1)^{N + j}\,(-1)^{i + j} = (-1)^{N + i}
= (-1)^{N + r + 1}
\]
using $i = 5 - r$. By question 18, $\sigma$ is solvable iff
$I(\sigma) = +1$ iff $N + r$ is odd. Check: solved, $N = 0$, $r
= 1$: odd, solvable; Loyd, $N = 1$, $r = 1$: even, unsolvable.

\textbf{20.} \emph{Action:} $e \cdot \sigma = \sigma \circ
\mathrm{id} = \sigma$ and $g \cdot (h \cdot \sigma) = \sigma
\circ h^{-1} \circ g^{-1} = \sigma \circ (gh)^{-1} = (gh) \cdot
\sigma$; and $\sigma \circ h^{-1}$ is again a blank-home
configuration ($h$ fixes cell $16$). \emph{Free:} $\sigma \circ
h^{-1} = \sigma$ gives $h^{-1} = \mathrm{id}$ (compose with
$\sigma^{-1}$). \emph{Orbits = program classes:} question 9
says the configurations reachable from $\sigma$ by programs are
exactly the $\sigma \circ \pi^{-1}$, $\pi \in H$: the orbit $H
\cdot \sigma$. \emph{Count:} freeness makes $h \mapsto h \cdot
\sigma$ injective, so every orbit has $\abs H = 15!/2$ elements;
the $15!$ blank-home configurations therefore split into
$15!\,/\,(15!/2) = 2$ orbits --- the blank-home shadow of the
two Johnson--Story classes.

\textbf{21.} The $3 \times 3$ grid is bipartite for the
checkerboard colouring: every step of a walk changes colour, so
every \emph{closed} walk has even length. A closed walk visiting
each of the $9$ cells exactly once would have length $9$, odd:
impossible. The grand-tour construction of Part III is therefore
unavailable on the eight puzzle.

\textbf{22.} \emph{Perimeter tour} $9 \to 8 \to 7 \to 4 \to 1
\to 2 \to 3 \to 6 \to 9$ (all steps adjacent; length $8$, even):
by the bookkeeping of question 11 with $c_1 = 8, c_2 = 7, c_3 =
4, c_4 = 1, c_5 = 2, c_6 = 3, c_7 = 6$, the effect is
\[
\zeta' = (8\ 6\ 3\ 2\ 1\ 4\ 7),
\]
a $7$-cycle fixing the center $5$ (content of $7$ moves to $8$,
of $4$ to $7$, of $1$ to $4$, of $2$ to $1$, of $3$ to $2$, of
$6$ to $3$, and of $8$ to $6$). \emph{Corner tour} $9 \to 6 \to
5 \to 8 \to 9$: effect $(6\ 8\ 5)$ (content of $5$ moves to
$6$, of $8$ to $5$, of $6$ --- parked at $9$ --- to $8$). Set
$y_t = \zeta'^{\,t}(8)$: $y_0 = 8, y_1 = 6, y_2 = 3, y_3 = 2,
y_4 = 1, y_5 = 4, y_6 = 7$. Conjugation (question 12):
\[
\zeta'^{\,t}\,(6\ 8\ 5)\,\zeta'^{-t}
= (y_{t+1}\ y_t\ 5) =: T_t \in H_{3\times3},
\]
since $\zeta'$ fixes $5$. The supports of $T_0 = (y_1\ y_0\ 5)$
and $T_1 = (y_2\ y_1\ 5)$ share exactly $\{y_1, 5\}$: Lemma A
gives all even permutations of $\{y_0, y_1, y_2, 5\}$. Then
$T_2 = (y_3\ y_2\ 5)$ adjoins $y_3$ by Lemma B (its letters
$y_2, 5$ lie in the current set, $k = 4$), and $T_3, T_4, T_5$
adjoin $y_4, y_5, y_6$ in turn: all even permutations of the
eight non-home cells lie in the program group, which also
consists of even permutations (the argument of question 9 is
board-independent). So $H_{3\times3} = A_8$, and the reasoning
of questions 6, 8, 18 --- also board-independent --- shows the
reachable configurations are exactly those with $I = +1$: half
of $9!$, i.e.\ $181\,440$.

\textbf{23.} Label the cells $0, \dots, n-1$ around the cycle.
A move swaps the blank with one of its two neighbours. Read the
tiles in cyclic order starting just after the blank: a word $w$
listing the $n - 1$ tiles. Moving the blank one step forward
replaces $(p, w)$ by $(p + 1, \rho w)$, where $p$ is the blank
cell and $\rho$ cyclically rotates the word by one; the backward
move is the inverse. The \emph{cyclic} order of the tiles (the
word up to rotation) is thus invariant. The reachable class of
$(p, w)$ is the orbit of the map $g \colon (p, w) \mapsto (p+1,
\rho w)$, an element of order $\operatorname{lcm}(n, n-1) =
n(n-1)$ in the product of the two cyclic groups (translations of
$\Z/n\Z$ and rotations of the $n-1$ word positions), the lcm
being $n(n-1)$ because $\gcd(n, n-1) = 1$: each class has
exactly $n(n-1)$ configurations, all with the same necklace.
Classes: $n!\,/\,\bigl(n(n-1)\bigr) = (n-2)!$. For $n \geq 5$,
$(n-2)! > 2$: the parity invariant (two classes at best) is
blind to almost all of the obstruction; the wealth of the $4
\times 4$ board --- where parity is the \emph{only} obstruction
--- is a genuinely geometric fact, not a formal one.

\textbf{24.} Both trays have the tiles in fully reversed order,
so $N = \binom{15}{2} = 105$ in both cases (every pair of tiles
is inverted). \emph{Blank home:} $r = 1$, $N + r = 106$ even:
unsolvable. \emph{Blank in cell $1$:} the blank is in the top
row, $r = 4$, $N + r = 109$ odd: solvable. Two trays that differ
only by where the hole sits fall on opposite sides of the wall.

\textbf{25.} \emph{Morphism property:} it converts ``one move =
one transposition'' into ``one move = one sign flip'' (questions
2, 4), making $I$ computable move by move. \emph{Lagrange:} it
forced $6 \mid \abs{\langle s, t\rangle}$ in Lemma A and sized
the cosets in the order-$6$ exclusion (question 14).
\emph{Generation by $3$-cycles:} it converted ``$H$ contains
enough $3$-cycles'' into ``$H$ contains all of $A_{15}$''
(question 16). \emph{Conjugation:} it manufactured the fifteen
consecutive $3$-cycles from a single $2 \times 2$ tour
transported by the grand tour (questions 12--13), and the
$3$-cycles $(a\ b\ w)$ in Lemma B. \emph{Meta-principle:} an
invariant proves impossibility, an explicit construction proves
possibility, and a problem is fully solved exactly when the two
bounds meet --- here, at one half.
\end{solution}
