\section*{Chapter \ref{ch:b2:reduction} --- Reduction of Endomorphisms}

\begin{solution}{exo:b2:reduction:1}
$A$: $\chi_A = X^2 - 2X - 3 = (X - 3)(X + 1)$. Eigenvectors: for
$3$: $(1,1)$; for $-1$: $(1,-1)$. So $P = \begin{pmatrix} 1 & 1\\ 1
& -1\end{pmatrix}$ gives $P^{-1}AP = \operatorname{diag}(3, -1)$.

$B = J - I$ where $J$ is the all-ones matrix. $J$ has rank $1$ with
$Jv = 3v$ for $v = (1,1,1)$ and $Jw = 0$ on the plane $x + y + z =
0$: spectrum of $B$ is $\{2, -1\}$ with eigenspaces
$\operatorname{Vect}(1,1,1)$ (dimension $1$) and $\{x + y + z = 0\}$
(dimension $2$, basis $(1,-1,0), (1,0,-1)$). $P$ with these three
columns gives $P^{-1}BP = \operatorname{diag}(2, -1, -1)$.
\end{solution}

\begin{solution}{exo:b2:reduction:2}
\emph{Eigenspaces:} $\chi_C = (X-1)^2$, single eigenvalue $1$;
$\ker(C - I) = \ker\begin{pmatrix} 0&1\\ 0&0\end{pmatrix}$ is the
line $\operatorname{Vect}(e_1)$: dimension $1 < 2 = m_1$, so not
diagonalizable (\cref{thm:b2:reduction:diagcrit}).

\emph{Minimal polynomial:} $\mu_C$ divides $(X-1)^2$ and $C \neq I$,
so $\mu_C = (X-1)^2$: a double root, so not diagonalizable
(\cref{cor:b2:reduction:minpolycrit}).
\end{solution}

\begin{solution}{exo:b2:reduction:3}
$X^2 - 5X + 6 = (X-2)(X-3)$: split with simple roots, so $u$ is
diagonalizable (\cref{cor:b2:reduction:minpolycrit}), with
$\operatorname{Sp}(u) \subseteq \{2, 3\}$. Possible spectra:
$\{2\}$ ($u = 2\,\mathrm{id}$), $\{3\}$ ($u = 3\,\mathrm{id}$), or
$\{2, 3\}$.

Powers: seek $u^k = a_k\,\mathrm{id} + b_k\,u$. On the eigenspaces,
this reads $2^k = a_k + 2b_k$ and $3^k = a_k + 3b_k$: solving, $b_k
= 3^k - 2^k$, $a_k = 3\cdot2^k - 2\cdot 3^k$:
\[
u^k = (3\cdot 2^k - 2\cdot 3^k)\,\mathrm{id} + (3^k - 2^k)\, u .
\]
(Valid for all three spectra: the identities hold eigenvalue-wise.)
\end{solution}

\begin{solution}{exo:b2:reduction:4}
$u$ diagonalizable: $P = \prod_{\lambda}(X - \lambda)$ over the
spectrum annihilates $u$, splits, simple roots. Then $P(u|_F) =
P(u)|_F = 0$: the restriction is annihilated by a split polynomial
with simple roots, hence diagonalizable
(\cref{cor:b2:reduction:minpolycrit}).
\end{solution}

\begin{solution}{exo:b2:reduction:5}
$\chi_A = X^2 - X - 1$, roots $\varphi$ and $\psi$ (distinct):
diagonalizable, with eigenvectors $(\varphi, 1)$ and $(\psi, 1)$.
The recurrence gives $\begin{pmatrix} F_{n+1}\\ F_n \end{pmatrix} =
A^n \begin{pmatrix}1\\ 0\end{pmatrix}$. Decompose $(1, 0)$ on the
eigenvectors: $(1,0) = \frac{1}{\varphi - \psi}\bigl((\varphi, 1) -
(\psi, 1)\bigr)$ with $\varphi - \psi = \sqrt5$. Applying $A^n$
multiplies each eigencomponent by its eigenvalue's $n$-th power;
reading the second coordinate:
\[
F_n = \frac{\varphi^n - \psi^n}{\sqrt 5} .
\]
(Check: $n = 1$ gives $\frac{\varphi - \psi}{\sqrt5} = 1$.)
\end{solution}

\begin{solution}{exo:b2:reduction:6}
Let $P = \prod_i (X - \mu_i)$ annihilate $u^2$, split with simple
roots $\mu_i$ (the spectrum of $u^2$). Since $u$ is invertible, $0$
is not an eigenvalue of $u^2$ ($\det u^2 = (\det u)^2 \neq 0$), so
all $\mu_i \neq 0$. Then
\[
Q(X) = \prod_i (X^2 - \mu_i) = \prod_i (X - \sqrt{\mu_i})(X +
\sqrt{\mu_i})
\]
annihilates $u$: $\;Q(u) = \prod_i (u^2 - \mu_i\,\mathrm{id}) =
P(u^2) = 0$. Its roots $\pm
\sqrt{\mu_i}$ (complex square roots) are pairwise distinct because
the $\mu_i$ are distinct and nonzero ($\sqrt{\mu_i} = -\sqrt{\mu_j}$
would give $\mu_i = \mu_j$). Split + simple roots: $u$ is
diagonalizable.

Counterexample without invertibility: $u = \begin{pmatrix} 0 & 1\\
0 & 0\end{pmatrix}$: $u^2 = 0$ is diagonalizable, $u$ is not.
\end{solution}

\begin{solution}{exo:b2:reduction:7}
$A = 2I + N$ with $N$ the shift ($N e_2 = e_1$, $Ne_3 = e_2$),
$N^3 = 0$, $N^2 = E_{13}$: this \emph{is} the Dunford decomposition
($2I$ diagonal, $N$ nilpotent, they commute; uniqueness makes it
the one). Binomial with commuting terms:
\[
A^k = 2^k I + k 2^{k-1} N + \binom k2 2^{k-2} N^2
= \begin{pmatrix}
2^k & k2^{k-1} & \binom k2 2^{k-2}\\
0 & 2^k & k2^{k-1}\\
0 & 0 & 2^k
\end{pmatrix},
\]
\[
\eu^{tA} = \eu^{2t}\Bigl(I + tN + \frac{t^2}{2}N^2\Bigr)
= \eu^{2t}\begin{pmatrix}
1 & t & t^2/2\\
0 & 1 & t\\
0 & 0 & 1
\end{pmatrix}.
\]
\end{solution}

\begin{solution}{exo:b2:reduction:8}
$X^k - 1$ annihilates $A$ and splits over $\C$ with the $k$ distinct
roots $\eu^{2\iu\pi j/k}$: $A$ is diagonalizable
(\cref{cor:b2:reduction:minpolycrit}) and its eigenvalues, roots of
$X^k - 1$, are $k$-th roots of unity.

If moreover $A$ is similar to a triangular matrix with unit
diagonal: all eigenvalues equal $1$, and $A$, diagonalizable with
sole eigenvalue $1$, is $P\,I\,P^{-1} = I$.
\end{solution}

\begin{solution}{exo:b2:reduction:9}
Write $E = \bigoplus_\lambda E_\lambda(u)$
(\cref{thm:b2:reduction:diagcrit}). Each $E_\lambda(u)$ is
$v$-stable: for $x \in E_\lambda$, $u(v(x)) = v(u(x)) = \lambda
v(x)$. The restriction of $v$ to $E_\lambda(u)$ is diagonalizable
(\cref{exo:b2:reduction:4}): choose a basis of $E_\lambda(u)$ made
of $v$-eigenvectors. Concatenating these bases over all $\lambda$
gives a basis of $E$ whose vectors are eigenvectors of \emph{both}
$u$ (by membership in $E_\lambda(u)$) and $v$ (by construction).
\end{solution}

\begin{solution}{exo:b2:reduction:10}
($\Rightarrow$) Let $u$ be diagonalizable and $F$ stable. Then
$u|_F$ is diagonalizable (\cref{exo:b2:reduction:4}): $F$ has a
basis of eigenvectors, which extends, inside each global eigenspace
$E_\lambda$, to a basis of $E_\lambda$ (incomplete basis theorem
within $E_\lambda$, starting from the part of $F$'s basis lying
there --- note $F = \bigoplus_\lambda (F \cap E_\lambda)$ since
$u|_F$ is diagonalizable). The added vectors span a stable
supplement (each lies in some $E_\lambda$, so their span is
$u$-stable).

($\Leftarrow$) Let $F = \sum_\lambda E_\lambda(u)$ (a stable
subspace) and $G$ a stable supplement. If $G \neq \{0\}$: $\chi_{u|_G}$
splits over $\C$, so $u|_G$ has an eigenvector $x \in G$
(\cref{thm:b2:reduction:trigonalization} or directly the existence
of a root); but every eigenvector of $u$ lies in $F$, so $x \in F
\cap G = \{0\}$: contradiction. Hence $G = \{0\}$ and $E = F$: the
eigenspaces fill $E$, i.e.\ $u$ is diagonalizable.
\end{solution}

\begin{solution}{exo:b2:reduction:11}
Trigonalize: $A = PTP^{-1}$, $T = \begin{pmatrix} \lambda & c\\ 0 &
\mu\end{pmatrix}$, $\abs\lambda, \abs\mu < 1$. Then $A^k =
PT^kP^{-1}$, and it suffices that $T^k \to 0$.

\emph{Distinct eigenvalues:} induction gives
\[
T^k = \begin{pmatrix}
\lambda^k & c\,\dfrac{\lambda^k - \mu^k}{\lambda - \mu}\\[4pt]
0 & \mu^k
\end{pmatrix},
\]
and each entry tends to $0$ ($\abs{\lambda}^k, \abs\mu^k \to 0$).

\emph{Equal eigenvalues ($\mu = \lambda$):} $T = \lambda I + cE_{12}$
and $T^k = \lambda^k I + k\lambda^{k-1}cE_{12}$; the entry
$k\lambda^{k-1} \to 0$ since $\abs\lambda < 1$ (geometric beats
polynomial). In both cases $T^k \to 0$ entrywise, hence $A^k =
PT^kP^{-1} \to 0$ (matrix multiplication by fixed $P, P^{-1}$ is
continuous in the entries --- each entry of the product is a fixed
linear combination).
\end{solution}

\begin{solution}{exo:b2:reduction:12}
$\ker u$ has dimension $n - 1$ (rank--nullity), so $0$ is an
eigenvalue of geometric multiplicity $n - 1$, and $\chi_u$ is
divisible by $X^{n-1}$
(\cref{def:b2:reduction:charpoly}: geometric $\leq$ algebraic).
Write $\chi_u = X^{n-1}(X - \alpha)$; the coefficient of
$X^{n-1}$ being $-\operatorname{tr} u$, we get $\alpha =
\operatorname{tr} u$: $\chi_u = X^{n-1}(X - \operatorname{tr}
u)$.

If $\operatorname{tr} u \neq 0$: the eigenvalue
$\operatorname{tr} u$ is a root of $\chi_u$, so it carries an
eigenvector; the eigenspaces for $0$ and $\operatorname{tr} u$
have dimensions $n - 1$ and $\geq 1$, summing to $\geq n$: they
fill $E$, and $u$ is diagonalizable
(\cref{thm:b2:reduction:diagcrit}). If $\operatorname{tr} u = 0$:
by \cref{exo:b2:linalg:5}, $u^2 = (\operatorname{tr} u)u = 0$
with $u \neq 0$: $u$ is a nonzero nilpotent, and a diagonalizable
nilpotent is zero (\cref{prop:b2:reduction:nilpotent}): not
diagonalizable.
\end{solution}

\begin{solution}{pb:b2:reduction:1}
\textbf{1.} The first $k - 1$ coordinates of $Cv_n$ are $u_{n+1},
\dots, u_{n+k-1}$ (the superdiagonal shifts), and the last is
$a_0 u_n + \dots + a_{k-1}u_{n+k-1}$. So $v_{n+1} = Cv_n$ holds
for all $n$ iff the last coordinates match for all $n$, i.e.\ iff
$(\mathcal R)$ holds. Iterating, $v_n = C^nv_0$.

\textbf{2.} Expand $D_k(X) = \det(XI_k - C)$ along the first
column: the two nonzero entries are $X$ (position $(1,1)$) and
$-a_0$ (position $(k,1)$). The first minor is $D_{k-1}$-shaped
for the coefficients $a_1, \dots, a_{k-1}$; the second minor is
upper triangular with diagonal $-1$: determinant $(-1)^{k-1}$,
with sign $(-1)^{k+1}$ from the position. Induction on $k$
(base $k = 1$: $X - a_0$) gives
\[
D_k(X) = X\bigl(X^{k-1} - a_{k-1}X^{k-2} - \dots - a_1\bigr) -
a_0 = P(X).
\]
For $\mu_C$: since $Q(C^{\mathsf T}) = Q(C)^{\mathsf T}$ for any
polynomial, $C$ and $C^{\mathsf T}$ have the same annihilators,
hence the same minimal polynomial. For $C^{\mathsf T}$: the
columns read $C^{\mathsf T}e_1 = e_2$, \dots, $C^{\mathsf
T}e_{k-1} = e_k$, so $(e_1, C^{\mathsf T}e_1, \dots, (C^{\mathsf
T})^{k-1}e_1)$ is the canonical basis: free. A polynomial $Q
\neq 0$ of degree $< k$ then has $Q(C^{\mathsf T})e_1 \neq 0$
(it is a nontrivial combination of basis vectors): $\deg\mu \geq
k$. As $\mu \mid \chi = P$ with $\deg P = k$: $\mu_C = P$.

\textbf{3.} For $v = (1, \lambda, \dots, \lambda^{k-1})^{\mathsf
T}$: rows $1$ to $k-1$ of $Cv$ give $\lambda, \lambda^2, \dots,
\lambda^{k-1}$, i.e.\ $\lambda$ times the first $k - 1$ entries
of $v$; the last row gives $\sum_m a_m\lambda^m = \lambda^k -
P(\lambda) = \lambda^k = \lambda\cdot\lambda^{k-1}$. So $Cv =
\lambda v$. Conversely, the equations $(Cx)_i = \lambda x_i$ for
$i < k$ read $x_{i+1} = \lambda x_i$: any eigenvector is
proportional to $v$ --- every eigenspace has dimension exactly
$1$. Diagonalizable iff the eigenspace dimensions sum to $k$
(\cref{thm:b2:reduction:diagcrit}) iff there are $k$ distinct
eigenvalues iff $P$ has $k$ distinct roots (the eigenvalues are
the roots of $\chi_C = P$).

\textbf{4.} Each $(\lambda_i^n)_n$ solves $(\mathcal R)$:
$\lambda_i^{n+k} = \lambda_i^n\,\lambda_i^k =
\lambda_i^n\sum_m a_m\lambda_i^m$. Freeness: a vanishing
combination $\sum_i c_i\lambda_i^n = 0$ for $n = 0, \dots, k-1$
is a Vandermonde system (\cref{exo:b2:linalg:11}) in the $c_i$:
all $c_i = 0$. The solution space has dimension $k$ (question 6,
whose proof is elementary and independent): $k$ free solutions
form a basis, and coordinates are unique.

\textbf{5.} $P = X^2 - X - 6 = (X - 3)(X + 2)$: general solution
$u_n = A\,3^n + B(-2)^n$. Initial conditions: $A + B = 1$, $3A -
2B = 8$: $A = 2$, $B = -1$:
\[
u_n = 2\cdot 3^n - (-2)^n .
\]
(Check: $u_2 = u_1 + 6u_0 = 14$ and $2\cdot9 - 4 = 14$.)

\textbf{6.} $P(S)\bigl((u_n)\bigr)$ is the sequence $n \mapsto
u_{n+k} - a_{k-1}u_{n+k-1} - \dots - a_0u_n$: it vanishes iff
$(\mathcal R)$ holds, so the solution set is $\ker P(S)$, a
subspace. The map $\ker P(S) \to \C^k$, $u \mapsto (u_0, \dots,
u_{k-1})$, is linear, injective (the recurrence determines
$u_{k}, u_{k+1}, \dots$ from the first $k$ values, by induction)
and surjective (define $u_n$ recursively from any initial data):
dimension $k$.

\textbf{7.} The proof of \cref{thm:b2:reduction:kernels} uses
only: the Bézout identity in $\C[X]$, and the fact that
polynomials in a fixed endomorphism commute. Neither mentions
the dimension of the ambient space: the lemma holds verbatim for
$S \in \mathcal{L}(\mathcal{S})$. Hence
\[
\ker P(S) = \bigoplus_{i=1}^{r}
\ker\,(S - \lambda_i\,\mathrm{id})^{m_i}.
\]

\textbf{8.} For $Q \in \C[X]$: $(S -
\lambda)\bigl(Q(n)\lambda^n\bigr)_n$ has $n$-th term
$Q(n{+}1)\lambda^{n+1} - \lambda Q(n)\lambda^n =
\lambda^{n+1}(\Delta Q)(n)$, with $\Delta Q = Q(X{+}1) - Q(X)$
of degree $\deg Q - 1$ (the leading terms cancel). Iterating,
$(S - \lambda)^m\bigl(Q(n)\lambda^n\bigr) =
\bigl(\lambda^{n+m}(\Delta^m Q)(n)\bigr)_n$, and $\Delta^m Q = 0$
when $\deg Q \leq m - 1$: the right-hand set is contained in the
kernel. It is a subspace of dimension $m$: the sequences
$(n^j\lambda^n)_n$, $0 \leq j < m$, are free, since
$\sum_j c_j n^j\lambda^n = 0$ for all $n$ forces (dividing by
$\lambda^n \neq 0$) the polynomial $\sum_j c_jX^j$ to vanish at
every $n \in \N$, hence to be zero. Conversely $\dim\ker(S -
\lambda)^m \leq m$: expanding $(S - \lambda)^m = \sum_j
\binom mj(-\lambda)^{m-j}S^j$, the equation $(S - \lambda)^m u =
0$ is a linear recurrence of order $m$ (leading coefficient
$1$), so $u$ is determined by $u_0, \dots, u_{m-1}$ as in
question 6. Equality of dimensions concludes.

\textbf{9.} Combine questions 7 and 8: every solution decomposes
uniquely as a sum of elements of the $\ker(S -
\lambda_i)^{m_i}$, i.e.\ $u_n = \sum_i Q_i(n)\lambda_i^n$ with
$\deg Q_i \leq m_i - 1$; the $Q_i$ are unique because the
decomposition is direct and, within each summand, the
coefficients of $Q_i$ are coordinates in the basis
$(n^j\lambda_i^n)_j$ (question 8). Sanity check on dimensions:
$\sum_i m_i = k$.

\textbf{10.} $P = X^2 - 4X + 4 = (X - 2)^2$: solutions $(a +
bn)2^n$. Initial data: $a = 1$, $2(a + b) = 0$, so $b = -1$:
\[
u_n = (1 - n)\,2^n .
\]
Check: $u_2 = 4u_1 - 4u_0 = -4$, and $(1 - 2)\cdot4 = -4$.

\textbf{11.} Write $u_n = \lambda_1^n\bigl(c_1 + \sum_{i\geq2}
c_i(\lambda_i/\lambda_1)^n\bigr)$; each ratio has modulus $< 1$,
so the bracket tends to $c_1 \neq 0$: $u_n \sim c_1\lambda_1^n$.
In particular $u_n \neq 0$ for large $n$, and
\[
\frac{u_{n+1}}{u_n} =
\lambda_1\,\frac{c_1 + o(1)}{c_1 + o(1)} \longrightarrow
\lambda_1 .
\]

\textbf{12.} Compute:
\[
q(a_{n+1}, b_{n+1}) = (a_n + 2b_n)^2 - 2(a_n + b_n)^2
= -a_n^2 + 2b_n^2 = -q(a_n, b_n).
\]
With $q(a_0, b_0) = 1 - 2 = -1$: $a_n^2 - 2b_n^2 = (-1)^{n+1}$.
Structurally: $q(a, b) = (a - \sqrt2\,b)(a + \sqrt2\,b)$ and
the linear map $M$ multiplies the factor $a + \sqrt2 b$ by $1 +
\sqrt2$ and the factor $a - \sqrt2 b$ by $1 - \sqrt2$ (compute:
$a_{n+1} + \sqrt2 b_{n+1} = (1 + \sqrt2)(a_n + \sqrt2 b_n)$);
the product is multiplied by $(1 + \sqrt2)(1 - \sqrt2) = -1 =
\det M$ each step.

\textbf{13.} Since $a_n^2 - 2b_n^2 = (a_n - \sqrt2 b_n)(a_n +
\sqrt2 b_n) = (-1)^{n+1}$,
\[
\Bigl|\frac{a_n}{b_n} - \sqrt2\Bigr|
= \frac{\abs{a_n^2 - 2b_n^2}}{b_n(a_n + \sqrt2 b_n)}
= \frac{1}{b_n(a_n + \sqrt2 b_n)} \leq \frac1{2b_n^2},
\]
using $a_n \geq b_n \geq 1$ (induction: both increase) so $a_n +
\sqrt2 b_n \geq (1 + \sqrt2)b_n \geq 2b_n$. Eigenvalues of $M$:
$\chi_M = X^2 - 2X - 1$, roots $1 \pm \sqrt2$; since $(a_0, b_0)$
has a nonzero component on the dominant eigenvector (all entries
positive), $b_n \sim c(1 + \sqrt2)^n$ with $c > 0$ (question
11). Hence the error is $\asymp (1 + \sqrt2)^{-2n} = (3 +
2\sqrt2)^{-n}$: geometric decay with ratio $1/(3 + 2\sqrt2) = 3
- 2\sqrt2 \approx 0.172$.

\textbf{14.} (a) From question 9: $\abs{u_n} \leq \sum_i
\abs{Q_i(n)}\abs{\lambda_i}^n \leq \bigl(\sum_i
\abs{Q_i(n)}\bigr)\rho^n$, and each $\abs{Q_i(n)} \leq C_i
n^{m_i - 1} \leq C_i n^{m-1}$ for $n \geq 1$: sum the constants.
(b) Let $\rho' = \max_{i \geq 2}\abs{\lambda_i} <
\abs{\lambda_1}$ and $d = \deg Q_1$, leading coefficient $c
\neq 0$. Then $u_n = Q_1(n)\lambda_1^n + R_n$ with $\abs{R_n}
\leq Cn^{m-1}\rho'^n$, and
\[
\frac{R_n}{Q_1(n)\lambda_1^n} = O\Bigl(n^{m-1-d}
\bigl(\rho'/\abs{\lambda_1}\bigr)^n\Bigr) \longrightarrow 0
\]
(geometric beats polynomial). So
\[
u_n \sim Q_1(n)\,\lambda_1^n \sim c\,n^d\lambda_1^n,
\qquad
\frac{u_{n+1}}{u_n} \longrightarrow \lambda_1
\quad\text{(since } Q_1(n{+}1)/Q_1(n) \to 1\text{)}.
\]
Check on question 10: for $u_n = (1-n)2^n$ the ratio is
\[
\frac{(-n)2^{n+1}}{(1-n)2^n} = 2\,\frac{-n}{1-n}
\longrightarrow 2 = \lambda_1 .
\]

\textbf{15.} Induction on $n$. For $n = 1$, $A_{ij}$ counts
walks of length $1$. Step: a walk of length $n + 1$ from $i$ to
$j$ is a walk of length $n$ from $i$ to some vertex $\ell$
followed by an edge $\ell j$:
\[
\#\{\text{walks}\} = \sum_{\ell} (A^n)_{i\ell}A_{\ell j} =
(A^{n+1})_{ij}.
\]

\textbf{16.} $J = 3\Pi$ where $\Pi = J/3$ is the projection onto
$\operatorname{Vect}(1,1,1)$ along the plane $x + y + z = 0$
($\Pi^2 = \Pi$ since $J^2 = 3J$). Then $A = J - I = 2\Pi - (I -
\Pi)$, and since $\Pi$ and $I - \Pi$ are complementary
projections,
\[
A^n = 2^n\,\Pi + (-1)^n (I - \Pi),
\qquad\text{i.e.}\qquad
(A^n)_{ij} = \frac{2^n}3 + (-1)^n\Bigl(\delta_{ij} -
\frac13\Bigr),
\]
which gives the two displayed formulas. At $n = 2$: diagonal
$(4 + 2)/3 = 2$ (walks $i \to \ell \to i$ for the two neighbours
$\ell$); off-diagonal $(4 - 1)/3 = 1$ (the single walk $i \to
\ell \to j$ through the third vertex).

\textbf{17.} Let $w_n^{(0)}, w_n^{(1)}$ count admissible words of
length $n$ ending in $0$, resp.\ $1$. Appending a letter: a $0$
may follow anything, a $1$ only a $0$:
\[
\begin{pmatrix} w_{n+1}^{(0)}\\ w_{n+1}^{(1)}\end{pmatrix}
= \begin{pmatrix} 1 & 1\\ 1 & 0\end{pmatrix}
\begin{pmatrix} w_n^{(0)}\\ w_n^{(1)}\end{pmatrix}.
\]
Summing, $w_{n+2} = w_{n+1} + w_n$ (or: condition on the first
letter). With $w_1 = 2$, $w_2 = 3$: $w_n = F_{n+2}$ by induction
($F_3 = 2$, $F_4 = 3$, same recurrence). Growth: the roots of
$X^2 - X - 1$ are $\varphi > \abs\psi$
(\cref{exo:b2:reduction:5}), and the $\varphi$-component is
nonzero (the $w_n$ are positive and $\psi^n \to 0$), so question
11 gives $w_{n+1}/w_n \to \varphi = \frac{1 + \sqrt5}2$.

\textbf{18.} $A = \begin{psmallmatrix} 0&1&0\\ 1&0&1\\
0&1&0\end{psmallmatrix}$. Check:
\[
A(1, \pm\sqrt2, 1)^{\mathsf T}
= (\pm\sqrt2, 2, \pm\sqrt2)^{\mathsf T}
= \pm\sqrt2\,(1, \pm\sqrt2, 1)^{\mathsf T},
\qquad
A(1, 0, -1)^{\mathsf T} = 0 :
\]
eigenvalues $\sqrt2, -\sqrt2, 0$ ($= 2\cos\frac\pi4,
2\cos\frac{3\pi}4, 2\cos\frac\pi2$). Decompose $e_1$ on the
eigenbasis and read the third coordinate, or use symmetry: with
$v_\pm = (1, \pm\sqrt2, 1)$, $v_0 = (1, 0, -1)$, one checks $e_1
= \frac14 v_+ + \frac14 v_- + \frac12 v_0$, so for $n \geq 1$
\[
(A^n)_{13} = \Bigl(\tfrac14(\sqrt2)^n v_+ +
\tfrac14(-\sqrt2)^n v_- + 0\Bigr)_{\!3}
= \frac{(\sqrt2)^n + (-\sqrt2)^n}{4},
\]
zero for odd $n$ (bipartite graph: ends are at even distance),
and $2\cdot 2^{n/2}/4 = 2^{n/2 - 1}$ for even $n$. At $n = 4$:
$2^{1} = 2$, matching the two walks $1\,2\,1\,2\,3$ and
$1\,2\,3\,2\,3$.

\textbf{19.} Closed walks of length $n$ from $i$ are
$(A^n)_{ii}$; summing over $i$ gives $\operatorname{tr}(A^n)$.
Trigonalizing $A$ (over $\C$), $A^n$ is triangular with diagonal
$\lambda_i^n$: $\operatorname{tr}(A^n) = \sum_i\lambda_i^n$.
Triangle: $\operatorname{tr}(A^n) = 3\,\frac{2^n + 2(-1)^n}3 =
2^n + 2(-1)^n = 2^n + (-1)^n + (-1)^n$: the spectrum $\{2, -1,
-1\}$, consistent with question 16.

\textbf{20.} The columns of $W^{\mathsf T}$: $W^{\mathsf T}e_i =
e_{i-1}$ for $i \geq 1$ and $W^{\mathsf T}e_0 = e_{k-1}$;
relabelling in the order $e_0, e_1, \dots$ this is exactly the
companion matrix of $X^k - 1$ ($a_0 = 1$, other $a_m = 0$).
Question 2: $\chi_{W} = \chi_{W^{\mathsf T}} = X^k - 1 =
\mu_{W}$. The roots $\omega^j$ ($j = 0, \dots, k-1$) are the $k$
distinct $k$-th roots of unity: $W$ is diagonalizable (question
3, or \cref{exo:b2:reduction:8}: $W^k = I$). Eigenvectors: $Wf_j
= \sum_m \omega^{-jm}e_{m+1} = \sum_{m'}\omega^{-j(m'-1)}e_{m'}
= \omega^j f_j$.

\textbf{21.} Circulants are polynomials in $W$, and polynomials
in a fixed matrix commute with one another. Each $f_j$ is an
eigenvector of every power: $W^m f_j = \omega^{jm}f_j$, so
\[
Cf_j = \sum_m c_m\omega^{jm} f_j = \widehat c(\omega^j)\,f_j :
\]
the basis $(f_0, \dots, f_{k-1})$ (free: Vandermonde in the
distinct $\omega^{-j}$, \cref{exo:b2:linalg:11}) diagonalizes
every circulant at once, with the stated eigenvalues.

\textbf{22.} The determinant is the product of the eigenvalues
(diagonalize): $\det C = \prod_{j}\widehat c(\omega^j)$. For $k
= 3$, $c_0 = a$, $c_1 = b$, $c_2 = c$ and $\omega = j =
\eu^{2\iu\pi/3}$:
\[
\det C = (a + b + c)(a + bj + cj^2)(a + bj^2 + cj^4),
\]
and $j^4 = j$: exactly the factorization of
\cref{exo:b2:linalg:8}.

\textbf{23.} $M = \frac12(W + W^{-1})$ is a circulant ($W^{-1} =
W^{k-1}$), with eigenvalues $\frac12(\omega^j + \omega^{-j}) =
\cos\frac{2\pi j}k$ on the same basis $f_j$. Coordinates: write
$x^{(0)} = \sum_j \alpha_j f_j$. The coordinates of $f_j$ sum to
$\sum_m \omega^{-jm}$, which is $k$ for $j = 0$ and $0$
otherwise (geometric sum with ratio $\omega^{-j} \neq 1$).
Summing the coordinates of $x^{(0)}$: $\sum_m x^{(0)}_m =
\alpha_0\,k$, so $\alpha_0 = \frac1k\sum_m x^{(0)}_m$, the mean.

\textbf{24.} $x^{(n)} = M^nx^{(0)} = \sum_j
\alpha_j\cos^n\bigl(\tfrac{2\pi j}k\bigr)f_j$. For odd $k$,
$\abs{\cos(2\pi j/k)} < 1$ for every $j \neq 0$ (the angle is
never $0$ or $\pi$), so all terms but $j = 0$ tend to $0$:
$x^{(n)} \to \alpha_0 f_0$, the constant vector equal to the
mean --- averaging on an odd ring equalizes. For $k = 4$ the
eigenvalues are $1, 0, -1, 0$: the $j = 2$ term
$\alpha_2(-1)^nf_2$ with $f_2 = (1, -1, 1, -1)^{\mathsf T}$
oscillates forever. The obstruction is the alternating mean:
multiplying the coordinates of $x^{(0)}$ by $(-1)^m$ and
summing, the same geometric-sum computation gives $\sum_m
(-1)^mx^{(0)}_m = 4\alpha_2$: the process converges iff
$x^{(0)}_0 - x^{(0)}_1 + x^{(0)}_2 - x^{(0)}_3 = 0$, and then
converges to the mean.

\textbf{25.} The companion matrix converts a scalar recurrence
of order $k$ into a first-order vector recurrence, so that
closed formulas become statements about $C^n$ --- reduction's
home ground (questions 1--5). The kernel decomposition lemma is
pure polynomial algebra (Bézout plus commutation), so it splits
$\ker P(S)$ even though $\mathcal{S}$ is infinite-dimensional
(questions 7--9). Dominant eigenvalues govern growth because
every other contribution is geometrically negligible after
normalization --- which is also why the Pell error decays at the
square of the dominant root (questions 11--14). Powers of the
adjacency matrix count walks because matrix multiplication sums
over intermediate vertices, so spectra count closed walks
(questions 15--19). Commuting matrices share an eigenbasis, and
one Fourier basis then diagonalizes the whole circulant algebra
in one stroke (questions 20--24). Summits: the fundamental
theorem of linear recurrences (question 9); and for nonnegative
matrices, the reason dominant roots like $\varphi$ or $1 +
\sqrt2$ are automatically real, positive and simple is the
Perron--Frobenius theorem, proved in the Year 3 volume.
\end{solution}
