\section*{Chapter \ref{ch:b2:metric} --- Topology of Metric Spaces}

\begin{solution}{exo:b2:metric:1}
$\delta$: symmetry and separation are clear; triangle inequality:
$\min(1, u + v) \leq \min(1,u) + \min(1,v)$ for $u, v \geq 0$ (if
either min is $1$, the right side is $\geq 1$; else it is $u + v$).
$d$: it is the pullback of $\abs{\cdot}$ by the injective $\arctan$:
the three axioms transfer.

Convergence: for $\delta$, $\delta(x_n, x) \to 0 \iff \abs{x_n - x}
\to 0$ (for small values the two distances agree): same convergent
sequences as usual. For $d$: $d(x_n, x) \to 0 \iff \arctan x_n \to
\arctan x \iff x_n \to x$ (continuity and strict monotonicity of
$\arctan$ and of its inverse on the relevant ranges): again the
usual convergence.

$(\R, d)$ is \emph{not} complete: $x_n = n$ satisfies $d(x_p, x_q)
= \abs{\arctan p - \arctan q} \to 0$ (both tend to $\frac\pi2$):
Cauchy; but $(x_n)$ does not converge for $d$ (its $d$-limit would
be an ordinary limit). Completeness is a property of the
\emph{distance}, not just of the convergent sequences.
\end{solution}

\begin{solution}{exo:b2:metric:2}
Convergent $\Rightarrow$ Cauchy: $d(x_p, x_q) \leq d(x_p, \ell) +
d(\ell, x_q)$. Bounded: beyond $N$, $d(x_n, \ell) \leq 1$; the
finitely many first terms are within some radius too.

Cauchy $+$ convergent subsequence $x_{\varphi(n)} \to \ell$: given
$\varepsilon$, for large $n$, $d(x_n, \ell) \leq d(x_n,
x_{\varphi(n)}) + d(x_{\varphi(n)}, \ell) \leq 2\varepsilon$ (the
first term by Cauchy, since $\varphi(n) \geq n$).

Compact $\Rightarrow$ complete: a Cauchy sequence has a convergent
subsequence (compactness), hence converges.
\end{solution}

\begin{solution}{exo:b2:metric:3}
$d_\infty(f, g) = \sup_{\intcc{0}{1}} \abs{x - x^2} = \frac14$
(maximum of $x - x^2$ at $x = \frac12$).

$\overline B(0, 1) = \{f : \sup\abs f \leq 1\}$: the continuous
functions with values in $\intcc{-1}{1}$.

$\{f : f(0) = 0\}$ is the preimage of $\{0\}$ under the
\emph{evaluation} $f \mapsto f(0)$, which is $1$-Lipschitz
($\abs{f(0) - g(0)} \leq d_\infty(f,g)$), hence continuous: the set
is closed (\cref{thm:b2:metric:globalcontinuity}). Likewise $\{f :
f(0) > 0\}$ is the preimage of the open $\intoo{0}{+\infty}$: open.
\end{solution}

\begin{solution}{exo:b2:metric:4}
\emph{Complete:} a Cauchy sequence with $\varepsilon = \frac12$ is
eventually constant, hence convergent.

\emph{Compact iff finite:} if $X$ is finite, any sequence takes some
value infinitely often (constant subsequence). If $X$ is infinite,
a sequence of pairwise distinct points has all mutual distances $1$:
no Cauchy subsequence, so no convergent one.

\emph{Connected subsets:} the singletons (and $\emptyset$). Any $A$
with two points $x \neq y$ splits as $\{x\} \cup (A
\setminus\{x\})$, both open in $A$ (every subset of a discrete
space is open --- balls of radius $\frac12$ are singletons).
\end{solution}

\begin{solution}{exo:b2:metric:5}
The function $g(x) = d(x, f(x))$ is continuous on the compact $X$
($\abs{g(x) - g(y)} \leq 2d(x,y)$ by two triangle inequalities), so
it attains its minimum at some $a$
(\cref{thm:b2:metric:compactprops}). If $f(a) \neq a$:
\[
g\bigl(f(a)\bigr) = d\bigl(f(a), f(f(a))\bigr) < d(a, f(a)) = g(a),
\]
contradicting minimality. So $f(a) = a$; uniqueness as usual (two
fixed points $a \neq b$ give $d(a,b) = d(f(a), f(b)) < d(a,b)$).

Non-compact example: $f(x) = x + \frac1x$ on $\intco{1}{+\infty}$:
$\abs{f(x) - f(y)} = \abs{x - y}\,\abs{1 - \frac{1}{xy}} < \abs{x -
y}$ for $x \neq y$ (as $xy > 1$), yet $f(x) > x$ everywhere.
\end{solution}

\begin{solution}{exo:b2:metric:6}
Pick $x_n \in K_n$. All terms from rank $n$ on lie in $K_n$; in
particular the whole sequence lies in the compact $K_0$: extract
$x_{\varphi(k)} \to \ell$. For each fixed $n$, the terms
$x_{\varphi(k)}$ with $\varphi(k) \geq n$ lie in the \emph{closed}
$K_n$, so the limit $\ell \in K_n$. Hence $\ell \in \bigcap K_n$.

Closed counterexample: $F_n = \intco{n}{+\infty}$ in $\R$: nested,
closed, nonempty, empty intersection.
\end{solution}

\begin{solution}{exo:b2:metric:7}
Continuity of $f^{-1}$ means: images $f(F)$ of closed sets $F
\subseteq K$ are closed (preimages under $f^{-1}$ are images under
$f$). A closed $F$ in the compact $K$ is compact
(\cref{thm:b2:metric:compactprops}~(1)); its continuous image
$f(F)$ is compact, hence closed. So $f^{-1}$ is continuous: $f$ is
a homeomorphism.

Counterexample: $\gamma(t) = (\cos t, \sin t)$ from
$\intco{0}{2\pi}$ (not compact) onto the unit circle is a
continuous bijection, but $\gamma^{-1}$ is discontinuous at $(1,0)$:
points on the circle just below the axis have parameters near
$2\pi$, not near $0$.
\end{solution}

\begin{solution}{exo:b2:metric:8}
$C = \bigcap_n C_n$ where each $C_n$ (union of $2^n$ closed
intervals of length $3^{-n}$) is closed: $C$ is closed and bounded
in $\R$, hence compact (\cref{thm:b2:metric:compactprops}~(2)).

Empty interior: $C$ contains no interval of length $> 3^{-n}$ (it
sits inside $C_n$, whose components have that length), for every
$n$.

Cardinality: the points of $C$ are exactly the reals $\sum_{n\geq1}
a_n 3^{-n}$ with digits $a_n \in \{0, 2\}$ (at each stage, the
deleted middle third removes the digit $1$); the map $(a_n) \mapsto
\sum a_n 3^{-n}$ is a bijection from $\{0,2\}^{\N^*}$ onto $C$
(injectivity as in \cref{exo:b2:structures:3}). So $C$ is equipotent
to $\{0,1\}^{\N}$: uncountable, though of ``length zero''.
\end{solution}

\begin{solution}{exo:b2:metric:9}
Suppose $a \notin f(X)$. The image $f(X)$ is compact (continuous
image), hence closed; so
\[
\varepsilon = d\bigl(a, f(X)\bigr)
= \inf_{y \in f(X)} d(a, y) > 0
\]
(the infimum of a continuous function on a compact set is attained;
if it were $0$, $a$ would be adherent to the closed $f(X)$, hence in
it).

Consider the orbit $x_n = f^n(a)$ ($x_0 = a$). For $p < q$:
\[
d(x_p, x_q) = d\bigl(f^p(a), f^p(f^{q-p}(a))\bigr)
= d\bigl(a, f^{q-p}(a)\bigr) \geq \varepsilon,
\]
(isometry iterated $p$ times; and $f^{q-p}(a) \in f(X)$ since $q - p
\geq 1$). A sequence with mutual distances $\geq \varepsilon$ has no
convergent subsequence --- contradicting compactness. Hence $f(X) =
X$.
\end{solution}

\begin{solution}{exo:b2:metric:10}
Let $A, B \in GL_n(\C)$ and $p(z) = \det\bigl((1-z)A + zB\bigr)$: a
polynomial in $z$ (each entry is affine in $z$; the determinant is a
polynomial in the entries). $p(0) = \det A \neq 0$: $p$ is not
identically zero, so it has finitely many roots $z_1, \dots, z_m$
(none equal to $0$ or $1$: $p(1) = \det B \neq 0$). The plane $\C$
minus finitely many points is path-connected: a path from $0$ to $1$
avoiding the $z_i$ exists (take a broken line through a point far
from all roots, or a circular arc; only finitely many obstacles).
Along such a path $\gamma$, $t \mapsto (1 - \gamma(t))A +
\gamma(t)B$ is a continuous path \emph{inside} $GL_n(\C)$ from $A$
to $B$ (the determinant never vanishes on it). Hence $GL_n(\C)$ is
path-connected --- unlike its real cousin
(\cref{ex:b2:metric:glnr}): the complex plane has room to walk
around obstacles.
\end{solution}

\begin{solution}{exo:b2:metric:11}
\emph{Lipschitz:} for $a \in A$, $d(x, a) \leq d(x, y) + d(y,
a)$; take the infimum over $a$: $d(x, A) \leq d(x, y) + d(y,
A)$, and swap $x, y$: $\abs{d(x,A) - d(y,A)} \leq d(x,y)$.

\emph{Vanishing:} $d(x, A) = 0$ iff there are $a_n \in A$ with
$d(x, a_n) \to 0$ iff $x$ is a limit of points of $A$ iff $x \in
\overline A$.

\emph{The switch:} for disjoint closed $A, B$: the denominator
$d(x,A) + d(x,B)$ never vanishes (it would force $x \in \overline
A \cap \overline B = A \cap B = \emptyset$), so $\varphi$ is well
defined, and continuous as a quotient of continuous functions
with nonvanishing denominator. $\varphi(x) = 0$ iff $d(x, A) = 0$
iff $x \in A$; $\varphi(x) = 1$ iff $d(x, B) = 0$ iff $x \in B$;
and $0 \leq \varphi \leq 1$ everywhere.
\end{solution}

\begin{solution}{exo:b2:metric:12}
Let $B(x_0, r_0)$ be any ball; we find a point of $\bigcap U_n$
in it. Since $U_1$ is dense and open, $U_1 \cap B(x_0, r_0)$ is
nonempty and open: it contains a closed ball $\overline B(x_1,
r_1)$ with $0 < r_1 \leq \frac{r_0}2$ (shrink the radius).
Inductively, $U_{n+1} \cap B(x_n, r_n)$ is nonempty open: pick
$\overline B(x_{n+1}, r_{n+1}) \subseteq U_{n+1} \cap B(x_n,
r_n)$ with $r_{n+1} \leq \frac{r_n}2$. For $p, q \geq n$, both
$x_p, x_q$ lie in $B(x_n, r_n)$ with $r_n \leq 2^{-n}r_0$: the
sequence is Cauchy, and converges to some $\ell$ by
completeness. For each $n$, the tail of the sequence lies in the
\emph{closed} ball $\overline B(x_{n+1}, r_{n+1}) \subseteq U_{n+1}
\cap B(x_0, r_0)$, so $\ell \in U_{n+1}$ for every $n$ and $\ell
\in \overline B(x_1, r_1) \subseteq B(x_0, r_0)$. Hence
$\bigcap_n U_n$ meets every ball: dense.

\emph{Application:} if $\R = \bigcup_n F_n$ with $F_n$ closed
with empty interior, then $U_n = \R \setminus F_n$ are dense
($\overline{U_n} = \R$ iff $F_n$ has empty interior) and open,
and Baire gives a point in $\bigcap U_n = \R \setminus \bigcup
F_n$: contradiction. In particular $\R \neq \bigcup_{x \in D}
\{x\}$ for a countable $D$ (singletons are closed with empty
interior): $\R$ is uncountable ---
\cref{thm:b2:structures:cantor} by another route.
\end{solution}

\begin{solution}{pb:b2:metric:1}
\textbf{1.} Let $F$ be closed in the complete $X$ and $(x_n)
\subseteq F$ Cauchy: it converges in $X$ to some $\ell$, and
$\ell \in F$ because $F$ is closed (limits of sequences of $F$
stay in $\overline F = F$): $F$ is complete. Conversely let $A
\subseteq X$ be complete and $x \in \overline A$: some sequence
of $A$ converges to $x$; it is Cauchy, so it converges \emph{in}
$A$; limits are unique, so $x \in A$: $A$ closed. Since
$\bigl(C(I), d_\infty\bigr)$ is complete
(\cref{thm:b2:metric:rncomplete}), its closed subsets are
complete.

\textbf{2.} For $y, z \in C(I)$ and $t \in I$:
\[
\Bigl|\int_{t_0}^{t}y - \int_{t_0}^{t}z\Bigr| \leq
\abs{t - t_0}\,\sup_I\abs{y - z} \leq h\,d_\infty(y, z),
\]
and take the sup over $t$.

\textbf{3.} Using the two fixed point equations and the triangle
inequality:
\[
d(\ell_g, \ell_{\widetilde g})
= d\bigl(g(\ell_g), \widetilde g(\ell_{\widetilde g})\bigr)
\leq d\bigl(g(\ell_g), g(\ell_{\widetilde g})\bigr) +
d\bigl(g(\ell_{\widetilde g}), \widetilde g(\ell_{\widetilde
g})\bigr)
\leq k\,d(\ell_g, \ell_{\widetilde g}) +
d\bigl(g(\ell_{\widetilde g}), \widetilde g(\ell_{\widetilde
g})\bigr),
\]
and solve for $d(\ell_g, \ell_{\widetilde g})$ (the coefficient
$1 - k$ is positive). The second inequality bounds the evaluated
gap by the uniform one.

\textbf{4.} $g^m$ is a contraction on a nonempty complete space:
it has a unique fixed point $\ell$
(\cref{thm:b2:metric:banach}). Then $g^m(g(\ell)) =
g(g^m(\ell)) = g(\ell)$: $g(\ell)$ is a fixed point of $g^m$,
so $g(\ell) = \ell$ by uniqueness. Any fixed point of $g$ is one
of $g^m$: uniqueness for $g$. Orbits: fix $r \in \{0, \dots,
m-1\}$; the subsequence $(x_{qm + r})_q$ is the $g^m$-orbit
started at $x_r$, so it converges to $\ell$ as $q \to \infty$
(Banach again). All $m$ subsequences converge to the same
$\ell$, hence $x_n \to \ell$: given $\varepsilon$, each residue
class is eventually within $\varepsilon$, and there are finitely
many classes.

\textbf{5.} If $y$ is continuous with values in $\intcc{y_0 -
b}{y_0 + b}$, the integrand $s \mapsto f(s, y(s))$ is continuous
on $I$ (composition), so the right-hand side is $C^1$ with
derivative $f(t, y(t))$ (fundamental theorem of calculus, Year 1
volume). If $y$ satisfies the integral equation, it is that
$C^1$ function, $y(t_0) = y_0$, and $y' = f(t, y)$. Conversely,
integrating $y' = f(s, y(s))$ from $t_0$ to $t$ gives the
integral equation.

\textbf{6.} $X_h$ contains the constant $y_0$; it is closed as
the preimage of $\intcc{0}{b}$ under the continuous map $y
\mapsto d_\infty(y, y_0)$ (distances are $1$-Lipschitz), hence
complete by question 1. Stability: for $y \in X_h$ and $t \in
I$,
\[
\abs{T(y)(t) - y_0} = \Bigl|\int_{t_0}^{t}f(s, y(s))\dd s\Bigr|
\leq M\abs{t - t_0} \leq Mh \leq b ,
\]
the last step by $h \leq b/M$ (or $M = 0$, trivial). And $T(y)$
is continuous ($C^1$ even, question 5): $T(y) \in X_h$.

\textbf{7.} For $t \in I$:
\[
\abs{T(y)(t) - T(z)(t)} \leq \int_{t_0}^{t}\abs{f(s, y(s)) -
f(s, z(s))}\,\abs{\dd s} \leq L\abs{t - t_0}\,d_\infty(y,z)
\leq Lh\,d_\infty(y,z).
\]
If $Lh < 1$: $T$ is a contraction of the nonempty complete
$X_h$, and Banach gives a unique fixed point --- by question 5,
the unique solution.

\textbf{8.} Induction; the case $n = 1$ is question 7's middle
inequality. Assuming the bound for $n$, for $t \geq t_0$ (the
case $t \leq t_0$ is symmetric):
\[
\abs{T^{n+1}(y)(t) - T^{n+1}(z)(t)}
\leq L\int_{t_0}^{t}\abs{T^n(y)(s) - T^n(z)(s)}\dd s
\leq L\int_{t_0}^{t}\frac{L^n(s - t_0)^n}{n!}\dd s\;
d_\infty(y,z),
\]
and the integral evaluates to $\frac{L^n(t -
t_0)^{n+1}}{(n+1)!}$: the bound with $n + 1$. Hence
$d_\infty(T^n y, T^n z) \leq \frac{(Lh)^n}{n!}d_\infty(y, z)$,
and $\frac{(Lh)^n}{n!} \to 0$ (the exponential series
converges): some $T^m$ is a contraction. Question 4 applies on
the complete $X_h$: $T$ has a unique fixed point, i.e.\ the
Cauchy problem has exactly one solution on $I$ with values in
$\intcc{y_0 - b}{y_0 + b}$.

\textbf{9.} Let $y$ solve the problem on $I$ and suppose the set
$E = \{t \in I, t > t_0 : \abs{y(t) - y_0} > b\}$ is nonempty
(the side $t < t_0$ is symmetric); let $\tau = \inf E$. By
continuity, $\abs{y(s) - y_0} \leq b$ for $s \in
\intcc{t_0}{\tau}$, so the graph lies in $R$ there, the integral
equation holds up to $\tau$, and
\[
\abs{y(\tau) - y_0} = \Bigl|\int_{t_0}^{\tau}f\bigl(s,
y(s)\bigr)\dd s\Bigr| \leq M(\tau - t_0) \leq Mh \leq b .
\]
If $\tau < t_0 + h$, points of $E$ arbitrarily close to $\tau$
from the right give, by continuity, $\abs{y(\tau) - y_0} \geq
b$, hence $= b$; but then the display forces $M(\tau - t_0) =
Mh$, i.e.\ $\tau = t_0 + h$: contradiction. So $\tau = t_0 + h$,
$E \subseteq \{t_0 + h\}$, and the display (at $\tau = t_0 + h$)
gives $\abs{y(t_0 + h) - y_0} \leq b$, contradicting membership
in $E$. Hence $E = \emptyset$: every solution on $I$ stays in
the band, is a fixed point of $T$ in $X_h$, and uniqueness is
unconditional.

\textbf{10.} $T(y)(t) = 1 + \int_0^t y$. From $y^{(0)} \equiv
1$:
\[
y^{(1)}(t) = 1 + t,\quad
y^{(2)}(t) = 1 + t + \frac{t^2}2,\quad\dots\quad
y^{(n)}(t) = \sum_{k=0}^{n}\frac{t^k}{k!}
\]
(induction: integrating the partial sum adds the next term).
These are the Taylor partial sums of $\eu^{\,t}$; on any bounded
$I$ they converge uniformly to $\eu^{\,t}$ (the tail is
dominated by the convergent numerical series $\sum h^k/k!$),
which is indeed the unique solution.

\textbf{11.} $y \equiv 0$ is a solution. For $y_c$: it is $C^1$
(both pieces are, and at $t = c$ the derivatives match: $0$ and
$2(t - c) \to 0$), and for $t > c$: $y_c' = 2(t - c) =
2\sqrt{(t-c)^2} = 2\sqrt{\abs{y_c}}$; for $t \leq c$ both sides
vanish. So the Cauchy problem $y(0) = 0$ has infinitely many
solutions ($c \geq 0$ arbitrary, and $y \equiv 0$). The field
$\varphi(y) = 2\sqrt{\abs y}$ is not Lipschitz near $0$:
$\frac{\varphi(y) - \varphi(0)}{y - 0} = \frac{2}{\sqrt y} \to
+\infty$ as $y \to 0^+$: no constant $L$ works on any
neighbourhood of $0$ --- exactly where all the solutions
branch.

\textbf{12.} Separating variables (or checking directly), the
unique local solution is $y(t) = \frac1{1 - t}$, defined on
$\intoo{-\infty}{1}$ and blowing up at $t = 1$. For the
rectangle $\intcc{-a}{a} \times \intcc{1 - b}{1 + b}$: $M =
\sup y^2 = (1 + b)^2$, so the certified half-width is $h =
\min\bigl(a, \frac{b}{(1+b)^2}\bigr)$. Maximizing
$\frac{b}{(1+b)^2}$: derivative zero at $b = 1$, value
$\frac14$. So the theorem guarantees life only on
$\intcc{-\frac14}{\frac14}$ --- correctly less than the true
lifespan $1$ forward, and infinitely less backward: the theorem
is local by nature, and blow-up shows it cannot be otherwise.

\textbf{13.} $g$ maps $\intcc12$ into itself: $g$ decreases on
$\intcc{1}{\sqrt2}$ and increases after (study $g'(x) = \frac12
- \frac1{x^2}$), with $g(1) = g(2) = \frac32$ and minimum
$g(\sqrt2) = \sqrt2 > 1$: $g(\intcc12) \subseteq
\intcc{\sqrt2}{\frac32} \subseteq \intcc12$; and $g$ maps
rationals to rationals. Contraction: $\abs{g'(x)} = \abs{\frac12
- \frac1{x^2}} \leq \frac12$ on $\intcc12$ ($\frac1{x^2} \in
\intcc{\frac14}{1}$), so the mean value inequality gives
$\abs{g(x) - g(y)} \leq \frac12\abs{x - y}$. A fixed point
satisfies $\frac x2 = \frac1x$, i.e.\ $x^2 = 2$: impossible in
$\Q$. The failing hypothesis is completeness of $X$ ($\Q \cap
\intcc12$ is not complete); in the completion $\intcc12$ the
fixed point is $\sqrt2$ --- Banach's theorem run on the
rationals \emph{creates} the irrational.

\textbf{14.} A posteriori: $d(x_n, \ell) \leq d(x_n, x_{n+1}) +
d(x_{n+1}, \ell) \leq k\,d(x_{n-1}, x_n) + k\,d(x_n, \ell)$,
whence $d(x_n, \ell) \leq \frac{k}{1-k}d(x_n, x_{n-1})$. Heron
from $x_0 = \frac32$: $x_1 = \frac{17}{12}$, $d(x_1, x_0) =
\frac1{12}$, $k = \frac12$: the a priori bound
$\frac{k^n}{1-k}d(x_1,x_0) = \frac{2^{-n+1}}{12}$ drops below
$10^{-6}$ first at $n = 18$. In reality $x_1 = \frac{17}{12}
\approx 1.41667$ (error $2.5\cdot10^{-3}$), $x_2 =
\frac{577}{408} \approx 1.4142157$ (error $2.1\cdot10^{-6}$),
$x_3 \approx 1.41421356237469$ (error $1.6\cdot10^{-12}$):
three steps suffice. Each Heron step roughly \emph{squares} the
error (quadratic convergence, a Newton phenomenon:
\cref{ch:b2:realfun}); the contraction estimate, which only
halves it, is honest for the worst case but pessimistic here.

\textbf{15.} Apply question 3 with $g = T_{y_0}$ (an
$Lh$-contraction, $Lh < 1$) and $\widetilde g = T_{z_0}$, whose
fixed point is $y[z_0]$. For any $y$,
\[
\abs{T_{y_0}(y)(t) - T_{z_0}(y)(t)} = \abs{y_0 - z_0},
\]
(the integrals are identical), so $\sup_y
d_\infty(T_{y_0}(y), T_{z_0}(y)) = \abs{y_0 - z_0}$, and
question 3 gives
\[
d_\infty(y[y_0], y[z_0]) \leq \frac{\abs{y_0 - z_0}}{1 - Lh} .
\]

\textbf{16.} On each piece, question 15 applied with the
left-endpoint values as initial data bounds the deviation at
the right endpoint:
\[
d_\infty \leq \frac{1}{1 - 1/2}\,(\text{left deviation})
= 2\,(\text{left deviation}) .
\] By
induction over the $m$ pieces, the final deviation is at most
$2^m\abs{y_0 - z_0}$, and the uniform deviation over the whole
segment obeys the same bound (each piece's sup is controlled at
its stage). With pieces of length $h \asymp \frac1{2L}$, the
factor is $2^m = 2^{\,\text{length}\cdot 2L}$: exponential in
the length of the interval, exactly as Gronwall's bound
$\eu^{L\abs{t-t_0}}$ predicts, with better constants.

\textbf{17.} Same scheme: for $y \in X_h$,
\[
\abs{T_f(y)(t) - T_g(y)(t)} \leq \int_{t_0}^t \abs{f(s,y(s)) -
g(s,y(s))}\,\abs{\dd s} \leq \varepsilon h ,
\]
so question 3 (with $T_f$ the contraction, $T_g$ the perturbed
map) yields $d_\infty \leq \frac{\varepsilon h}{1 - Lh}$.

\textbf{18.} By question 17 applied to $f = f_\lambda$, $g =
f_\mu$: $d_\infty(y_\lambda, y_\mu) \leq \frac{Ch}{1 -
Lh}\abs{\lambda - \mu}$: the solution map is Lipschitz with
constant $\frac{Ch}{1-Lh}$.

\textbf{19.} Every argument used only: the metric axioms, the
completeness of the stage, the bound $\abs{\int} \leq
\int\abs{\cdot}$, and the Lipschitz property of $f$ --- all
available for $\R^n$-valued functions with $d_\infty$ built on
any of the equivalent distances of \cref{ex:b2:metric:examples}
(complete by \cref{thm:b2:metric:rncomplete}). For $y' = Ay$
with the nilpotent $A$: $y^{(0)} \equiv (c_1, c_2)$,
\[
y^{(1)}(t) = (c_1, c_2) + \int_0^t (c_2, 0)\,\dd s
= (c_1 + tc_2,\; c_2),
\]
and $Ay^{(1)}(s) = (c_2, 0)$ again: $y^{(2)} = y^{(1)}$. The
iteration is stationary from $n = 1$ on, at the exact solution
$y(t) = (c_1 + tc_2, c_2) = \eu^{tA}y(0)$ --- nilpotency
truncates the exponential series, and Picard notices.

\textbf{20.} Fix $y \in X$ and let $g_y(x) = y - \eta(x)$: a
$k$-contraction of the complete $X$, so there is exactly one $x$
with $x + \eta(x) = y$: the map $\Phi = \mathrm{id} + \eta$ is
bijective. Lipschitz inverse: if $\Phi(x) = y$ and $\Phi(x') =
y'$,
\[
d(x, x') \leq d(y, y') + d\bigl(\eta(x), \eta(x')\bigr)
\leq d(y, y') + k\,d(x, x'),
\]
so $d(x, x') \leq \frac{1}{1-k}d(y, y')$. (Distances here come
from the norm structure, so $d(a - c, b - c) = d(a, b)$, which
the first inequality used.)

\textbf{21.} $g(x) = m + e\sin x$ is $e$-Lipschitz on the
complete $\R$ (mean value inequality, $\abs{g'} = \abs{e\cos x}
\leq e < 1$): Banach gives a unique solution and global
convergence of the iteration. For $e = \frac12$, $m = 1$, $x_0
= 1$: $x_1 = 1 + \frac12\sin 1 \approx 1.42074$, $d(x_1, x_0)
\approx 0.4207$, and the a priori bound
$\frac{(1/2)^n}{1/2}\cdot 0.4207 \leq 10^{-3}$ first holds at
$n = 10$: ten iterations certified (the true value $x \approx
1.4987$ is in fact reached to $10^{-3}$ already around $n =
5$).

\textbf{22.} Use the digit description
(\cref{exo:b2:metric:8}): $C$ is the set of sums
$\sum_{n\geq1}a_n3^{-n}$, $a_n \in \{0, 2\}$. Then $S_1(C) =
\{x/3 : x \in C\}$ is the subset with $a_1 = 0$, and $S_2(C) =
\{x/3 + 2/3\}$ the subset with $a_1 = 2$: their union, over the
free choice of $a_1$, is exactly $C$. Uniqueness in one
sentence: on the space of nonempty compact subsets of
$\intcc01$ metrized by the Hausdorff distance, $A \mapsto
S_1(A) \cup S_2(A)$ is a $\frac13$-contraction of a complete
space, so Banach allows only one fixed set --- the Year 3
volume makes the Hausdorff metric and this argument rigorous.

\textbf{23.} Let $Z = \{t \in J : y(t) = z(t)\}$: nonempty
($t_0 \in Z$), closed in $J$ (equalizer of two continuous maps:
preimage of $\{0\}$ under $y - z$). Open: if $t_1 \in Z$, apply
the local theorem (question 8) at the point $(t_1, y(t_1))$,
in a rectangle where $f$ is Lipschitz: on a small interval
around $t_1$ both $y$ and $z$ solve the same Cauchy problem, so
they coincide there (question 9's unconditional uniqueness):
a neighbourhood of $t_1$ lies in $Z$. A nonempty subset of the
interval $J$ that is both open and closed in $J$ is all of $J$
(intervals are connected,
\cref{thm:b2:metric:connectedness}): $y = z$ on $J$.

\textbf{24.} Here $f(t, y) = \alpha(t)y + \beta(t)$ is
$L$-Lipschitz in $y$ on all of $\intcc AB \times \R$ with $L =
\sup\abs\alpha$ (finite: $\alpha$ continuous on a segment), and
no band $\intcc{y_0 - b}{y_0 + b}$ is needed: take $X = C(\intcc
AB)$ entire, on which $T$ is well defined. The induction of
question 8 runs verbatim and gives $d_\infty(T^ny, T^nz) \leq
\frac{(L(B - A))^n}{n!}d_\infty(y, z)$: an iterate is a
contraction whatever the length $B - A$, and question 4
concludes: one and only one solution on the whole segment.
Linearity enters exactly once: it makes the Lipschitz bound
global in $y$, removing the rectangle and its $h \leq b/M$.

\textbf{25.} \emph{Completeness} turned the Cauchy sequence of
iterates into an actual solution (questions 1, 6, 8), and its
absence let $\sqrt2$ escape from $\Q$ (question 13).
\emph{Contraction} gave uniqueness, the algorithm, and the
error bars (questions 7, 14). \emph{The iterate trick} removed
the smallness condition $Lh < 1$, so the certified interval
depends only on $M$, not on $L$ --- and made linear equations
global (questions 8, 24). \emph{Connectedness} promoted local
uniqueness to global uniqueness (question 23). The
counterexamples: $2\sqrt{\abs y}$ guards Lipschitz (question
11), $y^2$ guards locality (question 12), $\Q$ guards
completeness (question 13). The summit is the
Picard--Lindel\"of theorem (question 8); when $f$ is merely
continuous, existence survives but uniqueness dies, and the
proof trades the contraction for compactness of function sets
--- Peano's theorem via Arzel\`a--Ascoli, in the Year 3
volume.
\end{solution}
