\section*{Chapter \ref{ch:b2:nvs} --- Normed Vector Spaces}

\begin{solution}{exo:b2:nvs:1}
Unit balls: a diamond ($\norm\cdot_1$), a disk ($\norm\cdot_2$), a
square ($\norm\cdot_\infty$), nested in that order. Inequalities:
$\norm x_\infty \leq \norm x_2$ (one square is at most the sum);
$\norm x_2 \leq \norm x_1$ (squaring: $x_1^2 + x_2^2 \leq (\abs{x_1}
+ \abs{x_2})^2$); $\norm x_1 \leq 2\norm x_\infty$ (two terms, each
$\leq \max$). Sharpness: $(1, 0)$ makes the first two equalities;
$(1, 1)$ makes $\norm x_1 = 2\norm x_\infty$ and also shows
$\norm x_2 = \sqrt2 \norm x_\infty$ and $\norm x_1 = \sqrt2 \norm
x_2$ are the extreme ratios in the other direction.
\end{solution}

\begin{solution}{exo:b2:nvs:2}
\omterm{def:b2:nvs:norm}{Norm} axioms: homogeneity and triangle inequality are inherited
termwise; separation: $N(f) = 0$ forces $f' = 0$ (so $f$ constant)
and $f(0) = 0$: $f = 0$. It is a \omterm{def:b2:nvs:norm}{norm}.

Comparison: $\norm f_\infty \leq N(f)$, since $\abs{f(x)} \leq
\abs{f(0)} + \abs{\int_0^x f'} \leq \abs{f(0)} +
\norm{f'}_\infty$. The converse fails: take $f_n(x) = \frac1n
\sin(nx)$: then $\norm{f_n}_\infty \leq \frac1n \to 0$ while
$N(f_n) = 0 + \norm{\cos(nx)}_\infty = 1$. No constant $C$ gives
$N \leq C\norm\cdot_\infty$.
\end{solution}

\begin{solution}{exo:b2:nvs:3}
$\abs{u(f)} \leq \norm f_\infty \int_0^1 \eu^t\,\dd t = (\eu -
1)\norm f_\infty$, with equality for $f \equiv 1$: $\vertiii u =
\eu - 1$.

Shift: $\norm{S(x)}_\infty = \max(\abs{x_2}, \dots, \abs{x_n}) \leq
\norm x_\infty$, equality at $x = e_2$: $\vertiii S = 1$ (for $n
\geq 2$).
\end{solution}

\begin{solution}{exo:b2:nvs:4}
Upper bound: for $\norm x_\infty \leq 1$,
\[
\abs{(Ax)_i} = \Bigl|\sum_j a_{ij}x_j\Bigr| \leq \sum_j
\abs{a_{ij}},
\]
so $\norm{Ax}_\infty \leq \max_i \sum_j \abs{a_{ij}}$. Attained: let
$i_0$ realize the max and take $x_j = \operatorname{sign}(a_{i_0
j})$ (modulus $1$ entries): then $(Ax)_{i_0} = \sum_j \abs{a_{i_0
j}}$. Hence the formula. For the given matrix: row sums $3$ and
$4$: $\vertiii A_\infty = 4$.
\end{solution}

\begin{solution}{exo:b2:nvs:5}
\emph{Geometric series:} for $\vertiii B < 1$, the series $\sum
B^k$ converges absolutely in the Banach $\mathcal{M}_n(K)$
(\cref{thm:b2:nvs:absoluteconvergence}, $\vertiii{B^k} \leq
\vertiii B^k$), and
\[
(I - B)\sum_{k=0}^{K} B^k = I - B^{K+1} \longrightarrow I :
\]
by \omterm{def:b2:metric:continuity}{continuity} of the product
(\cref{prop:b2:nvs:bilinear}), $(I - B)\sum_{k\geq0} B^k = I$, so
$I - B$ is invertible with inverse the sum (and
$\vertiii{(I-B)^{-1}} \leq \frac{1}{1 - \vertiii B}$).

\emph{Openness:} for $A$ invertible and $\vertiii H <
\frac{1}{\vertiii{A^{-1}}}$: $A + H = A(I + A^{-1}H)$ with
$\vertiii{A^{-1}H} \leq \vertiii{A^{-1}}\vertiii H < 1$: invertible.
So a ball around $A$ stays in $GL_n$.

\emph{\omterm{def:b2:metric:continuity}{Continuity} of inversion:} with $B = -A^{-1}H$,
\[
(A + H)^{-1} - A^{-1} = \bigl((I - B)^{-1} - I\bigr)A^{-1}
= \Bigl(\sum_{k \geq 1} B^k\Bigr) A^{-1},
\]
of \omterm{def:b2:nvs:norm}{norm} $\leq \frac{\vertiii B}{1 - \vertiii B}\vertiii{A^{-1}}
\to 0$ as $H \to 0$.
\end{solution}

\begin{solution}{exo:b2:nvs:6}
\omterm{def:b2:metric:continuity}{Continuous} $\Rightarrow$ closed kernel: preimage of the closed
$\{0\}$ (\cref{thm:b2:metric:globalcontinuity}).

Conversely, suppose $\ker\varphi$ closed and $\varphi \neq 0$. Pick
$a$ with $\varphi(a) = 1$; since $a \notin \ker\varphi$ and the
kernel is closed, some ball $B(a, r)$ misses it. Now let $h \in E$
with $\varphi(h) \neq 0$: the vector $a - \frac{h}{\varphi(h)}$
lies in $\ker\varphi$, hence outside $B(a, r)$:
\[
\Bigl\Vert \frac{h}{\varphi(h)} \Bigr\Vert \geq r
\quad\Longrightarrow\quad
\abs{\varphi(h)} \leq \frac{\norm h}{r},
\]
an inequality trivially true when $\varphi(h) = 0$ as well: bound
(4) of \cref{thm:b2:nvs:continuouslinear}: \omterm{def:b2:metric:continuity}{continuous}.
\end{solution}

\begin{solution}{exo:b2:nvs:7}
Let $f_n$ be $0$ on $\intcc{0}{\frac12 - \frac1n}$, affine up to
value $1$ at $\frac12$, then $1$ on $\intcc{\frac12}{1}$. For $m
\geq n$, $f_m - f_n$ is supported on an interval of length
$\frac1n$ with values in $\intcc{-1}{1}$: $\norm{f_m - f_n}_1 \leq
\frac1n$: Cauchy.

Suppose $f_n \to f$ in $\norm\cdot_1$ with $f$ \omterm{def:b2:metric:continuity}{continuous}. On
$\intcc{0}{\frac12 - \delta}$ (fixed $\delta$): $\int \abs{f} =
\int\abs{f - f_n} \leq \norm{f - f_n}_1 \to 0$ for $n >
\frac1\delta$, so $\int_0^{1/2 - \delta}\abs f = 0$, and by strict
positivity $f = 0$ there --- for every $\delta$: $f = 0$ on
$\intoo{0}{\frac12}$. Similarly $f = 1$ on $\intcc{\frac12}{1}$
(the $f_n$ all equal $1$ there). By \omterm{def:b2:metric:continuity}{continuity} at $\frac12$: $0 =
1$, absurd. No limit exists: the space is not \omterm{def:b2:metric:complete}{complete}.
\end{solution}

\begin{solution}{exo:b2:nvs:8}
Well defined and \omterm{def:b2:metric:continuity}{continuous}: $\abs{\varphi(f)} \leq \sum 2^{-n}
\norm f_\infty = \norm f_\infty$, so $\varphi$ is a linear form
with $\vertiii\varphi \leq 1$ (the series converges absolutely for
each $f$).

\omterm{def:b2:nvs:norm}{Norm} $1$: fix $K$; the points $1, \frac12, \dots, \frac1K$ are
pairwise distinct, so there is a \omterm{def:b2:metric:continuity}{continuous} $f_K$, $\norm{f_K}_\infty
\leq 1$, with $f_K(\frac1n) = (-1)^n$ for $n \leq K$ (piecewise
affine interpolation, constant near $0$). Then
\[
\varphi(f_K) \geq \sum_{n=1}^{K} 2^{-n} - \sum_{n > K} 2^{-n}
= 1 - 2^{-K+1} \xrightarrow[K \to \infty]{} 1 .
\]

Not attained: if $\norm f_\infty \leq 1$ and $\varphi(f) = 1$, each
term must contribute its maximum: $(-1)^n f(\frac1n) = 1$ for every
$n$ (otherwise the strict deficit of one term cannot be compensated,
all terms being $\leq 2^{-n}$). So $f(\frac1n) = (-1)^n$; but
$\frac1n \to 0$ and $f$ is \omterm{def:b2:metric:continuity}{continuous} at $0$, forcing the
contradictory convergence of $(-1)^n$. Hence the supremum is not a
maximum --- impossible in finite dimension, where the closed unit
ball is \omterm{def:b2:metric:compact}{compact}.
\end{solution}

\begin{solution}{exo:b2:nvs:9}
\emph{Bounded sequences:} a bounded sequence lies in some closed
ball $\overline B(0, R) = R\,\overline B(0,1)$, \omterm{def:b2:metric:compact}{compact} (image of
the \omterm{def:b2:metric:compact}{compact} unit ball under the homeomorphism $x \mapsto Rx$):
extract there.

\emph{Linear forms:} if $\dim E < \infty$, every linear map from
$E$ is \omterm{def:b2:metric:continuity}{continuous} (\cref{thm:b2:nvs:finitedim}). Conversely,
suppose $\dim E = \infty$ (which, by \cref{thm:b2:nvs:riesz},
is actually excluded by the \omterm{def:b2:metric:compact}{compactness} hypothesis --- the point of
this question is the implication between the two properties in
general normed spaces): choose a normalized linearly independent
sequence $(e_n)$, complete it into an algebraic basis (admitted),
and define $\varphi(e_n) = n$, $\varphi = 0$ on the other basis
vectors, extended linearly. Then $\abs{\varphi(e_n)} = n$ with
$\norm{e_n} = 1$: unbounded on the unit ball, discontinuous. So
``all forms \omterm{def:b2:metric:continuity}{continuous}'' forces finite dimension.
\end{solution}

\begin{solution}{exo:b2:nvs:10}
Cauchy--Schwarz with the constant function $1$: $\norm f_1 =
\int_0^1 \abs f\cdot 1 \leq \bigl(\int_0^1
f^2\bigr)^{1/2}\bigl(\int_0^1 1\bigr)^{1/2} = \norm f_2$. And
$\norm f_2^2 = \int f^2 \leq \norm f_\infty^2$. For $f_n(x) =
x^n$:
\[
\norm{f_n}_1 = \frac1{n+1}, \qquad
\norm{f_n}_2 = \frac1{\sqrt{2n+1}}, \qquad
\norm{f_n}_\infty = 1 .
\]
Then $\norm{f_n}_2/\norm{f_n}_1 = \frac{n+1}{\sqrt{2n+1}} \to
\infty$ and $\norm{f_n}_\infty/\norm{f_n}_2 = \sqrt{2n+1} \to
\infty$: no reverse inequalities, no pair is \omterm{def:b2:nvs:equivalent}{equivalent}.
\end{solution}

\begin{solution}{exo:b2:nvs:11}
\emph{Lower bound for the distance:} for $h \in \ker\varphi$,
$\abs{\varphi(x)} = \abs{\varphi(x - h)} \leq
\vertiii\varphi\,\norm{x - h}$; take the infimum over $h$:
$d(x, \ker\varphi) \geq \abs{\varphi(x)}/\vertiii\varphi$.

\emph{Upper bound:} we may assume $\varphi(x) \neq 0$. Given
$\varepsilon > 0$, pick a unit $u$ with $\abs{\varphi(u)} \geq
\vertiii\varphi - \varepsilon > 0$ and set $h = x -
\frac{\varphi(x)}{\varphi(u)}\,u$: then $\varphi(h) = 0$ and
\[
\norm{x - h} = \frac{\abs{\varphi(x)}}{\abs{\varphi(u)}}
\leq \frac{\abs{\varphi(x)}}{\vertiii\varphi - \varepsilon}.
\]
Let $\varepsilon \to 0$: $d(x, \ker\varphi) \leq
\abs{\varphi(x)}/\vertiii\varphi$; equality.

\emph{Non-attainment:} take $\varphi$ from \cref{exo:b2:nvs:8}
($\vertiii\varphi = 1$, not attained) and any $x$ with
$\varphi(x) \neq 0$. If some $h \in \ker\varphi$ realized
$\norm{x - h} = \abs{\varphi(x)}$, the unit vector $v = (x -
h)/\norm{x - h}$ would satisfy $\abs{\varphi(v)} =
\abs{\varphi(x)}/\norm{x - h} = 1 = \vertiii\varphi$: the
\omterm{thm:b2:nvs:continuouslinear}{operator norm} would be attained --- contradiction.
\end{solution}

\begin{solution}{exo:b2:nvs:12}
$N_1 \leq N_2$ is monotonicity of the sup in the domain, so the
identity $(E, N_2) \to (E, N_1)$ is $1$-Lipschitz. For $P_n(x) =
(x/2)^n$: $N_2(P_n) = 1$ (attained at $x = 2$) while $N_1(P_n) =
2^{-n}$: a bound $N_2 \leq CN_1$ would give $1 \leq C2^{-n}$ for
all $n$: impossible. So the \omterm{def:b2:nvs:norm}{norms} are not \omterm{def:b2:nvs:equivalent}{equivalent} and the
inverse identity is a discontinuous linear bijection.

\emph{Incompleteness:} let $S_n = \sum_{k=0}^{n}\frac{X^k}{k!}$.
For $m > n$, $N_1(S_m - S_n) \leq \sum_{k>n}\frac1{k!} \to 0$:
Cauchy for $N_1$. If $S_n \to P$ in $(E, N_1)$, then pointwise
$P(x) = \lim S_n(x) = \eu^x$ on $\intcc{0}{1}$; but a polynomial
of degree $d$ cannot equal $\eu^x$ on an interval (differentiate
$d + 1$ times: the left side dies, $\eu^x$ does not). No limit in
$E$: not \omterm{def:b2:metric:complete}{complete}.

In finite dimension all three phenomena are impossible: all \omterm{def:b2:nvs:norm}{norms}
are \omterm{def:b2:nvs:equivalent}{equivalent}, every normed space is \omterm{def:b2:metric:complete}{complete}, and the inverse
of a linear bijection is linear from a finite-dimensional space,
hence \omterm{def:b2:metric:continuity}{continuous} (\cref{thm:b2:nvs:finitedim}).
\end{solution}

\begin{solution}{pb:b2:nvs:1}
\textbf{1.} The candidates worth considering form $K = \{f \in F
: \norm{x - f} \leq \norm x\}$: nonempty ($0 \in K$), closed
(preimage of a closed interval under the \omterm{def:b2:metric:continuity}{continuous} $f \mapsto
\norm{x - f}$, intersected with the closed $F$,
\cref{cor:b2:nvs:closedsubspace}), bounded ($\norm f \leq
\norm{f - x} + \norm x \leq 2\norm x$). In the
finite-dimensional $F$, closed and bounded means \omterm{def:b2:metric:compact}{compact}
(\cref{thm:b2:nvs:finitedim}); the \omterm{def:b2:metric:continuity}{continuous} function $f
\mapsto \norm{x - f}$ attains on $K$ its infimum, which equals
the infimum over all of $F$ (any $f \notin K$ gives $\norm{x -
f} > \norm x \geq \inf$).

\textbf{2.} Parallelogram identity in $(\R^n,
\norm\cdot_2)$: $\norm{u + v}^2 + \norm{u - v}^2 = 2\norm u^2 +
2\norm v^2$ (expand the squares of sums of coordinates). For
unit $u \neq v$:
\[
\Bigl\Vert\frac{u+v}2\Bigr\Vert^2 = 1 - \frac{\norm{u -
v}^2}{4} < 1 .
\]
Not strictly convex: for $\norm\cdot_\infty$, take $u = (1, 1, 0,
\dots)$, $v = (1, -1, 0, \dots)$: unit vectors with midpoint
$(1, 0, \dots)$ of \omterm{def:b2:nvs:norm}{norm} $1$; for $\norm\cdot_1$, take $u = (1,
0, \dots)$, $v = (0, 1, 0, \dots)$: midpoint $(\frac12,
\frac12, 0, \dots)$ of \omterm{def:b2:nvs:norm}{norm} $1$.

\textbf{3.} Let $d = d(x, F)$. If $d = 0$: $x \in \overline F =
F$ and the only minimizer is $x$. If $d > 0$ and $f_1 \neq f_2$
both minimize: $u = \frac{x - f_1}{d}$ and $v = \frac{x -
f_2}{d}$ are distinct unit vectors, so
\[
\Bigl\Vert x - \frac{f_1 + f_2}2\Bigr\Vert
= d\,\Bigl\Vert\frac{u + v}2\Bigr\Vert < d ,
\]
with $\frac{f_1 + f_2}2 \in F$: contradicts the definition of
$d$. Unique minimizer.

\textbf{4.} $\norm{(0,1) - t(1,0)}_\infty = \max(\abs t, 1)
\geq 1$, with equality iff $\abs t \leq 1$: the minimizers form
the segment $\{t(1, 0) : t \in \intcc{-1}{1}\}$, all at distance
$1$ --- uniqueness fails exactly because the square ball has
flat sides (question 2).

\textbf{5.} Let $M = \max f$, $m = \min f$ (attained:
\omterm{def:b2:metric:compact}{compactness}). For any constant $c$: $\sup\abs{f - c} \geq
\max(M - c,\, c - m) \geq \frac{M - m}2$, the last inequality
because the two quantities average to $\frac{M-m}2$; equality in
both forces $M - c = c - m$, i.e.\ $c = c^* = \frac{M + m}2$.
Conversely $\sup\abs{f - c^*} = \max(M - c^*, c^* - m) =
\frac{M - m}2$. Unique best constant. For $f(x) = x^2$ on
$\intcc01$: $c^* = \frac12$, distance $\frac12$.

\textbf{6.} From $\cos(n{+}1)\theta + \cos(n{-}1)\theta =
2\cos\theta\cos n\theta$: the polynomials defined by $T_0 = 1$,
$T_1 = X$, $T_{n+1} = 2XT_n - T_{n-1}$ satisfy $T_n(\cos\theta)
= \cos n\theta$ by induction. Uniqueness: two polynomials
agreeing on $\intcc{-1}{1}$ (infinitely many points) are equal.
Induction again: $\deg T_n = n$ with leading coefficient
$2^{n-1}$ for $n \geq 1$ ($T_1$: coefficient $1 = 2^0$; the
recurrence doubles it).

\textbf{7.} Every $x \in \intcc{-1}{1}$ is $\cos\theta$, and
$\abs{\cos n\theta} \leq 1$. At $\eta_k = \cos\frac{k\pi}n$:
$T_n(\eta_k) = \cos k\pi = (-1)^k$, and $1 = \eta_0 > \eta_1 >
\dots > \eta_n = -1$. Roots: $\cos n\theta = 0$ iff $\theta =
\frac{(2k-1)\pi}{2n}$: the $n$ distinct points
$\cos\frac{(2k-1)\pi}{2n}$, and since $\frac{(k-1)\pi}n <
\frac{(2k-1)\pi}{2n} < \frac{k\pi}n$, each root lies strictly
between two consecutive extrema.

\textbf{8.} $T_2 = 2X^2 - 1$, $T_3 = 4X^3 - 3X$, $T_4 = 8X^4 -
8X^2 + 1$. For $T_3$: $T_3(1) = 1$, $T_3(\tfrac12) = \tfrac12 -
\tfrac32 = -1$, $T_3(-\tfrac12) = 1$, $T_3(-1) = -1$: perfect
alternation.

\textbf{9.} Let $u_\pm = x \pm \sqrt{x^2 - 1}$ for $x \geq 1$:
the roots of $z^2 - 2xz + 1$, with $u_+u_- = 1$. The sequence
$s_n = \frac{u_+^n + u_-^n}2$ satisfies $s_{n+1} = 2x\,s_n -
s_{n-1}$ (Newton-type recurrence from the quadratic), $s_0 = 1$,
$s_1 = x$: the same recurrence and initial values as $n \mapsto
T_n(x)$, so $s_n = T_n(x)$ for all $n$. Since $0 < u_- \leq 1
\leq u_+$ with $u_+ > 1$ for $x > 1$: $T_n(x) \geq
\frac{u_+^n}2 \to \infty$ and $T_n(x) \sim \frac12\bigl(x +
\sqrt{x^2-1}\bigr)^n$. (For $x \leq -1$ use the parity $T_n(-x)
= (-1)^nT_n(x)$, clear from the recurrence.)

\textbf{10.} For every $\theta$: $T_m\bigl(T_n(\cos\theta)\bigr)
= T_m(\cos n\theta) = \cos mn\theta =
T_{mn}(\cos\theta)$. The polynomials $T_m \circ T_n$ and
$T_{mn}$ agree on $\intcc{-1}{1}$, hence are equal.

\textbf{11.} $D = Q_n - P$ has degree $\leq n - 1$ (the monic
leading terms cancel). At the extrema: $(-1)^kD(\eta_k) =
2^{1-n} - (-1)^kP(\eta_k) \geq 2^{1-n} - \abs{P(\eta_k)} > 0$ by
hypothesis. So $D$ takes nonzero values of alternating signs at
the $n + 1$ decreasing points $\eta_0 > \dots > \eta_n$: by the
intermediate value theorem it has at least $n$ distinct roots,
one in each \omterm{def:b2:metric:topology}{open} interval $\intoo{\eta_{k}}{\eta_{k-1}}$. A
nonzero polynomial of degree $\leq n - 1$ cannot have $n$ roots;
and $D = 0$ contradicts the strict signs. Contradiction:
$\sup\abs P \geq 2^{1-n}$ for every monic $P$ of degree $n$.

\textbf{12.} Now $(-1)^kD(\eta_k) = 2^{1-n} - (-1)^kP(\eta_k)
\geq 0$ since $\abs{P} \leq 2^{1-n}$. On each
$\intcc{\eta_k}{\eta_{k-1}}$ ($k = 1, \dots, n$) the endpoint
values of $D$ have opposite weak signs: the intermediate value
theorem yields a zero $z_k$ in the closed interval. If the $z_k$
can be chosen pairwise distinct, $D \neq 0$ of degree $\leq n-1$
has $n$ roots: contradiction. Two consecutive intervals can only
share the zero $z_k = z_{k+1} = \eta_k$ with $0 < k < n$
(interior). There, $D(\eta_k) = 0$ means $P(\eta_k) =
(-1)^k2^{1-n}$, an extremal value of $P$ on $\intcc{-1}{1}$
attained at an \emph{interior} point: $P'(\eta_k) = 0$; and
$\eta_k$ is an interior extremum of $T_n$ too: $Q_n'(\eta_k) =
0$. Hence $D'(\eta_k) = 0$: $\eta_k$ is a root of multiplicity
$\geq 2$, compensating the shared interval. In all cases $D$ has
at least $n$ roots counted with multiplicity, degree $\leq n -
1$, so $D = 0$: $P = Q_n$. Chebyshev's extremal theorem is
proved: the unique monic minimizer is $2^{1-n}T_n$, of sup \omterm{def:b2:nvs:norm}{norm}
$2^{1-n}$.

\textbf{13.} Monic polynomials of degree $n$ are exactly the
$X^n - R$ with $R \in \R_{n-1}[X]$, so
\[
d_\infty\bigl(X^n, \R_{n-1}[X]\bigr) = \min_{P \text{
monic}}\ \sup_{\intcc{-1}{1}}\abs P = 2^{1-n},
\]
uniquely at $R^* = X^n - Q_n$. Substitution $x = \frac{1+t}2$:
if $P$ is monic of degree $n$ on $\intcc01$, then $t \mapsto
2^nP\bigl(\frac{1+t}2\bigr)$ is monic on $\intcc{-1}{1}$ with
sup equal to $2^n\sup_{\intcc01}\abs P$: hence
$\sup_{\intcc01}\abs P \geq 2^{-n}\cdot2^{1-n} = 2^{1-2n}$, with
equality exactly for $P^*(x) = 2^{-n}Q_n(2x - 1)$: on
$\intcc{0}{1}$ the distance is $2^{1-2n}$.

\textbf{14.} $\omega$ is monic of degree $n$, so
$\sup_{\intcc{-1}{1}}\abs\omega \geq 2^{1-n}$ with equality iff
$\omega = Q_n = 2^{1-n}T_n$, i.e.\ iff the nodes are the $n$
roots of $T_n$: $x_k = \cos\frac{(2k-1)\pi}{2n}$ --- the
\emph{Chebyshev nodes}. Minimal value: $2^{1-n}$. Equidistant
nodes are strictly worse; the interpolation error factor is
minimized by clustering nodes near the endpoints.

\textbf{15.} $n = 2$ by hand: $x^2$ ranges over $\intcc01$, so
$\sup_{\intcc{-1}{1}}\abs{x^2 - c} = \max(\abs c, \abs{1 - c})
\geq \frac12$, minimized at $c = \frac12$: minimal monic
quadratic $x^2 - \frac12 = \frac12(2x^2 - 1) = Q_2$, value
$\frac12 = 2^{1-2}$. For $n = 10$ on $\intcc{0}{1}$: $2^{1-20} =
2^{-19} \approx 1.9\cdot10^{-6}$. Some polynomial of degree $9$
stays within two millionths of $x^{10}$ on the whole of
$\intcc01$: at this scale the two graphs are indistinguishable
--- the flatness of $x^{10}$ near $0$ lets lower degrees do all
the work.

\textbf{16.} $\norm{g_k}_\infty = g_k(1) = 1$. For $j > k$ set
$a = 2^k$, $b = 2^j \geq 2a$, and evaluate at $x_0 = 2^{-1/a}$
(so $x_0^a = \frac12$):
\[
g_k(x_0) - g_j(x_0) = \frac12 - \Bigl(\frac12\Bigr)^{b/a}
\geq \frac12 - \frac14 = \frac14 .
\]
So $\norm{g_k - g_j}_\infty \geq \frac14$ for all $j \neq k$: no
subsequence is Cauchy, none converges. The closed unit ball of
$\bigl(C(\intcc01), \norm\cdot_\infty\bigr)$ is not \omterm{def:b2:metric:compact}{compact}.

\textbf{17.} $F$ is closed (\cref{cor:b2:nvs:closedsubspace})
and proper: pick $y \notin F$, so $\delta = d(y, F) > 0$. By
question 1 the distance is attained at some $f^* \in F$. Set $x
= \frac{y - f^*}{\delta}$, a unit vector ($\norm{y - f^*} =
\delta$). For every $g \in F$:
\[
\norm{x - g} = \frac{\norm{y - (f^* + \delta g)}}{\delta}
\geq \frac{\delta}{\delta} = 1 ,
\]
since $f^* + \delta g \in F$. Hence $d(x, F) \geq 1$; and $d(x,
F) \leq \norm{x - 0} = 1$: exactly $1$.

\textbf{18.} In an infinite-dimensional $E$, build unit vectors
inductively: $x_1$ arbitrary; given $x_1, \dots, x_k$, the
subspace $F_k = \operatorname{Vect}(x_1, \dots, x_k)$ is
finite-dimensional, hence proper, and question 17 gives a unit
$x_{k+1}$ with $d(x_{k+1}, F_k) = 1$: in particular
$\norm{x_{k+1} - x_i} \geq 1$ for $i \leq k$. The sequence has
pairwise distances $\geq 1$: the unit ball contains a sequence
with no convergent subsequence, hence is not \omterm{def:b2:metric:compact}{compact} ---
Riesz's theorem, with the sharp constant $1$.

\textbf{19.} Let $h_n$ be affine on each half of
$\bigl[\frac1{n+1}, \frac1n\bigr]$, from $0$ up to $1$ at the
midpoint and back to $0$, and zero elsewhere: \omterm{def:b2:metric:continuity}{continuous},
$\norm{h_n}_\infty = 1$. For $n \neq m$ the supports meet at
most at a common endpoint, where both vanish; at the peak of
$h_n$, $h_m = 0$: $\norm{h_n - h_m}_\infty = 1$ exactly. For
fixed $x > 0$: $h_n(x) = 0$ as soon as $\frac1n < x$, and
$h_n(0) = 0$ always: $h_n \to 0$ pointwise; but
$\norm{h_n - 0}_\infty = 1$: not uniformly. An explicit
constellation at mutual distance $1$ inside the unit ball.

\textbf{20.} A ball of radius $\frac13$ has diameter $\leq
\frac23 < 1$, so it contains at most one of the $h_n$ (two are
at distance $1$). Finitely many such balls contain finitely many
of the infinitely many $h_n$: they cannot cover the unit ball.
Total boundedness --- which \omterm{def:b2:metric:compact}{compact metric spaces} enjoy, by the
proof of \cref{thm:b2:metric:borellebesgue} --- fails as badly
as possible.

\textbf{21.} If $Ax = \lambda x$ with $x \neq 0$:
$\abs\lambda\,\norm x = \norm{Ax} \leq \vertiii A\,\norm x$,
so $\abs\lambda \leq \vertiii A$. For
$\begin{psmallmatrix}1 & -2\\ 3 & 1\end{psmallmatrix}$:
$\vertiii A_\infty = \max(1 + 2,\ 3 + 1) = 4$
(\cref{exo:b2:nvs:4}), so every \omterm{def:b2:reduction:eigen}{eigenvalue} has modulus $\leq
4$; in fact $\chi_A = X^2 - 2X + 7$ gives $\lambda = 1 \pm
\iu\sqrt6$ of modulus $\sqrt7 \approx 2.65$: the bound is
valid, not sharp.

\textbf{22.} $N_P$ is a \omterm{def:b2:nvs:norm}{norm}: $N_P(x) = 0$ forces $P^{-1}x = 0$,
so $x = 0$; homogeneity and triangle inequality are inherited
from $\norm\cdot_\infty$ through the linear $P^{-1}$. \omterm{thm:b2:nvs:continuouslinear}{Operator
norm}: with $y = P^{-1}x$ and $D = \mathrm{diag}(\lambda_i)$,
\[
N_P(Ax) = \norm{P^{-1}AP\,y}_\infty = \norm{Dy}_\infty,
\]
so $\vertiii A_{N_P}$ is the $\norm\cdot_\infty$-operator \omterm{def:b2:nvs:norm}{norm}
of $D$, which is its largest absolute row sum
(\cref{exo:b2:nvs:4}): $\max_i\abs{\lambda_i}$.

\textbf{23.} If all $\abs{\lambda_i} < 1$: $N_P(A^kx) \leq
\rho^kN_P(x)$ with $\rho = \max\abs{\lambda_i} < 1$, so $A^kx
\to 0$ for every $x$, and $A^k \to 0$ in any \omterm{def:b2:nvs:norm}{norm} on
$\mathcal{M}_n$ (all \omterm{def:b2:nvs:equivalent}{equivalent} in finite dimension,
\cref{thm:b2:nvs:finitedim}; convergence of $A^ke_j$ for each
$j$ is entrywise convergence). If some $\abs{\lambda} \geq 1$
with \omterm{def:b2:reduction:eigen}{eigenvector} $x$: $\norm{A^kx} = \abs\lambda^k\norm x
\not\to 0$. For $A = \frac14\begin{psmallmatrix}1 & 2\\ 2 &
1\end{psmallmatrix}$: \omterm{def:b2:reduction:eigen}{eigenvalues} $\frac14(1 \pm 2) =
\frac34, -\frac14$, both of modulus $< 1$: $A^k \to 0$.

\textbf{24.} Both are \omterm{def:b2:nvs:norm}{norms} on the finite-dimensional
$\R_n[X]$, hence \omterm{def:b2:nvs:equivalent}{equivalent} for each $n$. Take the minimal
monic $P^*_n$ of question 13 on $\intcc01$: its coefficient of
$X^n$ is $1$, so $N_c(P^*_n) \geq 1$, while
$\norm{P^*_n}_{\intcc01} = 2^{1-2n}$. Hence
\[
C_n \geq \frac{N_c(P^*_n)}{\norm{P^*_n}_{\intcc01}}
\geq 2^{2n-1} .
\]
The equivalence constants explode with the dimension: on the
union $\R[X]$ no single constant serves, which is exactly the
non-equivalence seen in \cref{exo:b2:nvs:12} --- ``all \omterm{def:b2:nvs:norm}{norms}
are \omterm{def:b2:nvs:equivalent}{equivalent}'' is a theorem about one dimension at a time,
and infinite dimension is where it dies.

\textbf{25.} \omterm{def:b2:metric:compact}{Compactness} of closed bounded sets in finite
dimension produced existence of best approximations (question
1) and powered the zero-counting at the $n + 1$ extremal points
(questions 11--12, via attained sups). Strict convexity governs
uniqueness of best approximation --- round balls give one
minimizer, flat-sided balls give segments of them (questions
2--4). The constellation of unit vectors at mutual distance
$1$ (questions 17--19) destroys \omterm{def:b2:metric:compact}{compactness} of the unit ball
and total boundedness with it (question 20). Question 24
quantifies the collapse: the constants relating two \omterm{def:b2:nvs:norm}{norms} on
$\R_n[X]$ grow like $4^n$, so no uniform comparison survives
the passage to $\R[X]$. The summit is Chebyshev's extremal
theorem (questions 11--12): the unique monic minimizer
$2^{1-n}T_n$. Best approximation finds its modern home in
Hilbert spaces, where the projection theorem replaces
\omterm{def:b2:metric:compact}{compactness} by \omterm{def:b2:metric:complete}{completeness} plus the parallelogram identity ---
proved honestly in the Year 3 volume.
\end{solution}
