\section*{Chapter \ref{ch:b2:funcseq} --- Sequences and Series of
Functions}

\begin{solution}{exo:b2:funcseq:1}
$f_n(x) = \frac{x}{1 + nx}$: pointwise limit $0$ on $\intcc{0}{1}$.
Uniformly: $f_n$ increases on $\intcc{0}{1}$ (derivative
$\frac{1}{(1+nx)^2} > 0$), so $\norm{f_n}_\infty = f_n(1) =
\frac{1}{1+n} \to 0$: \emph{uniform} on $\intcc{0}{1}$.

$g_n(x) = nx\,\eu^{-nx^2}$: pointwise limit $0$ (exponential
beats). Sup: $g_n' = n\eu^{-nx^2}(1 - 2nx^2)$ vanishes at $x_n =
\frac{1}{\sqrt{2n}}$, where $g_n(x_n) = \sqrt{\frac n2}\,\eu^{-1/2}
\to \infty$: not uniform on $\intcc{0}{1}$ --- but uniform on
$\intco{\delta}{1}$, since there $g_n(x) \leq n\,\eu^{-n\delta^2}
\to 0$.

$h_n(x) = x^n(1 - x^n)$: pointwise limit $0$ on $\intcc{0}{1}$
(both factors; at $x = 1$, $h_n = 0$). Sup: with $u = x^n \in
\intcc{0}{1}$, $u(1-u) \leq \frac14$ attained at $u = \frac12$,
i.e.\ $x = 2^{-1/n} \in \intoo{0}{1}$: $\norm{h_n}_\infty =
\frac14 \not\to 0$: not uniform on $\intcc{0}{1}$; uniform on
$\intcc{0}{a}$ ($\sup \leq a^n \to 0$).
\end{solution}

\begin{solution}{exo:b2:funcseq:2}
$\int_0^1 nx\,\eu^{-nx^2}\dd x = \bigl[-\tfrac12
\eu^{-nx^2}\bigr]_0^1 = \frac{1 - \eu^{-n}}{2} \to \frac12$, while
$\int_0^1 \lim g_n = 0$. No contradiction:
\cref{thm:b2:funcseq:integration} requires \emph{uniform}
convergence on the segment, which fails here (the bump of height
$\sim\sqrt n$ slides toward $0$).
\end{solution}

\begin{solution}{exo:b2:funcseq:3}
Normal convergence on $\intcc{-1}{1}$: $\norm{x^n/n^2}_\infty =
\frac{1}{n^2}$, summable: $S$ is continuous there
(\cref{thm:b2:funcseq:seriestransfer}).

Derivative: the derived series $\sum \frac{x^{n-1}}{n}$ converges
normally on every $\intcc{-a}{a}$, $a < 1$ ($\sup =
\frac{a^{n-1}}{n}$): $S$ is $C^1$ on $\intoo{-1}{1}$ with
\[
S'(x) = \sum_{n\geq1} \frac{x^{n-1}}{n} = \frac1x \sum_{n\geq1}
\frac{x^n}{n} = -\frac{\ln(1 - x)}{x}
\qquad (0 < \abs x < 1),
\]
the last identity being the Year 1 logarithmic series (re-derived
honestly in \cref{ch:b2:powerseries}).
\end{solution}

\begin{solution}{exo:b2:funcseq:4}
For fixed $x > 0$ the series is alternating with $\frac{1}{n + x}
\downarrow 0$: pointwise convergence, and the remainder bound
$\abs{R_N(x)} \leq \frac{1}{N + 1 + x} \leq \frac{1}{N+1}$ is
\emph{uniform} on $\intco{\delta}{\infty}$ (indeed on
$\intoo{0}{\infty}$): uniform convergence. (Not normal:
$\norm{\frac{(-1)^n}{n+x}}_\infty = \frac{1}{n + \delta}$,
divergent.) Continuity follows from
\cref{thm:b2:funcseq:seriestransfer}.

Functional equation: reindex $F(x + 1)$ with $m = n + 1$:
\[
F(x+1) = \sum_{n\geq0} \frac{(-1)^n}{n + 1 + x}
= \sum_{m\geq1}\frac{(-1)^{m-1}}{m+x} ,
\]
so, isolating the $m = 0$ term of $F(x)$,
\[
F(x) + F(x+1)
= \frac{1}{x} + \sum_{m\geq1}
\frac{(-1)^m + (-1)^{m-1}}{m+x} = \frac1x .
\]
\end{solution}

\begin{solution}{exo:b2:funcseq:5}
Set $g_n = f_n - f \geq 0$ (decreasing in $n$, by hypothesis; the
limit is $0$ pointwise); each $g_n$ is continuous. Fix $\varepsilon
> 0$ and let $U_n = \{x : g_n(x) < \varepsilon\}$: open (preimage
of an open set), increasing ($g_{n+1} \leq g_n$), and covering $K$
(pointwise convergence). By Borel--Lebesgue
(\cref{thm:b2:metric:borellebesgue}), finitely many $U_{n_1}
\subseteq \dots \subseteq U_{n_k}$ cover $K$: so $K = U_{n_k}$,
i.e.\ $\norm{g_{n_k}}_\infty \leq \varepsilon$, and by monotonicity
$\norm{g_n}_\infty \leq \varepsilon$ for all $n \geq n_k$: uniform
convergence. (Monotonicity is essential: the sliding bumps of
\cref{exo:b2:funcseq:2} converge pointwise on a compact without
uniformity.)
\end{solution}

\begin{solution}{exo:b2:funcseq:6}
Substitute $u = nx$:
\[
\int_0^1 \frac{n f(x)}{1 + n^2x^2}\dd x
= \int_0^n \frac{f(u/n)}{1 + u^2}\,\dd u .
\]
The integrands $h_n(u) = \frac{f(u/n)}{1+u^2}\mathbf{1}_{u \leq n}$
converge pointwise to $\frac{f(0)}{1+u^2}$ (continuity of $f$ at
$0$) and are dominated by $\frac{\norm f_\infty}{1 + u^2}$,
integrable on $\intco{0}{\infty}$: dominated convergence
(\cref{thm:b2:integration:dominated}) gives the limit
\[
\int_0^\infty \frac{f(0)}{1 + u^2}\dd u = \frac{\pi}{2} f(0) .
\]
(The kernels concentrate at $0$: an approximate identity.)
\end{solution}

\begin{solution}{exo:b2:funcseq:7}
For $f(x) = x^2$, use the second binomial identity family from the
proof: $\sum_k k^2 p_k = n(n-1)x^2 + nx$. Hence
\[
B_n(f)(x) = \sum_k \frac{k^2}{n^2}\,p_k
= \frac{n(n-1)x^2 + nx}{n^2}
= x^2 + \frac{x(1 - x)}{n} :
\]
$B_n(f) - f = \frac{x(1-x)}{n}$, of sup norm $\frac{1}{4n} =
O\bigl(\frac1n\bigr)$, as predicted.
\end{solution}

\begin{solution}{exo:b2:funcseq:8}
For $\varepsilon = 1$ there is $N$ with $\norm{P_n - P_m}_{\infty,
\R} \leq 1$ for $m, n \geq N$. A polynomial bounded on $\R$ is
constant (a nonconstant one tends to $\pm\infty$): $P_n - P_m =
c_{n,m}$, constants. So for $n \geq N$: $P_n = P_N + c_n$ with
$c_n = P_n(0) - P_N(0)$ convergent (pointwise convergence at $0$).
Hence $f = \lim P_n = P_N + \lim c_n$: a polynomial.
\end{solution}

\begin{solution}{exo:b2:funcseq:9}
(a) $\norm{(3/4)^n\varphi(4^n\cdot)}_\infty = \frac12
(3/4)^n$: normal convergence, so $W$ is continuous
(\cref{thm:b2:funcseq:seriestransfer}).

(b) Fix $x$, $m$; choose the sign of $h_m = \pm\frac12 4^{-m}$ so
that the segment $\intcc{4^mx}{4^m(x + h_m)}$ (of length $\frac12$)
contains no half-integer, making $\varphi$ affine of slope $\pm1$
on it (possible: an interval of length $\frac12$ meets at most one
half-integer point; pick the side avoiding it).

For $n > m$: $4^n h_m = \pm\frac12 4^{n-m}$ is an integer, and
$\varphi$ is $1$-periodic: the $n$-th term of the difference
vanishes.

For $n = m$: $\abs{\varphi(4^m x + 4^m h_m) - \varphi(4^m x)} =
\abs{4^m h_m} = \frac12$ ($\varphi$ affine of slope $\pm 1$ on the
segment), so the term contributes exactly $(3/4)^m \cdot
\frac{1/2}{\abs{h_m}} = (3/4)^m\,4^m = 3^m$ in the quotient.

For $n < m$: the $1$-Lipschitz $\varphi$ gives
$\bigl|(3/4)^n\bigl(\varphi(4^nx + 4^nh_m) -
\varphi(4^nx)\bigr)\bigr| \leq (3/4)^n 4^n\abs{h_m} =
3^n\abs{h_m}$: each contributes at most $3^n$ to the quotient.

Hence
\[
\Bigl|\frac{W(x + h_m) - W(x)}{h_m}\Bigr|
\geq 3^m - \sum_{n=0}^{m-1} 3^n
= 3^m - \frac{3^m - 1}{2} = \frac{3^m + 1}{2}
\longrightarrow \infty .
\]
If $W$ were differentiable at $x$, every difference quotient along
$h_m \to 0$ would converge to $W'(x)$: contradiction. $W$ is
continuous everywhere, differentiable nowhere.
\end{solution}

\begin{solution}{exo:b2:funcseq:10}
Pointwise: for $x \in \intco{0}{1}$ the series is geometric of
ratio $-x$,
\[
\sum_{n\geq0}(-1)^n x^n(1-x) = \frac{1-x}{1+x},
\]
and at $x = 1$ every term vanishes: sum $0 = \frac{1-1}{2}$,
consistent --- the sum is continuous on $\intcc{0}{1}$.
Uniformity: the remainder is a geometric tail,
\[
\abs{R_N(x)} = \frac{x^{N+1}(1-x)}{1+x} \leq x^{N+1}(1 - x)
\leq \max_{\intcc01} t^{N+1}(1-t)
= \frac{1}{N+2}\Bigl(\frac{N+1}{N+2}\Bigr)^{\!N+1}
\leq \frac{1}{N+2} \to 0 ,
\]
uniformly in $x$. Not normal: $\norm{u_n}_\infty =
\max x^n(1-x) = \frac{1}{n+1}\bigl(\frac{n}{n+1}\bigr)^n \sim
\frac{1}{\eu\,n}$, and $\sum \frac1{\eu n}$ diverges. Contrast:
$\sum x^n(1-x)$ has partial sums $1 - x^{N+1}$, converging
pointwise to the \emph{discontinuous} $\mathbf 1_{\intco01}$: by
\cref{thm:b2:funcseq:continuity}, that convergence cannot be
uniform on $\intcc{0}{1}$.
\end{solution}

\begin{solution}{exo:b2:funcseq:11}
The limit $f$ is continuous (\cref{thm:b2:funcseq:continuity}).
Then
\[
\abs{f_n(x_n) - f(x)}
\leq \abs{f_n(x_n) - f(x_n)} + \abs{f(x_n) - f(x)}
\leq \norm{f_n - f}_\infty + \abs{f(x_n) - f(x)} ,
\]
and both terms tend to $0$ (uniform convergence; continuity of
$f$ at $x$). Counterexample under mere pointwise convergence:
$g_n(x) = nx\,\eu^{-nx^2} \to 0$ pointwise on $\intcc{0}{1}$
with $g_n$ and the limit continuous, yet at $x_n =
\frac{1}{\sqrt{2n}} \to 0$:
\[
g_n(x_n) = \sqrt{\frac n2}\,\eu^{-1/2} \longrightarrow +\infty
\neq 0 = f(0) .
\]
\end{solution}

\begin{solution}{exo:b2:funcseq:12}
\begin{enumerate}
  \item Induction. $n = 1$ is the definition. Assume the formula
        for $n$ and set $g(x) = \int_0^x \frac{(x-t)^n}{n!}
        f(t)\dd t$. For an integrand continuous in $(x,t)$ and
        $C^1$ in $x$, the variable-limit parameter integral
        differentiates as
        \[
        g'(x) = \frac{(x-x)^n}{n!}f(x)
        + \int_0^x \frac{(x-t)^{n-1}}{(n-1)!}f(t)\dd t
        = T^nf(x)
        \]
        (split $g(x+h) - g(x)$ into the strip
        $\int_x^{x+h}$, which is $O(h\cdot\sup)$ with the
        integrand vanishing at $t = x$ like $h^n$, and the fixed
        integral of the $x$-increment, handled by the mean value
        inequality and continuity). Also $(T^{n+1}f)' = T^nf$
        (fundamental theorem of calculus) and $g(0) =
        T^{n+1}f(0) = 0$: two primitives of $T^nf$ vanishing at
        $0$ coincide, so $T^{n+1}f = g$. The bound:
        \[
        \abs{T^nf(x)} \leq \norm f_\infty
        \int_0^x \frac{(x-t)^{n-1}}{(n-1)!}\dd t
        = \norm f_\infty\,\frac{x^n}{n!}
        \leq \frac{\norm f_\infty}{n!} .
        \]
  \item $\sum_n \norm{T^nf}_\infty \leq \eu\,\norm f_\infty$:
        normal, hence uniform, convergence; $S$ is continuous.
        The partial sums satisfy $S_N = f + T S_{N-1}$, and $T$
        is $1$-Lipschitz for $\norm\cdot_\infty$
        ($\abs{Tg(x)} \leq x\norm g_\infty$): letting $N \to
        \infty$ in both sides gives $S = f + TS$.
  \item Set $V(x) = f(x) + \eu^x\int_0^x \eu^{-t}f(t)\dd t$.
        Then $V - f$ is $C^1$ with $(V-f)'(x) =
        \eu^x\int_0^x\eu^{-t}f + f(x) = V(x)$, and $(TV)' = V$
        with $(V - f)(0) = TV(0) = 0$: so $V - f = TV$, i.e.\
        $V$ solves the equation. Uniqueness: if $S_1, S_2$ are
        continuous solutions, $D = S_1 - S_2$ satisfies $D =
        TD$, hence $D = T^nD$ for all $n$ and $\norm D_\infty
        \leq \frac{\norm D_\infty}{n!} \to 0$: $D = 0$.
        Therefore
        \[
        \sum_{n\geq0} T^nf(x) = f(x) +
        \int_0^x \eu^{x-t}f(t)\,\dd t .
        \]
        (The series $\sum T^n$ is a geometric series of
        operators: a first taste of the resolvent
        $(\mathrm{Id} - T)^{-1}$, developed in the Year 3
        volume.)
\end{enumerate}
\end{solution}

\begin{solution}{pb:b2:funcseq:1}
\textbf{1.} Linearity is clear from the formula. Positivity:
the weights $p_k(x) \geq 0$, so $f \geq 0$ forces $B_nf \geq
0$; monotonicity follows applied to $g - f$. Bound: $\pm f \leq
\norm f_\infty$ gives $\pm B_nf \leq \norm f_\infty B_ne_0 =
\norm f_\infty$. Endpoints: $p_k(0) = \mathbf 1_{k=0}$ and
$p_k(1) = \mathbf 1_{k=n}$, so $B_nf(0) = f(0)$, $B_nf(1) =
f(1)$.

\textbf{2.} Differentiate $(x+y)^n = \sum_k\binom nk x^ky^{n-k}$
in $x$, multiply by $x$, and set $y = 1 - x$:
\[
nx = \sum_k k\,p_k(x) ;
\]
twice, multiplying by $x^2$: $n(n-1)x^2 = \sum_k k(k-1)p_k(x)$.
Hence $B_ne_0 = 1$ (binomial theorem), $B_ne_1(x) =
\frac{nx}{n} = x$, and
\[
B_ne_2(x) = \frac{\sum_k k^2p_k}{n^2}
= \frac{n(n-1)x^2 + nx}{n^2}
= x^2 + \frac{x(1-x)}{n} .
\]

\textbf{3.} Expand:
\[
\sum_k\Bigl(\frac kn - x\Bigr)^{\!2} p_k
= B_ne_2(x) - 2x\,B_ne_1(x) + x^2
= \frac{x(1-x)}{n} .
\]
Cauchy--Schwarz with the splitting $\abs{k/n - x}\sqrt{p_k}
\cdot \sqrt{p_k}$:
\[
\sum_k\Bigl|\frac kn - x\Bigr| p_k
\leq \Bigl(\sum_k\Bigl(\frac kn - x\Bigr)^2
p_k\Bigr)^{\!1/2}
= \sqrt{\frac{x(1-x)}{n}} \leq \frac{1}{2\sqrt n},
\]
using $x(1-x) \leq \frac14$.

\textbf{4.} The weights $p_k(x)$ are nonnegative with sum $1$
and barycenter $\sum_k \frac kn p_k(x) = x$ (question 2). The
finite Jensen inequality for the convex $f$ (induction from the
two-point definition, Year 1 volume) gives
\[
f(x) = f\Bigl(\sum_k \frac kn\,p_k\Bigr)
\leq \sum_k f\Bigl(\frac kn\Bigr)p_k = B_nf(x) .
\]

\textbf{5.} On $\{k : \abs{k/n - x} > \delta\}$ one has
$\bigl(\frac{k/n - x}{\delta}\bigr)^2 > 1$, so
\[
\sum_{\abs{k/n-x}>\delta} p_k
\leq \frac{1}{\delta^2}\sum_k\Bigl(\frac kn -
x\Bigr)^2p_k
= \frac{x(1-x)}{n\delta^2} \leq \frac{1}{4n\delta^2} .
\]
Reading: $p_k(x)$ is the law of a sample frequency $S_n/n$ of
$n$ coin flips of bias $x$; its mean is $x$, its variance
$\frac{x(1-x)}n \to 0$, and the display is Chebyshev's
inequality: the mass concentrates at $x$, so averaging $f$
against it reproduces $f(x)$ in the limit
(\cref{ch:b2:genfun} makes the vocabulary official).

\textbf{6.} $\omega \leq 2M < \infty$; monotonicity is clear
(sup over a larger set). Heine: $f$ continuous on a compact is
uniformly continuous, which says exactly $\omega(\delta) \to 0$
as $\delta \to 0^+$. Subadditivity: if $\abs{s - t} \leq
\delta_1 + \delta_2$, the point $u$ on the segment $\intcc st$
at distance $\min(\delta_1, \abs{s-t})$ from $s$ satisfies
$\abs{s-u} \leq \delta_1$, $\abs{u-t} \leq \delta_2$, and
$\abs{f(s)-f(t)} \leq \abs{f(s)-f(u)} + \abs{f(u)-f(t)}$.
Iterating, $\omega(p\delta) \leq p\,\omega(\delta)$ for $p \in
\N^*$; for $\lambda > 0$, with $p = \lceil\lambda\rceil \leq 1
+ \lambda$: $\omega(\lambda\delta) \leq \omega(p\delta) \leq
p\,\omega(\delta) \leq (1+\lambda)\omega(\delta)$.

\textbf{7.} Since $\sum p_k = 1$:
\[
\abs{B_nf(x) - f(x)}
= \Bigl|\sum_k\bigl(f(k/n) - f(x)\bigr)p_k\Bigr|
\leq \sum_k\omega\bigl(\abs{k/n - x}\bigr)p_k .
\]
For each $k$, question 6 with $\lambda = \abs{k/n - x}/\delta$
gives $\omega(\abs{k/n-x}) \leq \bigl(1 +
\frac{\abs{k/n-x}}\delta\bigr)\omega(\delta)$; summing against
the $p_k$ yields the master estimate.

\textbf{8.} Insert the bound of question 3:
\[
\abs{B_nf(x) - f(x)} \leq \Bigl(1 +
\frac{1}{2\delta\sqrt n}\Bigr)\omega(\delta),
\]
uniformly in $x$; with $\delta = n^{-1/2}$ the parenthesis is
$\frac32$: $\norm{B_nf - f}_\infty \leq
\frac32\omega(n^{-1/2}) \to 0$ by question 6 (Heine). This
reproves \cref{thm:b2:funcseq:weierstrass} with a rate.

\textbf{9.} $L$-Lipschitz means $\omega(\delta) \leq L\delta$:
rate $\frac{3L}{2\sqrt n}$. $\alpha$-H\"older means
$\omega(\delta) \leq C\delta^\alpha$: rate $\frac{3C}{2}
n^{-\alpha/2}$.

\textbf{10.} Using $k\binom{2m}k = 2m\binom{2m-1}{k-1}$:
\[
\sum_{k=m+1}^{2m}k\binom{2m}k
= 2m\sum_{j=m}^{2m-1}\binom{2m-1}{j}
= 2m\cdot 2^{2m-2},
\]
because $j \mapsto 2m-1-j$ bijects $\{m,\dots,2m-1\}$ onto
$\{0,\dots,m-1\}$, so the sum is half of $2^{2m-1}$. Also
$\sum_{k=m+1}^{2m}\binom{2m}k = \frac{2^{2m} -
\binom{2m}m}{2}$ (same symmetry). Hence
\[
\sum_{k=m+1}^{2m}(k-m)\binom{2m}k
= m\,2^{2m-1} - m\,\frac{2^{2m} - \binom{2m}m}{2}
= \frac m2\binom{2m}m .
\]
The substitution $k \mapsto 2m-k$ maps the terms with $k < m$
onto those with $k > m$ (equal binomials, equal $\abs{k-m}$):
the absolute sum is twice the one-sided sum, $m\binom{2m}m$.

\textbf{11.} At $x = \frac12$, $p_k(\tfrac12) =
\binom{2m}k2^{-2m}$ and $f(\tfrac12) = 0$:
\[
B_{2m}f\Bigl(\frac12\Bigr)
= \sum_k\Bigl|\frac{k}{2m} - \frac12\Bigr|
\binom{2m}k 2^{-2m}
= \frac{2^{-2m}}{2m}\,m\binom{2m}m
= \frac{\binom{2m}m}{2\cdot4^m}
\sim \frac{1}{2\sqrt{\pi m}}
\]
by \cref{ex:b2:comparison:centralbinomial}. Since $\omega_f
(\delta) = \delta$ here (the function is $1$-Lipschitz and the
bound is attained), question 8 predicts at most
$\frac32(2m)^{-1/2}$: the true error $\frac{1}{2\sqrt{\pi m}}$
has exactly the order $n^{-1/2}$ --- the rate is sharp up to
the constant.

\textbf{12.} $f \leq g$ gives $g - f \geq 0$, so $L(g-f) \geq
0$, i.e.\ $Lf \leq Lg$. From $-\abs f \leq f \leq \abs f$:
$-L\abs f \leq Lf \leq L\abs f$, i.e.\ $\abs{Lf} \leq
L\abs f$.

\textbf{13.} By Heine choose $\delta$ with $\abs{f(s)-f(x)}
\leq \varepsilon$ whenever $\abs{s-x} \leq \delta$. If
$\abs{s - x} > \delta$, then $\frac{(s-x)^2}{\delta^2} > 1$ and
$\abs{f(s)-f(x)} \leq 2M \leq \frac{2M}{\delta^2}(s-x)^2$. In
both cases the claimed bound holds.

\textbf{14.} Fix $x$; question 13 says, as functions of $s$:
\[
-\varepsilon e_0 - \frac{2M}{\delta^2}q_x
\;\leq\; f - f(x)e_0
\;\leq\; \varepsilon e_0 + \frac{2M}{\delta^2}q_x,
\qquad q_x = e_2 - 2x\,e_1 + x^2e_0 .
\]
Apply the monotone linear $L_n$ (question 12) and evaluate at
$x$:
\[
\abs{L_nf(x) - f(x)L_ne_0(x)}
\leq \varepsilon L_ne_0(x)
+ \frac{2M}{\delta^2}\bigl(L_ne_2(x) - 2xL_ne_1(x)
+ x^2L_ne_0(x)\bigr) .
\]

\textbf{15.} Write $\alpha_j = L_ne_j - e_j$, so
$\norm{\alpha_j}_\infty \to 0$. Since $e_2(x) - 2xe_1(x) +
x^2e_0(x) = 0$:
\[
L_ne_2(x) - 2xL_ne_1(x) + x^2L_ne_0(x)
= \alpha_2(x) - 2x\,\alpha_1(x) + x^2\alpha_0(x),
\]
of sup norm at most $\norm{\alpha_2} + 2\norm{\alpha_1} +
\norm{\alpha_0} \to 0$. Also $L_ne_0 \to e_0$ uniformly, so
$L_ne_0 \leq 2$ for large $n$, and $\abs{f(x)}\abs{L_ne_0(x) -
1} \leq M\norm{\alpha_0} \to 0$. Assembling with question 14:
for $n$ large, uniformly in $x$,
\[
\abs{L_nf(x) - f(x)} \leq 2\varepsilon +
\frac{2M}{\delta^2}\,o(1) + M\,o(1) \leq 3\varepsilon :
\]
$L_nf \to f$ uniformly --- Korovkin's theorem.

\textbf{16.} $B_ne_0 = e_0$ and $B_ne_1 = e_1$ exactly, and
$\norm{B_ne_2 - e_2}_\infty = \max_x\frac{x(1-x)}{n} =
\frac{1}{4n} \to 0$ (question 2): Korovkin applies, and
Weierstrass follows for the third time.

\textbf{17.} $I_nf$ is linear in $f$ (nodal values are), and on
each cell the affine interpolant of nonnegative nodal values is
nonnegative: positive. $I_ne_0 = e_0$ and $I_ne_1 = e_1$
because an affine function equals its own interpolant. On a
cell $\intcc ab$ ($b - a = \frac1n$), the affine interpolant of
$e_2$ is $L(t) = (a+b)t - ab$, and
\[
L(t) - t^2 = (t-a)(b-t) \in
\intcc{0}{\tfrac{(b-a)^2}{4}} ,
\]
with the maximum at the midpoint: $\norm{I_ne_2 - e_2}_\infty =
\frac{1}{4n^2} \to 0$. Korovkin: $I_nf \to f$ uniformly for
every continuous $f$ --- polygonal approximation, with no
further estimate needed.

\textbf{18.} Given $f$ and $\varepsilon$: Weierstrass provides
a polynomial $P = \sum_{j=0}^d a_jx^j$ with $\norm{f -
P}_\infty \leq \frac\varepsilon2$; replacing each $a_j$ by a
rational $b_j$ with $\abs{a_j - b_j} \leq
\frac{\varepsilon}{2(d+1)}$ moves the sup norm on
$\intcc{0}{1}$ by at most $\frac\varepsilon2$. The set of
rational-coefficient polynomials is a countable union (over
$d$) of countable sets, hence countable, and dense:
$C(\intcc{0}{1})$ is separable.

\textbf{19.} By linearity $\int_0^1 fP = 0$ for every
polynomial $P$. Choose polynomials $P_n \to f$ uniformly
(Weierstrass):
\[
\Bigl|\int_0^1 f^2\Bigr|
= \Bigl|\int_0^1 f\,(f - P_n)\Bigr|
\leq \norm f_\infty\,\norm{f - P_n}_\infty
\longrightarrow 0 ,
\]
so $\int_0^1 f^2 = 0$. If $f(x_0) \neq 0$, continuity gives
$f^2 \geq c > 0$ on a subinterval, contradicting the vanishing
integral: $f = 0$. Consequently two continuous functions with
the same moments $\int f t^n$ coincide.

\textbf{20.} With $p_{n,k}(x) = \binom nk x^k(1-x)^{n-k}$ and
the conventions $p_{n-1,-1} = p_{n-1,n} = 0$, the product rule
and $k\binom nk = n\binom{n-1}{k-1}$, $(n-k)\binom nk =
n\binom{n-1}{k}$ give
\[
p_{n,k}'(x) = n\bigl(p_{n-1,k-1}(x) - p_{n-1,k}(x)\bigr) .
\]
Summing against $f(k/n)$ and shifting the index in the first
sum (Abel summation):
\[
(B_nf)'(x) = n\sum_{j=0}^{n-1}\Bigl(f\Bigl(\frac{j+1}n\Bigr)
- f\Bigl(\frac jn\Bigr)\Bigr)p_{n-1,j}(x) .
\]

\textbf{21.} By the mean value theorem, $f(\frac{j+1}n) -
f(\frac jn) = \frac1n f'(\xi_j)$ with $\xi_j \in
\intoo{j/n}{(j+1)/n}$, so $(B_nf)'(x) = \sum_j
f'(\xi_j)\,p_{n-1,j}(x)$. The node $\frac{j}{n-1}$ also lies in
$\intcc{j/n}{(j+1)/n}$ (both inequalities reduce to $j \leq
n-1$), hence $\abs{\xi_j - \frac j{n-1}} \leq \frac1n$ and
\[
\bigl|(B_nf)'(x) - B_{n-1}(f')(x)\bigr|
\leq \sum_j\Bigl|f'(\xi_j) -
f'\Bigl(\frac{j}{n-1}\Bigr)\Bigr| p_{n-1,j}(x)
\leq \omega_{f'}\Bigl(\frac1n\Bigr) \longrightarrow 0
\]
uniformly. Since $B_{n-1}(f') \to f'$ uniformly
(\cref{thm:b2:funcseq:weierstrass} applied to the continuous
$f'$), the triangle inequality gives $(B_nf)' \to f'$
uniformly. The polynomials $P_n = B_nf$ then converge to $f$
in the $C^1$ sense.

\textbf{22.} Taylor--Lagrange at $x$: $f(\frac kn) - f(x) =
f'(x)(\frac kn - x) + \frac{f''(\xi_k)}2(\frac kn - x)^2$.
Summing against $p_k$, the linear term dies (question 2):
\[
\abs{B_nf(x) - f(x)}
\leq \frac{\norm{f''}_\infty}{2}\sum_k\Bigl(\frac kn -
x\Bigr)^2p_k
= \frac{\norm{f''}_\infty}{2}\cdot\frac{x(1-x)}{n}
\leq \frac{\norm{f''}_\infty}{8n} .
\]

\textbf{23.} Two more differentiations of $(x+y)^n$ give the
factorial moments, with $n_{(j)} = n(n-1)\cdots(n-j+1)$:
\[
\sum_k k_{(j)}\,p_k = n_{(j)}\,x^j \qquad (j = 3, 4),
\]
and $k^3 = k_{(3)} + 3k_{(2)} + k$, $k^4 = k_{(4)} + 6k_{(3)} +
7k_{(2)} + k$ convert them to power moments:
\[
\sum_k k^3p_k = n_{(3)}x^3 + 3n_{(2)}x^2 + nx,
\qquad
\sum_k k^4p_k = n_{(4)}x^4 + 6n_{(3)}x^3 + 7n_{(2)}x^2 + nx .
\]
Expanding $(k - nx)^4$ and collecting (a patient but purely
mechanical computation with the four power moments):
\[
\sum_k(k-nx)^4p_k = nx(1-x)\bigl(1 + 3(n-2)x(1-x)\bigr) .
\]
With $x(1-x) \leq \frac14$: the right side is at most
$\frac n4\bigl(1 + \frac{3n}4\bigr) = \frac{3n^2}{16} +
\frac n4 \leq n^2$ for $n \geq 1$.

\textbf{24.} The Peano form of Taylor at $x$ defines $\eta(t) =
\frac{f(t) - f(x) - f'(x)(t-x) - \frac12f''(x)(t-x)^2}
{(t-x)^2}$ for $t \neq x$, $\eta(x) = 0$: by Taylor--Lagrange
$\eta(t) = \frac12\bigl(f''(\xi) - f''(x)\bigr)$ for some $\xi$
between $t$ and $x$, so $\abs\eta \leq \norm{f''}_\infty$ and
$\eta(t) \to 0$ as $t \to x$ (continuity of $f''$). Summing
the expansion against $p_k$ and using questions 2--3:
\[
n\bigl(B_nf(x) - f(x)\bigr)
= \frac{x(1-x)}{2}f''(x)
+ n\sum_k\eta\Bigl(\frac kn\Bigr)\Bigl(\frac kn -
x\Bigr)^2p_k .
\]
Given $\varepsilon$, choose $\delta$ with $\abs\eta \leq
\varepsilon$ on $\abs{t - x}\leq\delta$. Near part: at most
$\varepsilon\,n\cdot\frac{x(1-x)}n \leq \varepsilon$. Far part:
with $C = \norm{f''}_\infty$ and question 23,
\[
n\,C\sum_{\abs{k/n-x}>\delta}\Bigl(\frac kn -
x\Bigr)^2p_k
\leq \frac{nC}{\delta^2}\sum_k\Bigl(\frac kn -
x\Bigr)^4p_k
= \frac{nC}{\delta^2 n^4}\sum_k(k-nx)^4p_k
\leq \frac{C}{\delta^2 n} \longrightarrow 0 .
\]
Hence $n(B_nf(x) - f(x)) \to \frac{x(1-x)}2f''(x)$ ---
Voronovskaya's theorem. For $f = e_2$ this is exact at every
$n$ (\cref{exo:b2:funcseq:7}): the $\frac1n$ ceiling is real.

\textbf{25.} (i) Positivity turned pointwise inequalities into
operator inequalities: it gave the norm bound, Jensen,
Chebyshev, and the whole of Korovkin --- linearity alone proves
nothing here. (ii) Compactness entered through Heine (questions
6, 13), through boundedness of $f$, and through the very norm
$\norm\cdot_\infty$ being finite. (iii) Three test functions
suffice because positivity reduces everything to controlling
$L_n$ on the single family $(s-x)^2 = e_2 - 2xe_1 + x^2e_0$,
whose span is that of $e_0, e_1, e_2$. (iv) Bernstein converges
at the honest rate $\omega(n^{-1/2})$ for every continuous $f$
(sharp, question 11) but saturates at $\frac1n$ for smooth $f$
(question 24); the Fourier chapter runs the same program for
periodic functions with Fej\'er's kernel --- another positive
operator with the same virtues and the same modesty.
\end{solution}
