\section*{Chapter \ref{ch:b2:quadratic} --- Quadratic Forms}

\begin{solution}{exo:b2:quadratic:1}
$q_1 = (x + 2y)^2 - 3y^2$: rank $2$, signature $(1, 1)$ (a
hyperbola-type form).

$q_2$: complete the square in $x$: $q_2 = (x + y)^2 + y^2 + 2yz +
3z^2 = (x+y)^2 + (y + z)^2 + 2z^2$: rank $3$, signature $(3, 0)$
--- positive definite.
\end{solution}

\begin{solution}{exo:b2:quadratic:2}
\omterm{def:b2:reduction:eigen}{Eigenvalues} $2$ (on $\operatorname{Vect}(1,1,1)$) and $-1$ (on the
plane $x + y + z = 0$). Orthonormalize: $e_1 =
\frac{1}{\sqrt3}(1,1,1)$; in the plane, Gram--Schmidt on $(1,-1,0),
(1,0,-1)$ gives $e_2 = \frac{1}{\sqrt2}(1,-1,0)$, $e_3 =
\frac{1}{\sqrt6}(1,1,-2)$. Then $P = (e_1\ e_2\ e_3)$ is orthogonal
with $P^{\mathsf T}AP = \operatorname{diag}(2,-1,-1)$.

The form $q = 2xy + 2yz + 2zx$ has matrix $A$: in the rotated
coordinates $q = 2X^2 - Y^2 - Z^2$ --- principal axes; signature
$(1,2)$, agreeing with \cref{ex:b2:quadratic:gaussexample}
(the same form!).
\end{solution}

\begin{solution}{exo:b2:quadratic:3}
$u^{**} = u$: $\langle u^{**}x, y\rangle = \langle x, u^*y\rangle =
\langle ux, y\rangle$ for all $y$. $(uv)^* = v^*u^*$: $\langle
uvx, y\rangle = \langle vx, u^*y\rangle = \langle x,
v^*u^*y\rangle$. Kernel: $y \in \ker u^* \iff \langle x, u^*y
\rangle = 0\ \forall x \iff \langle u(x), y\rangle = 0\ \forall x
\iff y \perp \operatorname{im} u$. Ranks: $\dim\ker u^* = n -
\operatorname{rk} u$ (orthogonal complement), so
$\operatorname{rk} u^* = \operatorname{rk} u$ by rank--nullity ---
the Euclidean avatar of the transpose-rank theorem.
\end{solution}

\begin{solution}{exo:b2:quadratic:4}
Spectral: $A = PDP^{\mathsf T}$, $D$ diagonal with entries
$\lambda_i$ satisfying $\lambda_i^3 = \lambda_i$: $\lambda_i \in
\{-1, 0, 1\}$. Then $A^2 = PD^2P^{\mathsf T}$ with $D^2$ diagonal
of entries $0/1$: $A^2$ is \omterm{def:b2:quadratic:adjoint}{symmetric} and idempotent
($(A^2)^2 = A^4 = A\cdot A^3 = A^2$) --- \omterm{def:b2:quadratic:adjoint}{symmetric} idempotent
$=$ orthogonal projection (it is the projection onto
$\ker(A^2 - I) = \ker(A-I)\oplus\ker(A+I)$ along $\ker A$, and
these are orthogonal by the spectral theorem).

Generally, $P(A) = P\!\left(\text{diag}\right)$: $P(A)$ has the
same \omterm{def:b2:reduction:eigen}{eigenvectors}, \omterm{def:b2:reduction:eigen}{eigenvalues} $P(\lambda_i)$ --- the ``spectral
mapping'' at the \omterm{def:b2:reduction:diag}{diagonalizable} level.
\end{solution}

\begin{solution}{exo:b2:quadratic:5}
Closed: $O(n) = g^{-1}(\{I\})$ for the \omterm{def:b2:metric:continuity}{continuous} $g(P) =
P^{\mathsf T}P$ (polynomial entries). Bounded: each column of $P
\in O(n)$ is a unit vector, so all entries lie in $\intcc{-1}{1}$.
Closed and bounded in $\mathcal{M}_n(\R) \simeq \R^{n^2}$: \omterm{def:b2:metric:compact}{compact}
(\cref{thm:b2:metric:compactprops}~(2)).

Not \omterm{def:b2:metric:connected}{connected}: $\det$ takes the two values $\pm1$ on $O(n)$, and a
\omterm{def:b2:metric:continuity}{continuous} surjection onto $\{-1, 1\}$ splits the space
(\cref{ex:b2:metric:glnr}'s argument).
\end{solution}

\begin{solution}{exo:b2:quadratic:6}
\emph{Existence:} $A = PDP^{\mathsf T}$ with $D =
\operatorname{diag}(\lambda_i)$, $\lambda_i \geq 0$; set $B =
P\sqrt D P^{\mathsf T}$ with $\sqrt D =
\operatorname{diag}(\sqrt{\lambda_i})$: \omterm{def:b2:quadratic:adjoint}{symmetric}, positive
semidefinite, $B^2 = A$.

\emph{Uniqueness:} let $B$ be \omterm{def:b2:quadratic:adjoint}{symmetric} psd with $B^2 = A$. $B$
commutes with $A$; hence $B$ preserves each \omterm{def:b2:reduction:eigen}{eigenspace}
$E_\lambda(A)$ (for $Ax = \lambda x$: $A(Bx) = BAx = \lambda Bx$).
On $E_\lambda(A)$, the restriction of $B$ is \omterm{def:b2:quadratic:adjoint}{symmetric} psd with
square $\lambda\,\mathrm{id}$; its \omterm{def:b2:reduction:eigen}{eigenvalues} $\mu$ satisfy $\mu^2
= \lambda$, $\mu \geq 0$: $\mu = \sqrt\lambda$ --- so the
restriction, being \omterm{def:b2:reduction:diag}{diagonalizable} with the single \omterm{def:b2:reduction:eigen}{eigenvalue}
$\sqrt\lambda$, \emph{is} $\sqrt\lambda\,\mathrm{id}$. Since $E =
\bigoplus E_\lambda(A)$, $B$ is determined: $B = \sqrt A$.
\end{solution}

\begin{solution}{exo:b2:quadratic:7}
In an orthonormal eigenbasis, $\norm{Ax}_2^2 = \sum \lambda_i^2
x_i^2 \leq (\max_i \lambda_i^2)\norm x_2^2$, with equality at the
corresponding \omterm{def:b2:reduction:eigen}{eigenvector}: $\vertiii A_2 = \max\abs{\lambda_i}$.
For the given matrix: \omterm{def:b2:reduction:eigen}{eigenvalues} $3, -1$
(\cref{ex:b2:quadratic:spectralexample}'s twin):
$\vertiii A_2 = 3$.
\end{solution}

\begin{solution}{exo:b2:quadratic:8}
($\Rightarrow$) The top-left $k \times k$ block $A_k$ is the matrix
of the restriction of the (definite) form to the span of the first
$k$ basis vectors: positive definite, so its \omterm{def:b2:reduction:eigen}{eigenvalues} are
positive and $\Delta_k = \det A_k > 0$.

($\Leftarrow$) Induction on $n$; $n = 1$ clear. Assume all
$\Delta_k > 0$. By induction, $A_{n-1}$ is positive definite: the
form $q$ restricted to $F = \operatorname{Vect}(e_1, \dots,
e_{n-1})$ is definite. Diagonalize $q|_F$ (Gauss): coordinates
$y_1, \dots, y_{n-1}$ with $q|_F = \sum y_i^2$. In the full space,
completing the square in the last variable,
\[
q = \sum_{i=1}^{n-1} \bigl(y_i + c_i x_n\bigr)^2 + c\,x_n^2
\]
for suitable constants (collect the cross terms into the squares).
The reduction exhibits signature $(n-1 + \epsilon, \cdot)$ with
$\epsilon$ the sign contribution of $c$; and the \omterm{def:b2:linalg:det}{determinant} keeps
the sign of the product of the diagonal coefficients under
\omterm{def:b2:quadratic:def}{congruence} ($\det(P^{\mathsf T}AP) = (\det P)^2\det A$): $\Delta_n
> 0$ forces $c > 0$. Hence $q$ is a sum of $n$ squares of
independent forms: positive definite.
\end{solution}

\begin{solution}{exo:b2:quadratic:9}
Let $(e_1, \dots, e_n)$ be an orthonormal eigenbasis for $\lambda_1
\geq \dots \geq \lambda_n$.

\emph{$\lambda_2 \leq$ the min-max:} for any hyperplane $H$, the
$2$-dimensional $V = \operatorname{Vect}(e_1, e_2)$ satisfies
$\dim(H \cap V) \geq 1$ (Grassmann): pick a unit $x \in H \cap V$,
$x = ae_1 + be_2$, $a^2 + b^2 = 1$:
\[
\langle u(x), x\rangle = \lambda_1 a^2 + \lambda_2 b^2 \geq
\lambda_2 :
\]
every hyperplane's max is $\geq \lambda_2$.

\emph{$\geq$:} for $H = e_1^{\perp}$, every unit $x = \sum_{i\geq2}
x_ie_i \in H$ has $\langle u(x), x\rangle = \sum_{i \geq 2}
\lambda_i x_i^2 \leq \lambda_2$, attained at $e_2$: this
hyperplane's max is exactly $\lambda_2$. The min over $H$ is
therefore $\lambda_2$.
\end{solution}

\begin{solution}{exo:b2:quadratic:10}
\emph{Algebraic road:} $2q = \bigl(\sum x_i\bigr)^2 - \sum
x_i^2$. On the hyperplane $H : \sum x_i = 0$ (dimension $n -
1$), $q = -\frac12\sum x_i^2$ is negative definite; on the line
$\R(1, \dots, 1)$, $q(t, \dots, t) = \binom n2 t^2 > 0$. A
subspace where $q$ is positive definite meets $H$ trivially, so
has dimension $\leq 1$: by Sylvester
(\cref{thm:b2:quadratic:sylvester}), $s = 1$, and $t \geq n-1$
from $H$; rank $\leq n$ forces signature $(1, n-1)$, rank $n$.

\emph{Spectral road:} the matrix is $\frac12(J - I)$; $J$ has
\omterm{def:b2:reduction:eigen}{eigenvalues} $n$ (on $(1,\dots,1)$) and $0$ (on $H$), so
$\frac12(J-I)$ has \omterm{def:b2:reduction:eigen}{eigenvalues} $\frac{n-1}{2}$ (once) and
$-\frac12$ ($n-1$ times): one positive, $n-1$ negative --- same
signature, by \cref{cor:b2:quadratic:principalaxes}.
\end{solution}

\begin{solution}{exo:b2:quadratic:11}
Write $B = C^{\mathsf T}C$ (\cref{exo:b2:quadratic:6} via $C =
\sqrt B$). Then, with $c_1, \dots, c_n$ the columns of
$C^{\mathsf T}$:
\[
\operatorname{tr}(AB) = \operatorname{tr}(AC^{\mathsf T}C)
= \operatorname{tr}(CAC^{\mathsf T})
= \sum_{i=1}^n \langle A c_i, c_i\rangle .
\]
By \cref{cor:b2:quadratic:principalaxes}~(2), each term lies
between $\lambda_{\min}(A)\norm{c_i}^2$ and
$\lambda_{\max}(A)\norm{c_i}^2$, and $\sum\norm{c_i}^2 =
\operatorname{tr}(C^{\mathsf T}C)^{\vphantom1} =
\operatorname{tr} B$: the double inequality follows. If $A$ is
also positive semidefinite, $\lambda_{\min}(A) \geq 0$:
$\operatorname{tr}(AB) \geq 0$.
\end{solution}

\begin{solution}{exo:b2:quadratic:12}
\begin{enumerate}
  \item $\varphi(M, N) = \operatorname{tr}(MN)$ is bilinear and
        \omterm{def:b2:quadratic:adjoint}{symmetric} ($\operatorname{tr}(MN) =
        \operatorname{tr}(NM)$), and $\varphi(M, M) = q(M)$:
        $q$ is the \omterm{def:b2:quadratic:def}{quadratic form} of $\varphi$.
  \item For $S$ \omterm{def:b2:quadratic:adjoint}{symmetric} and $K$ antisymmetric:
        $\operatorname{tr}(SK) =
        \operatorname{tr}\bigl((SK)^{\mathsf T}\bigr) =
        \operatorname{tr}(K^{\mathsf T}S^{\mathsf T}) =
        -\operatorname{tr}(KS) = -\operatorname{tr}(SK)$, so
        $\varphi(S, K) = 0$: the two subspaces are
        $\varphi$-orthogonal. Entrywise, $\operatorname{tr}
        (M^2) = \sum_{i,j} m_{ij}m_{ji}$: for \omterm{def:b2:quadratic:adjoint}{symmetric} $M$
        this is $\sum m_{ij}^2 > 0$ ($M \neq 0$); for
        antisymmetric $M$ it is $-\sum m_{ij}^2 < 0$.
  \item $\mathcal M_n(\R) = S_n \oplus A_n$ with dimensions
        $\frac{n(n+1)}2$ and $\frac{n(n-1)}2$; a \omterm{thm:b2:quadratic:gauss}{Gauss
        reduction} adapted to this $\varphi$-orthogonal
        splitting writes $q$ as $\frac{n(n+1)}2$ positive and
        $\frac{n(n-1)}2$ negative squares: signature
        $\bigl(\frac{n(n+1)}2, \frac{n(n-1)}2\bigr)$
        (Sylvester), rank $n^2$: the form is nondegenerate.
\end{enumerate}
\end{solution}

\begin{solution}{pb:b2:quadratic:1}
\textbf{1.} $G$ is \omterm{def:b2:quadratic:adjoint}{symmetric} by symmetry of the inner product,
and
\[
X^{\mathsf T}GX = \sum_{i,j}x_ix_j\langle v_i, v_j\rangle
= \Bigl\|\sum_i x_iv_i\Bigr\|^2 \geq 0 ,
\]
with equality iff $\sum x_iv_i = 0$: $G$ is definite iff the
only null combination is trivial, i.e.\ iff the family is
independent.

\textbf{2.} With $B = \sqrt A$ (\cref{exo:b2:quadratic:6}): $A
= B^2 = B^{\mathsf T}B$, the Gram matrix of the columns of $B$;
take $C = B$. And $X^{\mathsf T}AX = \norm{CX}^2$, so $A$ is
definite iff $CX \neq 0$ for $X \neq 0$, iff $C$ is invertible.

\textbf{3.} The restriction of the form to
$\operatorname{Vect}(e_i, e_j)$ has matrix
$\begin{pmatrix} a_{ii} & a_{ij}\\ a_{ij} & a_{jj}
\end{pmatrix}$, still positive semidefinite: its \omterm{def:b2:linalg:det}{determinant}
(product of its nonnegative \omterm{def:b2:reduction:eigen}{eigenvalues}) is $\geq 0$:
$a_{ij}^2 \leq a_{ii}a_{jj}$. This is Cauchy--Schwarz for the
vectors $v_i, v_j$ of a Gram realization.

\textbf{4.} \emph{Existence:} write $A$ as the Gram matrix of
an independent family $(v_1, \dots, v_n)$ (questions 1--2).
Gram--Schmidt produces an orthonormal $(e_1, \dots, e_n)$ with
\[
v_k = \sum_{i \leq k} t_{ik}\,e_i,
\qquad t_{kk} = \bigl\| v_k - \operatorname{proj}_{k-1}v_k
\bigr\| > 0 ,
\]
so $T = (t_{ik})$ is upper triangular with positive diagonal,
and
\[
a_{jk} = \langle v_j, v_k\rangle
= \sum_i t_{ij}t_{ik} = (T^{\mathsf T}T)_{jk} .
\]
\emph{Uniqueness:} if $T_1^{\mathsf T}T_1 = T_2^{\mathsf T}T_2$
then $R = T_1T_2^{-1}$ satisfies $R^{\mathsf T}R = I$: $R$ is
orthogonal, and also upper triangular with positive diagonal
(product of such). Then $R^{-1} = R^{\mathsf T}$ is
simultaneously upper (inverse of upper) and lower (\omterm{def:b2:linalg:transpose}{transpose} of
upper) triangular: diagonal; an orthogonal diagonal matrix has
entries $\pm1$, and positivity forces $R = I$: $T_1 = T_2$.

\textbf{5.} For $i, j \leq k$, $(T^{\mathsf T}T)_{ij} = \sum_m
t_{mi}t_{mj}$ only involves $m \leq \min(i,j) \leq k$: the
leading $k\times k$ block of $A$ is $T_k^{\mathsf T}T_k$ with
$T_k$ the leading block of $T$. Hence $\Delta_k = (\det T_k)^2
= (t_{11}\cdots t_{kk})^2$. Now the \omterm{thm:b2:quadratic:gauss}{Gauss reduction} of a
positive definite form in the natural order never meets a zero
square coefficient (the pivots are the diagonal entries of the
successively reduced positive definite blocks): it produces $q
= \sum_k d_k\ell_k^2$ with $\ell_k = x_k + (\text{terms in }
x_{k+1}, \dots)$, i.e.\ $A = L^{\mathsf T}DL$ with $L$
unipotent triangular; then $T = \sqrt D\,L$ is a Cholesky
factor, so by uniqueness $t_{kk}^2 = d_k$ and
\[
d_k = \frac{(t_{11}\cdots t_{kk})^2}
{(t_{11}\cdots t_{k-1,k-1})^2}
= \frac{\Delta_k}{\Delta_{k-1}} .
\]
On \cref{ex:b2:quadratic:tworoads}: $\Delta_1, \Delta_2,
\Delta_3 = 2, 3, 4$ and the pivots were $2, \frac32, \frac43$.

\textbf{6.} Each $a_{ii} = e_i^{\mathsf T}Ae_i > 0$. Let $D =
\operatorname{diag}(a_{ii}^{-1/2})$ and $B = DAD$: positive
definite (\omterm{def:b2:quadratic:def}{congruence}), with $b_{ii} = 1$, so $\operatorname{tr}
B = n$. Its \omterm{def:b2:reduction:eigen}{eigenvalues} $\mu_i > 0$ satisfy, by AM--GM,
\[
\det B = \prod_i\mu_i
\leq \Bigl(\frac{\sum\mu_i}{n}\Bigr)^{\!n} = 1 ,
\]
and $\det B = (\det D)^2\det A = \dfrac{\det A}{\prod
a_{ii}}$: $\det A \leq \prod a_{ii}$.

\textbf{7.} AM--GM is an equality iff all $\mu_i$ are equal
(to $1$); a \omterm{def:b2:quadratic:adjoint}{symmetric} matrix with sole \omterm{def:b2:reduction:eigen}{eigenvalue} $1$ is
$PIP^{\mathsf T} = I$. So equality iff $B = I$ iff $a_{ij} = 0$
for $i \neq j$: $A$ diagonal.

\textbf{8.} If $M$ is singular both sides are $\geq 0 =
\abs{\det M}$. Otherwise $A = M^{\mathsf T}M$ is positive
definite with $a_{ii} = \norm{c_i}^2$ and $\det A = (\det
M)^2$: question 6 gives $(\det M)^2 \leq \prod\norm{c_i}^2$.
Equality iff $A = M^{\mathsf T}M$ is diagonal (question 7),
i.e.\ iff the columns are pairwise orthogonal.

\textbf{9.} $\abs{\det M}$ is the volume of the parallelepiped
spanned by the columns: the volume is at most the product of
the edge lengths, with equality exactly for rectangular boxes.
If $\abs{m_{ij}} \leq 1$ then $\norm{c_i} \leq \sqrt n$, so
$\abs{\det M} \leq n^{n/2}$. (Attaining it forces orthogonal
columns of entries $\pm1$: a Hadamard matrix.)

\textbf{10.} $X^{\mathsf T}A^{\mathsf T}AX = \norm{AX}^2 > 0$
for $X \neq 0$ ($A$ invertible): $A^{\mathsf T}A$ is positive
definite. Its square root $S$ is positive definite (\omterm{def:b2:reduction:eigen}{eigenvalues}
$\sqrt{\lambda_i} > 0$), hence invertible, and $Q = AS^{-1}$
satisfies
\[
Q^{\mathsf T}Q = S^{-1}A^{\mathsf T}AS^{-1}
= S^{-1}S^2S^{-1} = I :
\]
$A = QS$ with $Q$ orthogonal, $S$ positive definite.

\textbf{11.} If $A = QS = Q'S'$ then $S'^{\,2} =
S'^{\mathsf T}Q'^{\mathsf T}Q'S' = A^{\mathsf T}A = S^2$; two
positive semidefinite matrices with the same square coincide
(\cref{exo:b2:quadratic:6}): $S' = S$, then $Q' = AS^{-1} =
Q$.

\textbf{12.} $\det(A + \varepsilon I)$ is a nonzero polynomial
in $\varepsilon$: it has finitely many roots, so some sequence
$\varepsilon_k \to 0$ avoids them. Write $A + \varepsilon_kI =
Q_kS_k$ (question 10). $O(n)$ is \omterm{def:b2:metric:compact}{compact}
(\cref{exo:b2:quadratic:5}): a subsequence gives $Q_{\varphi(k)}
\to Q \in O(n)$. Then
\[
S_{\varphi(k)} = Q_{\varphi(k)}^{\mathsf T}
\bigl(A + \varepsilon_{\varphi(k)}I\bigr)
\longrightarrow Q^{\mathsf T}A =: S,
\]
\omterm{def:b2:quadratic:adjoint}{symmetric} positive semidefinite as a limit of such (closed
conditions), and $A = QS$. Moreover $S^2 = S^{\mathsf T}S =
A^{\mathsf T}QQ^{\mathsf T}A = A^{\mathsf T}A$, so $S =
\sqrt{A^{\mathsf T}A}$ by uniqueness. For singular $A$, $S$ is
singular and $Q$ is not unique: it can be modified arbitrarily
on $(\operatorname{im} S)^{\perp}$ --- extreme case $A = 0$,
where every orthogonal $Q$ works.

\textbf{13.} Diagonalize $S = P\Sigma P^{\mathsf T}$ (spectral
theorem), $\Sigma = \operatorname{diag}(\sigma_i)$ with
$\sigma_i \geq 0$ the \omterm{def:b2:reduction:eigen}{eigenvalues} of $S =
\sqrt{A^{\mathsf T}A}$. Then
\[
A = QS = (QP)\,\Sigma\,P^{\mathsf T} = U\Sigma V^{\mathsf T},
\qquad U = QP,\ V = P \in O(n) .
\]

\textbf{14.} $\norm{Ax}^2 = x^{\mathsf T}S^2x \leq
\sigma_{\max}^2\norm x^2$ with equality at a top \omterm{def:b2:reduction:eigen}{eigenvector} of
$S$: $\vertiii A_2 = \sigma_{\max}$ --- for \omterm{def:b2:quadratic:adjoint}{symmetric} $A$,
$S = \sqrt{A^2}$ has \omterm{def:b2:reduction:eigen}{eigenvalues} $\abs{\lambda_i}$, recovering
\cref{exo:b2:quadratic:7}. \omterm{def:b2:linalg:det}{Determinant}: $\abs{\det A} =
\abs{\det U}\det\Sigma\abs{\det V} = \sigma_1\cdots\sigma_n$.
Sphere: writing $x = Vy$ with $\norm y = 1$, $Ax = U\Sigma y$
has coordinates $z_i = \sigma_iy_i$ in the orthonormal frame of
the columns of $U$: the image is $\{\sum z_i^2/\sigma_i^2 =
1\}$, an ellipsoid with semi-axes $\sigma_i$.

\textbf{15.} A rotation to principal axes
(\cref{cor:b2:quadratic:principalaxes}) turns $f$ into
$\lambda_1X^2 + \lambda_2Y^2 + \beta_1X + \beta_2Y + c$, with
$\lambda_1\lambda_2 = \det A$. If $\det A \neq 0$, absorb the
linear terms by the translation $X \mapsto X -
\frac{\beta_1}{2\lambda_1}$ (and likewise $Y$): $\lambda_1X'^2
+ \lambda_2Y'^2 = c'$. For $\det A > 0$ (equal signs): an
ellipse ($c'$ of the right sign), a point, or empty. For $\det
A < 0$: a hyperbola ($c' \neq 0$) or two crossing lines. If
$\det A = 0$ with rank $1$ (say $\lambda_2 = 0 \neq
\lambda_1$): $\lambda_1X'^2 + \beta_2Y + c''$, a parabola when
$\beta_2 \neq 0$; otherwise two parallel lines, one line, or
empty. Nondegenerate shapes: ellipse, hyperbola, parabola,
governed by the sign of $\det A$.

\textbf{16.} The quadratic part $x^2 + 4xy + y^2$ has matrix
$\begin{pmatrix}1 & 2\\ 2 & 1\end{pmatrix}$, \omterm{def:b2:reduction:eigen}{eigenvalues} $3$
and $-1$ with orthonormal directions $\frac{1}{\sqrt2}(1,1)$,
$\frac{1}{\sqrt2}(1,-1)$. In the rotated coordinates $u =
\frac{x+y}{\sqrt2}$, $v = \frac{x-y}{\sqrt2}$: $x^2 + y^2 = u^2
+ v^2$, $2xy = u^2 - v^2$, so the form is $3u^2 - v^2$, and $2x
- 2y = 2\sqrt2\,v$. The equation becomes
\[
3u^2 - v^2 + 2\sqrt2\,v = 4
\quad\Longleftrightarrow\quad
3u^2 - \bigl(v - \sqrt2\bigr)^2 = 2 :
\]
a hyperbola, center at $(u, v) = (0, \sqrt2)$, i.e.\ $(x, y) =
(1, -1)$, with axes along the rotated frame.

\textbf{17.} $f(x) = (x - x_0)^{\mathsf T}A(x - x_0) + f(x_0)$
whenever $Ax_0 = -\frac b2$, i.e.\ $x_0 = -\frac12A^{-1}b$: the
gradient of $f$ vanishes exactly there ($x_0$ is the center of
symmetry). With $M = \begin{pmatrix} I & x_0\\ 0 &
1\end{pmatrix}$:
\[
M^{\mathsf T}\widetilde QM
= \begin{pmatrix}
A & Ax_0 + \frac b2\\[2pt]
\bigl(Ax_0 + \frac b2\bigr)^{\mathsf T} &
x_0^{\mathsf T}Ax_0 + b^{\mathsf T}x_0 + c
\end{pmatrix}
= \begin{pmatrix} A & 0\\ 0 & f(x_0)\end{pmatrix},
\]
and $\det M = 1$: $\det\widetilde Q = f(x_0)\det A$. The
centered equation reads $q(X) = -f(x_0)$: for $f(x_0) = 0$ it
degenerates to $q(X) = 0$ (two lines through the center if the
signature is $(1,1)$, the single point $x_0$ if $q$ is
definite); for $f(x_0) \neq 0$ the conic is a genuine ellipse
or hyperbola.

\textbf{18.} $A^{-1} = -\frac13\begin{pmatrix} 1 & -2\\ -2 &
1\end{pmatrix}$, $\frac b2 = (1, -1)$: $x_0 =
-A^{-1}\frac b2 = (1, -1)$, as found in question 16. $f(x_0) =
q(1,-1) + 2 + 2 - 4 = (1 - 4 + 1) + 0 = -2 \neq 0$, and
$\det\widetilde Q = f(x_0)\det A = (-2)(-3) = 6 \neq 0$:
nondegenerate; $\det A = -3 < 0$: a hyperbola --- and indeed
the centered equation $3u^2 - (v - \sqrt2)^2 = -f(x_0) = 2$
matches question 16.

\textbf{19.} The matrix is $(1-\lambda)I + \lambda J$:
\omterm{def:b2:reduction:eigen}{eigenvalues} $1 + 2\lambda$ (direction $(1,1,1)$) and $1 -
\lambda$ (double, on $x + y + z = 0$). Cases:
\begin{itemize}
  \item $-\frac12 < \lambda < 1$: all \omterm{def:b2:reduction:eigen}{eigenvalues} positive:
        an ellipsoid of revolution about $(1,1,1)$ (a sphere
        for $\lambda = 0$);
  \item $\lambda = 1$: $q = (x+y+z)^2$: the equation gives the
        two parallel planes $x + y + z = \pm1$;
  \item $\lambda = -\frac12$: \omterm{def:b2:reduction:eigen}{eigenvalues} $0, \frac32,
        \frac32$: a circular cylinder of axis $(1,1,1)$;
  \item $\lambda > 1$: signature $(1, 2)$: a hyperboloid of
        two sheets;
  \item $\lambda < -\frac12$: signature $(2, 1)$: a
        hyperboloid of one sheet.
\end{itemize}

\textbf{20.} The polar form of $q$ is an inner product
$\langle\cdot,\cdot\rangle_q$ on $E$. For fixed $x$, $y \mapsto
\varphi'(x, y)$ (polar form of $q'$) is linear, hence equals
$\langle z_x, y\rangle_q$ for a unique $z_x$; $u(x) := z_x$ is
linear (uniqueness), and $\langle u(x), y\rangle_q =
\varphi'(x,y) = \varphi'(y,x) = \langle u(y), x\rangle_q$: $u$
is \omterm{def:b2:quadratic:adjoint}{symmetric} in the Euclidean space $(E,
\langle\cdot,\cdot\rangle_q)$. The spectral theorem
(\cref{thm:b2:quadratic:spectral}) gives a $q$-orthonormal
eigenbasis $(\varepsilon_i)$, $u(\varepsilon_i) =
\mu_i\varepsilon_i$: in it $q(x) = \sum x_i^2$ and $q'(x) =
\langle u(x), x\rangle_q = \sum\mu_ix_i^2$.

\textbf{21.} Let $P$ be the matrix of that basis: \omterm{def:b2:quadratic:def}{congruence}
gives $P^{\mathsf T}AP = I$ and $P^{\mathsf T}BP =
\operatorname{diag}(\mu_i)$. Then
\[
\det(B - \mu A) = \det(P^{-\mathsf T})
\det\bigl(\operatorname{diag}(\mu_i) - \mu I\bigr)
\det(P^{-1})
= (\det P)^{-2}\prod_i(\mu_i - \mu) :
\]
the $\mu_i$ are the roots of the pencil $\det(B - \mu A)$.

\textbf{22.} $\det(B - \mu A) =
\det\begin{pmatrix} -2\mu & 1-\mu\\ 1-\mu & -\mu\end{pmatrix}
= 2\mu^2 - (1-\mu)^2 = \mu^2 + 2\mu - 1$: roots $\mu_\pm = -1
\pm \sqrt2$. Solving $(B - \mu_\pm A)v = 0$: $v_\pm = (1 -
\mu_\pm,\ 2\mu_\pm)$ (the second row identity $(1-\mu)^2 =
2\mu^2$ at the roots confirms it). The $A$-norms come out
clean: $q_A(v_\pm) = 2(1 + \mu_\pm^2)$, and one checks
$\varphi_A(v_+, v_-) = 0$ using $\mu_+ + \mu_- = -2$,
$\mu_+\mu_- = -1$. The basis $\Bigl(\frac{v_+}{\sqrt{2(1 +
\mu_+^2)}}, \frac{v_-}{\sqrt{2(1+\mu_-^2)}}\Bigr)$ is
orthonormal for $A$ and diagonalizes $B$ with entries
$\mu_\pm$.

\textbf{23.} If some invertible $P$ diagonalized both forms,
question 21's computation would give $\det(B - \mu A) = (\det
P)^{-2}\prod(d_{2i} - \mu d_{1i})$, a real polynomial split
into real linear factors. But here
\[
\det(B - \mu A) = \det\begin{pmatrix} -\mu & 1\\ 1 &
\mu\end{pmatrix} = -\mu^2 - 1 ,
\]
of degree $2$ with no real root: contradiction. (The
Lorentz-signature $A$ admits $B$-``rotations'' with no real
axes.)

\textbf{24.} $A^{-1}B = A^{-1/2}\bigl(A^{-1/2}BA^{-1/2}\bigr)
A^{1/2}$ with $A^{1/2} = \sqrt A$ positive definite
(\cref{exo:b2:quadratic:6}): $A^{-1}B$ is similar to the
\emph{\omterm{def:b2:quadratic:adjoint}{symmetric}} $A^{-1/2}BA^{-1/2}$, hence \omterm{def:b2:reduction:diag}{diagonalizable} with
real \omterm{def:b2:reduction:eigen}{eigenvalues}. And $\det(B - \mu A) = \det A\cdot
\det(A^{-1}B - \mu I)$: the pencil roots $\mu_i$ of question
21 are exactly the \omterm{def:b2:reduction:eigen}{eigenvalues} of $A^{-1}B$.

\textbf{25.} (i) Every Part leaned on the spectral theorem:
through the square root (Cholesky, polar), the \omterm{def:b2:reduction:eigen}{eigenvalue}
bounds (Hadamard), the principal axes (conics), and the
$q$-adapted version (simultaneous reduction). (ii) Gram
realizations, Hadamard, and the \omterm{pb:b2:quadratic:1}{polar decomposition} survive in
the semidefinite world; Cholesky's uniqueness, the pivot
formula, and the simultaneous reduction need definiteness
(questions 12 and 23 show exactly how they fail). (iii)
Coupled small oscillations: kinetic energy (positive definite)
and potential energy are two \omterm{def:b2:quadratic:def}{quadratic forms}; the basis of
questions 20--22 is the system's \emph{normal modes}, and the
$\mu_i$ are the squared angular frequencies.
\end{solution}
