\section*{Chapter \ref{ch:b2:hermitian} --- Hermitian Forms}

\begin{solution}{exo:b2:hermitian:1}
$\langle x, y\rangle = \conj{1}\cdot\iu + \conj{\iu}\cdot 1 = \iu -
\iu = 0$: orthogonal. $\norm x = \norm y = \sqrt{1 + 1} = \sqrt2$.
Orthonormal basis: $\bigl(\frac{1}{\sqrt2}(1, \iu),\;
\frac{1}{\sqrt2}(\iu, 1)\bigr)$ --- the normalized pair itself.
\end{solution}

\begin{solution}{exo:b2:hermitian:2}
First: equals its conjugate transpose (real diagonal, $\conj{\iu} =
-\iu$ swapped): Hermitian (hence normal); not unitary ($A^\dagger A
\neq I$: columns not unit).

Second: $A^\dagger A = \frac12\begin{pmatrix} 1 & -\iu\\ -\iu &
1\end{pmatrix}\begin{pmatrix} 1 & \iu\\ \iu & 1\end{pmatrix} =
\frac12\begin{pmatrix} 2 & 0\\ 0 & 2\end{pmatrix} = I$: unitary
(hence normal); not Hermitian.

Third: $A^\dagger A = E_{22} \neq E_{11} = AA^\dagger$: not normal
(so neither Hermitian nor unitary) --- the standard nilpotent
counterexample.
\end{solution}

\begin{solution}{exo:b2:hermitian:3}
Uniqueness: $A = H + \iu K$ with $H^\dagger = H$, $K^\dagger = K$
forces $A^\dagger = H - \iu K$, so $H = \frac{A + A^\dagger}{2}$,
$K = \frac{A - A^\dagger}{2\iu}$; these formulas are Hermitian
(check: $\bigl(\frac{A - A^\dagger}{2\iu}\bigr)^\dagger = \frac{
A^\dagger - A}{-2\iu} = K$) and reconstruct $A$: existence.

Normality: $A^\dagger A - AA^\dagger = (H - \iu K)(H + \iu K) - (H
+ \iu K)(H - \iu K) = 2\iu(HK - KH)$: it vanishes iff $HK = KH$.
\end{solution}

\begin{solution}{exo:b2:hermitian:4}
$\chi_A = (X-2)^2 - 1$: eigenvalues $3$ and $1$. Eigenvectors, by
direct computation:
\[
A\begin{pmatrix}1\\ -\iu\end{pmatrix}
= \begin{pmatrix} 2 + \iu(-\iu)\\ -\iu + 2(-\iu)\end{pmatrix}
= \begin{pmatrix} 3\\ -3\iu \end{pmatrix}
= 3\begin{pmatrix}1\\ -\iu\end{pmatrix},
\qquad
A\begin{pmatrix}1\\ \iu\end{pmatrix}
= \begin{pmatrix} 2 + \iu\cdot\iu\\ -\iu + 2\iu\end{pmatrix}
= \begin{pmatrix}1\\ \iu\end{pmatrix}.
\]
Orthonormal eigenbasis: $u_1 = \frac{1}{\sqrt2}(1, -\iu)$
(eigenvalue $3$), $u_2 = \frac{1}{\sqrt2}(1, \iu)$ (eigenvalue
$1$); orthogonality as in \cref{exo:b2:hermitian:1}. Powers, via
the spectral projections $A^k = 3^k u_1u_1^\dagger + 1^k\,
u_2u_2^\dagger$:
\[
A^k = U\begin{pmatrix} 3^k & 0\\ 0 & 1\end{pmatrix}U^\dagger
= \frac{3^k}{2}\begin{pmatrix} 1 & \iu\\ -\iu & 1\end{pmatrix}
+ \frac{1}{2}\begin{pmatrix} 1 & -\iu\\ \iu & 1\end{pmatrix}
= \frac12\begin{pmatrix}
3^k + 1 & (3^k - 1)\iu\\
-(3^k-1)\iu & 3^k + 1
\end{pmatrix}.
\]
(Check $k = 1$: recovers $A$.)
\end{solution}

\begin{solution}{exo:b2:hermitian:5}
Compact: closed (preimage of $I$ under the continuous $U \mapsto
U^\dagger U$) and bounded (columns are unit vectors: entries of
modulus $\leq 1$) in $\mathcal{M}_n(\C) \simeq \R^{2n^2}$.

Eigenvalues: $\operatorname{diag}(\lambda, 1, \dots, 1)$ is unitary
for any $\abs\lambda = 1$. Determinant: $\abs{\det U}^2 = \det
U^\dagger \det U = \det(U^\dagger U) = 1$ (using $\det A^\dagger =
\conj{\det A}$): $\det U$ lies on the unit circle.
\end{solution}

\begin{solution}{exo:b2:hermitian:6}
By the spectral theorem, work in an orthonormal eigenbasis of $H$:
everything reduces to scalars $\lambda \in \R$ (the eigenvalues).
$H + \iu I$ has eigenvalues $\lambda + \iu \neq 0$: invertible. $U$
has eigenvalues $\mu = \frac{\lambda - \iu}{\lambda + \iu}$, of
modulus $1$ ($\abs{\lambda - \iu} = \abs{\lambda + \iu}$ for real
$\lambda$): $U^\dagger U = I$ holds since $U$ is unitarily
diagonalizable with unimodular eigenvalues (it is diagonal in the
chosen orthonormal basis). And $\mu = 1$ would force $-\iu = \iu$:
impossible, so $1 \notin \operatorname{Sp} U$. (The Cayley
transform maps Hermitian to unitary-minus-a-point --- the matrix
version of the map from $\R$ to the circle.)
\end{solution}

\begin{solution}{exo:b2:hermitian:7}
For an eigenpair $Ax = \lambda x$ ($x \neq 0$): $\lambda\norm x^2 =
\langle x, Ax\rangle > 0$, so $\lambda > 0$ (already real,
\cref{prop:b2:hermitian:eigenvalues}). Square root: in a spectral
basis, $B = U\operatorname{diag}(\sqrt{\lambda_i})U^\dagger$:
Hermitian, positive definite, $B^2 = A$. Determinant: product of
the positive eigenvalues.
\end{solution}

\begin{solution}{exo:b2:hermitian:8}
\begin{enumerate}
  \item $\norm{u(x)}^2 = \langle u(x), u(x)\rangle = \langle x,
        u^*u(x)\rangle = \langle x, uu^*(x)\rangle =
        \norm{u^*(x)}^2$. Applying this to the normal $u -
        \lambda\,\mathrm{id}$ (its adjoint is $u^* -
        \conj\lambda$, and normality is inherited): $\norm{(u -
        \lambda)x} = \norm{(u^* - \conj\lambda)x}$, so the kernels
        agree.
  \item For eigenvectors $u(x) = \lambda x$, $u(y) = \mu y$
        ($\lambda \neq \mu$): using (1), $u^*(x) = \conj\lambda x$;
        then
        \[
        \lambda\langle y, x\rangle = \langle y, u(x)\rangle
        = \langle u^*(y), x\rangle = \langle \conj\mu\, y, x\rangle
        = \mu \langle y, x\rangle ,
        \]
        so $\langle y, x\rangle = 0$. Stability of $E_\lambda^\perp$:
        for $x \perp E_\lambda$ and $z \in E_\lambda$, $\langle z,
        u(x)\rangle = \langle u^*(z), x\rangle =
        \langle\conj\lambda z, x\rangle = 0$.
  \item Induction on dimension: over $\C$, $u$ has an eigenvector
        $e_1$ (normalize); its orthogonal complement is stable
        under $u$ (by (2)) \emph{and} under $u^*$ (same argument
        with roles swapped), so the restriction is normal: induct
        and concatenate orthonormal eigenbases. Conversely, a
        unitarily diagonalizable $u = UDU^\dagger$ satisfies $u^*u
        = U\conj D D U^\dagger = U D\conj D U^\dagger = uu^*$:
        normal.
\end{enumerate}
\end{solution}

\begin{solution}{exo:b2:hermitian:9}
Eigenvalues: for $u(x) = \lambda x$, $x \neq 0$:
\[
\lambda\norm x^2 = \langle x, u(x)\rangle
= \langle u^*(x), x\rangle = -\langle u(x), x\rangle
= -\conj{\langle x, u(x)\rangle} = -\conj\lambda\,\norm x^2 ,
\]
so $\lambda = -\conj\lambda$: purely imaginary. Since $(\iu
u)^* = -\iu\,u^*$ (the adjoint is conjugate-linear in scalars),
$u^* = u$ gives $(\iu u)^* = -\iu u$: the map $u \mapsto \iu u$
sends Hermitian to skew-Hermitian, with inverse $w \mapsto -\iu
w$: a bijection. A skew-Hermitian $u$ satisfies $u^*u = -u^2 =
uu^*$: normal, hence unitarily diagonalizable by
\cref{exo:b2:hermitian:8}.
\end{solution}

\begin{solution}{exo:b2:hermitian:10}
\begin{enumerate}
  \item $S$ permutes an orthonormal basis: $\norm{Sx} = \norm
        x$, so $S$ is unitary. Indexing coordinates by $j = 0,
        \dots, n-1$ modulo $n$: $(Sx)_j = x_{j-1}$, so for
        $(f_k)_j = \frac{\omega^{jk}}{\sqrt n}$:
        \[
        (Sf_k)_j = \frac{\omega^{(j-1)k}}{\sqrt n}
        = \omega^{-k}\,(f_k)_j :
        \qquad Sf_k = \omega^{-k}f_k .
        \]
        Orthonormality: $\langle f_k, f_l\rangle = \frac1n
        \sum_j \omega^{j(l-k)} = \delta_{kl}$ (geometric sum
        of a nontrivial root of unity vanishes).
  \item $Cf_k = \sum_j c_j S^jf_k = \bigl(\sum_j
        c_j\omega^{-jk}\bigr)f_k = \widehat c(k)\,f_k$: every
        circulant is diagonal in the orthonormal Fourier basis,
        hence normal, with spectrum $\{\widehat c(k)\}$. (The
        change of basis is the discrete Fourier transform:
        convolution becomes multiplication.)
\end{enumerate}
\end{solution}

\begin{solution}{exo:b2:hermitian:11}
($\Leftarrow$) Let $P^2 = P = P^*$. Every $v$ splits as $v = Pv
+ (v - Pv)$ with $Pv \in \operatorname{im} P$ and $P(v - Pv) =
0$. The two pieces are orthogonal: for any $x, y$,
\[
\langle Px, (I - P)y\rangle = \langle x, P(I-P)y\rangle
= \langle x, (P - P^2)y\rangle = 0 :
\]
$\ker P \perp \operatorname{im} P$, so $P$ is the orthogonal
projection onto its image. ($\Rightarrow$) If $P$ is the
orthogonal projection onto $F = \operatorname{im}P$: for all
$x, y$, $\langle Px, y\rangle = \langle Px, Py\rangle$ (the
component $y - Py \perp F$ drops) and symmetrically $\langle
x, Py\rangle = \langle Px, Py\rangle$: $\langle Px, y\rangle =
\langle x, Py\rangle$, i.e.\ $P^* = P$. Matrices: onto $\C v$
($\norm v = 1$): $Px = v\,\langle v, x\rangle$, i.e.\ $P =
vv^\dagger$; onto $\operatorname{Vect}(v_1, \dots, v_k)$
orthonormal: $P = \sum_i v_iv_i^\dagger$.
\end{solution}

\begin{solution}{exo:b2:hermitian:12}
Diagonalize $A = U D U^\dagger$ (spectral theorem), $D$ diagonal
with entries among the $\lambda_i$. Then $P_i = L_i(A) = U
L_i(D)U^\dagger$, and $L_i(D)$ is diagonal with entries
$L_i(\lambda_j) = \delta_{ij}$: ones exactly at the slots of
$E_i$. So $P_i$ is Hermitian ($L_i$ real, $D$ real), idempotent,
with image $E_i$ and kernel $\bigoplus_{j\neq i}E_j = E_i^\perp$
(orthogonality of eigenspaces): the orthogonal projection onto
$E_i$ (\cref{exo:b2:hermitian:11}). Disjoint diagonal patterns
give $P_iP_j = 0$ ($i \neq j$); $\sum_i L_i = 1$ (degree $< p$,
value $1$ at $p$ points), so $\sum P_i = I$; and $\sum_i
\lambda_iL_i(\lambda_j) = \lambda_j$ gives $A = \sum
\lambda_iP_i$. For any polynomial $f$: $f(D)$ has diagonal
$f(\lambda_j)$, so
\[
f(A) = \sum_{i=1}^{p} f(\lambda_i)\,P_i :
\]
functions of $A$ are computed spectrally --- the calculus that
the Year 3 volume extends to continuous $f$ and beyond.
\end{solution}

\begin{solution}{pb:b2:hermitian:1}
\textbf{1.} $\conj{\langle x, Ax\rangle} = \langle Ax, x\rangle
= \langle x, A^*x\rangle = \langle x, Ax\rangle$: real. Writing
$x = \sum c_ie_i$:
\[
R_A(x) = \frac{\sum_i\lambda_i\abs{c_i}^2}
{\sum_i\abs{c_i}^2} ,
\]
a weighted average of the eigenvalues: it lies in
$\intcc{\lambda_n}{\lambda_1}$, with the bounds attained at
$e_1$ and $e_n$.

\textbf{2.} For real $t$ and any $v$, expand $R_A(x + tv) =
\frac{N(t)}{D(t)}$ with
\[
N(t) = \langle x, Ax\rangle + 2t\Re\langle v, Ax\rangle +
t^2\langle v, Av\rangle,
\quad
D(t) = \norm x^2 + 2t\Re\langle v, x\rangle + t^2\norm v^2 .
\]
The derivative at $t = 0$ is
\[
\frac{2}{\norm x^2}\,
\Re\bigl\langle v,\ Ax - R_A(x)\,x\bigr\rangle .
\]
It vanishes for all $v$ iff $\Re\langle v, w\rangle = 0$ for
all $v$, where $w = Ax - R_A(x)x$; replacing $v$ by $\iu v$
kills the imaginary part too: $w = 0$, i.e.\ $Ax = R_A(x)x$.
Critical points of $R_A$ are exactly eigenvectors, with
critical value the eigenvalue.

\textbf{3.} For $x = \sum_{i\leq k}c_ie_i \in V_k$: $R_A(x)$ is
a weighted average of $\lambda_1, \dots, \lambda_k$, hence
$\geq \lambda_k$, with equality at $e_k$: $\min_{V_k} R_A =
\lambda_k$. Symmetrically on $W_k$ the average involves
$\lambda_k, \dots, \lambda_n$: $\max_{W_k}R_A = \lambda_k$.

\textbf{4.} Let $\dim V = k$. Then $\dim V + \dim W_k = n + 1 >
n$, so there is a unit $x \in V \cap W_k$, and $R_A(x) \leq
\lambda_k$ (question 3): $\min_{V}R_A \leq \lambda_k$ for every
such $V$. Since $V_k$ attains $\lambda_k$, the max-min equals
$\lambda_k$. The min-max formula is the same argument with the
roles reversed ($\dim W = n - k + 1$ forces $W \cap V_k \neq
\{0\}$, so $\max_W R_A \geq \lambda_k$, attained at $W_k$).

\textbf{5.} $R_B(x) = R_A(x) + \frac{\langle x,
(B-A)x\rangle}{\norm x^2} \geq R_A(x)$ pointwise. Taking
$\min$ over any $k$-dimensional $V$ and then $\max$ over $V$:
$\lambda_k(B) \geq \lambda_k(A)$ by question 4.

\textbf{6.} Grassmann: $\dim(V\cap W) = \dim V + \dim W -
\dim(V + W) \geq \dim V + \dim W - n > 0$. Applying this twice:
\[
\dim(V_1\cap V_2\cap V_3)
\geq \dim(V_1\cap V_2) + \dim V_3 - n
\geq \dim V_1 + \dim V_2 + \dim V_3 - 2n .
\]

\textbf{7.} The subspaces $W_i(A)$, $W_j(B)$ (question 3, for
$A$ and $B$) and $V_{i+j-1}(A+B)$ have dimensions $(n-i+1) +
(n-j+1) + (i+j-1) = 2n + 1 > 2n$: by question 6 there is a unit
vector $x$ in all three. Then
\[
\lambda_{i+j-1}(A+B) \leq R_{A+B}(x)
= R_A(x) + R_B(x) \leq \lambda_i(A) + \lambda_j(B),
\]
the left inequality because $x \in V_{i+j-1}(A+B)$ (question
3), the right by the two $W$'s.

\textbf{8.} In a spectral basis of $E$: $\norm{Ex}^2 = \sum
\lambda_k(E)^2\abs{c_k}^2 \leq \max_k\lambda_k(E)^2\,\norm
x^2$, attained at the corresponding eigenvector: $\vertiii E_2
= \max_k\abs{\lambda_k(E)}$. Weyl with $j = 1$: $\lambda_k(A+E)
\leq \lambda_k(A) + \lambda_1(E) \leq \lambda_k(A) + \vertiii
E_2$; applying this to $(A+E) + (-E)$: $\lambda_k(A) \leq
\lambda_k(A+E) + \vertiii E_2$. Together:
$\abs{\lambda_k(A+E) - \lambda_k(A)} \leq \vertiii E_2$.

\textbf{9.} Lower bounds: $P \geq 0$ and question 5. Upper: $P$
has rank $1$, so $\lambda_2(P) = 0$; Weyl with $i = k-1$, $j =
2$:
\[
\lambda_k(A + P) \leq \lambda_{k-1}(A) + \lambda_2(P)
= \lambda_{k-1}(A) .
\]

\textbf{10.} $A + E = \begin{pmatrix} 2 & 1+\iu\\ 1-\iu &
2\end{pmatrix}$: characteristic polynomial $(2-\lambda)^2 -
\abs{1+\iu}^2 = (2-\lambda)^2 - 2$, spectrum $\{2 + \sqrt2,\ 2
- \sqrt2\}$. Against $\operatorname{Sp}A = \{3, 1\}$:
\[
\abs{(2+\sqrt2) - 3} = \abs{(2-\sqrt2) - 1} = \sqrt2 - 1
\approx 0.414 \leq 1 = \vertiii E_2 . \checkmark
\]

\textbf{11.} View $\C^{n-1} = \operatorname{Vect}(e_1, \dots,
e_{n-1})$ inside $\C^n$ (standard basis): for $x$ there,
$\langle x, Bx\rangle = \langle x, Ax\rangle$, so $R_B$ is the
restriction of $R_A$. Upper bound: the max-min for $\lambda_k
(B)$ ranges over $k$-dimensional subspaces \emph{of}
$\C^{n-1}$, a subfamily of those of $\C^n$: $\lambda_k(B) \leq
\lambda_k(A)$. Lower bound: the min-max for $\lambda_k(B)$
ranges over subspaces of $\C^{n-1}$ of dimension $(n-1)-k+1 =
n-k$; each is also a subspace of $\C^n$ of dimension $n -
(k+1) + 1$, so its max is $\geq \lambda_{k+1}(A)$:
$\lambda_k(B) \geq \lambda_{k+1}(A)$.

\textbf{12.} Remove the rows/columns one at a time and chain
question 11: each removal shifts the lower index by one, giving
$\lambda_{k+m}(A) \leq \lambda_k(B) \leq \lambda_k(A)$.

\textbf{13.} Conjugating $A$ by a permutation matrix (unitary)
neither changes the spectrum nor the multiset of diagonal
entries: assume $d_1, \dots, d_k$ occupy the leading positions.
The leading $k\times k$ principal submatrix $B$ then has
$\operatorname{tr} B = \sum_{i\leq k}d_i$, and its eigenvalues
satisfy $\mu_i(B) \leq \lambda_i(A)$ (question 12): summing,
$\sum_{i\leq k}d_i \leq \sum_{i\leq k}\lambda_i(A)$. At $k =
n$ both sides are $\operatorname{tr} A$.

\textbf{14.} Taking $x_i = e_i$ gives the value
$\sum_{i\leq k}\lambda_i$: the max is $\geq$. Conversely, an
orthonormal family $(x_1, \dots, x_k)$ extends to an
orthonormal basis, i.e.\ to a unitary $U$ with first columns
$x_i$; then $\sum_i\langle x_i, Ax_i\rangle$ is the sum of the
first $k$ diagonal entries of $U^\dagger AU$, which by question
13 is at most the sum of its $k$ largest eigenvalues ---
namely $\sum_{i\leq k}\lambda_i(A)$. Ky Fan's maximum
principle follows.

\textbf{15.} Interlacing ($\operatorname{Sp}A = \{2+\sqrt2, 2,
2-\sqrt2\}$, $\operatorname{Sp}B = \{3, 1\}$):
\[
2 \leq 3 \leq 2 + \sqrt2,
\qquad
2 - \sqrt2 \leq 1 \leq 2 . \checkmark
\]
Schur with diagonal $(2,2,2)$: $2 \leq 2+\sqrt2$; $4 \leq 4 +
\sqrt2$; $6 = 6$ (trace). \checkmark

\textbf{16.} $b_{ii} - a_{ii} = \langle e_i, (B-A)e_i\rangle
\geq 0$; summing gives the traces. For any $C$: $\langle x,
C^\dagger(B - A)Cx\rangle = \langle Cx, (B-A)(Cx)\rangle \geq
0$: $C^\dagger AC \leq C^\dagger BC$.

\textbf{17.} $A \geq 0$ clear; $B - A = \begin{pmatrix} 1 & 1\\
1 & 1\end{pmatrix}$ is positive semidefinite (eigenvalues $2,
0$): $A \leq B$. But
\[
B^2 = \begin{pmatrix} 5 & 3\\ 3 & 2\end{pmatrix},
\qquad
B^2 - A^2 = \begin{pmatrix} 4 & 3\\ 3 & 2\end{pmatrix},
\qquad \det(B^2 - A^2) = -1 < 0 :
\]
not positive semidefinite. Squaring does not respect the
Loewner order.

\textbf{18.} Let $S = \sqrt A$, $T = \sqrt B$ (Hermitian
positive semidefinite, \cref{exo:b2:hermitian:7} extended to
semidefinite by the same spectral formula). $T - S$ is
Hermitian; let $\mu$ be any eigenvalue, $v$ a unit eigenvector.
From $T^2 - S^2 = T(T - S) + (T - S)S$:
\[
0 \leq \langle v, (B - A)v\rangle
= \langle Tv, (T-S)v\rangle + \langle (T-S)v, Sv\rangle
= \mu\bigl(\langle v, Tv\rangle + \langle v,
Sv\rangle\bigr)
\]
($\mu$ is real). If $\langle v, Tv\rangle + \langle v,
Sv\rangle > 0$, then $\mu \geq 0$. If it vanishes, both
nonnegative terms vanish; $\langle v, Tv\rangle =
\norm{T^{1/2}v}^2 = 0$ forces $Tv = 0$, likewise $Sv = 0$, so
$\mu v = (T - S)v = 0$ and $\mu = 0$. All eigenvalues of $T -
S$ are $\geq 0$: $\sqrt A \leq \sqrt B$.

\textbf{19.} Congruence by $A^{-1/2}$ (question 16): $I \leq M
:= A^{-1/2}BA^{-1/2}$. So all eigenvalues of $M$ are $\geq 1$,
hence those of $M^{-1}$ lie in $\intoc{0}{1}$: $M^{-1} \leq I$.
But $M^{-1} = A^{1/2}B^{-1}A^{1/2}$; congruating $M^{-1} \leq
I$ by $A^{-1/2}$ gives $B^{-1} \leq A^{-1}$.

\textbf{20.} $\sqrt A^{-1}(AB)\sqrt A = \sqrt A\,B\sqrt A$: so
$AB$ is similar to the Hermitian positive definite $\sqrt
A\,B\sqrt A$ (definite: $\langle x, \sqrt AB\sqrt Ax\rangle =
\langle \sqrt Ax, B\sqrt Ax\rangle > 0$): its eigenvalues are
real and positive. Moreover
\[
\langle x, \sqrt AB\sqrt Ax\rangle
\leq \lambda_{\max}(B)\,\norm{\sqrt Ax}^2
= \lambda_{\max}(B)\,\langle x, Ax\rangle
\leq \lambda_{\max}(A)\lambda_{\max}(B)\norm x^2 ,
\]
so $\lambda_{\max}(AB) = \max R_{\sqrt AB\sqrt A} \leq
\lambda_{\max}(A)\lambda_{\max}(B)$.

\textbf{21.} With $v_k = (\sin j\theta_k)_j$ and the identity
$\sin((j-1)\theta) + \sin((j+1)\theta) =
2\sin(j\theta)\cos\theta$: for $2 \leq j \leq n-1$,
\[
(T_nv_k)_j = -\sin((j{-}1)\theta_k) + 2\sin(j\theta_k) -
\sin((j{+}1)\theta_k)
= (2 - 2\cos\theta_k)\sin(j\theta_k) .
\]
Row $1$ works because $\sin(0\cdot\theta_k) = 0$, row $n$
because $\sin((n+1)\theta_k) = \sin(k\pi) = 0$: the boundary
conditions select exactly $\theta_k = \frac{k\pi}{n+1}$. So
$T_nv_k = 4\sin^2\bigl(\frac{k\pi}{2(n+1)}\bigr)v_k$; the $n$
values are distinct in $\intoo04$ and the $v_k \neq 0$: this is
the whole spectrum, positive, so $T_n$ is positive definite.

\textbf{22.} $\min_id_i\,I \leq D \leq \max_id_i\,I$, so $T_n +
\min_id_i\,I \leq T_n + D \leq T_n + \max_id_i\,I$
(adding $T_n$ preserves the order); question 5 and
$\lambda_k(T_n + cI) = \lambda_k(T_n) + c$ give the sandwich.

\textbf{23.} $T_2 = \begin{pmatrix} 2 & -1\\ -1 &
2\end{pmatrix}$ has spectrum $\{3, 1\}$, and
\[
2 - \sqrt2 \;\leq\; 1 \;\leq\; 2 \;\leq\; 3 \;\leq\; 2 +
\sqrt2 :
\]
Cauchy interlacing, verified. (These are the same spectra as in
question 15: conjugating by $\operatorname{diag}(1,-1,1)$ flips
the sign of the off-diagonal.) Graph reading: $T_2$ is the
Laplacian-type matrix of the path with the last vertex deleted
--- a principal submatrix, exactly the situation of question
11.

\textbf{24.} The extreme eigenvalues behave as
\[
\lambda_{\min}(T_n) = 4\sin^2\frac{\pi}{2(n+1)}
\sim \frac{\pi^2}{(n+1)^2},
\qquad
\lambda_{\max}(T_n) = 4\cos^2\frac{\pi}{2(n+1)}
\longrightarrow 4 .
\]
Hence
\[
\kappa_n = \frac{\lambda_{\max}}{\lambda_{\min}}
\sim \frac{4(n+1)^2}{\pi^2} :
\]
the finer the grid, the worse conditioned the discrete second
derivative --- a fact that drives the design of numerical
linear algebra.

\textbf{25.} (i) Roots of the characteristic polynomial can
move wildly under perturbations of a general matrix, but the
min-max characterization pins each Hermitian eigenvalue between
explicit optimization values, forcing the $1$-Lipschitz
stability of questions 8--9. (ii) Questions 1, 5, 16--20 used
only the extreme Rayleigh values; Weyl, interlacing, Schur and
Ky Fan (questions 7--14) genuinely needed the full min-max
over subspaces. (iii) The Loewner order turns these scalar
inequalities into a calculus of matrix inequalities --- with
real traps (question 17) and real theorems (questions 18--19).
(iv) These tools are the daily bread of numerical analysis
(question 24's stiffness matrices), of quantum perturbation
theory (Weyl: energy levels move by at most the norm of the
perturbation), and of the Year 3 volume's min-max principle
for compact self-adjoint operators.
\end{solution}
