\section*{Chapter \ref{ch:b2:fourier} --- Fourier Series}

\begin{solution}{exo:b2:fourier:1}
Oddness kills the $a_n$. For $n \geq 1$:
\[
b_n = \frac{2}{\pi}\int_0^{\pi} \sin nt\,\dd t
= \frac{2}{\pi}\cdot\frac{1 - (-1)^n}{n}
= \begin{cases} \frac{4}{\pi n} & n \text{ odd},\\ 0 & n
\text{ even}. \end{cases}
\]
So $S(f)(t) = \frac{4}{\pi}\sum_{k\geq0} \frac{\sin\bigl((2k+1)t
\bigr)}{2k+1}$. Dirichlet at $t = \frac\pi2$ (a continuity point,
value $1$): $\sin\bigl((2k+1)\frac\pi2\bigr) = (-1)^k$, giving
\[
1 = \frac4\pi \sum_{k\geq0}\frac{(-1)^k}{2k+1}
\quad\Longrightarrow\quad
\sum_{k\geq0}\frac{(-1)^k}{2k+1} = \frac{\pi}{4}
\quad\text{(Leibniz)}.
\]
At the jump $t = 0$: the series sums to $0 = \frac{f(0^+) +
f(0^-)}{2}$, as Dirichlet prescribes (every term vanishes:
consistent).
\end{solution}

\begin{solution}{exo:b2:fourier:2}
Evenness kills the $b_n$; $a_0 = \frac{1}{\pi}\int_{-\pi}^\pi\abs
t\,\dd t = \pi$, and for $n \geq 1$:
\[
a_n = \frac{2}{\pi}\int_0^\pi t\cos nt\,\dd t
= \frac{2}{\pi}\cdot\frac{(-1)^n - 1}{n^2}
= \begin{cases} -\frac{4}{\pi n^2} & n \text{ odd},\\ 0 & n
\text{ even}, \end{cases}
\]
(one integration by parts). Hence
\[
\abs t = \frac{\pi}{2} - \frac{4}{\pi}\sum_{k\geq0}
\frac{\cos\bigl((2k+1)t\bigr)}{(2k+1)^2} ,
\]
with \emph{normal} convergence ($\sum (2k+1)^{-2} < \infty$) --- as
\cref{thm:b2:fourier:parseval}~(1) predicts for this continuous
piecewise-$C^1$ function. At $t = 0$:
\[
0 = \frac\pi2 - \frac4\pi\sum_{k\geq0}\frac{1}{(2k+1)^2}
\quad\Longrightarrow\quad
\sum_{k\geq0}\frac{1}{(2k+1)^2} = \frac{\pi^2}{8} .
\]
Splitting $\sum \frac{1}{n^2}$ into odd and even parts: $S =
\frac{\pi^2}{8} + \frac S4$, so $S = \frac{\pi^2}{6}$: Basel again.
\end{solution}

\begin{solution}{exo:b2:fourier:3}
Evenness: $b_n = 0$; $a_0 = \frac{1}{\pi}\int_{-\pi}^{\pi} t^2 =
\frac{2\pi^2}{3}$; two integrations by parts give $a_n =
\frac{4(-1)^n}{n^2}$ ($n \geq 1$). Hence
\[
t^2 = \frac{\pi^2}{3} + 4\sum_{n\geq1}
\frac{(-1)^n}{n^2}\cos nt
\qquad (\abs t \leq \pi),
\]
normally convergent. At $t = 0$: $0 = \frac{\pi^2}{3} +
4\sum\frac{(-1)^n}{n^2}$, i.e.\ $\sum_{n\geq1}
\frac{(-1)^{n+1}}{n^2} = \frac{\pi^2}{12}$. Parseval:
\[
\frac{1}{2\pi}\int_{-\pi}^{\pi} t^4\,\dd t = \frac{\pi^4}{5}
= \frac{a_0^2}{4} + \frac12\sum_{n\geq1} a_n^2
= \frac{\pi^4}{9} + 8\sum_{n\geq1}\frac{1}{n^4} ,
\]
so $\sum \frac{1}{n^4} = \frac18\bigl(\frac{\pi^4}{5} -
\frac{\pi^4}{9}\bigr) = \frac{\pi^4}{90}$.
\end{solution}

\begin{solution}{exo:b2:fourier:4}
If moreover $f$ is piecewise $C^1$: Parseval (proved) gives
$\norm f_2^2 = \sum\abs{c_n}^2 = 0$, and strict positivity of the
integral of the continuous $\abs f^2$ forces $f = 0$.

For merely continuous $f$, admit general Parseval: same one-line
proof. (Outline of the density route: Fej\'er/Weierstrass-type
trigonometric approximation shows trigonometric polynomials are
$\norm\cdot_2$-dense among continuous periodic functions; since $f
\perp$ all of them, $\norm f_2^2 = \langle f, f - P\rangle \leq
\norm f_2\norm{f - P}_2$ for approximants $P$, forcing $\norm f_2 =
0$.)
\end{solution}

\begin{solution}{exo:b2:fourier:5}
Evenness: $b_n = 0$;
\[
a_n = \frac{2}{\pi}\int_0^\pi \cos(\alpha t)\cos(nt)\,\dd t
= \frac{2}{\pi}\cdot
\frac{(-1)^n\,\alpha\sin(\pi\alpha)}{\alpha^2 - n^2}
\]
(product-to-sum, then integrate; $a_0 =
\frac{2\sin(\pi\alpha)}{\pi\alpha}$). Dirichlet at $t = \pi$ (a
continuity point of the periodic extension, whose one-sided values
agree by evenness):
\[
\cos(\pi\alpha)
= \frac{\sin(\pi\alpha)}{\pi\alpha}
+ \sum_{n\geq1} \frac{2\alpha\sin(\pi\alpha)}{\pi(\alpha^2 -
n^2)}\,(-1)^n\cos(n\pi)
= \frac{\sin(\pi\alpha)}{\pi}\Bigl(\frac{1}{\alpha} +
\sum_{n\geq1}\frac{2\alpha}{\alpha^2 - n^2}\Bigr),
\]
using $(-1)^n\cos n\pi = 1$. Dividing by $\sin(\pi\alpha)/\pi$
gives the cotangent expansion.
\end{solution}

\begin{solution}{exo:b2:fourier:6}
Iterating $c_n(f') = \iu n\,c_n(f)$ ($k$ times, integration by
parts across the $C^{k}$ pieces with matching boundary values):
$c_n(f^{(k)}) = (\iu n)^k c_n(f)$. The coefficients of the
piecewise continuous $f^{(k)}$ are bounded (indeed $\to 0$, Bessel):
\[
\abs{c_n(f)} = \frac{\abs{c_n(f^{(k)})}}{\abs n^k}
= O\bigl(\abs n^{-k}\bigr) .
\]
\end{solution}

\begin{solution}{exo:b2:fourier:7}
Parseval for $f$ and for $f'$ (both legitimate: $f$ is $C^1$, $f'$
piecewise continuous --- indeed continuous):
\[
\frac{1}{2\pi}\int \abs f^2 = \sum_{n\neq0} \abs{c_n}^2
\quad (c_0 = 0 \text{ by the mean-zero hypothesis}),
\qquad
\frac{1}{2\pi}\int \abs{f'}^2 = \sum_{n\neq0} n^2\abs{c_n}^2 .
\]
Termwise, $n^2\abs{c_n}^2 \geq \abs{c_n}^2$ for $\abs n \geq 1$:
the inequality follows. Equality forces $(n^2 - 1)\abs{c_n}^2 = 0$
for all $n$, i.e.\ $c_n = 0$ for $\abs n \geq 2$: $f(t) =
c_1\eu^{\iu t} + c_{-1}\eu^{-\iu t} = a\cos t + b\sin t$ (real
form); conversely such $f$ give equality.
\end{solution}

\begin{solution}{exo:b2:fourier:8}
From \cref{exo:b2:fourier:1}, $S_{2N-1}(x) =
\frac4\pi\sum_{k=0}^{N-1}\frac{\sin((2k+1)x)}{2k+1}$. At $x_N =
\frac{\pi}{2N}$, with $u_k = (2k+1)x_N = \frac{(2k+1)\pi}{2N}$:
\[
S_{2N-1}(x_N) = \frac{4}{\pi}\sum_{k=0}^{N-1}\frac{\sin u_k}{2k+1}
= \frac{4}{\pi}\sum_{k=0}^{N-1}\frac{\sin u_k}{u_k}\cdot
\frac{u_k}{2k+1}
= \frac{2}{\pi}\sum_{k=0}^{N-1}\frac{\sin u_k}{u_k}\cdot
\frac{\pi}{N},
\]
since $\frac{u_k}{2k+1} = \frac{\pi}{2N}$. The points $u_k$ are the
midpoints of the $N$ subintervals of $\intcc{0}{\pi}$ of length
$\frac{\pi}{N}$: the sum is a midpoint Riemann sum of the
continuous $u \mapsto \frac{\sin u}{u}$, hence converges to
\[
\frac{2}{\pi}\int_0^\pi \frac{\sin u}{u}\,\dd u
\approx \frac{2}{\pi}\times 1.8519 \approx 1.179 .
\]
The partial sums near the jump overshoot the value $1$ by $\approx
18\%$ of the half-jump forever: Gibbs's phenomenon, quantified.
\end{solution}

\begin{solution}{exo:b2:fourier:9}
From $\cos 3t = 4\cos^3t - 3\cos t$:
\[
\cos^3 t = \frac{3\cos t + \cos 3t}{4},
\qquad
\sin^2t\,\cos t = \cos t - \cos^3 t
= \frac{\cos t - \cos 3t}{4} .
\]
Each is a trigonometric polynomial, hence equal to its own
Fourier series (uniqueness of coefficients: two expansions
would differ by a trigonometric polynomial with all
coefficients zero). For $\cos^3t$: $a_1 = \frac34$, $a_3 =
\frac14$, all other $a_n$ and all $b_n$ zero; $c_{\pm1} =
\frac38$, $c_{\pm3} = \frac18$. For $\sin^2t\cos t$: $a_1 =
\frac14$, $a_3 = -\frac14$; $c_{\pm1} = \frac18$, $c_{\pm3} =
-\frac18$.
\end{solution}

\begin{solution}{exo:b2:fourier:10}
Direct computation:
\[
c_n = \frac{1}{2\pi}\int_{-\pi}^{\pi}\eu^{(a - \iu n)t}\dd t
= \frac{\eu^{(a-\iu n)\pi} - \eu^{-(a - \iu n)\pi}}
{2\pi(a - \iu n)}
= \frac{(-1)^n\sinh(a\pi)}{\pi(a - \iu n)} ,
\]
using $\eu^{\pm\iu n\pi} = (-1)^n$. At $t = \pi$ the periodic
extension jumps from $\eu^{a\pi}$ to $\eu^{-a\pi}$; Dirichlet
(symmetric partial sums) gives
\[
\cosh(a\pi) = \sum_{n\in\Z}(-1)^n c_n\,
= \frac{\sinh(a\pi)}{\pi}\Bigl(\frac1a +
\sum_{n\geq1}\Bigl(\frac{1}{a - \iu n} +
\frac{1}{a + \iu n}\Bigr)\Bigr)
= \frac{\sinh(a\pi)}{\pi}\Bigl(\frac1a +
\sum_{n\geq1}\frac{2a}{a^2 + n^2}\Bigr) ,
\]
the imaginary parts of the paired terms cancelling. Divide by
$\sinh(a\pi)$:
\[
\coth(\pi a) = \frac{1}{\pi a} +
\sum_{n\geq1}\frac{2a}{\pi(a^2 + n^2)} ,
\]
the hyperbolic twin of \cref{exo:b2:fourier:5}.
\end{solution}

\begin{solution}{exo:b2:fourier:11}
Commutativity: substitute $s = x - t$ and use periodicity of
the integrand. For $c_n(f * g)$, the integrand $(x, t) \mapsto
f(x-t)g(t)\eu^{-\iu nx}$ is continuous; the two iterated
integrals agree (both, as functions of the upper limit of the
outer variable, vanish at the left endpoint and have the same
derivative --- continuity plus
\cref{thm:b2:integration:continuity} justify differentiating
the iterated integral). Hence
\[
c_n(f*g) = \frac{1}{2\pi}\int_{-\pi}^{\pi} g(t)\,\eu^{-\iu nt}
\Bigl(\frac{1}{2\pi}\int_{-\pi}^{\pi}
f(x-t)\,\eu^{-\iu n(x-t)}\dd x\Bigr)\dd t
= c_n(f)\,c_n(g),
\]
the inner integral being $c_n(f)$ for every $t$ (substitution
and periodicity). Finally \cref{lem:b2:fourier:kernel} says
$S_N(f)(x) = \frac{1}{2\pi}\int f(x+u)D_N(u)\dd u$; the
substitution $u \mapsto -t$ and the evenness of $D_N$ turn this
into $(f * D_N)(x)$.
\end{solution}

\begin{solution}{exo:b2:fourier:12}
\begin{enumerate}
  \item With $w = r\eu^{\iu t}$:
        \[
        P_r(t) = 1 + 2\,\Re\frac{w}{1 - w}
        = \Re\frac{1 + w}{1 - w}
        = \frac{1 - \abs w^2}{\abs{1 - w}^2}
        = \frac{1 - r^2}{1 - 2r\cos t + r^2} > 0 .
        \]
        Mean $1$: term-by-term integration of the normally
        convergent series keeps only $n = 0$.
  \item For $\delta \leq \abs t \leq \pi$: $\cos t \leq
        \cos\delta$, so $P_r(t) \leq \frac{1 - r^2}{1 -
        2r\cos\delta + r^2}$, whose denominator tends to $2 -
        2\cos\delta > 0$ while the numerator tends to $0$:
        uniform convergence to $0$ off any neighborhood of
        $0$.
  \item Term-by-term integration (normal convergence in $t$)
        gives $(f * P_r)(x) = \sum_n r^{\abs
        n}c_n(f)\eu^{\iu nx}$. The approximate-identity
        argument: with mean $1$ and positivity,
        \[
        \abs{(f*P_r)(x) - f(x)}
        \leq \frac{1}{2\pi}\int_{-\pi}^{\pi}
        \abs{f(x-t) - f(x)}\,P_r(t)\,\dd t ,
        \]
        split at $\abs t = \delta$: at most $\varepsilon$
        (Heine) plus $2\norm f_\infty\sup_{\delta\leq\abs
        t\leq\pi}P_r \to \varepsilon$: uniform convergence as
        $r \to 1^-$. This is Abel summation of the Fourier
        series --- the Fourier twin of the power-series
        chapter's boundary theory.
\end{enumerate}
\end{solution}

\begin{solution}{pb:b2:fourier:1}
\textbf{1.} Averaging \cref{lem:b2:fourier:kernel} over $n = 0,
\dots, N-1$ (linearity of the integral) gives $\sigma_N(f)(x) =
\frac{1}{2\pi}\int f(x+u)F_N(u)\dd u$; each $D_n$ has mean
$1$, so $F_N$ has mean $1$.

\textbf{2.} With $\eu^{\iu u} - 1 = 2\iu\,\eu^{\iu
u/2}\sin\frac u2$:
\[
\sum_{n=0}^{N-1}\sin\Bigl(\Bigl(n + \frac12\Bigr)u\Bigr)
= \Im\Bigl[\eu^{\iu u/2}\,\frac{\eu^{\iu Nu} - 1}{\eu^{\iu u}
- 1}\Bigr]
= \Re\,\frac{1 - \eu^{\iu Nu}}{2\sin\frac u2}
= \frac{1 - \cos Nu}{2\sin\frac u2}
= \frac{\sin^2\frac{Nu}2}{\sin\frac u2} .
\]
Dividing by $N\sin\frac u2$:
\[
F_N(u) = \frac1N\sum_{n=0}^{N-1}
\frac{\sin\bigl((n+\frac12)u\bigr)}{\sin\frac u2}
= \frac{1}{N}\,
\frac{\sin^2\frac{Nu}{2}}{\sin^2\frac u2} \geq 0 .
\]

\textbf{3.} On $\delta \leq \abs u \leq \pi$: $\sin^2\frac u2
\geq \sin^2\frac\delta2$ and $\sin^2\frac{Nu}2 \leq 1$: $F_N
\leq \frac{1}{N\sin^2(\delta/2)} \to 0$, uniformly there.

\textbf{4.} By the unit mean, $\sigma_Nf(x) - f(x) =
\frac{1}{2\pi}\int\bigl(f(x+u) - f(x)\bigr)F_N(u)\dd u$. Given
$\varepsilon$, Heine yields $\delta$ with $\abs{f(x+u) - f(x)}
\leq \varepsilon$ for $\abs u \leq \delta$, uniformly in $x$.
Then, using $F_N \geq 0$ and its unit mean,
\[
\abs{\sigma_Nf(x) - f(x)}
\leq \varepsilon +
2\norm f_\infty\cdot\frac{1}{2\pi}
\int_{\delta\leq\abs u\leq\pi}F_N
\leq \varepsilon + \frac{2\norm
f_\infty}{N\sin^2\frac\delta2}
\leq 2\varepsilon
\]
for large $N$, uniformly in $x$: Fej\'er's theorem.

\textbf{5.} $F_N$ is even with mean $1$: each half
$\intcc{0}{\pi}$, $\intcc{-\pi}{0}$ carries mean $\frac12$.
Then
\[
\sigma_Nf(x) - \frac{f(x^+)+f(x^-)}{2}
= \frac{1}{2\pi}\int_0^\pi\bigl(f(x+u) -
f(x^+)\bigr)F_N\,\dd u
+ \frac{1}{2\pi}\int_{-\pi}^0\bigl(f(x+u) -
f(x^-)\bigr)F_N\,\dd u ;
\]
on each half, split at $\abs u = \delta$ where the one-sided
limit is $\varepsilon$-close, and let question 3 kill the far
part: both integrals tend to $0$. The bound: $F_N \geq 0$
gives $\abs{\sigma_Nf(x)} \leq
\frac{1}{2\pi}\int\abs{f(x+u)}F_N \leq \norm f_\infty$.

\textbf{6.} Each $\sigma_N(f)$ is a trigonometric polynomial
(an average of the $S_n(f)$, $n < N$), and $\sigma_N(f) \to f$
uniformly: the trigonometric Weierstrass theorem.

\textbf{7.} $c_n(f) = 0$ for all $n$ makes every $S_n(f) = 0$,
hence every $\sigma_N(f) = 0$; by Fej\'er, $f = \lim\sigma_Nf =
0$. Applying this to a difference: continuous periodic
functions are determined by their Fourier coefficients.

\textbf{8.} $S_Nf$ is the orthogonal projection of $f$ onto
$\mathcal T_N$ (\cref{prop:b2:fourier:bessel}), so it minimizes
$\norm{f - P}_2$ over $P \in \mathcal T_N$; since $\sigma_Nf
\in \mathcal T_{N-1} \subseteq \mathcal T_N$:
\[
\norm{f - S_Nf}_2 \leq \norm{f - \sigma_Nf}_2
\leq \norm{f - \sigma_Nf}_\infty \to 0
\]
(the middle inequality because the mean of $\abs\cdot^2$ is at
most the sup squared). Pythagoras $\norm f_2^2 = \sum_{\abs
n\leq N}\abs{c_n}^2 + \norm{f - S_Nf}_2^2$ then passes to the
limit: Parseval for every continuous periodic $f$.

\textbf{9.} Let $t_1, \dots, t_p$ be the jumps of $f$ in a
period, $M = \norm f_\infty$. For small $\eta$, define $g = f$
outside the intervals $\intoo{t_j - \eta}{t_j + \eta}$ and by
the affine chord across each such interval: $g$ is continuous,
periodic, $\norm g_\infty \leq M$, and
\[
\norm{f - g}_2^2 \leq \frac{1}{2\pi}\,p\cdot(2M)^2\cdot2\eta
\leq \varepsilon^2
\]
for $\eta$ small. Bessel makes $S_N$ a contraction for
$\norm\cdot_2$, so
\[
\norm{f - S_Nf}_2
\leq \norm{f - g}_2 + \norm{g - S_Ng}_2 + \norm{S_N(g -
f)}_2
\leq 2\varepsilon + \norm{g - S_Ng}_2 ,
\]
and question 8 gives $\limsup_N\norm{f - S_Nf}_2 \leq
2\varepsilon$ for every $\varepsilon$: $\norm{f - S_Nf}_2 \to
0$, and Pythagoras yields Parseval for every piecewise
continuous $f$: the ``admitted'' in
\cref{thm:b2:fourier:parseval} is now a theorem.

\textbf{10.} The square wave has $\norm f_\infty = 1$, so
question 5 gives $\abs{\sigma_N(f)} \leq 1$ everywhere and for
every $N$ --- no overshoot, ever --- while
\cref{exo:b2:fourier:8} shows $\sup_xS_{2N-1}(f)(x) \to
\approx 1.179$. The one responsible property: $F_N \geq 0$, so
$\sigma_Nf(x)$ is a weighted \emph{average} of values of $f$
and can never leave $\intcc{\min f}{\max f}$; $D_N$ takes
negative values, so $S_N$ can.

\textbf{11.} $\abs{\sin N\theta} \leq N\abs{\sin\theta}$ by
induction ($\abs{\sin(N{+}1)\theta} \leq
\abs{\sin N\theta}\abs{\cos\theta} +
\abs{\cos N\theta}\abs{\sin\theta} \leq
(N+1)\abs{\sin\theta}$): with $\theta = \frac u2$,
\[
F_N(u) = \frac{\sin^2\frac{Nu}2}{N\sin^2\frac u2}
\leq \frac{N^2\sin^2\frac u2}{N\sin^2\frac u2} = N .
\]
Concavity of $\sin$ on $\intcc{0}{\frac\pi2}$ gives
$\sin\frac u2 \geq \frac{u}{\pi}$ for $0 \leq u \leq \pi$, so
$F_N(u) \leq \frac{1}{N(u/\pi)^2} = \frac{\pi^2}{Nu^2}$.

\textbf{12.} By evenness and the split at $\frac1N$:
\[
\frac{1}{2\pi}\int_{-\pi}^{\pi}\abs uF_N
= \frac1\pi\int_0^\pi uF_N
\leq \frac1\pi\Bigl(\int_0^{1/N}uN\,\dd u +
\int_{1/N}^{\pi}\frac{\pi^2}{Nu}\,\dd u\Bigr)
= \frac{1}{2\pi N} + \frac{\pi\ln(\pi N)}{N} .
\]
For $N \geq 2$: $\ln(\pi N) \leq \bigl(1 +
\frac{\ln\pi}{\ln2}\bigr)\ln N \leq 2.66\ln N$ and
$\frac{1}{2\pi N} \leq \frac{\ln N}{N}$, so the moment is
$\leq \frac{9\ln N}{N}$.

\textbf{13.} For $L$-Lipschitz $f$:
\[
\abs{\sigma_Nf(x) - f(x)}
\leq \frac{1}{2\pi}\int\abs{f(x+u) - f(x)}F_N(u)\dd u
\leq L\cdot\frac{1}{2\pi}\int\abs uF_N
\leq \frac{9L\ln N}{N},
\]
uniformly in $x$.

\textbf{14.} $c_0(e_1) = 0$ gives $S_0(e_1) = 0$, while
$S_n(e_1) = e_1$ for $n \geq 1$: $\sigma_N(e_1) =
\frac{N-1}{N}e_1$, so $\norm{\sigma_Ne_1 - e_1}_\infty =
\frac1N$. Even for this entire, band-limited signal the rate is
$\frac1N$: Fej\'er saturates, exactly as Bernstein's operator
saturates at $\frac1n$ (Voronovskaya, in the
function-sequences chapter's weekend problem).

\textbf{15.} If $f$ vanishes on $\intoo{x-\delta}{x+\delta}$,
then $\sigma_Nf(x) = \frac{1}{2\pi}\int_{\delta \leq \abs u
\leq \pi}f(x+u)F_N(u)\dd u$, of absolute value at most
$\frac{\norm f_\infty}{N\sin^2(\delta/2)} \to 0$: the Ces\`aro
means at $x$ only see $f$ near $x$.

\textbf{16.} Given $\intcc ab$ and $\varepsilon$, choose
continuous $1$-periodic piecewise-affine $\varphi^\pm$ with
$\varphi^- \leq \mathbf 1_{\intcc ab} \leq \varphi^+$ and
$\int_0^1(\varphi^+ - \varphi^-) \leq \varepsilon$ (trapezoids
with slopes over intervals of total length $\varepsilon$).
Then
\[
\limsup_N\frac{\#\{n \leq N : x_n \in \intcc ab\}}{N}
\leq \lim_N\frac1N\sum_{n\leq N}\varphi^+(x_n)
= \int_0^1\varphi^+ \leq b - a + \varepsilon ,
\]
and symmetrically $\liminf \geq b - a - \varepsilon$: the
proportion tends to $b - a$.

\textbf{17.} For $f = \eu^{2\iu\pi k\cdot}$ with $k \neq 0$ the
hypothesis gives the limit $0 = \int_0^1f$; for $k = 0$ both
sides are $1$; linearity handles every trigonometric
polynomial. For continuous $1$-periodic $f$ and $\varepsilon >
0$, question 6 (transported by $t = 2\pi x$) provides a
trigonometric polynomial $P$ with $\norm{f - P}_\infty \leq
\varepsilon$:
\[
\Bigl|\frac1N\sum_{n\leq N}f(x_n) - \int_0^1f\Bigr|
\leq 2\varepsilon +
\Bigl|\frac1N\sum_{n\leq N}P(x_n) - \int_0^1P\Bigr|
\longrightarrow 2\varepsilon .
\]
With question 16: $(x_n)$ is equidistributed.

\textbf{18.} For $k \neq 0$ and irrational $\alpha$, $w =
\eu^{2\iu\pi k\alpha} \neq 1$:
\[
\Bigl|\sum_{n=1}^{N}w^n\Bigr|
= \Bigl|\frac{w(w^N - 1)}{w - 1}\Bigr|
\leq \frac{2}{\abs{1 - w}} ,
\]
a bound independent of $N$; dividing by $N$ gives Weyl's
criterion, and question 17 concludes: $(\{n\alpha\})$ is
equidistributed.

\textbf{19.} Every subinterval receives asymptotic proportion
equal to its length, in particular infinitely many points:
$(\{n\alpha\})$ is dense. Equidistribution is stronger: a
sequence can be dense while spending nearly all its time in one
corner (density says where the sequence \emph{goes},
equidistribution says \emph{how often}).

\textbf{20.} $2^n$ has leading digit $1$ iff $10^m \leq 2^n <
2\cdot10^m$ for some $m$, i.e.\ iff $\{n\log_{10}2\} \in
\intco{0}{\log_{10}2}$. Irrationality: $\log_{10}2 =
\frac pq$ would give $2^q = 10^p = 2^p5^p$, impossible for $p
\geq 1$ by unique factorization. Weyl's theorem (question 18
with $\alpha = \log_{10}2$; the half-open interval is squeezed
between closed ones of nearby lengths) gives the proportion
$\log_{10}2 \approx 0.301$: the first digits of $2^n$ follow
Benford's law.

\textbf{21.} $z$ is $C^1$, so its Fourier series converges
normally with sum $z$ (\cref{thm:b2:fourier:parseval}~(1)),
and $c_n(z') = \iu n\,c_n$. Since $\abs{z'} = \frac{L}{2\pi}$
is constant,
\[
\int_0^{2\pi}\abs{z'}^2\dd t
= 2\pi\Bigl(\frac{L}{2\pi}\Bigr)^{\!2}
= \frac{L^2}{2\pi} ,
\]
and Parseval applied to the continuous $z'$ gives
$\frac{1}{2\pi}\int_0^{2\pi}\abs{z'}^2 =
\sum_n\abs{\iu nc_n}^2$, i.e.\ $\frac{L^2}{2\pi} = 2\pi\sum_n
n^2\abs{c_n}^2$.

\textbf{22.} The polarized Parseval $\frac{1}{2\pi}\int\conj
fg = \sum\conj{c_n(f)}c_n(g)$ follows from Parseval applied to
$f + g$ and $f + \iu g$ (polarization identity), both
continuous. With $f = z$, $g = z'$:
\[
A = \frac12\,\Im\int_0^{2\pi}\conj z\,z'
= \pi\,\Im\sum_n\conj{c_n}(\iu n c_n)
= \pi\sum_n n\abs{c_n}^2 .
\]

\textbf{23.} Combining questions 21--22:
\[
L^2 - 4\pi A
= 4\pi^2\sum_n n^2\abs{c_n}^2 -
4\pi^2\sum_n n\abs{c_n}^2
= 4\pi^2\sum_{n\in\Z}(n^2 - n)\abs{c_n}^2 \geq 0 ,
\]
since $n^2 - n = n(n-1) \geq 0$ for every integer. Equality
forces $c_n = 0$ for all $n \notin \{0, 1\}$: $z(t) = c_0 +
c_1\eu^{\iu t}$, a circle of center $c_0$ and radius
$\abs{c_1} = \frac{L}{2\pi}$ (constant speed). Hurwitz's proof
of the isoperimetric inequality: $A \leq \frac{L^2}{4\pi}$,
circle only.

\textbf{24.} Circle of radius $R$: $L = 2\pi R$, $A = \pi
R^2$: $L^2 = 4\pi^2R^2 = 4\pi A$: equality. Square of side
$a$: $L^2 = 16a^2 > 4\pi a^2 = 4\pi A$ (as $16 > 4\pi \approx
12.57$). For negative $n$, $n^2 - n = n(n - 1)$ is a product
of two negative integers: positive --- so backward-winding
modes cost area twice. Constant speed entered in question 21,
converting $\int\abs{z'}^2$ into $\frac{L^2}{2\pi}$; for a
non-constant speed, Cauchy--Schwarz gives $\int\abs{z'}^2 \geq
\frac{(\int\abs{z'})^2}{2\pi} = \frac{L^2}{2\pi}$, so the
inequality survives, with the circle still the only equality
case.

\textbf{25.} (i) Everything flows from $F_N \geq 0$ (with unit
mean and concentration); $D_N$ has unit mean and concentration
of oscillation but not positivity, and Gibbs is the price.
(ii) Ces\`aro summability of the Fourier series implies its
Abel summability with the same sum (Frobenius, proved in the
power-series chapter's weekend problem) --- the Poisson-kernel
route of \cref{exo:b2:fourier:12} is exactly Abel's method.
(iii) Weyl's theorem needed only uniform approximation
(question 6); the isoperimetric inequality needed Parseval
itself (questions 8, 21--22). (iv) The Year 3 volume proves
completeness: the exponentials form a Hilbert basis of $L^2$,
Parseval becomes an isometry of Hilbert spaces, and Fej\'er's
theorem becomes the statement that this isometry is computable
by positive averages.
\end{solution}
