\section*{Chapter \ref{ch:b2:diffcalc} --- Differential Calculus}

\begin{solution}{exo:b2:diffcalc:1}
$J_f = \begin{pmatrix} 2x & -2y\\ 2y & 2x\end{pmatrix}$, $\det J_f
= 4(x^2 + y^2)$: invertible away from the origin. (This $f$ is $z
\mapsto z^2$ in complex disguise.)

$J_\Phi = \begin{pmatrix} \cos\theta & -r\sin\theta & 0\\
\sin\theta & r\cos\theta & 0\\ 0 & 0 & 1\end{pmatrix}$, $\det = r$:
invertible for $r \neq 0$ (cylindrical coordinates).
\end{solution}

\begin{solution}{exo:b2:diffcalc:2}
For $v = (a, b) \neq 0$: $\frac{f(tv) - 0}{t} = \frac{t^3a^3}{t\cdot
t^2(a^2+b^2)} = \frac{a^3}{a^2 + b^2}$: every directional
derivative exists, with value $D_v = \frac{a^3}{a^2+b^2}$. But $v
\mapsto D_v$ is not linear ($D_{(1,0)} = 1$, $D_{(0,1)} = 0$,
$D_{(1,1)} = \frac12 \neq 1$): no linear map can produce these
values, so $f$ is not \omterm{def:b2:diffcalc:differential}{differentiable} at $0$ (the \omterm{def:b2:diffcalc:differential}{differential} would
have to be $v \mapsto D_v$).
\end{solution}

\begin{solution}{exo:b2:diffcalc:3}
$f = x^3 + y^3 - 3xy$: critical points $(0,0)$ and $(1,1)$ (Year 1
computation). Hessians: $H = \begin{pmatrix} 6x & -3\\ -3 &
6y\end{pmatrix}$. At $(0,0)$: eigen-signs mixed ($\det = -9 < 0$):
saddle. At $(1,1)$: $\det = 27 > 0$, trace $> 0$: positive
definite, strict local minimum --- now justified by
\cref{thm:b2:diffcalc:extrema} rather than decreed.

$g = x^4 + y^4 - 2(x-y)^2$: $\nabla g = (4x^3 - 4(x - y),\; 4y^3 +
4(x-y))$; critical points $(0,0)$, $(\sqrt2, -\sqrt2)$,
$(-\sqrt2, \sqrt2)$ (Year 1). At $(\pm\sqrt2, \mp\sqrt2)$: $H =
\begin{pmatrix} 12\cdot2 - 4 & 4\\ 4 & 20\end{pmatrix} =
\begin{pmatrix} 20 & 4\\ 4 & 20\end{pmatrix}$: positive definite
(diagonally dominant; \omterm{def:b2:reduction:eigen}{eigenvalues} $24, 16$): strict local minima.
At $(0,0)$: $H = \begin{pmatrix} -4 & 4\\ 4 & -4\end{pmatrix}$,
singular negative semidefinite: the test is silent; the directional
study ($g(x,x) = 2x^4 > 0$, $g(x,-x) = 2x^4 - 8x^2 < 0$ small)
shows a saddle-like point --- no extremum.
\end{solution}

\begin{solution}{exo:b2:diffcalc:4}
Differentiate $t \mapsto f(tx)$ at $t = 1$: by the chain rule,
$\langle \nabla f(x), x\rangle$; by homogeneity the same function
is $t^pf(x)$, of derivative $pf(x)$ at $t = 1$: Euler's identity.

Converse: fix $x \neq 0$ and let $\varphi(t) = f(tx) - t^p f(x)$
on $t > 0$. Then $\varphi'(t) = \langle\nabla f(tx), x\rangle -
pt^{p-1}f(x) = \frac1t\bigl(\langle \nabla f(tx), tx\rangle -
p\,t^pf(x)\bigr)$. The hypothesis --- Euler's identity at the point
$tx$ --- evaluates the bracket as $p\,f(tx) - p\,t^pf(x) =
p\,\varphi(t)$. So $\varphi' = \frac{p}{t}\varphi$ with $\varphi(1)
= 0$: the linear ODE's unique solution is $\varphi \equiv 0$ (Year
1 uniqueness), i.e.\ $f(tx) = t^pf(x)$.
\end{solution}

\begin{solution}{exo:b2:diffcalc:5}
Expanding $f(x + h) - f(x) = \langle Ax - b, h\rangle +
\frac12\langle Ah, h\rangle$ (symmetry of $A$): $\nabla f(x) = Ax -
b$ and $H_f = A$ everywhere. $f$ is convex iff $A$ is positive
semidefinite (the Hessian test, global here since $H$ is constant:
the second-order Taylor formula is exact). If $A$ is positive
definite: the unique critical point is $x^* = A^{-1}b$, and $f(x^*
+ h) - f(x^*) = \frac12\langle Ah, h\rangle \geq
\frac{\lambda_{\min}}{2}\norm h^2 > 0$ for $h \neq 0$: strict
global minimum.
\end{solution}

\begin{solution}{exo:b2:diffcalc:6}
Since $\nabla g(a) \neq 0$, one partial, say $\frac{\partial
g}{\partial y}(a) \neq 0$: the implicit function theorem
(\cref{thm:b2:diffcalc:inverse}'s companion) parametrizes $\{g =
0\}$ near $a = (a_1, a_2)$ as $\gamma(t) = (t, y(t))$ with $y$
$C^1$, $y'(t) = -\frac{\partial_x g}{\partial_y g}(\gamma(t))$
(differentiate $g(t, y(t)) = 0$). The one-variable function $t
\mapsto f(\gamma(t))$ has a local extremum at $t = a_1$:
\[
0 = \frac{\dd}{\dd t}f(\gamma(t))\Big|_{a_1}
= \partial_x f(a) + \partial_y f(a)\,y'(a_1)
= \partial_x f(a) - \partial_yf(a)\frac{\partial_x
g(a)}{\partial_y g(a)} :
\]
the vectors $\nabla f(a)$ and $\nabla g(a)$ have proportional
coordinates: $\nabla f(a) = \lambda \nabla g(a)$ with $\lambda =
\frac{\partial_y f(a)}{\partial_y g(a)}$.

Application: on the circle, $\nabla(xy) = (y, x)$ parallel to
$(2x, 2y)$ forces $y^2 = x^2$; with the constraint, the candidates
are $\pm\bigl(\tfrac{1}{\sqrt2}, \tfrac{1}{\sqrt2}\bigr)$ (value
$\frac12$) and $\pm\bigl(\tfrac{1}{\sqrt2},
-\tfrac{1}{\sqrt2}\bigr)$ (value $-\frac12$): max $\frac12$, min
$-\frac12$ (attained: the circle is \omterm{def:b2:metric:compact}{compact}).
\end{solution}

\begin{solution}{exo:b2:diffcalc:7}
$\nabla f = (2x - y + 1,\; 2y - x - 1) = 0$: solving, $x =
-\frac13$, $y = \frac13$. Hessian $\begin{pmatrix} 2 & -1\\ -1 &
2\end{pmatrix}$, positive definite: global minimum of the quadratic
$f$, value $f\bigl(-\frac13, \frac13\bigr) = -\frac13$.

On the triangle $T$: the interior critical point $(-\frac13,
\frac13) \notin T$ (negative $x$). Edges: on $y = 0$, $x \in
\intcc{0}{1}$: $f = x^2 + x$, increasing: extremes $0$ and $2$. On
$x = 0$: $f = y^2 - y$, minimum $-\frac14$ at $y = \frac12$, values
$0$ and $0$ at the ends. On $x + y = 1$: substitute $y = 1 - x$,
$f = x^2 + (1-x)^2 - x(1-x) + x - (1-x) = 3x^2 - x$; on
$\intcc{0}{1}$: minimum $-\frac{1}{12}$ at $x = \frac16$, values
$0$ (at $x=0$) and $2$ (at $x=1$). Vertices: $f(0,0) = 0$, $f(1,0)
= 2$, $f(0,1) = 0$. Global on $T$: minimum $-\frac14$ at $(0,
\frac12)$, maximum $2$ at $(1, 0)$.
\end{solution}

\begin{solution}{exo:b2:diffcalc:8}
$J_f = \begin{pmatrix} 1 & 2y\\ 2x & 1\end{pmatrix}$, $\det J_f =
1 - 4xy$. At $0$: $\det = 1 \neq 0$: local diffeomorphism
(\cref{thm:b2:diffcalc:inverse}), with
\[
\dd(f^{-1})_{(0,0)} = (J_f(0))^{-1} = I_2 .
\]
Invertibility on a ball: need $4\abs{xy} < 1$ throughout; on
$\norm{(x,y)}_2 < r$, $\abs{xy} \leq \frac{x^2 + y^2}{2} <
\frac{r^2}{2}$, so $r = \frac{1}{\sqrt2}$ works; and it is largest:
at $(x, y) = \bigl(\tfrac12, \tfrac12\bigr)$, of \omterm{def:b2:nvs:norm}{norm}
$\frac{1}{\sqrt2}$, $\det J_f = 0$.
\end{solution}

\begin{solution}{exo:b2:diffcalc:9}
$f(t) = (\cos t, \sin t)$: $f(0) = f(2\pi) = (1, 0)$, yet $f'(t) =
(-\sin t, \cos t)$ has \omterm{def:b2:nvs:norm}{norm} $1$, never zero: no point where the
derivative vanishes --- Rolle has no vector analogue. The mean
value inequality holds comfortably: $\norm{f(2\pi) - f(0)} = 0 \leq
2\pi \cdot \sup\norm{f'} = 2\pi$.
\end{solution}

\begin{solution}{exo:b2:diffcalc:10}
$f(x+h) - f(x) = 2\langle x, h\rangle + \norm h^2$: $\dd f_x =
2\langle x, \cdot\rangle$, $\nabla f(x) = 2x$, Hessian $2I$
(constant). For $g$:
\[
g(x + h) - g(x) = \langle Ax, h\rangle + \langle Ah, x\rangle
+ \langle Ah, h\rangle
= \bigl\langle (A + A^{\mathsf T})x,\ h\bigr\rangle +
O(\norm h^2),
\]
so $\nabla g(x) = (A + A^{\mathsf T})x$ and $H_g = A +
A^{\mathsf T}$, constant. By \cref{exo:b2:diffcalc:11}, $g$ is
convex iff $A + A^{\mathsf T}$ is positive semidefinite ---
only the \omterm{def:b2:quadratic:adjoint}{symmetric} part of $A$ matters, as indeed $g(x) =
\langle \frac{A + A^{\mathsf T}}2 x, x\rangle$.
\end{solution}

\begin{solution}{exo:b2:diffcalc:11}
$f$ is convex iff its restriction to every segment is convex,
i.e.\ iff every $\varphi(t) = f(a + tv)$ is convex. By the chain
rule $\varphi''(t) = \langle H_{a+tv}\,v,\ v\rangle$.

If all Hessians are positive semidefinite: $\varphi'' \geq 0$,
so each $\varphi$ is convex (Year 1 volume) and $f$ is convex.
Conversely if $f$ is convex, each $\varphi$ is convex, so
$\varphi''(0) \geq 0$: $\langle H_a v, v\rangle \geq 0$ for
every $a$ and every direction $v$: all Hessians are positive
semidefinite.
\end{solution}

\begin{solution}{exo:b2:diffcalc:12}
\begin{enumerate}
  \item By the mean value inequality
        (\cref{thm:b2:diffcalc:mvi}) applied to $g$:
        $\norm{g(x) - g(y)} \leq k\norm{x-y}$, so
        \[
        \norm{f(x) - f(y)} \geq \norm{x - y} - \norm{g(x) -
        g(y)} \geq (1 - k)\norm{x - y} :
        \]
        $f$ is injective and $f^{-1}$ (defined on the image) is
        $\frac{1}{1-k}$-Lipschitz.
  \item Fix $y$; $T(x) = y - g(x)$ satisfies $\norm{T(x) -
        T(x')} = \norm{g(x') - g(x)} \leq k\norm{x - x'}$: a
        contraction of the \omterm{def:b2:metric:complete}{complete} space $\R^n$. Banach
        (\cref{thm:b2:metric:banach}) gives a fixed point $x^*
        = y - g(x^*)$, i.e.\ $f(x^*) = y$: $f$ is surjective.
  \item $f$ is thus a bijection of $\R^n$ with
        $\frac{1}{1-k}$-Lipschitz inverse: a global inverse
        theorem, where the smallness of $\dd g$ everywhere
        replaces the local invertibility hypothesis of
        \cref{thm:b2:diffcalc:inverse}.
\end{enumerate}
\end{solution}

\begin{solution}{pb:b2:diffcalc:1}
\textbf{1.} $\norm{ABx} \leq \vertiii A\norm{Bx} \leq
\vertiii A\vertiii B\norm x$: take the sup over $\norm x = 1$.
Matrix products, $\det$ and $\operatorname{tr}$ are polynomial
functions of the entries, hence \omterm{def:b2:metric:continuity}{continuous} ($\mathcal M_n(\R)
\simeq \R^{n^2}$, all \omterm{def:b2:nvs:norm}{norms} equivalent:
\cref{thm:b2:nvs:finitedim}).

\textbf{2.} $\sum\vertiii{X^k} \leq \sum\vertiii X^k <
\infty$: the series converges \omterm{def:b2:series:def}{absolutely}, hence converges
(\cref{thm:b2:nvs:absoluteconvergence}). From $(I -
X)\sum_{k\leq N}X^k = I - X^{N+1} \to I$: the sum is $(I -
X)^{-1}$. \omterm{def:b2:nvs:norm}{Norm}: $\leq \sum\vertiii X^k = \frac{1}{1 -
\vertiii X}$; and
\[
(I - X)^{-1} - I - X = \sum_{k\geq2}X^k = X^2(I - X)^{-1},
\qquad
\vertiii{X^2(I-X)^{-1}} \leq
\frac{\vertiii X^2}{1 - \vertiii X} = O(\vertiii X^2).
\]

\textbf{3.} $A + H = A(I + A^{-1}H)$ with $\vertiii{A^{-1}H}
\leq \vertiii{A^{-1}}\vertiii H < 1$: invertible by question 2:
the \omterm{def:b2:metric:topology}{open} ball of radius $\vertiii{A^{-1}}^{-1}$ around $A$ lies
in $GL_n(\R)$. Density: $\det(A + \varepsilon I)$ is a degree-$n$
polynomial in $\varepsilon$ with leading coefficient $1$: it
has finitely many roots, so there are $\varepsilon_k \to 0$
with $A + \varepsilon_kI$ invertible, converging to $A$.

\textbf{4.} For $\vertiii H < \frac{1}{2\vertiii{A^{-1}}}$:
\[
(A + H)^{-1} - A^{-1}
= \bigl[(I + A^{-1}H)^{-1} - I\bigr]A^{-1},
\]
of \omterm{def:b2:nvs:norm}{norm} at most $\frac{\vertiii{A^{-1}H}}{1 -
\vertiii{A^{-1}H}}\,\vertiii{A^{-1}} \leq
2\vertiii{A^{-1}}^2\vertiii H \to 0$: $\Phi$ is \omterm{def:b2:metric:continuity}{continuous} at
every $A \in GL_n(\R)$.

\textbf{5.} $(A+H)^2 = A^2 + AH + HA + H^2$: the map $H
\mapsto AH + HA$ is linear and the error $H^2$ is
$O(\vertiii H^2)$. Expanding $(A + H)^k$ and sorting by the
number of $H$-factors: the linear part is
$\sum_{i=0}^{k-1}A^iHA^{k-1-i}$, and the terms with $\geq 2$
factors $H$ are bounded by $\binom k2$-many products of \omterm{def:b2:nvs:norm}{norm}
$\leq \vertiii A^{k-2}\vertiii H^2$-scale: $O(\vertiii H^2)$.
One cannot collapse the sum to $kA^{k-1}H$ because $H$ and $A$
need not commute --- the sum is the correct noncommutative
derivative.

\textbf{6.} For small $H$:
\[
(A+H)^{-1} = (I + A^{-1}H)^{-1}A^{-1}
= \bigl(I - A^{-1}H + O(\vertiii H^2)\bigr)A^{-1}
= A^{-1} - A^{-1}HA^{-1} + O(\vertiii H^2) :
\]
$\dd\Phi_A(H) = -A^{-1}HA^{-1}$, linear in $H$. For $n = 1$:
$\dd(1/a)(h) = -h/a^2$, the familiar derivative.

\textbf{7.} Chain rule along the curve:
$\bigl(A(t)^{-1}\bigr)' = \dd\Phi_{A(t)}(A'(t)) =
-A(t)^{-1}A'(t)A(t)^{-1}$. At $A(t) = I + tB$, $t = 0$: $(I +
tB)^{-1} = I - tB + O(t^2)$.

\textbf{8.} By question 5 and cyclic invariance of the trace:
\[
\dd f_A(H) = \operatorname{tr}\Bigl(\sum_{i=0}^{k-1}
A^iHA^{k-1-i}\Bigr) = k\operatorname{tr}\bigl(A^{k-1}H\bigr) .
\]
Against the Frobenius product, $\dd f_A(H) =
\operatorname{tr}\bigl((\nabla f)^{\mathsf T}H\bigr)$ requires
$(\nabla f)^{\mathsf T} = kA^{k-1}$: $\nabla f(A) =
k\,(A^{k-1})^{\mathsf T}$.

\textbf{9.} $f(A + H) - f(A) = 2\operatorname{tr}
(A^{\mathsf T}H) + \operatorname{tr}(H^{\mathsf T}H)$: the
\omterm{def:b2:diffcalc:differential}{differential} is $H \mapsto 2\operatorname{tr}(A^{\mathsf T}H) =
2\langle A, H\rangle$, so $\nabla f(A) = 2A$; the second-order
term is exactly $\norm H_F^2$: the Hessian is twice the
identity \omterm{def:b2:quadratic:def}{quadratic form}, positive definite and constant, so
$\norm\cdot_F^2$ is strictly convex (the Taylor formula is
exact here).

\textbf{10.} By multilinearity in the columns, $\det(I + H) =
\sum_{S\subseteq\{1,\dots,n\}}\det(M_S)$ where $M_S$ has column
$h_j$ for $j \in S$ and $e_j$ otherwise. $S = \varnothing$
gives $1$; $S = \{j\}$ gives the \omterm{def:b2:linalg:det}{determinant} of $I$ with column
$j$ replaced by $h_j$, namely its $j$-th entry $h_{jj}$,
summing to $\operatorname{tr}H$; each term with $\abs S \geq 2$
is a \omterm{def:b2:linalg:det}{determinant} with at least two columns of size
$O(\vertiii H)$, hence $O(\vertiii H^2)$ (multilinear maps on a
finite-dimensional space are bounded), and there are finitely
many. So $\det(I + H) = 1 + \operatorname{tr}H +
O(\vertiii H^2)$: $\dd(\det)_I = \operatorname{tr}$.

\textbf{11.} $\det(A + H) = \det A\,\det(I + A^{-1}H) =
\det A\,\bigl(1 + \operatorname{tr}(A^{-1}H) +
O(\vertiii H^2)\bigr)$: the \omterm{def:b2:diffcalc:differential}{differential} is $H \mapsto
\det(A)\operatorname{tr}(A^{-1}H)$.

\textbf{12.} Laplace expansion along row $i$: $\det A =
\sum_j a_{ij}C_{ij}$, and the cofactors $C_{ij}$ do not involve
row $i$: $\frac{\partial\det}{\partial a_{ij}} = C_{ij}$. Hence
\[
\dd(\det)_A(H) = \sum_{i,j}C_{ij}h_{ij}
= \operatorname{tr}\bigl(\operatorname{com}(A)^{\mathsf T}
H\bigr)
= \operatorname{tr}\bigl(\operatorname{adj}(A)H\bigr),
\qquad
\nabla(\det)(A) = \operatorname{com}(A) .
\]
For invertible $A$, $\operatorname{adj}A = \det(A)A^{-1}$
recovers question 11. Along a $C^1$ curve, the chain rule reads
$(\det A(t))' = \operatorname{tr}(\operatorname{adj}
(A(t))\,A'(t))$: Jacobi's formula.

\textbf{13.} With $A' = MA$ and
$\operatorname{adj}(A)A = \det(A)I$:
\[
(\det A)' = \operatorname{tr}\bigl(\operatorname{adj}(A)MA
\bigr)
= \operatorname{tr}\bigl(A\operatorname{adj}(A)M\bigr)
= \det A\;\operatorname{tr}M
\]
(cyclicity; $A\operatorname{adj}A = \det(A) I$ as well). The
scalar linear ODE $y' = \operatorname{tr}(M(t))\,y$ has the
unique solution $y(t) = y(0)\exp\bigl(\int_0^t
\operatorname{tr}M\bigr)$ (Year 1): Liouville's formula.

\textbf{14.} At $A \in SL_n(\R)$, take $H = \frac1nA$:
$\dd(\det)_A\bigl(\tfrac1nA\bigr) = \frac1n\det(A)
\operatorname{tr}(A^{-1}A) = \frac1n\cdot1\cdot n = 1 \neq 0$:
the \omterm{def:b2:diffcalc:differential}{differential} is a nonzero linear form, hence surjective
onto $\R$ at every point of the level set: $SL_n(\R)$ is a
smooth level set (of dimension $n^2 - 1$).

\textbf{15.} $\sum_k\vertiii{A^k/k!} \leq
\sum\vertiii A^k/k! = \eu^{\vertiii A}$: absolute convergence
(question 2's \omterm{def:b2:metric:complete}{completeness} argument), with \omterm{def:b2:funcseq:series}{normal convergence}
on every ball $\vertiii A \leq R$ (bound $R^k/k!$ independent
of $A$). Each partial sum is \omterm{def:b2:metric:continuity}{continuous} (polynomial); the
uniform limit on balls is \omterm{def:b2:metric:continuity}{continuous}: $A \mapsto \eu^A$ is
\omterm{def:b2:metric:continuity}{continuous}, with $\vertiii{\eu^A} \leq \eu^{\vertiii A}$.

\textbf{16.} Both series converge \omterm{def:b2:series:def}{absolutely}, so the \omterm{thm:b2:series:fubini}{Cauchy
product} is legitimate (\cref{thm:b2:series:fubini}):
\[
\eu^A\eu^B = \sum_{n\geq0}\frac{1}{n!}\sum_{k=0}^n\binom
nkA^kB^{n-k}
= \sum_{n\geq0}\frac{(A+B)^n}{n!} = \eu^{A+B},
\]
the binomial identity requiring $AB = BA$. With $B = -A$:
$\eu^A\eu^{-A} = \eu^0 = I$: every $\eu^A \in GL_n(\R)$.

\textbf{17.} The series $\sum t^kA^k/k!$ and its derived series
$\sum t^{k-1}A^k/(k-1)! = A\sum t^{k-1}A^{k-1}/(k-1)!$ converge
\omterm{def:b2:funcseq:series}{normally} on every segment $\abs t \leq T$ (bounds
$T^k\vertiii A^k/k!$): the differentiation theorem for series
(\cref{thm:b2:funcseq:seriestransfer}, applied entry by entry)
gives $\frac{\dd}{\dd t}\eu^{tA} = A\eu^{tA}$; factoring $A$
on the right instead gives $\eu^{tA}A$. Iterating: $C^\infty$.

\textbf{18.} $A(t) = \eu^{tA}$ satisfies $A'(t) = A\,A(t)$:
Liouville's formula (question 13) with constant $M = A$ gives
$\det\eu^{tA} = \eu^{t\operatorname{tr}A}$ (value $1$ at $t =
0$); at $t = 1$, $\det\eu^A = \eu^{\operatorname{tr}A}$.
Checks: $n = 1$ is the exponential itself; a nilpotent $A$ has
$\operatorname{tr}A = 0$ and $\eu^A$ unipotent of \omterm{def:b2:linalg:det}{determinant}
$1$; $\operatorname{tr}A = 0$ gives $\det\eu^A = 1$: the
trace-zero matrices are sent into $SL_n(\R)$.

\textbf{19.} $\eu^H - I - H = \sum_{k\geq2}H^k/k!$, of \omterm{def:b2:nvs:norm}{norm}
$\leq \vertiii H^2\eu^{\vertiii H} = O(\vertiii H^2)$:
$\dd(\exp)_0 = \mathrm{id}$, invertible. Moreover $\exp$ is
$C^1$: by question 5, the \omterm{def:b2:diffcalc:differential}{differential} candidate $H \mapsto
\sum_k\frac1{k!}\sum_iA^iHA^{k-1-i}$ is a \omterm{def:b2:funcseq:series}{normally} convergent
series of linear maps depending \omterm{def:b2:metric:continuity}{continuously} on $A$ (bounds
$\vertiii A^{k-1}/(k-1)!$ on balls), so the partials exist and
are \omterm{def:b2:metric:continuity}{continuous} (\cref{thm:b2:diffcalc:c1} and the series
transfer theorem). The inverse function theorem
(\cref{thm:b2:diffcalc:inverse}) applies at $0$: $\exp$ is a
$C^1$ diffeomorphism of a neighborhood of $0$ onto a
neighborhood of $I$ --- matrices near $I$ have logarithms.

\textbf{20.} For \omterm{def:b2:quadratic:adjoint}{symmetric} $S = PDP^{\mathsf T}$ (spectral
theorem): $\eu^S = P\eu^DP^{\mathsf T}$ is \omterm{def:b2:quadratic:adjoint}{symmetric} with
\omterm{def:b2:reduction:eigen}{eigenvalues} $\eu^{\lambda_i} > 0$: positive definite.
Surjectivity: a positive definite $Q =
P\operatorname{diag}(\mu_i)P^{\mathsf T}$ ($\mu_i > 0$) is
$\eu^S$ for $S = P\operatorname{diag}(\ln\mu_i)P^{\mathsf T}$.
Injectivity: $\eu^S$ determines its \omterm{def:b2:reduction:eigen}{eigenspaces}, which are
exactly those of $S$ (on each \omterm{def:b2:reduction:eigen}{eigenspace} of $S$ for $\lambda$,
$\eu^S$ acts as $\eu^\lambda$; distinct $\lambda$ give
distinct $\eu^\lambda$), and taking $\ln$ of the \omterm{def:b2:reduction:eigen}{eigenvalues}
recovers $S$. So $\exp$ is a bijection from \omterm{def:b2:quadratic:adjoint}{symmetric} matrices
onto positive definite ones.

\textbf{21.} $F(A + H) = A^{\mathsf T}A + A^{\mathsf T}H +
H^{\mathsf T}A + H^{\mathsf T}H$: $\dd F_A(H) =
A^{\mathsf T}H + H^{\mathsf T}A$ (values in $S_n$; error
$O(\vertiii H^2)$). At $A \in O_n$ and for $S \in S_n$, the
choice $H = \frac12AS$ gives
\[
A^{\mathsf T}\cdot\tfrac12AS + \tfrac12(AS)^{\mathsf T}A
= \tfrac12 S + \tfrac12 S^{\mathsf T} = S :
\]
surjective. $O_n = F^{-1}(I)$ is a smooth level set of
dimension $n^2 - \dim S_n = \frac{n(n-1)}2$.

\textbf{22.} Differentiating $A(t)^{\mathsf T}A(t) = I$ at $t
= 0$ (with $A(0) = I$): $A'(0)^{\mathsf T} + A'(0) = 0$:
antisymmetric. Conversely, for $K^{\mathsf T} = -K$:
$(\eu^{tK})^{\mathsf T}\eu^{tK} = \eu^{tK^{\mathsf T}}
\eu^{tK} = \eu^{-tK}\eu^{tK} = I$ (\omterm{def:b2:linalg:transpose}{transpose} the series
termwise; the exponents commute): the curve stays in $O_n$,
with velocity $K$ at $t = 0$. Tangent space at $I$ $=$ the
antisymmetric matrices, of the expected dimension
$\frac{n(n-1)}2$.

\textbf{23.} $\operatorname{tr}K = 0$ for antisymmetric $K$,
so $\det\eu^K = \eu^0 = 1$ (question 18): the exponential
lands in $SO_n$. For $n = 2$: $J^2 = -I$, so $J^{2m} =
(-1)^mI$, $J^{2m+1} = (-1)^mJ$, and
\[
\eu^{\theta J}
= \Bigl(\sum_m\frac{(-1)^m\theta^{2m}}{(2m)!}\Bigr)I
+ \Bigl(\sum_m\frac{(-1)^m\theta^{2m+1}}{(2m+1)!}\Bigr)J
= \cos\theta\,I + \sin\theta\,J ,
\]
the rotation by $\theta$: the sine and cosine series live
inside the \omterm{ex:b2:nvs:matrixexp}{matrix exponential}.

\textbf{24.} For invertible $A, B$: $\operatorname{adj}(AB) =
\det(AB)(AB)^{-1} = \det(B)\det(A)B^{-1}A^{-1} =
\operatorname{adj}(B)\operatorname{adj}(A)$. Both sides of the
identity are polynomial (hence \omterm{def:b2:metric:continuity}{continuous}) maps of the entries
of $(A, B)$; they agree on the dense subset $GL_n\times GL_n$
of $\mathcal M_n\times\mathcal M_n$ (question 3: approximate
each factor), so they agree everywhere.

\textbf{25.} (i) Part I ran on \omterm{def:b2:metric:complete}{completeness} of finite-dimensional
normed spaces (\omterm{def:b2:series:def}{absolutely} convergent series converge), Part IV
on the same plus the spectral theorem for question 20, Part V's
local logarithm on the inverse function theorem. (ii) The
differential-equations chapter leans on Liouville's formula
(question 13) for the Wronskian of linear systems, and on
$\frac{\dd}{\dd t}\eu^{tA} = A\eu^{tA}$ (question 17), which is
the statement that $\eu^{tA}$ solves $X' = AX$. (iii)
$\dd(\det)_I = \operatorname{tr}$ says the trace is the
infinitesimal rate of volume change, and $\det\eu^A =
\eu^{\operatorname{tr}A}$ integrates that statement globally.
(iv) The Year 3 volume names the structures: $O_n$ and
$SL_n(\R)$ are Lie groups, their tangent spaces at $I$
(antisymmetric and trace-zero matrices) are Lie algebras, and
$\exp$ is the bridge between them.
\end{solution}
