\section*{Chapter \ref{ch:b2:diffeq} --- Differential Equations}

\begin{solution}{exo:b2:diffeq:1}
First matrix: \omterm{def:b2:reduction:eigen}{eigenvalues} $1, 2$, \omterm{def:b2:reduction:eigen}{eigenvectors} $(1,0)$ and $(1,1)$.
Decompose $X_0 = (1,0) = 1\cdot(1,0) + 0\cdot(1,1)$: the solution
is
\[
X(t) = \eu^{t}\begin{pmatrix}1\\ 0\end{pmatrix} .
\]
(The initial vector is itself an \omterm{def:b2:reduction:eigen}{eigenvector}.)

Second: $A = 2I + N$, $N = E_{12}$, $N^2 = 0$: $\eu^{tA} =
\eu^{2t}(I + tN)$, so
\[
X(t) = \eu^{2t}\begin{pmatrix} 1 & t\\ 0 & 1\end{pmatrix}
\begin{pmatrix}1\\ 0\end{pmatrix}
= \eu^{2t}\begin{pmatrix}1\\ 0\end{pmatrix}.
\]
\end{solution}

\begin{solution}{exo:b2:diffeq:2}
$y' = \sin(ty)$: $\abs{\sin(ty) - \sin(tz)} \leq \abs t\,\abs{y -
z}$ --- \omterm{def:b2:metric:continuity}{Lipschitz} in $y$ \omterm{def:b2:funcseq:def}{uniformly} on every segment of times:
unique global solutions on $\R$ (apply the theorem on every
segment).

$y' = y^2$: locally \omterm{def:b2:metric:continuity}{Lipschitz} only; no global theorem, and indeed
$y(0) = 1$ explodes at $t = 1$.

$y' = \abs y$: $\abs\cdot$ is $1$-Lipschitz: global existence and
uniqueness. With $y(0) = 0$: $y \equiv 0$ (uniqueness!). With $y(0)
= 1$: $y$ stays positive (cannot cross the zero solution), so $y' =
y$: $y = \eu^t$.
\end{solution}

\begin{solution}{exo:b2:diffeq:3}
If $y(t_1) = z(t_1)$ at some time, then $y$ and $z$ solve the same
Cauchy problem at $t_1$: uniqueness forces $y = z$ on their common
interval --- distinct solutions never meet.

For $y' = y(1-y)$: the constants $0$ and $1$ are solutions. A
solution starting in $\intoo{0}{1}$ can never reach $0$ or $1$
(it would collide with a constant solution): it stays in
$\intoo{0}{1}$, hence is global (bounded: no explosion --- e.g.\
by \cref{exo:b2:diffeq:9}'s criterion, or because the vector field
is bounded on the trapped strip).
\end{solution}

\begin{solution}{exo:b2:diffeq:4}
$(A - 2I)^2 = \begin{pmatrix}1 & 1\\ -1 & -1\end{pmatrix}^2 = 0$:
Dunford with $D = 2I$, $N = A - 2I$:
\[
\eu^{tA} = \eu^{2t}\,(I + tN)
= \eu^{2t}\begin{pmatrix} 1 + t & t\\ -t & 1 - t\end{pmatrix}.
\]
Variation of constants with $B(s) = (\eu^{2s}, 0)^{\mathsf T}$:
\[
X(t) = \int_0^t \eu^{(t-s)A}B(s)\,\dd s
= \eu^{2t}\int_0^t \begin{pmatrix} 1 + (t-s)\\ -(t-s)
\end{pmatrix}\dd s
= \eu^{2t}\begin{pmatrix} t + \frac{t^2}{2}\\[2pt] -\frac{t^2}{2}
\end{pmatrix},
\]
using $\eu^{(t-s)A}B(s) = \eu^{2(t-s)}(I + (t-s)N)\,(\eu^{2s},
0)^{\mathsf T} = \eu^{2t}\bigl(1 + (t-s),\, -(t-s)\bigr)^{\mathsf
T}$. (Check: $X(0) = 0$; $X' - AX = (\eu^{2t}, 0)^{\mathsf T}$
by differentiation.)
\end{solution}

\begin{solution}{exo:b2:diffeq:5}
\emph{Series:} $A = \alpha I + \beta J$ with $J =
\begin{pmatrix}0 & -1\\ 1 & 0\end{pmatrix}$, $J^2 = -I$; the two
summands commute, so $\eu^{tA} = \eu^{\alpha t}\,\eu^{\beta tJ}$,
and the series of $\eu^{\beta t J}$ splits along even/odd powers
into $\cos(\beta t)I + \sin(\beta t)J$: the stated rotation-scaling
matrix.

\emph{Complex:} identify $(x, y) \in \R^2$ with $z = x + \iu y$;
the system $X' = AX$ reads $z' = (\alpha + \iu\beta)z$, whose
solution $z(t) = \eu^{\alpha t}\eu^{\iu\beta t}z_0$ is exactly the
spiral: modulus $\eu^{\alpha t}$, argument advancing at speed
$\beta$.
\end{solution}

\begin{solution}{exo:b2:diffeq:6}
Let $v(t) = \eu^{-k(t - t_0)}\int_{t_0}^t u$. Then
\[
v'(t) = \eu^{-k(t-t_0)}\Bigl(u(t) - k\int_{t_0}^t u\Bigr)
\leq C\,\eu^{-k(t-t_0)},
\]
by the hypothesis. Integrating from $t_0$ to $t$ ($v(t_0) = 0$):
$v(t) \leq \frac{C}{k}\bigl(1 - \eu^{-k(t - t_0)}\bigr)$, i.e.\
$k\int_{t_0}^t u \leq C\bigl(\eu^{k(t-t_0)} - 1\bigr)$; feeding
this back into the hypothesis: $u(t) \leq C\eu^{k(t-t_0)}$.

Uniqueness/dependence: two solutions $y, z$ of the integral
equation satisfy
\[
\norm{y(t) - z(t)} \leq \norm{y_0 - z_0} +
k\int_{t_0}^{t}\norm{y - z},
\]
and Gronwall with $C = \norm{y_0 - z_0}$ gives the exponential
bound; $C = 0$ gives uniqueness.
\end{solution}

\begin{solution}{exo:b2:diffeq:7}
Substitute $y = y_1 z$ with $y_1 = \frac{\sin t}{t}$: the general
lowering formula (\cref{prop:b2:diffeq:secondorder}) gives, for $u
= z'$,
\[
y_1 u' + \Bigl(2y_1' + \frac{2}{t}\,y_1\Bigr)u = 0
\]
(the equation normalized as $y'' + \frac2t y' + y = 0$). Compute
$2y_1' + \frac2t y_1 = 2\,\frac{t\cos t - \sin t}{t^2} +
\frac{2\sin t}{t^2} = \frac{2\cos t}{t}$: so
\[
\frac{u'}{u} = -\frac{2\cos t}{t}\cdot\frac{t}{\sin t}
= -2\cot t
\quad\Longrightarrow\quad
u = \frac{1}{\sin^2 t}
\quad (\text{up to a constant}),
\]
and $z = -\cot t$, giving $y_2 = y_1 z = -\frac{\cos t}{t}$.
General solution on $\intoo{0}{\pi}$:
\[
y(t) = \alpha\,\frac{\sin t}{t} + \beta\,\frac{\cos t}{t} .
\]
(These are the spherical Bessel functions of order zero.)
\end{solution}

\begin{solution}{exo:b2:diffeq:8}
Dunford: $A = D + N$ commuting, $D$ \omterm{def:b2:reduction:diag}{diagonalizable} with the same
\omterm{def:b2:reduction:eigen}{eigenvalues}, $N$ nilpotent, so
\[
\eu^{tA} = \eu^{tD}\,\eu^{tN},
\qquad
\eu^{tN} = \sum_{k < n} \frac{t^kN^k}{k!}
\ \text{(a matrix polynomial in } t).
\]
Let $-2\alpha = \max_i \Re\lambda_i < 0$. In a basis diagonalizing
$D$, $\vertiii{\eu^{tD}} \leq \eu^{-2\alpha t}$ (entries
$\eu^{t\lambda_i}$ of modulus $\eu^{t\Re\lambda_i}$); \omterm{def:b2:nvs:norm}{norms} in
different bases differ by constants. Hence
\[
\norm{X(t)} \leq \vertiii{\eu^{tA}}\,\norm{X_0}
\leq C'\,\eu^{-2\alpha t}\,(1 + t)^{n-1}\,\norm{X_0}
\leq C\,\eu^{-\alpha t}\norm{X_0} ,
\]
absorbing the polynomial into one exponential factor
($\eu^{-\alpha t}(1+t)^{n-1} \to 0$, hence bounded).
\end{solution}

\begin{solution}{exo:b2:diffeq:9}
Let $y$ be a maximal solution on $\intco{0}{T}$, $T \leq \infty$,
and suppose $T < \infty$. The integral form gives, for $t < T$,
\[
\norm{y(t)} \leq \norm{y_0} + \int_0^t \bigl(a\norm{y(s)} +
b\bigr)\dd s
\leq \bigl(\norm{y_0} + bT\bigr) + a\int_0^t\norm{y(s)}\,\dd s ,
\]
and Gronwall bounds $\norm{y(t)} \leq (\norm{y_0} +
bT)\,\eu^{aT} =: M$ on $\intco{0}{T}$: the solution stays in a
\omterm{def:b2:metric:compact}{compact} ball. Then $y' = f(t, y)$ is bounded near $T$, so $y$ is
\omterm{def:b2:metric:continuity}{Lipschitz} near $T$ and extends \omterm{def:b2:metric:continuity}{continuously} to $T$ (Cauchy
criterion); solving the Cauchy problem at $(T, y(T))$ prolongs $y$
beyond $T$, contradicting maximality. Hence $T = \infty$: no
finite-time escape under linear growth.
\end{solution}

\begin{solution}{exo:b2:diffeq:10}
Characteristic roots of $r^2 - 3r + 2$: $1$ and $2$, so the
homogeneous solutions are $a\eu^t + b\eu^{2t}$. The guess
$\alpha\eu^t$ fails because $\eu^t$ already solves the
homogeneous equation (the root $r = 1$ ``resonates'' with the
right-hand side). With $y = \alpha t\,\eu^t$: $y' = \alpha(1 +
t)\eu^t$, $y'' = \alpha(2 + t)\eu^t$, and
\[
y'' - 3y' + 2y = \alpha\eu^t\bigl(2 + t - 3 - 3t + 2t\bigr)
= -\alpha\,\eu^t :
\]
$\alpha = -1$, $y_p = -t\,\eu^t$. General solution: $y = a\eu^t
+ b\eu^{2t} - t\eu^t$. Initial data $y(0) = y'(0) = 0$: $a + b
= 0$ and $a + 2b - 1 = 0$: $b = 1$, $a = -1$:
\[
y(t) = \eu^{2t} - (1 + t)\,\eu^{t} .
\]

\end{solution}

\begin{solution}{exo:b2:diffeq:11}
$A = \lambda I + N$ with $N = E_{12} + E_{23}$: $N^2 = E_{13}$,
$N^3 = 0$, and $\lambda I$ commutes with $N$:
\[
\eu^{tA} = \eu^{\lambda t}\Bigl(I + tN +
\frac{t^2}{2}N^2\Bigr)
= \eu^{\lambda t}\begin{pmatrix}
1 & t & \frac{t^2}{2}\\
0 & 1 & t\\
0 & 0 & 1
\end{pmatrix}.
\]
Solutions: $X(t) = \eu^{\lambda t}\bigl(X_0 + tNX_0 +
\frac{t^2}2N^2X_0\bigr)$ --- each component is $\eu^{\lambda
t}$ times a polynomial of degree $\leq 2$. The degree bound is
the nilpotency index minus one: the series of $\eu^{tN}$
truncates at $N^2$.
\end{solution}

\begin{solution}{exo:b2:diffeq:12}
\begin{enumerate}
  \item If $X(T) = X(0)$, then $Y(t) = X(t + T)$ solves $Y' =
        AY + B(t + T) = AY + B(t)$ with $Y(0) = X(0)$:
        uniqueness (\cref{thm:b2:diffeq:linear}) gives $Y = X$,
        i.e.\ $X$ is $T$-periodic. The converse is trivial.
  \item Trigonalize over $\C$: $A = PT'P^{-1}$ with $T'$ upper
        triangular, diagonal $(\lambda_i)$. Every power of a
        triangular matrix is triangular with diagonal
        $(\lambda_i^k)$, so $\eu^{TA} = P\eu^{TT'}P^{-1}$ is
        triangular in the same basis with diagonal
        $(\eu^{T\lambda_i})$: those are the \omterm{def:b2:reduction:eigen}{eigenvalues}. Then
        $I - \eu^{TA}$ is invertible iff $\eu^{T\lambda} \neq
        1$ for every \omterm{def:b2:reduction:eigen}{eigenvalue}, i.e.\ iff $T\lambda \notin
        2\iu\pi\Z$, which is the stated hypothesis.
  \item Variation of constants: $X(T) = \eu^{TA}X(0) +
        \int_0^T\eu^{(T-s)A}B(s)\dd s$, so $X(T) = X(0)$ reads
        \[
        \bigl(I - \eu^{TA}\bigr)X(0) =
        \int_0^{T}\eu^{(T-s)A}B(s)\,\dd s ,
        \]
        which has a unique solution $X(0)$ under the
        invertibility hypothesis: exactly one $T$-periodic
        solution. For the harmonic oscillator ($\lambda =
        \pm\iu\omega$), the excluded case is $\omega T \in
        2\pi\Z$: forcing whose period is a multiple of the
        natural period --- \omterm{pb:b2:diffeq:1}{resonance}, as the weekend problem
        quantifies.
\end{enumerate}
\end{solution}

\begin{solution}{pb:b2:diffeq:1}
\textbf{1.} The \omterm{def:b2:reduction:charpoly}{characteristic polynomial} is $\lambda^2 -
\tau\lambda + \delta$, with roots
$\frac{\tau\pm\sqrt\Delta}{2}$. If $\delta < 0$ then $\Delta =
\tau^2 - 4\delta > 0$ and the two real roots have product
$\delta < 0$: opposite signs. If $\delta > 0$ and $\Delta \geq
0$: real roots of product $> 0$ and sum $\tau$: both of the
sign of $\tau$. If $\Delta < 0$: conjugate pair $\alpha \pm
\iu\beta$ with $\alpha = \frac\tau2$, $\beta =
\frac{\sqrt{-\Delta}}2$.

\textbf{2.} $X(t) = a\,\eu^{\mu t}v_- + b\,\eu^{\lambda t}v_+$.
Orbits with $a = 0$ (resp.\ $b = 0$) run along the unstable
(resp.\ stable) eigenline; all others have $\norm X \to \infty$
in both time directions, asymptotic to $\R v_+$ as $t \to
+\infty$ and to $\R v_-$ as $t \to -\infty$: the saddle
picture. Boundedness on all of $\R$ forces $b = 0$ (else
explosion at $+\infty$) and $a = 0$ (at $-\infty$): only the
origin.

\textbf{3.} With $\mu < \lambda < 0$, both exponentials decay:
$X(t) \to 0$. If $b \neq 0$, factor $\eu^{\lambda t}$:
\[
X(t) = \eu^{\lambda t}\bigl(b\,v_\lambda + a\,\eu^{(\mu -
\lambda)t}v_\mu\bigr),
\qquad \eu^{(\mu-\lambda)t} \to 0 :
\]
the direction of $X(t)$ tends to $\R v_\lambda$, the
\emph{slow} eigendirection --- all orbits but the fast axis
arrive tangent to it (the right panel of the chapter's phase
portraits).

\textbf{4.} In the basis where $A = \begin{pmatrix} \alpha &
-\beta\\ \beta & \alpha\end{pmatrix}$
(\cref{exo:b2:diffeq:5}; for the oscillator systems of Part II
this form is reached by an explicit real change of basis), the
solution is $\eu^{\alpha t}$ times a rotation of angle $\beta
t$: logarithmic spirals, contracting when $\alpha = \frac\tau2
< 0$, expanding when $\tau > 0$, and closed curves (ellipses
in the original coordinates) when $\tau = 0$: the center.

\textbf{5.} $\Delta = 0$ gives the double \omterm{def:b2:reduction:eigen}{eigenvalue} $\lambda =
\frac\tau2$; by Cayley--Hamilton
(\cref{thm:b2:reduction:cayleyhamilton}), $(A - \lambda I)^2 =
0$, so $N = A - \lambda I$ is nilpotent, commutes with
$\lambda I$, and $\eu^{tA} = \eu^{\lambda t}(I + tN)$. If $N =
0$: $A = \lambda I$, all rays are orbits (star node). If $N
\neq 0$: $X(t) = \eu^{\lambda t}(X_0 + tNX_0)$, and for $NX_0
\neq 0$ the direction converges to the single eigendirection
$\operatorname{im}N$: improper node. This completes the
trace--determinant picture.

\textbf{6.} $\tau = -2\zeta\omega$, $\delta = \omega^2 > 0$,
$\Delta = 4\omega^2(\zeta^2 - 1)$. So: $0 < \zeta < 1$ gives
$\Delta < 0$, $\tau < 0$: stable spiral; $\zeta = 1$: $\Delta =
0$: degenerate stable node; $\zeta > 1$: $\Delta > 0$, $\tau <
0$, $\delta > 0$: stable node; $\zeta = 0$: $\tau = 0$,
$\delta > 0$: center. A vertical journey in the plane at
$\delta = \omega^2$.

\textbf{7.} Roots $r = -\zeta\omega \pm
\omega\sqrt{\zeta^2-1}$. For $\zeta < 1$: $r = -\zeta\omega
\pm \iu\omega_d$, $\omega_d = \omega\sqrt{1-\zeta^2}$:
\[
x(t) = \eu^{-\zeta\omega t}\bigl(a\cos\omega_dt +
b\sin\omega_dt\bigr)
= R\,\eu^{-\zeta\omega t}\cos(\omega_dt - \varphi) .
\]
For $\zeta = 1$: $x = (a + bt)\eu^{-\omega t}$. For $\zeta >
1$: $x = a\eu^{r_-t} + b\eu^{r_+t}$, both rates negative.
Successive maxima of $\abs x$ in the underdamped case occur at
times separated by the pseudo-period $\frac{2\pi}{\omega_d}$
(same phase of the cosine), and their ratio is
$\eu^{-\zeta\omega\cdot2\pi/\omega_d} =
\eu^{-2\pi\zeta/\sqrt{1-\zeta^2}}$: the logarithmic decrement,
a damping meter readable on an oscilloscope.

\textbf{8.} Rationalizing,
\[
\abs{\lambda_{\mathrm{slow}}} = \omega\bigl(\zeta -
\sqrt{\zeta^2-1}\bigr)
= \frac{\omega}{\zeta + \sqrt{\zeta^2 - 1}} ,
\]
whose denominator increases with $\zeta \geq 1$: the decay
rate is largest at $\zeta = 1$, where it equals $\omega$. An
overdamped door closes without slamming but \emph{slowly};
critical damping is the engineer's optimum.

\textbf{9.} $E' = x'x'' + \omega^2xx' =
x'\bigl(-2\zeta\omega x' - \omega^2x\bigr) + \omega^2xx' =
-2\zeta\omega\,x'^2 \leq 0$. If $x$ were periodic and
nonconstant, $E$ would be periodic and nonincreasing, hence
constant, forcing $x' \equiv 0$: $x$ constant, and then
$\omega^2x = 0$: $x \equiv 0$. So for $\zeta > 0$ the only
periodic solution is rest: damping kills every \omterm{def:b2:structures:sn}{cycle}.

\textbf{10.} The center $\zeta = 0$ lives on the line $\tau =
0$ of the trace--determinant plane --- a set with empty
interior: an arbitrarily small perturbation of the matrix
(any physical damping) moves $\tau$ off zero and turns the
closed orbits into spirals. Periodicity of the undamped
oscillator is thus a razor's-edge phenomenon, not a robust
one.

\textbf{11.} Substituting $x_p = \Re(z\eu^{\iu\gamma t})$ into
the equation:
\[
\bigl(-\gamma^2 + 2\iu\zeta\omega\gamma +
\omega^2\bigr)z = F
\quad\Longrightarrow\quad
z = \frac{F}{\omega^2 - \gamma^2 + 2\iu\zeta\omega\gamma},
\]
so $x_p = \abs z\cos(\gamma t - \varphi)$ with $\varphi =
\arg(\omega^2 - \gamma^2 + 2\iu\zeta\omega\gamma)$, i.e.\
$\tan\varphi = \frac{2\zeta\omega\gamma}{\omega^2 -
\gamma^2}$, and the stated $A(\gamma) = \abs z$.

\textbf{12.} The difference of two solutions solves the
homogeneous equation, which for $\zeta > 0$ decays to $0$
(question 7): every solution equals $x_p$ plus a transient
vanishing at infinity. The steady state is a global attractor:
initial conditions are forgotten, only $A(\gamma)$ and the
phase lag $\varphi$ remain.

\textbf{13.} Minimize $g(u) = (\omega^2 - u)^2 +
4\zeta^2\omega^2u$ over $u = \gamma^2 \geq 0$: $g'(u) =
-2(\omega^2 - u) + 4\zeta^2\omega^2 = 0$ at $u = \omega^2(1 -
2\zeta^2)$, interior iff $\zeta < \frac{1}{\sqrt2}$. There
\[
g(u_*) = 4\zeta^4\omega^4 + 4\zeta^2\omega^4(1 - 2\zeta^2)
= 4\zeta^2\omega^4(1 - \zeta^2),
\qquad
A(\gamma_*) = \frac{F}{2\zeta\omega^2\sqrt{1 - \zeta^2}} .
\]
Against the static response $A(0) = \frac{F}{\omega^2}$:
amplification $\frac{1}{2\zeta\sqrt{1-\zeta^2}} \approx
\frac{1}{2\zeta}$ for small $\zeta$ --- a lightly damped
system near $\gamma_* \approx \omega$ multiplies the input a
hundredfold when $\zeta = 0.005$.

\textbf{14.} The stated $x$ satisfies $x(0) = x'(0) = 0$ and
\[
x'' + \omega^2 x = \frac{F(\omega^2 -
\gamma^2)\cos\gamma t}{\omega^2 - \gamma^2} = F\cos\gamma t
\]
(the $\cos\omega t$ parts cancel). The product form follows
from $\cos p - \cos q = 2\sin\frac{q+p}{2}\sin\frac{q-p}2$
with $p = \gamma t$, $q = \omega t$. For $\gamma$ close to
$\omega$, the factor $\sin\frac{(\omega-\gamma)t}2$ is a slow
envelope modulating the fast oscillation
$\sin\frac{(\omega+\gamma)t}2$: beats, with amplitude
$\frac{2F}{\abs{\omega^2-\gamma^2}}$ --- large, but bounded.

\textbf{15.} For $x_p = \frac{F}{2\omega}t\sin\omega t$:
\[
x_p'' = \frac{F}{2\omega}\bigl(2\omega\cos\omega t -
\omega^2t\sin\omega t\bigr)
= F\cos\omega t - \omega^2x_p :
\]
a solution. And at fixed $t$, letting $\gamma \to \omega$ in
question 14:
\[
\frac{2F\sin\frac{(\omega-\gamma)t}2
\sin\frac{(\omega+\gamma)t}{2}}
{(\omega-\gamma)(\omega+\gamma)}
\longrightarrow
\frac{2F\cdot\frac{(\omega-\gamma)t}2\big/(\omega-\gamma)
\cdot\sin\omega t}{2\omega}
= \frac{F\,t\sin\omega t}{2\omega} .
\]
The amplitude grows linearly without bound: the \omterm{pb:b2:diffeq:1}{resonance}
catastrophe --- the reason soldiers break step on bridges.

\textbf{16.} The square wave carries harmonics at every odd
frequency $n = 1, 3, 5, \dots$; by linearity, each harmonic
$n$ is amplified by the oscillator's response at $\gamma = n$.
For $\omega = 3$ the third harmonic hits \omterm{pb:b2:diffeq:1}{resonance} exactly.
Engineers fear square (and sawtooth) inputs because they
excite \emph{all} odd harmonics at once: whatever the natural
frequency of the structure, some harmonic is waiting for it.

\textbf{17.} $W' = y_1y_2'' - y_1''y_2 = -qy_1y_2 + qy_1y_2 =
0$: $W$ is constant (Liouville with a trace-zero companion
matrix). $W = 0$ at one point makes the initial data of $y_2$
proportional to those of $y_1$, hence $y_2$ proportional to
$y_1$ (uniqueness); $W \neq 0$ iff independent. If $y(t_0) =
y'(t_0) = 0$ then $y \equiv 0$ (uniqueness): a nonzero
solution has simple zeros, and a simple zero is isolated
($y'$ of fixed sign nearby).

\textbf{18.} Between consecutive zeros $a < b$, $y_1$ keeps
one sign, say $y_1 > 0$ on $\intoo ab$: then $y_1'(a) > 0$ and
$y_1'(b) < 0$ (simple zeros). Evaluating the constant $W =
y_1y_2' - y_1'y_2$ at $a$ and $b$:
\[
W = -y_1'(a)\,y_2(a) = -y_1'(b)\,y_2(b) ,
\]
so $y_2(a)$ and $y_2(b)$ have opposite signs ($W \neq 0$
forbids either to vanish): $y_2$ vanishes in $\intoo ab$
(intermediate values). It cannot vanish twice there: two zeros
of $y_2$ would flank a zero of $y_1$ by the same argument with
the roles exchanged, contradicting consecutiveness: exactly
one zero --- interlacing.

\textbf{19.} Suppose $z$ has no zero in $\intoo ab$; replacing
$y, z$ by their negatives, assume $y > 0$ and $z > 0$ on
$\intoo ab$. Set $\varphi = yz' - y'z$: $\varphi' = yz'' -
y''z = (q_1 - q_2)\,yz \leq 0$ on $\intoo ab$: $\varphi$ is
nonincreasing. But $\varphi(a) = -y'(a)z(a) \leq 0$ (as
$y'(a) > 0$, $z(a) \geq 0$) and $\varphi(b) = -y'(b)z(b) \geq
0$ (as $y'(b) < 0$, $z(b) \geq 0$): a nonincreasing function
running from $\leq 0$ to $\geq 0$ vanishes identically, so
$(q_1 - q_2)yz \equiv 0$ on $\intoo ab$. If $q_1 < q_2$
somewhere in $\intoo ab$, this is absurd ($y, z > 0$ there):
$z$ must vanish strictly inside. In general ($q_1 \leq q_2$),
either $z$ vanishes in $\intoo ab$, or $\varphi \equiv 0$
forces $z$ proportional to $y$, which vanishes at $a$ and $b$:
in all cases $z$ has a zero in $\intcc ab$.

\textbf{20.} Upper bound: compare $y$ (coefficient $q \geq
m^2$) with $u(t) = \sin(m(t - a))$ (coefficient $m^2 \leq q$,
so $y$ plays the role of $z$ in question 19): if $y$ had no
zero in $\intoc{a}{a + \pi/m}$, the zeros $a$ and $a +
\frac\pi m$ of $u$ would be consecutive with $y \neq 0$
between them, contradicting question 19: consecutive zeros of
$y$ are at distance $\leq \frac\pi m$. Lower bound: if two
consecutive zeros $a < b$ of $y$ had $b - a < \frac\pi M$,
then $z(t) = \sin(M(t-a))$ (coefficient $M^2 \geq q$) would
have to vanish in $\intcc ab \subset \intoo{a}{a + \pi/M}
\cup\{a\}$, where its only zero is $a$ itself --- but
question 19 applied on $\intoo{a}{b}$ with strictness at the
endpoints gives a zero in $\intcc ab$, and $z > 0$ on
$\intoc ab$: contradiction. Hence $\frac\pi M \leq b - a \leq
\frac\pi m$; for $q \equiv \omega^2$ both bounds collapse to
the exact spacing $\frac\pi\omega$ of the harmonic
oscillator.

\textbf{21.} With $u = ty$: $u'' = ty'' + 2y'$, so $ty'' + 2y'
+ ty = u'' + u = 0$: $u = A\sin t + B\cos t = R\sin(t +
\varphi)$, and $y = \frac{u}{t}$ recovers $\frac{\sin t}t$ and
$\frac{\cos t}t$ (\cref{exo:b2:diffeq:7}) with no lowering of
order. The zeros of any nonzero solution are those of
$R\sin(t + \varphi)$: spaced exactly $\pi$ --- Sturm's
philosophy in action: the coefficient $q \equiv 1$ dictates
the zeros, formula or no formula.

\textbf{22.} Set $x(t) =
\frac1\omega\int_0^t\sin(\omega(t-s))F(s)\dd s$. Then $x(0) =
0$;
\[
x'(t) = \frac1\omega\sin(0)F(t) +
\int_0^t\cos(\omega(t-s))F(s)\dd s
= \int_0^t\cos(\omega(t-s))F(s)\dd s ,
\]
so $x'(0) = 0$; and $x''(t) = F(t) -
\omega\int_0^t\sin(\omega(t-s))F(s)\dd s = F(t) - \omega^2x(t)$
(differentiation of a \omterm{thm:b2:integration:continuity}{parameter integral} with variable limit,
as in the integration chapter). With $F(s) = F\cos\omega s$,
product-to-sum gives
\[
\int_0^t\sin(\omega(t-s))\cos(\omega s)\dd s
= \frac12\int_0^t\bigl(\sin\omega t + \sin(\omega t -
2\omega s)\bigr)\dd s
= \frac{t}{2}\sin\omega t ,
\]
(the second piece integrates to zero), so $x =
\frac{F}{2\omega}t\sin\omega t$: question 15 again, from
Duhamel.

\textbf{23.} In system form $X' = AX + (0, F(t))^{\mathsf T}$
with $\operatorname{Sp}A$ of negative real parts ($\zeta >
0$): variation of constants and \cref{exo:b2:diffeq:8}
($\vertiii{\eu^{tA}} \leq C\eu^{-\alpha t}$) give
\[
\norm{X(t)} \leq C\eu^{-\alpha t}\norm{X_0}
+ \int_0^t C\eu^{-\alpha(t-s)}\norm{F}_\infty\dd s
\leq C\norm{X_0} + \frac{C\norm F_\infty}{\alpha} :
\]
bounded input, bounded output --- uniformly in the initial
data after the transient.

\textbf{24.} For $\zeta = 0$ and $\gamma \neq \omega$,
question 14's solution is bounded, and adding any homogeneous
solution (bounded: the center's orbits are circles) keeps it
bounded; at $\gamma = \omega$, question 15 grows linearly. So
for the undamped oscillator, boundedness under periodic
forcing fails at exactly one frequency --- \omterm{pb:b2:diffeq:1}{resonance} ---
while question 23 shows any positive damping restores
boundedness for \emph{all} bounded inputs.

\textbf{25.} (i) The trace--determinant plane classifies all
autonomous planar linear dynamics, and Part II's oscillator
walks a vertical line of it; forcing leaves the plane
(non-autonomous), and Duhamel takes over. (ii) $\gamma_*$ is
the frequency the system prefers, $A(\gamma_*)$ the price of
exciting it, and $\frac{1}{2\zeta}$ the amplification factor
--- the \omterm{pb:b2:diffeq:1}{resonance} sharpness engineers call the quality
factor. (iii) Sturm's theorems read oscillation off the sign
and size of $q$ alone: they govern equations (Bessel,
Schr\"odinger) whose solutions have no elementary formulas.
(iv) Constant \omterm{def:b2:diffeq:wronskian}{wronskians} (question 17, via Liouville) and
bounded-input stability (question 23) are reused silently
whenever the book meets variable-coefficient equations or
perturbed systems.
\end{solution}
