\section*{Chapter \ref{ch:b2:affine} --- Affine Spaces}

\begin{solution}{exo:b2:affine:1}
$\{x + y + z = 1\}$: affine plane, direction the vector plane $\{x
+ y + z = 0\}$, dimension $2$. Adding $x - z = 3$: an affine line
(two independent equations), direction $\{x + y + z = 0,\ x = z\}
= \operatorname{Vect}\bigl((1, -2, 1)\bigr)$, dimension $1$. $\{x^2
+ y^2 = 1\}$: a cylinder --- not stable under barycenters (the
midpoint of $(1,0,0)$ and $(-1,0,0)$ is the origin, off the
cylinder): not affine. A compatible system $MX = B$: affine
subspace $X_0 + \ker M$ of dimension $\dim\ker M$, as recalled in
\cref{def:b2:affine:subspace}.
\end{solution}

\begin{solution}{exo:b2:affine:2}
$C = \operatorname{bar}(A, 1-t; B, t)$ means $\vect{AC} =
t\,\vect{AB}$: existence of such $t$ is exactly linkage of
$\vect{AC}$ with $\vect{AB} \neq 0$, i.e.\ alignment. Positions: $t
= \frac12$: midpoint; $t = 2$: beyond $B$, at $B$'s distance from
it ($\vect{AC} = 2\vect{AB}$); $t = -1$: the reflection of $B$
through $A$.
\end{solution}

\begin{solution}{exo:b2:affine:3}
$M$ is the projection matrix onto $\operatorname{Vect}(1,1)$ along
$(1,-1)$ (check $M^2 = M$). Image of $f$: $\{MX + C\} = C +
\operatorname{im} M$: the affine line through $(1,0)$ directed by
$(1,1)$. Fixed points: $X = MX + C$, i.e.\ $(I - M)X = C$; but $C =
(1, 0)^{\mathsf T}$ and $\operatorname{im}(I - M) =
\operatorname{Vect}(1,-1)$; is $(1,0)$ in it? $(1, 0) =
\alpha(1,-1)$ forces $\alpha = 1$ and $0 = -1$: no. No fixed
points. And
\[
f(f(X)) = M(MX + C) + C = MX + MC + C = f(X) + MC,
\qquad MC = \tfrac12(1,1)^{\mathsf T} :
\]
$f\circ f$ is $f$ followed by a translation along the image line
--- $f$ is a ``glide projection'': projection onto the line
composed with a slide.
\end{solution}

\begin{solution}{exo:b2:affine:4}
$I = \operatorname{bar}(B, 2; C, 1)$ (since $\vect{BI} =
\frac13\vect{BC}$ places $I$ closer to $B$: weights $2$ on $B$,
$1$ on $C$ --- check: $\vect{BI} = \frac{1}{3}\vect{BC}$).
Similarly $J = \operatorname{bar}(C, 2; A, 1)$, $K =
\operatorname{bar}(A, 2; B, 1)$. Summing the three weighted
systems, the barycenter of $(I, 1; J, 1; K, 1)$ (each of total
weight $3$, so replace $I$ by its system, etc.) is
\[
\operatorname{bar}\bigl(A, 1 + 2;\ B, 2 + 1;\ C, 1 + 2\bigr)
= \operatorname{bar}(A, 1; B, 1; C, 1) = G ,
\]
the centroid of $ABC$: the triangle $IJK$ has the same centroid ---
predictable, because the construction treats $A, B, C$ cyclically
and the centroid is the unique fixed point of the cyclic symmetry
of weights.
\end{solution}

\begin{solution}{exo:b2:affine:5}
Set $g(u) = f(O + u) - f(O)$ (working in $\R^n$ vectorialized at
$O$), $g(0) = 0$.

\emph{Additivity:} $\frac{(O + u) + (O + v)}{2} = O + \frac{u +
v}{2}$, so midpoint preservation gives $g\bigl(\frac{u+v}{2}\bigr)
= \frac{g(u) + g(v)}{2}$; with $v = 0$: $g(u/2) = g(u)/2$;
combining, $g(u + v) = 2g\bigl(\frac{u+v}{2}\bigr) = g(u) + g(v)$.

\emph{$\Q$-homogeneity:} additivity gives $g(nu) = ng(u)$ ($n \in
\N$, induction), then $g(-u) = -g(u)$ (add), then $g(\frac pq u) =
\frac pq g(u)$ (apply $q$, use injectivity of scaling).

\emph{$\R$-homogeneity:} for $t \in \R$, take rationals $t_n \to
t$: $g(t_nu) = t_ng(u)$, and continuity of $g$ (inherited from $f$)
passes to the limit: $g(tu) = tg(u)$. Hence $g$ is linear and $f =
f(O) + g$: affine.
\end{solution}

\begin{solution}{exo:b2:affine:6}
Linear part $\varphi(x,y) = (y, x)$: the reflection in the diagonal
$y = x$ (orthogonal, determinant $-1$). Fixed points: $(x, y) = (y
+ 1, x - 1)$ amounts to the single equation $y = x - 1$ (the two
components are equivalent): every point of the line $y = x - 1$ is
fixed. So $f$ fixes that line pointwise: $f$
is the \emph{reflection} in that axis (an isometry with a line of
fixed points and linear part a reflection). Consistently, $f \circ
f(x,y) = f(y+1, x-1) = (x - 1 + 1, y + 1 - 1) = (x, y)$: an
involution, as a reflection must be.
\end{solution}

\begin{solution}{exo:b2:affine:7}
The $n + 1$ vectors $\vect{A_1A_i}$ ($i \geq 2$) are linked in
dimension $n$: there are $\mu_i$, not all zero, with $\sum_{i\geq2}
\mu_i\vect{A_1A_i} = 0$; set $\mu_1 = -\sum_{i \geq 2}\mu_i$, so
$\sum_{i}\mu_i = 0$ and $\sum_i \mu_i\,\vect{OA_i} = 0$ for every
$O$, with not all $\mu_i$ zero. Split indices: $P = \{i : \mu_i >
0\}$, $N = \{i : \mu_i < 0\}$, both nonempty (the $\mu_i$ sum to
zero and are not all zero). With $s = \sum_{i\in P}\mu_i =
-\sum_{i \in N}\mu_i > 0$:
\[
\operatorname{bar}\bigl(A_i, \tfrac{\mu_i}{s}\bigr)_{i \in P}
= \operatorname{bar}\bigl(A_i, \tfrac{-\mu_i}{s}\bigr)_{i \in N},
\]
(both sides equal the point $X$ with $\vect{OX} = \frac1s\sum_{i\in
P}\mu_i\vect{OA_i}$, by the relation): a common point of the two
convex hulls, with disjoint index groups.
\end{solution}

\begin{solution}{exo:b2:affine:8}
\emph{Image = fixed points:} for $Y = f(X)$, $f(Y) = f(f(X)) =
f(X) = Y$: every image point is fixed; conversely fixed points are
images. So $\mathcal{F} = \operatorname{im} f =
\operatorname{Fix}(f)$ is nonempty, and it is an affine subspace
(image of an affine map), with direction $\operatorname{im}\vec f$.

\emph{Projection structure:} $\vec f$ is idempotent ($\vec{f\circ
f} = \vec f^{\,2} = \vec f$), so $E = \operatorname{im}\vec f
\oplus \ker \vec f$ (\cref{ex:b2:reduction:projections}). For any
point $X$, consider the vector $\vect{f(X)\,X}$; applying $\vec f$:
\[
\vec f\bigl(\vect{f(X)\,X}\bigr) = \vect{f(f(X))\,f(X)} = 0
\qquad (f \circ f = f),
\]
so $\vect{f(X)\,X} \in \ker\vec f$. Hence $X = f(X) +
\vect{f(X)X}$ displays $X$ as a point of $\mathcal{F}$ translated
by a vector of $\ker\vec f$: $f$ is exactly the projection onto
$\mathcal{F}$ along $\ker\vec f$. Conversely such projections
clearly satisfy $f \circ f = f$.
\end{solution}

\begin{solution}{exo:b2:affine:9}
Let $M = \operatorname{bar}(B, 2;\ C, 3)$, of total weight $5$.
Associativity gives $G = \operatorname{bar}(A, 1;\ M, 5)$, so
$\vect{AG} = \frac56\,\vect{AM}$: $G$ lies on the segment
$\intcc AM$ at five-sixths of it from $A$. Since $A \notin
(BC)$, the line $(AG) = (AM)$ meets $(BC)$ at the single point
$M$, with $\vect{BM} = \frac35\,\vect{BC}$. Likewise, with $N =
\operatorname{bar}(C, 3;\ A, 1)$ (total weight $4$, $\vect{CN}
= \frac14\,\vect{CA}$), associativity gives $G =
\operatorname{bar}(B, 2;\ N, 4)$: the line $(BG)$ meets $(CA)$
at $N$, and $\vect{BG} = \frac46\,\vect{BN} =
\frac23\,\vect{BN}$.
\end{solution}

\begin{solution}{exo:b2:affine:10}
Vectorialize at an origin $O$ and write points as vectors:
$h_{\Omega, \lambda}(x) = \omega + \lambda(x - \omega)$ with
$\omega = \vect{O\Omega}$. The composition $g = h_{\Omega',
\mu} \circ h_{\Omega, \lambda}$ is affine with linear part
$\mu\lambda\,\mathrm{id}$. If $\lambda\mu \neq 1$: $1 \notin
\operatorname{Sp}(\lambda\mu\,\mathrm{id})$, so
\cref{prop:b2:affine:fixedpoint} yields a unique fixed point
$\Omega''$ and, vectorialized there, $g =
\lambda\mu\,\mathrm{id}$: the homothety $h_{\Omega'',
\lambda\mu}$. If $\lambda\mu = 1$ the linear part is the
identity, so $g$ is a translation; expanding,
\[
g(x) = \omega' + \mu\bigl(\omega + \lambda(x - \omega) -
\omega'\bigr) = x + (1 - \mu)\,\omega' + \mu(1 -
\lambda)\,\omega .
\]
For $\lambda = \mu = -1$ (point reflections) the vector is
$2\omega' - 2\omega = 2\,\vect{\Omega\Omega'}$: the
composition of the point reflections in $\Omega$ then
$\Omega'$ is the translation by $2\,\vect{\Omega\Omega'}$.
\end{solution}

\begin{solution}{exo:b2:affine:11}
$\vect{A'B} = \alpha\,\vect{A'C}$ says exactly $1\cdot
\vect{A'B} - \alpha\,\vect{A'C} = 0$, i.e.\ $A' =
\operatorname{bar}(B, 1;\ C, -\alpha)$ (total weight $1 -
\alpha \neq 0$ since $B \neq C$); likewise $B' =
\operatorname{bar}(C, 1;\ A, -\beta)$ and $C' =
\operatorname{bar}(A, 1;\ B, -\gamma)$.

\emph{The alignment criterion.} Give each point $P$ its
normalized barycentric row $p = (p_A, p_B, p_C)$, $p_A + p_B +
p_C = 1$, with respect to $(A, B, C)$. If $\sum_i c_i p_i = 0$
with $(c_1, c_2, c_3) \neq 0$ for three points $P_1, P_2,
P_3$, then summing the entries gives $\sum c_i = 0$, and
$\sum_i c_i \vect{OP_i} = \sum_j \bigl(\sum_i
c_ip_{ij}\bigr)\vect{OV_j} = 0$: the $P_i$ are affinely
dependent, i.e.\ aligned. Conversely an affine dependence
$(t_i)$ gives $w = \sum t_ip_i$ with entries summing to $0$
and $\sum_j w_j\vect{OV_j} = 0$; expanding from $A$, $w_B
\vect{AB} + w_C\vect{AC} = 0$, so $w = 0$ by affine
independence of $(A, B, C)$: the rows are linearly dependent.
So alignment amounts to a vanishing $3 \times 3$ determinant,
and scaling rows by the nonzero factors $1 - \alpha$, $1 -
\beta$, $1 - \gamma$ changes nothing:
\[
\det\begin{pmatrix} 0 & 1 & -\alpha\\ -\beta & 0 & 1\\ 1 &
-\gamma & 0\end{pmatrix}
= 1 - \alpha\beta\gamma .
\]
Hence $A', B', C'$ are aligned iff $\alpha\beta\gamma = 1$:
Menelaus' theorem.
\end{solution}

\begin{solution}{exo:b2:affine:12}
Let $\Delta = \{\lambda \in \R^{n+1} : \lambda_i \geq 0,\
\sum\lambda_i = 1\}$: closed and bounded in $\R^{n+1}$, hence
compact, and $K^{n+1}$ is compact as a finite product. The map
\[
\Phi \colon \Delta \times K^{n+1} \to \R^n, \qquad
\Phi(\lambda, x_0, \dots, x_n) = \sum_{i=0}^n \lambda_i x_i
\]
is continuous, and \cref{thm:b2:affine:caratheodory} says
precisely that $\operatorname{conv}(K) = \Phi(\Delta \times
K^{n+1})$: a continuous image of a compact set
(\cref{thm:b2:metric:compactprops}), hence compact.

For a closed set: take $S = (\R \times \{0\}) \cup \{(0,
1)\}$, closed in $\R^2$. A convex combination putting weight
$t$ on $(0,1)$ and $1 - t$ on axis points has second
coordinate $t$, so
\[
\operatorname{conv}(S) = \bigl(\R \times \intco01\bigr) \cup
\{(0,1)\}
\]
(for $0 \leq t < 1$, $(x, t) = t\,(0,1) + (1-t)\,\bigl(\tfrac
x{1-t}, 0\bigr)$). The point $(1, 1) = \lim_{t \to 1}(1, t)$
is adherent but not in the hull: not closed.
\end{solution}

\begin{solution}{pb:b2:affine:1}
\textbf{1.} Rebase at $A_j$: for $i \neq j$, $\vect{A_jA_i} =
\vect{A_0A_i} - \vect{A_0A_j}$. If $\sum_{i \neq j}
c_i\vect{A_jA_i} = 0$, expanding gives $\sum_{i \notin \{0,
j\}} c_i\,\vect{A_0A_i} - \bigl(\sum_{i\neq j}
c_i\bigr)\vect{A_0A_j} = 0$; independence of the
$\vect{A_0A_i}$ forces $c_i = 0$ for $i \notin \{0, j\}$, then
$c_0 = 0$: independence at $A_j$. For the equivalence: two
weight families $(\lambda_i)$, $(\lambda_i')$ summing to $1$
with the same barycenter give, with $\nu = \lambda -
\lambda'$: $\sum\nu_i = 0$ and (origin $A_0$) $\sum_{i \geq
1}\nu_i\,\vect{A_0A_i} = 0$, so $\nu = 0$ under independence.
Conversely a nontrivial relation $\sum_{i\geq1}\mu_i
\vect{A_0A_i} = 0$, completed by $\mu_0 = -\sum_{i\geq1}
\mu_i$, lets one add $t(\mu_i)$ to any weight family without
moving the barycenter: non-uniqueness.

\textbf{2.} $(\vect{AB}, \vect{AC})$ is a basis of $\R^2$:
write $\vect{AM} = \beta\,\vect{AB} + \gamma\,\vect{AC}$
(unique) and set $\alpha = 1 - \beta - \gamma$; the barycenter
condition at origin $A$ reads exactly $\vect{AM} =
\beta\,\vect{AB} + \gamma\,\vect{AC}$. Uniqueness is question
1.

\textbf{3.} From $\alpha\vect{MA} + \beta\vect{MB} +
\gamma\vect{MC} = 0$ and Chasles, $\vect{MA} =
-\beta\,\vect{AB} - \gamma\,\vect{AC}$. With $D =
\det(\vect{AB}, \vect{AC})$:
\begin{align*}
\det(\vect{MB}, \vect{MC})
&= \det(\vect{MA} + \vect{AB},\ \vect{MA} + \vect{AC})\\
&= \det(\vect{MA}, \vect{AC}) + \det(\vect{AB}, \vect{MA}) +
D = -\beta D - \gamma D + D = \alpha D,
\end{align*}
using bilinearity and $\det(\vect{MA}, \vect{AC}) = -\beta D$,
$\det(\vect{AB}, \vect{MA}) = -\gamma D$. The other two
formulas follow by the same computation with the roles
permuted cyclically.

\textbf{4.} By definition $\operatorname{conv}\{A, B, C\}$ is
the set of barycenters with nonnegative weights; normalizing
the weights to sum $1$ and invoking uniqueness (question 2),
$M \in \operatorname{conv}\{A,B,C\}$ iff $\alpha, \beta,
\gamma \geq 0$. Each coordinate is an affine function of $M$
(question 3: a $2\times2$ determinant with one column affine
in $M$), so each open sign condition defines an open
half-plane. The pattern $(-,-,-)$ contradicts $\alpha + \beta
+ \gamma = 1$; each of the remaining seven patterns is
realized: scale a sign-respecting triple with at least one
$+$ entry so that the (positive) sum is $1$ --- e.g.\
$(-1, 1, 1)$, $(3, -1, -1)$, $(\frac13, \frac13, \frac13)$,
and permutations.

\textbf{5.} An affine map preserves barycenters
(\cref{def:b2:affine:subspace}), so $u(M) = \alpha u(A) +
\beta u(B) + \gamma u(C)$. Writing $u(x, y) = ax + by + c$
with $(a, b) \neq (0,0)$: $\{u = c'\}$ is a line, and every
line $ax + by = c'$ is such a level set. If $u(M), u(N) \geq
c$ and $t \in \intcc01$, then $u\bigl(\operatorname{bar}(M,
1-t; N, t)\bigr) = (1-t)u(M) + tu(N) \geq c$: half-planes are
convex.

\textbf{6.} The conditions $\sum\mu_i = 0$, $3\mu_2 + \mu_4 =
0$, $3\mu_3 + \mu_4 = 0$ give (taking $\mu_4 = 3$) the
dependence $(\mu_1, \mu_2, \mu_3, \mu_4) = (-1, -1, -1, 3)$.
Signs split as $\{A_4\} \mid \{A_1, A_2, A_3\}$, and
normalizing each side by $3$:
\[
A_4 = \operatorname{bar}\bigl(A_1, \tfrac13;\ A_2, \tfrac13;\
A_3, \tfrac13\bigr) = (1,1),
\]
the centroid of the triangle: the Radon point is $A_4$
itself, which indeed lies inside the triangle $A_1A_2A_3$.

\textbf{7.} The linear map $\Phi \colon \R^{n+2} \to \R
\times \R^n$, $\mu \mapsto (\sum\mu_i,\
\sum\mu_i\vect{OA_i})$, has rank $\leq n + 1$, so $\dim\ker
\Phi \geq 1$. If two independent dependences $\mu, \mu'$
existed, a suitable combination $\nu = \mu'_{n+2}\mu -
\mu_{n+2}\mu'$ (or $\mu$ itself if both last coefficients
vanish) would be a \emph{nonzero} dependence with $\nu_{n+2}
= 0$; restricting to $A_1, \dots, A_{n+1}$ and rebasing at
$A_1$, some $\nu_i \neq 0$ with $i \geq 2$ (a single nonzero
weight cannot sum to zero), giving a nontrivial relation
$\sum_{i\geq2}\nu_i\vect{A_1A_i} = 0$: the $n+1$ points would
be affinely dependent, against general position. So $\dim\ker
\Phi = 1$. The same restriction argument shows a nonzero
dependence has no vanishing coefficient.

\textbf{8.} Let $\mu \neq 0$ be a dependence, $P = \{i :
\mu_i > 0\}$ and $N = \{i : \mu_i < 0\}$: both nonempty
($\sum\mu_i = 0$, $\mu \neq 0$) and exhaustive (no zero
coefficient). Radon's construction (\cref{exo:b2:affine:7})
produces the common hull point from exactly this partition.
Since the dependence is unique up to a nonzero scalar
(question 7), the unordered pair $\{P, N\}$ --- hence the
Radon partition --- is unique.

\textbf{9.} Blocks are nonempty, so the type is $(1,3)$ or
$(2,2)$. Type $(1,3)$, block $\{j\}$: the Radon point lies in
$\operatorname{conv}\{A_j\} = \{A_j\}$, so $A_j \in
\operatorname{conv}$ of the other three; it cannot lie on an
edge (three of the points would be aligned, against general
position), so $A_j$ is interior to the triangle. Type
$(2,2)$, blocks $\{i,j\} \mid \{k,l\}$: the Radon point $z$
lies on $\intcc{A_i}{A_j} \cap \intcc{A_k}{A_l}$, and $z$ is
not an endpoint (that would align three points): the two
segments cross at an interior point. Moreover no point lies
in the hull of the others: such a containment $A_l =
\operatorname{bar}(A_i, \lambda_i)_{i \neq l}$ with
$\lambda_i \geq 0$ is an affine dependence with sign pattern
$(+,+,+,-)$, which by uniqueness (question 8) would make the
partition $(1,3)$. So in the $(2,2)$ case the four points are
in convex position and the crossing segments are the
diagonals.

\textbf{10.} The equations $\mu_2 + \mu_3 = 0$, $\mu_3 +
\mu_4 = 0$, $\sum\mu_i = 0$ give the dependence $(1, -1, 1,
-1)$: partition $\{(0,0), (1,1)\} \mid \{(1,0), (0,1)\}$, and
\[
\operatorname{bar}\bigl((0,0), \tfrac12;\ (1,1),
\tfrac12\bigr) = \bigl(\tfrac12, \tfrac12\bigr) =
\operatorname{bar}\bigl((1,0), \tfrac12;\ (0,1),
\tfrac12\bigr) :
\]
the Radon point is the center of the square, where the two
diagonals cross --- type $(2,2)$, as the picture predicts.

\textbf{11.} Radon applied to $x_1, \dots, x_4$ gives blocks
$I \mid J$ and a point $z \in \operatorname{conv}\{x_i : i
\in I\} \cap \operatorname{conv}\{x_j : j \in J\}$. Fix $k
\in \{1, \dots, 4\}$, say $k \in I$. Every $j \in J$
satisfies $j \neq k$, so $x_j \in C_k$ by the choice $x_j \in
\bigcap_{l \neq j}C_l$; since $C_k$ is convex, $z \in
\operatorname{conv}\{x_j : j \in J\} \subseteq C_k$. As $k$
was arbitrary, $z \in C_1 \cap C_2 \cap C_3 \cap C_4$.

\textbf{12.} Induction on $m$. For $m = 3$ the hypothesis is
the conclusion; $m = 4$ is question 11. Let $m \geq 4$,
assume the statement for $m$ sets, and take $C_1, \dots,
C_{m+1}$ with the triple-intersection property. Set $C_m' =
C_m \cap C_{m+1}$, convex. The family $C_1, \dots, C_{m-1},
C_m'$ has $m$ members; a triple avoiding $C_m'$ intersects by
hypothesis, and a triple $\{C_i, C_j, C_m'\}$ has
intersection $C_i \cap C_j \cap C_m \cap C_{m+1}$, nonempty
by question 11 applied to $C_i, C_j, C_m, C_{m+1}$ (any three
of these meet, by hypothesis). The induction hypothesis now
yields a common point of the new family, i.e.\ of all $m+1$
sets.

\textbf{13.} (a) The closed edges $\intcc PQ$, $\intcc QR$,
$\intcc RP$ of a nondegenerate triangle: any two share a
vertex, but a common point of all three would lie in $\intcc
PQ \cap \intcc RP = \{P\}$ and in $\intcc QR$, which excludes
$P$. (b) Any three of the sets $S_i = \{x_1, \dots, x_4\}
\setminus \{x_i\}$ omit three of the four points, leaving
exactly one common point; the total intersection omits every
point. The $S_i$ are finite, not convex: convexity is
essential. (c) Finitely many $H_k = \intco{k}{+\infty} \times
\R$ intersect in $\intco{k_{\max}}{+\infty} \times \R \neq
\emptyset$, yet no point has $x \geq k$ for all $k \in \N$:
for infinite families, compactness is essential.

\textbf{14.} Suppose $\bigcap_{i \in I}K_i = \emptyset$ and
fix $i_0$. Every $x \in K_{i_0}$ misses some $K_i$, so
$K_{i_0} \subseteq \bigcup_{i \in I}(\R^2 \setminus K_i)$, a
cover by open sets ($K_i$ is compact, hence closed). By
Borel--Lebesgue (\cref{thm:b2:metric:borellebesgue}) finitely
many suffice: $K_{i_0} \cap K_{i_1} \cap \dots \cap K_{i_N} =
\emptyset$. But any three members of this finite family of
convex sets intersect, so question 12 makes the intersection
nonempty: contradiction.

\textbf{15.} Set $D_p = \overline D(p, r)$ for $p \in S$:
compact convex sets. For $p, q, s \in S$, the hypothesis
gives a closed disk $\overline D(z, r)$ containing $p, q, s$;
then $\norm{\vect{zp}}, \norm{\vect{zq}}, \norm{\vect{zs}}
\leq r$, i.e.\ $z \in D_p \cap D_q \cap D_s$. By Helly
(question 12; the family is finite) there is $c \in
\bigcap_{p \in S}D_p$: every $p \in S$ satisfies
$\norm{\vect{cp}} \leq r$, so $S \subseteq \overline D(c,
r)$.

\textbf{16.} Let $c = x_{\lceil n/2\rceil}$. A closed
half-line containing $c$ is $\intoc{-\infty}{t}$ with $t \geq
c$ or $\intco{t}{+\infty}$ with $t \leq c$. The first
contains $x_1, \dots, x_{\lceil n/2\rceil}$: at least
$\lceil n/2\rceil \geq n/2$ points. The second contains
$x_{\lceil n/2\rceil}, \dots, x_n$: exactly $n - \lceil
n/2\rceil + 1 = \floor{n/2} + 1 > n/2$ points.

\textbf{17.} $\abs{A \cap B} = \abs A + \abs B - \abs{A \cup
B} \geq \abs A + \abs B - n > \tfrac{4n}3 - n = \tfrac n3$,
then
\[
\abs{A \cap B \cap C} \geq \abs{A \cap B} + \abs C - n >
\tfrac n3 + \tfrac{2n}3 - n = 0 .
\]

\textbf{18.} Note $m > 2n/3$. For $T_1, T_2, T_3 \subseteq S$
of cardinality $m$, question 17 provides a point $x \in T_1
\cap T_2 \cap T_3$; then $x \in \operatorname{conv}(T_i)$ for
each $i$: any three members of $\mathcal F$ meet. The family
is finite (finitely many subsets of $S$) and consists of
convex sets, so Helly (question 12) gives $c \in
\bigcap_{\abs T = m}\operatorname{conv}(T)$.

\textbf{19.} Suppose some closed half-plane $H \ni c$
contains fewer than $n/3$ points of $S$. Its complement $U$
is an open half-plane, convex, with $\abs{S \cap U} > 2n/3$,
hence $\abs{S \cap U} \geq \floor{2n/3} + 1 = m$; choose $T
\subseteq S \cap U$ with $\abs T = m$. Then
$\operatorname{conv}(T) \subseteq U$ by convexity of $U$, so
$c \in \operatorname{conv}(T) \subseteq U$: contradiction
with $c \in H$. Hence every closed half-plane containing $c$
contains at least $n/3$ points: $c$ is a centerpoint.

\textbf{20.} Take the triangle equilateral of side $L$ and
$\varepsilon = L/100$. Let $c$ be any point; we exhibit a
closed half-plane containing $c$ and at most $k$ points.

\emph{Case 1: $c$ is within $L/10$ of a vertex, say $B$.}
Directions from $c$ to $A$ and to $C$ deviate from the
directions $B \to A$, $B \to C$ by at most
$\arcsin\bigl(\tfrac{L/10}{9L/10}\bigr) = \arcsin\tfrac19$,
so they make an angle $\leq \tfrac\pi3 + 2\arcsin\tfrac19 <
\tfrac{2\pi}3$. \emph{Case 2: $c$ is at distance $> L/10$
from all vertices.} If $c$ is in the triangle, the three
angular gaps between the directions $u_A, u_B, u_C$ from $c$
to the vertices sum to $2\pi$, so some gap is $\leq 2\pi/3$;
if $c$ is outside, the three directions lie in an open
half-plane of directions and two of them make an angle $<
\pi/2$. In every case two directions, say toward $X$ and $Y$,
make an angle $\leq 2\pi/3$; let $w$ be their unit bisector,
so $\langle u_X, w\rangle, \langle u_Y, w\rangle \geq
\cos\tfrac\pi3 = \tfrac12$. For any point $b$ of the disk
around $X$:
\[
\langle \vect{cb}, w\rangle \geq
\tfrac12\norm{\vect{cX}} - \varepsilon > 0,
\]
since $\norm{\vect{cX}} \geq L/10 > 2\varepsilon$ (and
likewise for $Y$): the open half-plane $\{\langle \vect{cx},
w\rangle > 0\}$ swallows both clusters. Its closed
complement contains $c$ and at most the $k$ points of the
third cluster. So no point of the plane beats $n/3$: with
question 19, the centerpoint constant is exactly $1/3$.

\textbf{21.} Order the angles; the largest, $\theta$,
satisfies $\theta \geq \pi/3$ (the three sum to $\pi$). If
$\theta \geq \pi/2$, say at $R$, let $M$ be the midpoint of
the opposite side $\intcc PQ$. The median formula ($\vect{RM}
= \tfrac12(\vect{RP} + \vect{RQ})$, expand and eliminate
$\langle\vect{RP}, \vect{RQ}\rangle$ with the law of
cosines) gives
\[
\norm{\vect{RM}}^2 = \tfrac12\norm{\vect{RP}}^2 +
\tfrac12\norm{\vect{RQ}}^2 - \tfrac14\norm{\vect{PQ}}^2 \leq
\tfrac12\norm{\vect{PQ}}^2 - \tfrac14\norm{\vect{PQ}}^2 =
\tfrac14\norm{\vect{PQ}}^2,
\]
using $\norm{\vect{PQ}}^2 = \norm{\vect{RP}}^2 +
\norm{\vect{RQ}}^2 - 2\langle\vect{RP}, \vect{RQ}\rangle \geq
\norm{\vect{RP}}^2 + \norm{\vect{RQ}}^2$ (the inner product
is $\leq 0$). So the disk of diameter $\intcc PQ$, of radius
$\leq \tfrac12 < \tfrac1{\sqrt3}$, contains all three points
(degenerate aligned triples fall under $\theta = \pi$). If
$\theta < \pi/2$ the triangle is acute; by the law of sines
the circumradius is $R_c = a/(2\sin\theta)$ with $a \leq 1$
the side opposite $\theta$, and $\theta \in
\intco{\pi/3}{\pi/2}$ gives $\sin\theta \geq \sqrt3/2$, so
$R_c \leq 1/\sqrt3$: the circumdisk does the job.

\textbf{22.} For $p \in S$ let $K_p = \overline D(p,
1/\sqrt3)$: compact convex. Any three points $p, q, s$ of $S$
are pairwise at distance $\leq 1$, so question 21 gives a
disk of radius $1/\sqrt3$ containing them: its center lies in
$K_p \cap K_q \cap K_s$. By compact Helly (question 14,
arbitrary families allowed) there is $c \in \bigcap_{p\in
S}K_p$: every $p \in S$ is within $1/\sqrt3$ of $c$, i.e.\ $S
\subseteq \overline D(c, 1/\sqrt3)$. This is Jung's theorem
in the plane.

\textbf{23.} With $G$ the centroid, $\sum_i\vect{GA_i} = 0$,
so
\[
\sum_i\norm{\vect{OA_i}}^2 = \sum_i\norm{\vect{OG} +
\vect{GA_i}}^2 = 3\norm{\vect{OG}}^2 + 2\Bigl\langle
\vect{OG}, \sum_i\vect{GA_i}\Bigr\rangle +
\sum_i\norm{\vect{GA_i}}^2,
\]
and the middle term vanishes: the Leibniz identity. For the
equilateral triangle of side $1$, $\norm{\vect{GA_i}} =
1/\sqrt3$ (two thirds of the height $\sqrt3/2$), so
$\sum_i\norm{\vect{GA_i}}^2 = 1$. If $\overline D(O, r)$
contains the vertices, then $3r^2 \geq
\sum_i\norm{\vect{OA_i}}^2 = 3\norm{\vect{OG}}^2 + 1 \geq 1$:
$r \geq 1/\sqrt3$, with equality forcing $O = G$ and all
three distances equal to $r$ --- the circumdisk. Jung's
constant $1/\sqrt3$ is sharp.

\textbf{24.} \emph{Helly in $\R^n$: if $C_1, \dots, C_m$
($m \geq n + 1$) are convex subsets of $\R^n$ and any $n + 1$
of them intersect, then all of them do.} Base case $m = n +
2$: pick $x_i \in \bigcap_{j\neq i}C_j$; Radon's lemma
(\cref{exo:b2:affine:7}) splits $x_1, \dots, x_{n+2}$ into
blocks $I \mid J$ with a common hull point $z$, and for each
$k$, the block not containing $k$ consists of points of
$C_k$, so $z \in C_k$ by convexity, exactly as in question
11. Induction step for $m \geq n + 2$: replace $C_m,
C_{m+1}$ by $C_m \cap C_{m+1}$; an $(n+1)$-tuple of the new
family containing the intersected member amounts to $n + 2$
of the old sets, handled by the base case, and the other
tuples are covered by hypothesis. Conclude by the induction
hypothesis.

\textbf{25.} Linear algebra enters once: $n + 2$ vectors in
the $(n+1)$-dimensional space of pairs (total weight,
weighted position) must be dependent --- that is the affine
dependence. The signs of its coefficients split the points
into the two Radon blocks and turn one linear relation into
an equality of two nonnegative barycenters. Convexity is used
exactly twice: in Helly's step (the hull of points of $C_k$
stays in $C_k$) and in the applications (half-planes and
disks are convex). The dimension of the plane entered only
through the number $4 = 2 + 2$ of points fed to Radon, i.e.\
the ``$3 = 2 + 1$'' in Helly's hypothesis; everything else
was dimension-free, as question 24 confirms. In $\R^n$ the
constants become: Helly number $n + 1$; centerpoint constant
$\frac1{n+1}$ (every finite set has a point every closed
half-space through which contains a fraction $\geq
\frac1{n+1}$ of it); Jung radius
$\sqrt{\frac{n}{2(n+1)}}$ for sets of diameter $1$ --- equal
to $1/\sqrt3$ when $n = 2$.
\end{solution}
