\section*{Chapter \ref{ch:b2:surfaces} --- Surfaces}

\begin{solution}{exo:b2:surfaces:1}
With $\sigma(\theta, r) = (r\cos\theta,\ r\sin\theta,\ r)$:
\[
\sigma_\theta = (-r\sin\theta,\ r\cos\theta,\ 0),
\qquad
\sigma_r = (\cos\theta,\ \sin\theta,\ 1),
\]
independent for $r > 0$. The \omterm{def:b2:surfaces:tangent}{tangent plane} at $M_0 =
\sigma(\theta_0, r_0)$ passes through $M_0$ with directions
$\sigma_\theta, \sigma_r$. Now $M_0 - O = r_0(\cos\theta_0,
\sin\theta_0, 1) = r_0\,\sigma_r(\theta_0, r_0)$ is itself a
tangent direction: the apex $O$ lies on the \omterm{def:b2:surfaces:tangent}{tangent plane}. (This
is the general behaviour of cones: they are ruled by lines through
the apex, and a \omterm{def:b2:surfaces:tangent}{tangent plane} contains the ruling line through the
point of tangency.)
\end{solution}

\begin{solution}{exo:b2:surfaces:2}
Apply \cref{prop:b2:surfaces:gradient} to $f(x,y,z) =
\frac{x^2}{a^2} + \frac{y^2}{b^2} + \frac{z^2}{c^2}$: $\nabla
f(x_0,y_0,z_0) = 2\bigl(\frac{x_0}{a^2}, \frac{y_0}{b^2},
\frac{z_0}{c^2}\bigr) \neq 0$ on the surface. The \omterm{def:b2:surfaces:tangent}{tangent plane} is
\[
\frac{x_0}{a^2}(x - x_0) + \frac{y_0}{b^2}(y - y_0)
+ \frac{z_0}{c^2}(z - z_0) = 0,
\qquad\text{i.e.}\qquad
\frac{x_0\,x}{a^2} + \frac{y_0\,y}{b^2} + \frac{z_0\,z}{c^2} = 1,
\]
using that $(x_0, y_0, z_0)$ satisfies the ellipsoid equation ---
the ``split the squares'' rule generalizing the sphere's $\langle
M_0, M \rangle = R^2$.
\end{solution}

\begin{solution}{exo:b2:surfaces:3}
$\sigma_u = (-v\sin u,\ v\cos u,\ a)$ and $\sigma_v = (\cos u,\
\sin u,\ 0)$, so
\[
E = v^2 + a^2, \qquad F = 0, \qquad G = 1,
\qquad
\sqrt{EG - F^2} = \sqrt{v^2 + a^2} > 0
\]
(the helicoid is regular everywhere, including on its axis $v =
0$). \omterm{def:b2:surfaces:area}{Area} of the piece:
\[
\mathcal{A} = \int_0^{2\pi}\!\!\int_0^1
\sqrt{v^2 + a^2}\;\dd v\,\dd u
= 2\pi\cdot\frac12\Bigl[v\sqrt{v^2 + a^2}
+ a^2\ln\bigl(v + \sqrt{v^2 + a^2}\bigr)\Bigr]_0^1 ,
\]
i.e.\ $\mathcal{A} = \pi\Bigl(\sqrt{1 + a^2} +
a^2\ln\frac{1 + \sqrt{1 + a^2}}{a}\Bigr)$.
\end{solution}

\begin{solution}{exo:b2:surfaces:4}
Derivatives:
\[
\sigma_\theta = \bigl(-(R + r\cos\psi)\sin\theta,\
(R + r\cos\psi)\cos\theta,\ 0\bigr),
\qquad
\sigma_\psi = \bigl(-r\sin\psi\cos\theta,\
-r\sin\psi\sin\theta,\ r\cos\psi\bigr).
\]
Then
\[
E = (R + r\cos\psi)^2,
\qquad
F = r\sin\psi\cos\psi\,(R + r\cos\psi)
\bigl(\sin\theta\cos\theta - \sin\theta\cos\theta\bigr) = 0,
\qquad
G = r^2 ,
\]
so $\sqrt{EG - F^2} = r(R + r\cos\psi) \geq r(R - r) > 0$:
regular everywhere. \omterm{def:b2:surfaces:area}{Area}:
\[
\mathcal{A}
= \int_0^{2\pi}\!\!\int_0^{2\pi} r(R + r\cos\psi)
\,\dd\theta\,\dd\psi
= 2\pi r \Bigl(2\pi R + r\int_0^{2\pi}\cos\psi\,\dd\psi\Bigr)
= 4\pi^2 R r ,
\]
the $\cos\psi$ term integrating to zero --- Pappus' theorem: \omterm{def:b2:surfaces:area}{area}
$=$ (\omterm{def:b2:curves:length}{length} of the rotated circle) $\times$ (distance travelled by
its center).
\end{solution}

\begin{solution}{exo:b2:surfaces:5}
For $\sigma(x, y) = (x, y, f(x,y))$: $\sigma_x = (1, 0, f_x)$,
$\sigma_y = (0, 1, f_y)$, so $E = 1 + f_x^2$, $F = f_xf_y$, $G = 1
+ f_y^2$ and
\[
EG - F^2 = (1 + f_x^2)(1 + f_y^2) - f_x^2f_y^2
= 1 + f_x^2 + f_y^2 ,
\]
giving the stated \omterm{def:b2:surfaces:area}{area} formula. For $f = \frac12(x^2 + y^2)$ on
the unit disk: $1 + f_x^2 + f_y^2 = 1 + x^2 + y^2$, and in polar
coordinates ($x = \rho\cos\alpha$, $y = \rho\sin\alpha$, Jacobian
$\rho$, \cref{ch:b2:multint}):
\[
\mathcal{A}
= \int_0^{2\pi}\!\!\int_0^1 \sqrt{1 + \rho^2}\;\rho
\,\dd\rho\,\dd\alpha
= 2\pi\Bigl[\tfrac13(1 + \rho^2)^{3/2}\Bigr]_0^1
= \frac{2\pi}{3}\bigl(2\sqrt2 - 1\bigr).
\]
\end{solution}

\begin{solution}{exo:b2:surfaces:6}
From \cref{ex:b2:surfaces:spherearea}'s computation, $E =
R^2\cos^2\varphi$, $F = 0$, $G = R^2$, so by
\cref{prop:b2:surfaces:curvelength}
\[
L = \int_a^b R\sqrt{\cos^2\varphi(t)\,\theta'(t)^2
+ \varphi'(t)^2}\;\dd t .
\]
Let the endpoints be $(\theta_0, \varphi_1)$ and $(\theta_0,
\varphi_2)$, $\varphi_1 < \varphi_2$. For any joining curve,
\[
L \geq \int_a^b R\,\abs{\varphi'(t)}\,\dd t
\geq R\,\Bigl|\int_a^b \varphi'(t)\,\dd t\Bigr|
= R\,(\varphi_2 - \varphi_1),
\]
dropping the nonnegative $\cos^2\varphi\,\theta'^2$ term and using
the triangle inequality for integrals. The meridian arc
$\theta \equiv \theta_0$, $\varphi$ increasing from $\varphi_1$ to
$\varphi_2$, has \omterm{def:b2:curves:length}{length} exactly $R(\varphi_2 - \varphi_1)$: it is
shortest. (Meridians are great circles; this is the first,
elementary case of the fact that geodesics of the sphere are great
circles.)
\end{solution}

\begin{solution}{exo:b2:surfaces:7}
Fix a curve $\gamma$ drawn on $S$ and let $g(t) = \norm{\gamma(t)
- \Omega}^2$. Then $g'(t) = 2\langle \gamma'(t),\ \gamma(t) -
\Omega\rangle$. The normal line at $M = \gamma(t)$ passes through
$\Omega$ by hypothesis, so $\gamma(t) - \Omega$ is a \emph{normal}
vector, orthogonal to the \omterm{def:b2:surfaces:tangent}{tangent plane}, in particular to the
velocity $\gamma'(t)$ (\cref{prop:b2:surfaces:velocity}): $g' =
0$, and $g$ is constant along every drawn curve.

Now the set $S_c = \{M \in S : \norm{M - \Omega}^2 = c\}$ is
closed in $S$; it is also \omterm{def:b2:metric:topology}{open} in $S$: near any of its points, $S$
is a regular graph, so any nearby point of $S$ is joined to it by
a drawn curve (a lifted segment), along which $g$ is constant. As
$S$ is \omterm{def:b2:metric:connected}{connected} and $S_c$ nonempty for the right $c$, $S = S_c
\subseteq$ the sphere of center $\Omega$ and radius $\sqrt c$
(\cref{ch:b2:metric}: \omterm{def:b2:metric:connected}{connectedness} argument).
\end{solution}

\begin{solution}{exo:b2:surfaces:8}
Write $\Phi(u', v') = (u, v)$. By the chain rule,
\[
\tilde\sigma_{u'} = \frac{\partial u}{\partial u'}\sigma_u
+ \frac{\partial v}{\partial u'}\sigma_v,
\qquad
\tilde\sigma_{v'} = \frac{\partial u}{\partial v'}\sigma_u
+ \frac{\partial v}{\partial v'}\sigma_v ,
\]
partials of $\sigma$ evaluated at $\Phi(u',v')$. Expanding the
cross product bilinearly and using $\sigma_u \wedge \sigma_u =
\sigma_v \wedge \sigma_v = 0$, $\sigma_v \wedge \sigma_u =
-\sigma_u \wedge \sigma_v$:
\[
\tilde\sigma_{u'} \wedge \tilde\sigma_{v'}
= \Bigl(\frac{\partial u}{\partial u'}
\frac{\partial v}{\partial v'}
- \frac{\partial v}{\partial u'}
\frac{\partial u}{\partial v'}\Bigr)\,
\sigma_u \wedge \sigma_v
= \det J_\Phi \cdot (\sigma_u \wedge \sigma_v)\circ\Phi .
\]
Taking \omterm{def:b2:nvs:norm}{norms} gives the identity. Then, by the change-of-variables
formula (\cref{ch:b2:multint}) applied to the map $\Phi$ on $K' =
\Phi^{-1}(K)$:
\[
\iint_{K'} \norm{\tilde\sigma_{u'} \wedge \tilde\sigma_{v'}}
\,\dd u'\dd v'
= \iint_{K'} \norm{\sigma_u \wedge \sigma_v}\circ\Phi\;
\abs{\det J_\Phi}\,\dd u'\dd v'
= \iint_K \norm{\sigma_u \wedge \sigma_v}\,\dd u\,\dd v :
\]
the two parametrizations assign the same \omterm{def:b2:surfaces:area}{area} to the same piece of
surface.
\end{solution}

\begin{solution}{exo:b2:surfaces:9}
In the spherical chart, $z = R\sin\varphi$, and the zone
corresponds to $\varphi_1 \leq \varphi \leq \varphi_2$ with
$z_i = R\sin\varphi_i$. With the \omterm{def:b2:surfaces:area}{area} element
$R^2\cos\varphi\,\dd\theta\,\dd\varphi$
(\cref{ex:b2:surfaces:spherearea}):
\[
\mathcal A = \int_{\varphi_1}^{\varphi_2}\!\!\int_0^{2\pi}
R^2\cos\varphi\,\dd\theta\,\dd\varphi
= 2\pi R^2(\sin\varphi_2 - \sin\varphi_1)
= 2\pi R\,(z_2 - z_1) .
\]
The result depends only on the height $z_2 - z_1$: slices of
equal thickness carry equal \omterm{def:b2:surfaces:area}{areas}, whether cut at the equator
or at the pole --- Archimedes' hat-box theorem, and the reason
the lateral \omterm{def:b2:surfaces:area}{area} of the circumscribed cylinder ($2\pi R \cdot
2R = 4\pi R^2$) equals the \omterm{def:b2:surfaces:area}{area} of the sphere.
\end{solution}

\begin{solution}{exo:b2:surfaces:10}
$\sigma_\theta = (-r\sin\theta,\ r\cos\theta,\ 0)$ and
$\sigma_z = (r'\cos\theta,\ r'\sin\theta,\ 1)$, so
\[
\sigma_\theta \wedge \sigma_z
= (r\cos\theta,\ r\sin\theta,\ -r\,r') ,
\]
a normal vector at $M = (r\cos\theta, r\sin\theta, z)$. The
normal line is
\[
t \mapsto \bigl(r(1 + t)\cos\theta,\ r(1 + t)\sin\theta,\
z - t\,r\,r'\bigr),
\]
which at $t = -1$ reaches $(0,\ 0,\ z + r(z)\,r'(z))$: every
normal line meets the axis, at the height $z + rr'$. (This is
the three-dimensional reason rotational symmetry survives in
the normal field.)
\end{solution}

\begin{solution}{exo:b2:surfaces:11}
$\sigma_u = (-\sin u, \cos u, 0)$, $\sigma_v = (0, 0, 1)$: $E
= 1$, $F = 0$, $G = 1$. By
\cref{prop:b2:surfaces:curvelength} the \omterm{def:b2:curves:length}{length} of a drawn
curve is $\int\sqrt{u'^2 + v'^2}\,\dd t$ --- the \omterm{def:b2:curves:length}{length} of its
parameter shadow $(u(t), v(t))$ in the plane: the chart is a
local isometry (the unrolling of the cylinder). The helix
$t \mapsto (\cos t, \sin t, ct)$, $t \in \intcc0{2\pi}$, has
shadow the segment from $(0,0)$ to $(2\pi, 2\pi c)$, of \omterm{def:b2:curves:length}{length}
$2\pi\sqrt{1 + c^2}$. Any drawn curve from $(1,0,0)$ to $(1,
0, 2\pi c)$ making one full turn has a \omterm{def:b2:metric:continuity}{continuous} shadow
joining $(0, 0)$ to $(2\pi, 2\pi c)$, of plane \omterm{def:b2:curves:length}{length} $\geq$
the straight segment; since \omterm{def:b2:curves:length}{lengths} agree, the helix is
shortest.
\end{solution}

\begin{solution}{exo:b2:surfaces:12}
The function $g(M) = \norm{M}^2$ is \omterm{def:b2:metric:continuity}{continuous} on the \omterm{def:b2:metric:compact}{compact}
$S$, so it attains its maximum at some $M_0$. For every curve
$\gamma$ drawn on $S$ with $\gamma(0) = M_0$, the function $t
\mapsto \norm{\gamma(t)}^2$ has a maximum at $t = 0$, so its
derivative $2\langle\gamma'(0), M_0\rangle$ vanishes: $M_0$ is
orthogonal to every tangent vector, i.e.\ normal to $S$ at
$M_0$. As $\nabla f(M_0) \neq 0$ also directs the normal line
(\cref{prop:b2:surfaces:gradient}), $\nabla f(M_0)$ and
$\vect{OM_0}$ are collinear. For the ellipsoid, radiality
means
\[
\Bigl(\frac{x_0}{a^2}, \frac{y_0}{b^2},
\frac{z_0}{c^2}\Bigr) = \mu\,(x_0, y_0, z_0) :
\]
each coordinate satisfies $x_0(\frac1{a^2} - \mu) = 0$, etc.;
since $a^{-2}, b^{-2}, c^{-2}$ are distinct, at most one
coordinate is nonzero, and the solutions on the surface are
the six axis endpoints $(\pm a, 0, 0)$, $(0, \pm b, 0)$, $(0,
0, \pm c)$. The farthest points are $(\pm a, 0, 0)$, at
distance $a = \max(a,b,c)$.
\end{solution}

\begin{solution}{pb:b2:surfaces:1}
\textbf{1.} Expanding, and using the symmetry of $A$
($t^{\mathsf T}\!APY = (At)^{\mathsf T}PY$):
\[
q(PY + t) = Y^{\mathsf T}P^{\mathsf T}\!APY +
2\,(At + b)^{\mathsf T}PY + \bigl(t^{\mathsf T}\!At +
2b^{\mathsf T}t + c\bigr),
\]
which is the displayed triple. $A' = P^{\mathsf T}\!AP =
P^{-1}AP$ is similar to $A$: same \omterm{def:b2:reduction:charpoly}{characteristic polynomial},
\omterm{def:b2:reduction:eigen}{spectrum}, rank, signature. Finally $\{q = 0\} = \{sq = 0\}$
for $s \neq 0$, so only the equation \emph{up to a scalar} is
attached to the set; scaling by $s$ multiplies all \omterm{def:b2:reduction:eigen}{eigenvalues}
by $s$.

\textbf{2.} The spectral theorem provides $P \in O(3)$ with
$P^{\mathsf T}\!AP = \operatorname{diag}(\lambda_1, \lambda_2,
\lambda_3)$; question 1 with $t = 0$ turns the equation into
$\sum\lambda_iy_i^2 + 2\sum\beta_iy_i + c = 0$, where $\beta =
P^{\mathsf T}b$.

\textbf{3.} For $\lambda_i \neq 0$: $\lambda_iy_i^2 +
2\beta_iy_i = \lambda_i\bigl(y_i +
\frac{\beta_i}{\lambda_i}\bigr)^2 -
\frac{\beta_i^2}{\lambda_i}$; the translation $z_i = y_i +
\beta_i/\lambda_i$ (and $z_i = y_i$ for $i > r$) gives
\[
\sum_{i \leq r}\lambda_iz_i^2 + 2\sum_{i > r}\beta_iz_i +
c'' = 0, \qquad
c'' = c - \sum_{i\leq r}\frac{\beta_i^2}{\lambda_i}.
\]

\textbf{4.} $q(2\Omega - X) = (2\Omega - X)^{\mathsf T}
A(2\Omega - X) + 2b^{\mathsf T}(2\Omega - X) + c$; expanding
and subtracting $q(X)$, the quadratic terms cancel and
\[
q(2\Omega - X) - q(X) = -4\,\langle A\Omega + b,\ X\rangle +
4\,\langle A\Omega + b,\ \Omega\rangle .
\]
If $A\Omega + b = 0$ this vanishes identically: the point
reflection preserves $q$, hence the \omterm{pb:b2:surfaces:1}{quadric}. Conversely,
``$q(2\Omega - X) = q(X)$ for all $X$'' says that the \omterm{def:b2:affine:subspace}{affine}
function above vanishes on all of $\R^3$, which forces its
linear part $A\Omega + b$ to be zero. So centers $=$ solutions of
$A\Omega = -b$: a nonempty set iff $b \in \operatorname{im}A$
(an \omterm{def:b2:affine:subspace}{affine} subspace directed by $\ker A$), and a single point
iff $A$ is invertible.

\textbf{5.} Signature $(3,0)$ (all $\lambda_i > 0$): $\delta >
0$ gives $\frac{x^2}{a^2} + \frac{y^2}{b^2} + \frac{z^2}{c^2}
= 1$ with $a = \sqrt{\delta/\lambda_1}$, etc.\ --- an
ellipsoid; $\delta = 0$: the single point $O$; $\delta < 0$:
empty. Signature $(2,1)$ ($\lambda_1, \lambda_2 > 0 >
\lambda_3$): $\delta > 0$: $\frac{x^2}{a^2} + \frac{y^2}{b^2}
- \frac{z^2}{c^2} = 1$, the one-sheet hyperboloid; $\delta =
0$: the cone $\frac{x^2}{a^2} + \frac{y^2}{b^2} =
\frac{z^2}{c^2}$; $\delta < 0$: $\frac{z^2}{c^2} -
\frac{x^2}{a^2} - \frac{y^2}{b^2} = 1$, the two-sheet
hyperboloid ($\abs z \geq c$: two components).

\textbf{6.} The quadratic part has matrix $A$ with diagonal
$1$ and off-diagonal $2$: $A = 2J - I$. Since $J$ has \omterm{def:b2:reduction:eigen}{spectrum}
$\{3, 0, 0\}$ (\omterm{def:b2:reduction:eigen}{eigenvector} $(1,1,1)$ for $3$), $A$ has
\omterm{def:b2:reduction:eigen}{spectrum} $\{5, -1, -1\}$, the \omterm{def:b2:reduction:eigen}{eigenvalue} $5$ carried by
$\R(1,1,1)$. In the rotated coordinates: $5u^2 - v^2 - w^2 =
1$, i.e.\ $\frac{u^2}{1/5} - v^2 - w^2 = 1$: a two-sheet
hyperboloid, of revolution (equal \omterm{def:b2:reduction:eigen}{eigenvalues} $-1$) about the
axis $\R(1,1,1)$.

\textbf{7.} The surface is $\bigl(\frac xa - \frac
zc\bigr)\bigl(\frac xa + \frac zc\bigr) = \bigl(1 - \frac
yb\bigr)\bigl(1 + \frac yb\bigr)$. For $(\lambda : \mu) \neq
(0:0)$ define the line
\[
D_{\lambda:\mu} :\quad
\lambda\Bigl(\frac xa - \frac zc\Bigr) = \mu\Bigl(1 - \frac
yb\Bigr), \qquad
\mu\Bigl(\frac xa + \frac zc\Bigr) = \lambda\Bigl(1 + \frac
yb\Bigr)
\]
(two independent \omterm{def:b2:affine:subspace}{affine} equations: a line). Multiplying the
two equations shows every point of $D_{\lambda:\mu}$ lies on
the surface when $\lambda\mu \neq 0$; the cases $\lambda = 0$
or $\mu = 0$ are checked directly (e.g.\ $\lambda = 0$: $y =
b$, $\frac xa = -\frac zc$, which satisfies the equation). The
second family $D'_{\lambda:\mu}$ swaps the two right-hand
factors.

\textbf{8.} Fix $M$ on the surface. The conditions for $M \in
D_{\lambda:\mu}$ form a $2\times2$ homogeneous linear system
in $(\lambda, \mu)$ whose \omterm{def:b2:linalg:det}{determinant} is
\[
\Bigl(\frac{x^2}{a^2} - \frac{z^2}{c^2}\Bigr) - \Bigl(1 -
\frac{y^2}{b^2}\Bigr) = 0
\]
precisely because $M$ lies on the \omterm{pb:b2:surfaces:1}{quadric}: a nontrivial
solution $(\lambda : \mu)$ exists. The coefficient matrix is
never zero (that would force $1 - \frac yb = 1 + \frac yb =
0$), so its rank is $1$ and the solution is unique up to
scale: exactly one line of the family passes through $M$.
The same holds for the second family, and the two lines are
distinct (at $(a, 0, 0)$ they are $\{x = a,\ \frac zc = \frac
yb\}$ and $\{x = a,\ \frac zc = -\frac yb\}$): the one-sheet
hyperboloid is doubly ruled.

\textbf{9.} Let $M = (x, y, z)$ on the hyperboloid, $\rho^2 =
\frac{x^2}{a^2} + \frac{y^2}{b^2} = 1 + \frac{z^2}{c^2} \geq
1$. The point $N = (x,\ y,\ \varepsilon c\rho)$ with
$\varepsilon$ the sign of $z$ satisfies $\frac{x^2}{a^2} +
\frac{y^2}{b^2} - \frac{(c\rho)^2}{c^2} = 0$: $N \in C$, and
\[
d(M, C) \leq \abs{z - \varepsilon c\rho}
= c\bigl(\rho - \sqrt{\rho^2 - 1}\bigr)
= \frac{c}{\rho + \sqrt{\rho^2 - 1}} .
\]
If $\norm M \to \infty$ then $\rho \to \infty$ (all three
coordinates are bounded by multiples of $\rho$), so $d(M, C)
\to 0$. The section $x = a$: $\frac{y^2}{b^2} -
\frac{z^2}{c^2} = 0$, the pair of crossing lines of question
8 --- and likewise at $x = -a$.

\textbf{10.} With $\lambda_3 = 0$, question 3 leaves
$\lambda_1z_1^2 + \lambda_2z_2^2 + 2\beta_3z_3 + c'' = 0$. In
the eigenbasis, $\operatorname{im}A = \operatorname{Vect}(e_1,
e_2)$, so by question 4 centers exist iff $\beta_3 = 0$. If
$\beta_3 \neq 0$: the translation $z_3 \mapsto z_3 -
c''/(2\beta_3)$ removes the constant, leaving $\lambda_1z_1^2
+ \lambda_2z_2^2 + 2\beta_3z_3 = 0$, i.e.\ $z_3 = px^2 +
qy^2$ after renaming: an \emph{elliptic paraboloid} if
$\lambda_1\lambda_2 > 0$, a \emph{hyperbolic paraboloid} if
$\lambda_1\lambda_2 < 0$ --- and indeed no center. If
$\beta_3 = 0$: the equation $\lambda_1z_1^2 + \lambda_2z_2^2 +
c'' = 0$ does not involve $z_3$: the \omterm{pb:b2:surfaces:1}{quadric} is a cylinder
over the corresponding plane conic --- elliptic cylinder,
line, or empty set when $\lambda_1\lambda_2 > 0$; hyperbolic
cylinder or pair of intersecting planes when
$\lambda_1\lambda_2 < 0$ --- with a whole line of centers
$\{(z_1^*, z_2^*)\} \times \R$.

\textbf{11.} $xy - z = 0$. Substituting $x = \frac{u +
v}{\sqrt2}$, $y = \frac{u - v}{\sqrt2}$ (rotation by $\pi/4$):
$xy = \frac{u^2 - v^2}2$, so the equation becomes $z =
\frac12(u^2 - v^2)$: a hyperbolic paraboloid --- the saddle
of the figure, up to the factor $\frac12$.

\textbf{12.} The line $\{x = x_0,\ z = x_0y\}$ (parametrized
by $y$) clearly lies on $z = xy$, as does $\{y = y_0,\ z =
xy_0\}$; through $(x_0, y_0, x_0y_0)$ pass the two of them.
Uniqueness: if $t \mapsto (x_0 + tv_1, y_0 + tv_2, z_0 +
tv_3)$ stays on the surface, the coefficient of $t^2$ in $(x_0
+ tv_1)(y_0 + tv_2) - z_0 - tv_3$ gives $v_1v_2 = 0$, so
$v_1 = 0$ or $v_2 = 0$, landing in one of the two families:
one line of each through each point --- the second doubly
ruled \omterm{pb:b2:surfaces:1}{quadric}.

\textbf{13.} Completing squares: $(x - 1)^2 + (y + 2)^2 = 2$,
with no condition on $z$: a right circular cylinder of radius
$\sqrt2$ and axis the vertical line $\{(1, -2, z) : z \in
\R\}$ --- a line of centers, as question 10 predicts.

\textbf{14.} With $\lambda_2 = \lambda_3 = 0$ the reduced
equation is $\lambda_1z_1^2 + 2\beta_2z_2 + 2\beta_3z_3 + c''
= 0$. A rotation of the kernel plane $(z_2, z_3)$ aligns the
linear form: $2\beta_2z_2 + 2\beta_3z_3 = 2\beta w$ with
$\beta = \sqrt{\beta_2^2 + \beta_3^2}$. If $\beta \neq 0$,
translate $w$ to absorb $c''$: $\lambda_1z_1^2 + 2\beta w =
0$, a \emph{parabolic cylinder}; if $\beta = 0$:
$\lambda_1z_1^2 = -c''$ gives two parallel planes ($c''
\lambda_1 < 0$), a double plane ($c'' = 0$), or the empty
set. For $(x + y)^2 = z$: with $u = \frac{x + y}{\sqrt2}$ the
equation reads $z = 2u^2$: a parabolic cylinder, invariant
under translations along $(1, -1, 0)$.

\textbf{15.} Every \omterm{pb:b2:surfaces:1}{quadric} is carried by a rotation plus
translations onto one of: (rank 3) ellipsoid, point, empty
set, one-sheet hyperboloid, cone, two-sheet hyperboloid;
(rank 2) elliptic paraboloid, hyperbolic paraboloid, elliptic
cylinder, line, empty set, hyperbolic cylinder, pair of
intersecting planes; (rank 1) parabolic cylinder, pair of
parallel planes, double plane, empty set. Identifying the
three empty variants as distinct \emph{\omterm{def:b2:affine:subspace}{affine} types} of
equations, the count is seventeen; among them nine are honest
surfaces: ellipsoid, the two hyperboloids, the cone, the two
paraboloids, and the three cylinders.

\textbf{16.} Algorithm. (i) Read off $(A, b, c)$; compute the
\omterm{def:b2:reduction:charpoly}{characteristic polynomial} of $A$, its \omterm{def:b2:reduction:eigen}{eigenvalues} (real, by
the spectral theorem) and $r = \operatorname{rank}A$. (ii)
Solve $A\Omega = -b$ (Gaussian elimination): solvable or not
--- centers or not. (iii) If solvable, translate to a center:
the equation becomes $\sum\lambda_iz_i^2 + c'' = 0$ with $c''
= c + \langle b, \Omega\rangle$; sort by $r$, the signature,
and the sign of $c''$ using questions 5, 10, 14. (iv) If not
solvable ($r \leq 2$), rotate and reduce as in questions 10
and 14: paraboloid ($r = 2$) or parabolic cylinder ($r = 1$),
elliptic/hyperbolic according to the sign of
$\lambda_1\lambda_2$. Each step is a finite computation:
roots of a cubic with real roots, ranks, linear systems.

\textbf{17.} $A$ has diagonal $1$, off-diagonal $-1$: $A = 2I
- J$, \omterm{def:b2:reduction:eigen}{spectrum} $\{2 - 3,\ 2,\ 2\} = \{-1, 2, 2\}$ with $-1$
on $\R(1,1,1)$. Signature $(2,1)$, $b = 0$, right side
$\delta = 1 > 0$: $2u^2 + 2v^2 - w^2 = 1$, a one-sheet
hyperboloid of revolution about the axis $\R(1,1,1)$.

\textbf{18.} Complete squares: $(x-1)^2 + (y+2)^2 - (z-1)^2 +
(-1 - 4 + 1 + 4) = 0$, i.e.
\[
(x-1)^2 + (y+2)^2 = (z-1)^2 :
\]
the constant vanished --- a right circular \emph{cone} with
vertex (and unique center) $(1, -2, 1)$ and axis parallel to
$Oz$.

\textbf{19.} The quadratic part $x^2 + 4xy + y^2$ has matrix
$\begin{psmallmatrix}1 & 2\\ 2 & 1\end{psmallmatrix}$ (in the
$xy$-plane), \omterm{def:b2:reduction:eigen}{eigenvalues} $3$ (on $(1,1)$) and $-1$ (on
$(1,-1)$): with $u = \frac{x+y}{\sqrt2}$, $v =
\frac{x-y}{\sqrt2}$ it equals $3u^2 - v^2$, and the \omterm{pb:b2:surfaces:1}{quadric}
is
\[
z = \tfrac32u^2 - \tfrac12v^2 :
\]
a hyperbolic paraboloid ($A$ has rank $2$ and $b$ has a
component along $\ker A = \R e_z$: no center).

\textbf{20.} A rigid motion transforms the equation's data by
$A \mapsto P^{\mathsf T}\!AP$ (same \omterm{def:b2:reduction:eigen}{eigenvalues}) and the
normal forms have no residual freedom except permuting
coordinates and multiplying the whole equation by a scalar
($> 0$ to preserve the writing): two central normal forms
coincide up to isometry iff the coefficient lists agree up to
permutation and common positive factor --- for the ellipsoid,
iff the semiaxes $(a, b, c)$ agree. Affinely, one may also
scale each coordinate separately ($z_i \mapsto z_i\sqrt{\abs
{\lambda_i}}$), which erases the \omterm{def:b2:reduction:eigen}{eigenvalues} and leaves only
their signs: every ellipsoid becomes $u^2 + v^2 + w^2 = 1$,
the sphere. Sylvester's inertia theorem
(\cref{thm:b2:quadratic:sylvester}) guarantees the signature
survives any invertible linear change: the \omterm{def:b2:affine:subspace}{affine} types of
question 15 are genuinely distinct.

\textbf{21.} Parametrize the plane affinely: $M = P + su +
tv$. Then $q(P + su + tv)$ is a polynomial of degree $\leq 2$
in $(s, t)$ (expand the \omterm{def:b2:quadratic:def}{quadratic form} bilinearly), so the
section $\{q = 0\}$ is a conic of the plane, possibly
degenerate. For $z = xy$ and the plane $z = c$: $xy = c$, a
hyperbola for $c \neq 0$, and for $c = 0$ the two coordinate
lines --- the pair of rulings through the origin.

\textbf{22.} \emph{Ellipsoid}: bounded, contains no line.
\emph{Two-sheet hyperboloid} $\frac{z^2}{c^2} -
\frac{x^2}{a^2} - \frac{y^2}{b^2} = 1$: restrict to $p + tv$;
the $t^2$ coefficient $\frac{v_3^2}{c^2} - \frac{v_1^2}{a^2}
- \frac{v_2^2}{b^2}$ must vanish, so $v_3 \neq 0$ (else $v =
0$); the $t$ coefficient gives $\frac{p_3v_3}{c^2} =
\frac{p_1v_1}{a^2} + \frac{p_2v_2}{b^2}$, and Cauchy--Schwarz
yields
\[
\frac{p_3^2}{c^2}
= \frac{c^2}{v_3^2}\Bigl(\frac{p_1v_1}{a^2} +
\frac{p_2v_2}{b^2}\Bigr)^2
\leq \frac{c^2}{v_3^2}\Bigl(\frac{p_1^2}{a^2} +
\frac{p_2^2}{b^2}\Bigr)\frac{v_3^2}{c^2}
= \frac{p_1^2}{a^2} + \frac{p_2^2}{b^2},
\]
so the constant term is $\leq 0 \neq 1$: no line.
\emph{Elliptic paraboloid} $z = \frac{x^2}{a^2} +
\frac{y^2}{b^2}$: the $t^2$ coefficient forces $v_1 = v_2 =
0$, then the equation is linear nonconstant in $t$: no line.
\emph{Cone} $x^2 + y^2 = z^2$: a line on it satisfies $q(p) =
q(v) = B(p, v) = 0$ for the Lorentz form; equality in the
plane Cauchy--Schwarz $\abs{p_1v_1 + p_2v_2} = \abs{p_3v_3} =
\sqrt{(p_1^2 + p_2^2)(v_1^2 + v_2^2)}$ forces $(p_1, p_2)
\parallel (v_1, v_2)$ and then $p = kv$: all lines pass
through the vertex --- one family. \emph{Cylinders}: for the
elliptic and parabolic cylinders the $t^2$ coefficient forces
$v_1 = v_2 = 0$ (rulings only); for the hyperbolic cylinder
$\frac{x^2}{a^2} - \frac{y^2}{b^2} = 1$, $\frac{v_1}a = \pm
\frac{v_2}b$ with $v_2 \neq 0$ leads via the $t$ coefficient
to $\frac{p_1^2}{a^2} = \frac{p_2^2}{b^2}$, contradicting the
constant term $1$: again only the vertical rulings. So the
doubly ruled \omterm{pb:b2:surfaces:1}{quadric} surfaces are exactly the one-sheet
hyperboloid and the hyperbolic paraboloid.

\textbf{23.} Gauss: $x^2 - 2xy - 2zx = (x - y - z)^2 - y^2 -
z^2 - 2yz$, so
\[
q = (x - y - z)^2 - 4yz = (x - y - z)^2 + (y - z)^2 - (y +
z)^2 :
\]
three independent squares with signs $(+, +, -)$ --- signature
$(2, 1)$, matching question 17, with no \omterm{def:b2:reduction:eigen}{eigenvalue}
computation. Gauss loses the metric data: the new coordinates
are not orthonormal, so the \omterm{def:b2:reduction:eigen}{eigenvalues} (the shape of the
hyperboloid, its axes and their \omterm{def:b2:curves:length}{lengths}) are gone; only the
\omterm{def:b2:affine:subspace}{affine} type remains.

\textbf{24.} Let $p$, $n$, $z$ be the numbers of positive,
negative and zero \omterm{def:b2:reduction:eigen}{eigenvalues}, $p + n + z = 3$. Descartes'
rule bounds $p$ by the number $V$ of sign changes of
$\chi_A$, and $n$ by the number $V'$ of sign changes of
$\chi_A(-\lambda)$; moreover each pair of consecutive nonzero
coefficients produces a change in exactly one of the two
polynomials, so $V + V' \leq 3 - z$ (zero roots are visible
as vanishing trailing coefficients). Then $p + n = 3 - z \geq
V + V' \geq p + n$: equality, so $p = V$ exactly --- the
signs of the \omterm{def:b2:reduction:eigen}{eigenvalues} can be read off. For $\chi_A(\lambda)
= \lambda^3 - 3\lambda^2 - 9\lambda - 5$: signs $+,-,-,-$
give $V = 1$, and $\chi_A(-\lambda) = -\lambda^3 - 3\lambda^2
+ 9\lambda - 5$ has signs $-,-,+,-$: $V' = 2$. Signature $(1,
2)$ --- consistent with the exact factorization $\chi_A =
(\lambda - 5)(\lambda + 1)^2$ of question 6.

\textbf{25.} Algorithm: diagonalize the quadratic part
orthonormally (spectral theorem: this is the only
\omterm{def:b2:powerseries:analytic}{analytically} deep step, and it is what makes the
classification \emph{Euclidean}); solve $A\Omega = -b$ to
decide central versus parabolic types and to translate away
the linear part where possible (\omterm{def:b2:affine:subspace}{affine} geometry); read the
type from rank, signature and the constant (Sylvester
guarantees these are invariants); when only the \omterm{def:b2:affine:subspace}{affine} type
matters, Gauss's reduction replaces the spectral theorem at
the cost of the metric information. In $\R^2$ the same
machine classifies conics: ellipse, hyperbola, parabola, plus
pairs of lines, a line, a point, and empty sets. In $\R^n$
nothing changes but bookkeeping: the types are indexed by the
signature of $A$, the position of $b$ relative to
$\operatorname{im}A$, and one constant --- with the
$(n{+}2)$-dimensional bordered matrix of $(A, b, c)$
providing a \omterm{def:b2:metric:compact}{compact} invariant.
\end{solution}
