\section*{Chapter \ref{ch:b2:proba} --- Probability on Countable Spaces}

\begin{solution}{exo:b2:proba:1}
By symmetry: the drawing order induces a uniformly random relative
order on balls $1$ and $2$, so $\P(1 \text{ before } 2) =
\frac12$. Formally: exchanging the positions of balls $1$ and $2$
in a drawing sequence is a bijection of the (equiprobable)
outcomes that swaps the event with its complement. Among balls
$1, \dots, k$: the relative order of these $k$ balls is uniform
among the $k!$ orderings, and ball $1$ is first in $(k-1)!$ of
them: probability $\frac{(k-1)!}{k!} = \frac1k$.
\end{solution}

\begin{solution}{exo:b2:proba:2}
$\frac{1}{k(k+1)} = \frac1k - \frac1{k+1}$, so $\sum_{k\geq1} p_k$
telescopes to $1$: a probability measure. Even outcomes:
\[
\P(2\N^*) = \sum_{j=1}^{\infty}\frac{1}{2j(2j+1)}
= \sum_{j=1}^{\infty}\Bigl(\frac{1}{2j} - \frac{1}{2j+1}\Bigr)
= \frac12 - \frac13 + \frac14 - \frac15 + \cdots
\]
This is the alternating harmonic series with its first term
removed and signs flipped: since $\ln 2 = 1 - \frac12 + \frac13 -
\frac14 + \cdots$ (\cref{ch:b2:series}),
\[
\P(2\N^*) = -\bigl(\ln 2 - 1\bigr) = 1 - \ln 2 \approx 0.307 .
\]
\end{solution}

\begin{solution}{exo:b2:proba:3}
Let $S$ = sick, $+$ = positive test. Bayes
(\cref{thm:b2:proba:bayes}) with the partition $\{S, S^c\}$:
\[
\P(S \mid +)
= \frac{0.99 \times 10^{-4}}
       {0.99 \times 10^{-4} + 0.01 \times 0.9999}
= \frac{0.000099}{0.000099 + 0.009999}
\approx 0.0098 ,
\]
below $1\%$. Although the test is ``99\% accurate'', a positive
result leaves you about $99\%$ likely to be healthy: the false
positives among the vast healthy majority swamp the true
positives from the tiny sick minority. Screening tests for rare
conditions must always be read through this base-rate
computation.
\end{solution}

\begin{solution}{exo:b2:proba:4}
Pointwise on $\Omega$: $\omega \in \bigcup A_i$ iff some factor
$1 - \mathbf{1}_{A_i}(\omega)$ vanishes, so
\[
\mathbf{1}_{\bigcup A_i}
= 1 - \prod_{i=1}^n\bigl(1 - \mathbf{1}_{A_i}\bigr)
= \sum_{\emptyset \neq J \subseteq \{1,\dots,n\}}
(-1)^{\abs J + 1}\prod_{i \in J}\mathbf{1}_{A_i} ,
\]
expanding the product and moving the $1$ across. Now
$\prod_{i\in J}\mathbf{1}_{A_i} = \mathbf{1}_{\bigcap_{i \in J}
A_i}$, and summing against the weights $\P(\{\omega\})$ ---
legitimate: finitely many bounded terms, each family summable ---
turns each indicator into the probability of its event, giving
the formula.
\end{solution}

\begin{solution}{exo:b2:proba:5}
Let $A_i$ = ``letter $i$ is in the right envelope''. For $J$ of
size $k$, $\P\bigl(\bigcap_{i\in J}A_i\bigr) =
\frac{(n-k)!}{n!}$ (fix $k$ letters, permute the rest). By
inclusion--exclusion,
\[
\P\Bigl(\bigcup A_i\Bigr)
= \sum_{k=1}^n (-1)^{k+1}\binom nk \frac{(n-k)!}{n!}
= \sum_{k=1}^n \frac{(-1)^{k+1}}{k!} ,
\]
so
\[
\P(\text{no match})
= 1 - \P\Bigl(\bigcup A_i\Bigr)
= \sum_{k=0}^{n}\frac{(-1)^k}{k!}
\xrightarrow[n\to\infty]{} e^{-1} \approx 0.368 .
\]
Exactly one match: a permutation with exactly one fixed point is
determined by the choice of the fixed letter ($n$ ways) and a
\emph{derangement} (no-match arrangement) of the other $n - 1$;
writing $D_{n-1} = (n-1)!\sum_{k=0}^{n-1}\frac{(-1)^k}{k!}$ for
the number of derangements (the first part, scaled by
$(n-1)!$),
\[
\P(\text{exactly one match})
= \frac{n\,D_{n-1}}{n!}
= \frac{D_{n-1}}{(n-1)!}
= \sum_{k=0}^{n-1}\frac{(-1)^k}{k!}
\xrightarrow[n\to\infty]{} e^{-1} :
\]
in the limit, ``no match'' and ``exactly one match'' are equally
likely, each with probability $e^{-1}$.
\end{solution}

\begin{solution}{exo:b2:proba:6}
Condition on the start (chain rule /
\cref{thm:b2:proba:bayes}):
\begin{itemize}
\item first toss T (probability $1 - p$): the game restarts
afresh; lasting more than $n$ means lasting more than $n - 1$ from
there: contribution $(1-p)\,q_{n-1}$;
\item first tosses HT (probability $p(1-p)$): restart after two
tosses: contribution $p(1-p)\,q_{n-2}$;
\item first tosses HH: the game has ended (within $n$ tosses, $n
\geq 2$): contributes $0$.
\end{itemize}
Hence $q_n = (1-p)q_{n-1} + p(1-p)q_{n-2}$. The characteristic
equation $r^2 = (1-p)r + p(1-p)$ has roots
\[
r_\pm = \frac{(1-p) \pm \sqrt{(1-p)^2 + 4p(1-p)}}{2},
\]
with $\abs{r_\pm} < 1$: indeed the polynomial $\chi(r) = r^2 -
(1-p)r - p(1-p)$ satisfies $\chi(1) = 1 - (1-p) - p(1-p) = p^2 >
0$ and $\chi(-1) = 1 + (1-p) - p(1-p) > 0$, while $\chi(0) =
-p(1-p) < 0$: one root in $\intoo{-1}{0}$, one in $\intoo{0}{1}$.
So $q_n = \alpha r_+^n + \beta r_-^n \to 0$. The events ``game
lasts more than $n$'' decrease to ``game never ends''; monotone
continuity (\cref{thm:b2:proba:continuity}) gives
$\P(\text{never ends}) = \lim q_n = 0$: the game ends almost
surely.
\end{solution}

\begin{solution}{exo:b2:proba:7}
\emph{$\P(R_n) = 1/n$:} among the first $n$ draws, each of the $n$
relative positions of the last draw is equally likely (uniformity
of the relative order), and $R_n$ is the event that it is the
largest: probability $1/n$.

\emph{Infinitely many records:} $\sum_n \P(R_n) = \sum 1/n =
\infty$ and the $R_n$ are independent (admitted), so
Borel--Cantelli~2 (\cref{thm:b2:proba:borelcantelli}) gives
$\P(\limsup R_n) = 1$: records never stop, almost surely --- but
they thin out logarithmically.

\emph{Consecutive records:} by independence,
\[
\sum_n \P(R_n \cap R_{n+1})
= \sum_n \frac{1}{n(n+1)} < \infty ,
\]
so Borel--Cantelli~1 applies: almost surely, only finitely many
times is a record immediately followed by another record. The two
halves of the lemma work in tandem: infinitely many records, but
(a.s.) eventually never two in a row.
\end{solution}

\begin{solution}{exo:b2:proba:8}
\emph{First part:} $\sum \P(A_n) = \sum\frac{1}{n+1} = \infty$
with independence: Borel--Cantelli~2 gives $\P(\limsup A_n) = 1$.
Each individual $A_n$ is increasingly unlikely, yet almost every
$\omega$ belongs to infinitely many of them.

\emph{Counterexample without independence:} take $\Omega = \N^*$
with the weights $p_k = \frac{1}{k(k+1)}$ of
\cref{exo:b2:proba:2}, and $A_n = \{k \in \N^* : k \geq n\}$.
Then
\[
\P(A_n) = \sum_{k \geq n}\Bigl(\frac1k - \frac1{k+1}\Bigr)
= \frac1n ,
\qquad
\sum_n \P(A_n) = \infty ,
\]
but the $A_n$ are decreasing, so $\limsup_n A_n = \bigcap_n A_n =
\emptyset$: $\P(\limsup A_n) = 0$. Divergence of $\sum\P(A_n)$
alone guarantees nothing when the events pile up on a shrinking
part of the space --- independence is what forbids that
conspiracy.
\end{solution}

\begin{solution}{exo:b2:proba:9}
With $q = 1 - p$, the first head falls at rank $2j + 1$ with
probability $q^{2j}p$, so
\[
\P(\text{odd rank}) = \sum_{j\geq0}q^{2j}p
= \frac{p}{1 - q^2} = \frac{1}{1 + q} .
\]
For a fair coin: $\frac1{1 + 1/2} = \frac23$. (Sanity check:
odd ranks should be likelier, since rank $1$ comes first ---
and indeed $\frac1{1+q} > \frac12$ always.)
\end{solution}

\begin{solution}{exo:b2:proba:10}
The events $B_N = \bigcap_{n=1}^N A_n^c$ decrease to
$\bigcap_nA_n^c$, and by independence of the complements
$\P(B_N) = \prod_{n=1}^N(1 - p_n)$; monotone continuity
(\cref{thm:b2:proba:continuity}) gives the displayed limit.
Taking logarithms, $\prod(1 - p_n) > 0$ iff $\sum-\ln(1 -
p_n) < \infty$. If $\sum p_n < \infty$ then $p_n \to 0$ and
$-\ln(1 - p_n) \sim p_n$: the log-series converges. If
$\sum p_n = \infty$, then $-\ln(1 - p_n) \geq p_n$ forces
divergence, so the product is $0$. This matches
Borel--Cantelli~2: for $\sum p_n = \infty$, not only is
$\P(\text{no }A_n\text{ occurs}) = 0$, but almost surely
infinitely many $A_n$ occur.
\end{solution}

\begin{solution}{exo:b2:proba:11}
Say box $A$ is the one first discovered empty, with the other
box holding $k$. This means: among the first $2n - k$
reaches, exactly $n$ went to $A$ and $n - k$ to $B$ (in some
order), and reach number $2n - k + 1$ went to $A$ again,
finding it empty. The reaches are independent fair choices,
so this event has probability
$\binom{2n-k}{n}2^{-(2n-k)}\cdot\frac12$; doubling (the empty
box can be either one) gives
\[
\P(\text{other box has }k) = \binom{2n-k}{n}\,2^{-(2n-k)} .
\]
For $n = 1$: $k = 1$ gives $\binom11 2^{-1} = \frac12$ and $k
= 0$ gives $\binom21 2^{-2} = \frac12$: total $1$, as it must
be.
\end{solution}

\begin{solution}{exo:b2:proba:12}
(a) If $\P(\{n\}) = c$ for all $n$, $\sigma$-additivity
forces $1 = \sum_nc$: impossible, whether $c = 0$ (sum $0$)
or $c > 0$ (sum infinite). There is no uniform probability on
$\N$.

(b) If $A \cap B = \emptyset$ and $d(A)$, $d(B)$ exist, then
$\abs{(A \sqcup B)\cap\intint1n} = \abs{A\cap\intint1n} +
\abs{B\cap\intint1n}$, so $d(A \sqcup B) = d(A) + d(B)$:
finite additivity on such pairs. Each singleton has counting
function eventually constant, so density $0$, while
$d(\N^*) = 1$. Were $d$ $\sigma$-additive, $\N^* =
\bigsqcup_k\{k\}$ would give $1 = \sum_k 0 = 0$: density is
finitely additive but not $\sigma$-additive --- the axiom has
content.

(c) Let $A = \bigcup_{k\geq0}\intint{4^k}{2\cdot4^k - 1}$
(blocks from $4^k$ to $2\cdot4^k - 1$). At $n = 2\cdot4^K -
1$ the count is $\sum_{k\leq K}4^k \sim \frac43 4^K$, giving
ratio $\to \frac23$; at $n = 4^{K+1} - 1$ the count is
unchanged, giving ratio $\to \frac13$. The ratio oscillates
between limits $\frac13$ and $\frac23$: no density.
\end{solution}

\begin{solution}{pb:b2:proba:1}
\textbf{1.} A path of length $n$ is determined by the set of
its up-steps; ending at $k$ means $u$ up-steps and $n - u$
down-steps with $u - (n - u) = k$, i.e.\ $u = \frac{n+k}2$:
possible iff $n + k$ is even and $\abs k \leq n$, in
$\binom{n}{(n+k)/2}$ ways. Each specific path is one point of
the fair product measure on $n$ tosses: probability $2^{-n}$.
Hence $\P(S_n = k) = N_n(k)2^{-n}$.

\textbf{2.} $S_n$ has the parity of $n$, so $S_{2n+1} \neq
0$; and $u_n = N_{2n}(0)4^{-n} = \binom{2n}n4^{-n}$. Values:
$u_1 = \frac12$, $u_2 = \frac6{16} = \frac38$, $u_3 =
\frac{20}{64} = \frac5{16}$.

\textbf{3.} $\dfrac{u_n}{u_{n-1}} =
\dfrac{\binom{2n}n}{4\binom{2n-2}{n-1}} =
\dfrac{(2n)(2n-1)}{4n^2} = \dfrac{2n-1}{2n} < 1$: decreasing.
By \cref{ex:b2:comparison:centralbinomial}, $\binom{2n}n \sim
\frac{4^n}{\sqrt{\pi n}}$, so $u_n \sim \frac1{\sqrt{\pi n}}
\to 0$, and $\sum u_n$ diverges by comparison with $\sum
n^{-1/2}$.

\textbf{4.} Given a path from $1$ to $k$ touching $0$,
reflect its initial segment (up to the \emph{first} visit to
$0$) through the horizontal axis: the result is a path from
$-1$ to $k$, and the operation is an involution --- every
path from $-1$ to $k \geq 1$ must cross $0$, and reflecting
its initial segment back recovers the original. Hence the
touching paths number $N_{n-1}(k + 1)$ (from $-1$ to $k$ the
displacement is $k + 1$). A path from $0$ to $k$ staying $>
0$ after time $0$ starts with an up-step and then goes from
$1$ to $k$ in $n - 1$ steps without touching $0$: there are
$N_{n-1}(k-1) - N_{n-1}(k+1)$ of them.

\textbf{5.} With $m = \frac{n+k}2$, using
$\binom{n-1}{m-1} = \frac mn\binom nm$ and $\binom{n-1}{m} =
\frac{n-m}n\binom nm$:
\[
\frac{N_{n-1}(k-1) - N_{n-1}(k+1)}{N_n(k)}
= \frac{\binom{n-1}{m-1} - \binom{n-1}{m}}{\binom nm}
= \frac{m - (n - m)}{n} = \frac kn .
\]
For $n = 3$, $k = 1$: $N_3(1) = 3$ paths ($++-$, $+-+$,
$-++$), of which only $++-$ stays positive ($+-+$ returns to
$0$ at time $2$): one out of three, and $\frac kn = \frac13$.

\textbf{6.} By symmetry the probability is $2\P(S_i > 0\ \forall
i \leq 2n)$. Summing over the endpoint $2k$ and using
question 4 (with $n$ replaced by $2n$):
\[
\P(S_i > 0\ \forall i) = 2^{-2n}\sum_{k\geq1}
\bigl(N_{2n-1}(2k-1) - N_{2n-1}(2k+1)\bigr)
= 2^{-2n}\,N_{2n-1}(1),
\]
a telescoping sum. Now $N_{2n-1}(1) = \binom{2n-1}{n}$ and
$2\binom{2n-1}n = \binom{2n}n$ (Pascal), so the displayed
probability is $2\cdot2^{-2n}\binom{2n-1}n =
\binom{2n}n4^{-n} = u_n$.

\textbf{7.} The events $D_n = \{S_i \neq 0,\ i \leq 2n\}$
decrease, with intersection ``no return ever''; by monotone
continuity and question 6, $\P(\text{no return}) = \lim u_n =
0$: the walk returns almost surely. Moreover $f_n = \P(D_{n-1})
- \P(D_n) = u_{n-1} - u_n$, and by question 3
\[
u_{n-1} - u_n = u_n\Bigl(\frac{2n}{2n-1} - 1\Bigr) =
\frac{u_n}{2n-1};
\qquad
\sum_{n\geq1}f_n = u_0 - \lim u_n = 1 .
\]

\textbf{8.} $2n\,f_n = \frac{2n}{2n-1}u_n \geq u_n$, and
$\sum u_n = \infty$ (question 3): the series $\sum 2nf_n$
diverges. The first return is certain but has no finite mean
waiting time --- the walk is \emph{null recurrent}, in the
vocabulary that \cref{ch:b2:randomvar} will provide.

\textbf{9.} The event ``at least $k$ returns'' is the
disjoint countable union, over $0 < n_1 < \dots < n_k$, of
the events ``the first $k$ returns happen exactly at times
$2n_1, \dots, 2n_k$''. Such an event is the intersection of
$k$ events depending on the disjoint toss blocks
$\intint1{2n_1}$, $\intint{2n_1+1}{2n_2}$, \dots, each block
requiring a fresh walk to make its first return after
exactly the allotted number of steps; by independence of the
blocks its probability is $f_{n_1}f_{n_2-n_1}\cdots
f_{n_k-n_{k-1}}$. Summing by packets (\cref{ch:b2:series},
all terms nonnegative):
\[
\P(\text{at least }k\text{ returns})
= \Bigl(\sum_{n\geq1}f_n\Bigr)^{\!k} = 1^k = 1 .
\]
The events decrease in $k$, so by monotone continuity
$\P(\text{infinitely many returns}) = 1$: recurrence.

\textbf{10.} By question 9 the walk makes infinitely many
excursions away from $0$. The first step of each excursion is
a fresh coin, independent of everything before: the
probability that the first $m$ excursions all start downward
is $2^{-m}$. To reach $1$ the walk only needs one upward
excursion start (from $<0$ it must pass through $0$ before
reaching $1$, steps being $\pm1$), so $\P(\text{never hit }1)
\leq 2^{-m}$ for every $m$: the walk hits $1$ almost surely.
Decomposing over the (almost surely finite) hitting time, the
walk restarted there is a fresh walk started at $1$: by
induction it hits every $k \geq 1$ almost surely, and by
symmetry every $k \leq -1$. Finally, restarting at the first
visit to $k$, question 9 applies to the fresh walk: every
site is visited infinitely often, almost surely.

\textbf{11.} The events $A_n = \{S_{2n} = 0\}$ are far from
independent (being at $0$ at time $2n$ makes being at $0$ at
time $2n + 2$ far likelier than $u_{n+1}$), so
Borel--Cantelli~2 is unavailable, and indeed the whole work
of Part II was to replace it. The other direction needs no
independence: \emph{if} $\sum\P(A_n)$ converges,
Borel--Cantelli~1 yields finitely many returns almost surely.
That implication is the engine of every transience proof
below.

\textbf{12.} A return at time $2n$ requires $n$ up- and $n$
down-steps: $\P(S_{2n} = 0) = \binom{2n}np^nq^n =
u_n(4pq)^n$, and $4pq = 1 - (p - q)^2 < 1$ for $p \neq
\frac12$. Since $u_n \leq 1$, the series $\sum\P(S_{2n} = 0)$
is dominated by the geometric $\sum(4pq)^n$: convergent. By
Borel--Cantelli~1, $\P(S_{2n} = 0 \text{ infinitely often}) =
0$: finitely many returns, almost surely.

\textbf{13.} For $n + k$ even, $\P(S_n = k) =
\binom{n}{\frac{n+k}2}p^{\frac{n+k}2}q^{\frac{n-k}2}$; the
binomial coefficient is at most the central one, and
$p^{\frac{n+k}2}q^{\frac{n-k}2} = (pq)^{n/2}(p/q)^{k/2}$,
giving the stated bound $\leq
2^n(pq)^{n/2}(p/q)^{k/2} = (4pq)^{n/2}(p/q)^{k/2}$, summable
in $n$ since $\sqrt{4pq} < 1$. Borel--Cantelli~1: site $k$ is
visited finitely often almost surely; the union over $k \in
\Z$ of the exceptional null events is still null (countable
subadditivity). Almost surely every site is visited finitely
often, so the integer sequence $(S_n)$ leaves every bounded
window for good: $\abs{S_n} \to \infty$.

\textbf{14.} $\P(S_n \neq 0,\ 1 \leq n \leq 200) = u_{100} =
\binom{200}{100}4^{-100} \approx \frac1{\sqrt{100\pi}}
\approx 0.056$: more than a one-in-twenty chance that $200$
fair tosses never tie. The decay $1/\sqrt{\pi n}$ is
excruciatingly slow: certainty of a tie (question 7) is
compatible with very long tie-free stretches --- a first
taste of the arcsine phenomena of Part IV.

\textbf{15.} Condition on the first step. If $X_1 = +1$ then
$T_1 = 1$, and $f_1 = \frac12$ matches. If $X_1 = -1$, the
walk must climb from $-1$ to $1$; by the block decomposition,
returning to $0$ for the first time at time $2n$ splits as:
one step down, then a fresh walk started at $-1$ first
reaching $0$ --- equivalently a fresh walk first reaching
$+1$ --- in $2n - 1$ steps, or the symmetric event upward.
Both signs contribute equally:
\[
f_n = 2\cdot\tfrac12\,\P(T_1 = 2n - 1) = \P(T_1 = 2n-1) .
\]
Hence $\P(T_1 < \infty) = \sum f_n = 1$, while $\sum_n(2n -
1)f_n = \sum_n u_n = \infty$ by question 7: the walk reaches
$1$ almost surely, in infinite mean time.

\textbf{16.} Partition $\{M_n \geq k\}$ by the terminal value
$S_n = m$. For $m \geq k$ the condition $M_n \geq k$ is
automatic. For $m < k$, reflect the path after its
\emph{first} visit to level $k$: this is a bijection between
$\{M_n \geq k, S_n = m\}$ and $\{S_n = 2k - m\}$ (every path
ending at $2k - m > k$ visits $k$; reflecting back is the
inverse). Hence
\[
\P(M_n \geq k)
= \sum_{m > k}\P(S_n = m) + \P(S_n = k)
+ \sum_{m < k}\P(S_n = 2k - m)
= 2\P(S_n > k) + \P(S_n = k).
\]

\textbf{17.} At even time $2n$ with $k = 1$: $\P(S_{2n} = 1)
= 0$ and $\P(S_{2n} > 1) = \P(S_{2n} \geq 2)$, so
\[
\P(M_{2n} \geq 1) = 2\P(S_{2n} \geq 2)
= \P(S_{2n} \geq 2) + \P(S_{2n} \leq -2)
= 1 - u_n .
\]
Thus $\P(S_i \leq 0\ \forall i \leq 2n) = u_n$: the walk
never leads in the first $2n$ steps exactly as often as it
never ties (question 6) --- two quite different events,
carried by the same $u_n$.

\textbf{18.} $\{L_{2n} = 2k\} = \{S_{2k} = 0\} \cap
\{\text{the walk of the tosses } 2k+1, \dots, 2n \text{ has
no zero}\}$. The two events depend on disjoint toss blocks,
so they are independent; the first has probability $u_k$, the
second $u_{n-k}$ by question 6 applied to the fresh
$(2n-2k)$-step walk. Hence $\P(L_{2n} = 2k) = u_ku_{n-k}$.
Since $L_{2n}$ takes exactly the values $0, 2, \dots, 2n$,
these probabilities sum to $1$:
$\sum_{k=0}^nu_ku_{n-k} = 1$, a binomial identity delivered
by a probabilistic partition.

\textbf{19.} Symmetry is immediate: $u_ku_{n-k} =
u_{n-k}u_k$. As $u_j$ decreases in $j$, the product
$u_ku_{n-k}$ is smallest for central $k$ and largest at the
extremes $k \in \{0, n\}$, where it equals $u_n$;
quantitatively $u_ku_{n-k} \approx \frac1{\pi\sqrt{k(n-k)}}$
in the bulk, against $u_n \approx \frac1{\sqrt{\pi n}}$ at
the edges. For $n = 5$: $\P(L_{10} = 0) = \P(L_{10} = 10) =
u_5 = \frac{63}{256} \approx 0.246$, while $\P(L_{10} = 4) =
u_2u_3 = \frac38\cdot\frac5{16} = \frac{15}{128} \approx
0.117$. In a long fair game the last equalization is most
likely near the very beginning or the very end: one player
typically leads for enormous stretches, with no bias in the
coin.

\textbf{20.} The picture: at time $n$ the walk lives at scale
$\sqrt n$ (the binomial spread of question 3 --- $u_n \sim
1/\sqrt{\pi n}$ is the height of the central peak); it
returns to $0$ infinitely often with probability $1$ (Part
II), yet the waiting time between returns has divergent mean
(question 8), which is why single excursions can occupy a
positive fraction of any horizon; correspondingly the last
tie of a $2n$-step game is spread out with extreme values
most likely (questions 18--19), and never-leading has the
same slowly decaying probability $u_n$ as never-tying
(question 17). Certainty in the limit, persistence at every
finite horizon: that is the fair walk.

\textbf{21.} Partition $\{S_{2n} = 0\}$ ($n \geq 1$) by the
first return time $2k$, $1 \leq k \leq n$: the first block of
$2k$ tosses realizes a first return, the remaining $2n - 2k$
tosses realize a return of a fresh walk, and the blocks are
independent: $u_n = \sum_{k=1}^nf_ku_{n-k}$. Both series
$U(x) = \sum u_nx^n$, $F(x) = \sum f_nx^n$ have radius $\geq
1$ (coefficients in $\intcc01$), and the Cauchy product
(\cref{ch:b2:powerseries}) gives, for $0 \leq x < 1$,
\[
U(x) - 1 = \sum_{n\geq1}\Bigl(\sum_{k=1}^n
f_ku_{n-k}\Bigr)x^n = F(x)\,U(x),
\qquad\text{i.e.}\qquad
U(x)\bigl(1 - F(x)\bigr) = 1 .
\]

\textbf{22.} As $x \uparrow 1$, $U(x)$ and $F(x)$ increase
(nonnegative coefficients); every partial sum $\sum_{n\leq
N}u_n$ is a limit of $\sum_{n\leq N}u_nx^n \leq U(x)$, so
$U(x) \uparrow \sum u_n \in \intoc0{+\infty}$, and likewise
$F(x) \uparrow f = \sum f_n$. If $\sum u_n = \infty$: $1 -
F(x) = 1/U(x) \to 0$, so $f = 1$. If $\sum u_n = S < \infty$:
$1 - f = 1/S > 0$, so $f < 1$. Checks: fair walk, $\sum u_n =
\infty$ and $f = 1$ (questions 3, 7); biased walk, $\sum
u_n(4pq)^n < \infty$ and correspondingly $f = 1 -
1/\sum_{n\geq0}u_n(4pq)^n < 1$, consistent with the
almost-sure finiteness of the number of returns (question
12).

\textbf{23.} For the four steps $(\pm1, 0), (0, \pm1)$ of the
$\Z^2$ walk, the increments of $U = X + Y$ and $V = X - Y$
are: $(+,+)$ for $(1,0)$, $(+,-)$ for $(0,1)$, $(-,+)$ for
$(0,-1)$, $(-,-)$ for $(-1,0)$ --- each pair of signs with
probability $\frac14 = \frac12\cdot\frac12$: the two
coordinate walks $(U_n)$ and $(V_n)$ are independent fair
walks on $\Z$. Since $S^{(2)}_{2n} = (0,0)$ iff $U_{2n} = 0$
and $V_{2n} = 0$,
\[
\P\bigl(S^{(2)}_{2n} = (0,0)\bigr) = u_n^2 \sim \frac1{\pi
n}, \qquad \sum_nu_n^2 = \infty .
\]
The renewal identity of question 21 and the dichotomy of
question 22 used nothing one-dimensional (only the
decomposition over the first return and disjoint-block
independence), so $\sum u_n^{(2)} = \infty$ gives $f^{(2)} =
1$, and the argument of question 9 upgrades it: the walk on
$\Z^2$ returns to the origin infinitely often almost surely.

\textbf{24.} With the admitted bound $\P(S^{(3)}_{2n} = 0)
\leq Cn^{-3/2}$, the series converges, and Borel--Cantelli~1
gives finitely many returns almost surely: the walk on
$\Z^3$ is transient (and the same bound with exponent $-d/2$
handles every $d \geq 3$). Altogether: \emph{P\'olya's
theorem} --- the simple random walk is recurrent on $\Z$ and
$\Z^2$, transient on $\Z^d$ for $d \geq 3$. A drunk man finds
his way home; a drunk bird may not.

\textbf{25.} Path counting and reflection produced the exact
laws ($u_n$, the ballot theorem, $f_n$, the maximum, the last
zero); monotone continuity converted every limiting
statement (``returns at least once'', ``infinitely often'')
into a limit of finite-horizon probabilities;
disjoint-block independence powered the renewal
decompositions (questions 9, 18, 21) --- it is the countable
skeleton of the Markov property; Borel--Cantelli~1 was the
transience weapon (questions 12--13, 24), needing no
independence; the renewal identity organized everything into
the dichotomy $\sum u_n = \infty \iff$ recurrence. The single
analytic input is the local estimate $u_n \sim 1/\sqrt{\pi
n}$: its square $1/(\pi n)$ still diverges (dimension $2$,
recurrent), while $n^{-3/2}$ converges (dimension $3$,
transient) --- P\'olya's theorem is, in the end, a statement
about the divergence of $\sum n^{-d/2}$.
\end{solution}
