\section*{Chapter \ref{ch:b2:randomvar} --- Discrete Random Variables}

\begin{solution}{exo:b2:randomvar:1}
\emph{Binomial:} $X = \sum_{i=1}^n X_i$ with \omterm{def:b2:proba:independence}{independent} Bernoulli
$X_i$; $V(X_i) = \E(X_i^2) - \E(X_i)^2 = p - p^2$, and \omterm{def:b2:randomvar:variance}{variances}
of \omterm{def:b2:proba:independence}{independent} variables add
(\cref{thm:b2:randomvar:variancerules}):
\[
\E(X) = np, \qquad V(X) = np(1-p) .
\]
\emph{Poisson:} $\E\bigl(X(X-1)\bigr) =
\sum_{k\geq2}k(k-1)e^{-\lambda}\frac{\lambda^k}{k!} = \lambda^2
e^{-\lambda}\sum_{j\geq0}\frac{\lambda^j}{j!} = \lambda^2$, so
\[
V(X) = \E(X^2) - \E(X)^2
= \lambda^2 + \lambda - \lambda^2 = \lambda .
\]
\emph{Geometric} ($q = 1 - p$): differentiating
$\sum_{k\geq0}q^k = \frac{1}{1-q}$ twice inside the disk
(\cref{ch:b2:powerseries}), $\sum_{k\geq2}k(k-1)q^{k-2} =
\frac{2}{(1-q)^3}$, so
\[
\E\bigl(X(X-1)\bigr) = pq\sum_{k\geq2}k(k-1)q^{k-2}
= \frac{2q}{p^2},
\qquad
V(X) = \frac{2q}{p^2} + \frac1p - \frac{1}{p^2}
= \frac{q}{p^2} = \frac{1-p}{p^2} .
\]
\end{solution}

\begin{solution}{exo:b2:randomvar:2}
\emph{Sum:} for $n \in \N$, by disjointness and \omterm{def:b2:proba:independence}{independence},
\[
\P(X + Y = n)
= \sum_{k=0}^n \P(X = k)\P(Y = n - k)
= e^{-(\lambda + \mu)}\frac{1}{n!}
\sum_{k=0}^n \binom nk \lambda^k\mu^{n-k}
= e^{-(\lambda+\mu)}\frac{(\lambda + \mu)^n}{n!}
\]
by the binomial theorem: $X + Y \sim \mathcal{P}(\lambda + \mu)$.
\emph{Conditional \omterm{def:b2:randomvar:law}{law}:} for $0 \leq k \leq n$,
\[
\P(X = k \mid X + Y = n)
= \frac{\P(X = k)\P(Y = n - k)}{\P(X + Y = n)}
= \binom nk
\Bigl(\frac{\lambda}{\lambda+\mu}\Bigr)^{k}
\Bigl(\frac{\mu}{\lambda+\mu}\Bigr)^{n-k} ,
\]
the binomial \omterm{def:b2:randomvar:law}{law} $\mathcal{B}\bigl(n,
\frac{\lambda}{\lambda+\mu}\bigr)$: given the total count, each
\omterm{def:b2:proba:space}{event} independently ``chooses'' the first source with probability
proportional to its rate.
\end{solution}

\begin{solution}{exo:b2:randomvar:3}
When $k$ toys are still missing, each new box brings a new toy
with probability $\frac kn$, independently of the past: the
waiting time $W_k$ for the next new toy is geometric
$\mathcal{G}\bigl(\frac kn\bigr)$, with $\E(W_k) = \frac nk$, and
$T_n = W_n + W_{n-1} + \dots + W_1$ (first box always gives a new
toy: $W_n = 1$, consistent with $\E = n/n$). By linearity,
\[
\E(T_n) = \sum_{k=1}^n \frac nk = n\sum_{k=1}^n\frac1k
\sim n\ln n ,
\]
using $\sum_{k\leq n}\frac1k = \ln n + \gamma + o(1)$
(\cref{ch:b2:comparison}). Collecting the last few toys is what
costs: half the boxes go to the final handful.
\end{solution}

\begin{solution}{exo:b2:randomvar:4}
\omterm{def:b2:funcseq:def}{Pointwise}, $X(\omega) = \#\{n \geq 1 : X(\omega) \geq n\} =
\sum_{n\geq1}\mathbf{1}_{X \geq n}(\omega)$. The double family
$\bigl(\mathbf{1}_{X \geq n}(\omega)\,\P(\{\omega\})\bigr)_{n,
\omega}$ is nonnegative, so Fubini for families
(\cref{ch:b2:series}) applies unconditionally: summing first in
$n$ gives $\E(X)$, summing first in $\omega$ gives $\sum_n \P(X
\geq n)$; the two are simultaneously finite and equal. For $X \sim
\mathcal{G}(p)$: $\P(X \geq n) = q^{n-1}$ ($q = 1-p$), so $\E(X) =
\sum_{n\geq1}q^{n-1} = \frac{1}{1 - q} = \frac1p$.
\end{solution}

\begin{solution}{exo:b2:randomvar:5}
\emph{Symmetry:} the $i$-th ball drawn is a uniformly random ball
of the urn (any of the $N$ balls is equally likely to land in
position $i$ of the drawing order), so $\P(Y_i = 1) = \frac MN =
p$ and $\E(X) = np$ by linearity --- no \omterm{def:b2:proba:independence}{independence} needed.

\emph{\omterm{def:b2:randomvar:variance}{Covariance}:} for $i \neq j$, $\E(Y_iY_j) = \P(\text{draws }
i, j \text{ both white}) = \frac{M(M-1)}{N(N-1)}$ (ordered pairs
of distinct positions get an ordered pair of distinct balls,
uniformly). Hence
\[
\operatorname{Cov}(Y_i, Y_j)
= \frac{M(M-1)}{N(N-1)} - \frac{M^2}{N^2}
= \frac{M(N - M)}{N^2}\cdot\frac{-1}{N-1}
= -\frac{p(1-p)}{N-1} < 0 :
\]
drawing a white ball makes whites scarcer for the other draws.
By \cref{thm:b2:randomvar:variancerules},
\[
V(X) = np(1-p) + n(n-1)\Bigl(-\frac{p(1-p)}{N-1}\Bigr)
= np(1-p)\,\frac{N - n}{N - 1} \leq np(1-p) :
\]
sampling without replacement has the same mean but \emph{smaller}
\omterm{def:b2:randomvar:variance}{variance} than with replacement (equality only for $n = 1$), the
negative correlations acting as a stabilizer. For $n = N$ the
\omterm{def:b2:randomvar:variance}{variance} vanishes: the count is then deterministic.
\end{solution}

\begin{solution}{exo:b2:randomvar:6}
Expanding around $m = \E(X)$:
\[
\E\bigl((X - c)^2\bigr)
= \E\bigl((X - m)^2\bigr) + 2(m - c)\,\E(X - m) + (m - c)^2
= V(X) + (m - c)^2 ,
\]
minimal exactly at $c = m$ with value $V(X)$ --- \omterm{def:b2:randomvar:expectation}{expectation} is
the best constant predictor in mean square.

If $\P(X = m) = 1$ then $(X - m)^2$ vanishes with probability
$1$, so $V(X) = 0$ (the defining family has null terms except on a
null set). Conversely, if $V(X) = 0$, Chebyshev
(\cref{thm:b2:randomvar:markov}) gives $\P\bigl(\abs{X - m} \geq
\frac1n\bigr) \leq n^2\,V(X) = 0$ for every $n$; the \omterm{def:b2:proba:space}{events}
$\bigl\{\abs{X - m} \geq \frac1n\bigr\}$ increase to $\{X \neq
m\}$, so monotone \omterm{def:b2:metric:continuity}{continuity} (\cref{thm:b2:proba:continuity})
yields $\P(X \neq m) = 0$.
\end{solution}

\begin{solution}{exo:b2:randomvar:7}
$\E(S_n) = \frac n2$ and $V(S_n) = \frac n4$.
\emph{Markov:} $\P\bigl(S_n \geq \frac{3n}4\bigr) \leq
\frac{n/2}{3n/4} = \frac23$ --- a constant bound, useless for
large $n$.
\emph{Chebyshev:} the \omterm{def:b2:proba:space}{event} implies $\abs{S_n - \frac n2} \geq
\frac n4$, so the probability is $\leq \frac{n/4}{(n/4)^2} =
\frac4n$ --- decays, but only polynomially.
\emph{Chernoff:} by \omterm{def:b2:proba:independence}{independence}, $\E(e^{tS_n}) =
\prod_{i=1}^n\E(e^{tX_i}) = \bigl(\frac{1 + e^t}{2}\bigr)^n$, and
Markov applied to $e^{tS_n} \geq e^{3nt/4}$ gives, for every $t >
0$,
\[
\P\Bigl(S_n \geq \frac{3n}4\Bigr)
\leq \Bigl(\frac{1 + e^t}{2}\Bigr)^n e^{-3nt/4}
= \exp\Bigl(n\bigl(\ln\tfrac{1 + e^t}{2} - \tfrac{3t}4\bigr)\Bigr).
\]
Minimize the exponent: $\frac{\dd}{\dd t}\ln\frac{1+e^t}{2} =
\frac{e^t}{1 + e^t} = \frac34$ at $e^t = 3$, i.e.\ $t = \ln 3$,
giving
\[
\P\Bigl(S_n \geq \frac{3n}4\Bigr)
\leq \Bigl(\frac{4}{2}\Bigr)^n 3^{-3n/4}
= \bigl(2 \cdot 3^{-3/4}\bigr)^n \approx (0.877)^n ,
\]
exponentially small. The hierarchy Markov $\to$ Chebyshev $\to$
Chernoff is the standard ladder: each rung applies Markov to a
faster-growing function of the variable.
\end{solution}

\begin{solution}{exo:b2:randomvar:8}
By the transfer theorem (\cref{thm:b2:randomvar:transfer})
applied to $f\bigl(\frac{S_n}{n}\bigr)$ with $S_n \sim
\mathcal{B}(n, x)$:
\[
\E\Bigl[f\Bigl(\frac{S_n}{n}\Bigr)\Bigr]
= \sum_{k=0}^n f\Bigl(\frac kn\Bigr)\binom nk x^k(1-x)^{n-k}
= B_nf(x) .
\]
Fix $\delta > 0$ and split $\abs{f(S_n/n) - f(x)}$ on the \omterm{def:b2:proba:space}{event} $D
= \bigl\{\abs{\frac{S_n}{n} - x} \geq \delta\bigr\}$:
off $D$, the difference is at most the modulus of \omterm{def:b2:metric:continuity}{continuity}
$\omega_f(\delta) = \sup_{\abs{s - t}\leq\delta}\abs{f(s) -
f(t)}$; on $D$, at most $2\norm f_\infty$. Taking \omterm{def:b2:randomvar:expectation}{expectations} and
using Chebyshev with $V\bigl(\frac{S_n}{n}\bigr) =
\frac{x(1-x)}{n} \leq \frac{1}{4n}$:
\[
\abs{B_nf(x) - f(x)}
\leq \E\,\abs{f(S_n/n) - f(x)}
\leq \omega_f(\delta)
+ 2\norm f_\infty\,\P(D)
\leq \omega_f(\delta) + \frac{2\norm f_\infty}{4n\delta^2} .
\]
Uniform \omterm{def:b2:metric:continuity}{continuity} of $f$ on $[0, 1]$ makes $\omega_f(\delta) \to
0$: choose $\delta$ then $n$, and $B_nf \to f$ \omterm{def:b2:funcseq:def}{uniformly} --- the
Weierstrass approximation theorem of \cref{ch:b2:funcseq}, whose
``counting lemma'' is now recognizable as Chebyshev's inequality
for the binomial \omterm{def:b2:randomvar:law}{law}.
\end{solution}

\begin{solution}{exo:b2:randomvar:9}
Expand $S_n^4 = \sum_{i,j,k,l}X_iX_jX_kX_l$ and take
\omterm{def:b2:randomvar:expectation}{expectations}. By \omterm{def:b2:proba:independence}{independence} and centering, any term containing
an index appearing exactly once vanishes ($\E(X_i) = 0$ factors
out). Surviving terms: the $n$ diagonal terms $\E(X_i^4)$, and the
terms pairing two pairs of equal indices, $\E(X_i^2X_j^2) =
\E(X_1^2)^2$ for $i \neq j$, appearing $3n(n-1)$ times: choose
the unordered pair of values ($\binom n2$ ways), then the
$\frac{4!}{2!\,2!} = 6$ ways to place them in the four slots ---
$6\binom n2 = 3n(n-1)$. Hence, with $\E(X_1^2)^2
\leq \E(X_1^4)$ (Jensen or Cauchy--Schwarz),
\[
\E(S_n^4) = n\,\E(X_1^4) + 3n(n-1)\,\E(X_1^2)^2
\leq C n^2,
\qquad C = 4\,\E(X_1^4) .
\]
Markov at order 4:
\[
\P\Bigl(\Bigl|\frac{S_n}{n}\Bigr| \geq \varepsilon\Bigr)
= \P\bigl(S_n^4 \geq n^4\varepsilon^4\bigr)
\leq \frac{Cn^2}{n^4\varepsilon^4}
= \frac{C}{n^2\varepsilon^4} ,
\]
a \omterm{def:b2:series:summable}{summable} series. By Borel--Cantelli~1
(\cref{thm:b2:proba:borelcantelli}), for each $j$ the \omterm{def:b2:proba:space}{event}
$B_j = \limsup_n\bigl\{\abs{S_n/n} \geq \frac1j\bigr\}$ has
probability $0$, so $\P\bigl(\bigcup_j B_j\bigr) = 0$ by
\omterm{def:b2:structures:countable}{countable} subadditivity. On the complement --- of probability $1$
--- for every $j$ there is $N$ with $\abs{S_n/n} < \frac1j$ for
all $n \geq N$: precisely $S_n/n \to 0$. The strong law of large
numbers holds under a fourth moment; removing that hypothesis
(Kolmogorov's theorem) is Year 3 work.
\end{solution}

\begin{solution}{exo:b2:randomvar:10}
$\P(M \leq k) = \bigl(\frac k6\bigr)^2$ (both dice at most
$k$, independently), so $\P(M \geq k) = 1 -
\bigl(\frac{k-1}6\bigr)^2$ and
\[
\E(M) = \sum_{k=1}^6\P(M \geq k)
= 6 - \frac{0 + 1 + 4 + 9 + 16 + 25}{36}
= 6 - \frac{55}{36} = \frac{161}{36} \approx 4.47 ,
\]
comfortably above the mean $3.5$ of a single die, as a
maximum should be.
\end{solution}

\begin{solution}{exo:b2:randomvar:11}
With $I_i = \mathbf 1_{\sigma(i) = i}$: $\P(\sigma(i) = i) =
\frac{(n-1)!}{n!} = \frac1n$, so $\E(F_n) = n\cdot\frac1n =
1$. For $i \neq j$: $\P(\sigma(i) = i, \sigma(j) = j) =
\frac{(n-2)!}{n!} = \frac1{n(n-1)}$, hence
\[
\operatorname{Cov}(I_i, I_j) = \frac1{n(n-1)} - \frac1{n^2}
= \frac{1}{n^2(n-1)} .
\]
By the \omterm{def:b2:randomvar:variance}{variance} toolkit
(\cref{thm:b2:randomvar:variancerules}),
\[
V(F_n) = n\cdot\frac1n\Bigl(1 - \frac1n\Bigr)
+ n(n-1)\cdot\frac1{n^2(n-1)}
= 1 - \frac1n + \frac1n = 1 .
\]
Mean $1$, \omterm{def:b2:randomvar:variance}{variance} $1$, independent of $n$ --- consistent
with the Poisson limit of the matching problem
(\cref{exo:b2:proba:5}).
\end{solution}

\begin{solution}{exo:b2:randomvar:12}
$T_n = \sum_{k=1}^nG_k$ where $G_k \sim \mathcal G(k/n)$ is
the time to see a new toy when $k$ are missing, the stages
being \omterm{def:b2:proba:independence}{independent}. Hence
\[
V(T_n) = \sum_{k=1}^n\frac{1 - k/n}{(k/n)^2}
\leq \sum_{k=1}^n\frac{n^2}{k^2}
\leq \frac{\pi^2}6\,n^2 ,
\]
by \cref{ex:b2:fourier:basel}. With $\E(T_n) = nH_n$, $H_n =
\sum_1^n\frac1k$ (\cref{exo:b2:randomvar:3}), Chebyshev gives,
for $\varepsilon > 0$,
\[
\P\bigl(\abs{T_n - nH_n} \geq \varepsilon\,n\ln n\bigr)
\leq \frac{\pi^2n^2/6}{\varepsilon^2n^2\ln^2 n}
= \frac{\pi^2}{6\,\varepsilon^2\ln^2n}
\xrightarrow[n\to\infty]{} 0 .
\]
Since $H_n \sim \ln n$, dividing by $n\ln n$ shows
$T_n/(n\ln n) \to 1$ in probability: the fluctuations of
$T_n$ are of order $n$, negligible against the mean $n\ln n$.
\end{solution}

\begin{solution}{pb:b2:randomvar:1}
\textbf{1.} $\E(S_n) = \frac n2$ and Markov
(\cref{thm:b2:randomvar:markov}) give $\P(S_n \geq an) \leq
\frac{n/2}{an} = \frac1{2a}$. Markov only knows the mean: it
cannot distinguish a variable concentrated at $n/2$ from one
spread between $0$ and $n$, so it prices the tail as if all
the mass could sit there.

\textbf{2.} The fair binomial is \omterm{def:b2:quadratic:adjoint}{symmetric} about $n/2$
($S_n$ and $n - S_n$ have the same \omterm{def:b2:randomvar:law}{law}), so with $x = n(a -
\frac12) > 0$ the two \omterm{def:b2:proba:space}{events} $\{S_n - \frac n2 \geq x\}$ and
$\{S_n - \frac n2 \leq -x\}$ are disjoint and equiprobable:
$\P(S_n \geq an) = \frac12\P(\abs{S_n - \frac n2} \geq x)$.
Chebyshev with $V(S_n) = \frac n4$:
\[
\P(S_n \geq an)
\leq \frac12\cdot\frac{n/4}{n^2(a - 1/2)^2}
= \frac1{8n(a - 1/2)^2},
\]
which is $\frac2n$ at $a = \frac34$.

\textbf{3.} By \omterm{def:b2:proba:independence}{independence} and the product theorem,
$\E(\eu^{tS_n}) = \bigl(\E \eu^{tX_1}\bigr)^n = \bigl(\frac{1
+ \eu^t}2\bigr)^n$. Markov applied to $\eu^{tS_n}$:
\[
\P(S_n \geq an) \leq \eu^{-tan}\Bigl(\frac{1 +
\eu^t}2\Bigr)^{\!n} = \exp\Bigl(n\bigl(\ln\tfrac{1 +
\eu^t}2 - ta\bigr)\Bigr).
\]
The exponent's derivative in $t$ is $\frac{\eu^t}{1 + \eu^t}
- a$, vanishing at $\eu^t = \frac a{1-a}$, i.e.\ $t^* =
\ln\frac a{1-a} > 0$; there $\frac{1 + \eu^{t^*}}2 =
\frac1{2(1-a)}$ and the exponent equals
\[
n\Bigl(-\ln 2 - \ln(1-a) - a\ln\frac a{1-a}\Bigr)
= -n\bigl(\ln2 + a\ln a + (1-a)\ln(1-a)\bigr) = -n\,I(a),
\]
with $I(\frac12) = 0$ and $I'(a) = \ln\frac a{1-a} > 0$ on
$\intoo{\frac12}1$: $I(a) > 0$. At $a = \frac34$:
$\eu^{-I(3/4)} = \frac12(\tfrac34)^{-3/4}(\tfrac14)^{-1/4} =
2\cdot3^{-3/4}$, the bound of \cref{exo:b2:randomvar:7}.

\textbf{4.} The $n + 1$ numbers $\binom nja^j(1-a)^{n-j}$ sum
to $1$, and the largest is the one at $j = k = an$ (the mode
of $\mathcal B(n, a)$ is $\floor{(n+1)a} = k$ here). A
maximum of $n + 1$ numbers summing to $1$ is at least
$\frac1{n+1}$:
\[
\binom nk a^k(1-a)^{n-k} \geq \frac1{n+1}
\quad\Longrightarrow\quad
\binom nk \geq \frac{a^{-an}(1-a)^{-n(1-a)}}{n+1}
= \frac{\eu^{nH(a)}}{n+1}.
\]
Hence $\P(S_n \geq an) \geq \binom{n}{an}2^{-n} \geq
\eu^{n(H(a) - \ln2)}/(n+1) = \eu^{-nI(a)}/(n+1)$: up to the
polynomial factor $n + 1$, Chernoff's exponent is the truth.

\textbf{5.} $n = 100$, $a = \frac34$: Markov $\frac23$;
Chebyshev $\frac2{100} = 0.02$; Chernoff
$(2\cdot3^{-3/4})^{100} = \eu^{-100\,I(3/4)} \approx
2.1\cdot10^{-6}$, against the exact $2.8\cdot10^{-7}$. Moral:
each moment of information divides the bound polynomially;
the exponential moment changes its \emph{nature}.

\textbf{6.} $\cosh t = \sum_{k\geq0}\frac{t^{2k}}{(2k)!}$ and
$\eu^{t^2/2} = \sum_{k\geq0}\frac{t^{2k}}{2^kk!}$; the claim
follows term by term from $(2k)! \geq 2^kk!$, which holds by
induction: $(2k)! = 2k(2k-1)\cdot(2k-2)! \geq 2k\cdot
2^{k-1}(k-1)! = 2^kk!\cdot(2k-1) \geq 2^kk!$.

\textbf{7.} By \omterm{def:b2:proba:independence}{independence}, $\E\bigl(\eu^{t\sum\varepsilon_i}
\bigr) = (\cosh t)^n \leq \eu^{nt^2/2}$, so Markov gives
$\P(\sum\varepsilon_i \geq s) \leq \eu^{nt^2/2 - ts}$;
minimizing at $t = s/n$ yields $\eu^{-s^2/(2n)}$.

\textbf{8.} With $X_i = \frac{1 + \varepsilon_i}2$, $\widehat
p_n - \frac12 = \frac1{2n}\sum\varepsilon_i$, so
$\{\widehat p_n - \frac12 \geq \delta\} =
\{\sum\varepsilon_i \geq 2n\delta\}$ and question 7 gives the
bound $\eu^{-(2n\delta)^2/(2n)} = \eu^{-2n\delta^2}$. The
symmetric event has the same bound, whence the factor $2$ for
$\abs{\widehat p_n - \frac12} \geq \delta$.

\textbf{9.} $\E(\eu^{tX}) = \sum_x\eu^{tx}\P(X = x)$ is a
series of smooth functions of $t$ whose term-by-term
derivatives are dominated, on every \omterm{def:b2:metric:compact}{compact} $t$-interval, by
$\eu^{\abs t}\P(X = x)$ (as $0 \leq x \leq 1$): by the
differentiation theorem for \omterm{def:b2:funcseq:series}{normally} convergent series
(\cref{thm:b2:funcseq:differentiation}) it is twice
\omterm{def:b2:diffcalc:differential}{differentiable}, and the quotient rule gives $\psi' = \E_t(X)$
and $\psi'' = \E_t(X^2) - \E_t(X)^2$, where $\E_t$ is the
\omterm{def:b2:randomvar:expectation}{expectation} for the reweighted weights
$\eu^{tx}\P(X{=}x)/\E(\eu^{tX})$ --- nonnegative, summing to
$1$, carried by the same values $x \in \intcc01$. A \omterm{def:b2:randomvar:variance}{variance}
of a $\intcc01$-valued variable is at most $\frac14$: by
\cref{exo:b2:randomvar:6}, it is $\min_c\E_t((X - c)^2) \leq
\E_t\bigl((X - \tfrac12)^2\bigr) \leq \tfrac14$. Taylor with
integral remainder, using $\psi(0) = 0$, $\psi'(0) = p$:
\[
\psi(t) = tp + \int_0^t(t - s)\,\psi''(s)\,\dd s
\leq tp + \frac{t^2}2\cdot\frac14,
\]
i.e.\ $\E(\eu^{t(X - p)}) \leq \eu^{t^2/8}$ for all real $t$.

\textbf{10.} By \omterm{def:b2:proba:independence}{independence}, $\E\bigl(\eu^{t(S_n -
np)}\bigr) \leq \eu^{nt^2/8}$; Markov and the optimization $t
= 4\delta$ give
\[
\P(\widehat p_n - p \geq \delta)
\leq \eu^{nt^2/8 - tn\delta}\Big|_{t = 4\delta}
= \eu^{-2n\delta^2};
\]
applying this to the variables $1 - X_i$ (also in
$\intcc01$) bounds the other tail, whence the two-sided
$2\eu^{-2n\delta^2}$.

\textbf{11.} Chebyshev needs only a second moment and gives
$\frac{p(1-p)}{n\delta^2}$; Hoeffding needs
\emph{boundedness} and gives $2\eu^{-2n\delta^2}$. At $p =
\frac12$, $\delta = 0.03$: the bounds are
$\frac{278}{n}$ (approximately) versus
$2\eu^{-0.0018n}$; they cross near $n \approx 1200$, after
which the exponential bound wins, and wildly so ($n = 5000$:
$0.056$ versus $2.5\cdot10^{-4}$).

\textbf{12.} By Hoeffding (question 10), $\P(\abs{\widehat
p_n - p} \geq \delta) \leq 2\eu^{-2n\delta^2} \leq \alpha$ as
soon as $2n\delta^2 \geq \ln\frac2\alpha$, i.e.\ $n \geq
\frac{\ln(2/\alpha)}{2\delta^2}$.

\textbf{13.} $\delta = 0.03$, $\alpha = 0.05$: $n \geq
\frac{\ln 40}{2\cdot0.0009} \approx 2049.4$: $2050$ people.
For $\delta = 0.01$: $n \geq \frac{\ln40}{0.0002} \approx
18\,445$. The population size never appears because each
sampled voter is modelled as a fresh Bernoulli$(p)$ draw: the
poll's difficulty is the \omterm{def:b2:randomvar:variance}{variance} of a coin, not the size of
the country. Halving the margin costs four times the sample
--- the $1/\delta^2$ law.

\textbf{14.} Chebyshev: $\P(\abs{\widehat p_n - p} \geq
\delta) \leq \frac{p(1-p)}{n\delta^2} \leq
\frac1{4n\delta^2} \leq \alpha$ for $n \geq
\frac1{4\alpha\delta^2}$, i.e.\ $5556$ at three points ---
about $2.7$ times the Hoeffding requirement. Without
replacement, the \omterm{def:b2:randomvar:variance}{variance} is multiplied by $\frac{N -
n}{N-1} < 1$ (\cref{exo:b2:randomvar:5}), so the same $n$ can
only do better: the with-replacement computation is the
conservative one.

\textbf{15.} $\{\widehat p_n \leq \frac12\} \subseteq
\{\widehat p_n - 0.52 \leq -0.02\}$, so by the one-sided
Hoeffding bound $\P(\widehat p_n \leq \tfrac12) \leq
\eu^{-2n(0.02)^2} \leq 0.01$ as soon as $n \geq \frac{\ln
100}{2\cdot0.0004} \approx 5756.5$: $5757$ voters. The cost
scales like the inverse square of the \emph{lead}, not of the
desired precision: close races are expensive.

\textbf{16.} Used: the sample is drawn uniformly and
independently from the electorate; every sampled person
answers, honestly, and $p$ does not move during polling. Real
polls violate all three: reachable and willing respondents
are not a uniform sample (selection and non-response bias),
and answers can be untruthful or unstable. These are
\emph{bias} errors: they shift $\E(\widehat p_n)$ away from
$p$ by an amount independent of $n$, so no sample size
reduces them --- the mathematics of this Part controls only
the fluctuation term.

\textbf{17.} Chebyshev for one group of size $m$: $\P(\abs{
\widehat p^{(i)} - p} \geq \delta) \leq \frac{1}{4m\delta^2}
\leq \frac18$ for $m \geq \frac2{\delta^2}$. If fewer than
$k/2$ groups err, then more than $k/2$ of the values
$\widehat p^{(i)}$ lie in the \omterm{def:b2:metric:topology}{open} interval $\intoo{p -
\delta}{p + \delta}$, and so does their median; hence
$\{\abs{M - p} \geq \delta\}$ forces at least $\lceil
k/2\rceil$ errors among $k$ \omterm{def:b2:proba:independence}{independent} groups. The union
bound over the $\binom k{\lceil k/2\rceil}$ possible sets of
erring groups gives
\[
\P(\abs{M - p} \geq \delta)
\leq \binom{k}{\lceil k/2\rceil}
\Bigl(\frac18\Bigr)^{k/2}
\leq 2^k\,8^{-k/2} = 2^{-k/2} :
\]
exponential decay in the number of groups, bought with
nothing but \omterm{def:b2:randomvar:variance}{variances} --- useful precisely when the summands
are unbounded and Hoeffding is unavailable.

\textbf{18.} Cauchy--Schwarz
(\cref{thm:b2:randomvar:jensen}):
\[
\E(X) = \E(X\,\mathbf 1_{X>0})
\leq \sqrt{\E(X^2)}\sqrt{\E(\mathbf 1_{X>0}^2)}
= \sqrt{\E(X^2)\,\P(X > 0)} ;
\]
square and divide.

\textbf{19.} Let $h(a) = I(a) - 2(a - \tfrac12)^2$. Then
$h(\tfrac12) = 0$, $h'(a) = \ln\frac a{1-a} - 4(a -
\tfrac12)$ vanishes at $\tfrac12$, and
\[
h''(a) = \frac1a + \frac1{1-a} - 4 = \frac{1}{a(1-a)} - 4
\geq 0
\]
since $a(1-a) \leq \frac14$. So $h'$ increases from $0$ on
$\intco{\frac12}1$, hence $h' \geq 0$ and $h \geq 0$: $I(a)
\geq 2(a - \tfrac12)^2$.

\textbf{20.} $I(\tfrac12) = I'(\tfrac12) = 0$, $I''(a) =
\frac1{a(1-a)}$ gives $I''(\tfrac12) = 4$, and $I'''(\tfrac12)
= 0$ (the function is \omterm{def:b2:quadratic:adjoint}{symmetric} about $\tfrac12$), so
$I(\tfrac12 + \delta) = 2\delta^2 + O(\delta^4)$. Question 4
then bounds the true tail \emph{below} by
$\eu^{-n(2\delta^2 + O(\delta^4))}/(n+1)$: for small $\delta$
the Hoeffding exponent $2n\delta^2$ is asymptotically exact
--- only polynomial-in-$n$ improvements are possible.

\textbf{21.} Markov: one moment, decay $1/a$, useful only as
the engine behind the others (question 1 shows it flat).
Chebyshev: two moments, decay $\frac{V}{n\delta^2}$, sharp
without further hypotheses
(\cref{ex:b2:randomvar:chebsharp}), and the best tool at
question 23. Fourth moment (\cref{exo:b2:randomvar:9}):
decay $C/n^2$, just enough \omterm{def:b2:series:summable}{summability} for a strong law.
Hoeffding: bounded variables, decay $2\eu^{-2n\delta^2}$,
the workhorse of Part III. Chernoff with the exact rate
$I(a)$: full exponential moments, unbeatable exponent
(questions 4, 20), the reference point for everything else.

\textbf{22.} If the coin is fair: $\P(\widehat p_n > 0.525)
\leq \P(\widehat p_n - \tfrac12 \geq 0.025) \leq
\eu^{-2n(0.025)^2}$. If $p = 0.55$: $\P(\widehat p_n \leq
0.525) \leq \P(\widehat p_n - 0.55 \leq -0.025) \leq
\eu^{-2n(0.025)^2}$. Both errors are below $0.01$ when
$2n(0.025)^2 \geq \ln 100$, i.e.\ $n \geq 3684.2$: $3685$
flips. (Distinguishing hypotheses $2.5$ points apart costs
what estimating to $\pm2.5$ points costs.)

\textbf{23.} Hoeffding: $n \geq \frac{\ln 40}{2(0.005)^2}
\approx 73\,778$. Chebyshev with the true \omterm{def:b2:randomvar:variance}{variance} $p(1-p)
= 0.0099$: $n \geq \frac{0.0099}{0.05\cdot(0.005)^2} =
7920$ --- nine times cheaper. Hoeffding's exponent
$2n\delta^2$ prices the \omterm{def:b2:randomvar:variance}{variance} at its worst case $\frac14$,
absurdly pessimistic when $p = 0.01$; the humble second
moment knows better. The missing tool is a variance-aware
exponential bound (Bernstein's inequality, Year 3) --- or,
for rare \omterm{def:b2:proba:space}{events}, the Poisson approximation proved in
\cref{ch:b2:genfun}, which works on the natural relative
scale.

\textbf{24.} Fix $\delta > 0$: $\sum_n 2\eu^{-2n\delta^2} <
\infty$ (geometric-type series), so Borel--Cantelli~1
(\cref{thm:b2:proba:borelcantelli}) gives
$\P(\abs{\widehat p_n - p} \geq \delta \text{ infinitely
often}) = 0$, i.e.\ the \omterm{def:b2:proba:space}{event} $E_j =
\bigcup_N\bigcap_{n\geq N}\{\abs{\widehat p_n - p} <
\tfrac1j\}$ has probability $1$ for each $j$. The \omterm{def:b2:structures:countable}{countable}
intersection $\bigcap_jE_j$ still has probability $1$
(subadditivity on the complements), and on it $\widehat p_n
\to p$: the strong law of large numbers for coin flips, with
boundedness playing the role that the fourth moment played in
\cref{exo:b2:randomvar:9}.

\textbf{25.} One moment buys a flat bound; two buy
$1/(n\delta^2)$, and no more (the sharpness example); four
buy $1/n^2$, enough to telescope into an almost-sure law;
boundedness buys $\eu^{-2n\delta^2}$; and the full
exponential moment buys the exact rate $I$, which no method
beats. Polling $2050$ people suffices for any country because
the sample's fluctuation is governed by the coin's \omterm{def:b2:randomvar:variance}{variance},
not the population's size --- the $1/\delta^2$ and
$\ln(1/\alpha)$ price tags are universal. The Year 3 volume's
central limit theorem replaces these inequalities, on the
$\sqrt n$ scale, by an exact limit \omterm{def:b2:randomvar:law}{law} with explicit
constants --- turning every bound of this problem into an
asymptotic equality.
\end{solution}
