\section*{Chapter \ref{ch:b2:genfun} --- Probability Generating Functions}

\begin{solution}{exo:b2:genfun:1}
Fair die: $G(t) = \frac16(t + t^2 + \dots + t^6) = \frac t6(1 + t
+ \dots + t^5)$. If the sum of two fair dice were uniform on
$\{2, \dots, 12\}$, then
\[
G(t)^2 = \frac{t^2}{36}\,h(t)^2
= \frac{t^2}{11}\sum_{k=0}^{10}t^k ,
\qquad h(t) = 1 + t + \dots + t^5 .
\]
Now $h$ is a real polynomial of odd degree $5$, so it has a real
root (intermediate value theorem; concretely $h(-1) = 0$), hence
$h^2$ has a real root. But $\sum_{k=0}^{10}t^k$ has none: it is
positive for $t \geq 0$, and for $t < 0$ it equals
$\frac{t^{11} - 1}{t - 1}$, a quotient of two negative numbers.
Contradiction --- the sum of two fair dice is never uniform (as
the familiar triangular \omterm{def:b2:randomvar:law}{distribution} of dice sums confirms).
\end{solution}

\begin{solution}{exo:b2:genfun:2}
\emph{Binomial:} $G(t) = (1 - p + pt)^n$, $G'(t) = np(1 - p +
pt)^{n-1}$, $G''(t) = n(n-1)p^2(1 - p + pt)^{n-2}$, so
\[
\E(X) = G'(1) = np,
\qquad
V(X) = G''(1) + G'(1) - G'(1)^2
= n(n-1)p^2 + np - n^2p^2 = np(1-p).
\]
\emph{Geometric} ($q = 1 - p$): $G(t) = \frac{pt}{1 - qt}$, so
$G'(t) = \frac{p}{(1 - qt)^2}$ and $G''(t) = \frac{2pq}{(1 -
qt)^3}$; at $t = 1$ (using $1 - q = p$):
\[
\E(X) = \frac{p}{p^2} = \frac1p,
\qquad
V(X) = \frac{2q}{p^2} + \frac1p - \frac{1}{p^2}
= \frac{2q + p - 1}{p^2}
= \frac{q}{p^2} ,
\]
matching \cref{exo:b2:randomvar:1} with less work.
\end{solution}

\begin{solution}{exo:b2:genfun:3}
No, even with different loadings. Suppose $X, Y$ are \omterm{def:b2:randomvar:law}{laws} on
$\{1, \dots, 6\}$ with uniform sum. Then $G_X(t) = t\,a(t)$ and
$G_Y(t) = t\,b(t)$ with $a, b$ real polynomials of degree \emph{at
most} $5$ --- and their degrees must sum to $10$ (the sum reaches
$12$ with positive probability), so $\deg a = \deg b = 5$, both
odd. As in \cref{exo:b2:genfun:1},
\[
a(t)\,b(t) = \frac{1}{11}\sum_{k=0}^{10}t^k
\]
would force a real root on the left (each odd-degree real
polynomial has one) and none on the right. So no loading of two
\omterm{def:b2:proba:independence}{independent} dice --- equal or not --- produces a uniform sum.
\end{solution}

\begin{solution}{exo:b2:genfun:4}
By \cref{thm:b2:genfun:compound} with $G_N(s) = e^{\lambda(s-1)}$
and $G_X(t) = 1 - p + pt$:
\[
G_S(t) = e^{\lambda(1 - p + pt - 1)} = e^{\lambda p(t - 1)} :
\]
$S \sim \mathcal{P}(\lambda p)$. The discarded count $D = N - S$
counts the same items kept with probability $1 - p$, so by the
same computation $D \sim \mathcal{P}(\lambda(1 - p))$.
\omterm{def:b2:proba:independence}{Independence}, directly: for $j, k \in \N$,
\begin{align*}
\P(S = j,\ D = k)
&= \P(N = j + k)\,\binom{j+k}{j}p^jq^k
= e^{-\lambda}\frac{\lambda^{j+k}}{(j+k)!}\,
\frac{(j+k)!}{j!\,k!}\,p^jq^k\\
&= \Bigl(e^{-\lambda p}\frac{(\lambda p)^j}{j!}\Bigr)
\Bigl(e^{-\lambda q}\frac{(\lambda q)^k}{k!}\Bigr)
\end{align*}
with $q = 1 - p$: the joint \omterm{def:b2:randomvar:law}{law} factors as
$\mathcal{P}(\lambda p) \otimes \mathcal{P}(\lambda q)$. A Poisson
flow split at random yields \emph{\omterm{def:b2:proba:independence}{independent}} Poisson flows ---
a small miracle constantly used in queueing theory.
\end{solution}

\begin{solution}{exo:b2:genfun:5}
The waiting times between consecutive heads are \omterm{def:b2:proba:independence}{independent}
geometric $\mathcal{G}(p)$ variables (memorylessness: after each
head the game restarts), so $T_r = W_1 + \dots + W_r$ and
multiplicativity (\cref{thm:b2:genfun:product}) gives
\[
G_{T_r}(t) = \Bigl(\frac{pt}{1 - qt}\Bigr)^{r},
\qquad
\E(T_r) = r\,\E(W_1) = \frac rp,
\qquad
V(T_r) = r\,V(W_1) = \frac{rq}{p^2}
\]
($q = 1 - p$; \omterm{def:b2:randomvar:variance}{variances} add by \omterm{def:b2:proba:independence}{independence}). Expansion: by the
generalized binomial series
(\cref{ch:b2:powerseries}), $(1 - qt)^{-r} = \sum_{m\geq0}
\binom{m + r - 1}{r - 1}q^mt^m$, so the coefficient of $t^n$ in
$p^rt^r(1 - qt)^{-r}$ is (with $m = n - r$)
\[
\P(T_r = n) = \binom{n-1}{r-1}p^r(1-p)^{n-r},
\qquad n \geq r ,
\]
the \emph{negative binomial} \omterm{def:b2:randomvar:law}{law} --- combinatorially: the $r$-th
head falls at toss $n$ iff the previous $r - 1$ heads choose their
places among the first $n - 1$ tosses.
\end{solution}

\begin{solution}{exo:b2:genfun:6}
$m = 1\cdot\frac38 + 2\cdot\frac38 + 3\cdot\frac18 = \frac{3 + 6 +
3}{8} = \frac32 > 1$: supercritical. The \omterm{ex:b2:powerseries:fibonacci}{generating function} is
\[
G(t) = \frac{1 + 3t + 3t^2 + t^3}{8} = \frac{(1 + t)^3}{8} ,
\]
so fixed points solve $(1 + t)^3 = 8t$, i.e.\ $t^3 + 3t^2 - 5t + 1
= 0$. Factoring out the guaranteed root $t = 1$:
\[
t^3 + 3t^2 - 5t + 1 = (t - 1)\bigl(t^2 + 4t - 1\bigr),
\]
and $t^2 + 4t - 1 = 0$ gives $t = -2 \pm \sqrt5$. The root in
$\intco{0}{1}$ is $\sqrt5 - 2 \approx 0.236$: by
\cref{thm:b2:genfun:extinction},
\[
q = \sqrt 5 - 2 .
\]
(A pleasant check: the offspring \omterm{def:b2:randomvar:law}{law} is that of $3$ \omterm{def:b2:proba:independence}{independent}
fair coins, $Z_1 \sim \mathcal{B}(3, \frac12)$.)
\end{solution}

\begin{solution}{exo:b2:genfun:7}
\emph{\omterm{def:b2:randomvar:expectation}{Expectation}.} First $\E(Z_n) = m^n$: by
\cref{thm:b2:genfun:compound}, $\E(Z_{n+1}) = \E(Z_n)\,m$, and
$\E(Z_0) = 1$. The family $\bigl(Z_n(\omega)\P(\{\omega\})
\bigr)_{n, \omega}$ is nonnegative, so Fubini for families applies
unconditionally:
\[
\E(Y) = \sum_{n=0}^{\infty}\E(Z_n)
= \sum_{n=0}^\infty m^n = \frac{1}{1 - m} < \infty
\]
(in particular $Y$ is almost surely finite: consistent with
certain extinction in the subcritical case).

\emph{Functional equation.} Decompose the population by the
ancestor's children: if the ancestor has $Z_1 = k$ children, the
total progeny is $Y = 1 + Y_1 + \dots + Y_k$, where $Y_i$ is the
total progeny of the $i$-th child's line --- and the $Y_i$ are
\omterm{def:b2:proba:independence}{independent} copies of $Y$, independent of $Z_1$ (distinct lines
use disjoint, \omterm{def:b2:proba:independence}{independent} reproduction \omterm{def:b2:proba:space}{events}). Conditioning on
$Z_1$ as in \cref{thm:b2:genfun:compound}:
\[
H(t) = \E\bigl(t^Y\bigr)
= t\sum_{k=0}^\infty \P(Z_1 = k)\,H(t)^k
= t\,G\bigl(H(t)\bigr),
\]
the factor $t$ accounting for the ancestor itself. (For the
\omterm{def:b2:randomvar:law}{law} $p_0 = 1 - p$, $p_2 = p$ of binary branching, this quadratic
equation in $H$ can be solved explicitly and expanded --- the
\omterm{ex:b2:powerseries:catalan}{Catalan numbers} of \cref{ch:b2:powerseries} count the family
trees.)
\end{solution}

\begin{solution}{exo:b2:genfun:8}
Let $R > 1$ be the radius and fix $t \in \intoo{1}{R}$. Then
$\E(t^X) = G(t) < \infty$, and Markov's inequality
(\cref{thm:b2:randomvar:markov}) applied to the nonnegative
variable $t^X$ at level $t^n$:
\[
\P(X \geq n) = \P\bigl(t^X \geq t^n\bigr)
\leq \frac{G(t)}{t^n} = C\rho^n,
\qquad C = G(t),\quad \rho = \frac1t \in \intoo{0}{1}.
\]
\emph{Converse:} if $\P(X \geq n) \leq C\rho^n$, then $p_n \leq
\P(X \geq n) \leq C\rho^n$, so for $\abs t < \frac1\rho$ the
series $\sum p_n\abs t^n$ is dominated by the convergent geometric
series $C\sum(\rho\abs t)^n$: the radius is at least
$\frac1\rho > 1$. Radius of the \omterm{ex:b2:powerseries:fibonacci}{generating function} and geometric
decay of the tail are two faces of the same property.
\end{solution}

\begin{solution}{exo:b2:genfun:9}
Write $p_k^{(n)} = \P(X_n = k)$, $p_k = \P(X = k)$.

\emph{Case $k = 0$.} For $t \in \intoo{0}{1}$ and any \omterm{def:b2:randomvar:law}{law}
$(q_j)$ with $\sum_j q_j \leq 1$:
\[
\Bigl|\,q_0 - \sum_j q_jt^j\Bigr|
= \sum_{j \geq 1} q_j t^j
\leq \sum_{j\geq1}t^j = \frac{t}{1 - t} .
\]
Hence
\[
\abs{p_0^{(n)} - p_0}
\leq \frac{2t}{1 - t}
+ \abs{G_{X_n}(t) - G_X(t)} .
\]
Given $\varepsilon > 0$, choose $t$ with $\frac{2t}{1-t} <
\frac\varepsilon2$, then $n_0$ such that the last term is $<
\frac\varepsilon2$ for $n \geq n_0$: so $p_0^{(n)} \to p_0$.

\emph{Induction step.} Assume $p_j^{(n)} \to p_j$ for $j < k$.
Consider the \emph{shifted} functions
\[
g_n(t) = \frac{G_{X_n}(t) - p^{(n)}_0}{t}
= \sum_{j\geq0} p^{(n)}_{j+1}t^j,
\qquad
g(t) = \frac{G_X(t) - p_0}{t} ,
\]
\omterm{ex:b2:powerseries:fibonacci}{generating functions} of the sub-probability sequences
$(p^{(n)}_{j+1})_j$ (total mass $\leq 1$, which is all the
$k = 0$ argument used). For fixed $t \in \intoo{0}{1}$, $g_n(t)
\to g(t)$ by hypothesis and the case $k = 0$. Applying the $k =
0$ argument to $g_n$ gives $p_1^{(n)} \to p_1$; iterating the
shift $k$ times gives $p_k^{(n)} \to p_k$ for every $k$. (This is
the discrete, elementary instance of Lévy's \omterm{def:b2:metric:continuity}{continuity} theorem,
whose general form --- for characteristic functions --- is a Year
3 landmark.)
\end{solution}

\begin{solution}{exo:b2:genfun:10}
\omterm{def:b2:funcseq:def}{Pointwise}, $\frac{1 + (-1)^X}{2}$ equals $1$ when $X$ is even
and $0$ when odd, so taking \omterm{def:b2:randomvar:expectation}{expectations} (transfer),
\[
\P(X \text{ even}) = \frac{1 + \E\bigl((-1)^X\bigr)}2 =
\frac{1 + G_X(-1)}2 .
\]
Poisson: $\frac{1 + \eu^{-2\lambda}}2 \to \frac12$ as
$\lambda$ grows. Binomial: $\frac{1 + (1 - 2p)^n}2$. In both
cases $G_X(-1) \to 0$ says the parity of $X$ becomes a fair
coin: the \omterm{def:b2:randomvar:law}{law} spreads over many integers and forgets its
parity.
\end{solution}

\begin{solution}{exo:b2:genfun:11}
$t + \dots + t^6 = t\,\frac{1 - t^6}{1 - t}$ and $1 - t^6 =
(1 - t)(1 + t)(1 + t + t^2)(1 - t + t^2)$, giving the stated
factorization. For the first die, $(1 + t)(1 + t + t^2) = 1 +
2t + 2t^2 + t^3$, so $\frac{t(1+t)(1+t+t^2)}6 = \frac{t +
2t^2 + 2t^3 + t^4}6$: faces $\{1, 2, 2, 3, 3, 4\}$. For the
second, expanding
\[
(1 + 2t + 2t^2 + t^3)(1 - t + t^2)^2 = 1 + t^2 + t^3 + t^4 +
t^5 + t^7,
\]
so $\frac{t(1+t)(1+t+t^2)(1-t+t^2)^2}6 = \frac{t + t^3 + t^4
+ t^5 + t^6 + t^8}6$: faces $\{1, 3, 4, 5, 6, 8\}$. The
product of the two \omterm{ex:b2:powerseries:fibonacci}{generating functions} regroups the six
factors into $\bigl(\frac{t(1+t)(1+t+t^2)(1-t+t^2)}6
\bigr)^2$, the square of the standard die's function: the
Sicherman pair has exactly the standard \omterm{def:b2:randomvar:law}{law} for the total ---
\omterm{ex:b2:powerseries:fibonacci}{generating functions} classify all such regroupings.
\end{solution}

\begin{solution}{exo:b2:genfun:12}
Let $A$ and $B$ be the \omterm{ex:b2:powerseries:fibonacci}{generating functions} of the remaining
duration started from ``no current head'' and ``one current
head''. One toss is spent, then: from state $0$, tails
returns to state $0$, heads moves to state $1$; from state
$1$, heads ends the game, tails returns to state $0$:
\[
A(t) = t\bigl(q\,A(t) + p\,B(t)\bigr),
\qquad
B(t) = t\bigl(p + q\,A(t)\bigr).
\]
Substituting: $A(1 - qt) = pt\,B = pt(pt + qtA)$, so
\[
G_T(t) = A(t) = \frac{p^2t^2}{1 - qt - pq\,t^2} .
\]
At $t = 1$ the denominator is $1 - q - pq = p(1 - q) = p^2$:
$G_T(1) = 1$, the game ends almost surely (as
\cref{exo:b2:proba:6} showed by recursion). Logarithmic
differentiation at $1$: $\E(T) = 2 - \frac{D'(1)}{D(1)}$ with
$D(t) = 1 - qt - pqt^2$, $D'(1) = -q - 2pq$:
\[
\E(T) = 2 + \frac{q + 2pq}{p^2} = \frac{2p^2 + q + 2pq}{p^2}
= \frac{1 + p}{p^2},
\]
which is $6$ for $p = \frac12$.
\end{solution}

\begin{solution}{pb:b2:genfun:1}
\textbf{1.} For $t \in \intoo01$, the chain rule on $G_n = G
\circ G_{n-1}$ gives $G_n'(t) =
G'\bigl(G_{n-1}(t)\bigr)G_{n-1}'(t)$. As $t \to 1^-$,
$G_{n-1}(t) \uparrow 1$, and $G'$ is nondecreasing with
left limit $m$ at $1$, so the first factor tends to $m$; by
induction the second tends to $m^{n-1}$. By
\cref{thm:b2:genfun:moments}, $\E(Z_n) = G_n'(1^-) = m^n$.

\textbf{2.} Differentiating once more,
\[
G_n'' = G''(G_{n-1})\,(G_{n-1}')^2 +
G'(G_{n-1})\,G_{n-1}'',
\]
and letting $t \to 1^-$: $a_n = G''(1)m^{2(n-1)} + m\,
a_{n-1}$ with $a_n = G_n''(1)$, $a_1 = G''(1)$. For $m \neq
1$ one checks by induction that $a_n = G''(1)\,m^{n-1}
\frac{m^n - 1}{m - 1}$ (the recursion adds $G''(1)m^{2n-2}$
to $m\cdot G''(1)m^{n-2}\frac{m^{n-1}-1}{m-1}$, and
$m^{n-1} + \frac{m^{n-1}-1}{m-1} = \frac{m^n - 1}{m-1}$);
for $m = 1$, $a_n = a_{n-1} + G''(1) = n\,G''(1)$.

\textbf{3.} $V(Z_n) = a_n + m^n - m^{2n}$ and $G''(1) =
\sigma^2 + m^2 - m$. For $m \neq 1$, the piece $(m^2 -
m)m^{n-1}\frac{m^n-1}{m-1} = m^n(m^n - 1)$ cancels $m^n -
m^{2n}$ exactly, leaving $V(Z_n) =
\sigma^2m^{n-1}\frac{m^n-1}{m-1}$. For $m = 1$: $V(Z_n) =
nG''(1) = n\sigma^2$.

\textbf{4.} $Z_n$ is a nonnegative integer variable, so
$\P(Z_n > 0) = \P(Z_n \geq 1) \leq \E(Z_n) = m^n$ by Markov
(\cref{thm:b2:randomvar:markov}). For $m < 1$ this decays
geometrically --- and \omterm{def:b2:series:summable}{summably}, so Borel--Cantelli even gives
that only finitely many generations are nonempty, which is
extinction again.

\textbf{5.} Cauchy--Schwarz: $\E(Z_n)^2 = \E(Z_n\mathbf
1_{Z_n>0})^2 \leq \E(Z_n^2)\,\P(Z_n > 0)$. With question 3
and $m < 1$:
\[
\E(Z_n^2) = V(Z_n) + m^{2n}
\leq \frac{\sigma^2m^{n-1}}{1-m} + m^{2n},
\]
so, dividing $m^{2n}$ by this bound and simplifying by
$m^n$,
\[
\P(Z_n > 0) \geq \frac{m^n}{\frac{\sigma^2}{m(1-m)} + m^n}
\geq \Bigl(\frac{\sigma^2}{m(1-m)} + 1\Bigr)^{-1}m^n ,
\]
using $m^n \leq 1$ in the denominator. With question 4:
$\P(Z_n > 0) \asymp m^n$.

\textbf{6.} $G(t) = q\sum_k(pt)^k = \frac{q}{1 - pt}$, and $m
= G'(1) = \frac{pq}{(1-p)^2} = \frac pq$. Subcritical for $p
< \frac12$, critical for $p = \frac12$, supercritical for $p
> \frac12$.

\textbf{7.} $G(t) = t$ reads $pt^2 - t + q = 0$, with roots
$\frac{1 \pm \abs{p - q}}{2p}$, i.e.\ $1$ and $\frac qp =
\frac1m$. The extinction probability is the smallest fixed
point in $\intcc01$ (\cref{thm:b2:genfun:extinction}):
$q_{\mathrm{ext}} = 1$ if $m \leq 1$, and $\frac1m$ if $m >
1$.

\textbf{8.} For $m \neq 1$, with $p = \frac m{m+1}$, $q =
\frac1{m+1}$: if $q_n = \frac{m^n - 1}{m^{n+1} - 1}$, then
\[
1 - p\,q_n = \frac{(m+1)(m^{n+1} - 1) - m(m^n - 1)}
{(m+1)(m^{n+1} - 1)} = \frac{m^{n+2} - 1}{(m+1)(m^{n+1} -
1)},
\]
so $q_{n+1} = \frac{q}{1 - pq_n} = \frac{m^{n+1} -
1}{m^{n+2} - 1}$; the base case $q_0 = 0$ holds. For $m = 1$:
$G(t) = \frac1{2 - t}$ and $q_{n+1} = \frac1{2 -
\frac{n}{n+1}} = \frac{n+1}{n+2}$, with $q_0 = 0$.

\textbf{9.} $1 - q_n = \frac{m^n(m - 1)}{m^{n+1} - 1}$. For
$m < 1$ the denominator tends to $-1$: $1 - q_n \sim (1 -
m)\,m^n$. For $m > 1$:
\[
q_{\mathrm{ext}} - q_n = \frac1m - \frac{m^n - 1}{m^{n+1} -
1} = \frac{m - 1}{m\,(m^{n+1} - 1)} \sim \frac{m -
1}{m^{2}}\;m^{-n} .
\]
And $G'(t) = \frac{pq}{(1 - pt)^2}$ evaluated at $t = \frac
qp$ (where $1 - pt = 1 - q = p$) gives $G'(q_{\mathrm{ext}})
= \frac qp = \frac1m$: the observed ratio $m^{-1}$ is exactly
the derivative at the attracting fixed point.

\textbf{10.} For $p = \frac12$: $G''(t) = \frac{1/4}{(1 -
t/2)^3}$, so $G''(1) = 2$ and $\sigma^2 = G''(1) + m - m^2 =
2$. The closed form gives $1 - q_n = \frac1{n+1}$: the
survival probability decays like $1/n$ --- too slowly to be
\omterm{def:b2:series:summable}{summable}, unlike any subcritical rate.

\textbf{11.} Induction: $G_1(t) = \frac1{2-t}$ matches the
formula for $n = 1$, and
\[
G(G_n(t)) = \cfrac{1}{2 - \cfrac{n - (n-1)t}{n+1 - nt}}
= \frac{n + 1 - nt}{2(n+1) - 2nt - n + (n-1)t}
= \frac{n+1 - nt}{n + 2 - (n+1)t} .
\]
Then
\[
\frac{G_n(t) - q_n}{1 - q_n}
= (n+1)\,\Bigl(\frac{n - (n-1)t}{n+1 - nt} -
\frac{n}{n+1}\Bigr)
= \frac{t}{n + 1 - nt}
= \frac{\frac{t}{n+1}}{1 - \frac{n}{n+1}t} ,
\]
the pgf of the geometric \omterm{def:b2:randomvar:law}{law} $\mathcal G\bigl(\frac1{n+1}
\bigr)$ on $\N^*$ (\cref{ex:b2:genfun:classical}): given
survival, $\P(Z_n = k \mid Z_n > 0) =
\frac1{n+1}\bigl(\frac n{n+1}\bigr)^{k-1}$, with conditional
mean $n + 1$. The unconditional mean $1 = \E(Z_n)$ is the
product of a vanishing survival probability and a linearly
growing conditional size.

\textbf{12.} If the line goes extinct at generation $n$, then
$Y = Z_0 + \dots + Z_{n-1}$ is finite; if it never goes
extinct, $Y \geq \sum_n 1 = \infty$. So $\{Y < \infty\}$ is
the extinction \omterm{def:b2:proba:space}{event} and $\P(Y < \infty) = q_{\mathrm{ext}}$.
The derivation of $H(t) = tG(H(t))$ in
\cref{exo:b2:genfun:7} --- the ancestor contributes the
factor $t$, its children found \omterm{def:b2:proba:independence}{independent} copies of $Y$
counted through $G$ --- used only \cref
{thm:b2:genfun:compound}, valid in every regime.

\textbf{13.} With $G(s) = \frac{1 + s^2}2$ the equation reads
$tH^2 - 2H + t = 0$, so $H = \frac{1 - \sqrt{1 - t^2}}{t}$
(the root with $H(0) = 0$). Comparing with the Catalan series
$C(x) = \frac{1 - \sqrt{1 - 4x}}{2x}$
(\cref{ex:b2:powerseries:catalan}): $H(t) = \frac
t2\,C\bigl(\frac{t^2}4\bigr) =
\sum_{k\geq0}C_k\,\frac{t^{2k+1}}{2^{2k+1}}$, i.e.\ $\P(Y =
2k+1) = C_k2^{-2k-1}$. Checks: $\P(Y = 1) = C_0/2 = \frac12$
(the ancestor has no child); $\P(Y = 3) = C_1/8 = \frac18$
(two children, both childless: $\frac12\cdot\frac12\cdot
\frac12$).

\textbf{14.} Differentiating $H = tG(H)$ on $\intoo01$ and
letting $t \to 1^-$ (monotone limits as in
\cref{thm:b2:genfun:moments}): $H'(1)\bigl(1 - G'(H(1))\bigr)
= G(H(1))$. In the subcritical case $H(1) = 1$ and $\E(Y) =
H'(1) = \frac1{1 - m}$. In the critical case $G'(1) = 1$
makes the left factor vanish while the right side is $1$: no
finite $H'(1)$ can exist, so $\E(Y) = \infty$ --- yet $\P(Y <
\infty) = q = 1$.

\textbf{15.} $C_k = \frac1{k+1}\binom{2k}k \sim
\frac{4^k}{\sqrt\pi\,k^{3/2}}$ by
\cref{ex:b2:comparison:centralbinomial}, so
\[
\P(Y = 2k+1) = \frac{C_k}{2\cdot4^{k}} \sim
\frac1{2\sqrt\pi\,k^{3/2}} .
\]
Summing the tail (comparison with $\int_K^\infty
k^{-3/2}\dd k = 2K^{-1/2}$, above and below): $\P(Y > 2K)
\asymp K^{-1/2}$, i.e.\ $\P(Y > n) \asymp n^{-1/2}$ --- a
heavy tail with infinite mean, quantifying question 14.

\textbf{16.} Both critical objects --- the fair walk's return
time (the weekend problem of \cref{ch:b2:proba}) and the
critical total progeny --- are almost surely finite with
infinite mean, with local laws of exponent $-3/2$ and tails
of exponent $-1/2$. This is no coincidence: exploring a
family tree child by child produces a $\pm1$ path (one step
up per birth, one down per death) which is exactly a fair
walk, and $Y$ becomes a first passage time. Criticality means
zero drift: the process is always on the verge of both
extinction and explosion, and the $\sqrt{}$-scale
fluctuations of zero-drift randomness produce precisely these
exponents.

\textbf{17.} $G''(t) = \sum_{n\geq2}n(n-1)p_nt^{n-2}$ has
nonnegative terms, so it is nondecreasing on $\intco01$ with
finite limit $G''(1) = \sigma^2$ (criticality makes
$\E Z_1(Z_1 - 1) = \sigma^2$); a nondecreasing function with
limit equal to the boundary value is \omterm{def:b2:metric:continuity}{continuous} at $1$.
Taylor with integral remainder at the point $1$:
\[
G(t) = 1 + (t - 1) + \int_1^t(t - s)G''(s)\,\dd s
= t + \frac{G''(1)}2(1-t)^2 + o\bigl((1-t)^2\bigr),
\]
since $G''(s) = G''(1) + o(1)$ as $s \to 1^-$.

\textbf{18.} Reducing to a common denominator, $h(t) =
\frac{G(t) - t}{(1 - G(t))(1 - t)}$. By question 17 the
numerator is $b(1-t)^2 + o((1-t)^2)$ and $1 - G(t) = (1 -
t)\bigl(1 - b(1-t) + o(1-t)\bigr)$, so $h(t) \to b$.

\textbf{19.} By definition of $h$ at $t = q_j$ and $G(q_j) =
q_{j+1}$: $\frac1{1 - q_{j+1}} - \frac1{1-q_j} = h(q_j)$;
summing from $j = 0$ ($q_0 = 0$) gives the display. Since the
critical process dies out, $q_j \uparrow 1$, so $h(q_j) \to
b$ and the Cesaro mean $\frac1n\sum_{j<n}h(q_j) \to b$:
$\frac1{1-q_n} \sim bn$, i.e.
\[
\P(Z_n > 0) \sim \frac1{bn} = \frac{2}{\sigma^2 n} .
\]

\textbf{20.} Geometric critical case: $\sigma^2 = 2$
(question 10), so Kolmogorov predicts $1 - q_n \sim \frac1n$
--- and the exact value is $\frac1{n+1}$.

\textbf{21.} Since $Z_n\mathbf 1_{Z_n > 0} = Z_n$, $\E(Z_n
\mid Z_n > 0) = \frac{\E(Z_n)}{\P(Z_n > 0)} = \frac1{1 -
q_n} \sim \frac{\sigma^2n}2$. In the geometric case this is
$n + 1$, matching question 11 exactly ($\sigma^2 = 2$). The
critical picture: extinction is certain, the mean size is
frozen at $1$, and the rare surviving lines have size growing
linearly --- each factor balancing the other.

\textbf{22.} For $\mathcal P(\lambda)$ offspring, $G(t) =
\eu^{\lambda(t-1)}$ and the extinction probability is the
smallest root of $q = \eu^{\lambda(q-1)}$. Numerically:
$\lambda = 1.5$ gives $q \approx 0.417$ (iterate $q \mapsto
\eu^{1.5(q-1)}$: $0, 0.223, 0.312, 0.356, \dots \to 0.4172$);
$\lambda = 2$ gives $q \approx 0.203$. So one index case
sparks a major outbreak with probability $58\%$
($\lambda = 1.5$) or $80\%$ ($\lambda = 2$) --- likely, not
certain. The iteration from $q_0 = 0$ converges to the
\emph{smallest} root because $G$ is nondecreasing: by
induction $q_n \leq r$ for any fixed point $r$, and $(q_n)$
increases (it is $\P(Z_n = 0)$), so its limit is a fixed
point below all others.

\textbf{23.} The $k$ ancestors found \omterm{def:b2:proba:independence}{independent} family
trees, and total extinction is the intersection of $k$
\omterm{def:b2:proba:independence}{independent} extinction \omterm{def:b2:proba:space}{events}: probability $q^k$. For
$\lambda = 1.5$: outbreak probability $1 - q^k \geq 0.99$
requires $q^k \leq 0.01$, i.e.\ $k \geq
\frac{\ln 0.01}{\ln 0.417} \approx 5.3$: six initial cases
make the outbreak $99\%$ certain.

\textbf{24.} \emph{$G'(q) < 1$:} $G - \mathrm{id}$ is \omterm{def:b2:affine:convex}{convex}
and vanishes at $q$ and $1$, so it is $\leq 0$ on $\intcc
q1$; if $G'(q) = 1$, the tangent at $q$ (which convexity
places below $G$) would force $G(t) \geq t$ on $\intcc q1$,
hence $G \equiv \mathrm{id}$ there, killing all coefficients
$p_n$ ($n \geq 2$) and contradicting $m > 1$. \emph{Geometric
convergence:} $q_n < q$ for all $n$ (induction, $G$
increasing), and the mean value theorem gives $q - q_{n+1} =
G'(c_n)(q - q_n)$ with $c_n \in \intoo{q_n}q$, so $G'(c_n)
\leq G'(q) < 1$ and $q - q_n \leq q\,G'(q)^n$.
\emph{Companion process:} $\widehat G(t) = G(qt)/q =
\sum_kp_kq^{k-1}t^k$ has nonnegative coefficients and
$\widehat G(1) = G(q)/q = 1$: a pgf; its mean is $\widehat
G'(1) = G'(q) < 1$: subcritical. Geometric family: $G(t) =
\frac{q}{1-pt}$, $q_{\mathrm{ext}} = \frac qp$, and
\[
\widehat G(t) = \frac pq\cdot\frac{q}{1 - p\frac qp t}
= \frac{p}{1 - qt} :
\]
the geometric offspring \omterm{def:b2:randomvar:law}{law} with $p$ and $q$ exchanged ---
the supercritical process seen on its extinction \omterm{def:b2:proba:space}{event} is the
mirror subcritical one.

\textbf{25.} The table: $m < 1$: $q = 1$, $\P(Z_n > 0)
\asymp m^n$ (questions 4--5), $\E(Y) = \frac1{1-m}$,
surviving generations of bounded conditional mean. $m = 1$:
$q = 1$, $\P(Z_n > 0) \sim \frac2{\sigma^2n}$ (Kolmogorov),
$\E(Y) = \infty$ with $\P(Y > n) \asymp n^{-1/2}$, survivors
of size $\sim \frac{\sigma^2n}2$. $m > 1$: $q < 1$ is the
smallest fixed point, $q - q_n = O(G'(q)^n)$, $\E(Z_n) = m^n$
growth, and conditioned on dying the process is the
subcritical companion (question 24). The tools: composition
of pgfs turned population recursion into function iteration;
convexity fixed the geometry of the fixed points; Taylor at
$1^-$ converted moment hypotheses into local expansions; and
Cesaro averaging extracted Kolmogorov's $1/n$ from a
telescoping sum. The Year 3 volume adds the martingale $Z_n/
m^n$ --- whose almost-sure limit refines $\E(Z_n) = m^n$ into
a trajectory-by-trajectory growth rate --- and Yaglom's
theorem, the limit \omterm{def:b2:randomvar:law}{law} behind the conditional geometry
observed in question 11.
\end{solution}
