\section*{Chapter \ref{ch:g7:fractions} --- Fractions: Comparing and Adding}

\begin{solution}{exo:g7:fractions:1}
$\dfrac{2}{5} = \dfrac{8}{20}$; \quad
$\dfrac{18}{24} = \dfrac{3}{4}$; \quad
$\dfrac{7}{3} = \dfrac{28}{12}$; \quad
$\dfrac{5}{9} = \dfrac{20}{36}$.
\end{solution}

\begin{solution}{exo:g7:fractions:2}
$\dfrac{12}{16} = \dfrac34$; \quad
$\dfrac{30}{45} = \dfrac23$; \quad
$\dfrac{27}{36} = \dfrac34$; \quad
$\dfrac{48}{12} = 4$.
\end{solution}

\begin{solution}{exo:g7:fractions:3}
$\dfrac{15}{35} = \dfrac37$ and $\dfrac{21}{49} = \dfrac37$: equal.

$\dfrac{16}{28} = \dfrac47$ and $\dfrac{20}{36} = \dfrac59$: to
compare, $\frac47 = \frac{36}{63}$ and $\frac59 = \frac{35}{63}$ ---
not equal.
\end{solution}

\begin{solution}{exo:g7:fractions:4}
$\dfrac{7}{11} < \dfrac{9}{11}$ (same \omterm{def:g4:fractions:def}{denominator}).

$\dfrac56 = \dfrac{15}{18} > \dfrac{11}{18}$.

$\dfrac{13}{12} > 1$ while $\dfrac{19}{20} < 1$, so
$\dfrac{13}{12} > \dfrac{19}{20}$.
\end{solution}

\begin{solution}{exo:g7:fractions:5}
$\dfrac59 + \dfrac29 = \dfrac79$; \quad
$\dfrac{11}{7} - \dfrac47 = \dfrac77 = 1$; \quad
$\dfrac58 + \dfrac78 = \dfrac{12}{8} = \dfrac32$.
\end{solution}

\begin{solution}{exo:g7:fractions:6}
$\dfrac23 + \dfrac56 = \dfrac46 + \dfrac56 = \dfrac96 = \dfrac32$.

$\dfrac{7}{10} - \dfrac25 = \dfrac{7}{10} - \dfrac{4}{10} =
\dfrac{3}{10}$.

$\dfrac34 + \dfrac{5}{12} = \dfrac{9}{12} + \dfrac{5}{12} =
\dfrac{14}{12} = \dfrac76$.
\end{solution}

\begin{solution}{exo:g7:fractions:7}
$3 - \dfrac54 = \dfrac{12}{4} - \dfrac54 = \dfrac74$; \quad
$1 + \dfrac38 = \dfrac{11}{8}$; \quad
$2 - \dfrac76 = \dfrac{12}{6} - \dfrac76 = \dfrac56$.
\end{solution}

\begin{solution}{exo:g7:fractions:8}
$5 \times \dfrac{3}{10} = \dfrac{15}{10} = \dfrac32$; \quad
$\dfrac78 \times 4 = \dfrac{28}{8} = \dfrac72$; \quad
$\dfrac23$ of $45$ is $\dfrac{2 \times 45}{3} = \dfrac{90}{3} = 30$.
\end{solution}

\begin{solution}{exo:g7:fractions:9}
Together: $\dfrac14 + \dfrac38 = \dfrac28 + \dfrac38 = \dfrac58$ of
the pizza. Left: $1 - \dfrac58 = \dfrac38$.
\end{solution}

\begin{solution}{exo:g7:fractions:10}
Poured out: $2 \times \dfrac16 = \dfrac26 = \dfrac13$ L. Remaining:
\[
\frac34 - \frac13 = \frac{9}{12} - \frac{4}{12} = \frac{5}{12}
\text{ L}.
\]
\end{solution}

\begin{solution}{exo:g7:fractions:11}
Adding the positive quantity $\frac12$ to $\frac35$ must give a result
\emph{larger} than $\frac35$. But $\frac47 < \frac35$ (indeed
$\frac47 = \frac{20}{35}$ and $\frac35 = \frac{21}{35}$): the claimed
\omterm{def:g1:addition:def}{sum} is smaller than one of its terms --- impossible.
\end{solution}

\begin{solution}{exo:g7:fractions:12}
Step by step:
\[
\frac12 + \frac14 = \frac34, \qquad
\frac34 + \frac18 = \frac68 + \frac18 = \frac78, \qquad
\frac78 + \frac1{16} = \frac{14}{16} + \frac1{16} = \frac{15}{16}.
\]
The partial \omterm{def:g1:addition:def}{sums} $\frac12, \frac34, \frac78, \frac{15}{16}, \dots$ are
each time half-way closer to $1$: at every step, the missing part is
cut in \omterm{def:g3:division:half}{half}. The \omterm{def:g1:addition:def}{sums} approach $1$ without ever reaching it.
\end{solution}

\begin{solution}{pb:g7:fractions:1}
\textbf{1.} (a) $\frac12 + \frac14 + \frac14 = \frac24 + \frac14
+ \frac14 = \frac44 = 1$: legal. (b) $\frac14 + \frac18 +
\frac18 + \frac12 = \frac28 + \frac18 + \frac18 + \frac48 =
\frac88 = 1$: legal.

\textbf{2.} $\frac12 + \frac14 + \frac18 = \frac48 + \frac28 +
\frac18 = \frac78$: one eighth note is missing.

\textbf{3.} A measure is $1 = \frac{16}{16}$: sixteen sixteenth
notes. An eighth is $\frac18 = \frac{2}{16}$: two sixteenths.

\textbf{4.} Dotted \omterm{def:g3:division:half}{half}: $\frac12 + \frac14 = \frac34$. Dotted
\omterm{def:g3:division:half}{quarter}: $\frac14 + \frac18 = \frac28 + \frac18 = \frac38$.

\textbf{5.} The dotted \omterm{def:g3:division:half}{half} is worth $\frac34$ (question 4):
exactly one waltz measure. Other fillings, for instance: \omterm{def:g3:division:half}{quarter}
$+$ \omterm{def:g3:division:half}{quarter} $+$ \omterm{def:g3:division:half}{quarter} ($\frac14 + \frac14 + \frac14 =
\frac34$), or \omterm{def:g3:division:half}{quarter} $+$ eighth $+$ eighth $+$ \omterm{def:g3:division:half}{quarter}
($\frac14 + \frac18 + \frac18 + \frac14 = \frac{2 + 1 + 1 +
2}{8} = \frac68 = \frac34$).

\textbf{6.} $\frac38 + \frac18 + \frac14 = \frac38 + \frac18 +
\frac28 = \frac68 = \frac34$: one \omterm{def:g3:division:half}{quarter} note ($\frac14$) is
missing.

\textbf{7.} Dotted eighth: $\frac18 + \frac{1}{16} =
\frac{2}{16} + \frac{1}{16} = \frac{3}{16}$. \omterm{def:g3:division:half}{Quarter}:
$\frac{4}{16}$. The \omterm{def:g3:division:half}{quarter} is longer.

\textbf{8.} $\frac12 + \frac18 = \frac48 + \frac18 = \frac58$,
that is $\frac{10}{16}$: ten sixteenths.

\textbf{9.} $1 - \frac{11}{16} = \frac{16}{16} - \frac{11}{16}
= \frac{5}{16}$ of the measure is silence.

\textbf{10.} Both examples total $\frac{2 + 1 + 1}{16} =
\frac{4}{16}$: legal. All rhythms filling four sixteenths with
pieces of $2$ and $1$:
\[
2{+}2, \quad 2{+}1{+}1, \quad 1{+}2{+}1, \quad 1{+}1{+}2, \quad
1{+}1{+}1{+}1 :
\]
five rhythms.

\textbf{11.} A rhythm filling $n$ sixteenths ends either with a
sixteenth --- and what precedes fills $n - 1$ --- or with an
eighth --- and what precedes fills $n - 2$. Every rhythm is
counted exactly once this way, so
$\text{ways}(n) = \text{ways}(n-1) + \text{ways}(n-2)$. Table:
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{c|cccccccc}
$n$ & $1$ & $2$ & $3$ & $4$ & $5$ & $6$ & $7$ & $8$ \\
\hline
ways & $1$ & $2$ & $3$ & $5$ & $8$ & $13$ & $21$ & $34$ \\
\end{tabular}
\end{center}
A \omterm{def:g3:division:half}{half} note ($8$ sixteenths) can be drummed in $34$ ways.
($n = 4$ recovers the five rhythms of question 10.)

\textbf{12.} Each number is the \omterm{def:g1:addition:def}{sum} of the two before it:
$3 + 5 = 8$, $5 + 8 = 13$, $8 + 13 = 21$, $13 + 21 = 34$ ---
the rule question 11 proved, a thousand years old and now known
as the Fibonacci rule.

\textbf{13.} $\frac16 + \frac{1}{12} = \frac{2}{12} +
\frac{1}{12} = \frac{3}{12} = \frac14$: the same duration as
two straight eighths ($\frac18 + \frac18 = \frac14$). The swing
swap is legal.

\textbf{14.} Each triplet note lasts $\frac{1}{12}$: check,
$3 \times \frac{1}{12} = \frac{3}{12} = \frac14$
(\cref{prop:g7:fractions:mult}). For a \omterm{def:g3:division:half}{half} note: each lasts
$\frac16$, since $3 \times \frac16 = \frac36 = \frac12$.

\textbf{15.} Dotted \omterm{def:g3:division:half}{quarter} $+$ \omterm{def:g3:division:half}{quarter} $+$ \omterm{def:g3:division:half}{quarter}: in eighths,
$3 + 2 + 2 = 7$ eighths $= \frac78$: legal. Halves and \omterm{def:g3:division:half}{quarters}
are worth $4$ and $2$ eighths --- both \emph{\omterm{def:g3:numbers:evenodd}{even}} numbers. Any
\omterm{def:g1:addition:def}{sum} of \omterm{def:g3:numbers:evenodd}{even numbers} is \omterm{def:g3:numbers:evenodd}{even}, but a seven-eight measure needs
$7$ eighths, an \omterm{def:g3:numbers:evenodd}{odd number}: impossible. To fill an \omterm{def:g3:numbers:evenodd}{odd} measure,
something \omterm{def:g3:numbers:evenodd}{odd} --- an eighth or a dotted note --- must appear.
\end{solution}
