\section*{Chapter \ref{ch:g7:central} --- Central Symmetry and Parallelograms}

\begin{solution}{exo:g7:central:1}
Half-turn image: $M'$ is four squares \emph{left} of $O$ and one
square \emph{down} (both directions reversed). Reflection image
across the vertical line: four squares left but still one square
\emph{up}. The two images differ: they are vertical mirror images of
each other.
\end{solution}

\begin{solution}{exo:g7:central:2}
$A'B' = AB$ (half-turns preserve lengths) and
$(A'B') \parallel (AB)$ (the image of a line is a \omterm{def:g6:lines:perp}{parallel} line,
\cref{prop:g7:central:preserved}).
\end{solution}

\begin{solution}{exo:g7:central:3}
Calculator digits with a \omterm{def:g7:central:center}{center of symmetry}: $0$, $2$, $5$, $8$ (and
$1$ if drawn as a plain bar). Letters: H, N, S, Z, O have one; A and
T do not (they have an axis instead).
\end{solution}

\begin{solution}{exo:g7:central:4}
Opposite sides equal: $MN = KL = 7$ cm and $NK = LM = 4$ cm. Opposite
angles equal: $\widehat M = \widehat K = 115^\circ$. The diagonals
cut each other at their common midpoint, and that point is $O$: the
midpoint of $[LN]$ is $O$.
\end{solution}

\begin{solution}{exo:g7:central:5}
Draw a \omterm{def:g6:lines:objects}{segment} $[AC]$ of $8$ cm, mark its midpoint $O$, draw through
$O$ a line making $60^\circ$ with $(AC)$, and place $B$ and $D$ on it
at $2.5$ cm from $O$ on either side. Join $A$, $B$, $C$, $D$: the
diagonals cut at their midpoints, so $ABCD$ is a \omterm{def:g7:central:parallelogram}{parallelogram} by
definition.
\end{solution}

\begin{solution}{exo:g7:central:6}
Consecutive angles are supplementary: next to $115^\circ$ sits
$180 - 115 = 65^\circ$. Opposite angles are equal, so the four angles
are $115^\circ$, $65^\circ$, $115^\circ$, $65^\circ$ (\omterm{def:g1:addition:def}{sum}
$360^\circ$).
\end{solution}

\begin{solution}{exo:g7:central:7}
From $A(1,1)$ to $O(2.5, 2.5)$: $1.5$ right and $1.5$ up; continuing
the same: $A'(4, 4)$. From $B(4,2)$ to $O$: $1.5$ left and $0.5$ up;
continuing: $B'(1, 3)$.
\end{solution}

\begin{solution}{exo:g7:central:8}
No. An \omterm{def:g6:shapes:triangles}{isosceles} trapezoid has its two slanted sides equal without
being a \omterm{def:g7:central:parallelogram}{parallelogram} (the two \omterm{def:g6:lines:perp}{parallel} sides have different
lengths). What suffices is criterion 3 of
\cref{thm:g7:central:recognize}: two opposite sides equal
\emph{and \omterm{def:g6:lines:perp}{parallel}}.
\end{solution}

\begin{solution}{exo:g7:central:9}
$O$ is the midpoint of $[BC]$ by choice, and it is the midpoint of
$[AA']$ by construction of the \omterm{def:g7:central:def}{symmetric}. The \omterm{def:g4:geometry:polygon}{quadrilateral} $ABA'C$
has diagonals $[AA']$ and $[BC]$ sharing the midpoint $O$: criterion
1 (the definition) makes it a \omterm{def:g7:central:parallelogram}{parallelogram} at no extra cost.
\end{solution}

\begin{solution}{exo:g7:central:10}
``Opposite angles equal $\Rightarrow$ \omterm{def:g7:central:parallelogram}{parallelogram}'': \emph{true}
(this is another recognition criterion; with the angle \omterm{def:g1:addition:def}{sum}
$360^\circ$, equal opposite pairs force supplementary consecutive
angles, hence \omterm{def:g6:lines:perp}{parallel} sides by
\cref{thm:g7:triangles:equalities}).

``\omterm{def:g6:lines:perp}{Perpendicular} diagonals $\Rightarrow$ \omterm{def:g6:shapes:quadrilaterals}{rhombus}'': \emph{true} for a
\omterm{def:g7:central:parallelogram}{parallelogram} (the diagonals already cut at their midpoints).

``One \omterm{def:g3:shapes:rightangle}{right angle} $\Rightarrow$ rectangle'': \emph{true} ---
consecutive angles are supplementary, so all four become right.
\end{solution}

\begin{solution}{exo:g7:central:11}
Whatever the \omterm{def:g4:geometry:polygon}{quadrilateral} --- even a very irregular one --- the
midpoints $I$, $J$, $K$, $L$ always form a \emph{\omterm{def:g7:central:parallelogram}{parallelogram}}. The
proof with the midpoint theorem is in \cref{ch:g8:midpoints}
(\cref{exo:g8:midpoints:9}).
\end{solution}

\begin{solution}{pb:g7:central:1}
\textbf{1.} $M(3, 1) \to (-3, -1)$; $N(-2, 4) \to (2, -4)$;
$P(0, -5) \to (0, 5)$. Rule: the half-turn about the origin
sends $(x, y)$ to $(-x, -y)$ --- both coordinates change sign.

\textbf{2.} From $M(5, 3)$: three squares right of $C$ becomes
three squares left, two up becomes two down: image
$(-1, -1)$. Recipe check: twice the center minus the point,
$(2 \times 2 - 5,\ 2 \times 1 - 3) = (-1, -1)$.

\textbf{3.} The half-turn about the card's center swaps the pips
in pairs. With an \omterm{def:g3:numbers:evenodd}{odd number} of pips, one pip has no partner: it
must be its own image, and the only point that is its own image
is the center itself. Hence the middle pip of the 3, the 5, the
7\,\dots\ sits dead center (look at a real card: it does).

\textbf{4.} About $O_1(2,0)$: $A(1,1) \to (3,-1)$; about
$O_2(5,0)$: $(3,-1) \to (7,1)$. Start $A(1,1)$, finish
$(7,1)$: slid $6$ to the right, not flipped. Same for
$B(0,3) \to (4,-3) \to (6,3)$: slid $6$ right. The two
half-turns amount to a \emph{translation}.

\textbf{5.} The slide is $6$, and $O_1 O_2 = 3$: the translation
covers \emph{twice} the distance between the centers (in the
direction from $O_1$ to $O_2$) --- just as two \omterm{def:g6:lines:perp}{parallel} mirrors
translated by twice their gap in \cref{pb:g6:symmetry:1}.

\textbf{6.} The half-turn about $O$ sends the \omterm{def:g7:central:parallelogram}{parallelogram} onto
itself (that is what ``\omterm{def:g7:central:center}{center of symmetry}'' means for it,
\cref{thm:g7:central:properties}). A line through $O$ is also
sent onto itself, so the two points where it crosses the
boundary swap with each other: they are \omterm{def:g7:central:def}{symmetric} about $O$. The
two pieces of the \omterm{def:g7:central:parallelogram}{parallelogram} on either side of the line swap
too, so they are exact copies (\cref{prop:g7:central:preserved})
--- equal \omterm{def:g6:measure:area}{areas}.

\textbf{7.} Cut along the line through the \emph{two centers}:
the center of the cake rectangle and the center of the hole.
By question 6 applied to the cake rectangle, the cut halves the
cake; applied to the hole rectangle (the line passes through its
center too), it halves the hole. Each part gets \omterm{def:g3:division:half}{half} the cake
minus \omterm{def:g3:division:half}{half} the hole: perfectly fair.

\textbf{8.} $D = (0 + 8 - 6,\ 0 + 5 - 1) = (2, 4)$ (the
diagonals share their midpoint, so $D$ is the half-turn image of
$B$ about the center). Center: midpoint of $[AC]$,
$O(4,\ 2.5)$.

\textbf{9.} In the \omterm{def:g7:central:parallelogram}{parallelogram} $ABA'C$, opposite sides are
equal: $BA' = AC$. The triangle inequality in $ABA'$ gives
$AA' < AB + BA' = AB + AC$. Since $O$ is the midpoint of
$[AA']$, $AA' = 2\,AO$, so $2\,AO < AB + AC$, i.e.\
$AO < \frac{AB + AC}{2}$.

\textbf{10.} Upper bound: $AO < \frac{5 + 7}{2} = 6$ cm. Lower
bound: in $ABA'$, $AA' > BA' - AB = 7 - 5 = 2$, so
$AO > 1$ cm. The median from $A$ lies strictly between $1$ and
$6$ cm.

\textbf{11.} \omterm{def:g6:lines:objects}{Segment}: yes, its midpoint. Line: yes --- every one
of its points is a center. \omterm{def:g6:shapes:triangles}{Equilateral triangle}: none. Square:
yes, the crossing of the diagonals. Circle: yes, its center.
S and Z: yes, their middle point (turn the page upside down:
they read the same).

\textbf{12.} The half-turn would pair the three \omterm{def:g5:solids:def}{vertices} among
themselves; three being \omterm{def:g3:numbers:evenodd}{odd}, one \omterm{def:g5:solids:def}{vertex} $A$ would be its own
image, forcing $A$ to be the center of the half-turn. The two
other \omterm{def:g5:solids:def}{vertices} $B$ and $C$ would then be \omterm{def:g7:central:def}{symmetric} about $A$ ---
so $A$ would be the midpoint of $[BC]$, putting the three
\omterm{def:g5:solids:def}{vertices} on one line. A triangle has no three aligned \omterm{def:g5:solids:def}{vertices}:
contradiction. No center exists.

\textbf{13.} The same parity argument: a half-turn preserving
the \omterm{def:g4:geometry:polygon}{polygon} pairs up its \omterm{def:g5:solids:def}{vertices}; an \omterm{def:g3:numbers:evenodd}{odd number} of \omterm{def:g5:solids:def}{vertices}
forces a fixed \omterm{def:g5:solids:def}{vertex}, which must be the center and the midpoint
of the \omterm{def:g6:lines:objects}{segment} joining two other \omterm{def:g5:solids:def}{vertices} --- three aligned
\omterm{def:g5:solids:def}{vertices}, impossible in a \omterm{def:g4:geometry:polygon}{polygon}\,\dots\ so no \omterm{def:g4:geometry:polygon}{polygon} with an
\omterm{def:g3:numbers:evenodd}{odd number} of \omterm{def:g5:solids:def}{vertices} has a \omterm{def:g7:central:center}{center of symmetry}.

\textbf{14.} Performing the half-turn about $O_1$, then about
$O_2$, sends the figure onto itself both times --- so their
combination does too. By question 5, that combination is a
translation by twice the distance $O_1 O_2$, which is not \omterm{def:g1:counting:zero}{zero}
since the centers differ. A figure fitting on a sheet cannot
equal itself slid by a fixed nonzero amount (slide it enough
times and it leaves the sheet entirely). So a bounded figure has
at most one \omterm{def:g7:central:center}{center of symmetry}.

\textbf{15.} On the endless strip, a half-turn about the point
midway between a left footprint and the following right one
sends the pattern onto itself; the point one full step further
does too: two distinct centers. Composing the two half-turns
gives the translation by twice the half-step --- exactly the
one-step slide that visibly maps the infinite frieze onto
itself, as question 14 predicts. Unbounded patterns live by
different rules: that is the mathematics of wallpaper.
\end{solution}
