\section*{Chapter \ref{ch:g7:areas} --- Areas and Volumes}

\begin{solution}{exo:g7:areas:1}
$8 \times 5 = 40$ cm$^2$; \qquad $6.5 \times 4 = 26$ m$^2$.
\end{solution}

\begin{solution}{exo:g7:areas:2}
$A = 10 \times 4 = 40$ cm$^2$. The answer is not
$10 \times 6 = 60$ because the $6$ cm side is slanted: the formula
uses the \emph{height} (\omterm{def:g6:lines:perp}{perpendicular} distance), which is $4$ cm.
\end{solution}

\begin{solution}{exo:g7:areas:3}
$\frac{12 \times 7}{2} = 42$ cm$^2$; \quad
$\frac{9 \times 6}{2} = 27$; \quad
$\frac{5.5 \times 2}{2} = 5.5$ m$^2$.
\end{solution}

\begin{solution}{exo:g7:areas:4}
Disk: $\pi \times 5^2 = 25\pi \approx 79$ cm$^2$. Half-disk of \omterm{def:g3:shapes:shapes}{radius}
$10$: $\frac{\pi \times 10^2}{2} = 50\pi \approx 157$ cm$^2$.
\end{solution}

\begin{solution}{exo:g7:areas:5}
\omterm{def:g6:measure:perimeter}{Perimeter}: $2\pi \times 3 = 6\pi \approx 18.8$ cm (a length, in cm).
\omterm{def:g6:measure:area}{Area}: $\pi \times 3^2 = 9\pi \approx 28.3$ cm$^2$ (a surface, in
cm$^2$).
\end{solution}

\begin{solution}{exo:g7:areas:6}
\omterm{def:g7:areas:prism}{Prism}: $24 \times 10 = 240$ cm$^3$. \omterm{def:g7:areas:prism}{Cylinder}:
$\pi \times 3^2 \times 8 = 72\pi \approx 226$ cm$^3$.
\end{solution}

\begin{solution}{exo:g7:areas:7}
$\frac{8 \times h}{2} = 28$, so $4h = 28$ and $h = 7$ cm.
\end{solution}

\begin{solution}{exo:g7:areas:8}
The three \omterm{def:g2:mult:def}{products} agree because each one computes the \emph{same}
quantity --- the \omterm{def:g6:measure:area}{area} of the triangle. This gives a practical check
of constructions, and a way of computing a height from another
base--height pair (used in \cref{exo:g7:areas:12}).
\end{solution}

\begin{solution}{exo:g7:areas:9}
The two triangles have \omterm{def:g6:measure:area}{areas} $\frac{30 \times 20}{2} = 300$ m$^2$
and $\frac{18 \times 20}{2} = 180$ m$^2$: total $480$ m$^2$. General
formula suggested: for \omterm{def:g6:lines:perp}{parallel} sides $a$ and $b$ at distance $h$,
\[
A = \frac{(a + b) \times h}{2}
\]
--- the trapezoid formula (here $\frac{(30+18)\times 20}{2} = 480$).
\end{solution}

\begin{solution}{exo:g7:areas:10}
Water \omterm{def:g6:measure:volume}{volume}: $\pi \times 6^2 \times 15 = 540\pi \approx 1\,696$
cm$^3$. \omterm{def:g5:solids:def}{Box} base: $18 \times 12 = 216$ cm$^2$. Height reached:
$\frac{540\pi}{216} = 2.5\pi \approx 7.9$ cm.
\end{solution}

\begin{solution}{exo:g7:areas:11}
Radii $15$ and $20$ cm. \omterm{def:g6:measure:area}{Areas}: $225\pi \approx 707$ cm$^2$ and
$400\pi \approx 1257$ cm$^2$. Price per $100$ cm$^2$:
$\frac{9}{7.07} \approx 1.27$ euros and
$\frac{14}{12.57} \approx 1.11$ euros. The large pizza is the better
deal --- \omterm{def:g6:measure:area}{areas} grow with the \emph{square} of the \omterm{def:g4:geometry:circle}{diameter}.
\end{solution}

\begin{solution}{exo:g7:areas:12}
With the legs: $A = \frac{6 \times 8}{2} = 24$ cm$^2$. With the
hypotenuse as base: $24 = \frac{10 \times h}{2} = 5h$, so
$h = 4.8$ cm.
\end{solution}

\begin{solution}{pb:g7:areas:1}
\textbf{1.} Square: $8 \times 8 = 64$. Rectangle:
$5 \times 13 = 65$. The same four pieces appear to cover $64$
square units in one figure and $65$ in the other: one unit of
\omterm{def:g6:measure:area}{area} has been created from nothing --- supposedly.

\textbf{2.} Each triangle: $\frac{8 \times 3}{2} = 12$. Each
trapezoid: $\frac{(5 + 3) \times 5}{2} = 20$.

\textbf{3.} $12 + 12 + 20 + 20 = 64$: the pieces total $64$,
so they can fill the square exactly --- and can \emph{not} fill
the $65$-unit rectangle. The rectangle picture must contain a
hole.

\textbf{4.} Same steepness would mean $3$ across $8$
\omterm{def:g6:propdata:def}{proportional} to $2$ across $5$; cross \omterm{def:g2:mult:def}{products}:
$3 \times 5 = 15$ and $2 \times 8 = 16$. Not equal: the
hypotenuse and the slanted edge do \emph{not} line up. The
``diagonal'' of the rectangle is a lie --- two slightly
different slopes meeting at a shallow angle.

\textbf{5.} Along the false diagonal the four pieces leave a
long, extremely thin gap --- a sliver in the shape of a very
flat \omterm{def:g7:central:parallelogram}{parallelogram} --- whose \omterm{def:g6:measure:area}{area} is exactly the missing
$65 - 64 = 1$ square unit. Stretched along a diagonal about
$13$ units long, its average width is about
$1 \div 13 \approx 0.08$ units: on a real drawing, thinner than
the pencil line that hides it.

\textbf{6.} Each triangle has base $6$ cm and height $4$ cm ---
the distance between the \omterm{def:g6:lines:perp}{parallels}, wherever the apex sits ---
so each \omterm{def:g6:measure:area}{area} is $\frac{6 \times 4}{2} = 12$ cm$^2$
(\cref{thm:g7:areas:triangle}). Sliding the apex along the
\omterm{def:g6:lines:perp}{parallel} changes the shape, never the base or the height:
equal \omterm{def:g6:measure:area}{areas} forever.

\textbf{7.} The diagonal cuts the trapezoid into a triangle
with base $B$ and one with base $b$, both of height $h$ (the
distance between the \omterm{def:g6:lines:perp}{parallel} sides):
\[
A = \frac{B \times h}{2} + \frac{b \times h}{2}
= \frac{(B + b) \times h}{2} .
\]

\textbf{8.} $A = \frac{(9 + 5) \times 4}{2} = \frac{56}{2} =
28$ cm$^2$.

\textbf{9.} The average of the \omterm{def:g6:lines:perp}{parallel} sides is
$\frac{B + b}{2}$, so (average) $\times h =
\frac{(B+b) \times h}{2}$: the same formula, factored
differently. Check: average $= \frac{9+5}{2} = 7$, and
$7 \times 4 = 28$ cm$^2$.

\textbf{10.} The slanted stack contains exactly the same cards
--- the same layers, each of the same \omterm{def:g6:measure:area}{area} --- so the same
\omterm{def:g6:measure:volume}{volume}. In $V = B \times h$, the height must be measured
\emph{\omterm{def:g6:lines:perp}{perpendicular}} to the base (the vertical height of the
stack), not along the slant: exactly as the \omterm{def:g7:central:parallelogram}{parallelogram}'s
\omterm{def:g6:measure:area}{area} wants the \omterm{def:g6:lines:perp}{perpendicular} height, not the slanted side
(\cref{exo:g7:areas:2}), one dimension up.

\textbf{11.} Ring $= \pi \times 5^2 - \pi \times 3^2 = 25\pi -
9\pi = 16\pi \approx 50$ m$^2$ (\cref{thm:g7:areas:disk}).

\textbf{12.} A disk of \omterm{def:g6:measure:area}{area} $16\pi$ has \omterm{def:g3:shapes:shapes}{radius} $4$, since
$\pi \times 4^2 = 16\pi$. The radii $3$, $4$, $5$: the most
famous triple in mathematics, starring in
\cref{ch:g8:pythagoras}.

\textbf{13.} One $40$ cm pizza: \omterm{def:g3:shapes:shapes}{radius} $20$,
\omterm{def:g6:measure:area}{area} $\pi \times 400 \approx 1\,257$ cm$^2$. Two $28$ cm
pizzas: \omterm{def:g3:shapes:shapes}{radius} $14$, together
$2 \times \pi \times 196 = 392\pi \approx 1\,232$ cm$^2$. The
single large pizza is \emph{more} pizza than the two mediums.

\textbf{14.} Scaling the \omterm{def:g3:shapes:shapes}{radius} by $1.4$ \omterm{def:g7:prop:scale}{scales} the \omterm{def:g6:measure:area}{area} by
$1.4^2 = 1.96$ --- almost the double. The exact factor is the
number whose square is $2$, about $1.414$: the diagonal number
of \cref{ex:g8:pythagoras:diagonal}. (Rule of thumb: \omterm{def:g3:division:half}{half} again
as wide is twice the pizza, nearly.)

\textbf{15.} $13 \times 13 = 169$, but $8 \times 21 = 168$:
this time one square unit \emph{disappears} --- the pieces
overlap in a thin sliver instead of leaving a gap. With
consecutive terms $3, 5, 8, 13, 21$ of the
Hemachandra--Fibonacci sequence, the \omterm{def:g2:mult:def}{products} flip between one
more and one less than the square ($5 \times 13 = 65 =
8^2 + 1$, then $8 \times 21 = 168 = 13^2 - 1$): the magician
alternately ``gains'' and ``loses'' a unit. \omterm{def:g6:measure:area}{Areas} never lie:
pieces totalling $64$ cover $64$, wherever they slide --- every
missing or extra unit is hiding in a sliver, waiting for a slope
check to expose it.
\end{solution}
