\chapter{Probability and Sampling}\label{ch:g10:proba}

\omterm{def:g10:proba:distribution}{Probability} assigns numbers to chance. Starting from experiments with
equally likely outcomes --- dice, coins, cards --- this chapter builds the
basic rules for computing probabilities of \omterm{def:g10:proba:events}{events}, introduces \omterm{met:g10:proba:tree}{tree
diagrams} for two-step experiments, and ends with the link between
probabilities and observed frequencies. The theory continues in
\cref{ch:g11:prob}.

\section{Random experiments and events}

\begin{definition}[Outcomes and events]\label{def:g10:proba:events}
A \emph{random experiment}\index{random experiment} is an experiment whose
result cannot be predicted (rolling a die). The set of all possible
results is the \emph{sample space}\index{sample space}
$\Omega = \{\omega_1, \dots, \omega_n\}$; its elements are the
\emph{outcomes}. An \emph{event}\index{event} is a set of outcomes; an
event \emph{occurs} when the result of the experiment belongs to it.
\end{definition}

\begin{definition}[Probability distribution]\label{def:g10:proba:distribution}
Assigning to each outcome $\omega_i$ a number $p_i \geq 0$, with
$p_1 + \dots + p_n = 1$, defines a \emph{probability}\index{probability}:
the probability $\P(A)$ of an \omterm{def:g10:proba:events}{event} $A$ is the sum of the $p_i$ of the
outcomes in $A$. In particular $\P(\Omega) = 1$ and the impossible \omterm{def:g10:proba:events}{event}
$\varnothing$ has probability $0$.
\end{definition}

\begin{proposition}[Equally likely outcomes]\label{prop:g10:proba:uniform}
If the $n$ outcomes are equally likely (each $p_i = \frac1n$), then for
every \omterm{def:g10:proba:events}{event} $A$:
\[
\P(A) = \frac{\text{number of outcomes in } A}{n}
= \frac{\text{favorable cases}}{\text{possible cases}} .
\]
\end{proposition}

\begin{proof}
$A$ contains some number $k$ of outcomes, each of \omterm{def:g10:proba:distribution}{probability} $\frac1n$,
so $\P(A) = \frac kn$.
\end{proof}

\begin{example}\label{ex:g10:proba:die}
Roll a fair die: $\Omega = \{1, 2, 3, 4, 5, 6\}$, each outcome having
\omterm{def:g10:proba:distribution}{probability} $\frac16$. The \omterm{def:g10:proba:events}{event} $A$: ``the result is even'' is
$A = \{2, 4, 6\}$, so $\P(A) = \frac36 = \frac12$. The \omterm{def:g10:proba:events}{event} $B$: ``the
result is at least $5$'' is $B = \{5, 6\}$, so $\P(B) = \frac26 =
\frac13$.
\end{example}

\section{Combining events}

\begin{definition}[Intersection, union, complement]\label{def:g10:proba:operations}
Let $A$ and $B$ be \omterm{def:g10:proba:events}{events}.
\begin{itemize}
  \item $A \cap B$ (``$A$ and $B$'') occurs when both occur;
  \item $A \cup B$ (``$A$ or $B$'') occurs when at least one occurs;
  \item the \emph{complement}\index{complement} $\bar A$ (``not $A$'')
        occurs exactly when $A$ does not;
  \item $A$ and $B$ are \emph{incompatible}\index{incompatible} when they cannot occur
        together: $A \cap B = \varnothing$.
\end{itemize}
\end{definition}

\begin{theorem}[Addition rules]\label{thm:g10:proba:addition}
For all \omterm{def:g10:proba:events}{events} $A$ and $B$:
\[
\P(A \cup B) = \P(A) + \P(B) - \P(A \cap B),
\qquad
\P(\bar A) = 1 - \P(A) .
\]
For \omterm{def:g10:proba:operations}{incompatible} \omterm{def:g10:proba:events}{events}, the first rule simplifies to
$\P(A \cup B) = \P(A) + \P(B)$.
\end{theorem}

\begin{proof}
Summing $\P(A) + \P(B)$ adds the \omterm{def:g10:proba:distribution}{probability} of every outcome of
$A \cup B$ once, \emph{except} those of $A \cap B$, which are counted
twice --- once in $A$ and once in $B$. Subtracting $\P(A \cap B)$ repairs
the double counting. For the \omterm{def:g10:proba:operations}{complement}: $A$ and $\bar A$ are \omterm{def:g10:proba:operations}{incompatible}
and $A \cup \bar A = \Omega$, so $\P(A) + \P(\bar A) = \P(\Omega) = 1$.
\end{proof}

\begin{omfigure}
\begin{tikzpicture}[scale=0.9]
  \draw[omExo, thick, rounded corners] (-3.1,-1.7) rectangle (3.4,1.7);
  \node[omExo, font=\small] at (3.0,1.45) {$\Omega$};
  \begin{scope}
    \clip (-0.8,0) circle (1.25);
    \fill[omProp!25] (0.8,0) circle (1.25);
  \end{scope}
  \draw[omDef, thick] (-0.8,0) circle (1.25);
  \draw[omThm, thick] (0.8,0) circle (1.25);
  \node[omDef, font=\small] at (-1.6,0.9) {$A$};
  \node[omThm, font=\small] at (1.7,0.9) {$B$};
  \node[font=\small] at (0,0) {$A \cap B$};
\end{tikzpicture}

{\small Two overlapping \omterm{def:g10:proba:events}{events}: in $\P(A) + \P(B)$, the shaded
\omterm{def:g10:numbers:interunion}{intersection} $A \cap B$ is counted twice, which explains the subtraction
in the addition rule.}
\end{omfigure}

\begin{example}\label{ex:g10:proba:cards}
Draw one card from a standard $52$-card deck. Let $A$: ``the card is a
heart'' ($13$ cards) and $B$: ``the card is a king'' ($4$ cards). Then
$A \cap B$ is ``the king of hearts'' ($1$ card), and
\[
\P(A \cup B) = \frac{13}{52} + \frac{4}{52} - \frac{1}{52} = \frac{16}{52}
= \frac{4}{13}.
\]
The \omterm{def:g10:proba:operations}{complement} rule is often the fastest route: the \omterm{def:g10:proba:distribution}{probability} that the
card is \emph{not} a heart is $1 - \frac{13}{52} = \frac34$.
\end{example}

\section{Tree diagrams}

\begin{method}[Two-step experiments]\label{met:g10:proba:tree}
For an experiment performed in two steps:
\begin{enumerate}
  \item draw a \emph{tree}\index{tree diagram}: one branch per possible
        result of the first step, then from each of these, one branch per
        result of the second step, writing the \omterm{def:g10:proba:distribution}{probability} on each branch
        (the branches leaving any node must sum to $1$);
  \item the \omterm{def:g10:proba:distribution}{probability} of a \emph{path} (a leaf of the \omterm{met:g10:proba:tree}{tree}) is the
        product of the probabilities along its branches;
  \item the \omterm{def:g10:proba:distribution}{probability} of an \omterm{def:g10:proba:events}{event} is the sum of the probabilities of
        the paths that realize it.
\end{enumerate}
\end{method}

\begin{example}\label{ex:g10:proba:tree}
An urn contains $3$ red balls and $2$ blue balls. Draw one ball, put it
back, draw again. Each draw gives red (R) with \omterm{def:g10:proba:distribution}{probability} $\frac35$ and
blue (B) with \omterm{def:g10:proba:distribution}{probability} $\frac25$.

\begin{omfigure}
\begin{tikzpicture}[scale=1.0]
  \coordinate (root) at (0,0);
  \coordinate (R) at (2.3,1.1);
  \coordinate (B) at (2.3,-1.1);
  \coordinate (RR) at (5.0,1.7);
  \coordinate (RB) at (5.0,0.5);
  \coordinate (BR) at (5.0,-0.5);
  \coordinate (BB) at (5.0,-1.7);
  \draw (root) -- (R) node[midway, above, font=\small, sloped] {$\frac35$};
  \draw (root) -- (B) node[midway, below, font=\small, sloped] {$\frac25$};
  \draw (R) -- (RR) node[midway, above, font=\small, sloped] {$\frac35$};
  \draw (R) -- (RB) node[midway, below, font=\small, sloped] {$\frac25$};
  \draw (B) -- (BR) node[midway, above, font=\small, sloped] {$\frac35$};
  \draw (B) -- (BB) node[midway, below, font=\small, sloped] {$\frac25$};
  \node[omThm, font=\small] at (R) [above left=-2pt] {R};
  \node[omDef, font=\small] at (B) [below left=-2pt] {B};
  \node[right, font=\small] at (RR)
    {RR:\ $\frac35 \times \frac35 = \frac{9}{25}$};
  \node[right, font=\small] at (RB)
    {RB:\ $\frac35 \times \frac25 = \frac{6}{25}$};
  \node[right, font=\small] at (BR)
    {BR:\ $\frac25 \times \frac35 = \frac{6}{25}$};
  \node[right, font=\small] at (BB)
    {BB:\ $\frac25 \times \frac25 = \frac{4}{25}$};
\end{tikzpicture}

{\small The \omterm{met:g10:proba:tree}{tree} of two draws with replacement. The four path
probabilities sum to $1$.}
\end{omfigure}

The \omterm{def:g10:proba:distribution}{probability} of getting two balls of the same color is
$\P(\text{RR}) + \P(\text{BB}) = \frac{9}{25} + \frac{4}{25} =
\frac{13}{25}$; the \omterm{def:g10:proba:distribution}{probability} of at least one blue is
$1 - \P(\text{RR}) = 1 - \frac{9}{25} = \frac{16}{25}$ (\omterm{def:g10:proba:operations}{complement} rule).
\end{example}

\begin{remark}
Without replacement, the probabilities of the second draw depend on the
first: with the urn above, after drawing a red ball, only $2$ red and $2$
blue remain, so the second branch carries $\frac24$ and $\frac24$. The
\omterm{met:g10:proba:tree}{tree} method works unchanged; only the numbers on the second-level branches
change.
\end{remark}

\section{Frequencies and sampling}

When a \omterm{def:g10:proba:events}{random experiment} is repeated many times, the observed \omterm{def:g10:stats:series}{frequency}
of an \omterm{def:g10:proba:events}{event} approaches its \omterm{def:g10:proba:distribution}{probability} --- this is why \omterm{def:g10:proba:distribution}{probability} is
useful for describing the real world.

\begin{definition}[Sample]\label{def:g10:proba:sample}
A \emph{sample}\index{sample} of size $n$ is the result of repeating the
same experiment $n$ times independently. The \emph{observed \omterm{def:g10:stats:series}{frequency}} of
an \omterm{def:g10:proba:events}{event} in the sample is the number of occurrences divided by $n$.
\end{definition}

\begin{example}\label{ex:g10:proba:sampling}
A fair coin gives heads with \omterm{def:g10:proba:distribution}{probability} $0.5$. Tossing it $100$ times
rarely gives exactly $50$ heads: \omterm{def:g10:proba:sample}{samples} \emph{fluctuate}. Typical runs of
$100$ tosses give frequencies like $0.46$, $0.53$, $0.49$ --- close to
$0.5$, but not equal. Larger \omterm{def:g10:proba:sample}{samples} fluctuate less: with $10\,000$
tosses, observed frequencies are usually within about $0.01$ of $0.5$.
\end{example}

\begin{remark}[Fluctuation interval]
A useful rule of thumb, justified in \cref{ch:g11:binom}: for a \omterm{def:g10:proba:sample}{sample} of
size $n$ of an \omterm{def:g10:proba:events}{event} of \omterm{def:g10:proba:distribution}{probability} $p$ (with $n$ reasonably large), the
observed \omterm{def:g10:stats:series}{frequency} lies in the \omterm{def:g10:numbers:interval}{interval}
\[
\intcc{p - \frac{1}{\sqrt n}}{\,p + \frac{1}{\sqrt n}}
\]
in roughly $95\%$ of \omterm{def:g10:proba:sample}{samples}. If an observed \omterm{def:g10:stats:series}{frequency} falls far outside
this \omterm{def:g10:numbers:interval}{interval}, doubt the value of $p$: this is the starting point of
statistical testing.
\end{remark}

\begin{example}\label{ex:g10:proba:test}
A factory claims that only $10\%$ of its parts are defective
($p = 0.1$). In a \omterm{def:g10:proba:sample}{sample} of $n = 400$ parts, $64$ are defective: observed
\omterm{def:g10:stats:series}{frequency} $\frac{64}{400} = 0.16$. The fluctuation \omterm{def:g10:numbers:interval}{interval} is
$\intcc{0.1 - \frac{1}{20}}{0.1 + \frac{1}{20}} = \intcc{0.05}{0.15}$,
and $0.16$ falls outside it: the \omterm{def:g10:proba:sample}{sample} casts serious doubt on the claim.
\end{example}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:g10:proba:1}
A fair die is rolled. Compute the \omterm{def:g10:proba:distribution}{probability} of each \omterm{def:g10:proba:events}{event}: ``getting a
$6$''; ``getting an odd number''; ``getting at most $4$''; ``getting a
$7$''.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:proba:2}
A bag contains $5$ green, $3$ yellow and $2$ black marbles; one marble is
drawn at random. Compute the \omterm{def:g10:proba:distribution}{probability} that it is green; that it is not
black; that it is green or yellow.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:proba:3}
In a class of $30$ students, $18$ study Spanish, $10$ study German, and
$4$ study both. A student is picked at random. Using the addition rule,
compute the \omterm{def:g10:proba:distribution}{probability} that this student studies at least one of the two
languages, then the \omterm{def:g10:proba:distribution}{probability} that they study neither.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g10:proba:4}
A fair coin is tossed twice. Draw the \omterm{met:g10:proba:tree}{tree} of the experiment and compute
the \omterm{def:g10:proba:distribution}{probability} of getting two heads; exactly one head; at least one
head.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:proba:5}
An urn contains $4$ red and $6$ white balls. Two balls are drawn one after
the other \emph{without} replacement. Draw the \omterm{met:g10:proba:tree}{tree} with the correct
probabilities on the second level, and compute the \omterm{def:g10:proba:distribution}{probability} of drawing
two red balls, then of drawing two balls of different colors.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:proba:6}
Two fair dice are rolled and their sum is recorded.
\begin{enumerate}
  \item Explain why the \omterm{def:g10:proba:events}{sample space} can be taken as the $36$ equally
        likely pairs $(1,1), (1,2), \dots, (6,6)$.
  \item Compute the \omterm{def:g10:proba:distribution}{probability} that the sum is $7$; that it is $12$;
        that it is at most $4$.
  \item Which sum is the most likely?
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:proba:7}
A multiple-choice question offers $4$ answers, one correct. A student
answers $2$ independent such questions completely at random. Compute the
\omterm{def:g10:proba:distribution}{probability} of answering both correctly, exactly one correctly, none
correctly. Check that the three probabilities sum to $1$.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g10:proba:8}
A politician claims $50\%$ approval. In a poll of $n = 900$ randomly
chosen voters, $396$ approve.
\begin{enumerate}
  \item Compute the observed \omterm{def:g10:stats:series}{frequency} and the fluctuation \omterm{def:g10:numbers:interval}{interval}
        around $p = 0.5$ for $n = 900$.
  \item Does the poll cast doubt on the claim?
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g10:proba:9}
A game costs $2$ to play: you roll a fair die and win $6$ if you roll a
$6$, $3$ if you roll a $5$, nothing otherwise.
\begin{enumerate}
  \item Compute the \omterm{def:g10:proba:distribution}{probability} of each amount won ($6$, $3$, $0$).
  \item Over $600$ games, about how many of each outcome do you expect?
        Compute the total expected gain and loss: is the game worth
        playing?
\end{enumerate}
\end{exercise}
