\section*{Chapter \ref{ch:b3:banach} --- Banach Spaces and the
Fundamental Theorems}

\begin{solution}{exo:b3:banach:1}
(a) $\norm{Sx}_2 = \norm x_2$: $S$ is an isometry, $\vertiii S =
1$. For the backward shift: $\norm{S^*x}_2^2 = \sum_{n \geq
2}\abs{x_n}^2 \leq \norm x_2^2$, with equality for $x = e_2$:
$\vertiii{S^*} = 1$.

(b) $\norm{M_ax}_2^2 = \sum\abs{a_n}^2\abs{x_n}^2 \leq \norm
a_\infty^2\norm x_2^2$; testing $x = e_n$ gives $\vertiii{M_a}
\geq \abs{a_n}$ for every $n$: $\vertiii{M_a} = \norm a_\infty$.

(c) $\abs{\Lambda(f)} \leq \int_0^1\abs f \leq \norm f_\infty$:
$\norm\Lambda \leq 1$. For $\varepsilon > 0$ let $f_\varepsilon$
be $1$ on $[0, \frac12 - \varepsilon]$, $-1$ on $[\frac12 +
\varepsilon, 1]$, affine between: $\norm{f_\varepsilon}_\infty =
1$ and $\Lambda(f_\varepsilon) \geq 1 - 2\varepsilon$:
$\norm\Lambda = 1$. Not attained: $\Lambda(f) = 1$ with $\norm
f_\infty \leq 1$ forces $\int_0^{1/2}f = \frac12$ and
$\int_{1/2}^1 f = -\frac12$, i.e.\ (continuity, $\abs f \leq 1$)
$f \equiv 1$ on $[0, \frac12]$ and $f \equiv -1$ on $[\frac12,
1]$: contradiction at $\frac12$.
\end{solution}

\begin{solution}{exo:b3:banach:2}
Write $S = T\bigl(I - T^{-1}(T - S)\bigr)$ with
$\vertiii{T^{-1}(T-S)} \leq \vertiii{T^{-1}}\,\vertiii{T - S} =
\theta < 1$: by \cref{prop:b3:banach:neumann}, $S$ is invertible
with $S^{-1} = \sum_{n\geq0}\bigl(T^{-1}(T -
S)\bigr)^nT^{-1}$, whence
\[
\vertiii{S^{-1} - T^{-1}} \leq
\sum_{n\geq1}\theta^n\,\vertiii{T^{-1}}
= \frac{\vertiii{T^{-1}}^2\,\vertiii{T - S}}{1 - \theta}.
\]
For the linear system: $x_T = T^{-1}b$ and $x_S = S^{-1}b$
differ by at most that bound times $\norm b$ --- small
perturbations of an invertible system remain uniquely solvable,
with Lipschitz dependence of the solution on the operator.
\end{solution}

\begin{solution}{exo:b3:banach:3}
(a) $\ell^1$: let $(x^{(k)})$ be Cauchy. Each coordinate is
Cauchy ($\abs{x^{(k)}_n - x^{(l)}_n} \leq \norm{x^{(k)} -
x^{(l)}}_1$): let $x_n = \lim_kx^{(k)}_n$. Given $\varepsilon$,
for $k, l \geq K$: $\sum_{n \leq N}\abs{x^{(k)}_n - x^{(l)}_n}
\leq \varepsilon$ for every $N$; let $l \to \infty$, then $N \to
\infty$: $\norm{x^{(k)} - x}_1 \leq \varepsilon$, and $x =
x^{(k)} - (x^{(k)} - x) \in \ell^1$. $\ell^\infty$: Cauchy for
$\norm\cdot_\infty$ is uniformly Cauchy: converges uniformly to
a bounded sequence. $c_0$ is closed in $\ell^\infty$: if
$x^{(k)} \to x$ uniformly with $x^{(k)}_n \to_n 0$, then
$\abs{x_n} \leq \norm{x - x^{(k)}}_\infty + \abs{x^{(k)}_n}$
gives $\limsup_n\abs{x_n} \leq \varepsilon$: $x \in c_0$; a
closed subspace of a Banach space is Banach. Finite sequences:
their closure contains every $x \in c_0$ (truncations converge:
$\sup_{n>N}\abs{x_n} \to 0$) and is contained in the closed
$c_0$.

(b) By homogeneity assume $\norm x_p = 1$: then $\abs{x_n} \leq
1$ for all $n$, so $\abs{x_n}^q \leq \abs{x_n}^p$ and $\norm
x_q \leq 1 = \norm x_p$; for $q = \infty$, $\abs{x_n} \leq
\norm x_p$ directly. Strictness: $x_n = n^{-\alpha}$ with
$\frac1q < \alpha \leq \frac1p$ lies in $\ell^q \setminus
\ell^p$ (Riemann series).
\end{solution}

\begin{solution}{exo:b3:banach:4}
($\leq$) If $\norm f \leq 1$ and $f\restriction_F = 0$: for
every $y \in F$, $\abs{f(x)} = \abs{f(x - y)} \leq \norm{x -
y}$; take the infimum. ($\geq$, attained)
\cref{cor:b3:banach:hbnormed}(3) produces $f$ with
$f\restriction_F = 0$, $\norm f \leq 1$, $f(x) = d(x, F)$: the
supremum is a maximum. Consequence: $F \subseteq \bigcap\{\ker
f : f\restriction_F = 0\}$ trivially, and a point $x \notin F$
is excluded from the intersection by the functional above
($f(x) = d(x,F) > 0$, $F$ being closed).
\end{solution}

\begin{solution}{exo:b3:banach:5}
For $y \in \ell^\infty$: $\abs{\Lambda_y(x)} =
\abs{\sum x_ny_n} \leq \norm y_\infty\norm x_1$, so
$\norm{\Lambda_y} \leq \norm y_\infty$; testing on $e_n$:
$\abs{y_n} = \abs{\Lambda_y(e_n)} \leq \norm{\Lambda_y}$:
equality. Conversely, given $\Lambda \in (\ell^1)'$, set $y_n =
\Lambda(e_n)$: $\abs{y_n} \leq \norm\Lambda$, so $y \in
\ell^\infty$; $\Lambda$ and $\Lambda_y$ agree on finite
sequences, which are dense in $\ell^1$ ($\norm{x - \sum_{n\leq
N}x_ne_n}_1 = \sum_{n>N}\abs{x_n} \to 0$): $\Lambda =
\Lambda_y$. The map $y \mapsto \Lambda_y$ is linear, isometric,
onto. For $(\ell^\infty)'$ the same start produces a sequence
$y_n = \Lambda(e_n)$, but finite sequences are \emph{not} dense
in $\ell^\infty$
(the constant sequence $\mathbf 1$ is at distance $1$ from all
of them), so $\Lambda$ is not determined by the $y_n$ --- and
indeed $(\ell^\infty)' \neq \ell^1$
(\cref{pb:b3:banach:1}).
\end{solution}

\begin{solution}{exo:b3:banach:6}
For each fixed $x$, $y \mapsto B(x, y)$ is continuous linear:
$\sup_{\norm y \leq 1}\norm{B(x,y)} < \infty$. So the family
$\{B(\cdot, y) : \norm y \leq 1\} \subseteq \mathcal L(E, G)$
(each member continuous, by continuity in $x$) is pointwise
bounded on the Banach space $E$: Banach--Steinhaus
(\cref{thm:b3:banach:banachsteinhaus}) yields $C$ with
$\norm{B(x,y)} \leq C\norm x$ for all $\norm y \leq 1$;
homogeneity in $y$ finishes: $\norm{B(x,y)} \leq
C\norm x\norm y$.
\end{solution}

\begin{solution}{exo:b3:banach:7}
(a) $\norm f_1 \leq \norm f_\infty$: the identity is bounded,
and bijective. Its inverse is unbounded: $f_n(x) = x^n$ has
$\norm{f_n}_1 = \frac1{n+1} \to 0$ but $\norm{f_n}_\infty = 1$.
No contradiction with \cref{cor:b3:banach:equivnorms}: $(\mathcal
C(\intcc01), \norm\cdot_1)$ is \emph{not} complete
(\cref{exo:b3:complete:1}); the corollary requires completeness
on both sides.

(b) On the space $E_0$ of finite sequences (dense in $\ell^2$),
$\varphi(x) = \sum_n n\,x_n$ is linear and unbounded
($\varphi(e_n) = n$ with $\norm{e_n}_2 = 1$). The closed graph
theorem does not apply: $E_0$ is not complete --- and $\varphi$
has no continuous extension to $\ell^2$, illustrating that
density without uniform continuity is powerless
(\cref{thm:b3:complete:extension}).
\end{solution}

\begin{solution}{exo:b3:banach:8}
We verify the closed-graph hypothesis. Let $x_k \to x$ and
$Tx_k \to z$ in $\ell^2$. For every $y$:
\[
\langle z, y\rangle = \lim_k\langle Tx_k, y\rangle
= \lim_k \langle x_k, Ty\rangle = \langle x, Ty\rangle
= \langle Tx, y\rangle,
\]
using continuity of the inner product in each slot
(Cauchy--Schwarz) and symmetry twice. So $z - Tx$ is orthogonal
to every $y$, in particular to itself: $z = Tx$. The graph is
closed and $\ell^2$ is Banach: $T$ is bounded
(\cref{thm:b3:banach:closedgraph}). Hence a symmetric operator
defined on \emph{all} of $\ell^2$ is automatically bounded;
genuinely unbounded symmetric operators (position, momentum,
Hamiltonians) must live on proper dense subspaces.
\end{solution}

\begin{solution}{exo:b3:banach:9}
First, $\norm{\Lambda_n} = \sum_i\abs{w_{i,n}}$: $\leq$ is the
triangle inequality; $\geq$ by testing a piecewise linear $f$
with $\norm f_\infty \leq 1$ and $f(x_{i,n}) =
\operatorname{sign}(w_{i,n})$ (interpolate linearly between the
finitely many nodes; where nodes coincide the signs agree).

($\Rightarrow$) Pointwise convergence at every $f$ implies (i),
and pointwise boundedness, so Banach--Steinhaus
(\cref{thm:b3:banach:banachsteinhaus}) on the Banach $\mathcal
C(\intcc01)$ gives (ii).

($\Leftarrow$) Let $M = \sup_n\norm{\Lambda_n} + 1$. Given $f$
and $\varepsilon$, choose a polynomial $P$ with $\norm{f -
P}_\infty < \varepsilon/(2M)$
(\cref{cor:b3:complete:weierstrass}); then
\[
\Bigl|\Lambda_n f - \int_0^1 f\Bigr|
\leq \abs{\Lambda_n(f - P)} + \Bigl|\Lambda_nP -
\int_0^1P\Bigr| + \Bigl|\int_0^1(P - f)\Bigr|
\leq \varepsilon + \Bigl|\Lambda_nP - \int_0^1P\Bigr| \to
\varepsilon .
\]
Positive weights, exactness on constants: $\sum_i\abs{w_{i,n}}
= \sum_iw_{i,n} = \Lambda_n(\mathbf 1) = \int_0^1 1 = 1$ for
rules exact on constants --- (ii) holds with constant $1$.
\end{solution}

\begin{solution}{exo:b3:banach:10}
Linearity of $J$ is formal; $\norm{J(x)} = \sup_{\norm f \leq
1}\abs{f(x)} = \norm x$ by
\cref{cor:b3:banach:hbnormed}(2). If $\dim E = n$: $\dim E' =
n$ (a basis gives coordinate functionals), so $\dim E'' = n$,
and the injective (isometric) $J$ is onto. Separability: let
$(f_n)$ be dense in $E'$ and choose $\norm{x_n} = 1$ with
$\abs{f_n(x_n)} \geq \frac12\norm{f_n}$. Let $F =
\overline{\operatorname{Vect}}(x_n)$; if $F \neq E$, take $g
\in E'$, $g \neq 0$, vanishing on $F$
(\cref{cor:b3:banach:hbnormed}(3)); choose $f_{n_k} \to g$:
\[
\norm{f_{n_k} - g} \geq \abs{(f_{n_k} - g)(x_{n_k})}
= \abs{f_{n_k}(x_{n_k})} \geq \tfrac12\norm{f_{n_k}}
\geq \tfrac12\bigl(\norm g - \norm{f_{n_k} - g}\bigr),
\]
so $\norm{f_{n_k} - g} \geq \frac13\norm g > 0$: contradiction.
Hence $F = E$, and rational (or $\Q + \iu\Q$) combinations of
the $x_n$ form a countable dense set.
\end{solution}

\begin{solution}{exo:b3:banach:11}
(a) Well defined: $d(x, F)$ depends only on $\bar x$
(translating $x$ by $F$ does not change the distance).
Homogeneity and triangle inequality pass from $\norm\cdot$
through the infimum. Separation needs closedness: $\norm{\bar
x} = 0$ means $d(x, F) = 0$, i.e.\ $x \in \bar F = F$, i.e.\
$\bar x = 0$. $\vertiii\pi \leq 1$: $\norm{\bar x} \leq
\norm x$. Open ball onto open ball: if $\norm{\bar x} < 1$,
some representative has $\norm{x - y} < 1$; conversely
$\pi(B_E(0,1)) \subseteq B_{E/F}(0,1)$ by the norm
inequality --- so $\pi$ is open, the model case of the open
mapping theorem.

(b) Let $\sum_k\norm{\bar x_k} < \infty$ and lift with
$\norm{x_k} \leq \norm{\bar x_k} + 2^{-k}$: then
$\sum\norm{x_k} < \infty$, so $s = \sum_kx_k$ converges in
the Banach $E$ (\cref{exo:b3:complete:1}(b)), and continuity
of $\pi$ gives $\sum_k\bar x_k = \bar s$: every absolutely
convergent series of $E/F$ converges, which is equivalent to
completeness (same exercise).

(c) The map $\lambda(x) = \lim_nx_n$ is linear $c \to K$,
vanishes exactly on $c_0$, so it induces a linear bijection
$c/c_0 \to K$. Isometry: $d(x, c_0) = \abs{\lambda(x)}$ ---
$\leq$: subtract from $x$ the sequence $x - \lambda(x)\mathbf
1 \in c_0$, leaving $\lambda(x)\mathbf 1$ of norm
$\abs{\lambda(x)}$; $\geq$: for $y \in c_0$, $\norm{x -
y}_\infty \geq \limsup_n\abs{x_n - y_n} = \abs{\lambda(x)}$.
\end{solution}

\begin{solution}{exo:b3:banach:12}
(a) $W = \ker P$ is closed (preimage of $0$ under a
continuous map); $V = \operatorname{im}P = \ker(I - P)$
(indeed $Px = x$ iff $x \in \operatorname{im}P$, using $P^2 =
P$), closed likewise. Every $x$ splits as $Px + (x - Px)$
with $Px \in V$, $x - Px \in W$, and $V \cap W = 0$ ($x = Px
= 0$): $E = V \oplus W$.

(b) The graph argument: let $x_n \to x$ and $Px_n \to z$.
Then $z \in V$ ($V$ closed, $Px_n \in V$) and $x_n - Px_n \to
x - z \in W$ ($W$ closed). So $x = z + (x - z)$ with $z \in
V$, $x - z \in W$; by uniqueness of the decomposition, $z =
Px$. The graph of $P$ is closed, $E$ is Banach: $P$ is
bounded (\cref{thm:b3:banach:closedgraph}).

(c) (a) and (b) together are the equivalence. In a Hilbert
space every closed $V$ has the closed complement $V^\perp$
(\cref{ch:b3:hilbert}): every closed subspace is
complemented. In general Banach spaces this fails --- $c_0$
has no closed complement in $\ell^\infty$ (Phillips'
theorem, beyond our tools) --- so bounded projections are a
privilege, and the closed graph theorem is exactly the
bookkeeping that converts geometric splittings into bounded
operators.
\end{solution}

\begin{solution}{pb:b3:banach:1}
\textbf{1.} For $a, b > 0$: by concavity of $\log$,
$\log\bigl(\tfrac{a^p}p + \tfrac{b^q}q\bigr) \geq \tfrac1p\log
a^p + \tfrac1q\log b^q = \log(ab)$; exponentiate. (If $ab = 0$
the inequality is trivial.) Equality iff $a^p = b^q$.

\textbf{2.} We may assume $\norm x_p = \norm y_q = 1$
(homogeneity; zero cases trivial). Then
\[
\abs{\langle x, y\rangle} \leq \sum_n\abs{x_n}\abs{y_n}
\leq \sum_n\Bigl(\frac{\abs{x_n}^p}p +
\frac{\abs{y_n}^q}q\Bigr) = \frac1p + \frac1q = 1 =
\norm x_p\norm y_q .
\]
Equality requires $\abs{x_n}^p = \abs{y_n}^q$ for all $n$
(Young's equality case) and alignment of the phases of
$x_ny_n$.

\textbf{3.} $\abs{x_n + y_n}^p \leq \bigl(\abs{x_n} +
\abs{y_n}\bigr)\abs{x_n + y_n}^{p-1}$; summing and applying
H\"older ($p$ against $q$, noting $(p - 1)q = p$):
\[
\norm{x + y}_p^p \leq \bigl(\norm x_p + \norm
y_p\bigr)\,\Bigl(\sum_n\abs{x_n + y_n}^{p}\Bigr)^{1/q}
= \bigl(\norm x_p + \norm y_p\bigr)\,\norm{x+y}_p^{p/q};
\]
if $\norm{x + y}_p \neq 0$ (else trivial), divide by
$\norm{x+y}_p^{p/q}$ and use $p - \frac pq = 1$. (Finiteness of
$\norm{x+y}_p$ first: $\abs{x_n+y_n}^p \leq
2^p(\abs{x_n}^p + \abs{y_n}^p)$.)

\textbf{4.} Completeness: as for $\ell^1$
(\cref{exo:b3:banach:3}), coordinatewise limits plus the
uniform tail bound $\sum_{n\leq N}\abs{x^{(k)}_n - x^{(l)}_n}^p
\leq \varepsilon^p$, letting $l$ then $N$ tend to infinity.
Density of finite sequences: $\norm{x - \sum_{n\leq
N}x_ne_n}_p^p = \sum_{n > N}\abs{x_n}^p \to 0$.

\textbf{5.} H\"older gives $\abs{\Lambda_y(x)} \leq \norm
x_p\norm y_q$: $\norm{\Lambda_y} \leq \norm y_q$. Testing: let
$x^{(N)}_n = \abs{y_n}^{q-1}\overline{\operatorname{sign}}(y_n)$
for $n \leq N$, $0$ beyond (with $\operatorname{sign}$ the
unimodular phase, so that $x_ny_n = \abs{y_n}^q$). Then
$\Lambda_y(x^{(N)}) = \sum_{n\leq N}\abs{y_n}^q$ and
$\norm{x^{(N)}}_p = \bigl(\sum_{n\leq
N}\abs{y_n}^{q}\bigr)^{1/p}$ (as $(q-1)p = q$), so
\[
\norm{\Lambda_y} \geq \Bigl(\sum_{n\leq
N}\abs{y_n}^q\Bigr)^{1 - 1/p} \xrightarrow[N\to\infty]{}
\norm y_q .
\]

\textbf{6.} Set $y_n = \Lambda(e_n)$. With the same test
vectors, $\sum_{n \leq N}\abs{y_n}^q = \Lambda(x^{(N)}) \leq
\norm\Lambda\,\bigl(\sum_{n\leq N}\abs{y_n}^q\bigr)^{1/p}$,
whence $\bigl(\sum_{n\leq N}\abs{y_n}^q\bigr)^{1/q} \leq
\norm\Lambda$ for every $N$: $y \in \ell^q$, $\norm y_q \leq
\norm\Lambda$. The functionals $\Lambda$ and $\Lambda_y$ agree
on the dense finite sequences (question 4): $\Lambda =
\Lambda_y$. With question 5, $y \mapsto \Lambda_y$ is an
isometric isomorphism $\ell^q \cong (\ell^p)'$.

\textbf{7.} Let $\xi \in (\ell^p)''$. Composing with the
isometry $\ell^q \cong (\ell^p)'$ of question 6, $\xi$ defines
an element of $(\ell^q)'$, which (question 6 with $p, q$
swapped) is $\Lambda_z$ for a unique $z \in \ell^p$: for every
$y \in \ell^q$, $\xi(\Lambda_y) = \sum_nz_ny_n$. On the other
hand $J(z)(\Lambda_y) = \Lambda_y(z) = \sum_ny_nz_n$: the same
value. Since every element of $(\ell^p)'$ is some $\Lambda_y$,
$\xi = J(z)$: $J$ is onto --- $\ell^p$ is reflexive.

\textbf{8.} Finite sequences with rational (real and imaginary)
entries are countable and dense in $\ell^p$ ($p < \infty$) and
in $c_0$. In $\ell^\infty$: the family $\{\mathbf 1_A : A
\subseteq \N\}$ is uncountable with $\norm{\mathbf 1_A -
\mathbf 1_B}_\infty = 1$ for $A \neq B$; the balls $B(\mathbf
1_A, \frac12)$ are pairwise disjoint, and a dense set must meet
each: no countable dense set exists.

\textbf{9.} If $\ell^1$ were reflexive, then $(\ell^\infty)'
\cong ((\ell^1)')' = J(\ell^1)$ would be separable (isometric
image of the separable $\ell^1$); by
\cref{exo:b3:banach:10}, the separability of the dual
$(\ell^\infty)'$ would force $\ell^\infty$ separable ---
contradicting question 8. So $\ell^1$ is not reflexive (and
$(\ell^\infty)'$ is strictly larger than $\ell^1$, as question
11 makes concrete).

\textbf{10.} Homogeneity of $p$ is clear; subadditivity:
averages are linear, and $\limsup(u_n + v_n) \leq \limsup u_n +
\limsup v_n$. On $c$: the Ces\`aro means of a convergent
sequence converge to its limit, so $p(x) = \lim x =
\mathrm{LIM}(x)$ there; in particular $\mathrm{LIM} \leq p$ on
$c$. Hahn--Banach (\cref{thm:b3:banach:hahnbanach}) extends
$\mathrm{LIM}$ to $\ell^\infty_\R$ with $\mathrm{LIM} \leq p$
globally. Positivity: for $x \geq 0$, $-\mathrm{LIM}(x) =
\mathrm{LIM}(-x) \leq p(-x) = \limsup\text{avg}(-x) \leq 0$.
Shift-invariance: the averages of $x - Sx$ telescope to
$\frac{x_1 - x_{n+1}}n \to 0$, so $p(\pm(x - Sx)) = 0$ and
$\mathrm{LIM}(x - Sx) = 0$. Bounds: $\mathrm{LIM}(x) \leq p(x)
\leq \limsup x$ (averages lag behind sups), and applying this
to $-x$ gives the lower bound.

\textbf{11.} $e_n \in c_0$, so $\mathrm{LIM}(e_n) = 0$ for all
$n$. If $\mathrm{LIM} = \Lambda_y$ with $y \in \ell^1$, then
$y_n = \Lambda_y(e_n) = 0$ for all $n$: $\Lambda_y = 0$; but
$\mathrm{LIM}(\mathbf 1) = 1$. So $\mathrm{LIM} \in
(\ell^\infty)' \setminus \{\Lambda_y : y \in \ell^1\}$:
$(\ell^\infty)' \neq \ell^1$, again.

\textbf{12.} For $x = (0,1,0,1,\dots)$: $x + Sx = \mathbf 1$,
so $2\,\mathrm{LIM}(x) = \mathrm{LIM}(x) + \mathrm{LIM}(Sx) =
1$: $\mathrm{LIM}(x) = \frac12$. If $\varphi$ were a
shift-invariant multiplicative extension of the limit: $x\cdot
Sx = 0$ gives $\varphi(x)\varphi(Sx) = \varphi(x)^2 = 0$, so
$\varphi(x) = 0$; but $\varphi(x) + \varphi(Sx) =
\varphi(\mathbf 1) = 1$ gives $2\varphi(x) = 1$: contradiction.
Averaging and multiplication cannot coexist.

\textbf{13.} Let $\norm{x^{(k)}}_p \leq M$. Coordinates are
bounded by $M$: a diagonal extraction gives a subsequence
(still written $x^{(k)}$) with $x^{(k)}_n \to x_n$ for every
$n$. Then $x \in \ell^p$: $\sum_{n\leq N}\abs{x_n}^p =
\lim_k\sum_{n\leq N}\abs{x^{(k)}_n}^p \leq M^p$ for all $N$.
Weak convergence: for $y \in \ell^q$ and $N$ arbitrary,
\[
\abs{\Lambda_y(x^{(k)} - x)} \leq
\Bigl|\sum_{n \leq N}(x^{(k)}_n - x_n)y_n\Bigr|
+ 2M\Bigl(\sum_{n>N}\abs{y_n}^q\Bigr)^{1/q},
\]
where the first term tends to $0$ as $k \to \infty$ (finitely
many coordinates) and the second is small for $N$ large
(H\"older on the tail): $\Lambda_y(x^{(k)}) \to \Lambda_y(x)$
for every $y$ --- weak convergence, since every functional is a
$\Lambda_y$ (question 6). In $\ell^1$ this fails: consider
$(e_n)$, bounded. Any subsequence $(e_{n_k})$ converges
coordinatewise to $0$, so its only weak limit candidate is $0$;
but testing against $y \in \ell^\infty$ defined by $y_{n_k} =
(-1)^k$ (and $0$ elsewhere), $\Lambda_y(e_{n_k}) = (-1)^k$
diverges. No weakly convergent subsequence: weak sequential
compactness of balls characterizes the reflexive world.

\textbf{14.} For $y \in \ell^1$, $\Lambda_y(x) = \sum x_ny_n$
is defined on $c_0$ with $\abs{\Lambda_y(x)} \leq \norm
x_\infty\norm y_1$, and testing on $x^{(N)} =
(\operatorname{sign}y_1, \dots, \operatorname{sign}y_N, 0,
\dots) \in c_0$ gives $\Lambda_y(x^{(N)}) =
\sum_{n\leq N}\abs{y_n} \to \norm y_1$:
$\norm{\Lambda_y} = \norm y_1$. Conversely, for $\Lambda \in
(c_0)'$ put $y_n = \Lambda(e_n)$; the same tests give
$\sum_{n \leq N}\abs{y_n} = \Lambda(x^{(N)}) \leq
\norm\Lambda$, so $y \in \ell^1$, and $\Lambda = \Lambda_y$ on
the dense finite sequences, hence everywhere. The chain of
duals: $(c_0)' = \ell^1$, $(\ell^1)' = \ell^\infty$
(\cref{exo:b3:banach:10}), $(\ell^\infty)' \supsetneq \ell^1$
(questions 9--11): reflexivity fails at the very first step
--- $c_0'' = \ell^\infty \neq c_0$ --- and never recovers.

\textbf{15.} $J x^{(k)} \in E'' = (E')'$ is a family of
bounded functionals on the \emph{Banach} space $E'$ (duals
are complete, \cref{prop:b3:banach:llcomplete}); for each
$\Lambda \in E'$, the sequence $Jx^{(k)}(\Lambda) =
\Lambda(x^{(k)})$ converges, hence is bounded.
Banach--Steinhaus on $E'$ gives $\sup_k\norm{Jx^{(k)}} <
\infty$, and $J$ is isometric
(\cref{rem:b3:banach:bidual}): $\sup_k\norm{x^{(k)}} <
\infty$.

\textbf{16.} ($\Rightarrow$) Boundedness is question 15;
coordinates are the functionals $\Lambda_{e_n}$.
($\Leftarrow$) Let $M = \sup_k\norm{x^{(k)}}_p$, $y \in
\ell^q$, $\varepsilon > 0$; choose $N$ with
$\bigl(\sum_{n>N}\abs{y_n}^q\bigr)^{1/q} < \varepsilon$. Then
\[
\abs{\Lambda_y(x^{(k)} - x)} \leq
\sum_{n\leq N}\abs{x^{(k)}_n - x_n}\,\abs{y_n} +
\norm{x^{(k)} - x}_p\,\varepsilon
\leq \sum_{n\leq N}\abs{x^{(k)}_n - x_n}\abs{y_n} +
(M + \norm x_p)\,\varepsilon,
\]
and the finite sum tends to $0$: $\limsup \leq (M + \norm
x_p)\varepsilon$ for every $\varepsilon$. (That $x \in
\ell^p$ with $\norm x_p \leq M$ follows from Fatou-style
finite-section bounds: $\sum_{n \leq N}\abs{x_n}^p =
\lim_k\sum_{n\leq N}\abs{x^{(k)}_n}^p \leq M^p$.) For $e_k$
in $\ell^2$: bounded, coordinatewise $\to 0$, so $e_k
\rightharpoonup 0$, yet $\norm{e_k} = 1$: the unit of mass
escapes to infinite index, invisible to every fixed
functional.

\textbf{17.} Take $\Lambda$ with $\norm\Lambda = 1$ and
$\Lambda(x) = \norm x$ (\cref{cor:b3:banach:hbnormed}). Then
$\norm x = \lim\Lambda(x^{(k)}) \leq
\liminf\norm\Lambda\,\norm{x^{(k)}} =
\liminf\norm{x^{(k)}}$. (With $e_k \rightharpoonup 0$: $0
\leq \liminf 1$, and the inequality can be strict.)

\textbf{18.} In $\ell^2$, $\norm{x^{(k)} - x}_2^2 =
\norm{x^{(k)}}_2^2 - 2\operatorname{Re}\langle x^{(k)},
x\rangle + \norm x_2^2$ (real case: $-2\langle x^{(k)},
x\rangle$). Weak convergence applied to the functional
$\Lambda_x$ gives $\langle x^{(k)}, x\rangle \to \norm
x_2^2$, and the norms converge by hypothesis: the right side
tends to $\norm x^2 - 2\norm x^2 + \norm x^2 = 0$.
Counterexample without norm convergence: $e_k
\rightharpoonup 0$, $\norm{e_k - 0} = 1 \not\to 0$.

\textbf{19.} Coordinate convergence: apply the functionals
$\Lambda_{e_n} \in (\ell^1)' = \ell^\infty$ ($y = e_n$).
Construction: having chosen $k_{j-1}, N_{j-1}$, pick $k_j >
k_{j-1}$ so large that $\sum_{n \leq
N_{j-1}}\abs{x^{(k_j)}_n} < \frac\delta{20}$ (finitely many
coordinates, each $\to 0$), then $N_j > N_{j-1}$ so large
that the tail satisfies $\sum_{n > N_j}\abs{x^{(k_j)}_n} <
\frac\delta{20}$ (convergence of the series defining
$\norm{x^{(k_j)}}_1$). The block $B_j = \intoc{N_{j-1}}{N_j}$
then carries all but $\frac\delta{10}$ of the mass of
$x^{(k_j)}$.

\textbf{20.} With $y$ as defined ($\abs{y_n} \leq 1$
everywhere):
\[
\langle x^{(k_j)}, y\rangle
= \sum_{n \in B_j}\abs{x^{(k_j)}_n}
+ \sum_{n \notin B_j}x^{(k_j)}_ny_n
\geq \Bigl(\norm{x^{(k_j)}}_1 - \frac\delta{10}\Bigr) -
\frac\delta{10}
\geq \delta - \frac{2\delta}{10} = \frac{4\delta}5 > 0,
\]
the middle inequality because the mass off the block is at
most $\frac\delta{10}$ (question 19). But $y \in
\ell^\infty = (\ell^1)'$ and $x^{(k)} \rightharpoonup 0$
force $\langle x^{(k_j)}, y\rangle \to 0$: contradiction.
Hence weakly null sequences of $\ell^1$ are norm-null, and by
translation weakly convergent ones converge in norm: Schur's
theorem.

\textbf{21.} A weakly convergent subsequence of $(e_k)$ would
have limit $0$ (coordinates), hence by Schur $\norm{e_{k_j}}_1
\to 0$ --- but the norms are $1$. So no weakly convergent
subsequence exists, as found by hand in question 13. No
paradox: Schur says sequences cannot distinguish the weak
from the norm topology in $\ell^1$ (the topologies themselves
do differ --- weak neighborhoods are never norm-bounded), and
weak sequential compactness of the unit ball is a different,
stronger property, equivalent to reflexivity
(Eberlein--\v Smulian, beyond our tools; the failure, at
least, we have proved).

\textbf{22.} \emph{The census.}
\begin{center}
\begin{tabular}{l|l|c|c|c}
$E$ & $E'$ & sep. & refl. & weak seq.\ cpt.\ balls\\
\hline
$c_0$ & $\ell^1$ & yes & no & no ($e_1{+}\dots{+}e_k$)\\
$\ell^1$ & $\ell^\infty$ & yes & no & no ($e_k$, q.~21)\\
$\ell^p$ & $\ell^q$ & yes & yes & yes (q.~13)\\
$\ell^\infty$ & $\supsetneq\ell^1$ & no & no & no\\
\end{tabular}
\end{center}
Signatures: $c_0$ --- its bidual is $\ell^\infty$: the first
non-reflexive step (question 14); $\ell^1$ --- Schur's
property (question 20); $\ell^p$ --- reflexivity and weak
compactness (questions 7, 13); $\ell^\infty$ ---
non-separability and Banach limits: functionals no sequence
can represent (questions 8, 10--11). One family of spaces,
four different worlds.

\textbf{23.} Let $(f_k) \subseteq F$ with $\norm{x - f_k}
\to d$. Then $\norm{f_k} \leq \norm x + \sup_k\norm{x -
f_k}$: bounded, so by question 13 a subsequence $f_{k_j}
\rightharpoonup f$. If $f \notin F$, then $\delta =
\operatorname{dist}(f, F) > 0$ ($F$ closed); on $F \oplus \R
f$ the linear form $\lambda(g + tf) = t$ satisfies
$\abs{\lambda(u)} \leq \norm u/\delta$ (because $\norm{g +
tf} \geq \abs t\,\delta$), and Hahn--Banach extends it to
$\Lambda \in (\ell^p)'$ with $\Lambda\restriction_F = 0$,
$\Lambda(f) = 1$; but then $0 = \Lambda(f_{k_j}) \to
\Lambda(f) = 1$: contradiction. So $f \in F$, and $x -
f_{k_j} \rightharpoonup x - f$ gives, by question 17,
\[
d \leq \norm{x - f} \leq \liminf_j\,\norm{x - f_{k_j}} = d :
\]
the distance is attained at $f$. In $c_0$: $\abs{\Lambda(x)}
\leq \sum_n2^{-n}\abs{x_n} < \norm x_\infty$ for every $x
\neq 0$ (a nonzero null sequence cannot satisfy $\abs{x_n} =
\norm x_\infty$ for all $n$), while the truncated ones $(1,
\dots, 1, 0, \dots)$ give $\Lambda = 1 - 2^{-N} \to 1$: so
$\norm\Lambda = 1$, never attained. Distance formula: for $f
\in \ker\Lambda$, $\abs{\Lambda(x)} = \abs{\Lambda(x - f)}
\leq \norm{x - f}$, so $\operatorname{dist}(x, \ker\Lambda)
\geq \abs{\Lambda(x)}$; conversely, for $u$ in the unit ball
with $\Lambda(u) \geq 1 - \varepsilon$, the vector $f = x -
\frac{\Lambda(x)}{\Lambda(u)}u$ lies in $\ker\Lambda$ with
$\norm{x - f} \leq \frac{\abs{\Lambda(x)}}{1 - \varepsilon}$:
equality. If some $f \in \ker\Lambda$ attained it, $z = x -
f$ would satisfy $\abs{\Lambda(z)} = \abs{\Lambda(x)} =
\norm z \neq 0$, so $\Lambda$ would attain its norm at
$z/\norm z$: impossible. A closed hyperplane of $c_0$ with no
nearest points anywhere --- reflexivity was not decorative.

\textbf{24.} Set $y_1 = x^{(1)}$. Given $y_1, \dots, y_j$,
each map $k \mapsto \langle y_i, x^{(k)}\rangle$ tends to $0$
($y_i \in \ell^2 = (\ell^2)'$), so there is $k_{j+1}$ beyond
the previous index with $\abs{\langle y_i,
x^{(k_{j+1})}\rangle} \leq \frac1{j+1}$ for $i = 1, \dots,
j$; call the choice $y_{j+1}$. Then
\[
\Bigl\lVert\sum_{j=1}^my_j\Bigr\rVert_2^2
= \sum_{j=1}^m\norm{y_j}_2^2
+ 2\sum_{j=2}^m\sum_{i<j}\langle y_i, y_j\rangle
\leq mC^2 + 2\sum_{j=2}^m\frac{j-1}j
\leq mC^2 + 2m,
\]
and dividing by $m^2$: $\norm{\frac1m\sum_jy_j}_2^2 \leq
\frac{C^2 + 2}m \to 0$. (For a weak limit $x \neq 0$, apply
this to $x^{(k)} - x$.) On the orthonormal $(e_k)$ no
extraction is even needed: $\norm{\frac1m(e_1 + \dots +
e_m)}_2 = \frac{\sqrt m}m = \frac1{\sqrt m}$. Averages
convert weak convergence into norm convergence: the
Banach--Saks property of $\ell^2$.

\textbf{25.} $A_mx = \frac1m\sum_{i=0}^{m-1}S^ix$, so
linearity and shift-invariance give $L(A_mx) = L(x)$. For any
bounded $u$ and $\varepsilon > 0$, pick $N$ with $u_n \leq
\limsup u + \varepsilon$ for $n \geq N$; positivity applied
to $(\limsup u + \varepsilon)\mathbf 1 - S^Nu \geq 0$ and
$L(S^Nu) = L(u)$ give $L(u) \leq \limsup u + \varepsilon$,
and symmetrically $L(u) \geq \liminf u - \varepsilon$: hence
$\liminf A_mx \leq L(x) \leq \limsup A_mx$ for every $m$. If
$x$ is $T$-periodic, $A_Tx$ is the constant sequence equal to
the period mean $\mu$: $L(x) = \mu$ for \emph{every} Banach
limit --- $\frac13$ on $(1,0,0,\dots)$, $\frac12$ on
$(0,1,0,1,\dots)$ as in question 12. For the block sequence:
at $N = 3^j$ with $j$ even the last block is all ones, so the
Ces\`aro mean is $\geq \frac{3^j - 3^{j-1}}{3^j} = \frac23$;
at $N = 3^j$ with $j$ odd all the ones sit in $\intoc0
{3^{j-1}}$, so the mean is $\leq \frac13$. Hence $p(x) \geq
\frac23$ and $-p(-x) = \liminf_n\frac{x_1 + \dots + x_n}n
\leq \frac13$. On $M = c \oplus \R x$ define $\Lambda_+(y +
tx) = \lim y + t\,p(x)$. Domination by $p$: for $t > 0$,
sublinearity gives $p(tx) \leq p(y + tx) + p(-y)$, i.e.\
$p(y + tx) \geq t\,p(x) + \lim y$ (note $p(\pm y) = \pm\lim
y$ for $y \in c$: Ces\`aro means of a convergent sequence
converge to its limit); for $t = -s < 0$, $p(y) \leq p(y -
sx) + p(sx)$ gives $p(y - sx) \geq \lim y - s\,p(x)$; for $t
= 0$ there is equality. So $\Lambda_+ \leq p$ on $M$, and
Hahn--Banach extends it to $L_+ \leq p$ on $\ell^\infty_\R$,
which is a Banach limit exactly as in question 10 (domination
by $p$ yields positivity, shift-invariance, and the value
$\lim$ on $c$), with $L_+(x) = p(x) \geq \frac23$. The same
computation with $\Lambda_-(y + tx) = \lim y - t\,p(-x)$
(using $p(-sx) \leq p(y - sx) + p(-y)$ for the case $t = -s
< 0$) yields a Banach limit $L_-$ with $L_-(x) = -p(-x) \leq
\frac13$. Two Banach limits, one sequence, two values:
outside the periodic (and, more generally, \emph{almost
convergent}) world, a Banach limit is a genuine choice.
\end{solution}
