\section*{Chapter \ref{ch:b3:measure} --- Measure Theory}

\begin{solution}{exo:b3:measure:1}
(a) Complementation swaps the two defining cases. A countable
union of countable sets is countable; if one member is
co-countable, the union is co-countable: stability holds. It
contains the singletons, and any $\sigma$-algebra containing
them contains all countable sets (countable unions) and their
complements: it is $\sigma(\{\text{singletons}\})$.

(b) Write $\mathcal B = \sigma(\text{opens})$. Every open
subset of $\R$ is a countable union of open intervals with
rational endpoints (around each rational point of the open set,
a rational-radius interval inside it), so opens $\in
\sigma(\text{open intervals}) \subseteq \sigma(\text{rational
data})$. Conversions: $\intoo ab = \bigcup_n\intcc{a +
\frac1n}{b - \frac1n}$; $\intcc ab = \bigcap_n \intoo{a -
\frac1n}{b + \frac1n}$; $\intoc{-\infty}a = \bigcap_n
\intoo{-\infty}{a + \frac1n}$ and conversely $\intoo ab =
\intoo{-\infty}b \setminus \intoc{-\infty}a$; rational rays:
$\intoc{-\infty}a =
\bigcap_{q \in \Q,\, q > a}\intoc{-\infty}q$. Each family
generates the others by countable operations: all four generate
$\mathcal B$.

(c) With finite endpoints only, the family is not even an
algebra: the complement of $\intoc01$ contains unbounded rays.
Allowing infinite endpoints ($\intoc{-\infty}b$, $\intoo
a{+\infty}$) it becomes an algebra (complements and finite
unions of such unions are such), but not a $\sigma$-algebra:
$\{0\} = \bigcap_n\intoc{-\frac1n}0$ is not a finite union of
nondegenerate intervals.
\end{solution}

\begin{solution}{exo:b3:measure:2}
(a) $A \cup B = A \sqcup (B \setminus (A\cap B))$, so
$\mu(A\cup B) = \mu(A) + \mu(B) - \mu(A\cap B)$ (finiteness
permits the subtraction). Three sets: apply the two-set formula
twice,
\[
\mu(A\cup B\cup C) = \sum\mu(A) - \sum\mu(A\cap B) +
\mu(A\cap B\cap C)
\]
(sums over the obvious index sets).

(b) For Lebesgue measure, $A_n = [n, +\infty) \downarrow
\varnothing$, but $\lambda(A_n) = \infty \not\to 0$.

(c) A point lies in an interval of length $\varepsilon$:
$\lambda(\{x\}) = 0$; countable subadditivity kills countable
sets. Hence $\lambda(\Q \cap \intcc01) = 0$ and, by additivity,
$\lambda(\intcc01\setminus\Q) = 1$: the irrationals carry all
the length.
\end{solution}

\begin{solution}{exo:b3:measure:3}
The rays $\intoc{-\infty}t$ form a $\pi$-system (the
intersection of two is the smaller) generating $\mathcal B(\R)$
(\cref{exo:b3:measure:1}). The sets $P_k = \intoc{-\infty}k$
increase to $\R$ with $\mu(P_k) \leq 1 < \infty$:
\cref{thm:b3:measure:uniqueness} applies, and $\mu = \nu$ on
$\mathcal B(\R)$. Thus the distribution function determines the
measure.
\end{solution}

\begin{solution}{exo:b3:measure:4}
For every $N$, $\limsup A_n \subseteq \bigcup_{n \geq N}A_n$,
so $\mu(\limsup A_n) \leq \sum_{n\geq N}\mu(A_n)$, the tail of
a convergent series: let $N \to \infty$. Application: with
$A_n = \{x \in \intcc01 : \abs{x - r_n} \leq 4^{-n}\}$
($r_n$ the $n$-th rational), $\lambda(A_n) \leq 2\cdot4^{-n}$
is summable: $\lambda(\limsup A_n) = 0$, i.e.\ almost every $x$
belongs to only finitely many $A_n$. (Yet every $x$ is a limit
of rationals: the point is the \emph{speed} $4^{-n}$.)
\end{solution}

\begin{solution}{exo:b3:measure:5}
$K = \bigcap K_n$ is an intersection of finite unions of closed
intervals: compact. At stage $n$ there remain $2^n$ intervals
of common length $\ell_n \leq 2^{-n}$ (each stage halves and
shrinks); an interval $I \subseteq K$ would lie inside a single
stage-$n$ interval for every $n$, forcing $\ell(I) = 0$: empty
interior. The measure removed is $\sum_{n\geq1}2^{n-1}\cdot
4^{-n} = \frac12\sum_{n\geq1}2^{-n} = \frac12$, all removals
being disjoint open intervals: $\lambda(K) = \frac12$.

Variant: removing central intervals of length $\varepsilon
4^{-n}$ leaves a nowhere dense compact $K^{(\varepsilon)}$ of
measure $1 - \frac\varepsilon2$. Then $\bigcup_m
K^{(1/m)}$ is meagre (countable union of nowhere dense sets)
of measure $\geq \sup_m(1 - \frac1{2m}) = 1$: a meagre set of
full measure --- and its complement in $\intoo01$ is a dense
$G_\delta$ of measure $0$ (topologically fat, metrically
null). The complement of $K^{(\varepsilon)}$ in $\intoo01$ is
open, dense, of measure $\frac\varepsilon2 < \varepsilon$.
\end{solution}

\begin{solution}{exo:b3:measure:6}
Cutting $\intoc01$ into $2^n$ translates of $\intoc0{2^{-n}}$:
$c = 2^n\,\mu(\intoc0{2^{-n}})$, so $\mu(\intoc0{2^{-n}}) =
c\,2^{-n} = c\,\lambda(\intoc0{2^{-n}})$. By translation
invariance and additivity, $\mu = c\lambda$ on every interval
$\intoc ab$ with $b - a$ a dyadic rational and any $a$; a
general $\intoc ab$ is an increasing union of such
($b_k \uparrow b$ dyadic steps from $a$), and continuity from
below extends the equality. The intervals $\intoc ab$ form a
$\pi$-system generating $\mathcal B(\R)$, with $\intoc{-k}k
\uparrow \R$ of finite measure ($\mu(\intoc{-k}k) = 2kc$):
\cref{thm:b3:measure:uniqueness} gives $\mu = c\lambda$ on
$\mathcal B(\R)$.
\end{solution}

\begin{solution}{exo:b3:measure:7}
By regularity (\cref{thm:b3:measure:regularity}) choose $K
\subseteq A \subseteq U$ with $K$ compact, $U$ open,
$\lambda(U\setminus K) < \varepsilon$ (both approximations
within $\varepsilon/2$, and $\lambda(U \setminus K) =
\lambda(U\setminus A) + \lambda(A \setminus K)$). Write $U$ as
a countable disjoint union of open intervals $(I_n)$ (the
components of the open set); the compact $K$ is covered by
finitely many, $K \subseteq B = I_1\cup\dots\cup I_N \subseteq
U$. Then $A \setminus B \subseteq A\setminus K \subseteq
U\setminus K$ and $B \setminus A \subseteq U \setminus A
\subseteq U\setminus K$:
$\lambda(A\,\triangle\,B) \leq 2\lambda(U\setminus K)$ ---
start from $\varepsilon/2$ to land below $\varepsilon$.
\end{solution}

\begin{solution}{exo:b3:measure:8}
Replacing $A$ by $A \cap [-M, M]$ of positive measure (some $M$
works, by continuity from below), assume $0 < \lambda(A) <
\infty$. Take $U \supseteq A$ open with $\lambda(U) <
\frac43\lambda(A)$ and decompose $U = \bigsqcup_nI_n$ into
disjoint open intervals: $\lambda(A) = \sum_n\lambda(A\cap
I_n)$. If every $n$ had $\lambda(A\cap I_n) \leq
\frac34\ell(I_n)$, summing would give $\lambda(A) \leq
\frac34\lambda(U) < \lambda(A)$: some interval $I$ satisfies
$\lambda(A\cap I) > \frac34\ell(I)$. Set $B = A \cap I$ and let
$\abs t < \frac12\ell(I)$: both $B$ and $B + t$ lie in the
interval $I \cup (I + t)$, of length $< \frac32\ell(I)$. If
they were disjoint: $\lambda(B) + \lambda(B + t) = 2\lambda(B)
> \frac32\ell(I)$ would exceed the containing interval's
measure --- impossible. So $B \cap (B + t) \neq \varnothing$:
some $x \in B$ writes $x = y + t$ with $y \in B$, and $t = x -
y \in A - A$.
Hence $\intoo{-\frac{\ell(I)}2}{\frac{\ell(I)}2} \subseteq A -
A$.
\end{solution}

\begin{solution}{exo:b3:measure:9}
The Vitali translates $(V + q)_{q\in\Q}$ partition $\R$ (every
real is equivalent to exactly one representative). Suppose all
the sets $B_q = A \cap (V + q)$ were measurable. Any two
elements of $V + q$ differ by an irrational or zero (two
distinct representatives are inequivalent), so $B_q - B_q$
meets $\Q$ only in $\{0\}$: it contains no interval, and
Steinhaus (\cref{exo:b3:measure:8}) forces $\lambda(B_q) = 0$.
Then $\lambda(A) \leq \sum_{q}\lambda(B_q) = 0$, contradicting
$\lambda(A) > 0$. So some $B_q \subseteq A$ is non-measurable.
\end{solution}

\begin{solution}{exo:b3:measure:10}
\emph{Measurable $\Rightarrow$ $\varepsilon$-approximation}: by
outer regularity (\cref{thm:b3:measure:regularity}) pick open
$U \supseteq A$ with $\lambda(U) \leq \lambda(A) +
\varepsilon$; measurability allows the subtraction
$\lambda(U\setminus A) = \lambda(U) - \lambda(A) \leq
\varepsilon$. \emph{$\varepsilon$-version $\Rightarrow$
$G_\delta$-version}: take $U_n$ with $\lambda^*(U_n\setminus
A) < \frac1n$ and $G = \bigcap U_n$: a $G_\delta$ with
$\lambda^*(G\setminus A) \leq \lambda^*(U_n\setminus A) \to 0$.
\emph{$G_\delta$-version $\Rightarrow$ measurable}: $G\setminus
A$ is $\lambda^*$-null, hence measurable by completeness
(\cref{thm:b3:measure:caratheodory}), and $A = G \setminus
(G\setminus A)$ is measurable ($G$ is Borel). So Lebesgue sets
are exactly ``Borel modulo null''.
\end{solution}

\begin{solution}{exo:b3:measure:11}
(a) $\liminf A_n = \bigcup_N\bigcap_{n\geq N}A_n$ is an
increasing union of the sets $B_N = \bigcap_{n\geq N}A_n$,
so $\mu(\liminf A_n) = \lim_N\mu(B_N)$ (continuity from
below); and $\mu(B_N) \leq \inf_{n \geq N}\mu(A_n)$, whose
limit is $\liminf\mu(A_n)$. For the $\limsup$: apply the
same to the complements inside the finite-measure ambient
$U = \bigcup A_n$ --- continuity from above on the
decreasing $C_N = \bigcup_{n\geq N}A_n$ requires $\mu(C_1)
\leq \mu(U) < \infty$, and gives $\mu(\limsup A_n) =
\lim\mu(C_N) \geq \limsup\mu(A_n)$.

(b) The moving interval $A_n = \intcc n{n+1}$: every point
belongs to at most two of the $A_n$ and to none eventually,
so $\limsup A_n = \varnothing$; yet $\mu(A_n) = 1$. Thus
$\limsup\mu(A_n) = 1 > 0 = \mu(\limsup A_n)$: without a
finite-measure envelope, the second inequality of (a) fails
--- the mass escapes to infinity, where no fixed set can
catch it.

(c) If $\sum\mu(A_n) < \infty$: $\mu(C_N) \leq
\sum_{n\geq N}\mu(A_n) \to 0$ and $\limsup A_n =
\bigcap C_N$ has measure $\leq \inf\mu(C_N) = 0$. Monotone
cases: increasing is continuity from below; decreasing with
$\mu(A_1) < \infty$ is continuity from above ---
both proved in \cref{ch:b3:measure}'s basic properties; the
counterexample $A_n = \intco n\infty$ (decreasing to
$\varnothing$ with $\mu \equiv \infty$) shows the finiteness
is again essential.
\end{solution}

\begin{solution}{exo:b3:measure:12}
(a) The sets $E_{k,N}$ increase with $N$ (fewer
constraints), and every $x$ eventually satisfies $\abs{f_n(x)
- f(x)} \leq \frac1k$ for all $n \geq N(x)$ (pointwise
convergence): $\bigcup_NE_{k,N} = X$. Continuity from below:
$\mu(E_{k,N}) \to \mu(X) < \infty$, so $\mu(X\setminus
E_{k,N}) \to 0$; choose $N_k$ accordingly.

(b) Let $A = \bigcap_kE_{k,N_k}$: $\mu(X\setminus A) \leq
\sum_k\varepsilon2^{-k} = \varepsilon$. On $A$: for every
$k$, all $n \geq N_k$ satisfy $\sup_A\abs{f_n - f} \leq
\frac1k$ --- exactly uniform convergence on $A$.

(c) For the moving bump, uniform convergence on $A$ requires
$A$ to eventually avoid every $\intcc n{n+1}$ --- more
precisely $\sup_A\abs{f_n} < \frac12$ forces $A \cap \intcc
n{n+1}$ to be empty for large $n$, so $X \setminus A$
contains a tail $\bigcup_{n\geq n_0}\intcc n{n+1}$, of
infinite measure. In (a), finiteness converted
``$E_{k,N}\nearrow X$'' into ``the complements' measures
tend to $0$'': continuity from above needs a finite start,
and on infinite-measure spaces the escape to infinity is
precisely what it cannot see.
\end{solution}

\begin{solution}{pb:b3:measure:1}
\textbf{1.} Induction. Continuity: the three formulas agree at
the junctions ($\frac12c_n(1) = \frac12$ and $\frac12 +
\frac12c_n(0) = \frac12$); each piece is continuous.
Monotonicity and the boundary values are inherited. For the
contraction estimate: on $[0,\frac13]$, $\abs{c_{n+1} - c_n}(x)
= \frac12\abs{c_n - c_{n-1}}(3x) \leq \frac12\norm{c_n -
c_{n-1}}_\infty$; on the middle third the difference is $0$; on
the right third, the same as the left.

\textbf{2.} $\norm{c_{n+1} - c_n}_\infty \leq
2^{-n}\norm{c_1 - c_0}_\infty$: the series of increments
converges uniformly, so $c_n \to c$ uniformly; $c$ is
continuous, nondecreasing, $c(0) = 0$, $c(1) = 1$ (all
preserved by uniform limits), and passing to the limit in the
defining recursion shows $c$ itself satisfies the three
self-similar identities.

\textbf{3.} By the middle identity, $c \equiv \frac12$ on
$\intcc{\frac13}{\frac23}$, the first gap. Every gap of $C$ is
the image of the first gap under a composition of the two
affine contractions $x \mapsto \frac x3$, $x\mapsto\frac{x +
2}3$; the identities transport constancy accordingly (with
values the dyadic rationals). For the digit formula, take $x =
\sum_n 2b_n3^{-n} \in C$: if $b_1 = 0$ then $x \in
[0,\frac13]$ and $c(x) = \frac12c(3x)$ with $3x$ having digits
$(b_2, b_3, \dots)$; if $b_1 = 1$ then $x \in [\frac23, 1]$
and $c(x) = \frac12 + \frac12c(3x - 2)$, same shift. By
induction, the first $N$ binary digits of $c(x)$ are $b_1,
\dots, b_N$ for every $N$: $c(x) = \sum_nb_n2^{-n}$.

\textbf{4.} Off $C$, $c$ is locally constant: differentiable
with derivative $0$. Since $\lambda(C) = 0$
(\cref{ex:b3:measure:cantor}), $c' = 0$ almost everywhere. Yet
$c(1) - c(0) = 1$: the fundamental theorem in its $\mathcal
C^1$ form requires $c$ to be differentiable \emph{everywhere}
with continuous (or at least integrable, plus absolute
continuity --- see \cref{ch:b3:lebesgue}) derivative; $c$ is
not differentiable at points of $C$, and more fundamentally $c$
fails \emph{absolute continuity}: it climbs on a null set.

\textbf{5.} Given $y = \sum_n\beta_n2^{-n} \in \intcc01$
($\beta_n \in \{0,1\}$), the point $x = \sum_n2\beta_n3^{-n}
\in C$ has $c(x) = y$ by question 3: $c(C) = \intcc01$, a set
of measure $1$ --- the null set $C$ carries, through $c$, the
whole interval.

\textbf{6.} $h$ is continuous, and strictly increasing ($x$ is,
$c$ is nondecreasing); $h(0) = 0$, $h(1) = 1$, so by the
intermediate value theorem $h$ is a continuous bijection of
$\intcc01$; a continuous bijection from a compact space to a
Hausdorff one is a homeomorphism
(\cref{cor:b3:topology:compacthomeo}).

\textbf{7.} On a gap $(u, v)$ (length $\ell$), $c$ is constant,
so $h$ is affine with slope $\frac12$: $h((u,v))$ is an
interval of length $\frac\ell2$. The gaps are disjoint and $h$
is injective: the images are disjoint, of total measure
$\frac12\sum\ell = \frac12(1 - \lambda(C)) = \frac12$.

\textbf{8.} $h(\intcc01) = \intcc01$ and $h(C)$ is compact
(continuous image), hence measurable, with
\[
\lambda\bigl(h(C)\bigr) = 1 -
\lambda\bigl(h(\intcc01\setminus C)\bigr) = 1 - \tfrac12 =
\tfrac12 .
\]
A homeomorphism can inflate a null set to measure $\frac12$:
``topological size'' (category, dimension) and ``measure'' are
transported by homeomorphisms very differently --- only the
former is a topological invariant.

\textbf{9.} $Z = h^{-1}(W) \subseteq h^{-1}(h(C)) = C$ has
$\lambda^*(Z) \leq \lambda(C) = 0$: a null set, hence
Lebesgue-measurable by completeness
(\cref{thm:b3:measure:caratheodory}).

\textbf{10.} Let $\varphi$ be continuous and $\mathcal D = \{B
: \varphi^{-1}(B) \in \mathcal B\}$. Preimages commute with
complements and countable unions, so $\mathcal D$ is a
$\sigma$-algebra; it contains the open sets (continuity):
$\mathcal D \supseteq \mathcal B$ --- preimages of Borel sets
under continuous maps are Borel.

\textbf{11.} If $Z$ were Borel, apply question 10 to the
continuous $\varphi = h^{-1}$: $\varphi^{-1}(Z) = h(Z) = W$
would be Borel, hence Lebesgue-measurable --- contradicting the
choice of $W$. So $Z \in \mathcal L \setminus \mathcal B$:
Lebesgue's $\sigma$-algebra strictly contains Borel's.

\textbf{12.} $g = \mathbf 1_Z$ is Lebesgue-measurable ($Z \in
\mathcal L$) and $\varphi = h^{-1}$ is continuous, but
$(g\circ\varphi)^{-1}(\{1\}) = \varphi^{-1}(Z) = W$ is not
measurable: $g \circ \varphi$ is not Lebesgue-measurable. The
care needed: ``Lebesgue-measurable function'' means preimages
of \emph{Borel} sets land in $\mathcal L$; composing requires
preimages of \emph{Lebesgue} sets to be Lebesgue, which
continuity does not grant (here $\varphi^{-1}(Z) \notin
\mathcal L$ even though $\varphi$ is a homeomorphism).

\textbf{13.} The chains, with witnesses for strictness:
\[
\{\text{countable}\} \subsetneq \{\text{Borel null}\}
\subsetneq \{\text{Lebesgue null}\} \subsetneq \mathcal L
\subsetneq \mathcal P(\R),
\qquad
\{\text{Borel null}\} \subsetneq \mathcal B \subsetneq
\mathcal L .
\]
Witnesses: $C$ is Borel, null, uncountable (first gap); $Z$ is
Lebesgue null but not Borel (second and, inside $\mathcal B
\subsetneq \mathcal L$, sixth); a fat Cantor set is Borel,
nowhere dense, of positive measure (separating null sets from
Borel sets); Vitali's $V$ is not in $\mathcal L$ (last gap).
Measure, topology and cardinality slice $\mathcal P(\R)$ along
genuinely different lines.

\textbf{14.} Outer measure: $\mu_F^*(\varnothing) = 0$ (cover
by a vanishing interval), monotonicity is clear, and countable
subadditivity follows by splicing $\varepsilon2^{-k}$-optimal
covers, exactly as for $\lambda^*$. The one-interval cover
gives $\mu_F^*(\intoc ab) \leq F(b) - F(a)$. Conversely let
$\intoc ab \subseteq \bigcup_k\intoc{a_k}{b_k}$. By continuity
of $F$, pick $b_k' > b_k$ with $F(b_k') \leq F(b_k) +
\varepsilon2^{-k}$, and $a' \in \intoo ab$ with $F(a') \leq
F(a) + \varepsilon$. The compact $\intcc{a'}b$ is covered by
the open $\intoo{a_k}{b_k'}$: finitely many suffice, and the
chaining argument of \cref{thm:b3:measure:lebesgue} (walk from
$a'$ to $b$ through overlapping intervals, telescoping
$F$-increments, monotonicity absorbing overlaps) yields $F(b)
- F(a') \leq \sum_k(F(b_k') - F(a_k)) \leq \sum_k(F(b_k) -
F(a_k)) + \varepsilon$. Let $\varepsilon \to 0$: equality.
(Continuity of $F$ is what allowed opening the intervals at
arbitrarily small $F$-cost.)

\textbf{15.} It suffices to prove each half-line $H_t =
\intoc{-\infty}t$ Carath\'eodory-measurable, since the
measurable sets form a $\sigma$-algebra
(\cref{thm:b3:measure:caratheodory}) and half-lines generate
$\mathcal B$. Given $A$ and an $\varepsilon$-optimal cover
$(\intoc{a_k}{b_k})$ of $A$: each interval splits as
$\intoc{a_k}{t\wedge b_k} \cup \intoc{t \vee a_k}{b_k}$ (one
piece possibly empty), with $F$-costs adding up exactly to
$F(b_k) - F(a_k)$; the first pieces cover $A \cap H_t$, the
second $A \setminus H_t$. Hence $\mu_F^*(A\cap H_t) +
\mu_F^*(A\setminus H_t) \leq \mu_F^*(A) + \varepsilon$, and
the reverse inequality is subadditivity. Restricting the
resulting measure to $\mathcal B$: the Lebesgue--Stieltjes
measure $\mu_F$.

\textbf{16.} $\mu(\R) = \lim_n(F(n) - F(-n)) = 1 - 0 = 1$
(continuity of the measure along $\intoc{-n}n$). On a gap
$\intoo uv$ of $C$, $c$ is constant, so every half-open
subinterval is $\mu$-null and so is the gap (countable
union); outside $\intcc01$, $c$ is constant too. Hence
$\mu(\R\setminus C) = 0$, $\mu(C) = 1$, while $\lambda(C) =
0$ (\cref{ex:b3:measure:cantor}): each of $\mu, \lambda$ is
carried by a set the other declares null --- mutually
singular.

\textbf{17.} $\mu(\{x\}) = \lim_{\delta\downarrow0}
\mu(\intoc{x-\delta}x) = \lim(c(x) - c(x-\delta)) = 0$ by
continuity of $c$: no atoms. So $\mu$ is an atomless
probability measure carried by a Lebesgue-null compact ---
neither diffuse-with-density like $\lambda$'s restrictions,
nor atomic like counting measures: a third species.

\textbf{18.} The piece $C_\varepsilon$ spans a ternary
interval $I_\varepsilon$ of length $3^{-m}$, and Part I
question 3 shows that across $I_\varepsilon$ the staircase
climbs exactly $2^{-m}$ (the first $m$ binary digits of $c$
are frozen to $\varepsilon$, the rest sweep everything).
Hence $\mu(C_\varepsilon) = \mu(I_\varepsilon) =
c(\text{right end}) - c(\text{left end}) = 2^{-m}$: depth-$m$
digit cylinders all have mass $2^{-m}$, the law of $m$ fair
coins.

\textbf{19.} The right-hand side defines the Borel measure
$\nu = \frac12\,(x \mapsto \tfrac x3)_*\mu + \frac12\,(x
\mapsto \tfrac{x+2}3)_*\mu$ evaluated at $A$ --- probability
measure. Pointwise, with $c$ globally extended, one checks
case by case ($x \leq 0$; the three thirds; $x \geq 1$) the
identity
\[
c(x) = \tfrac12\,c(3x) + \tfrac12\,c(3x - 2),
\]
e.g.\ on $\intcc{1/3}{2/3}$: $\frac12\cdot1 + \frac12\cdot0 =
\frac12 = c(x)$. Evaluating $\nu$ on $\intoc ab$ therefore
gives $c(b) - c(a) = \mu(\intoc ab)$, and two finite measures
agreeing on the $\pi$-system of half-open intervals coincide
on $\mathcal B$ (\cref{thm:b3:measure:uniqueness}): $\nu =
\mu$.

\textbf{20.} By induction on $n$: $c_0(1-x) = 1 - c_0(x)$,
and if $c_n$ has the symmetry, then for $x \in
\intcc0{1/3}$: $c_{n+1}(1 - x) = \frac12 + \frac12c_n(3(1-x)
- 2) = \frac12 + \frac12c_n(1 - 3x) = \frac12 + \frac12(1 -
c_n(3x)) = 1 - c_{n+1}(x)$; the middle third mirrors around
$\frac12$; the right third is the left case reflected. In the
limit $c(1-x) = 1 - c(x)$. Pushforward: $(s_*\mu)(\intoc ab)
= \mu(\intco{1-b}{1-a}) = c(1-a) - c(1-b)$ ($\mu$ atomless,
question 17, so boundary conventions cost nothing) $= (1 -
c(a)) - (1 - c(b)) = \mu(\intoc ab)$: $s_*\mu = \mu$ by
uniqueness.

\textbf{21.} Let $X \sim \mu$ (integrals of continuous
functions against $\mu$ exist as limits of Riemann-type sums
over the depth-$m$ pieces, each of mass $2^{-m}$, with
sampling error $\leq \operatorname{osc} \leq
\norm{f'}_\infty3^{-m}$; \cref{ch:b3:lebesgue} will
systematize this). Symmetry: $1 - X \sim X$, so $\E X =
\frac12$. Self-similarity: $X$ has the law of $\frac Y3$ with
probability $\frac12$ and of $\frac{Y+2}3$ with probability
$\frac12$, $Y \sim \mu$, so
\[
\E X^2 = \frac12\,\frac{\E Y^2}9 + \frac12\,
\frac{\E Y^2 + 4\E Y + 4}{9} = \frac{2\E X^2 + 6}{18},
\]
whence $\E X^2 = \frac38$ and $\operatorname{Var}X = \frac38
- \frac14 = \frac18$. The uniform law has variance
$\frac1{12} < \frac18$: Cantor mass hugs the endpoints.

\textbf{22.} $C$ is closed and $\mu(C) = 1$. If an open $I$
meets $C$ at $x$, the depth-$m$ pieces containing $x$ shrink
to $x$, so some $I_\varepsilon \subseteq I$ and $\mu(I) \geq
2^{-m} > 0$: no smaller closed set can carry $\mu$. Points
off $C$ have gap neighborhoods of $\mu$-measure $0$. Hence
$\operatorname{supp}\mu = C$ exactly.

\textbf{23.} One phenomenon, two dialects. Part I says the
growth of $c$ is invisible to its derivative: $c' = 0$ a.e.,
all the climbing concentrated on the null set $C$. Part V
says the associated measure $\mu_c$ puts all its mass on that
same null set: $\mu_c \perp \lambda$, so no density $f \geq
0$ can satisfy $\mu_c(A) = \int_Af\,\dd\lambda$ --- a density
forces vanishing on $\lambda$-null sets. The dictionary:
nondecreasing bounded $F \leftrightarrow$ finite measure
$\mu_F$ (questions 14--15); $F$ an integral of its derivative
$\leftrightarrow$ $\mu_F$ has a density (the ``absolutely
continuous'' case); and in general $F'$, which exists a.e.\
for monotone $F$ (Lebesgue's differentiation theorem, beyond
this chapter), recovers only the density part. The staircase
is the extreme: continuous, with derivative $0$ a.e.\ ---
its measure is \emph{purely singular}, and the fundamental
theorem of calculus, far from failing by accident, fails by
the exact amount $\mu_c(\R) = 1$ of singular mass.

\textbf{24.} Fix $x < y$ in $\intcc01$ and choose $m \geq 0$
with $3^{-(m+1)} < y - x \leq 3^{-m}$. Across any triadic
interval $\intcc{k3^{-m}}{(k+1)3^{-m}}$ the staircase climbs
at most $2^{-m}$: such an interval is either one of the $2^m$
pieces of $C_m$, where the climb is exactly $2^{-m}$ (question
18), or it is contained in the closure of a single gap of some
stage $\leq m$, where $c$ is constant. Since $y - x \leq
3^{-m}$, the interval $\intcc xy$ meets at most two
consecutive depth-$m$ triadic intervals, so
\[
c(y) - c(x) \leq 2\cdot2^{-m}
= 2\bigl(3^{-m}\bigr)^{s}
< 2\bigl(3(y - x)\bigr)^{s}
= 4\,(y - x)^{s},
\]
using $3^{s} = 2$ and $3^{-m} < 3(y - x)$. Optimality: the
endpoints $u < v$ of a depth-$m$ Cantor piece satisfy $v - u =
3^{-m}$ and $c(v) - c(u) = 2^{-m} = (v - u)^{s}$; a H\"older
bound $\abs{c(v) - c(u)} \leq K(v - u)^{t}$ with $t > s$ would
force $K \geq 2^{-m}3^{mt} = (3^{t}/2)^{m} \to \infty$ (since
$3^{t} > 3^{s} = 2$), and every subinterval of $\intcc01$
contains such pieces, so the failure is local as well.
Measure form: $\mu\bigl(\intcc{x-r}{x+r}\bigr) \leq
c(x + r) - c(x - r) \leq 4(2r)^{s} = 4\cdot2^{s}r^{s} \leq
8r^{s}$ (values of $c$ extended to $\R$ as in question 16;
$\mu$ has no atoms, question 17).

\textbf{25.} Write $S_0(x) = \frac x3$ and $S_1(x) =
\frac{x+2}3$; the hypothesis says $\nu = \frac12(S_0)_*\nu +
\frac12(S_1)_*\nu$. Iterating $m$ times,
\[
\nu = 2^{-m}\sum_{w \in \{0,1\}^m}(S_w)_*\nu,
\qquad S_w = S_{w_1}\circ\dots\circ S_{w_m}.
\]
Since $\nu$ is carried by $\intcc01$ and $S_w(\intcc01) =
I_w$, the depth-$m$ Cantor piece indexed by $w$, each
$(S_w)_*\nu$ is a probability measure carried by $I_w$; the
$2^m$ pieces are pairwise disjoint closed intervals of length
$3^{-m}$. Fix $\intoc ab$ and let $N_m$ be the number of
pieces $I_w \subseteq \intoc ab$. A piece meeting $\intoc ab$
without being contained in it must contain $a$ or $b$, and a
point lies in at most one piece, so
\[
2^{-m}N_m \leq \nu(\intoc ab) \leq 2^{-m}N_m + 2\cdot2^{-m}.
\]
The same double inequality holds for $\mu$ (question 19 gives
the identical iteration), with the same count $N_m$. Hence
$\abs{\nu(\intoc ab) - \mu(\intoc ab)} \leq 2^{-m+1} \to 0$:
$\nu$ and $\mu$ agree on the $\pi$-system of half-open
intervals, and both are probability measures, so
\cref{thm:b3:measure:uniqueness} gives $\nu = \mu$. The
staircase measure is thus \emph{the} fixed point of the
two-map averaging scheme --- the measure-level statement of
the self-similarity of $C$.
\end{solution}
