\section*{Chapter \ref{ch:b3:lebesgue} --- The Lebesgue
Integral}

\begin{solution}{exo:b3:lebesgue:1}
(a) If $f$ is nondecreasing, $\{f > t\}$ is $\varnothing$, $\R$,
or a ray $\intoo a{+\infty}$ / $\intco a{+\infty}$: Borel in
every case; nonincreasing likewise. A derivative: $f'(x) =
\lim_n n\bigl(f(x + \frac1n) - f(x)\bigr)$ is a pointwise limit
of continuous (hence measurable) functions:
\cref{prop:b3:lebesgue:stability}(c).

(b) $\{f > t\} = \bigcup_{q \in \Q,\, q > t}\{f > q\}$: if the
rational levels are measurable, all levels are, and the rays
generate $\mathcal B(\R)$.
\end{solution}

\begin{solution}{exo:b3:lebesgue:2}
First: $(1 + x/n)^n \nearrow \eu^x$ for $x \geq 0$, so the
integrand tends pointwise to $\eu^{-x}\cos x$; for $n \geq 2$,
$(1 + x/n)^n \geq (1 + x/2)^2$, giving the integrable dominator
$(1 + x/2)^{-2}$. DCT:
\[
\lim_n\int_0^\infty\frac{\cos x}{(1 + x/n)^n}\dd x
= \int_0^\infty \eu^{-x}\cos x\,\dd x
= \operatorname{Re}\int_0^\infty\eu^{-(1 - \iu)x}\dd x
= \operatorname{Re}\frac{1}{1 - \iu} = \frac12 .
\]
Second: substitute $u = x^n$ (a $\mathcal C^1$ bijection of
$\intoo01$):
\[
\int_0^1\frac{nx^{n-1}}{1 + x}\dd x = \int_0^1\frac{\dd u}{1 +
u^{1/n}} \longrightarrow \int_0^1\frac{\dd u}{2} = \frac12,
\]
by DCT: for $u \in \intoo01$, $u^{1/n} \to 1$, and the
integrand is bounded by $1$ on a finite measure space.
\end{solution}

\begin{solution}{exo:b3:lebesgue:3}
(a) $f_n = n\,\mathbf 1_{\intoo0{1/n}}$: $\liminf f_n = 0$
pointwise, $\int f_n = 1$: $0 < 1$.

(b) Height: $n\mathbf 1_{\intoo0{1/n}}$; width: $\frac1n\mathbf
1_{\intoo0n}$; translation: $\mathbf 1_{\intoo n{n+1}}$. All
tend to $0$ pointwise with $\int = 1$. In each case the
\emph{domination} hypothesis fails: $\sup_nf_n$ is
$\approx 1/x$ near $0$, $\approx$ a nonintegrable constant
profile, $\mathbf 1_{\intoo1\infty}$-like --- never integrable.

(c) ``$\int\limsup f_n \geq \limsup\int f_n$'' fails for the
translating bump: left side $0$, right side $1$.
``$\limsup\int \leq \int\limsup$'' is the same statement. And
Fatou for $\limsup$ with $\leq$ reversed
(``reverse Fatou'') requires a dominator --- the same bump is
the counterexample.
\end{solution}

\begin{solution}{exo:b3:lebesgue:4}
(a) For $x > 0$: $\frac1{\eu^x - 1} = \frac{\eu^{-x}}{1 -
\eu^{-x}} = \sum_{n\geq1}\eu^{-nx}$, so $\frac{x}{\eu^x - 1} =
\sum_{n\geq1}x\eu^{-nx}$, a series of nonnegative measurable
functions: \cref{cor:b3:lebesgue:additivity} allows term-by-term
integration:
\[
\int_0^\infty\frac{x\,\dd x}{\eu^x - 1}
= \sum_{n\geq1}\int_0^\infty x\eu^{-nx}\dd x
= \sum_{n\geq1}\frac1{n^2} = \frac{\pi^2}6
\]
($\int_0^\infty x\eu^{-nx}\dd x = n^{-2}$ by parts; Basel from
the Year 2 volume, or \cref{exo:b3:hilbert:5} to come).

(b) On $\intoo01$, $-x\ln x \geq 0$, so $x^{-x} =
\eu^{-x\ln x} = \sum_k\frac{(-x\ln x)^k}{k!}$ is a series of
nonnegative terms: interchange again. Substituting $x =
\eu^{-u/(k+1)}$:
\[
\int_0^1(-x\ln x)^k\dd x
= \int_0^\infty\Bigl(\frac{u}{k+1}\Bigr)^{k}
\eu^{-\frac{ku}{k+1}}\;\frac{\eu^{-\frac u{k+1}}}{k+1}\,\dd u
= \frac{1}{(k+1)^{k+1}}\int_0^\infty u^k\eu^{-u}\dd u
= \frac{k!}{(k+1)^{k+1}} .
\]
Hence $\int_0^1x^{-x}\dd x = \sum_{k\geq0}\frac{1}{(k+1)^{k+1}}
= \sum_{n\geq1}n^{-n}$: the sophomore's dream, rigorously.
\end{solution}

\begin{solution}{exo:b3:lebesgue:5}
(a) Markov: $a\,\mathbf 1_{\{f \geq a\}} \leq f$, integrate:
$\mu(\{f \geq a\}) \leq \frac1a\int f$. If $\int f = 0$:
$\mu(\{f \geq \frac1n\}) = 0$ for every $n$, and $\{f > 0\} =
\bigcup_n\{f \geq \frac1n\}$ is null.

(b) $\mu(\{\abs f = \infty\}) \leq \mu(\{\abs f \geq n\}) \leq
\frac1n\int\abs f \to 0$.

(c) Take $A = \{f > 0\}$: $\int f^+\dd\mu = \int_Af\,\dd\mu =
0$, so $f^+ = 0$ a.e.\ by (a); likewise $f^- = 0$ a.e.
\end{solution}

\begin{solution}{exo:b3:lebesgue:6}
(a) $\mathbf 1_\Q$: discontinuous everywhere, not
Riemann-integrable (\cref{thm:b3:lebesgue:riemann}); its
Lebesgue integral is $\lambda(\Q) = 0$. Thomae: continuous at
every irrational (given $\varepsilon$, only finitely many
rationals in $\intcc01$ have denominator $\leq
1/\varepsilon$; avoid them by a small neighborhood),
discontinuous at rationals (density of irrationals): continuous
a.e., Riemann-integrable, integral $0$ (it vanishes a.e.).
$\mathbf 1_K$, $K$ a fat Cantor set: the discontinuity set is
$\partial K = K$ (closed with empty interior), of measure
$\frac12 > 0$: \emph{not} Riemann-integrable --- yet
Lebesgue-integrable with integral $\lambda(K) = \frac12$.

(b) $\int_{k\pi}^{(k+1)\pi}\frac{\abs{\sin x}}x\dd x \geq
\frac1{(k+1)\pi}\int_{k\pi}^{(k+1)\pi}\abs{\sin x}\dd x =
\frac{2}{(k+1)\pi}$: the series diverges. Convergence of the
improper integral: for $A > \pi$,
\[
\int_\pi^A\frac{\sin x}x\dd x = \Bigl[\frac{-\cos
x}x\Bigr]_\pi^A - \int_\pi^A\frac{\cos x}{x^2}\dd x,
\]
and both terms converge as $A \to \infty$ ($\frac1{x^2}$ is
integrable): conditional convergence without absolute
integrability.
\end{solution}

\begin{solution}{exo:b3:lebesgue:7}
Domination: $\abs{\partial_t(\eu^{-x^2}\cos(tx))} =
\abs{x\eu^{-x^2}\sin(tx)} \leq x\eu^{-x^2}$, integrable and
independent of $t$: \cref{thm:b3:lebesgue:paramdiff} applies
globally,
\[
F'(t) = -\int_0^\infty x\eu^{-x^2}\sin(tx)\,\dd x
= \Bigl[\tfrac12\eu^{-x^2}\sin(tx)\Bigr]_0^\infty
- \frac t2\int_0^\infty\eu^{-x^2}\cos(tx)\dd x
= -\frac t2F(t)
\]
(integration by parts with $\dd v = x\eu^{-x^2}\dd x$). The
linear ODE gives $F(t) = F(0)\eu^{-t^2/4}$; with $F(0) =
\frac{\sqrt\pi}2$ (\cref{pb:b3:lebesgue:1}), the Gaussian
reproduces itself under this cosine transform.
\end{solution}

\begin{solution}{exo:b3:lebesgue:8}
The integral converges: near $0$ the integrand tends to $b - a$
(bounded), and it decays like $\eu^{-ax}$ at infinity. Fix $a$
and view $I(b) = \int_0^\infty\frac{\eu^{-ax} -
\eu^{-bx}}x\dd x$ as a function of $b \in \intco a{+\infty}$.
For all $b \geq a$: $\abs{\partial_b(\text{integrand})} =
\eu^{-bx} \leq \eu^{-ax}$, and $\int_0^\infty\eu^{-ax}\dd x =
\frac1a < \infty$: an integrable dominator. So
\cref{thm:b3:lebesgue:paramdiff} gives $I'(b) =
\int_0^\infty\eu^{-bx}\dd x = \frac1b$, and $I(a) = 0$:
\[
I(b) = \int_a^b\frac{\dd t}t = \ln\frac ba .
\]
(Equivalently, the hint's route: the integrand is
$\int_a^b\eu^{-xt}\dd t \geq 0$ and the interchange is the
continuous analogue of \cref{cor:b3:lebesgue:additivity},
i.e.\ Tonelli --- proved in \cref{ch:b3:product}; the
parameter route stays within this chapter.)
\end{solution}

\begin{solution}{exo:b3:lebesgue:9}
$\nu(\varnothing) = 0$; for disjoint $(A_n)$, $f\mathbf
1_{\bigsqcup A_n} = \sum_nf\mathbf 1_{A_n}$ (pointwise, all
terms $\geq 0$), and \cref{cor:b3:lebesgue:additivity} gives
$\sigma$-additivity. The formula $\int g\,\dd\nu = \int
gf\,\dd\mu$: for $g = \mathbf 1_A$ it is the definition of
$\nu$; for simple $g$, linearity; for $g \geq 0$ measurable,
take simple $s_n \nearrow g$
(\cref{thm:b3:lebesgue:approximation}): $s_nf \nearrow gf$, and
MCT on both sides passes to the limit. (This ``indicator
$\to$ simple $\to$ MCT'' escalator is the standard machine of
the theory.)
\end{solution}

\begin{solution}{exo:b3:lebesgue:10}
(a) $\abs{\sin(tx)} \leq \abs tx$ gives
$\abs{\frac{\sin(tx)}{x(1+x^2)}} \leq \frac{\abs
t}{1+x^2}$: the integral converges, and on $\abs t \leq T$ the
dominator $\frac{T}{1+x^2}$ yields continuity
(\cref{thm:b3:lebesgue:paramcont}). Differentiation:
$\abs{\partial_t} = \abs{\frac{\cos(tx)}{1+x^2}} \leq
\frac1{1+x^2}$, integrable: $f'(t) =
\int_0^\infty\frac{\cos(tx)}{1+x^2}\dd x$ for all $t$. A second
differentiation would require integrating $\frac{x\sin(tx)}{1 +
x^2}$, whose absolute value behaves like $\frac{\abs{\sin(tx)}}
x$ at infinity: \emph{not} integrable --- no dominator exists
and \cref{thm:b3:lebesgue:paramdiff} cannot be applied again.

(b) Admitting $f'(t) = \frac\pi2\eu^{-t}$ for $t > 0$: by
oddness of $f$, $f'$ is even, so $f'(t) =
\frac\pi2\eu^{-\abs t}$ --- which is \emph{not differentiable
at $0$}: $f$ is $\mathcal C^1$ but not $\mathcal C^2$,
confirming that the blocked second differentiation was not a
technical accident. For $t > 0$ the improper integral
$\int_0^\infty\frac{x\sin(tx)}{1+x^2}\dd x$ equals $-f''(t) =
\frac\pi2\eu^{-t}$ (differentiating the admitted formula where
it is legitimate, i.e.\ on $\intoo0\infty$) --- an improper,
non-Lebesgue value.
\end{solution}

\begin{solution}{exo:b3:lebesgue:11}
(a) Let $g_n = (f - f_n)^+$: then $0 \leq g_n \leq f$
(positivity of $f_n$), $g_n \to 0$ a.e., and $f$ is an
integrable dominator: $\int g_n \to 0$ (DCT). Since
$\abs{f_n - f} = 2(f - f_n)^+ - (f - f_n)$,
\[
\int\abs{f_n - f} = 2\int g_n - \Bigl(\int f - \int
f_n\Bigr) \longrightarrow 0 + 0 .
\]

(b) The sliding bump $f_n = \mathbf 1_{\intcc n{n+1}}$ has
$f_n \to 0$ a.e.\ and $\int f_n = 1 \not\to 0$: without the
convergence of integrals, $L^1$ convergence fails (and so
does the hypothesis). The signed example: $f_n \to 0$ at
every $x \neq 0$, $\int f_n = 0$, but $\int\abs{f_n} = 2$:
for signed sequences the theorem is genuinely about
$\abs{f_n}$-type control, and positivity was used exactly in
$g_n \leq f$.

(c) Densities satisfy $\int p_n = 1 = \int p$: the
hypothesis of (a) is automatic, so $p_n \to p$ a.e.\ forces
$\norm{p_n - p}_{L^1} \to 0$ --- and hence convergence of
the probabilities $\int_Ap_n \to \int_Ap$ \emph{uniformly}
over all measurable $A$ ($\abs{\int_A(p_n - p)} \leq
\norm{p_n - p}_1$): Scheff\'e turns pointwise convergence of
densities into total-variation convergence of laws.
\end{solution}

\begin{solution}{exo:b3:lebesgue:12}
(a) On $\intoo0n$, $\varphi_n(x) = n\log(1 - \frac xn)$
increases in $n$ to $-x$ (the map $t \mapsto
\frac{\log(1 - xt)}{t}$ decreases as $t = \frac1n
\downarrow 0$; or expand: $\varphi_{n}' \geq 0$ in $n$ via
$\log(1-u) + \frac{u}{1-u} \geq 0$). So $(1 -
\frac xn)^n\eu^{x/2}\mathbf 1_{x<n} \nearrow
\eu^{-x/2}$, and MCT gives
\[
\lim_n\int_0^n\Bigl(1 - \frac xn\Bigr)^n\eu^{x/2}\dd x =
\int_0^\infty\eu^{-x/2}\dd x = 2 .
\]

(b) Substitute $u = x^n$ ($x = u^{1/n}$, $n x^{n-1}\dd x =
\dd u$):
\[
\int_0^1\frac{nx^{n-1}}{1 + x}\dd x =
\int_0^1\frac{\dd u}{1 + u^{1/n}} .
\]
For $u \in \intoo01$: $u^{1/n} \to 1$, so the integrand
tends to $\frac12$, dominated by $1$: the limit is
$\frac12$ (DCT). (The mass of $nx^{n-1}$ concentrates at $x
= 1$, where $\frac1{1+x} = \frac12$: the substitution makes
the concentration visible.)

(c) Pointwise, $(1 + x/n)^n \nearrow \eu^x$ and $x^{1/n} \to
1$ ($x > 0$): the integrand tends to $\eu^{-x}$. Dominator
for $n \geq 2$: on $\intoc01$, $x^{-1/n} \leq x^{-1/2}$ and
$(1 + x/n)^{-n} \leq 1$: bound $x^{-1/2}$, integrable; on
$\intoo1\infty$, $x^{-1/n} \leq 1$ and $(1 + x/n)^n \geq 1 +
\binom n2\frac{x^2}{n^2} \geq 1 + \frac{x^2}4$: bound
$\frac{4}{4 + x^2}$, integrable. DCT:
\[
\lim_n\int_0^\infty\frac{\dd x}{(1 + x/n)^nx^{1/n}} =
\int_0^\infty\eu^{-x}\dd x = 1 .
\]
\end{solution}

\begin{solution}{pb:b3:lebesgue:1}
\textbf{1.} $A$ is $\mathcal C^1$ by the fundamental theorem
of calculus and the chain rule: $A'(t) =
2\eu^{-t^2}\int_0^t\eu^{-x^2}\dd x$. For $B$:
$\partial_t\bigl(\frac{\eu^{-t^2(1+x^2)}}{1+x^2}\bigr) =
-2t\,\eu^{-t^2(1+x^2)}$, continuous and bounded on $[t_0, T]
\times \intcc01$ for any $0 < t_0 < T$ (bounded domination on a
finite measure space suffices): $B$ is $\mathcal C^1$ on
$\intoo0{+\infty}$ with
\[
B'(t) = -2t\int_0^1 \eu^{-t^2(1+x^2)}\dd x
= -2\eu^{-t^2}\int_0^1 t\,\eu^{-(tx)^2}\dd x
= -2\eu^{-t^2}\int_0^t\eu^{-u^2}\dd u = -A'(t)
\]
(substitution $u = tx$).

\textbf{2.} $A(0) = 0$ and $B(0) = \int_0^1\frac{\dd x}{1+x^2}
= \frac\pi4$. $A + B$ has zero derivative on
$\intoo0{+\infty}$ and is continuous at $0$ ($B$ by
domination $\frac1{1+x^2}$ and
\cref{thm:b3:lebesgue:paramcont}): $A + B \equiv \frac\pi4$.
As $t\to+\infty$: $0 \leq B(t) \leq \eu^{-t^2} \to 0$, and
$A(t) \to \bigl(\int_0^\infty\eu^{-x^2}\dd x\bigr)^2$ (MCT or
just monotone convergence of the inner integral).

\textbf{3.} Hence $\bigl(\int_0^\infty\eu^{-x^2}\dd x\bigr)^2
= \frac\pi4$: $\int_0^\infty\eu^{-x^2}\dd x =
\frac{\sqrt\pi}2$, and by evenness $G = \sqrt\pi$. Also
$\Gamma(\frac12) = \int_0^\infty x^{-1/2}\eu^{-x}\dd x
\overset{x = u^2}{=} 2\int_0^\infty\eu^{-u^2}\dd u =
\sqrt\pi$.

\textbf{4.} For $t > 0$: $\abs{\eu^{-tx}\frac{\sin x}x} \leq
\eu^{-tx}$, integrable. For $t = 0$, improper convergence: on
$[\pi, A]$,
\[
\int_\pi^A\frac{\sin x}x\dd x
= \Bigl[-\frac{\cos x}x\Bigr]_\pi^A -
\int_\pi^A\frac{\cos x}{x^2}\dd x,
\]
both terms convergent as $A \to \infty$; near $0$ the integrand
extends continuously by $1$.

\textbf{5.} On $[t_0, +\infty)$ ($t_0 > 0$):
$\abs{\partial_t\bigl(\eu^{-tx}\tfrac{\sin x}x\bigr)} =
\eu^{-tx}\abs{\sin x} \leq \eu^{-t_0x}$, integrable:
\cref{thm:b3:lebesgue:paramdiff} applies on every such
interval, so on all of $\intoo0{+\infty}$:
\[
F'(t) = -\int_0^\infty\eu^{-tx}\sin x\,\dd x
= -\operatorname{Im}\int_0^\infty\eu^{-(t - \iu)x}\dd x
= -\operatorname{Im}\frac{1}{t - \iu} = -\frac{1}{1 + t^2}.
\]

\textbf{6.} $\abs{F(t)} \leq \int_0^\infty\eu^{-tx}\dd x =
\frac1t \to 0$. Integrating $F' = -\frac1{1+t^2}$: $F(t) = C -
\arctan t$, and $t \to \infty$ forces $C = \frac\pi2$: $F(t) =
\frac\pi2 - \arctan t$ on $\intoo0{+\infty}$.

\textbf{7.} Integrate by parts on $[A, R]$ with $\sin x =
(-\cos x)'$ and let $R \to \infty$:
\[
\int_A^{\infty}\eu^{-tx}\frac{\sin x}x\dd x
= \frac{\cos A\;\eu^{-tA}}{A}
- \int_A^\infty \cos x\,\Bigl(\frac tx +
\frac1{x^2}\Bigr)\eu^{-tx}\dd x .
\]
Bounding $\abs{\cos} \leq 1$: the first term is $\leq \frac1A$;
the integral is at most $\int_A^\infty t\eu^{-tx}\frac{\dd x}x
+ \int_A^\infty\frac{\dd x}{x^2} \leq \frac1A\int_0^\infty
t\eu^{-tx}\dd x + \frac1A = \frac2A$. Total: $\leq \frac 3A$,
uniformly for $t \in [0, 1]$ (the case $t = 0$ included). Now
\[
\abs{F(t) - D} \leq
\Bigl|\int_0^A(\eu^{-tx} - 1)\,\frac{\sin x}x\,\dd x\Bigr| +
\frac6A .
\]
On $[0, A]$: $\abs{\eu^{-tx} - 1} \leq tx \leq tA$ and
$\abs{\frac{\sin x}x} \leq 1$, so the first term is at most
$tA^2$. Choose $A$ with $\frac6A < \varepsilon$, then $t <
\varepsilon/A^2$: $\abs{F(t) - D} < 2\varepsilon$.

\textbf{8.} Therefore $D = \lim_{t\to0^+}F(t) =
\lim_{t\to0^+}\bigl(\frac\pi2 - \arctan t\bigr) =
\frac\pi2$.

\textbf{9.} By parts ($u = \sin^2x$, $v' = x^{-2}$):
\[
\int_0^\infty\frac{\sin^2x}{x^2}\dd x
= \Bigl[-\frac{\sin^2x}{x}\Bigr]_0^\infty +
\int_0^\infty\frac{2\sin x\cos x}{x}\dd x
= \int_0^\infty\frac{\sin 2x}{x}\dd x = D = \frac\pi2
\]
(substitute $u = 2x$ in the last step; boundary terms vanish:
$\sin^2 x/x \to 0$ at both ends).

\textbf{10.} By parts ($u = 1 - \cos x$, $v' = x^{-2}$):
$\int_0^\infty\frac{1 - \cos x}{x^2}\dd x =
\int_0^\infty\frac{\sin x}x\dd x = \frac\pi2$. Consistency:
$1 - \cos x = 2\sin^2\frac x2$, and the substitution $x = 2u$
turns $\int\frac{2\sin^2(x/2)}{x^2}\dd x$ into
$\int\frac{\sin^2u}{u^2}\dd u$: the two computations agree.

\textbf{11.} $\int_0^\infty\frac{\sin(ax)}x\dd x = \frac\pi2$
for every $a > 0$ (substitute $u = ax$: the integral is
scale-invariant); $\int_\R\eu^{-ax^2}\dd x = \sqrt{\pi/a}$
(substitute $u = \sqrt a\,x$). Scaling in the argument leaves
the Dirichlet integral fixed and divides the Gaussian by
$\sqrt a$.

\textbf{12.} A dominator valid for all $t \in [0,1]$ must
dominate $\sup_{t\in[0,1]}\abs{\eu^{-tx}\frac{\sin
x}x} = \abs{\frac{\sin x}x}$, which is not integrable
(\cref{exo:b3:lebesgue:6}): DCT cannot cross $t = 0$. The
substitute of question 7 --- tails uniformly small in the
parameter, compact part handled by DCT --- is the standard
``uniform integrability'' pattern, and reappears whenever
conditional convergence meets limit interchange.

\textbf{13.} If $g = \log f$ is convex then $f = \exp\circ g$
is convex (exp is convex increasing: $f(\lambda x + (1-\lambda)
y) \leq \eu^{\lambda g(x) + (1-\lambda)g(y)} \leq \lambda f(x)
+ (1-\lambda)f(y)$, the last step by convexity of exp between
the points $g(x), g(y)$). Products and affine substitutions:
logarithms turn them into sums and affine substitutions of
convex functions. H\"older ($\frac1p + \frac1q = 1$): for
$\int\abs u^p = \int\abs v^q = 1$, Young gives $\abs{uv} \leq
\frac{\abs u^p}p + \frac{\abs v^q}q$, integrate: $\int\abs{uv}
\leq 1$; the general case by homogeneity. Then, for $\lambda
\in \intoo01$, apply it with $p = \frac1\lambda$ to the
factorization
\[
t^{\lambda x + (1-\lambda)y - 1}\eu^{-t}
= \bigl(t^{x-1}\eu^{-t}\bigr)^{\lambda}
\bigl(t^{y-1}\eu^{-t}\bigr)^{1-\lambda} :
\qquad
\Gamma(\lambda x + (1{-}\lambda)y) \leq
\Gamma(x)^\lambda\,\Gamma(y)^{1-\lambda} .
\]

\textbf{14.} For a convex $g$, the slope of a chord increases
with its endpoints (three-chord inequality). Comparing the
chords over $[n-1, n]$, $[n, n+x]$, $[n, n+1]$:
\[
\log(n-1) = \frac{g(n) - g(n-1)}1 \leq
\frac{g(n+x) - g(n)}x \leq \frac{g(n+1) - g(n)}1 = \log n,
\]
and multiplying by $x > 0$ gives the claim.

\textbf{15.} With $f(n) = (n-1)!$ (recursion from $f(1) = 1$)
and $f(n + x) = x(x+1)\cdots(x+n-1)\,f(x)$, question 14 reads
\[
(n-1)^x\,(n-1)! \;\leq\; x(x+1)\cdots(x+n-1)\,f(x)
\;\leq\; n^x\,(n-1)! .
\]
The upper bound rewrites as $f(x) \leq
\frac{n!\,n^x}{x(x+1)\cdots(x+n)}\cdot\frac{x+n}n$, and the
lower bound at rank $n+1$ as $f(x) \geq
\frac{n!\,n^x}{x(x+1)\cdots(x+n)}$. The correction factor
$\frac{x+n}n \to 1$: the sandwich forces
\[
f(x) = \lim_n\frac{n!\,n^x}{x(x+1)\cdots(x+n)}
\qquad (x \in \intoc01),
\]
an expression independent of $f$: uniqueness on $\intoc01$,
hence everywhere by the recursion. Since $\Gamma$ satisfies
all three hypotheses (question 13), $f = \Gamma$ and Gauss's
formula holds --- for all $x > 0$, as both sides obey the same
recursion.

\textbf{16.} Near $0$, the integrand is $\sim t^{x-1}$,
integrable iff $x > 0$; near $1$, symmetric with $y$.
Integration by parts on $\intcc\varepsilon{1-\varepsilon}$,
letting $\varepsilon \to 0$ (boundary terms vanish for $x, y >
0$):
\[
B(x{+}1, y) = \Bigl[-t^x\frac{(1-t)^y}y\Bigr]_0^1 +
\frac xy\int_0^1t^{x-1}(1-t)^y\,\dd t
= \frac xy\bigl(B(x, y) - B(x{+}1, y)\bigr),
\]
using $(1-t)^y = (1-t)^{y-1}(1 - t)$; solving, $B(x+1, y) =
\frac{x}{x+y}B(x, y)$. And $B(1, y) =
\int_0^1(1-t)^{y-1}\dd t = \frac1y$.

\textbf{17.} $f(1) = \frac1y\cdot\frac{\Gamma(1+y)}{\Gamma(y)}
= 1$; $f(x+1) = \frac{x}{x+y}B(x,y)\cdot\frac{(x+y)\Gamma(x+y)}
{\Gamma(y)} = x\,f(x)$; and $f$ is log-convex in $x$ as a
product of the log-convex $B(\cdot, y)$ (H\"older on the
factorization $t^{(\lambda x_1 + (1-\lambda)x_2)-1}(1-t)^{y-1}
= (\cdots)^\lambda(\cdots)^{1-\lambda}$, as in question 13)
and $\Gamma(\cdot + y)$ (affine shift). Bohr--Mollerup: $f =
\Gamma$, i.e.\ $B(x, y) =
\frac{\Gamma(x)\Gamma(y)}{\Gamma(x+y)}$.

\textbf{18.} With $t = \sin^2\theta$, $\dd t =
2\sin\theta\cos\theta\,\dd\theta$ and $t^{-1/2}(1-t)^{-1/2} =
\frac1{\sin\theta\cos\theta}$:
\[
B\Bigl(\frac12, \frac12\Bigr) =
\int_0^{\pi/2}2\,\dd\theta = \pi
= \frac{\Gamma(\frac12)^2}{\Gamma(1)}
\quad\Longrightarrow\quad
\Gamma\Bigl(\frac12\Bigr) = \sqrt\pi .
\]
Part I reached the same constant by differentiating a
parameter and racing two functions to their limits; here
convexity alone rigidified the problem until only one value
survived. Analysis by motion versus analysis by shape.

\textbf{19.} $g(1) = \frac{2^0}{\sqrt\pi}\Gamma(\frac12)
\Gamma(1) = 1$. Recursion:
\[
g(x+1) = \frac{2^{x}}{\sqrt\pi}\,
\Gamma\Bigl(\frac{x+1}2\Bigr)\Gamma\Bigl(\frac x2 + 1\Bigr)
= \frac{2^{x}}{\sqrt\pi}\cdot\frac x2\,
\Gamma\Bigl(\frac x2\Bigr)\Gamma\Bigl(\frac{x+1}2\Bigr)
= x\,g(x) .
\]
Log-convexity: product of $\eu^{(x-1)\log2}$ (log-affine) and
two affine reparametrizations of the log-convex $\Gamma$.
Bohr--Mollerup gives $g = \Gamma$; setting $x = 2z$:
$\Gamma(2z) = \frac{2^{2z-1}}{\sqrt\pi}\Gamma(z)\Gamma(z +
\frac12)$.

\textbf{20.} From $\Gamma(\frac12) = \sqrt\pi$ and the
recursion, $\Gamma(n + \frac12) = (n - \frac12)(n -
\frac32)\cdots\frac12\,\sqrt\pi =
\frac{(2n-1)(2n-3)\cdots1}{2^n}\sqrt\pi =
\frac{(2n)!}{4^nn!}\sqrt\pi$ (complete the odd product with
the evens). Asymptotics: question 14 with $g = \log\Gamma$
and $x = \frac12$ gives $\sqrt{n-1} \leq
\frac{\Gamma(n+\frac12)}{\Gamma(n)} \leq \sqrt n$, so the
ratio to $\sqrt n$ is squeezed between $\sqrt{1 - \frac1n}$
and $1$.

\textbf{21.} From question 20, $(2n)! =
\frac{4^nn!}{\sqrt\pi}\Gamma(n + \tfrac12)$, so
\[
\binom{2n}n = \frac{(2n)!}{(n!)^2}
= \frac{4^n\,\Gamma(n+\frac12)}{\sqrt\pi\;n!}
= \frac{4^n}{\sqrt\pi}\cdot
\frac{\Gamma(n+\frac12)}{n\,\Gamma(n)}
\sim \frac{4^n}{\sqrt\pi}\cdot\frac{\sqrt n}{n}
= \frac{4^n}{\sqrt{\pi n}} .
\]
Numerically, $4^{10}/\sqrt{10\pi} = 1048576/5.6050 \approx
187078.6$, against $\binom{20}{10} = 184756$: ratio $0.9876$
--- the error is $O(1/n)$, visible at $n = 10$.

\textbf{22.} Equivalences: $\Gamma(\frac12) = \sqrt\pi
\Leftrightarrow G = \sqrt\pi$ (the substitution $x = u^2$ of
question 3) $\Leftrightarrow B(\frac12, \frac12) = \pi$
(Euler's formula); duplication at $z = n$ \emph{is} the
closed form of $\Gamma(n + \frac12)$, which \emph{is} the
central binomial estimate up to the slope lemma. The
parameter technique (Part I--II) computes limits of moving
quantities and is indispensable when a genuine deformation is
present (Dirichlet's integral has no convexity proof); the
convexity technique computes nothing but forbids everything
--- it excels at uniqueness and functional equations
(Gauss, Euler, Legendre in three strokes), where
differentiation would drown in computation. A complete
analyst carries both.

\textbf{23.} With $t = \sin^2\theta$, $\dd t =
2\sin\theta\cos\theta\,\dd\theta =
2\,t^{1/2}(1-t)^{1/2}\,\dd\theta$, so
\[
W_p = \int_0^1 t^{p/2}\,
\frac{\dd t}{2\,t^{1/2}(1-t)^{1/2}}
= \frac12\int_0^1 t^{\frac{p+1}2 - 1}(1-t)^{\frac12 - 1}\dd t
= \frac12\,B\Bigl(\frac{p+1}2, \frac12\Bigr).
\]
The Beta recursion with $x = \frac{n+1}2$, $y = \frac12$
gives
\[
W_{n+2} = \frac12\,
\frac{(n+1)/2}{(n+2)/2}\,B\Bigl(\frac{n+1}2, \frac12\Bigr)
= \frac{n+1}{n+2}\,W_n .
\]
Starting from $W_0 = \frac\pi2$, $W_1 = 1$:
\[
W_{2n} = \frac\pi2\prod_{k=1}^n\frac{2k-1}{2k},
\qquad
W_{2n+1} = \prod_{k=1}^n\frac{2k}{2k+1} .
\]
Since $\sin^{n+1} \leq \sin^n$ on $\intcc0{\pi/2}$, the
sequence $(W_n)$ is nonincreasing, so
\[
1 \geq \frac{W_{2n+1}}{W_{2n}} \geq
\frac{W_{2n+1}}{W_{2n-1}} = \frac{2n}{2n+1}
\longrightarrow 1 .
\]
But the closed forms give
\[
\frac{W_{2n+1}}{W_{2n}} = \frac2\pi
\prod_{k=1}^n\frac{(2k)^2}{(2k-1)(2k+1)}
= \frac2\pi\prod_{k=1}^n\frac{4k^2}{4k^2-1},
\]
and letting $n \to \infty$ yields Wallis's product. (Via
Euler's formula, $W_p = \frac{\sqrt\pi}2\,
\Gamma(\frac{p+1}2)/\Gamma(\frac p2 + 1)$: Wallis is the
Gaussian integral in yet another costume.)

\textbf{24.} The $b$-derivative of the integrand is
$-2x\,\eu^{-x^2}\sin(2bx)$, dominated by $2\abs
x\,\eu^{-x^2} \in L^1(\R)$ uniformly in $b$: $\Phi$ is
$\mathcal C^1$ with
\[
\Phi'(b) = -\int_{-\infty}^{+\infty}
2x\,\eu^{-x^2}\sin(2bx)\,\dd x .
\]
Integrating by parts with $u = \sin(2bx)$, $\dd v =
-2x\,\eu^{-x^2}\dd x$ (so $v = \eu^{-x^2}$), the boundary
terms vanish and
\[
\Phi'(b) = -2b\int_{-\infty}^{+\infty}
\eu^{-x^2}\cos(2bx)\,\dd x = -2b\,\Phi(b).
\]
Hence $\bigl(\Phi(b)\,\eu^{b^2}\bigr)' = 0$ and $\Phi(b) =
\Phi(0)\,\eu^{-b^2} = \sqrt\pi\,\eu^{-b^2}$ by Part I. Up to
normalization this says the Fourier transform of
$\eu^{-x^2}$ is again a Gaussian --- the fixed point on which
the inversion theory of \cref{ch:b3:fouriertransform} pivots.

\textbf{25.} For $0 < a < b$ and $x > 0$,
\[
0 \leq \frac{\eu^{-ax} - \eu^{-bx}}{x}
= \int_a^b \eu^{-sx}\,\dd s \leq (b - a)\,\eu^{-ax},
\]
so the integral $I(a)$ converges (Lebesgue); the pointwise
bound also shows the integrand extends continuously by $b -
a$ at $x = 0$. Fix $b$; on $\intco{a_0}{+\infty}$ the
$a$-derivative of the integrand is $-\eu^{-ax}$, dominated by
$\eu^{-a_0x} \in L^1(\intoo0{+\infty})$, so $I$ is $\mathcal
C^1$ on $\intoo0b$ with
\[
I'(a) = -\int_0^{+\infty}\eu^{-ax}\dd x = -\frac1a,
\qquad\text{hence}\qquad
I(a) = -\log a + c .
\]
The two-sided bound gives $0 \leq I(a) \leq (b-a)/a \to 0$ as
$a \to b^-$, so $c = \log b$ and $I(a) = \log\frac ba$. The
removable singularity is needed \emph{at $0$}: each term
$\eu^{-ax}/x$ separately has a divergent (logarithmic)
integral near $0$, and only the first-order cancellation
$\eu^{-ax} - \eu^{-bx} = O(x)$ makes the difference
integrable there; at infinity each term is harmless on its
own.
\end{solution}
