\section*{Chapter \ref{ch:b3:lp} --- The $L^p$ Spaces}

\begin{solution}{exo:b3:lp:1}
(a) $p = q = 2$ in \cref{thm:b3:lp:holder}: $\abs{\int f\bar
g\,\dd\mu} \leq \int\abs{fg} \leq \norm f_2\norm g_2$ ---
Cauchy--Schwarz, with equality iff $\abs f, \abs g$
proportional and the phases aligned.

(b) On a probability space, for $p \leq q$: apply Jensen
(\cref{exo:b3:lp:10}) with the convex $\Phi(t) =
\abs t^{q/p}$ to the function $\abs f^p$:
$\bigl(\int\abs f^p\bigr)^{q/p} \leq \int\abs f^q$, i.e.\
$\norm f_p \leq \norm f_q$.

(c) $\int\abs{fg} = \norm f_1\norm g_\infty$ iff $\abs{g} =
\norm g_\infty$ a.e.\ on $\{f \neq 0\}$ (the inequality
$\abs{fg} \leq \abs f\norm g_\infty$ must be an a.e.\
equality).
\end{solution}

\begin{solution}{exo:b3:lp:2}
$\int_0^1 x^{-p/2}\dd x < \infty$ iff $p < 2$: the first is in
$L^p$ for $p \in \intco12$. $\int_1^\infty x^{-p/2}\dd x <
\infty$ iff $p > 2$: the second for $p \in \intoo2\infty$ (and
$p = \infty$: it is bounded --- include it). Third: for $p <
2$, dominated by $x^{-p/2}$: \omterm{def:b3:lebesgue:l1}{integrable}; for $p = 2$,
substitute $u = -\ln x$: $\int_0^1\frac{\dd x}{x(1 + \abs{\ln
x})^2} = \int_0^\infty\frac{\dd u}{(1 + u)^2} < \infty$; for
$p > 2$ the power dominates: divergent. So $p \in \intcc12$.
Fourth: $\int_\R\frac{\dd x}{(1 + \abs x)^p} < \infty$ iff $p >
1$; bounded, so also $L^\infty$: $p \in
\intoc1\infty$. Moral: \omterm{def:b3:lebesgue:l1}{integrability} at $0$ likes small $p$, at
$\infty$ large $p$; combining both obstructions, no inclusion
between $L^p(\R)$ spaces.
\end{solution}

\begin{solution}{exo:b3:lp:3}
(a) $\norm{f_n}_p^p = \lambda(I_n) \to 0$ (at dyadic level $k$
the length is $2^{-k}$). But every $x$ lies in one interval of
each dyadic level: $f_n(x) = 1$ infinitely often and $= 0$
infinitely often (intervals of the same level not containing
$x$): no convergence at any point.

(b) $f_{n_k} = \mathbf 1_{\intcc0{2^{-k}}}$ (the first interval
of each level) converges to $0$ at every $x > 0$: a.e.

(c) a.e.\ but not $L^1$: $n\mathbf 1_{\intoo0{1/n}} \to 0$
a.e., integral $1$. In $L^1$ but in no $L^p$ ($p > 1$): $g_n =
\eu^n\,\mathbf 1_{(0,\ \eu^{-n}/n)}$: $\norm{g_n}_1 = \frac1n
\to 0$, while $\norm{g_n}_p^p = \eu^{(p-1)n}/n \to \infty$ for
every $p > 1$.
\end{solution}

\begin{solution}{exo:b3:lp:4}
Upper: $\norm f_p \leq \mu(X)^{1/p}\norm f_\infty$
(\cref{prop:b3:lp:inclusions}(a) with $q = \infty$), and
$\mu(X)^{1/p} \to 1$. Lower: for $\varepsilon > 0$, $A =
\{\abs f > \norm f_\infty - \varepsilon\}$ has $\mu(A) > 0$
(definition of the essential sup), and
\[
\norm f_p \geq \Bigl(\int_A \abs f^p\Bigr)^{1/p}
\geq (\norm f_\infty - \varepsilon)\,\mu(A)^{1/p}
\xrightarrow[p\to\infty]{} \norm f_\infty - \varepsilon .
\]
\end{solution}

\begin{solution}{exo:b3:lp:5}
(a) Twice: the conversion $\norm{\tau_hg - g}_p \leq
C^{1/p}\norm{\tau_hg - g}_\infty$ (\omterm{def:b3:measure:measure}{finite-measure} support)
degenerates for $p = \infty$ only in that the \emph{\omterm{ex:b3:lebesgue:gamma}{density} of
$\mathcal C_c$} fails there --- that is the real gap: step (2)
of \cref{thm:b3:lp:density} has no $L^\infty$ analogue.

(b) If $f$ has a uniformly \omterm{def:b3:topology:continuity}{continuous} representative $g$:
$\norm{\tau_hf - f}_\infty = \sup_x\abs{g(x - h) - g(x)} \to
0$. Conversely, suppose $\norm{\tau_hf - f}_\infty \to 0$. The
mollifications $f_\varepsilon = f * \rho_\varepsilon$ are
\omterm{def:b3:topology:continuity}{continuous}, and
\[
\norm{f_\varepsilon - f}_\infty
\leq \sup_{\norm y \leq \varepsilon}\norm{\tau_yf -
f}_\infty \longrightarrow 0
\]
(the convolution is an average of translates). Each
$f_\varepsilon$ is uniformly \omterm{def:b3:topology:continuity}{continuous}
($\norm{\tau_hf_\varepsilon - f_\varepsilon}_\infty \leq
\norm{\tau_hf - f}_\infty$, by averaging), and a uniform limit
of uniformly \omterm{def:b3:topology:continuity}{continuous} functions is one: $f$ agrees a.e.\
with a uniformly \omterm{def:b3:topology:continuity}{continuous} function.
\end{solution}

\begin{solution}{exo:b3:lp:6}
By H\"older, for every $x$ the integrand $y \mapsto f(x -
y)g(y)$ is in $L^1$ with $\abs{(f*g)(x)} \leq \norm f_p\norm
g_q$: everywhere defined and bounded. Uniform \omterm{def:b3:topology:continuity}{continuity} ($p <
\infty$):
\[
\abs{(f*g)(x + h) - (f*g)(x)}
= \Bigl|\int\bigl(\tau_{-h}f - f\bigr)(x - y)\,g(y)\dd y\Bigr|
\leq \norm{\tau_{-h}f - f}_p\,\norm g_q
\xrightarrow[h\to0]{} 0,
\]
uniformly in $x$ (\cref{thm:b3:lp:density}(3)). If $p =
\infty$, then $q = 1$: write $f * g = g * f$ and run the same
bound with the translation acting on $g \in L^1$.
\end{solution}

\begin{solution}{exo:b3:lp:7}
Fix $\chi \in \mathcal C_c^\infty(\intoo ab)$ with $\int\chi =
1$, and set $c = \int f\chi$. Let $\varphi \in \mathcal
C_c^\infty$ be arbitrary and $\psi = \varphi -
\bigl(\int\varphi\bigr)\chi$: then $\int\psi = 0$, so $\Phi(x)
= \int_a^x\psi$ defines $\Phi \in \mathcal C_c^\infty(\intoo
ab)$ (it vanishes near both ends: near $a$ trivially, near $b$
because the total integral is $0$) with $\Phi' = \psi$. The
hypothesis gives $\int f\psi = \int f\Phi' = 0$, hence
\[
\int f\varphi = \Bigl(\int\varphi\Bigr)\int f\chi = \int
c\,\varphi
\quad\text{for every } \varphi:
\qquad \int(f - c)\varphi = 0 .
\]
By \cref{cor:b3:lp:fundlemma} (localized on $\intoo ab$), $f =
c$ a.e.
\end{solution}

\begin{solution}{exo:b3:lp:8}
Let $3\delta < d(K, \R^d\setminus U)$ (positive:
\cref{exo:b3:topology:6}(b); if $U = \R^d$ any $\delta$
works), $K_\delta = \{x : d(x, K) \leq \delta\}$, and
\[
\varphi = \mathbf 1_{K_\delta} * \rho_{\delta/2} .
\]
Then $\varphi \in \mathcal C^\infty$
(\cref{thm:b3:lp:regularization}(1); the indicator is $L^1$),
$0 \leq \varphi \leq 1$ ($\int\rho = 1$), $\varphi = 1$ on $K$
(for $x \in K$, $\bar B(x, \delta/2) \subseteq K_\delta$, so
the convolution integrates $\rho$ fully), and
$\operatorname{supp}\varphi \subseteq K_{3\delta/2} \subseteq
U$: compactly supported ($K_\delta$ is bounded). Partition of
unity: given $K \subseteq U_1\cup\dots\cup U_m$, choose (by
compactness) \omterm{def:b3:topology:compact}{compacts} $K_i \subseteq U_i$ with $K \subseteq
\bigcup \mathring K_i$, take $\varphi_i$ as above for $(K_i,
U_i)$, and set $\psi_i = \varphi_i\prod_{j <i}(1 -
\varphi_j)$: each $\psi_i \in \mathcal C_c^\infty(U_i)$, and
$\sum_i\psi_i = 1 - \prod_i(1 - \varphi_i) = 1$ on $K$.
\end{solution}

\begin{solution}{exo:b3:lp:9}
(a) The interpolation exponent for $(p_0, q_0) = (1, \infty)$
at $r = p$ is $\alpha = \frac1p$:
\cref{prop:b3:lp:inclusions}(c) gives $\norm f_p \leq \norm
f_1^{1/p}\norm f_\infty^{1 - 1/p}$.

(b) Taking logarithms in
\cref{prop:b3:lp:inclusions}(c): $\ln\norm f_r \leq
\alpha\ln\norm f_p + (1-\alpha)\ln\norm f_q$ where $\frac1r$
is the same convex combination of $\frac1p, \frac1q$: $\frac1p
\mapsto \ln\norm f_p$ is convex. Example with $L^p$-membership
exactly on $\intoo{p_0}{p_1}$:
\[
f(x) = x^{-1/p_1}\,\mathbf 1_{\intoo01}(x) +
x^{-1/p_0}\,\mathbf 1_{\intco1\infty}(x):
\]
the first term is $L^p$ iff $p < p_1$, the second iff $p >
p_0$.
\end{solution}

\begin{solution}{exo:b3:lp:10}
Let $m = \int f\,\dd\mu \in \R$. Convexity provides a support
line at $m$: there is $s$ with $\Phi(t) \geq \Phi(m) + s(t -
m)$ for all $t$ (take $s$ between the one-sided derivatives,
which exist for convex functions). Substitute $t = f(x)$ and
integrate against the probability $\mu$:
\[
\int\Phi\circ f\,\dd\mu \geq \Phi(m) + s\Bigl(\int f - m\Bigr)
= \Phi\Bigl(\int f\,\dd\mu\Bigr)
\]
(\omterm{def:b3:lebesgue:measurable}{measurability}: $\Phi$ is \omterm{def:b3:topology:continuity}{continuous}; \omterm{def:b3:lebesgue:l1}{integrability} of the
negative part of $\Phi\circ f$ is guaranteed by the support
line). AM--GM: on a finite set with weights $w_i$, take
$\Phi = \exp$ and $f = \sum(\ln a_i)\mathbf 1_i$:
$\exp\bigl(\sum w_i\ln a_i\bigr) \leq \sum w_ia_i$, i.e.\
$\prod a_i^{w_i} \leq \sum w_ia_i$. The norm monotonicity is
\cref{exo:b3:lp:1}(b).
\end{solution}

\begin{solution}{exo:b3:lp:11}
(a) Assume first $p, q, r < \infty$ and $f, g \geq 0$
(replace by absolute values). The three exponents $r$,
$\alpha = \frac{pr}{r-p}$, $\beta = \frac{qr}{r-q}$
satisfy $\frac1r + \frac1\alpha + \frac1\beta = \frac1r +
\frac1p - \frac1r + \frac1q - \frac1r = 1$ (the scaling
relation). Split, for fixed $x$,
\[
f(y)g(x-y) = \bigl[f(y)^pg(x-y)^q\bigr]^{1/r}\cdot
f(y)^{1 - p/r}\cdot g(x-y)^{1 - q/r},
\]
and H\"older with the three exponents gives
\[
(f*g)(x) \leq \Bigl(\int f^pg(x-\cdot)^q\Bigr)^{1/r}
\norm f_p^{\,p/\alpha\cdot\alpha/p}\cdots
\]
more precisely: the second factor is
$\bigl(\int f^{(1-p/r)\alpha}\bigr)^{1/\alpha} =
\norm f_p^{p(1/p - 1/r)}$ since $(1 - \frac pr)\alpha = p$,
and likewise the third is $\norm g_q^{q(1/q - 1/r)}$. Raise
to the $r$-th power and integrate in $x$ (Tonelli on the
first factor):
\[
\norm{f*g}_r^r \leq \norm f_p^p\,\norm g_q^q\cdot
\norm f_p^{\,rp(1/p - 1/r)}\,\norm g_q^{\,rq(1/q - 1/r)}
= \norm f_p^r\,\norm g_q^r .
\]
The endpoint cases ($r = \infty$ or an exponent equal to
its bound) are plain H\"older or direct estimates.

(b) $r = \infty$ forces $q = p'$: $\abs{f*g(x)} \leq \norm
f_p\norm g_{p'}$ --- H\"older after translation-reflection.
$q = 1$ gives $r = p$: $\norm{f*g}_p \leq \norm
g_1\norm f_p$, the mollification workhorse
(\cref{thm:b3:lp:regularization}'s engine). $p = q = 1$
gives $r = 1$: the convolution algebra
(\cref{thm:b3:product:convolution}).

(c) Replace $f, g$ by $f_\lambda = f(\lambda\cdot)$,
$g_\lambda = g(\lambda\cdot)$: then $f_\lambda*g_\lambda =
\lambda^{-d}(f*g)(\lambda\cdot)$, and comparing norms,
\[
\text{LHS} \sim \lambda^{-d - d/r}, \qquad
\text{RHS} \sim \lambda^{-d/p - d/q} :
\]
an inequality valid for all $f, g$ forces the two scaling
exponents to match, i.e.\ $1 + \frac1r = \frac1p + \frac1q$
exactly. Any other combination dies at $\lambda \to 0$ or
$\infty$: Young's relation is not a convenience but a
scaling law.
\end{solution}

\begin{solution}{exo:b3:lp:12}
(a) Normalize $\norm f_p = \norm g_q = 1$. The proof of
H\"older integrates Young's inequality $\abs{fg} \leq
\frac{\abs f^p}p + \frac{\abs g^q}q$; equality of the
integrals forces equality a.e.\ in Young, which (strict
convexity of $\exp$; equality iff $a^p = b^q$) means $\abs
f^p = \abs g^q$ a.e. Undoing the normalization:
$\abs f^p\norm g_q^q = \abs g^q\norm f_p^p$ a.e.\ ---
proportionality.

(b) Minkowski is two H\"olders applied to $\abs{f + g}^{p-1}
\abs f$ and $\abs{f+g}^{p-1}\abs g$; equality forces (a)'s
proportionality in both: $\abs f^p$ and $\abs g^p$ each
proportional to $\abs{f+g}^{(p-1)q} = \abs{f+g}^p$, so
$\abs g = t\abs f$ a.e.\ for a constant $t \geq 0$; and the
initial pointwise triangle inequality $\abs{f + g} \leq
\abs f + \abs g$ must also be an a.e.\ equality, which for
complex values means $f$ and $g$ have a.e.\ the same
argument where both are nonzero. Combining: $g = tf$ a.e.,
$t > 0$ (both nonzero).

(c) $p = 1$: equality in $\int\abs{f + g} = \int\abs f +
\int\abs g$ holds whenever $f, g$ have the same sign
pattern (same argument a.e.) --- no proportionality needed:
$f = \mathbf 1_{\intcc01}$ and $g = \mathbf 1_{\intcc02}$
work. $p = \infty$: $\norm{f+g}_\infty = \norm f_\infty +
\norm g_\infty$ as soon as the two functions peak
compatibly at a common point (or along a common sequence):
$f = g\,$ near one point suffices regardless of behavior
elsewhere. The strict convexity of the $L^p$ balls for $1 <
p < \infty$ --- and its failure at the endpoints --- is
exactly what these equality cases witness.
\end{solution}

\begin{solution}{pb:b3:lp:1}
\textbf{1.} $f$ has support in some $[\alpha, \beta]
\subseteq \intoo0\infty$, so $F = 0$ on $[0, \alpha]$ and $F
\equiv F(\beta)$ on $[\beta, \infty)$: $Hf$ vanishes near $0$
and is $O(1/x)$ at infinity; $\int_1^\infty x^{-p}\dd x <
\infty$ for $p > 1$, and $Hf$ is \omterm{def:b3:topology:continuity}{continuous}: $Hf \in L^p$.

\textbf{2.} $\bigl(x^{1-p}F(x)^p\bigr)' = (1-p)x^{-p}F^p +
p\,x^{1-p}F^{p-1}f$. Both boundary values vanish: at $0$
because $F = 0$ near $0$; at $\infty$ because $x^{1-p}F^p \leq
F(\beta)^p x^{1-p} \to 0$ ($p > 1$). Integrating the identity
over $\intoo0\infty$:
\[
0 = (1 - p)\int_0^\infty\Bigl(\frac Fx\Bigr)^p\dd x
+ p\int_0^\infty\Bigl(\frac Fx\Bigr)^{p-1}f(x)\,\dd x,
\]
which is the displayed relation.

\textbf{3.} H\"older with exponents $q = \frac p{p-1}$ and $p$:
\[
\int\Bigl(\frac Fx\Bigr)^{p-1}f
\leq \Bigl(\int\Bigl(\frac
Fx\Bigr)^{p}\Bigr)^{1 - 1/p}\,\norm f_p ,
\]
so $\norm{Hf}_p^p \leq \frac p{p-1}\norm{Hf}_p^{p-1}\norm
f_p$; divide (finite by question 1, and if $0$ there is
nothing to prove).

\textbf{4.} Let $f \in L^p$, $f \geq 0$. Choose $\varphi_k \in
\mathcal C_c(\intoo0\infty)$ with $\varphi_k \to f$ in $L^p$
(\cref{thm:b3:lp:density}(2), intersected with the open
half-line --- approximate $f\mathbf 1_{[1/k, k]}$ and
diagonalize), and replace $\varphi_k$ by $\abs{\varphi_k}$
(still \omterm{def:b3:topology:continuity}{continuous}, and closer to $f \geq 0$). For every fixed
$x > 0$, H\"older on $\intoo0x$ gives
\[
\abs{H\varphi_k(x) - Hf(x)} \leq
\frac1x\,x^{1 - 1/p}\,\norm{\varphi_k - f}_p \to 0 :
\]
$H\varphi_k \to Hf$ pointwise. Fatou and question 3:
\[
\int (Hf)^p \leq \liminf_k\int(H\varphi_k)^p
\leq \Bigl(\frac p{p-1}\Bigr)^p\liminf_k\norm{\varphi_k}_p^p
= \Bigl(\frac p{p-1}\Bigr)^p\norm f_p^p .
\]
For signed or complex $f$: $\abs{Hf} \leq H\abs f$ pointwise,
and the nonnegative case applies to $\abs f$.

\textbf{5.} $\norm{f_A}_p^p = \int_1^A t^{-1}\dd t = \ln A$.
For $1 \leq x \leq A$:
\[
(Hf_A)(x) = \frac1x\int_1^x t^{-1/p}\dd t
= \frac{x^{1 - 1/p} - 1}{(1 - \tfrac1p)\,x}
= \frac p{p-1}\,x^{-1/p}\bigl(1 - x^{-(1 - 1/p)}\bigr).
\]

\textbf{6.} Fix $\varepsilon > 0$ and $X_0$ with $(1 -
x^{-(1-1/p)})^p \geq 1 - \varepsilon$ for $x \geq X_0$. Then
\[
\norm{Hf_A}_p^p \geq \int_{X_0}^A\Bigl(\frac
p{p-1}\Bigr)^p\frac{1 - \varepsilon}{x}\,\dd x
= \Bigl(\frac p{p-1}\Bigr)^p(1 - \varepsilon)\,(\ln A - \ln
X_0),
\]
so $\dfrac{\norm{Hf_A}_p^p}{\norm{f_A}_p^p} \geq \bigl(\frac
p{p-1}\bigr)^p(1 - \varepsilon)\bigl(1 - \frac{\ln X_0}{\ln
A}\bigr) \to \bigl(\frac p{p-1}\bigr)^p(1 - \varepsilon)$ as
$A \to \infty$: the constant cannot be improved.

\textbf{7.} Equality in question 3 forces equality in
H\"older: $f^p$ proportional to $\bigl(\frac Fx\bigr)^{(p-1)q}
= \bigl(\frac Fx\bigr)^p$ a.e., i.e.\ $f = \gamma\,\frac Fx$
a.e.\ for some $\gamma \geq 0$. Since $F(x) = \int_0^xf$ is
absolutely \omterm{def:b3:topology:continuity}{continuous} with $F' = f$ a.e., $F$ solves $F' =
\gamma F/x$: on any interval where $F > 0$, $(\ln F)' =
\gamma/x$, so $F = c\,x^{\gamma}$ and $f = c\gamma
x^{\gamma - 1}$ there. But no nonzero power $x^{\gamma-1}$
belongs to $L^p(\intoo0\infty)$ ($p(\gamma - 1) < -1$ needed
at $\infty$ and $> -1$ at $0$: incompatible), and $F$ cannot
vanish identically unless $f = 0$. So equality demands $f =
0$.

\textbf{8.} Given $(a_n)$ nonincreasing $\geq 0$, define the
step function $g(t) = a_{\lceil t\rceil}$ on
$\intoo0{+\infty}$: nonincreasing, with $\int_0^\infty g^p =
\sum_na_n^p$ and $\int_0^ng = a_1 + \dots + a_n$, so
$(Hg)(n) = \frac{a_1 + \dots + a_n}n$. The average of a
nonincreasing function is nonincreasing, so $Hg$ is, and
\[
\sum_{n\geq1}\Bigl(\frac{a_1{+}\dots{+}a_n}n\Bigr)^p
= \sum_{n\geq1}(Hg)(n)^p
\leq \sum_{n\geq1}\int_{n-1}^n (Hg)(t)^p\,\dd t
= \norm{Hg}_p^p
\leq \Bigl(\frac p{p-1}\Bigr)^p\sum_n a_n^p
\]
by Part I. For a general nonnegative sequence, let $(a_n^*)$
be its nonincreasing rearrangement (possible when $a_n \to 0$,
which we may assume --- otherwise both sides are infinite):
the right side is unchanged, and each partial sum $a_1 + \dots
+ a_n$ is at most $a_1^* + \dots + a_n^*$ (the $n$ largest
terms): the left side only grows. Hence the inequality for all
$(a_n)$.

\textbf{9.} If $(a_n) \in \ell^p$, the sequence of Ces\`aro
means is in $\ell^p$ with norm $\leq \frac p{p-1}\norm
a_p$. Example ($p = 2$): $a_n = \frac1{\sqrt n\,\ln n}$ ($n
\geq 2$): $\sum a_n^2 = \sum\frac1{n\ln^2n} < \infty$, yet
$\sum a_n = \infty$ (integral test); Hardy still guarantees
$\sum_n\bigl(\frac{a_1 + \dots + a_n}n\bigr)^2 < \infty$.

\textbf{10.} For $f = \mathbf 1_{\intcc01}$: $Hf(x) = 1$ on
$\intoc01$ and $= \frac1x$ for $x \geq 1$: $\int_0^\infty Hf =
1 + \int_1^\infty\frac{\dd x}x = \infty$, while $\norm f_1 =
1$. The proof collapses at two points: the constant
$\frac p{p-1}$ blows up as $p \to 1$, and the boundary term
$x^{1-p}F^p$ no longer vanishes at infinity for $p = 1$.
Hardy's inequality is an honest $p > 1$ phenomenon.

\textbf{11.} Pick the longest interval $B_{i_1}$; discard
every interval meeting it; pick the longest survivor
$B_{i_2}$; iterate (finitely many intervals). The chosen ones
are disjoint by construction, and every discarded $B$ met a
chosen interval at least as long: an interval meeting a
longer-or-equal one is contained in its triple, $B \subseteq
3B_{i_l}$.

\textbf{12.} $\{Mf > t\}$ is open: each average $x \mapsto
\frac1{2r}\int_{x-r}^{x+r}\abs f$ is \omterm{def:b3:topology:continuity}{continuous} (dominated
convergence in $x$), and a supremum of \omterm{def:b3:topology:continuity}{continuous} functions
is lower semicontinuous. Each $x$ in it owns $B_x =
\intoo{x-r_x}{x+r_x}$ with $\int_{B_x}\abs f >
t\,\lambda(B_x)$. For \omterm{def:b3:topology:compact}{compact} $K \subseteq \{Mf > t\}$:
finitely many $B_x$ cover $K$, Vitali (question 11) extracts
disjoint $B_1', \dots, B_k'$ with $K \subseteq
\bigcup_l3B_l'$, so
\[
\lambda(K) \leq 3\sum_l\lambda(B_l') <
\frac3t\sum_l\int_{B_l'}\abs f \leq \frac3t\,\norm f_1
\]
by disjointness; inner regularity
(\cref{thm:b3:measure:regularity}) concludes.

\textbf{13.} For $x > 1$: with $r \in \intcc{x-1}x$ the
average is $\frac{r - x + 1}{2r}$, increasing in $r$; for $r
\geq x$ it is $\frac1{2r}$, decreasing: the supremum is
$\frac1{2x}$, attained at $r = x$. So $M\mathbf 1_{\intcc01}
\notin L^1$. In general, if $\int_I\abs f = c > 0$ on a
bounded interval $I \subseteq \intcc{-C}C$, then $Mf(x) \geq
\frac{c}{2(\abs x + C)}$ for all $x$: never \omterm{def:b3:lebesgue:l1}{integrable}
unless $f = 0$ a.e.

\textbf{14.} With $f_t = f\,\mathbf 1_{\abs f > t/2}$:
$M(f - f_t) \leq \frac t2$, so $\{Mf > t\} \subseteq \{Mf_t
> \frac t2\}$ and question 12 gives $\lambda(Mf > t) \leq
\frac6t\int_{\abs f > t/2}\abs f$. Layer cake
(\cref{prop:b3:product:layercake}) and Tonelli:
\[
\norm{Mf}_p^p = p\int_0^\infty t^{p-1}\lambda(Mf > t)\dd t
\leq 6p\int\abs f\int_0^{2\abs f}t^{p-2}\,\dd t\,\dd\lambda
= \frac{6p\,2^{p-1}}{p-1}\,\norm f_p^p .
\]

\textbf{15.} Write $A_rf(x) = \frac1{2r}\int_{x-r}^{x+r}f$.
For \omterm{def:b3:topology:continuity}{continuous} $g$: $A_rg(x) \to g(x)$ everywhere. Given
$\varepsilon > 0$, split $f = g + h$ with $g$ \omterm{def:b3:topology:continuity}{continuous} of
\omterm{def:b3:topology:compact}{compact} support and $\norm h_1 < \varepsilon$
(\cref{thm:b3:lp:density}); then
\[
\limsup_{r\to0}\,\abs{A_rf(x) - f(x)} \leq Mh(x) +
\abs{h(x)},
\]
so $\Omega_\delta = \{\limsup_r\abs{A_rf - f} > \delta\}
\subseteq \{Mh > \tfrac\delta2\} \cup \{\abs h >
\tfrac\delta2\}$ has \omterm{def:b3:measure:measure}{measure} $\leq \frac{6\varepsilon}\delta
+ \frac{2\varepsilon}\delta$ (question 12; Markov).
$\varepsilon$ arbitrary: $\lambda(\Omega_\delta) = 0$; union
over $\delta = \frac1k$: $A_rf \to f$ a.e.

\textbf{16.} (a) For each $q \in \Q$, question 15 applied to
$\abs{f - q}$ gives $A_r\abs{f - q}(x) \to \abs{f(x) - q}$
a.e.; on the intersection of these \omterm{def:b3:measure:measure}{full-measure} sets, choose
$q$ with $\abs{f(x) - q} < \eta$:
$\limsup_rA_r\abs{f - f(x)}(x) \leq 2\eta$ for every $\eta$:
almost every $x$ is a Lebesgue point. (b) At a Lebesgue
point,
\[
\Bigl|\frac{F(x + h) - F(x)}h - f(x)\Bigr|
= \Bigl|\frac1h\int_x^{x+h}\bigl(f - f(x)\bigr)\Bigr|
\leq 2\,A_{\abs h}\abs{f - f(x)}(x) \to 0 :
\]
$F' = f$ a.e.\ --- primitives of $L^1$ functions do
differentiate back; the staircase (\cref{pb:b3:measure:1})
is the counterexample to the \emph{converse} direction only.

\textbf{17.} Apply question 15 to $\mathbf 1_{A\cap[-n,n]}$
and let $n$ grow: for a.e.\ $x \in A$ the \omterm{ex:b3:lebesgue:gamma}{density}
$\frac{\lambda(A\cap\intcc{x-r}{x+r})}{2r} \to 1$. Steinhaus:
around a \omterm{ex:b3:lebesgue:gamma}{density} point take $r$ with \omterm{ex:b3:lebesgue:gamma}{density} $> \frac34$;
for $\abs t < \frac r2$, $A$ and $A + t$ each fill more than
$\frac32r$ of an interval of length $\leq \frac52r$, hence
intersect: $\intoo{-r/2}{r/2} \subseteq A - A$.

\textbf{18.} Let $G(x) = x^{\alpha+1-p}F(x)^p$: $G$ vanishes
at $0$ ($F$ vanishes near $0$) and at $\infty$ ($F$ bounded,
$\alpha + 1 - p < 0$), so $\int_0^\infty G' = 0$ with
\[
G' = (\alpha + 1 - p)\,x^{\alpha-p}F^p +
p\,x^{\alpha+1-p}F^{p-1}f
= (\alpha + 1 - p)\Bigl(\frac Fx\Bigr)^px^\alpha +
p\Bigl(\frac Fx\Bigr)^{p-1}f\,x^\alpha .
\]
Hence $\int(F/x)^px^\alpha = \frac{p}{p-1-\alpha}
\int(F/x)^{p-1}f\,x^\alpha$; H\"older for the \omterm{def:b3:measure:measure}{measure}
$x^\alpha\dd x$ (exponents $\frac p{p-1}$ and $p$) finishes
as in question 3. Borderline $\alpha = p - 1$: with $f(t) =
\frac1t\mathbf 1_{\intcc1A}$, the right side is $\ln A$
while the left contains $\int_1^A\frac{(\ln x)^p}x\dd x =
\frac{(\ln A)^{p+1}}{p+1}$: no constant survives $A \to
\infty$.

\textbf{19.} Tonelli on $\{0 < t < x\}$:
\[
\langle Hf, g\rangle =
\int_0^\infty\frac{g(x)}x\int_0^xf(t)\,\dd t\,\dd x
= \int_0^\infty f(t)\int_t^\infty\frac{g(x)}x\,\dd x\,\dd t
= \langle f, H^*g\rangle .
\]
Duality: $\norm{H^*f}_p = \sup\{\langle f, Hg\rangle : g
\geq 0, \norm g_q \leq 1\} \leq \norm f_p\cdot
\sup\norm{Hg}_q \leq \frac q{q-1}\norm f_p = p\,\norm f_p$,
Hardy being invoked in $L^q$, whose constant $\frac q{q-1}$
equals $p$.

\textbf{20.} Split along the (null) diagonal. On $\{y \leq
x\}$:
\[
\iint_{y\leq x}\frac{f(x)g(y)}{x}\,\dd y\,\dd x
= \int f(x)\,(Hg)(x)\,\dd x \leq \norm f_p\,\norm{Hg}_q
\leq p\,\norm f_p\norm g_q,
\]
since Hardy in $L^q$ carries the constant $\frac q{q-1} =
p$. Symmetrically, $\iint_{y>x} = \int g\,(Hf) \leq
\norm g_q\norm{Hf}_p \leq q\,\norm f_p\norm g_q$ ($\frac
p{p-1} = q$). Total: $(p + q)\,\norm f_p\norm g_q$.

\textbf{21.} For $a_n = n^{-1/p}$, $n \leq N$: the right
side is $\bigl(\frac p{p-1}\bigr)^p\sum_{n\leq N}\frac1n =
\bigl(\frac p{p-1}\bigr)^p\ln N + O(1)$. On the left, for $n
\leq N$: $a_1 + \dots + a_n \geq \int_1^{n+1}t^{-1/p}\dd t =
\frac{p}{p-1}\bigl((n+1)^{1-1/p} - 1\bigr)$, so the $n$-th
Ces\`aro mean is $\geq \frac p{p-1}n^{-1/p}(1 -
o(1))$ uniformly for $n$ in any range $n \geq n_0(\eta)$;
raising to the $p$ and summing, the left side is $\geq
\bigl(\frac p{p-1}\bigr)^p(1 - \eta)\ln N + O_\eta(1)$.
Dividing and letting $N \to \infty$, then $\eta \to 0$: no
constant smaller than $\bigl(\frac p{p-1}\bigr)^p$ can work.

\textbf{22.} $H$ bounded on $L^p$: cumulative averages do
not inflate $p$-norms (constant $\frac p{p-1}$); Ces\`aro on
$\ell^p$: the same, discretized; $M$ bounded on $L^p$: even
the best local average stays under control (constant
$O(\frac1{p-1})$). All three fail at $p = 1$, and for one
reason: averaging a concentrated unit of mass produces a
$\frac1x$ tail (questions 10 and 13), and $\frac1x$ belongs
to every $L^p$ near infinity except $L^1$. Smoothing spreads
mass exactly to the harmonic frontier of \omterm{def:b3:lebesgue:l1}{integrability}.

\textbf{23.} Set $b_n = a_n^{1/p}$, so $\sum b_n^p = \sum
a_n$. Question 8 gives
\[
\sum_{n\geq1}\Bigl(\frac{b_1 + \dots + b_n}n\Bigr)^{p}
\leq \Bigl(\frac p{p-1}\Bigr)^{p}\sum_{n\geq1}a_n,
\]
and AM--GM bounds each summand from below:
\[
\Bigl(\frac{b_1 + \dots + b_n}n\Bigr)^{p}
\geq \bigl(b_1\cdots b_n\bigr)^{p/n}
= \bigl(a_1\cdots a_n\bigr)^{1/n}.
\]
Hence $\sum(a_1\cdots a_n)^{1/n} \leq \bigl(\frac
p{p-1}\bigr)^p\sum a_n$ for \emph{every} $p > 1$. With $m =
p - 1$,
\[
\Bigl(\frac p{p-1}\Bigr)^{p} =
\Bigl(1 + \frac1m\Bigr)^{m+1},
\]
which decreases to $\eu$ as $m \to \infty$ (the classical
monotone upper sequence for $\eu$). Taking the infimum over
$p$ gives Carleman's inequality with constant $\eu$.

\textbf{24.} For $a_n = \frac1n$, $(a_1\cdots a_n)^{1/n} =
(n!)^{-1/n}$. The bracketing of \cref{pb:b3:product:1} gives
$\sqrt{2\pi n}\,(n/\eu)^n \leq n! \leq \sqrt{2\pi
n}\,(n/\eu)^n\eu^{1/(12n)}$, so
\[
(n!)^{1/n} = \frac n\eu\,(2\pi n)^{1/(2n)}
\eu^{O(1/n^2)}
= \frac n\eu\Bigl(1 +
O\Bigl(\frac{\ln n}{n}\Bigr)\Bigr),
\]
since $(2\pi n)^{1/(2n)} = \exp\bigl(\frac{\ln(2\pi
n)}{2n}\bigr)$. Inverting, $(n!)^{-1/n} = \frac\eu n(1 +
O(\frac{\ln n}n))$, and summing over $n \leq N$:
\[
\sum_{n\leq N}(n!)^{-1/n}
= \eu\ln N + O(1),
\qquad
\eu\sum_{n\leq N}\frac1n = \eu\ln N + O(1)
\]
(the error series $\sum\frac{\ln n}{n^2}$ converges). A
Carleman inequality with constant $c$ would force $\eu\ln N
+ O(1) \leq c\,(\ln N + O(1))$, hence $c \geq \eu$ upon
dividing by $\ln N$. The optimizing sequences line up:
Hardy's constant is approached by $n^{-1/p}$ (question 21),
Carleman's by $n^{-1}$ --- in each case the harmonic-type
sequence sitting just outside the space being averaged.

\textbf{25.} For $f = \mathbf 1_{\intcc01}$: $Hf(x) = 1$ on
$\intoc01$ and $Hf(x) = \frac1x$ for $x > 1$, so
$\norm{Hf}_2^2 = 1 + \int_1^\infty x^{-2}\dd x = 2$ and the
ratio is $\sqrt2 \approx 1.414$, about $71\%$ of the sharp
bound. For $f_A$ ($p = 2$): $F(x) = 2(\sqrt x - 1)$ on
$\intcc1A$, so on that range $Hf_A = 2x^{-1/2} - 2x^{-1}$
and, for $x > A$, $Hf_A(x) = 2(\sqrt A - 1)/x$. Squaring
and integrating,
\begin{align*}
\int_1^A\bigl(2x^{-1/2} - 2x^{-1}\bigr)^2\dd x
&= 4\ln A - 12 + \frac{16}{\sqrt A} - \frac4A,\\
\int_A^{\infty}\frac{4(\sqrt A - 1)^2}{x^2}\,\dd x
&= 4 - \frac{8}{\sqrt A} + \frac4A,
\end{align*}
whence $\norm{Hf_A}_2^2 = 4\ln A - 8 + 8A^{-1/2}$; dividing
by $\norm{f_A}_2^2 = \ln A$ gives the stated identity. At $A
= \eu^{10}$: $4 - \frac{8(1 - \eu^{-5})}{10} = 3.2054$, so
the ratio is $\sqrt{3.2054} \approx 1.790 < 2$. The defect
$4 - \norm{Hf_A}_2^2/\norm{f_A}_2^2 \sim 8/\ln A$ decays
only logarithmically: to reach ratio $1.99$ one would need
$\ln A \approx 200$, i.e.\ $A \approx 10^{87}$. Sharp
constants are theorems, not experiments.
\end{solution}
