\section*{Chapter \ref{ch:b3:fouriertransform} --- The Fourier
Transform}

\begin{solution}{exo:b3:fouriertransform:1}
$\widehat{\mathbf 1_{\intcc{-a}a}}(\xi) =
\int_{-a}^a\eu^{-\iu\xi x}\dd x = \frac{2\sin(a\xi)}{\xi}$
(value $2a$ at $0$).
$\widehat{\eu^{-a\abs x}}(\xi) = \int_0^\infty\eu^{-(a +
\iu\xi)x} + \eu^{-(a - \iu\xi)x}\,\dd x = \frac1{a + \iu\xi} +
\frac1{a - \iu\xi} = \frac{2a}{a^2 + \xi^2}$.
Tent: $\max(0, 1 - \abs x) = \mathbf 1_{\intcc{-1/2}{1/2}} *
\mathbf 1_{\intcc{-1/2}{1/2}}$, so its transform is
$\bigl(\frac{2\sin(\xi/2)}\xi\bigr)^2 =
\bigl(\frac{\sin(\xi/2)}{\xi/2}\bigr)^2$.
Last: $\frac{2a}{a^2+\xi^2} \in L^1$, so inversion
(\cref{thm:b3:fouriertransform:inversion}(2)) applied to
$\eu^{-a\abs x}$ gives, after renaming variables,
\[
\widehat{\Bigl(\frac1{x^2 + a^2}\Bigr)}(\xi) =
\frac{\pi}{a}\,\eu^{-a\abs\xi} .
\]
\end{solution}

\begin{solution}{exo:b3:fouriertransform:2}
From \cref{prop:b3:fouriertransform:basic}:
$\widehat{f(\cdot - a)} = \eu^{-\iu a\xi}\hat f(\xi)$;
$\widehat{f\cos(b\cdot)} = \frac12\bigl(\hat f(\xi - b) + \hat
f(\xi + b)\bigr)$;
$\widehat{f(a\cdot + b)}(\xi) = \frac1a\,\eu^{\iu
b\xi/a}\,\hat f(\xi/a)$ ($a > 0$);
$\widehat{\overline{f(-\cdot)}} = \overline{\hat f}$;
$\widehat{f * f} = \hat f^2$. On the Gaussian
($\widehat{\eu^{-x^2}} = \sqrt\pi\eu^{-\xi^2/4}$) each rule is
a one-line check --- e.g.\ $\eu^{-(x-a)^2}$ has transform
$\sqrt\pi\,\eu^{-\iu a\xi}\eu^{-\xi^2/4}$, which the direct
computation (complete the square) confirms.
\end{solution}

\begin{solution}{exo:b3:fouriertransform:3}
(a) $h = \mathbf 1_{\intcc{-1}1}*\mathbf 1_{\intcc{-1}1}$ has
$\hat h = \bigl(\frac{2\sin\xi}\xi\bigr)^2 \in L^1$; inversion
at $x = 0$, where $h(0) = \lambda(\intcc{-1}1\cap\intcc{-1}1)
= 2$:
\[
2 = \frac1{2\pi}\int_\R\Bigl(\frac{2\sin\xi}\xi\Bigr)^2
\dd\xi
\ \Longrightarrow\
\int_\R\Bigl(\frac{\sin\xi}\xi\Bigr)^2\dd\xi = \pi ,
\]
consistent with $\int_0^\infty\frac{\sin^2}{\xi^2} =
\frac\pi2$ (\cref{pb:b3:lebesgue:1}).

(b) Plancherel for $f = \eu^{-\abs x}$: $\int\abs{\hat f}^2 =
2\pi\int\abs f^2$ reads $\int\frac{4\,\dd\xi}{(1 + \xi^2)^2} =
2\pi\int\eu^{-2\abs x}\dd x = 2\pi$:
$\int_\R\frac{\dd\xi}{(1+\xi^2)^2} = \frac\pi2$.
\end{solution}

\begin{solution}{exo:b3:fouriertransform:4}
(a) $g_t(x) = \frac1{2\sqrt{\pi t}}\eu^{-x^2/4t}$: by
\cref{ex:b3:fouriertransform:gaussian} with $a = \frac1{4t}$,
$\hat g_t(\xi) = \frac1{2\sqrt{\pi t}}\sqrt{4\pi
t}\,\eu^{-t\xi^2} = \eu^{-t\xi^2}$.
(b) $\widehat{g_t * g_s} = \eu^{-t\xi^2}\eu^{-s\xi^2} =
\widehat{g_{t+s}}$, and the transform is injective on $L^1$
(\cref{thm:b3:fouriertransform:inversion}(3)): $g_t * g_s =
g_{t+s}$.
(c) $\norm{g_t}_1 = 1$ (Gaussian integral);
$\norm{g_t}_2^2 = \frac1{4\pi t}\int\eu^{-x^2/2t}\dd x =
\frac{\sqrt{2\pi t}}{4\pi t} = \frac1{\sqrt{8\pi t}}$.
\end{solution}

\begin{solution}{exo:b3:fouriertransform:5}
For real even $f$: $\hat f(\xi) = \int f\cos(\xi x)\dd x$ (the
sine part cancels): real and even. Odd: $\hat f(\xi) =
-\iu\int f\sin(\xi x)$: odd, purely imaginary. $\hat f(0) =
\int f$: the total mass. If $f \geq 0$: $\abs{\hat f(\xi)}
\leq \int\abs f = \int f = \hat f(0)$, so the sup is attained
at $0$. For a probability density, $\hat f(-\xi)$ is the
characteristic function of \cref{ch:b3:clt}: modulus $\leq 1$
everywhere, $= 1$ at the origin.
\end{solution}

\begin{solution}{exo:b3:fouriertransform:6}
(i) Injectivity is
\cref{thm:b3:fouriertransform:inversion}(3); continuity is
$\norm{\hat f}_\infty \leq \norm f_1$ (with values in
$\mathcal C_0$ by Riemann--Lebesgue); $\mathcal C_0$ is closed
in the sup norm (uniform limits of vanishing-at-infinity
functions vanish at infinity): Banach.

(ii) A continuous bijection between Banach spaces has
continuous inverse (\cref{thm:b3:banach:openmapping}): there
would be $C$ with $\norm f_1 \leq C\norm{\hat f}_\infty$.

(iii) Let $f_n(x) = \frac{\sin x}x\cdot\frac{\sin(x/n)}{x/n}$:
a product of two $L^2$ functions, and $O(x^{-2})$ at
infinity, so $f_n \in L^1\cap L^2$. Since
$\bigl(\frac{\sin(ax)}{ax}\bigr)$ has $L^2$-transform
$\frac\pi a\mathbf 1_{\intcc{-a}a}$, the product formula
$\widehat{gh} = \frac1{2\pi}\hat g * \hat h$ (valid for $g, h
\in L^2$ with $gh \in L^1$; check it on Schwartz functions by
Fubini and extend by $L^2$-continuity of both sides via
Plancherel) gives
\[
\hat f_n = \frac1{2\pi}\,\bigl(\pi\mathbf
1_{\intcc{-1}1}\bigr) * \bigl(\pi n\,\mathbf
1_{\intcc{-1/n}{1/n}}\bigr):
\]
a trapezoid of height $\frac{\pi n}2\cdot\frac2n = \pi$:
$\norm{\hat f_n}_\infty = \pi$ for every $n$. But on $[1,
n]$, $\frac{\sin(x/n)}{x/n} \geq \sin 1 > 0$, so
\[
\norm{f_n}_1 \geq \sin 1\int_1^n\frac{\abs{\sin x}}x\dd x
\geq c\ln n
\]
(arch-counting, as in \cref{thm:b3:banach:fourierdiverge}).
The bound $\norm{f_n}_1 \leq C\pi$ fails for large $n$: not
surjective. (The image is a dense --- by
Stone--Weierstrass-type arguments --- but proper subspace of
$\mathcal C_0$.)
\end{solution}

\begin{solution}{exo:b3:fouriertransform:7}
If $\hat f(\xi) = O(\abs\xi^{-k-1-\delta})$: then $\xi^j\hat f
\in L^1$ for $0 \leq j \leq k$ (integrable at infinity by the
decay, locally by continuity of $\hat f$). Inversion
(\cref{thm:b3:fouriertransform:inversion}(2)) represents $f$
a.e.\ by $x \mapsto \frac1{2\pi}\int\hat f(\xi)\eu^{\iu
x\xi}\dd\xi$, and differentiation under the integral
(dominators $\abs{\xi^j\hat f}$) makes this representative
$\mathcal C^k$. Conversely for $f \in \mathcal C_c^k$:
iterating \cref{prop:b3:fouriertransform:basic}(4),
$(\iu\xi)^k\hat f = \widehat{f^{(k)}}$, so $\abs{\hat f} \leq
\norm{f^{(k)}}_1\abs\xi^{-k}$. Tent function: continuous with
compact support ($k = 0$: transform bounded), and its
transform $\sim \xi^{-2} = O(\abs\xi^{-0-1-1})$ returns, by
the first direction, a $\mathcal C^0$ representative --- both
sharp: the tent is not $\mathcal C^1$, and its transform
decays no faster than $\xi^{-2}$.
\end{solution}

\begin{solution}{exo:b3:fouriertransform:8}
Integration by parts ($f \in \mathcal S$; boundary terms
vanish):
\[
1 = \int f^2 = \bigl[xf^2\bigr]_{-\infty}^{\infty} - \int
x\,(f^2)' = -2\int xff' \leq 2\,\norm{xf}_2\,\norm{f'}_2 .
\]
Plancherel and $\widehat{f'} = \iu\xi\hat f$:
$\norm{f'}_2^2 = \frac1{2\pi}\int\xi^2\abs{\hat f}^2$. Squaring
the display:
\[
\frac14 \leq \norm{xf}_2^2\cdot\frac1{2\pi}
\int\xi^2\abs{\hat f}^2\dd\xi .
\]
Equality requires equality in Cauchy--Schwarz: $f' =
\lambda xf$ with $\lambda < 0$ (integrability), i.e.\ $f(x) =
c\,\eu^{\lambda x^2/2}$: Gaussians. A signal concentrated in
$x$ (small $\norm{xf}_2$) must have spread-out spectrum, and
conversely: the uncertainty principle.
\end{solution}

\begin{solution}{exo:b3:fouriertransform:9}
The Gaussian $f(x) = \eu^{-\pi tx^2}$ is Schwartz, so
\cref{thm:b3:fouriertransform:poisson} applies, and
$\hat f(\xi) = t^{-1/2}\eu^{-\xi^2/4\pi t}$; at $\xi = 2\pi
k$ the right side becomes $t^{-1/2}\eu^{-\pi k^2/t}$: the
theta identity. At $t = 1$:
\[
\sum_{n\in\Z}\eu^{-\pi n^2} = 1 + 2\eu^{-\pi} + 2\eu^{-4\pi} +
\cdots \approx 1 + 0.0864278 + 0.0000070 = 1.0864348,
\]
accurate to $6$ decimals with three terms ($\eu^{-9\pi}
\approx 5\cdot10^{-13}$). At $t = 10^{-2}$: the defining
series needs $\eu^{-\pi n^2/100} < 10^{-7}$, i.e.\ $n \gtrsim
23$ --- about $47$ terms --- while the transformed series is
$10\sum_k\eu^{-100\pi k^2}$, where already the $k = 1$ term is
$\sim 10^{-136}$: one term suffices.
\end{solution}

\begin{solution}{exo:b3:fouriertransform:10}
$\mathcal Ff \in L^2(\intcc{-\pi}\pi) \subseteq
L^1(\intcc{-\pi}\pi)$ (finite measure), so inversion gives the
continuous representative $f(x) =
\frac1{2\pi}\int_{-\pi}^\pi\mathcal Ff(\xi)\eu^{\iu
x\xi}\dd\xi$, with
\[
f(n) = \frac1{2\pi}\int_{-\pi}^{\pi}\mathcal
Ff(\xi)\,\eu^{\iu n\xi}\dd\xi = \langle e_{-n}, \mathcal
Ff\rangle
\]
in the notation of \cref{thm:b3:hilbert:fourier}. Expanding in
that Hilbert basis: $\mathcal Ff = \sum_nf(n)\,\eu^{-\iu
n\xi}$ in $L^2(\intcc{-\pi}\pi)$. Apply the
$L^2$-continuous $\mathcal F^{-1}$ term by term:
\[
\mathcal F^{-1}\bigl(\mathbf
1_{\intcc{-\pi}\pi}\eu^{-\iu n\xi}\bigr)(x)
= \frac1{2\pi}\int_{-\pi}^{\pi}\eu^{\iu\xi(x - n)}\dd\xi
= \frac{\sin\bigl(\pi(x-n)\bigr)}{\pi(x - n)} ,
\]
giving $f = \sum_nf(n)\operatorname{sinc}(\cdot - n)$ in
$L^2$: a band-limited signal is determined by its integer
samples --- Shannon's sampling theorem.
\end{solution}

\begin{solution}{exo:b3:fouriertransform:11}
(a) Inversion on $\mathcal S$: $f(x) =
\frac1{2\pi}\int\hat f(\xi)\eu^{\iu x\xi}\dd\xi =
\frac1{2\pi}(\mathcal F\hat f)(-x)$, i.e.\ $(\mathcal
F^2f)(x) = 2\pi f(-x)$. Applying twice: $\mathcal F^4f =
2\pi\,\mathcal F^2f(-\cdot) = (2\pi)^2f$.

(b) If $\mathcal Ff = \lambda f$ with $f \neq 0$:
$(2\pi)^2f = \mathcal F^4f = \lambda^4f$, so $\lambda^4 =
(2\pi)^2$: $\lambda \in \{\pm\sqrt{2\pi},
\pm\iu\sqrt{2\pi}\}$. The Gaussian $g(x) = \eu^{-x^2/2}$ has
$\hat g = \sqrt{2\pi}\,g$
(\cref{ex:b3:fouriertransform:gaussian} at $a = \frac12$):
eigenfunction for $+\sqrt{2\pi}$.

(c) $\mathcal F^2f = 2\pi f(-\cdot)$ equals $\pm2\pi f$
according to parity. For $h(x) = x\eu^{-x^2/2}$:
differentiating $\hat g(\xi) = \sqrt{2\pi}\eu^{-\xi^2/2}$
with the rule $\widehat{xf} = \iu\frac{\dd}{\dd\xi}\hat f$:
\[
\hat h(\xi) = \iu\,\frac{\dd}{\dd\xi}\bigl(\sqrt{2\pi}
\eu^{-\xi^2/2}\bigr) = -\iu\sqrt{2\pi}\,\xi\eu^{-\xi^2/2}
= -\iu\sqrt{2\pi}\,h(\xi) :
\]
an eigenfunction for $-\iu\sqrt{2\pi}$. (The Hermite
functions continue the pattern, cycling through the four
eigenvalues --- the discrete Fourier clock.)

\end{solution}

\begin{solution}{exo:b3:fouriertransform:12}
(a) $\tilde f \in L^2$ with $\norm{\tilde f}_2 = \norm f_2$;
\cref{exo:b3:lp:6} (conjugate exponents $p = q = 2$) makes
$A_f = f * \tilde f$ bounded and uniformly continuous, with
\[
A_f(x) = \int f(y)\,\overline{f(y - x)}\,\dd y,
\qquad A_f(0) = \norm f_2^2,
\qquad \abs{A_f(x)} \leq \norm f_2\,\norm{f(\cdot -
x)}_2 = A_f(0)
\]
by Cauchy--Schwarz.

(b) For $f \in L^1\cap L^2$: $\tilde f \in L^1$ too, and the
convolution theorem gives $\widehat{A_f} = \hat f\,
\widehat{\tilde f}$; computing, $\widehat{\tilde f}(\xi) =
\int\overline{f(-x)}\eu^{-\iu\xi x}\dd x =
\overline{\int f(u)\eu^{-\iu\xi u}\dd u} =
\overline{\hat f(\xi)}$: $\widehat{A_f} = \abs{\hat f}^2
\geq 0$.

(c) When $\abs{\hat f}^2 \in L^1$, inversion applies to the
continuous $A_f$ (its transform is integrable;
\cref{thm:b3:fouriertransform:inversion}):
\[
A_f(x) = \frac1{2\pi}\int\abs{\hat f(\xi)}^2
\eu^{\iu x\xi}\,\dd\xi .
\]
For $f = \mathbf 1_{\intcc{-1/2}{1/2}}$: $\hat f(\xi) =
\frac{2\sin(\xi/2)}\xi = \frac{\sin(\xi/2)}{\xi/2}$, and at
$x = 0$:
\[
1 = \norm f_2^2 = \frac1{2\pi}\int_\R
\Bigl(\frac{\sin(\xi/2)}{\xi/2}\Bigr)^2\dd\xi
= \frac1{2\pi}\cdot2\int_\R\Bigl(\frac{\sin u}u\Bigr)^2\dd u
\]
($\xi = 2u$), i.e.\ $\int_\R\bigl(\frac{\sin u}u\bigr)^2\dd
u = \pi$ --- Plancherel's favorite integral, recovered by
autocorrelation.
\end{solution}

\begin{solution}{pb:b3:fouriertransform:1}
\textbf{1.} Transforming the equation in $x$ (formally):
$\partial_t\hat u(t,\xi) = \widehat{\partial^2_{xx}u} =
(\iu\xi)^2\hat u = -\xi^2\hat u$, an ODE in $t$ for each
frequency: $\hat u(t,\xi) = \eu^{-t\xi^2}\hat f(\xi)$. Since
$\eu^{-t\xi^2} = \hat g_t$
(\cref{exo:b3:fouriertransform:4}), the product is the
transform of $g_t * f$.

\textbf{2.} $\abs{u(t,x)} \leq \norm f_\infty\int g_t = \norm
f_\infty$: well defined. On $[t_0, T]\times[-A, A]$: each
mixed derivative $\partial^m_t\partial^n_xg_t(x - y)$ is a
polynomial in $(x - y)$ and $t^{-1}$ times
$\eu^{-(x-y)^2/4t}$, bounded for $\abs y \geq 2A$ by
$C\,(1 + y^2)^N\eu^{-(\abs y - A)^2/4T}$, an integrable
dominator independent of $(t, x)$ in the window (and bounded
for $\abs y \leq 2A$): repeated differentiation under the
integral (\cref{thm:b3:lebesgue:paramdiff}) applies: $u \in
\mathcal C^\infty(\intoo0\infty\times\R)$.

\textbf{3.} With $g_t(x) = \frac1{2\sqrt{\pi
t}}\eu^{-x^2/4t}$:
\[
\partial_tg_t = g_t\Bigl(\frac{x^2}{4t^2} - \frac1{2t}\Bigr)
= \partial^2_{xx}g_t
\]
(differentiate twice in $x$: $\partial_xg_t = -\frac
x{2t}g_t$, $\partial^2_{xx}g_t = \bigl(\frac{x^2}{4t^2} -
\frac1{2t}\bigr)g_t$). By question 2 the derivatives pass
under the integral: $\partial_tu = \partial^2_{xx}u$.

\textbf{4.} $u(t,x) - f(x) = \int g_t(y)\bigl(f(x - y) -
f(x)\bigr)\dd y$. Given a compact $K$ and $\varepsilon$:
uniform continuity of $f$ on a neighborhood of $K$ gives
$\delta$ with $\abs{f(x-y) - f(x)} < \varepsilon$ for $x \in
K$, $\abs y \leq \delta$; the tail contributes $\leq 2\norm
f_\infty\int_{\abs y > \delta}g_t(y)\dd y = 2\norm
f_\infty\,\P$-mass beyond $\delta$, which is
$\frac2{\sqrt\pi}\int_{\delta/2\sqrt t}^\infty\eu^{-z^2}\dd z
\to 0$ as $t \to 0$. For $f \in L^p$: $\norm{u(t) - f}_p \leq
\int g_t(y)\norm{\tau_yf - f}_p\dd y$ (Minkowski/Jensen as in
\cref{thm:b3:lp:regularization}), split the same way using
\cref{thm:b3:lp:density}(3).

\textbf{5.} Instant smoothing is question 2 ($u(t)$ is
$\mathcal C^\infty$ for $t > 0$ with no smoothness of $f$
used). For $f = \mathbf 1_{\intoo0\infty}$:
\[
u(t, x) = \int_0^\infty g_t(x - y)\dd y
= \frac1{\sqrt\pi}\int_{-x/2\sqrt t}^{\infty}\eu^{-z^2}\dd z
= \frac12\Bigl(1 +
\operatorname{erf}\Bigl(\frac{x}{2\sqrt t}\Bigr)\Bigr),
\qquad \operatorname{erf}(s) =
\frac2{\sqrt\pi}\int_0^s\eu^{-z^2}\dd z :
\]
a smoothed step whose transition zone widens like $\sqrt t$
(profiles at $t_1 < t_2 < t_3$: ever flatter ramps through
$(0, \frac12)$).

\textbf{6.} The integrand $g_t(x-y)f(y)$ is $\geq 0$ and the
kernel is strictly positive: $u(t,x) = 0$ would force $f = 0$
a.e. Monotonicity in $f$ is monotonicity of the integral. A
spot where $f = 0$ on an interval still has $u(t, \cdot) > 0$
there for every $t > 0$: heat propagates at infinite speed
(any positivity anywhere is felt everywhere instantly).

\textbf{7.} Tonelli ($g_t(x-y)\abs{f(y)}$ is integrable on
$\R^2$): $\int u(t,x)\dd x = \int f(y)\bigl(\int g_t(x -
y)\dd x\bigr)\dd y = \int f$.

\textbf{8.} Plancherel: $2\pi\norm{u(t)}_2^2 =
\int\eu^{-2t\xi^2}\abs{\hat f(\xi)}^2\dd\xi$, nonincreasing in
$t$ (pointwise), strictly unless $\hat f = 0$ a.e.\ ($= f =
0$), with limit $0$ as $t\to\infty$ by DCT. And
$\norm{u(t)}_\infty \leq \norm{g_t}_\infty\norm f_1 =
\frac{\norm f_1}{2\sqrt{\pi t}} \to 0$.

\textbf{9.} For $\varphi \in \mathcal C_c^\infty$, $t \mapsto
\langle\varphi, \hat u(t)\rangle$ is $\mathcal C^1$ with
derivative $\langle\varphi, \partial_t\hat u\rangle =
\langle\varphi, \widehat{\partial_{xx}u}\rangle =
\langle\xi^2\varphi\dots\rangle$ --- precisely,
$\widehat{\partial^2_{xx}u} = -\xi^2\hat u$ transfers the
equation. Then for a.e.\ $\xi$, the absolutely continuous
function $t \mapsto \eu^{t\xi^2}\hat u(t,\xi)$ has derivative
$\eu^{t\xi^2}(\xi^2\hat u + \partial_t\hat u) = 0$ in the
integrated sense: it is constant, and letting $t \to 0$
($\hat u(t) \to \hat f$ in $L^2$, a.e.\ along a subsequence):
$\hat u(t, \xi) = \eu^{-t\xi^2}\hat f(\xi)$ a.e. Two
solutions in the class have the same transform: they are
equal.

\textbf{10.} $f = g_s * h$ with $h \in L^2$ forces $\hat f =
\eu^{-s\xi^2}\hat h$, i.e.\ $\hat h = \eu^{s\xi^2}\hat f \in
L^2$. Take $\hat f(\xi) = \eu^{-\abs\xi}$: then $f(x) =
\frac1\pi\cdot\frac1{1 + x^2}$
(\cref{exo:b3:fouriertransform:1}, inversion), a perfectly
smooth $L^2$ function; but $\eu^{2s\xi^2 - 2\abs\xi} \to
\infty$: $\eu^{s\xi^2}\hat f \notin L^2$ for every $s > 0$.
The Cauchy profile is \emph{never} the result of prior
diffusion.

\textbf{11.} The heat semigroup multiplies transforms by
$\eu^{-t\xi^2}$, which vanishes nowhere: injective ---
formally, no information is destroyed. But its range consists
of functions whose transforms decay like $\eu^{-t\xi^2}$: a
tiny, dense-but-proper subspace of $L^2$ (question 10 shows
even excellent functions lie outside). Inverting would
amplify the frequency $\xi$ by $\eu^{t\xi^2}$: unbounded,
hence unstable against any perturbation. Diffusion is
irreversible not because the map forgets, but because its
inverse cannot be continuous --- an arrow of time made of
functional analysis.

\textbf{12.} $\hat f \in L^2(\intcc{-\Omega}\Omega)
\subseteq L^1$ (Cauchy--Schwarz on a bounded interval), so
$F(x) = \frac1{2\pi}\int_{-\Omega}^\Omega\hat
f(\xi)\eu^{\iu x\xi}\dd\xi$ is defined everywhere, and
differentiation under the integral (dominated by
$\Omega^k\abs{\hat f} \in L^1$ on the band) makes it
$\mathcal C^\infty$ with $\abs{F^{(k)}} \leq
\frac{\Omega^k}{2\pi}\norm{\hat f}_{L^1}$ everywhere. And $F
= f$ a.e.: both sides have the same transform, and the
transform is injective on $L^2$
(\cref{thm:b3:fouriertransform:plancherel} and its
$L^2$ extension).

\textbf{13.} The exponentials $\xi \mapsto \eu^{-\iu
n\pi\xi/\Omega}$, $n \in \Z$, form a Hilbert basis of
$L^2(\intcc{-\Omega}\Omega)$
(\cref{thm:b3:hilbert:fourier}, rescaled). The coefficient
of $\hat f$ along the $n$-th one is
\[
\frac1{2\Omega}\int_{-\Omega}^\Omega\hat f(\xi)\,
\eu^{\iu n\pi\xi/\Omega}\dd\xi
= \frac{2\pi}{2\Omega}\cdot
\frac1{2\pi}\int_{-\Omega}^{\Omega}\hat f(\xi)\,
\eu^{\iu(n\pi/\Omega)\xi}\dd\xi
= \frac\pi\Omega\,f\Bigl(\frac{n\pi}\Omega\Bigr),
\]
by question 12's formula at $x = \frac{n\pi}\Omega$: the
stated expansion holds in $L^2$ of the band.

\textbf{14.} Insert the expansion into the inversion formula
of question 12; the exchange of sum and integral is the
continuity of the $L^2$ pairing against $\frac1{2\pi}
\eu^{\iu x\xi}\mathbf 1_{\abs\xi\leq\Omega}$ (of $L^2$ norm
$\frac{\sqrt{2\Omega}}{2\pi}$, independent of $x$ ---
whence uniformity):
\[
f(x) = \sum_nf\Bigl(\frac{n\pi}\Omega\Bigr)\cdot
\frac1{2\Omega}\int_{-\Omega}^\Omega
\eu^{\iu(x - n\pi/\Omega)\xi}\dd\xi
= \sum_nf\Bigl(\frac{n\pi}\Omega\Bigr)
\operatorname{sinc}(\Omega x - n\pi),
\]
since $\frac1{2\Omega}\int_{-\Omega}^\Omega\eu^{\iu
u\xi}\dd\xi = \frac{\sin(\Omega u)}{\Omega u}$.

\textbf{15.} Reading question 14's computation backwards,
the transform of $s_n = \operatorname{sinc}(\Omega\cdot -
n\pi)$ is $\hat s_n = \frac\pi\Omega\,\eu^{-\iu
n\pi\xi/\Omega}\,\mathbf 1_{\intcc{-\Omega}\Omega}$.
Plancherel:
\[
\langle s_n, s_m\rangle = \frac1{2\pi}
\Bigl(\frac\pi\Omega\Bigr)^2\int_{-\Omega}^\Omega
\eu^{\iu(n-m)\pi\xi/\Omega}\dd\xi =
\frac\pi\Omega\,\delta_{nm} :
\]
an orthogonal family of constant norm
$\sqrt{\pi/\Omega}$. Taking norms in question 14's
expansion: $\norm f_2^2 =
\frac\pi\Omega\sum_n\abs{f(n\pi/\Omega)}^2$.

\textbf{16.} $g(x) = \sin(\Omega x)\operatorname{sinc}
(\Omega x) = \frac{\sin^2(\Omega x)}{\Omega x}$ vanishes at
every grid point $\frac{n\pi}\Omega$ (including $0$, by the
limit) and is not identically zero. Its band: write $g =
\frac1{2\iu}\bigl(\eu^{\iu\Omega x} - \eu^{-\iu\Omega
x}\bigr)\operatorname{sinc}(\Omega x)$; modulation by
$\eu^{\pm\iu\Omega x}$ shifts the transform by
$\mp\Omega$, so $\hat g$ is supported in
$\intcc{-2\Omega}{2\Omega}$ (indeed in the union of two
shifted bands): $g \in PW_{2\Omega}$, invisible to
$\Omega$-rate sampling --- aliasing incarnate.

\textbf{17.} By question 15 the samples carry the energy
democratically: $\norm f^2 = \frac\pi\Omega\sum
\abs{f(n\pi/\Omega)}^2$. If the signal's energy outside the
time window $\intcc{-T/2}{T/2}$ is $\leq \varepsilon^2\norm
f^2$, the samples outside the window satisfy (up to boundary
terms controlled by the uniform bound of question 12)
$\frac\pi\Omega\sum_{\abs{n\pi/\Omega} > T/2}
\abs{f(n\pi/\Omega)}^2 \approx \norm{f\,\mathbf 1_{\abs x >
T/2}}^2 \leq \varepsilon^2\norm f^2$: truncating the
sampling series to the $\approx \frac{\Omega T}\pi$
in-window indices reconstructs $f$ up to relative error
$\approx\varepsilon$. Hence the time--bandwidth product
$\frac{\Omega T}{\pi}$ counts the effective real degrees of
freedom of the signal --- the rule behind every audio
format.

\textbf{18.} (a) $\operatorname{sinc}(\Omega x)$ has
samples $f(n\pi/\Omega) = \operatorname{sinc}(n\pi) =
\delta_{n0}$: the series reduces to its $n = 0$ term,
$\operatorname{sinc}(\Omega x)$ --- the theorem reproduces
its own kernel. (b) If $\hat f$ is supported in
$\intcc{-\Omega'}{\Omega'} \subseteq
\intcc{-\Omega}\Omega$, every step of questions 13--14 runs
verbatim with the larger band $\Omega$ (the expansion of
$\hat f$ on the bigger interval is still legitimate):
sampling faster than one's own Nyquist rate changes nothing
in the reconstruction --- oversampling is harmless, and in
practice beneficial (faster-decaying reconstruction kernels
can then be used).

\textbf{19.} Expand $\eu^{-\iu\xi x} =
\sum_k\frac{(-\iu\xi x)^k}{k!}$ inside the integral; on
$\intcc{-A}A$ the series converges normally
($\sum_k\frac{\abs{\xi}^kA^k}{k!}\abs f \in L^1$), so
integration term by term is legitimate:
\[
\hat f(\xi) = \sum_{k\geq0}\frac{(-\iu\xi)^k}{k!}m_k,
\qquad \abs{m_k} \leq A^k\norm f_1 .
\]
The bound makes the series converge for every complex $\xi$;
around any point $\xi_0$, regrouping (absolute convergence)
gives a power series in $\xi - \xi_0$: $\hat f$ is
real-analytic with infinite radius everywhere.

\textbf{20.} Let $g$ be real-analytic on $\R$ (Taylor series
converging to $g$ near each point) and $Z = \{\xi :
g^{(k)}(\xi) = 0\ \forall k\}$. $Z$ is closed (intersection
of closed sets); it is open, since at $\xi_0 \in Z$ the local
Taylor expansion of $g$ is the zero series, so $g$ vanishes
identically near $\xi_0$, together with all derivatives. If
$g$ vanishes on an interval, $Z \neq \varnothing$; by
connectedness of $\R$, $Z = \R$: $g \equiv 0$. Now if $f
\neq 0$ had compact support together with $\hat f$: question
19 makes $\hat f$ real-analytic, vanishing outside a compact,
hence on intervals: $\hat f \equiv 0$, so $f = 0$ a.e.\ by
injectivity --- contradiction. Likewise a nonzero $f \in
PW_\Omega$ cannot be compactly supported (swap the roles of
$f$ and $\hat f$ via inversion): band-limited signals never
die, time-limited signals occupy unbounded spectrum.

\textbf{21.} For $f = \eu^{-ax^2}$: $\norm f_2^2 =
\sqrt{\frac\pi{2a}}$ and $\int x^2\abs f^2 =
\frac1{4a}\sqrt{\frac{\pi}{2a}}$ (Gaussian second moment);
$\hat f = \sqrt{\frac\pi a}\,\eu^{-\xi^2/4a}$
(\cref{ex:b3:fouriertransform:gaussian}) and
\[
\frac1{2\pi}\int\xi^2\abs{\hat f}^2\dd\xi =
\frac1{2\pi}\cdot\frac\pi a\int\xi^2
\eu^{-\xi^2/2a}\dd\xi = \frac1{2a}\cdot a\sqrt{2\pi a}
= \frac{\sqrt{2\pi a}}2 .
\]
Normalized product: $\frac1{4a}\sqrt{\frac\pi{2a}}\cdot
\frac{\sqrt{2\pi a}}2\big/\frac{\pi}{2a} = \frac14$,
independent of $a$. Scaling explains the constancy: replacing
$f$ by $f(\lambda\cdot)$ multiplies $\int x^2\abs f^2/\norm
f^2$ by $\lambda^{-2}$ and $\frac1{2\pi}\int\xi^2\abs{\hat
f}^2/\norm f^2$ by $\lambda^{2}$: the product is a dilation
invariant, and the Gaussians form one dilation orbit.

\textbf{22.} With $\abs f^2$ the position density and
$\frac1{2\pi}\abs{\hat f}^2$ the momentum density of a
quantum state (physical units insert $\hbar$),
\cref{exo:b3:fouriertransform:8} reads $\sigma_x\sigma_p
\geq \frac\hbar2$: no state is sharp in both observables.
Across the chapter, one law wears five suits: heat instantly
smooths because $\eu^{-t\xi^2}$ annihilates high frequencies
(Part II); the flow cannot run backward because restoring
them is unbounded (Part IV); a band-limited signal is rigid
enough to live on a countable grid (Part V); no function
beats Heisenberg's floor; and no function is compactly
supported on both sides of the transform (questions 19--20).
What $\hat f$ does at infinity governs what $f$ may do
anywhere.

\textbf{23.} Both $g_t$ and $g_s$ are in $L^1$ with
$\widehat{g_t}(\xi) = \eu^{-t\xi^2}$ (question 1's
computation), so the convolution theorem gives
$\widehat{g_t * g_s} = \eu^{-t\xi^2}\eu^{-s\xi^2} =
\eu^{-(t+s)\xi^2} = \widehat{g_{t+s}}$; two $L^1$ functions
with the same transform agree a.e.\ (injectivity, via the
inversion theorem --- here both sides are continuous, so
they agree everywhere): $g_t * g_s = g_{t+s}$. Consequently
$u(t + s) = g_{t+s} * f = g_s * (g_t * f) = g_s * u(t)$
(associativity of convolution, Tonelli). Log-convexity: let
$N(t) = \norm{u(t)}_2^2 = \frac1{2\pi}\int
\eu^{-2t\xi^2}\abs{\hat f}^2\dd\xi$ (Plancherel, question
8). For $t = \frac{t_1 + t_2}2$, write
\[
\eu^{-2t\xi^2}\abs{\hat f}^2
= \Bigl(\eu^{-2t_1\xi^2}\abs{\hat f}^2\Bigr)^{1/2}
\Bigl(\eu^{-2t_2\xi^2}\abs{\hat f}^2\Bigr)^{1/2},
\]
and Cauchy--Schwarz gives $N\bigl(\frac{t_1+t_2}2\bigr) \leq
\sqrt{N(t_1)\,N(t_2)}$: $\ln N$ is midpoint-convex, and
being continuous (dominated convergence in $t$), convex; so
is $\ln\norm{u(t)}_2 = \frac12\ln N(t)$. Decay with a convex
logarithm: the heat flow cannot lose energy in a burst and
then stall.

\textbf{24.} The kernel's moments: $\int g_t = 1$ (question
7 with $f = g_s$, or directly the Gaussian integral), $\int
x\,g_t(x)\dd x = 0$ (odd integrand), and, substituting $x =
2\sqrt t\,v$,
\[
\int_\R x^2g_t(x)\,\dd x
= \frac{4t}{\sqrt\pi}\int_\R v^2\eu^{-v^2}\dd v = 2t .
\]
Substituting $x = z + y$ in the convolution and noting
$\iint(\abs z + \abs y)^2g_t(z)f(y)\,\dd z\,\dd y < \infty$
(each of $\int\abs z^kg_t$, $\int\abs y^kf$ is finite for $k
\leq 2$, using $\abs y \leq \frac{1 + y^2}2$), Fubini and
Tonelli apply to the moment integrals below:
\[
\int x\,u(t,x)\dd x = \iint (z + y)\,g_t(z)f(y)\,\dd z\,\dd
y = 0\cdot\!\int\! f + 1\cdot\!\int\! yf(y)\dd y,
\]
which is the first claim; and
\[
\iint (z+y)^2g_t(z)f(y)\,\dd z\,\dd y
= 2t\int f + 2\cdot0\cdot\!\int\! yf + \int y^2f(y)\dd y ,
\]
the second. Means add, variances add, and the kernel
contributes mean $0$ and variance $2t$: after time $t$ the
heat has spread over a width of order $\sqrt{2t}$ ---
distance grows like the square root of time, the signature
of diffusion (and of the Brownian paths of
\cref{ch:b3:probability}).

\textbf{25.} Transform side: $\hat f(\xi) =
\sqrt\pi\,\eu^{-\xi^2/4}$, so $\hat u(t,\xi) =
\sqrt\pi\,\eu^{-(t + \frac14)\xi^2}$, which is the transform
of $(1 + 4t)^{-1/2}\exp\bigl(-x^2/(1+4t)\bigr)$ (the
Gaussian dictionary $\eu^{-ax^2} \mapsto
\sqrt{\pi/a}\,\eu^{-\xi^2/4a}$ with $a = \frac1{1+4t}$):
the closed form. Direct check, with $\sigma = 1 + 4t$:
\[
\partial_tu = \sigma^{-1/2}\eu^{-x^2/\sigma}
\Bigl(-\frac2\sigma + \frac{4x^2}{\sigma^2}\Bigr)
= \partial^2_{xx}u ,
\]
both sides computed from $\partial_xu =
-\frac{2x}\sigma\,u$. Conservation: $\int u(t) =
\sigma^{-1/2}\sqrt{\pi\sigma} = \sqrt\pi$ for all $t$.
Dissipation:
\[
\norm{u(t)}_2^2 = \frac1\sigma\int\eu^{-2x^2/\sigma}\dd x
= \frac1\sigma\sqrt{\frac{\pi\sigma}2}
= \sqrt{\frac\pi2}\,(1+4t)^{-1/2},
\]
so $\norm{u(t)}_2 = (\pi/2)^{1/4}(1+4t)^{-1/4}$,
nonincreasing with convex logarithm (question 23); the
$L^2$ norm decays like $t^{-1/4}$, exactly half the
$t^{-1/2}$ exponent of $\norm{u(t)}_\infty$ --- consistent
with $\norm u_2^2 \leq \norm u_\infty\norm u_1$ and
conservation of $\norm u_1$. Variance: $\int x^2u(t) =
\frac1\sigma\cdot\frac{\sigma^{3/2}\sqrt\pi}2 =
\frac{\sqrt\pi}2\,(1 + 4t) = \int x^2f + 2t\sqrt\pi$, as
question 24 predicts ($\int x^2f = \frac{\sqrt\pi}2$, $\int
f = \sqrt\pi$). At $t = 6$: $\sigma = 25$, peak height
$u(6, 0) = \frac15$ versus $u(0,0) = 1$, width scale
$\sqrt\sigma = 5$ times the initial one, and $\int u =
\sqrt\pi \approx 1.7725$ throughout: the spot is five times
lower, five times wider, and not a calorie is missing.
\end{solution}
