\section*{Chapter \ref{ch:b3:residues} --- Laurent Series and
the Residue Theorem}

\begin{solution}{exo:b3:residues:1}
$\frac{\sin z}z = 1 - \frac{z^2}6 + \cdots$: removable,
\omterm{def:b3:residues:singularities}{residue} $0$. $\frac{\eu^z - 1}{z^2} = \frac1z + \frac12 +
\frac z6 + \cdots$: simple pole, \omterm{def:b3:residues:singularities}{residue} $1$.
$\frac1{z(z-1)^2}$: simple pole at $0$ with \omterm{def:b3:residues:singularities}{residue}
$\frac1{(0-1)^2} = 1$; double pole at $1$ with \omterm{def:b3:residues:singularities}{residue}
$\frac{\dd}{\dd z}\bigl(\frac1z\bigr)\big|_{z=1} = -1$.
$\frac{\cos z}{z^3} = \frac1{z^3} - \frac1{2z} + \cdots$: pole
of order $3$, \omterm{def:b3:residues:singularities}{residue} $-\frac12$. $\eu^{1/z} = \sum_{n\geq0}
\frac{z^{-n}}{n!}$: essential, \omterm{def:b3:residues:singularities}{residue} $1$. $\frac1{\sin z}$:
simple poles at $n\pi$; \omterm{def:b3:residues:singularities}{residue} $\frac1{\cos 0} = 1$ at $0$,
$\frac1{\cos\pi} = -1$ at $\pi$
(\cref{met:b3:residues:compute}, $g/h'$).
\end{solution}

\begin{solution}{exo:b3:residues:2}
With $z = \eu^{\iu t}$, $\cos t = \frac{z + z^{-1}}2$, $\dd t
= \frac{\dd z}{\iu z}$:
\[
\int_0^{2\pi}\frac{\dd t}{a + \cos t}
= \oint_{\abs z = 1}\frac{2\,\dd z}{\iu\,(z^2 + 2az + 1)} .
\]
The roots $z_\pm = -a \pm \sqrt{a^2 - 1}$ satisfy $z_+z_- =
1$ with $\abs{z_+} < 1 < \abs{z_-}$; the \omterm{def:b3:residues:singularities}{residue} at $z_+$ is
$\frac{1}{z_+ - z_-} = \frac1{2\sqrt{a^2-1}}$, so the integral
is $\frac2\iu\cdot2\iu\pi\cdot\frac1{2\sqrt{a^2-1}} =
\frac{2\pi}{\sqrt{a^2-1}}$. As $a \to 1^+$ it blows up (the
integrand peaks at $t = \pi$); as $a \to \infty$ it behaves
like $\frac{2\pi}a$, matching $\int\frac{\dd t}a$.
\end{solution}

\begin{solution}{exo:b3:residues:3}
First integral: upper poles of $\frac{z^2}{1+z^6}$ at $p =
\eu^{\iu\pi/6}, \iu, \eu^{5\iu\pi/6}$; at each,
$\operatorname{Res} = \frac{p^2}{6p^5} = \frac{p^3}{6p^6} =
-\frac{p^3}6$, and $p^3$ takes the values $\iu, -\iu, \iu$:
sum of \omterm{def:b3:residues:singularities}{residues} $-\frac{\iu}{6}$. The arc is
$O(R^{-4})\cdot O(R) \to 0$:
\[
\int_\R\frac{x^2\,\dd x}{1 + x^6} =
2\iu\pi\Bigl(-\frac \iu6\Bigr) = \frac\pi3 .
\]
Second: double pole at $\iu$ of $\frac1{(1+z^2)^2} =
\frac1{(z-\iu)^2(z+\iu)^2}$:
\[
\operatorname{Res} = \frac{\dd}{\dd z}\,(z +
\iu)^{-2}\Big|_{z=\iu} = \frac{-2}{(2\iu)^3} =
\frac{-2}{-8\iu} = -\frac\iu4,
\qquad
\int_\R\frac{\dd x}{(1+x^2)^2} =
2\iu\pi\cdot\Bigl(-\frac\iu4\Bigr) = \frac\pi2 ,
\]
consistent with \cref{exo:b3:fouriertransform:3}(b).
\end{solution}

\begin{solution}{exo:b3:residues:4}
Close $\frac{\eu^{\iu tz}}{z^2 + a^2}$ in the upper half-plane
($t \geq 0$): there $\abs{\eu^{\iu tz}} =
\eu^{-t\operatorname{Im}z} \leq 1$, so the arc contributes
$O(R^{-2})\cdot O(R) \to 0$. The single enclosed pole $\iu a$
is simple with \omterm{def:b3:residues:singularities}{residue} $\frac{\eu^{-at}}{2\iu a}$:
\[
\int_\R\frac{\eu^{\iu tx}}{x^2 + a^2}\dd x =
\frac{\pi}{a}\,\eu^{-at},
\qquad\text{hence}\qquad
\int_\R\frac{\cos(tx)}{x^2+a^2}\dd x = \frac\pi
a\,\eu^{-a\abs t}
\]
(real part; even in $t$). This is the inversion counterpart
of $\widehat{\eu^{-a\abs x}} = \frac{2a}{a^2+\xi^2}$
(\cref{exo:b3:fouriertransform:1}): the two computations
confirm each other through
\cref{thm:b3:fouriertransform:inversion}.
\end{solution}

\begin{solution}{exo:b3:residues:5}
Partial fractions: $f = \frac1{z-2} - \frac1{z-1}$.
On $\abs z < 1$ (Taylor):
$f = \sum_{n\geq0}\bigl(1 - 2^{-n-1}\bigr)z^n$.
On $1 < \abs z < 2$: $\frac1{z-2} =
-\sum_{n\geq0}\frac{z^n}{2^{n+1}}$ and $-\frac1{z-1} =
-\sum_{n\geq1}z^{-n}$: a genuine two-sided series.
On $\abs z > 2$: $f = \sum_{n\geq1}\bigl(2^{n-1} -
1\bigr)z^{-n}$. The three differ because Laurent expansions
are attached to \emph{annuli}, not points: each region has its
own geometric expansions. The $z^{-1}$-coefficients ($-1$ and
$0$ respectively) are integrals over circles encircling
\emph{the singularities inside}, not \omterm{def:b3:residues:singularities}{residues} at $0$ ($f$ is
\omterm{def:b3:holomorphic:holo}{holomorphic} at $0$): for $1 < \abs z < 2$ the coefficient
$-1$ is $\operatorname{Res}(f, 1)$; for $\abs z > 2$ the
coefficient $0$ is $\operatorname{Res}(f,1) +
\operatorname{Res}(f,2) = -1 + 1$. The \omterm{def:b3:residues:singularities}{residues} of $f$: $-1$
at $1$ and $+1$ at $2$.
\end{solution}

\begin{solution}{exo:b3:residues:6}
(a) The \omterm{thm:b3:residues:laurent}{Laurent series} $\sum_nz^{-n}/n!$ has infinitely many
negative terms: essential. Solving $\eu^{1/z} = w$ ($w \neq
0$): $\frac1z = \log\abs w + \iu\arg w + 2\iu\pi k$, so
\[
z_k = \frac1{\log\abs w + \iu\arg w + 2\iu\pi k}
\xrightarrow[k\to\infty]{} 0 :
\]
every nonzero value is attained infinitely often near $0$ ---
stronger than \omterm{ex:b3:lebesgue:gamma}{density}.
(b) Along $z = 1/x$, $x \to +\infty$: $\eu^x \to \infty$;
along $z = \iu/y$: modulus $1$. A pole requires
$\abs{f(z)} \to \infty$ \emph{along every approach}: here the
limit simply does not exist, even in $[0, +\infty]$.
\end{solution}

\begin{solution}{exo:b3:residues:7}
(a) On $\abs z = 1$: $\abs{z^7 + z - 1} \leq 3 < 4 =
\abs{-4z^3}$. Rouch\'e with $f = -4z^3$, $g = z^7 + z - 1$:
three zeros in the disc.
(b) On $\abs z = 1$: $\abs{z^4 + 1} \leq 2 < 5 = \abs{5z}$:
one zero in $\abs z < 1$. On $\abs z = 2$: $\abs{5z + 1} \leq
11 < 16 = \abs{z^4}$: four zeros in $\abs z < 2$. Hence three
zeros in the annulus.
(c) For $P = z^n + a_{n-1}z^{n-1} + \dots$: on $\abs z = R >
1 + \sum\abs{a_k}$, $\abs{P - z^n} \leq
\bigl(\sum\abs{a_k}\bigr)R^{n-1} < R^n = \abs{z^n}$: $P$ has
exactly $n$ zeros in $D(0, R)$ --- d'Alembert--Gauss with
multiplicity, by pure counting.
\end{solution}

\begin{solution}{exo:b3:residues:8}
(a) Suppose $f(a) = 0$, $f \not\equiv 0$: choose $r$ with $f$
zero-free on the circle $C = \partial D(a, r)$ (\omterm{thm:b3:holomorphic:identity}{isolated
zeros}) and $m = \min_C\abs f > 0$. By
\cref{thm:b3:holomorphic:weierstrassconv}, $f_n \to f$ and
$f_n' \to f'$ uniformly on $C$; for large $n$, $\abs{f_n}
\geq m/2$ on $C$, so
\[
\frac1{2\iu\pi}\int_C\frac{f_n'}{f_n}
\longrightarrow \frac1{2\iu\pi}\int_C\frac{f'}{f} \geq 1
\]
(the limit counts the zero $a$; convergence because
numerators converge uniformly and denominators are uniformly
bounded below). The left side is an integer counting zeros of
$f_n$ in the disc: it must be $\geq 1$ eventually ---
contradicting zero-freeness. So $f$ is zero-free.

(b) Fix $w \in \Omega$ and apply (a) on the \omterm{def:b3:topology:connected}{connected} \omterm{def:b3:topology:topology}{open set}
$\Omega\setminus\{w\}$ (removing a point of an open \omterm{def:b3:topology:connected}{connected}
subset of $\C$ preserves \omterm{def:b3:topology:connected}{connectedness}) to $g_n(z) = f_n(z) -
f_n(w)$, zero-free there by injectivity, converging to $g = f
- f(w)$. If $f$ is nonconstant, $g \not\equiv 0$ on
$\Omega\setminus\{w\}$, so $g$ is zero-free there: $f(z) \neq
f(w)$ for all $z \neq w$. As $w$ was arbitrary, $f$ is
injective.
\end{solution}

\begin{solution}{exo:b3:residues:9}
The sector boundary consists of $[0, R]$, the arc $A_R$, and
the ray $\eu^{2\iu\pi/n}[0, R]$ reversed. Inside lies the
single pole $p = \eu^{\iu\pi/n}$ of $\frac1{1+z^n}$, with
\omterm{def:b3:residues:singularities}{residue} $\frac1{np^{n-1}} = \frac{p}{np^n} = -\frac pn$. On
the return ray, $z = \eu^{2\iu\pi/n}x$ gives $z^n = x^n$ and
$\dd z = \eu^{2\iu\pi/n}\dd x$; the arc is
$O(R^{-n})\cdot O(R) \to 0$. Hence
\[
\bigl(1 - \eu^{2\iu\pi/n}\bigr)
\int_0^\infty\frac{\dd x}{1 + x^n}
= 2\iu\pi\Bigl(-\frac{\eu^{\iu\pi/n}}n\Bigr),
\quad\text{so}\quad
\int_0^\infty\frac{\dd x}{1+x^n}
= \frac{2\iu\pi}{n\,\bigl(\eu^{\iu\pi/n} -
\eu^{-\iu\pi/n}\bigr)} = \frac{\pi}{n\sin(\pi/n)} .
\]
$n = 2$: $\frac\pi{2\sin(\pi/2)} = \frac\pi2 =
[\arctan]_0^\infty$. As $n \to \infty$: the value tends to
$1$, and indeed the integrand tends to $\mathbf 1_{\intco01}$
(with DCT domination $\min(1, x^{-2})$ for $n \geq 2$).
\end{solution}

\begin{solution}{exo:b3:residues:10}
On $\abs z = R$ large, $\abs f \leq C R^{-2}$: $\abs{\oint_{C_R}
f} \leq 2\pi R\cdot CR^{-2} \to 0$. But for $R$ beyond all
poles, the \omterm{def:b3:residues:singularities}{residue} theorem gives $\oint_{C_R}f =
2\iu\pi\sum_{\text{all }p}\operatorname{Res}(f, p)$: the total
sum vanishes. For $f = \frac1{z(z-1)(z-2)}$: \omterm{def:b3:residues:singularities}{residues}
$\frac1{(-1)(-2)} = \frac12$ at $0$, $\frac1{1\cdot(-1)} = -1$
at $1$, $\frac1{2\cdot1} = \frac12$ at $2$ --- summing to $0$
as predicted, and
\[
\frac1{z(z-1)(z-2)} = \frac{1/2}{z} - \frac1{z - 1} +
\frac{1/2}{z-2} :
\]
the \omterm{def:b3:residues:singularities}{residues} \emph{are} the partial fraction coefficients, and
the zero-sum identity supplies a free consistency check (or
determines the last coefficient from the others).
\end{solution}

\begin{solution}{exo:b3:residues:11}
On the keyhole, with the chosen determination:
just above the cut, $\log z = \ln x$; just below, $\log z =
\ln x + 2\iu\pi$. The four pieces give
\[
\Bigl(1 - \eu^{2\iu\pi(a-1)}\Bigr)\int_\varepsilon^R
\frac{x^{a-1}}{1+x}\dd x + \int_{C_R} + \int_{C_\varepsilon}
= 2\iu\pi\operatorname{Res}(f, -1) .
\]
Arcs: $\abs{f} \leq \frac{R^{a-1}}{R - 1}$ on $C_R$, length
$2\pi R$: contribution $O(R^{a-1}) \to 0$ ($a < 1$);
$\abs f \leq \frac{\varepsilon^{a-1}}{1 - \varepsilon}$ on
$C_\varepsilon$, length $2\pi\varepsilon$: $O(\varepsilon^a)
\to 0$ ($a > 0$). \omterm{def:b3:residues:singularities}{Residue}: at $z = -1 = \eu^{\iu\pi}$,
$\operatorname{Res} = \eu^{(a-1)\iu\pi}$. Hence
\[
I(a) = \frac{2\iu\pi\,\eu^{\iu\pi(a-1)}}{1 -
\eu^{2\iu\pi(a-1)}}
= \frac{2\iu\pi}{\eu^{-\iu\pi(a-1)} - \eu^{\iu\pi(a-1)}}
= \frac{\pi}{-\sin(\pi(a-1))} = \frac{\pi}{\sin\pi a} .
\]
Gamma reflection: $B(a, 1-a) =
\int_0^1t^{a-1}(1-t)^{-a}\dd t$; the substitution $t =
\frac x{1+x}$, $1 - t = \frac1{1+x}$, $\dd t =
\frac{\dd x}{(1+x)^2}$ turns it into $\int_0^\infty
\frac{x^{a-1}}{1+x}\dd x = I(a)$, and Euler's formula
$B(a, 1-a) = \Gamma(a)\Gamma(1-a)/\Gamma(1) $
(\cref{pb:b3:lebesgue:1}) gives
$\Gamma(a)\Gamma(1-a) = \frac\pi{\sin\pi a}$ --- in
particular $\Gamma(\tfrac12) = \sqrt\pi$ once more.

\end{solution}

\begin{solution}{exo:b3:residues:12}
(a) On $\abs z = 1$: $\abs{z^4} = 1 < 7 \leq \abs{8z + 1}$
($\abs{8z} - 1 = 7$): $P$ has as many zeros in $\mathbb D$
as $8z + 1$, namely \emph{one} (at $-\frac18$).

(b) On $\abs z = 3$: $\abs{8z + 1} \leq 25 < 81 =
\abs{z^4}$: Rouch\'e against $z^4$ gives all four zeros in
$\abs z < 3$, hence $4 - 1 = 3$ zeros in the annulus $1 <
\abs z < 3$. Sharpening: a zero with $\abs z = r \geq 2.1$
would satisfy $r^4 = \abs{8z + 1} \leq 8r + 1$, but $r^4 -
8r - 1$ is increasing for $r \geq 2$ and equals $19.45 -
16.8 - 1 = 1.65 > 0$ at $r = 2.1$: impossible. So the three
outer zeros lie in $1 < \abs z < 2.1$. (Numerically: a real
zero near $-1.95$ and a conjugate pair near $1.04 \pm
1.73\iu$, of modulus $2.02$ --- which is why a Rouch\'e
attempt at radius exactly $2$ must fail: the theorem demands
strict domination, and the zeros sit just outside.)

(c) No zeros on $\iu\R$: $\operatorname{Re}P(\iu t) = t^4 +
1 \geq 1$. Zeros in the right half-plane: use the argument
principle on the boundary of the half-disc $\{\abs z \leq
R,\ \operatorname{Re}z \geq 0\}$. On the large arc, $\arg P
\approx \arg z^4$ turns by $4\cdot\pi = 2\pi\cdot2$ (the
arc spans angle $\pi$). Along the imaginary axis from
$\iu R$ down to $-\iu R$: $P(\iu t) = (t^4 + 1) + 8\iu t$
stays in the right half-plane ($\operatorname{Re} > 0$), so
$\arg P$ varies within $\intoo{-\pi/2}{\pi/2}$ and returns
with net change $\to 0$ as $R \to \infty$ (endpoints both
$\approx \arg t^4 = 0$). Total winding: $\frac{4\pi + 0}
{2\pi} = 2$: \emph{two} zeros in the right half-plane ---
consistent with the numerics: the conjugate pair $\approx
1.04 \pm 1.73\iu$ has positive real part, the real zeros
$\approx -0.125$ and $\approx -1.96$ negative.
\end{solution}

\begin{solution}{pb:b3:residues:1}
\textbf{1.} $\sin(\pi z)$ has simple zeros exactly at $\Z$
($\sin\pi z = 0$ iff $z \in \Z$, and $(\sin\pi z)' = \pi\cos\pi
z \neq 0$ there), and $\cos(\pi n) \neq 0$: $\pi\cot(\pi z) =
\pi\cos(\pi z)/\sin(\pi z)$ has simple poles at $\Z$ with
\[
\operatorname{Res}(\pi\cot\pi z,\ n)
= \frac{\pi\cos(\pi n)}{\pi\cos(\pi n)} = 1
\]
($g/h'$ rule, \cref{met:b3:residues:compute}).

\textbf{2.} $\frac{\sin(\pi z)}{\pi z} = 1 - \frac{(\pi z)^2}6
+ \frac{(\pi z)^4}{120} - \cdots$ is \omterm{def:b3:holomorphic:holo}{holomorphic} and nonzero
near $0$: its reciprocal is \omterm{def:b3:holomorphic:holo}{holomorphic}
(\cref{def:b3:holomorphic:holo}: quotient), with series $1 +
\frac{(\pi z)^2}{6} + \frac{7(\pi z)^4}{360} + \cdots$
(identify: $(1 - u)^{-1}$-style coefficients from $u =
\frac{(\pi z)^2}6 - \frac{(\pi z)^4}{120}$: the $z^4$
coefficient is $\frac1{36} - \frac1{120} = \frac{7}{360}$).
Multiply by $\cos(\pi z) = 1 - \frac{(\pi z)^2}2 + \frac{(\pi
z)^4}{24} - \cdots$ and divide by $z$:
\[
\pi\cot(\pi z) = \frac1z\Bigl[1 + \pi^2z^2\Bigl(\frac16 -
\frac12\Bigr) + \pi^4z^4\Bigl(\frac7{360} - \frac1{12} +
\frac1{24}\Bigr)\Bigr] + \cdots
= \frac1z - \frac{\pi^2}3\,z - \frac{\pi^4}{45}\,z^3 -
\cdots
\]
($\frac7{360} - \frac{30}{360} + \frac{15}{360} =
-\frac8{360} = -\frac1{45}$).

\textbf{3.} Vertical sides $z = \pm(N + \frac12) + \iu y$:
by $\pi$-periodicity of $\cot$, $\cot(\pi z) = \cot(\pm\frac\pi2
+ \iu\pi y) = -\tan(\iu\pi y) = -\iu\tanh(\pi y)$, of modulus
$\leq 1$. Horizontal sides $z = x \pm \iu(N + \frac12)$: from
$\abs{\cot(a + \iu b)}^2 = \frac{\cos^2a +
\sinh^2b}{\sin^2a + \sinh^2b} \leq \frac{1 +
\sinh^2b}{\sinh^2b} = \coth^2 b$,
\[
\abs{\cot(\pi z)} \leq \coth\bigl(\pi(N + \tfrac12)\bigr)
\leq \coth(\pi/2) < 1.1 .
\]
Both bounds are $\leq 2$.

\textbf{4.} Apply \cref{thm:b3:residues:residue} to $F(z) =
\pi\cot(\pi z)f(z)$ on $C_N$ ($N$ beyond all poles of $f$):
\[
\frac1{2\iu\pi}\oint_{C_N}F = \sum_{n=-N}^{N}f(n) +
\sum_p\operatorname{Res}(F, p),
\]
the integer poles contributing $f(n)$ (question 1; $f$
\omterm{def:b3:holomorphic:holo}{holomorphic} there). On $C_N$: $\abs F \leq 2\pi\cdot
C\abs z^{-2} \leq 2\pi C N^{-2}$, and the perimeter is $8(N +
\frac12)$: the integral is $O(1/N) \to 0$. Let $N \to
\infty$: the displayed summation formula.

\textbf{5.} The decay $\abs f = O(\abs z^{-2})$ killed the
\omterm{def:b3:holomorphic:contour}{contour integral} \emph{and} made $\sum\abs{f(n)}$ converge.
For $f(z) = \frac1{z + \frac12}$: the symmetric sums
$\sum_{-N}^N\frac1{n + \frac12}$ telescope to $0$ (the terms
$n$ and $-n - 1$ cancel), and the right side is
$-\operatorname{Res}\bigl(\frac{\pi\cot\pi z}{z + \frac12},
-\frac12\bigr) = -\pi\cot(-\frac\pi2) = 0$: consistent ---
but $\sum\abs{f(n)}$ diverges; the method computes the
symmetric (principal value) limit only.

\textbf{6.} $g(z) = \frac{\pi\cot(\pi z)}{z^2}$: poles at the
nonzero integers with \omterm{def:b3:residues:singularities}{residues} $\frac1{n^2}$, and at $0$
where, by question 2,
\[
g(z) = \frac1{z^3} - \frac{\pi^2}{3z} - \frac{\pi^4}{45}z -
\cdots :
\qquad \operatorname{Res}(g, 0) = -\frac{\pi^2}3 .
\]
The \omterm{def:b3:holomorphic:contour}{contour integral} over $C_N$ tends to $0$ as in question 4
($\abs{g} = O(N^{-2})$ on $C_N$). Hence $0 =
\sum_{n\neq0}\frac1{n^2} - \frac{\pi^2}3$: $2\zeta(2) =
\frac{\pi^2}3$, $\zeta(2) = \frac{\pi^2}6$.

\textbf{7.} $g(z) = \frac{\pi\cot(\pi z)}{z^4} = \frac1{z^5}
- \frac{\pi^2}{3z^3} - \frac{\pi^4}{45z} - \cdots$: \omterm{def:b3:residues:singularities}{residue}
at $0$ equal to $-\frac{\pi^4}{45}$, and $0 =
2\zeta(4) - \frac{\pi^4}{45}$: $\zeta(4) =
\frac{\pi^4}{90}$.

\textbf{8.} With $g = \pi\cot(\pi z)/z^{2k}$: the \omterm{def:b3:residues:singularities}{residue} at
$0$ is the coefficient $a_{2k-1}$ of $z^{2k-1}$ in the
expansion of $\pi\cot(\pi z)$, and the vanishing contour gives
$2\zeta(2k) + a_{2k-1} = 0$: $\zeta(2k) = -a_{2k-1}/2$, a
rational multiple of $\pi^{2k}$ since the cotangent's
coefficients are. One more division step yields $a_5 =
-\frac{2\pi^6}{945}$, whence $\zeta(6) = \frac{\pi^6}{945}$.
For $\zeta(3)$: the natural kernel $g = \pi\cot(\pi z)/z^3$
produces $\sum_{n\neq0}\frac1{n^3} = 0$ by oddness --- the
method proves $0 = 0$ and is structurally blind to odd zeta
values (no closed form for $\zeta(3)$ is known; its
irrationality, Ap\'ery 1978, needed entirely different
ideas).

\textbf{9.} $F(z) = \frac{\pi\cot(\pi z)}{(z - w)(z + w)}$
has poles at the integers (\omterm{def:b3:residues:singularities}{residues} $\frac1{n^2 - w^2}$,
noting the sign: $f(n) = \frac1{(n-w)(n+w)} = \frac1{n^2 -
w^2}$) and simple poles at $\pm w$ with \omterm{def:b3:residues:singularities}{residues}
$\frac{\pi\cot(\pm\pi w)}{\pm2w} = \frac{\pi\cot(\pi
w)}{2w}$ each ($\cot$ is odd). Question 4's argument
($\abs f = O(\abs z^{-2})$) gives
\[
\sum_{n\in\Z}\frac1{n^2 - w^2} + \frac{\pi\cot(\pi w)}{w} =
0,
\qquad\text{i.e.}\qquad
\pi\cot(\pi w) = \frac1w + \sum_{n\geq1}\frac{2w}{w^2 -
n^2},
\]
(the $n = 0$ term is $-\frac1{w^2}$; regroup $\pm n$). Normal
convergence on \omterm{def:b3:topology:compact}{compacts} of $\C\setminus\Z$: for $\abs w \leq
R$ and $n \geq 2R$, $\abs{\frac{2w}{w^2 - n^2}} \leq
\frac{2R}{n^2 - R^2} \leq \frac{8R}{3n^2}$.

\textbf{10.} For $\abs w \leq r < 1$: $\frac{2w}{w^2 - n^2} =
-\frac{2w}{n^2}\cdot\frac1{1 - w^2/n^2} =
-2\sum_{k\geq0}\frac{w^{2k+1}}{n^{2k+2}}$, with
$\abs{\text{terms}} \leq 2r^{2k+1}/n^{2k+2}$, summable over
$(n, k)$: Fubini for series rearranges
\[
\pi\cot(\pi w) = \frac1w -
2\sum_{k\geq0}\zeta(2k+2)\,w^{2k+1} .
\]
Matching with question 2: $-2\zeta(2) = -\frac{\pi^2}3$ and
$-2\zeta(4) = -\frac{\pi^4}{45}$ --- the same values. Three
roads to $\frac{\pi^2}6$: \omterm{thm:b3:hilbert:parseval}{Parseval} (Fourier series), the
trace of the string's Green operator, and the cotangent's two
expansions; that a sum over frequencies, an operator trace,
and a \omterm{def:b3:holomorphic:contour}{contour integral} agree is no accident --- each is a
face of the same spectral identity.

\textbf{11.} Fix $R \geq 1$ and let $n_0$ be the smallest
integer $\geq 2R$. For $\abs z \leq R$ and $n \geq n_0$:
$\abs{z^2/n^2} \leq \frac14$, so $1 - z^2/n^2 \in
\bar D(1, \frac14)$, where the principal logarithm is
\omterm{def:b3:holomorphic:holo}{holomorphic}, and
\[
\Bigl|\log\Bigl(1 - \frac{z^2}{n^2}\Bigr)\Bigr|
\leq \sum_{k\geq1}\frac1k\,\Bigl|\frac{z^2}{n^2}\Bigr|^k
\leq \frac{\abs{z^2/n^2}}{1 - \abs{z^2/n^2}}
\leq \frac{2R^2}{n^2} :
\]
the sum $S(z) = \sum_{n\geq n_0}\log(1 - z^2/n^2)$ converges
normally on $\bar D(0, R)$, with \omterm{def:b3:holomorphic:holo}{holomorphic} partial sums
$S_N$ and $\abs{S_N} \leq 2R^2\zeta(2)$ uniformly. Since
$\abs{\eu^a - \eu^b} \leq
\eu^{\max(\abs a,\abs b)}\abs{a - b}$ (mean value bound on
the segment), the tail products $\prod_{n_0\leq n\leq N} =
\eu^{S_N}$ converge uniformly on $\bar D(0, R)$ to the
zero-free $\eu^S$. Multiplying by the fixed polynomial
$z\prod_{n<n_0}(1 - z^2/n^2)$: $P_N \to P$ uniformly on
$\bar D(0, R)$, and
\cref{thm:b3:holomorphic:weierstrassconv} makes $P$
\omterm{def:b3:holomorphic:holo}{holomorphic} there; $R$ being arbitrary, $P$ is entire. On
$\bar D(0, R)$ the zeros of $P$ are those of the polynomial
prefactor --- the integers of modulus $\leq R$, each simple
($(1 - z/n)(1 + z/n)$ has distinct simple zeros, $\eu^S$
none): the zero set of $P$ is $\Z$, all zeros simple.

\textbf{12.} Logarithmic differentiation of the finite
product, away from its zeros:
\[
\frac{P_N'(z)}{P_N(z)} = \frac1z +
\sum_{n=1}^{N}\frac{-2z/n^2}{1 - z^2/n^2}
= \frac1z + \sum_{n=1}^{N}\frac{2z}{z^2 - n^2} .
\]
On a \omterm{def:b3:topology:compact}{compact} $K \subseteq \C\setminus\Z$: $P_N \to P$ and
$P_N' \to P'$ uniformly
(\cref{thm:b3:holomorphic:weierstrassconv}), and $\min_K
\abs P > 0$ ($P$ vanishes only on $\Z$), so eventually
$\abs{P_N} \geq \frac12\min_K\abs P$ and $P_N'/P_N \to
P'/P$ uniformly on $K$. The middle member converges to
$\frac1z + \sum_{n\geq1}\frac{2z}{z^2-n^2} = \pi\cot(\pi z)$
by question 9: hence $P'/P = \pi\cot(\pi z)$ on
$\C\setminus\Z$.

\textbf{13.} $\sin(\pi z)$ and $P$ are entire with the same
zero set $\Z$, all zeros simple (questions 1 and 11). Near
$m \in \Z$ write $\sin(\pi z) = (z - m)\,\sigma(z)$ and
$P(z) = (z - m)\,\psi(z)$ with $\sigma, \psi$ \omterm{def:b3:holomorphic:holo}{holomorphic}
and nonvanishing at $m$ (factor the power series): $Q =
\sin(\pi z)/P = \sigma/\psi$ extends \omterm{def:b3:holomorphic:holo}{holomorphically} and
zero-free across each integer, and is zero-free on
$\C\setminus\Z$ as a quotient of zero-free functions. There,
\[
\frac{Q'}{Q} = \frac{(\sin\pi z)'}{\sin\pi z} -
\frac{P'}{P} = \pi\cot(\pi z) - \pi\cot(\pi z) = 0 ,
\]
so the entire function $Q'$ vanishes on $\C\setminus\Z$,
hence everywhere by \omterm{def:b3:topology:continuity}{continuity}: $Q$ is constant. As $z \to
0$: $\sin(\pi z)/z \to \pi$ and $P(z)/z \to 1$, so $Q =
\pi$:
\[
\sin(\pi z) = \pi z\prod_{n\geq1}
\Bigl(1 - \frac{z^2}{n^2}\Bigr) .
\]

\textbf{14.} At $z = \frac12$: $1 = \sin\frac\pi2 =
\frac\pi2\prod_{n\geq1}\bigl(1 - \frac1{4n^2}\bigr) =
\frac\pi2\prod\frac{4n^2-1}{4n^2}$, so
\[
\frac\pi2 = \prod_{n\geq1}\frac{4n^2}{4n^2 - 1}
= \lim_{N\to\infty}\prod_{n=1}^N
\frac{(2n)(2n)}{(2n-1)(2n+1)}
= \lim_{N\to\infty}
\frac{2\cdot2\cdot4\cdot4\cdots(2N)(2N)}
{1\cdot3\cdot3\cdot5\cdots(2N-1)(2N+1)} :
\]
Wallis's product, a one-line corollary of Euler's
factorization.

\textbf{15.} For $\abs z \leq r < 1$ every factor lies in
$D(1, r^2) \subseteq D(1, 1)$, so $P(z)/z = \exp\bigl(
\sum_{n\geq1}\log(1 - z^2/n^2)\bigr)$: each partial product
is the exponential of a partial sum, and both sides pass to
the limit by \omterm{def:b3:topology:continuity}{continuity} of $\exp$. The double series
\[
\sum_{n\geq1}\log\Bigl(1 - \frac{z^2}{n^2}\Bigr)
= -\sum_{n\geq1}\sum_{k\geq1}\frac{z^{2k}}{k\,n^{2k}}
= -\sum_{k\geq1}\frac{\zeta(2k)}k\,z^{2k}
= -\zeta(2)\,z^2 + O(z^4)
\]
rearranges by Fubini for series: $\sum_{n,k}
\frac{r^{2k}}{kn^{2k}} \leq \sum_k\zeta(2k)r^{2k} \leq
\zeta(2)\frac{r^2}{1-r^2} < \infty$. Hence
\[
P(z) = z\,\exp\bigl(-\zeta(2)z^2 + O(z^4)\bigr)
= z - \zeta(2)\,z^3 + O(z^5) ,
\]
and question 13 compares this with $\sin(\pi z) = \pi z -
\frac{\pi^3}6z^3 + O(z^5)$: $\pi\zeta(2) = \frac{\pi^3}6$,
i.e.\ $\zeta(2) = \frac{\pi^2}6$. The additive face
(question 10) and the multiplicative face compute the same
number.

\textbf{16.} On a \omterm{def:b3:topology:compact}{compact} $K \subseteq \C\setminus\Z$ the
partial sums $S_N = \frac1z + \sum_{n\leq N}\bigl(
\frac1{z-n} + \frac1{z+n}\bigr)$ (question 9, terms
regrouped as $\frac{2z}{z^2-n^2} = \frac1{z-n} +
\frac1{z+n}$) converge uniformly to $\pi\cot(\pi z)$, so
\cref{thm:b3:holomorphic:weierstrassconv} gives $S_N' \to
(\pi\cot\pi z)' = -\pi^2/\sin^2(\pi z)$ uniformly on $K$.
Since $S_N' = -\sum_{\abs n\leq N}(z - n)^{-2}$:
\[
\frac{\pi^2}{\sin^2(\pi z)} =
\sum_{n\in\Z}\frac1{(z - n)^2} ,
\]
the convergence normal on \omterm{def:b3:topology:compact}{compacts} of $\C\setminus\Z$
(terms $O(n^{-2})$).

\textbf{17.} At $z = \frac12$ the left side is $\pi^2$; on
the right, $(\frac12 - n)^2 = \frac{(2n-1)^2}4$ with $2n -
1$ running over all odd integers exactly once as $n$ runs
over $\Z$:
\[
\pi^2 = \sum_{n\in\Z}\frac{4}{(2n-1)^2}
= 8\sum_{m\geq0}\frac1{(2m+1)^2},
\qquad
\sum_{m\geq0}\frac1{(2m+1)^2} = \frac{\pi^2}8 .
\]
Parity split: $\zeta(2) = \frac{\pi^2}8 +
\sum_{n\geq1}\frac1{(2n)^2} = \frac{\pi^2}8 +
\frac{\zeta(2)}4$, so $\frac34\zeta(2) = \frac{\pi^2}8$ and
$\zeta(2) = \frac{\pi^2}6$ once more.

\textbf{18.} With $c = \cot\theta$ and $\cot(2\theta) =
\frac{c^2-1}{2c}$: $\cot\theta - 2\cot(2\theta) = c -
\frac{c^2-1}c = \frac1c = \tan\theta$. Hence $\pi\tan(\pi z)
= \pi\cot(\pi z) - 2\pi\cot(2\pi z)$, and question 9 at $z$
and at $2z$ gives
\[
\pi\cot(\pi z) = \frac1z +
\sum_{n\geq1}\frac{2z}{z^2 - n^2},
\qquad
2\pi\cot(2\pi z) = \frac1z +
\sum_{n\geq1}\frac{8z}{4z^2 - n^2} .
\]
Both series converge absolutely at each fixed $z$ off the
poles, so the difference may be regrouped at will: in the
second series the even terms $n = 2m$ give $\frac{8z}{4z^2 -
4m^2} = \frac{2z}{z^2 - m^2}$ and cancel the first series
entirely, leaving
\[
\pi\tan(\pi z) = -\sum_{m\geq0}\frac{8z}{4z^2 - (2m+1)^2}
= \sum_{m\geq0}\frac{8z}{(2m+1)^2 - 4z^2} ,
\]
normally on \omterm{def:b3:topology:compact}{compacts} avoiding $\frac12 + \Z$ (terms
$O(m^{-2})$).

\textbf{19.} For $\abs z \leq r < \frac12$:
\[
\frac{8z}{(2m+1)^2 - 4z^2} = \frac{8z}{(2m+1)^2}
\sum_{k\geq0}\Bigl(\frac{4z^2}{(2m+1)^2}\Bigr)^{k},
\]
with $\sum_{m,k}8r\,(4r^2)^k(2m+1)^{-2k-2} < \infty$ since
$4r^2 < 1$: Fubini rearranges the double sum into
\[
\pi\tan(\pi z) =
\sum_{k\geq0}8\cdot4^k\,\lambda(2k+2)\,z^{2k+1} .
\]
Against $\pi\tan(\pi z) = \pi^2z + \frac{\pi^4}3z^3 +
O(z^5)$: the coefficient of $z$ gives $8\lambda(2) = \pi^2$
--- question 17 again --- and that of $z^3$ gives
$32\lambda(4) = \frac{\pi^4}3$, i.e.\ $\lambda(4) =
\frac{\pi^4}{96}$. Removing the even denominators,
$\lambda(4) = \zeta(4) - 2^{-4}\zeta(4) =
\frac{15}{16}\zeta(4)$: $\zeta(4) = \frac{16}{15}\cdot
\frac{\pi^4}{96} = \frac{\pi^4}{90}$, matching question 7.

\textbf{20.} $\cot\frac\theta2 - \cot\theta =
\frac{\cos\frac\theta2\,\sin\theta -
\cos\theta\,\sin\frac\theta2}{\sin\frac\theta2\,\sin\theta}
= \frac{\sin\frac\theta2}{\sin\frac\theta2\,\sin\theta} =
\frac1{\sin\theta}$, the numerator being $\sin(\theta -
\frac\theta2)$. With $\theta = \pi z$, question 9 at
$\frac z2$ reads $\pi\cot\frac{\pi z}2 = \frac2z +
\sum_{n\geq1}\frac{4z}{z^2 - 4n^2}$, so
\[
\frac{\pi}{\sin(\pi z)}
= \pi\cot\frac{\pi z}2 - \pi\cot(\pi z)
= \frac1z + \sum_{n\geq1}\frac{4z}{z^2 - 4n^2}
- \sum_{n\geq1}\frac{2z}{z^2 - n^2} .
\]
Absolute convergence permits the parity regrouping: even $n
= 2m$ in the subtracted series contributes
$\frac{2z}{z^2-4m^2}$, leaving $+\frac{2z}{z^2-4m^2}$ from
the first sum, while odd $n$ survives with sign $-$:
\[
\frac{\pi}{\sin(\pi z)} = \frac1z +
\sum_{n\geq1}(-1)^n\,\frac{2z}{z^2 - n^2} .
\]
Signs check: $\operatorname{Res}\bigl(\pi/\sin(\pi z),
n\bigr) = \pi/(\pi\cos\pi n) = (-1)^n$
(\cref{met:b3:residues:compute}), and $(-1)^n\frac{2z}{z^2 -
n^2} = \frac{(-1)^n}{z-n} + \frac{(-1)^n}{z+n}$ carries
exactly that \omterm{def:b3:residues:singularities}{residue} at $\pm n$.

\textbf{21.} For $\abs z \leq r < 1$, expanding each term as
in question 19 ($(-1)^n\frac{2z}{z^2-n^2} =
(-1)^{n-1}\frac{2z}{n^2}\sum_k\frac{z^{2k}}{n^{2k}}$) and
applying Fubini:
\[
\frac{\pi}{\sin(\pi z)} = \frac1z +
2\sum_{k\geq0}\eta(2k+2)\,z^{2k+1},
\qquad
\eta(s) = \sum_{n\geq1}\frac{(-1)^{n-1}}{n^s} .
\]
Taylor side: $\sin(\pi z) = \pi z(1 - \frac{(\pi z)^2}6 +
O(z^4))$ gives $\frac\pi{\sin\pi z} = \frac1z +
\frac{\pi^2}6z + O(z^3)$: $2\eta(2) = \frac{\pi^2}6$, so
$\eta(2) = \frac{\pi^2}{12}$. Consistency: $\eta(2) =
\zeta(2) - 2\sum_n(2n)^{-2} = (1 - 2^{1-2})\zeta(2) =
\frac{\zeta(2)}2 = \frac{\pi^2}{12}$.

\textbf{22.} At $z = \frac12$, question 20 gives
\[
\pi = 2 + \sum_{n\geq1}(-1)^n\frac{1}{\frac14 - n^2}
= 2 + 4\sum_{n\geq1}\frac{(-1)^{n-1}}{(2n-1)(2n+1)} .
\]
With $\frac1{(2n-1)(2n+1)} = \frac12\bigl(\frac1{2n-1} -
\frac1{2n+1}\bigr)$, and both alternating series convergent
(Leibniz criterion), the sum splits as $\frac12\bigl[L -
(1 - L)\bigr] = L - \frac12$, where $L = 1 - \frac13 +
\frac15 - \cdots$ and $\frac13 - \frac15 + \frac17 - \cdots
= 1 - L$. Hence $\pi = 2 + 4(L - \frac12) = 4L$:
\[
\frac\pi4 = 1 - \frac13 + \frac15 - \frac17 + \cdots
\]
The moral: each kernel is \omterm{def:b3:residues:singularities}{meromorphic} with poles on an
arithmetic progression and prescribed \omterm{def:b3:residues:singularities}{residues}. The
cotangent puts \omterm{def:b3:residues:singularities}{residue} $1$ at every integer and sums
$f(n)$; its derivative squares the poles and prices
$\lambda(2)$; the tangent moves the poles to $\frac12 + \Z$
and prices the odd denominators; $\pi/\sin$ keeps the
integer poles but alternates the \omterm{def:b3:residues:singularities}{residues} $(-1)^n$, hence
the alternating series. The first-order kernels are all
odd functions: pairing $n$ with $-n$ doubles the even-power
coefficients and annihilates the odd ones, so $\zeta(2k)$
pours out mechanically while $\zeta(3)$ never appears. The
blindness is parity, not lack of technique.

\textbf{23.} With $w = 2\iu\pi z$:
\[
\iu\pi z + \frac{2\iu\pi z}{\eu^{2\iu\pi z} - 1}
= \iu\pi z\,\frac{\eu^{2\iu\pi z} + 1}{\eu^{2\iu\pi z} - 1}
= \iu\pi z\,\frac{\eu^{\iu\pi z} +
\eu^{-\iu\pi z}}{\eu^{\iu\pi z} - \eu^{-\iu\pi z}}
= \pi z\,\frac{\cos\pi z}{\sin\pi z} = \pi z\cot(\pi z) .
\]
Hence, using the generating function on $w = 2\iu\pi z$ (and
$B_1 = -\frac12$ cancelling the $\iu\pi z$ term, odd $B$'s
vanishing beyond):
\[
\pi z\cot(\pi z) = \sum_{k\geq0}\frac{B_{2k}}{(2k)!}
(2\iu\pi z)^{2k}
= 1 + \sum_{k\geq1}(-1)^k\frac{(2\pi)^{2k}B_{2k}}{(2k)!}
z^{2k} .
\]
On the other hand the partial fraction expansion (Part IV)
gives $\pi z\cot(\pi z) = 1 - 2\sum_{k\geq1}\zeta(2k)z^{2k}$
(expand each $\frac{2z^2}{z^2 - n^2} =
-2\sum_k\frac{z^{2k}}{n^{2k}}$ and sum over $n$, normal
convergence justifying the interchange for $\abs z < 1$).
Comparing coefficients: $-2\zeta(2k) =
(-1)^k\frac{(2\pi)^{2k}B_{2k}}{(2k)!}$, the stated formula.

\textbf{24.} $k = 1$: $\frac{(2\pi)^2}{2\cdot2}\cdot\frac16
= \frac{\pi^2}6$. $k = 2$: $-\frac{(2\pi)^4}{2\cdot24}
\cdot\bigl(-\frac1{30}\bigr) = \frac{16\pi^4}{48\cdot30} =
\frac{\pi^4}{90}$. $k = 3$: $\frac{(2\pi)^6}{2\cdot720}
\cdot\frac1{42} = \frac{64\pi^6}{1440\cdot42} =
\frac{\pi^6}{945}$.

\textbf{25.} Write $C(z) = \pi z\cot(\pi z) = 1 -
2\sum_{k\geq1}\zeta(2k)z^{2k}$ (question 23). Direct
differentiation of $C = \pi z\cot(\pi z)$, using $(\cot u)'
= -1 - \cot^2u$:
\[
zC'(z) = \pi z\cot(\pi z) - \pi^2z^2\bigl(1 +
\cot^2(\pi z)\bigr) = C - \pi^2z^2 - C^2 .
\]
Now expand both sides in powers of $z^2$. Left side:
$\sum_k(-4k)\,\zeta(2k)\,z^{2k}$. Right side: squaring the
series,
\[
C - C^2 = 2\sum_{k\geq1}\zeta(2k)z^{2k}
- 4\sum_{k\geq2}\Bigl(\sum_{j=1}^{k-1}\zeta(2j)
\zeta(2k-2j)\Bigr)z^{2k},
\]
and the term $-\pi^2z^2 = -6\zeta(2)z^2$ only adjusts $k =
1$. Comparing the coefficients of $z^{2k}$ for $k \geq 2$:
\[
-4k\,\zeta(2k) = 2\,\zeta(2k) -
4\sum_{j=1}^{k-1}\zeta(2j)\,\zeta(2k-2j),
\]
i.e.\ $\bigl(k + \frac12\bigr)\zeta(2k) =
\sum_{j=1}^{k-1}\zeta(2j)\zeta(2k-2j)$. (At $k = 1$ the
identity reads $-4\zeta(2) = 2\zeta(2) - 6\zeta(2)$: a
consistency check, not new information.) Applications: $k =
2$: $\frac52\zeta(4) = \zeta(2)^2 = \frac{\pi^4}{36}$, so
$\zeta(4) = \frac{\pi^4}{90}$; $k = 3$: $\frac72\zeta(6) =
2\zeta(2)\zeta(4) = \frac{\pi^6}{270}$, so $\zeta(6) =
\frac{2}{7}\cdot\frac{\pi^6}{270} = \frac{\pi^6}{945}$. One
transcendental seed ($\zeta(2) = \frac{\pi^2}6$), and pure
algebra generates every even zeta value.
\end{solution}
