\section*{Chapter \ref{ch:b3:submanifolds} --- Submanifolds of
$\R^n$}

\begin{solution}{exo:b3:submanifolds:1}
(a) Each is $F^{-1}(0)$ for a submersion: $\norm x^2 - 1$ on
$\R^n\setminus\{0\}$ (gradient $2x \neq 0$): dimension $n-1$;
$x^2 + y^2 - z^2 - 1$ (gradient $(2x, 2y, -2z) \neq 0$ on the
hyperboloid, where $x^2 + y^2 = 1 + z^2 > 0$): dimension $2$;
the torus function $G = (\rho - R)^2 + z^2 - r^2$, $\rho =
\sqrt{x^2+y^2}$, is $\mathcal C^\infty$ near the torus ($\rho
\geq R - r > 0$ there) with $\nabla G \neq 0$ (its
$z$-component is $2z$, and where $z = 0$ the radial component
is $2(\rho - R)\ne0$ since $\abs{\rho - R} = r$): dimension
$2$.

(b) For small $\varepsilon$,
$(C\setminus\{0\})\cap B(0,\varepsilon)$ has exactly $2$
\omterm{def:b3:topology:components}{connected components} (upper and lower punctured nappes, each
\omterm{def:b3:topology:pathconnected}{path-connected}: connect through circles and rays). If $C$
were a $2$-submanifold at $0$, a straightening would give a
\omterm{def:b3:topology:continuity}{homeomorphism} from $C\cap\Omega$ onto an open piece of a
plane sending $0$ to a point $p$; small punctured
\omterm{def:b3:topology:topology}{plane-neighborhoods} of $p$ have \emph{one} component, and
\omterm{def:b3:topology:continuity}{homeomorphisms} preserve the number of components of punctured
\omterm{def:b3:topology:topology}{neighborhoods}: contradiction.
\end{solution}

\begin{solution}{exo:b3:submanifolds:2}
(a) $T_aS^{n-1} = \ker\bigl(2a^{\mathsf T}\bigr) = a^\perp$.
(b) At $p = ((R+r)\cos\theta, (R+r)\sin\theta, 0)$: $\nabla G
= (2r\cos\theta, 2r\sin\theta, 0)$, so the tangent plane is
$\operatorname{Vect}\bigl((-\sin\theta, \cos\theta, 0),\ (0,
0, 1)\bigr)$: the vertical plane tangent to the outer
equator.
(c) The helix is an embedded curve with $\varphi'(t_0) =
(-\sin t_0, \cos t_0, 1) \neq 0$: the tangent line at
$\varphi(t_0)$ is $\varphi(t_0) + \R\,\varphi'(t_0)$, as
\cref{prop:b3:submanifolds:tangent}(2) prescribes.
\end{solution}

\begin{solution}{exo:b3:submanifolds:3}
(a) $Df(x,y) = \bigl(\begin{smallmatrix}2x & -2y\\ 2y &
2x\end{smallmatrix}\bigr)$, $\det = 4(x^2 + y^2)$: the
theorem applies at every point except the origin.
(b) $f(-z) = f(z)$: never injective on a set symmetric about
$0$; on $\R^2\setminus\{0\}$ it is a local diffeomorphism
everywhere yet $2$-to-$1$ globally. Near $(1, 0) = f(\pm(1,
0))$, the two inverses are the two square-root branches: in
complex notation $w \mapsto \pm\sqrt w$ (principal branch),
i.e.
\[
(u, v) \longmapsto \pm\Bigl(\sqrt{\tfrac{u +
\sqrt{u^2+v^2}}{2}},\ \frac{v}{2\sqrt{(u +
\sqrt{u^2+v^2})/2}}\Bigr).
\]
(c) Jacobian $r > 0$: local diffeomorphism on
$\intoo0\infty\times\R$, but $\theta \mapsto \theta + 2\pi$
gives the same point: locally invertible (angle determined up
to $2\pi$ on a half-plane), never globally.
\end{solution}

\begin{solution}{exo:b3:submanifolds:4}
(a) $\nabla F = 3(x^2 - y,\ y^2 - x)$ vanishes iff $y = x^2$
and $x = y^2$, i.e.\ $x^4 = x$: at $(0,0)$ and $(1,1)$; only
$(0,0)$ lies on $\mathcal C$ ($F(1,1) = -1$). So on $\mathcal
C\setminus\{0\}$, $F$ is a submersion: a $1$-submanifold,
locally a graph $y = \psi(x)$ where $F_y = 3(y^2 - x) \neq 0$
and $x = \chi(y)$ where $F_x = 3(x^2 - y) \neq 0$ (at least
one holds off the origin).

(b) At $(\frac32, \frac32)$: $\nabla F = 3(\frac94 - \frac32)
(1, 1) = \frac94(1,1)$: tangent line $x + y = 3$.

(c) The rational parametrization $x = \frac{3t}{1 + t^3}$,
$y = \frac{3t^2}{1+t^3}$ passes through $0$ at $t = 0$ with
velocity $(3, 0)$; exchanging $x \leftrightarrow y$ (the
curve is symmetric, or reparametrize by $1/t$) gives a second
$\mathcal C^1$ curve through $0$ with velocity $(0, 3)$. Two
independent tangent directions are impossible for a
$1$-submanifold (its \omterm{def:b3:submanifolds:tangent}{tangent space} is a line,
\cref{prop:b3:submanifolds:tangent}): $\mathcal C$ is not a
submanifold at the origin --- a transverse self-crossing.
\end{solution}

\begin{solution}{exo:b3:submanifolds:5}
(a) $F(M + H) = M^{\mathsf T}M + M^{\mathsf T}H + H^{\mathsf
T}M + H^{\mathsf T}H$: $DF(M)(H) = M^{\mathsf T}H +
H^{\mathsf T}M$. For $M \in O_n$ and $S$ symmetric, $H =
\frac12MS$ gives $DF(M)(H) = \frac12(S + S^{\mathsf T}) = S$:
surjective onto $S_n$.

(b) $O_n = F^{-1}(I)$ with $F$ a submersion (onto $S_n$, of
dimension $\frac{n(n+1)}2$) at each of its points:
a submanifold of dimension $n^2 - \frac{n(n+1)}2 =
\frac{n(n-1)}2$. \omterm{def:b3:topology:compact}{Compact}: closed ($F$ \omterm{def:b3:topology:continuity}{continuous}) and bounded
(columns are unit vectors). Tangent at $I$: $\ker DF(I) =
\{H : H + H^{\mathsf T} = 0\}$.

(c) $\bigl(\eu^{tH}\bigr)^{\mathsf T}\eu^{tH} =
\eu^{tH^{\mathsf T}}\eu^{tH} = \eu^{-tH}\eu^{tH} = I$
(transpose passes through the series; the exponentials of
commuting matrices multiply,
\cref{thm:b3:ode:matrixexp}).
\end{solution}

\begin{solution}{exo:b3:submanifolds:6}
(a) $\det(M + H) = \det M\,\det(I + M^{-1}H) = \det M\bigl(1
+ \operatorname{tr}(M^{-1}H) + O(\norm H^2)\bigr)$ for
invertible $M$ (expansion of $\det$ near $I$: the linear term
of $\prod(1 + \lambda_i)$); with $\det M\cdot M^{-1} =
\operatorname{com}(M)^{\mathsf T}$:
$D\det(M)(H) = \operatorname{tr}\bigl(\operatorname{com}
(M)^{\mathsf T}H\bigr)$, and the formula extends to all $M$
by \omterm{ex:b3:lebesgue:gamma}{density} and \omterm{def:b3:topology:continuity}{continuity}. On $SL_n$, $\det M = 1$:
$D\det(M) \ne 0$ (its value on $H = M$ is
$\operatorname{tr}(I)\cdot\dots = n\det M = n$).

(b) $SL_n = \det^{-1}(1)$ with $\det$ a submersion there
(values in $\R$): dimension $n^2 - 1$; $T_ISL_n = \ker
D\det(I) = \{H : \operatorname{tr}H = 0\}$.

(c) $GL_n$ is an \emph{open} subset of $M_n(\R)$
(\cref{exo:b3:topology:8}): a submanifold of full dimension
$n^2$ (straightening: the identity chart).
\end{solution}

\begin{solution}{exo:b3:submanifolds:7}
(a) $(y, x) = \lambda(2x, 2y)$ and $x^2 + y^2 = 1$: $y =
2\lambda x$, $x = 2\lambda y$ give $x^2 = y^2 = \frac12$.
Values of $xy$: $\pm\frac12$: maximum $\frac12$ at
$\pm\frac1{\sqrt2}(1,1)$, minimum $-\frac12$ at
$\pm\frac1{\sqrt2}(1,-1)$ (the constraint set is \omterm{def:b3:topology:compact}{compact}:
extrema exist).

(b) On the \omterm{def:b3:topology:interior}{interior} of the simplex ($p_i > 0$), Lagrange for
$H(p) = -\sum p_i\ln p_i$ with constraint $\sum p_i = 1$:
$-\ln p_i - 1 = \lambda$ for all $i$: all $p_i$ equal, $p_i =
\frac1n$, with $H = \ln n$. The maximum over the \omterm{def:b3:topology:compact}{compact}
simplex is attained; if it were attained on the boundary
(some $p_i = 0$), the distribution lives on $\leq n - 1$
points and by induction $H \leq \ln(n-1) < \ln n$: the
\omterm{def:b3:topology:interior}{interior} critical point is the global maximum --- uniform
ignorance maximizes entropy.

(c) Minimize $(x-1)^2 + y^2$ on the \omterm{def:b3:topology:compact}{compact} ellipse:
$(2(x{-}1), 2y) = \lambda(\frac x2, 2y)$. If $y \neq 0$:
$\lambda = 1$, then $2(x - 1) = \frac x2$ gives $x =
\frac43$, $y^2 = 1 - \frac49 = \frac59$: distance$^2$ $=
\frac19 + \frac59 = \frac23$. If $y = 0$: $x = \pm2$,
distances $1$ and $3$. Closest points: $\bigl(\frac43,
\pm\frac{\sqrt5}3\bigr)$, at distance $\sqrt{2/3} < 1$. The
multiplier equation says the segment from $(1,0)$ to the
closest point is parallel to $\nabla$(ellipse): it meets the
ellipse orthogonally, as geometry demands.
\end{solution}

\begin{solution}{exo:b3:submanifolds:8}
Induction on $n$; $n = 1$ trivial. The Rayleigh function
$f(x) = \langle Ax, x\rangle$ attains its maximum $\lambda_1$
on the \omterm{def:b3:topology:compact}{compact} $S^{n-1}$ at some $v_1$; Lagrange
(\cref{thm:b3:submanifolds:lagrange}, sphere as level set)
gives $2Av_1 = 2\lambda v_1$, and $\lambda = \langle Av_1,
v_1\rangle = \lambda_1$. The hyperplane $v_1^\perp$ is
$A$-invariant (symmetry: $\langle Av, v_1\rangle = \langle
v, Av_1\rangle = 0$); the restriction is symmetric, and
induction yields an orthonormal eigenbasis $v_2, \dots, v_n$
of $v_1^\perp$ with eigenvalues $\lambda_2 \geq \dots \geq
\lambda_n$, each the maximum of $f$ on the sphere of the
remaining orthocomplement. Courant--Fischer follows exactly
as in \cref{exo:b3:spectral:8}: expand $x = \sum c_iv_i$;
on a $k$-dimensional test space intersect with
$\operatorname{Vect}(v_k, \dots, v_n)$ (dimension count in
$\R^n$) to get $\min \leq \lambda_k$, and
$\operatorname{Vect}(v_1, \dots, v_k)$ achieves $\min =
\lambda_k$.
\end{solution}

\begin{solution}{exo:b3:submanifolds:9}
Scaling each column to unit norm divides $\abs{\det}$ by
$\prod\norm{c_i}$: it suffices to prove $\abs{\det M} \leq 1$
when all columns are unit, with equality iff $M \in O_n$. The
function $\det$ is \omterm{def:b3:topology:continuity}{continuous} on the \omterm{def:b3:topology:compact}{compact}
$(S^{n-1})^n$: it attains a maximum $m \geq \det I = 1 > 0$
at some $M$. Fixing all columns but the $i$-th, $\det$ is
linear in $c_i$ with gradient the $i$-th column of
$\operatorname{com}(M)$; Lagrange on the $i$-th sphere:
$\operatorname{com}(M)_{\cdot i} = \lambda_i c_i$. The
identity $M^{\mathsf T}\operatorname{com}(M) = \det(M)\,I$
reads $\langle c_j, \operatorname{com}(M)_{\cdot i}\rangle =
\det(M)\,\delta_{ij}$, i.e.\ $\lambda_i\langle c_j,
c_i\rangle = \det(M)\delta_{ij}$; taking $j = i$: $\lambda_i
= \det M = m \neq 0$, and then $j \neq i$ gives $\langle c_i,
c_j\rangle = 0$: the columns are orthonormal, $M \in O_n$,
$m = \abs{\det M} = 1$. Hence $\abs{\det} \leq
\prod\norm{c_i}$ always, with equality exactly for orthogonal
columns (rescale back): a parallelepiped's volume is largest,
for given edge lengths, when the edges are perpendicular.
\end{solution}

\begin{solution}{exo:b3:submanifolds:10}
$y = \pm\sqrt{1 - x^2}$ works near every point with $y \neq
0$; $x = \pm\sqrt{1 - y^2}$ near every point with $x \neq 0$;
at $(1, 0)$: the graph $x = \sqrt{1 - y^2}$ over $y \in
\intoo{-1}1$, which is
\cref{thm:b3:submanifolds:characterizations}(3) with the
coordinates swapped. Some permutation always works because
the tangent line, being one-dimensional, cannot be
simultaneously vertical and horizontal --- but it can be
either, so no fixed choice of ``dependent'' coordinate
serves at every point.
\end{solution}

\begin{solution}{exo:b3:submanifolds:11}
(a) The defining map $F(M) = M^{\mathsf T}M - I$ is
\omterm{def:b3:topology:continuity}{continuous}: $O_n = F^{-1}(0)$ is closed; columns of an
orthogonal matrix are unit vectors, so the Euclidean
(\omterm{thm:b3:galois:finitefields}{Frobenius}) norm is exactly $\sqrt n$: bounded. \omterm{def:b3:topology:compact}{Compact} by
Heine--Borel in $M_n(\R) \cong \R^{n^2}$.

(b) $O_n$ is the level set $F = 0$ studied in the chapter:
$DF(A)H = A^{\mathsf T}H + H^{\mathsf T}A$, surjective onto
symmetric matrices at each $A \in O_n$ (given symmetric $S$,
take $H = \frac12AS$), so $O_n$ is a submanifold of
dimension $n^2 - \frac{n(n+1)}2 = \frac{n(n-1)}2$ with
\[
T_AO_n = \ker DF(A) = \{H : A^{\mathsf T}H \text{
antisymmetric}\} = \{AK : K^{\mathsf T} = -K\} ;
\]
at $A = I$ these are the antisymmetric matrices.

(c) $\exp(tK)^{\mathsf T}\exp(tK) = \exp(tK^{\mathsf T})
\exp(tK) = \exp(-tK)\exp(tK) = I$ (the two matrices $\pm tK$
commute, so the product of exponentials is the exponential
of the sum): $\exp(tK) \in O_n$, and $\gamma(t) =
A\exp(tK)$ is a curve in $O_n$ with $\gamma(0) = A$,
$\gamma'(0) = AK$. As $K$ runs over antisymmetric matrices,
$AK$ sweeps $T_AO_n$: the exponential realizes the whole
\omterm{def:b3:submanifolds:tangent}{tangent space} by explicit curves --- the Lie-group shortcut
that \cref{pb:b3:submanifolds:1} exploits for $SO(3)$.
\end{solution}

\begin{solution}{exo:b3:submanifolds:12}
Existence: intersect $M$ with a large closed ball around
$p$ to get a nonempty \omterm{def:b3:topology:compact}{compact}; the \omterm{def:b3:topology:continuity}{continuous} distance
attains its minimum there, and points outside the ball are
farther. First-order condition: for a curve $\gamma$ in $M$
with $\gamma(0) = x_0$, the function $h(t) = \norm{\gamma(t)
- p}^2$ is differentiable with a minimum at $0$:
\[
0 = h'(0) = 2\,\langle\gamma'(0),\ x_0 - p\rangle,
\]
and $\gamma'(0)$ sweeps $T_{x_0}M$: $p - x_0 \perp
T_{x_0}M$. Sphere $S(c, r)$: the \omterm{def:b3:submanifolds:tangent}{tangent space} at $x_0$ is
$(x_0 - c)^\perp$, so $p - x_0 \parallel x_0 - c$: $x_0$
lies on the line through $c$ and $p$, at distance $r$ from
$c$ --- the ray point, as geometry insists. Parabola: at
$x_0 = (x, x^2)$ the tangent is spanned by $(1, 2x)$;
orthogonality to $p - x_0 = (2 - x, -x^2)$ reads
\[
(2 - x) - 2x^3 = 0,
\qquad\text{i.e.}\qquad 2x^3 + x - 2 = 0,
\]
with unique real root ($x \mapsto 2x^3 + x$ is strictly
increasing) $x \approx 0.835$; then $x_0 \approx (0.835,
0.698)$ and $d(p, M) = \sqrt{(2 - 0.835)^2 + 0.698^2}
\approx 1.358$.
\end{solution}

\begin{solution}{pb:b3:submanifolds:1}
\textbf{1.} Map $q = t + x\mathrm i + y\mathrm j + z\mathrm k
\mapsto \bigl(\begin{smallmatrix}\alpha & \beta\\ -\bar\beta
& \bar\alpha\end{smallmatrix}\bigr)$ with $\alpha = t + \iu
x$, $\beta = y + \iu z$: one checks that $1, \mathrm i,
\mathrm j, \mathrm k$ go to $I$, $\bigl(\begin{smallmatrix}
\iu & 0\\ 0 & -\iu\end{smallmatrix}\bigr)$,
$\bigl(\begin{smallmatrix}0 & 1\\ -1 &
0\end{smallmatrix}\bigr)$, $\bigl(\begin{smallmatrix}0 &
\iu\\ \iu & 0\end{smallmatrix}\bigr)$, whose products
reproduce the quaternion table: the map is an injective
algebra morphism, so $\mathbb H$ inherits associativity;
$N(q) = \abs\alpha^2 + \abs\beta^2 = \det$ is multiplicative,
and conjugation corresponds to the adjugate-transpose,
giving $\overline{pq} = \bar q\bar p$. \omterm{ex:b3:groups:actions}{Center}: commuting
with $\mathrm i$ forces $y = z = 0$, with $\mathrm j$ forces
$x = 0$: $\R$.

\textbf{2.} $q\bar q = N(q)$: for $q \neq 0$, $q^{-1} = \bar
q/N(q)$: a division algebra. On $S^3$: $N(pq) = 1$ and
$N(q^{-1}) = 1$: a group; and $S^3 \subseteq \R^4$ is the
unit sphere: a \omterm{def:b3:topology:compact}{compact} $3$-submanifold.

\textbf{3.} $v$ is pure iff $\bar v = -v$; then
$\overline{qv\bar q} = q\bar v\bar q = -qv\bar q$: $\rho_q$
preserves $P$. Linearity is clear; $N(qv\bar q) =
N(q)N(v)N(q) = N(v)$: an isometry of $(P, N) \cong (\R^3,
\norm\cdot^2)$: $\rho_q \in O(3)$. And $\rho_{pq}(v) =
pqv\overline{pq} = p(qv\bar q)\bar p =
\rho_p(\rho_q(v))$: a morphism.

\textbf{4.} $\rho_q = \mathrm{id}$ iff $qv = vq$ for all pure
$v$, iff $q$ commutes with $\mathrm i, \mathrm j, \mathrm k$,
iff $q$ is central (question 1): $q \in \R\cap S^3 =
\{\pm1\}$.

\textbf{5.} Write $q = t + p$ ($t \in \R$, $p$ pure): $1 =
N(q) = t^2 + N(p)$, so $t = \cos\frac\theta2$ and $p =
\sin\frac\theta2\,u$ with $N(u) = 1$ for some $\theta$
(if $p = 0$, $q = \pm1$ acts trivially). Since $u^2 = -N(u)
= -1$, $q$ and $u$ commute, and $\rho_q(u) = qu\bar q =
uq\bar q = u$: the axis. For pure units $w \perp u$: the
product rule $vw = -\langle v, w\rangle + v\wedge w$
(expand in coordinates from the table) gives $uw = u\wedge w
= -wu$. Then
\[
\rho_q(w) = \bigl(\cos\tfrac\theta2 +
\sin\tfrac\theta2u\bigr)\,w\,\bigl(\cos\tfrac\theta2 -
\sin\tfrac\theta2u\bigr)
= \cos\theta\,w + \sin\theta\,(u\wedge w),
\]
using $uwu = -u^2w = w$ and the double-angle formulas: the
rotation of angle $\theta$ in the oriented plane $(w,
u\wedge w)$.

\textbf{6.} Each $\rho_q$ is a rotation about $u$ by
$\theta$: in the orthonormal basis $(u, w, u\wedge w)$ its
matrix has determinant $+1$: $\operatorname{im}\rho \subseteq
SO(3)$. Surjectivity: a matrix $R \in SO(3)$ has $1$ as an
eigenvalue, since
\[
\det(R - I) = \det R\,\det(I - R^{\mathsf T}) = \det(I - R)
= (-1)^3\det(R - I),
\]
so $\det(R - I) = 0$. Take a unit eigenvector $u$; $R$
preserves $u^\perp$ and restricts there to a rotation of
some angle $\theta$ (planar orthogonal, determinant $1$):
$R = \rho_q$ for $q = \cos\frac\theta2 +
\sin\frac\theta2\,u$. With question 4 and the first
isomorphism theorem (\cref{thm:b3:groups:firstiso}): $SO(3)
\cong S^3/\{\pm1\}$.

\textbf{7.} $O_3$ is a \omterm{def:b3:topology:compact}{compact} $3$-dimensional submanifold
(\cref{exo:b3:submanifolds:5}); $\det$ is \omterm{def:b3:topology:continuity}{continuous} on it
with values in $\{\pm1\}$, so $SO(3) = O_3\cap\{\det = 1\}$
is open and closed in $O_3$: a union of \omterm{def:b3:topology:components}{connected components},
hence itself a \omterm{def:b3:topology:compact}{compact} $3$-submanifold, with the same \omterm{def:b3:submanifolds:tangent}{tangent
space} at $I$: the antisymmetric matrices.

\textbf{8.} $A_u^2v = u\wedge(u\wedge v) = \langle u,
v\rangle u - v$ (unit $u$), so $A_u^3v = u\wedge(\langle
u,v\rangle u - v) = -u\wedge v$: $A_u^3 = -A_u$. Splitting
the exponential series by \omterm{def:b3:residues:singularities}{residues} of powers mod the relation
$A^3 = -A$:
\[
\eu^{\theta A_u} = I + \Bigl(\theta - \frac{\theta^3}{3!} +
\cdots\Bigr)A_u + \Bigl(\frac{\theta^2}{2!} -
\frac{\theta^4}{4!} + \cdots\Bigr)A_u^2
= I + \sin\theta\,A_u + (1 - \cos\theta)\,A_u^2 .
\]
On $u$: $A_uu = 0$: fixed. On $w \perp u$: $\eu^{\theta
A_u}w = w + \sin\theta\,u\wedge w + (1 - \cos\theta)(-w) =
\cos\theta\,w + \sin\theta\,u\wedge w$: the rotation of axis
$u$, angle $\theta$ --- Rodrigues. Every rotation has this
form (question 6): $\exp$ is onto $SO(3)$ from the
antisymmetric matrices.

\textbf{9.} With $q(t) = \cos\frac t2 + \sin\frac t2\,u$:
question 5 shows $\rho_{q(t)}$ is the rotation of axis $u$
and angle $t$, i.e.\ $\rho_{q(t)} = \eu^{tA_u}$, whose
derivative at $t = 0$ is $A_u$. The quaternion runs at half
the angle --- the analytic trace of the double cover.

\textbf{10.} $\rho_{q(t)}$ is the rotation about $\mathrm k$
by angle $t$: at $t = 2\pi$ it returns to the identity ---
a loop in $SO(3)$. Its lift satisfies $q(2\pi) = \cos\pi =
-1 = -q(0)$: the lifted \omterm{def:b3:topology:pathconnected}{path} is \emph{not} closed; only at
$t = 4\pi$ does $q$ return to $1$. Interpretation: the loop
of full rotations is not contractible in $SO(3)$ --- its
lift ends at the other sheet of the cover --- while the
double loop is; a body attached to its surroundings by
straps (the belt trick) returns to an untwisted state after
$4\pi$ but not after $2\pi$. Rotation groups remember the
parity of full turns; $S^3$, being \omterm{def:b3:conformal:simplyconnected}{simply connected}, is
where that memory lives.

\textbf{11.} Quarter turns: $q_{\mathrm i} =
\cos\frac\pi4 + \sin\frac\pi4\,\mathrm i$, $q_{\mathrm j} =
\cos\frac\pi4 + \sin\frac\pi4\,\mathrm j$. Product
(applying the $\mathrm j$-turn first):
\[
q_{\mathrm i}q_{\mathrm j} = \tfrac12(1 + \mathrm i)(1 +
\mathrm j) = \tfrac12\bigl(1 + \mathrm i + \mathrm j +
\mathrm k\bigr),
\]
of norm $1$, with $\cos\frac\theta2 = \frac12$: $\theta =
\frac{2\pi}3$, and axis $u = \frac{\mathrm i + \mathrm j +
\mathrm k}{\sqrt3}$ (the pure part normalized). Two
successive quarter-turns about orthogonal axes make a
$120^\circ$ rotation about the cube's main diagonal --- four
real multiplications' worth of bookkeeping, orthogonality
preserved exactly: why flight software and graphics engines
compose rotations through quaternions.

\textbf{12.} From the table, $\mathrm{ji} = -\mathrm k$ and
$\mathrm{ki} = \mathrm j$, so
\[
q\,\mathrm i = a\mathrm i - b + c(\mathrm{ji}) +
d(\mathrm{ki}) = -b + a\mathrm i + d\mathrm j - c\mathrm k .
\]
Multiplying by $\bar q = a - b\mathrm i - c\mathrm j -
d\mathrm k$ with the scalar--vector rule $(t_1 + p_1)(t_2 +
p_2) = t_1t_2 - \langle p_1, p_2\rangle + t_1p_2 + t_2p_1 +
p_1\wedge p_2$, where $p_1 = (a, d, -c)$ and $p_2 = (-b, -c,
-d)$: the scalar part is $-ab - (-ab) = 0$ (pure, as it must
be), and the vector part is
\[
b(b, c, d) + a(a, d, -c) + (-c^2 - d^2,\ bc + ad,\ bd - ac)
= \bigl(a^2 + b^2 - c^2 - d^2,\ 2(bc + ad),\ 2(bd -
ac)\bigr):
\]
the first column of $R_q$. The cyclic map $\sigma(\mathrm i)
= \mathrm j$, $\sigma(\mathrm j) = \mathrm k$,
$\sigma(\mathrm k) = \mathrm i$ preserves the relations
$\mathrm i^2 = \mathrm j^2 = \mathrm k^2 = \mathrm{ijk} =
-1$ (the word $\mathrm{ijk}$ is cyclically invariant up to
the relation $\mathrm{ijk} = \mathrm{jki}$, which holds in
any ring: conjugating $\mathrm{ijk} = -1$ by the invertible
$\mathrm i$), so $\sigma$ extends to an $\R$-algebra
automorphism, and $\sigma(\rho_q(v)) =
\rho_{\sigma(q)}(\sigma(v))$. Unwinding, the image of
$\mathrm j$ is the first-column formula after the
substitution $(b, c, d) \to (c, d, b)$ with the basis
relabeled $\mathrm i \to \mathrm j \to \mathrm k \to \mathrm
i$, which is exactly the second column displayed; one more
turn gives the third.

\textbf{13.} Summing the diagonal, $\operatorname{tr}R_q =
3a^2 - (b^2 + c^2 + d^2) = 4a^2 - 1$ (unit norm), and with
$a = \cos\frac\theta2$: $4\cos^2\frac\theta2 - 1 = 1 +
2\cos\theta$. Antisymmetric part: the three independent
entries of $R_q - R_q^{\mathsf T}$ are $4ab, 4ac, 4ad$ (in
positions $(3,2), (1,3), (2,1)$), so $\frac12(R_q -
R_q^{\mathsf T}) = A_m$ with $m = 2a\,(b, c, d) =
2\cos\frac\theta2\sin\frac\theta2\,u = \sin\theta\,u$.
Algorithm: $\theta = \arccos\frac{\operatorname{tr}R -
1}{2} \in \intcc0\pi$; if $0 < \theta < \pi$, read $u$ off
$\frac{R - R^{\mathsf T}}{2\sin\theta}$ and set $q =
\pm(\cos\frac\theta2 + \sin\frac\theta2 u)$; if $\theta =
0$, $q = \pm1$. For $\theta = \pi$: $a = 0$, and Rodrigues
(question 8) gives $R = I + 2A_u^2 = I + 2(uu^{\mathsf T} -
I) = 2uu^{\mathsf T} - I$, i.e.\ $R + I = 2uu^{\mathsf T}$;
any nonzero column of $R + I$, normalized, is $\pm u$, and
$q = \pm u$.

\textbf{14.} With $a = b = c = d = \frac12$: all diagonal
entries vanish, $2(bc - ad) = 0$, $2(bd + ac) = 1$, $2(bc +
ad) = 1$, $2(cd - ab) = 0$, $2(bd - ac) = 0$, $2(cd + ab) =
1$:
\[
R_q = \begin{pmatrix} 0 & 0 & 1\\ 1 & 0 & 0\\ 0 & 1 & 0
\end{pmatrix},
\]
the cyclic permutation $e_1 \to e_2 \to e_3 \to e_1$. Its
trace is $0 = 1 + 2\cos\theta$, so $\theta = \frac{2\pi}3$,
and it fixes $(1,1,1)$: the rotation by $120^\circ$ about
the main diagonal --- precisely question 11's answer, now
visible as the matrix that cycles the coordinate axes.

\textbf{15.} On the basis one checks $\Phi(\bar q) =
\Phi(q)^*$ (the matrix of $\bar q$ has $\alpha' =
\bar\alpha$, $\beta' = -\beta$, which is the conjugate
transpose of $\bigl(\begin{smallmatrix}\alpha & \beta\\
-\bar\beta & \bar\alpha\end{smallmatrix}\bigr)$). Hence
$\Phi(q)^*\Phi(q) = \Phi(\bar qq) = N(q)I$ and $\det\Phi(q)
= \abs\alpha^2 + \abs\beta^2 = N(q)$: for $q \in S^3$,
$\Phi(q) \in SU(2)$, and $\Phi$ is an injective morphism
(question 1). Surjectivity: let $U =
\bigl(\begin{smallmatrix}\alpha & \beta\\ \gamma &
\delta\end{smallmatrix}\bigr)$ with $\det U = 1$; then
$U^{-1} = \bigl(\begin{smallmatrix}\delta & -\beta\\
-\gamma & \alpha\end{smallmatrix}\bigr)$, and $U^{-1} = U^*
= \bigl(\begin{smallmatrix}\bar\alpha & \bar\gamma\\
\bar\beta & \bar\delta\end{smallmatrix}\bigr)$ forces
$\delta = \bar\alpha$, $\gamma = -\bar\beta$, and then $1 =
\det U = \abs\alpha^2 + \abs\beta^2$: $U = \Phi(q)$ for the
unit quaternion $q$ with coordinates $\alpha = a + \iu b$,
$\beta = c + \iu d$. So $S^3 \cong SU(2)$.

\textbf{16.} Real scalars are central and $N(p) = 1$ gives
$\overline{pq\bar p} = p\bar q\bar p$, so $pq\bar p +
\overline{pq\bar p} = p(q + \bar q)\bar p = q + \bar q$:
the real part is invariant. Conversely let
$\operatorname{Re}q = \operatorname{Re}q' = a$; then the
pure parts have the same norm $\sqrt{1 - a^2} = s$. If $s =
0$, $q = q' = \pm1$. If $s > 0$, write $q = a + su$, $q' =
a + su'$ with $u, u'$ unit pure; question 6 provides a
rotation carrying $u'$ to $u$, i.e.\ $p \in S^3$ with
$\rho_p(u') = u$, and then $pq'\bar p = a + s\rho_p(u') =
q$. The classes of $S^3$ are therefore $\{1\}$, $\{-1\}$,
and for each $a \in \intoo{-1}1$ the $2$-sphere $\{a + su :
u \text{ unit pure}\}$ of radius $s$. Under $\Phi$,
$\operatorname{tr}\Phi(q) = \alpha + \bar\alpha =
2\operatorname{Re}q$: the classes of $SU(2)$ are the level
sets of the trace. Projecting: if $q' = pq\bar p$ then
$\rho_{q'} = \rho_p\rho_q\rho_p^{-1}$; conversely
$\rho_{q'} = \rho_p\rho_q\rho_p^{-1} = \rho_{pq\bar p}$
forces $q' = \pm pq\bar p$ (kernel), so
$\operatorname{Re}q' = \pm\operatorname{Re}q$, i.e.\
$\cos\frac{\theta'}2 = \abs{\operatorname{Re}q'} =
\abs{\operatorname{Re}q} = \cos\frac\theta2$ for the angles
in $\intcc0\pi$: conjugate rotations have equal angles.
Conversely, equal angles allow representatives with the
same nonnegative real part, conjugate by the above: in
$SO(3)$, the \omterm{ex:b3:groups:actions}{conjugacy class} of a rotation is exactly its
angle.

\textbf{17.} $N$ is multiplicative, so $\abs\cdot$ is a
multiplicative norm on $\mathbb H \cong \R^4$ and
$\abs{q^k} = \abs q^k$: the series $\sum q^k/k!$ converges
absolutely in the finite-dimensional (hence \omterm{def:b3:complete:complete}{complete})
space, dominated by $\sum\abs q^k/k! = \eu^{\abs q}$. For a
unit pure $u$: $u^2 = -1$, so $(\theta u)^{2m} =
(-1)^m\theta^{2m}$ and $(\theta u)^{2m+1} =
(-1)^m\theta^{2m+1}u$; splitting the series,
\[
\exp(\theta u) = \sum_m\frac{(-1)^m\theta^{2m}}{(2m)!} +
u\sum_m\frac{(-1)^m\theta^{2m+1}}{(2m+1)!} = \cos\theta +
\sin\theta\,u .
\]
Any $q \in S^3$ is $\cos\alpha + \sin\alpha\,u$ with
$\alpha \in \intcc0\pi$ (question 5): $q = \exp(\alpha u)$,
so $\exp(P) = S^3$. Finally $\exp(su) = \cos s + \sin s\,u
= q(2s)$ in question 9's notation, and $\rho_{q(t)} =
\eu^{tA_u}$ there: $\rho_{\exp(su)} = \eu^{2sA_u}$.

\textbf{18.} Expanding $vw$ coordinatewise with the table:
the products $\mathrm i\cdot\mathrm i = -1$, \dots\ give
the scalar $-(v_1w_1 + v_2w_2 + v_3w_3)$, and the mixed
products ($\mathrm{ij} = \mathrm k$, $\mathrm{ji} =
-\mathrm k$, \dots) give the vector $(v_2w_3 - v_3w_2,\
v_3w_1 - v_1w_3,\ v_1w_2 - v_2w_1)$: $vw = -\langle v,
w\rangle + v\wedge w$. Subtracting the reversed product:
$vw - wv = 2\,v\wedge w$ (the scalar parts cancel, the
cross products add). For the matrices, with $a\wedge(b\wedge
c) = b\langle a, c\rangle - c\langle a, b\rangle$:
\[
[A_v, A_w]x = v\wedge(w\wedge x) - w\wedge(v\wedge x)
= w\langle v, x\rangle - v\langle w, x\rangle
= (v\wedge w)\wedge x = A_{v\wedge w}x .
\]
Derivative: $\overline{\exp(tv)} = \exp(-tv)$ (conjugation
is \omterm{def:b3:topology:continuity}{continuous} and negates pure quaternions), so
\[
\frac{\dd}{\dd t}\Bigr|_{t=0}\exp(tv)\,x\,\exp(-tv) = vx -
xv = 2\,v\wedge x = 2A_vx :
\]
the differential is $v \mapsto 2A_v$, a linear bijection
from $P$ onto the antisymmetric matrices, and $[2A_v, 2A_w]
= 4A_{v\wedge w} = 2A_{2v\wedge w} = 2A_{[v,w]}$ shows it
carries the quaternion \omterm{def:b3:groups:derived}{commutator} to the matrix
\omterm{def:b3:groups:derived}{commutator}.

\textbf{19.} Applying $\rho$: $\rho(\varepsilon(t)) =
\rho(s(R(t)))\,\rho(q(t))^{-1} = R(t)R(t)^{-1} =
\operatorname{id}$, so $\varepsilon(t) \in \ker\rho =
\{\pm1\}$ (question 4). As a product of the \omterm{def:b3:topology:continuity}{continuous maps}
$t \mapsto s(R(t))$ and $t \mapsto q(t)^{-1} =
\bar{q(t)}$, $\varepsilon$ is \omterm{def:b3:topology:continuity}{continuous} on the \omterm{def:b3:topology:connected}{connected}
interval $\intcc0{2\pi}$ with values in the discrete pair
$\{\pm1\}$: it is constant, say $\varepsilon(t) \equiv
\varepsilon$. But $R(0) = R(2\pi) = I$, so $s(R(0)) =
s(R(2\pi))$, while $s(R(0)) = \varepsilon\,q(0) =
\varepsilon$ and $s(R(2\pi)) = \varepsilon\,q(2\pi) =
-\varepsilon$: contradiction. No \omterm{def:b3:topology:continuity}{continuous} section
exists: the sign ambiguity $\pm q$ is global, not a defect
of a particular formula.

\textbf{20.} \emph{Onto:} every $R \in SO(3)$ is
$\eu^{\theta A_u}$ for some unit $u$ and $\theta \in
\intcc0{2\pi}$ (questions 6 and 8); if $\theta > \pi$,
Rodrigues gives $\eu^{\theta A_u} = \eu^{(2\pi -
\theta)A_{-u}}$ (both equal $I + \sin\theta A_u + (1 -
\cos\theta)A_u^2$, and $A_{-u} = -A_u$ with $\sin(2\pi -
\theta) = -\sin\theta$, $\cos(2\pi - \theta) =
\cos\theta$), so $R = E(v)$ with $\norm v \leq \pi$.
\emph{Injective inside:} if $E(v) = E(v') \neq I$ with
$\norm v, \norm{v'} < \pi$, question 13 recovers the same
angle $\theta = \norm v = \norm{v'} \in \intoo0\pi$ from
the trace and, since $\sin\theta \neq 0$, the same axis
from the antisymmetric part: $v = v'$; and $E(v) = I$
forces $\theta \in \{0\}$ on the open ball. \emph{Boundary:}
$E(\pi u) = I + 2A_u^2 = 2uu^{\mathsf T} - I$ depends on
$u$ only through $uu^{\mathsf T}$, whence $E(\pi u) =
E(-\pi u)$; conversely $2uu^{\mathsf T} - I =
2u'u'^{\mathsf T} - I$ applied to $u$ gives $u = \langle
u', u\rangle u'$, so $u' = \pm u$. \omterm{def:b3:topology:interior}{Interior} and boundary
never collide (trace $> -1$ versus $= -1$). So $E$ induces
a \omterm{def:b3:topology:continuity}{continuous} bijection from the ball-with-antipodal-gluing
--- \omterm{def:b3:topology:compact}{compact} --- onto $SO(3)$: a \omterm{def:b3:topology:continuity}{homeomorphism}, and $SO(3)
\cong \mathbb{RP}^3$. A diameter from $\pi u$ to $-\pi u$
has glued endpoints: it is a loop in $SO(3)$, and its
$E$-description matches question 10's family of rotations
about $u$ sweeping a full turn.

\textbf{21.} $\rho$ is a morphism and $\rho_p^{-1} =
\rho_{p^{-1}} = \rho_{\bar p}$, so $\rho_p\rho_q\rho_p^{-1}
= \rho_{pq\bar p}$; by question 16, conjugating a rotation
preserves its angle and rotates its axis by $\rho_p$.
Involutions: $\rho_q^2 = \rho_{q^2} = \operatorname{id}$
iff $q^2 = \pm1$. If $q^2 = 1$ then $(q - 1)(q + 1) = q^2 -
1 = 0$ (central scalars, so this factorization is valid) and
$q = \pm1$ in the division ring $\mathbb H$, giving $\rho_q
= I$, excluded; $q^2 = -1$ with $q = a +
su$ gives $a^2 - s^2 + 2as\,u = -1$, so $a = 0$, $s = 1$:
$q$ is a unit pure $w$, and $\rho_w$ is the half-turn about
$w$ (angle $\pi$, question 5). \omterm{ex:b3:groups:actions}{Center}: if $\rho_q$ commutes
with every $\rho_p$, then $\rho_{pq\bar p} = \rho_q$, so
$pq\bar p = \varepsilon(p)\,q$ with $\varepsilon(p) \in
\{\pm1\}$; $p \mapsto \varepsilon(p) = (pq\bar p)q^{-1}$ is
\omterm{def:b3:topology:continuity}{continuous} on the \omterm{def:b3:topology:connected}{connected} $S^3$ and equals $1$ at $p =
1$, hence $\equiv 1$: $q$ commutes with all of $S^3$, hence
with all of $\mathbb H$ (rescale), so $q \in \R \cap S^3 =
\{\pm1\}$ (question 1) and $\rho_q = I$: the \omterm{ex:b3:groups:actions}{center} is
trivial.

\textbf{22.} Since $u \perp w$ are unit pures, $uw =
u\wedge w$ is pure (question 18), so
\[
w' = qw = \cos\tfrac\theta2\,w + \sin\tfrac\theta2\,u\wedge
w
\]
is pure, of norm $\abs q\abs w = 1$. Then $w'w = qw\cdot w
= qw^2 = -q$, and
\[
\rho_{w'}\rho_w = \rho_{w'w} = \rho_{-q} = \rho_q .
\]
Both axes $w$ and $w' = \cos\frac\theta2 w +
\sin\frac\theta2(u\wedge w)$ lie in the plane $u^\perp$
orthogonal to the rotation axis, and $\langle w', w\rangle
= \cos\frac\theta2$: they make the half-angle
$\frac\theta2$. This is the classical generation: two
half-turns about axes meeting at angle $\frac\theta2$
compose to the rotation of angle $\theta$ about their
common perpendicular.

\textbf{23.} Compactness is question 7. \omterm{def:b3:topology:connected}{Path-connectedness}:
$SO(3) = \rho(S^3)$ is the \omterm{def:b3:topology:continuity}{continuous} image of the
\omterm{def:b3:topology:pathconnected}{path-connected} sphere; alternatively, for $R = \eu^{A}$
with $A$ antisymmetric (question 8), $t \mapsto \eu^{tA}$
is a \omterm{def:b3:topology:pathconnected}{path} in $SO(3)$ from $I$ to $R$ (orthogonal since
$(\eu^{tA})^{\mathsf T} = \eu^{-tA}$, determinant $1$ by
\omterm{def:b3:topology:continuity}{continuity} from $t = 0$). For $O(3)$: $\det$ is \omterm{def:b3:topology:continuity}{continuous}
onto $\{\pm1\}$, so $O(3)$ is disconnected, $O(3) = SO(3)
\sqcup D\,SO(3)$ for any fixed $D$ with $\det D = -1$
(e.g.\ $D = -I$), and left multiplication by $D$ is a
\omterm{def:b3:topology:continuity}{homeomorphism}: exactly two components, each a copy of
$SO(3)$. The two-to-one $\rho$, section-free by question
19, is thus an honest double cover of a \omterm{def:b3:topology:connected}{connected} \omterm{def:b3:topology:compact}{compact}
group by the \omterm{def:b3:conformal:simplyconnected}{simply connected} $S^3$ --- the geometry behind
the belt trick.

\textbf{24.} (i) Matrices:
\[
R_1 = \begin{pmatrix} 0 & -1 & 0\\ 1 & 0 & 0\\ 0 & 0 & 1
\end{pmatrix},
\quad
R_2 = \begin{pmatrix} 1 & 0 & 0\\ 0 & 0 & -1\\ 0 & 1 & 0
\end{pmatrix},
\quad
R_2R_1 = \begin{pmatrix} 0 & -1 & 0\\ 0 & 0 & -1\\ 1 & 0 &
0\end{pmatrix}.
\]
Trace $0 = 1 + 2\cos\theta$ gives $\cos\theta = -\frac12$:
$\theta = \frac{2\pi}3$. Antisymmetric part $\frac{R -
R^{\mathsf T}}2$ has entries encoding $\sin\theta\,(v_3,
-v_2, v_1)$-wise the axis: here $\frac{R - R^{\mathsf T}}2 =
\frac12\bigl(\begin{smallmatrix}0 & -1 & -1\\ 1 & 0 & -1\\ 1
& 1 & 0\end{smallmatrix}\bigr)$, which reads (Part V's
dictionary $A_v$) $v\sin\theta = \frac12(1, -1, 1)$; with
$\sin\frac{2\pi}3 = \frac{\sqrt3}2$: $v = \frac1{\sqrt3}(1,
-1, 1)$.
(ii) Quaternions: $q_1 = \cos\frac\pi4 + \sin\frac\pi4\,k =
\frac{\sqrt2}2(1 + k)$, $q_2 = \frac{\sqrt2}2(1 + i)$, and
\[
q_2q_1 = \tfrac12(1 + i)(1 + k) = \tfrac12(1 + k + i + ik)
= \tfrac12\bigl(1 + i - j + k\bigr)
\]
($ik = -j$). So $\cos\frac\theta2 = \frac12$: $\theta =
\frac{2\pi}3$, and the vector part $\frac12(i - j + k)$ has
direction $\frac1{\sqrt3}(1, -1, 1)$ --- the same answer,
with the quaternion route requiring one line of
multiplication instead of a matrix product: the practical
reason flight software composes attitudes in $S^3$.

\textbf{25.} $I + K$ invertible: $(I + K)v = 0$ gives $0 =
\langle v, v\rangle + \langle v, Kv\rangle = \norm v^2$
(antisymmetry kills the second term): $v = 0$. Orthogonality
of $C = C(K)$: using $(I \pm K)^{\mathsf T} = I \mp K$ and
the fact that all four matrices $I \pm K$, $(I \pm K)^{-1}$
commute (polynomial expressions in $K$, plus limits):
\[
C^{\mathsf T}C = (I + K)^{-\mathsf T}(I - K)^{\mathsf T}
(I - K)(I + K)^{-1}
= (I - K)^{-1}(I + K)(I - K)(I + K)^{-1} = I .
\]
Determinant: $\det(I - K) = \det\bigl((I - K)^{\mathsf
T}\bigr) = \det(I + K)$, so $\det C = 1$: $C \in SO(n)$.
No eigenvalue $-1$: $Cv = -v$ means $(I - K)w = -(I + K)w$
for $w = (I + K)^{-1}v$, i.e.\ $2w = 0$: $v = 0$.
Inversion: from $C(I + K) = I - K$, solve $K(I + C) = I -
C$; since $-1 \notin \operatorname{Sp}C$, $I + C$ is
invertible and $K = (I - C)(I + C)^{-1}$, which is
antisymmetric whenever $C$ is orthogonal without eigenvalue
$-1$ (transpose the expression and use $C^{\mathsf T} =
C^{-1}$: $K^{\mathsf T} = (I - C^{-1})(I + C^{-1})^{-1} =
(C - I)(C + I)^{-1} = -K$). The two maps are mutually
inverse by construction: a global rational parametrization
of the dense open piece of $SO(n)$ avoiding eigenvalue $-1$
--- no series, no trigonometry, and in dimension $3$ it is
the half-angle substitution $K = \tan\frac\theta2\,A_v$ in
disguise.
\end{solution}
