\section*{Chapter \ref{ch:b3:clt} --- Characteristic
Functions and the Central Limit Theorem}

\begin{solution}{exo:b3:clt:1}
Uniform on $\intcc{-1}1$: $\varphi(\xi) =
\frac12\int_{-1}^1\eu^{\iu\xi x}\dd x =
\frac{\sin\xi}{\xi}$ (equal to $1$ at $\xi = 0$).
Exponential $\mathcal E(\lambda)$: $\varphi(\xi) =
\lambda\int_0^\infty\eu^{(\iu\xi - \lambda)x}\dd x =
\frac{\lambda}{\lambda - \iu\xi}$ (the antiderivative
vanishes at $+\infty$ since $\operatorname{Re}(\iu\xi -
\lambda) < 0$). Poisson $\mathcal P(\lambda)$: by the
transfer theorem for discrete laws,
\[
\varphi(\xi) = \sum_{k\geq0}\eu^{\iu\xi
k}\,\eu^{-\lambda}\frac{\lambda^k}{k!}
= \eu^{-\lambda}\exp\bigl(\lambda\eu^{\iu\xi}\bigr)
= \exp\bigl(\lambda(\eu^{\iu\xi} - 1)\bigr).
\]
Binomial $\mathcal B(n, p)$: a sum of $n$ independent
Bernoulli variables, each with cf $1 - p + p\eu^{\iu\xi}$,
so $\varphi(\xi) = \bigl(1 - p + p\eu^{\iu\xi}\bigr)^n$
(\cref{prop:b3:clt:cfbasics}(b)). Poisson additivity: if $X
\sim \mathcal P(\lambda)$, $Y \sim \mathcal P(\mu)$ are
independent,
\[
\varphi_{X+Y}(\xi) = \eu^{\lambda(\eu^{\iu\xi}-1)}
\eu^{\mu(\eu^{\iu\xi}-1)} =
\eu^{(\lambda+\mu)(\eu^{\iu\xi}-1)},
\]
the cf of $\mathcal P(\lambda + \mu)$; injectivity
(\cref{thm:b3:clt:injectivity}) identifies the law.
\end{solution}

\begin{solution}{exo:b3:clt:2}
(a) $\overline{\varphi_X(\xi)} = \E\eu^{-\iu\xi X} =
\varphi_{-X}(\xi)$. So $\varphi_X$ is real if and only if
$\varphi_X = \varphi_{-X}$, if and only if (injectivity,
\cref{thm:b3:clt:injectivity}) $X$ and $-X$ have the same
law.
(b) Write $\varphi_X(\xi_0) = \eu^{\iu\theta}$. Then
\[
\E\bigl[1 - \cos(\xi_0X - \theta)\bigr]
= 1 - \operatorname{Re}\bigl(\eu^{-\iu\theta}
\varphi_X(\xi_0)\bigr) = 1 - 1 = 0 .
\]
The integrand is nonnegative, so $\cos(\xi_0X - \theta) = 1$
almost surely (a nonnegative variable with zero expectation
vanishes a.s.), i.e.\ $\xi_0X - \theta \in 2\pi\Z$ a.s.: $X$
takes its values in the arithmetic progression
$\frac{\theta}{\xi_0} + \frac{2\pi}{\xi_0}\Z$ almost surely.
If $X$ has a density, this countable set is Lebesgue-null,
so it carries probability $0$ --- contradiction; therefore
$\abs{\varphi_X(\xi)} < 1$ for every $\xi \neq 0$.
\end{solution}

\begin{solution}{exo:b3:clt:3}
By independence and \cref{prop:b3:clt:cfbasics}:
\[
\varphi_{X+Y}(\xi) = \eu^{\iu m_1\xi - \sigma_1^2\xi^2/2}\,
\eu^{\iu m_2\xi - \sigma_2^2\xi^2/2}
= \eu^{\iu(m_1+m_2)\xi - (\sigma_1^2+\sigma_2^2)\xi^2/2},
\]
the cf of $\mathcal N(m_1 + m_2, \sigma_1^2 + \sigma_2^2)$;
injectivity concludes. Stability under affine maps is the
affine rule ($aX + b \sim \mathcal N(am_1 + b,
a^2\sigma_1^2)$, allowing the degenerate case $a = 0$), and
stability under independent sums follows by induction on the
computation above. For \emph{dependent} Gaussians the sum
need not be Gaussian: in \cref{exo:b3:clt:9}, $X$ and $Y =
\varepsilon X$ are each standard Gaussian but $X + Y$
vanishes with probability $\frac12$ without being a.s.\
zero, so it is not Gaussian.
\end{solution}

\begin{solution}{exo:b3:clt:4}
(a) \emph{Direct implication.} Let $t$ be a continuity point
of $F_X$ and $\delta > 0$. Take the continuous ramps
$f^-$ ($= 1$ on $\intoc{-\infty}{t-\delta}$, $0$ from $t$
on, affine between) and $f^+$ ($= 1$ on $\intoc{-\infty}t$,
$0$ from $t + \delta$ on, affine between); then $f^- \leq
\mathbf 1_{\intoc{-\infty}t} \leq f^+$, so
\[
\E f^-(X_n) \leq F_{X_n}(t) \leq \E f^+(X_n),
\]
and the outer terms converge to $\E f^\pm(X)$, themselves
squeezed between $F_X(t - \delta)$ and $F_X(t + \delta)$.
Letting $n \to \infty$ then $\delta \to 0$ and using
continuity of $F_X$ at $t$: $F_{X_n}(t) \to F_X(t)$.

\emph{Converse.} Let $f$ be bounded continuous, $M =
\sup\abs f$, $\varepsilon > 0$. The continuity points of
$F_X$ are dense ($F_X$ has at most countably many jumps), so
choose continuity points $a < b$ with $F_X(a) < \varepsilon$
and $1 - F_X(b) < \varepsilon$. On the compact $\intcc ab$
the function $f$ is uniformly continuous: choose continuity
points $a = t_0 < t_1 < \dots < t_m = b$ of $F_X$ with
oscillation of $f$ at most $\varepsilon$ on each
$\intoc{t_{j-1}}{t_j}$, and set $g = \sum_j
f(t_j)\,\mathbf 1_{\intoc{t_{j-1}}{t_j}}$. Then $\abs{f - g}
\leq \varepsilon$ on $\intoc ab$, $\abs g \leq M$, and for
$T = X_n$ or $X$:
\[
\bigl|\E f(T) - \E g(T)\bigr| \leq \varepsilon +
2M\bigl(F_T(a) + 1 - F_T(b)\bigr).
\]
Moreover $\E g(X_n) = \sum_j f(t_j)\bigl(F_{X_n}(t_j) -
F_{X_n}(t_{j-1})\bigr) \to \E g(X)$ (a finite sum of
converging terms, all $t_j$ being continuity points), and
$F_{X_n}(a) \to F_X(a) < \varepsilon$, $1 - F_{X_n}(b) \to 1
- F_X(b) < \varepsilon$. Assembling:
$\limsup_n\abs{\E f(X_n) - \E f(X)} \leq 2\varepsilon +
8M\varepsilon$; let $\varepsilon \to 0$.

(b) The distribution function of the constant $c$ is
$\mathbf 1_{\intco c\infty}$, continuous except at $c$. For
$\varepsilon > 0$, the points $c - \varepsilon$ and $c +
\frac\varepsilon2$ are continuity points, so
\[
\P(\abs{X_n - c} > \varepsilon) \leq F_{X_n}(c -
\varepsilon) + 1 - F_{X_n}\Bigl(c + \frac\varepsilon2\Bigr)
\longrightarrow 0 + 1 - 1 = 0 .
\]
\end{solution}

\begin{solution}{exo:b3:clt:5}
Let $z_n = p_n(\eu^{\iu\xi} - 1)$, so $\varphi_{X_n}(\xi) =
(1 + z_n)^n$ (\cref{exo:b3:clt:1}) and $\abs{z_n} \leq 2p_n
\to 0$ (note $p_n = \frac{np_n}n \to 0$). Both $1 + z_n$ and
$\eu^{z_n}$ have modulus at most $1$: $\abs{1 + z_n} =
\abs{(1 - p_n) + p_n\eu^{\iu\xi}} \leq 1$ by the triangle
inequality, and $\abs{\eu^{z_n}} = \eu^{p_n(\cos\xi - 1)}
\leq 1$. The telescoping inequality $\abs{a^n - b^n} \leq
n\abs{a - b}$ (proof of \cref{thm:b3:clt:clt}) and the power
series bound $\abs{\eu^z - 1 - z} \leq
\abs z^2\eu^{\abs z}$ give
\[
\bigl|(1 + z_n)^n - \eu^{nz_n}\bigr| \leq n\bigl|1 + z_n -
\eu^{z_n}\bigr| \leq n\,\abs{z_n}^2\,\eu^{\abs{z_n}} \leq
4\eu^2\,np_n^2 = 4\eu^2\,(np_n)\,p_n \longrightarrow 0 .
\]
Since $nz_n = np_n(\eu^{\iu\xi} - 1) \to
\lambda(\eu^{\iu\xi} - 1)$, we conclude $\varphi_{X_n}(\xi)
\to \exp\bigl(\lambda(\eu^{\iu\xi} - 1)\bigr)$ for every
$\xi$: the cf of $\mathcal P(\lambda)$, and L\'evy
(\cref{thm:b3:clt:levy}) gives $X_n \Rightarrow \mathcal
P(\lambda)$. Numerically: $\P\bigl(\mathcal B(100, 0.02) =
0\bigr) = 0.98^{100} = \eu^{100\ln 0.98} \approx
\eu^{-2.020} \approx 0.1326$, while $\P\bigl(\mathcal P(2) =
0\bigr) = \eu^{-2} \approx 0.1353$: two percent apart
already at this coarse $n$.
\end{solution}

\begin{solution}{exo:b3:clt:6}
(a) One roll has mean $\frac72$ and variance
$\frac{35}{12}$, so $S$ has mean $3500$, variance
$\frac{35000}{12} \approx 2916.7$ and standard deviation
$\approx 54.0$. By the CLT,
\[
\P(S > 3600) = \P\Bigl(\frac{S - 3500}{54.0} > 1.85\Bigr)
\approx 1 - \Phi(1.85) \approx 0.032 :
\]
about a $3\%$ chance.
(b) $S \sim \mathcal B(100, \frac12)$: mean $50$, standard
deviation $5$. With the continuity correction,
\[
\P(45 \leq S \leq 55) \approx
\Phi\Bigl(\frac{55.5 - 50}{5}\Bigr) -
\Phi\Bigl(\frac{44.5 - 50}{5}\Bigr) = 2\Phi(1.1) - 1
\approx 0.729,
\]
against the exact value $0.7287$; without the correction,
$2\Phi(1) - 1 \approx 0.683$, off by almost five points. The
correction matters because $S$ is a lattice variable: the
atom $\P(S = k)$ is well approximated by the Gaussian mass
of $\intcc{k - \frac12}{k + \frac12}$, and clipping the
interval at the integers $45$ and $55$ discards half an atom
at each end.
\end{solution}

\begin{solution}{exo:b3:clt:7}
The variables $g(U_k)$ are i.i.d.\ (measurable images of
i.i.d.\ variables), square-integrable, with mean $I$
(transfer theorem, \cref{exo:b3:product:9}) and variance
$\sigma^2$. If $\sigma > 0$, \cref{thm:b3:clt:clt} applied
to them is exactly the stated convergence
\[
\sqrt n\,\Bigl(\frac1n\sum_{k\leq n}g(U_k) - I\Bigr) =
\frac{\sum_{k\leq n}\bigl(g(U_k) - I\bigr)}{\sqrt n}
\Longrightarrow \mathcal N(0, \sigma^2)
\]
(if $\sigma = 0$, $g$ is a.s.\ constant and the left-hand
side vanishes identically). Hence
$\P\bigl(\abs{\frac1n\sum g(U_k) - I} \leq
1.96\,\sigma/\sqrt n\bigr) \to 0.95$: the error bar $\pm
1.96\,\sigma/\sqrt n$ sees the dimension $d$ only through
the constant $\sigma$, never through the rate in $n$. The
midpoint rule with $n$ nodes in dimension $d$ has mesh
$n^{-1/d}$ and error of order $n^{-2/d}$ for $\mathcal C^2$
integrands. Monte Carlo's $n^{-1/2}$ decays faster than
$n^{-2/d}$ exactly when $\frac12 > \frac2d$, i.e.\ $d > 4$:
from dimension $5$ on, random sampling asymptotically beats
the grid --- the curse of dimensionality spares
probabilistic methods, which is why Monte Carlo rules
high-dimensional integration.
\end{solution}

\begin{solution}{exo:b3:clt:8}
\emph{Sum.} For fixed $\xi$:
\[
\bigl|\E\eu^{\iu\xi(X_n+Y_n)} -
\eu^{\iu\xi c}\,\E\eu^{\iu\xi X_n}\bigr|
= \bigl|\E\bigl[\eu^{\iu\xi X_n}\bigl(\eu^{\iu\xi Y_n} -
\eu^{\iu\xi c}\bigr)\bigr]\bigr|
\leq \E\bigl|\eu^{\iu\xi(Y_n - c)} - 1\bigr| .
\]
Split on the event $\{\abs{Y_n - c} \leq \delta\}$: there,
$\abs{\eu^{\iu\xi(Y_n-c)} - 1} \leq \abs\xi\,\delta$ (the
chord is shorter than the arc); the complement contributes
at most $2\,\P(\abs{Y_n - c} > \delta) \to 0$. Hence the
$\limsup$ is $\leq \abs\xi\,\delta$ for every $\delta > 0$:
the difference tends to $0$. Since $\E\eu^{\iu\xi X_n} \to
\varphi_X(\xi)$, we get $\varphi_{X_n+Y_n}(\xi) \to
\eu^{\iu\xi c}\varphi_X(\xi) = \varphi_{X+c}(\xi)$, and
L\'evy (\cref{thm:b3:clt:levy}) yields $X_n + Y_n
\Rightarrow X + c$.

\emph{Product.} First, $cX_n \Rightarrow cX$:
$\varphi_{cX_n}(\xi) = \varphi_{X_n}(c\xi) \to
\varphi_X(c\xi) = \varphi_{cX}(\xi)$. Next, $(Y_n - c)X_n
\to 0$ in probability: the laws of the $X_n$ are tight (their
cfs converge to a cf; see the tightness step of
\cref{thm:b3:clt:levy}), so given $\varepsilon > 0$ pick $M$
with $\P(\abs{X_n} > M) \leq \varepsilon$ for all $n$; then
\[
\P\bigl(\abs{(Y_n - c)X_n} > \varepsilon\bigr) \leq
\P(\abs{X_n} > M) + \P\Bigl(\abs{Y_n - c} >
\frac{\varepsilon}{M}\Bigr) \leq \varepsilon + o(1) .
\]
Writing $Y_nX_n = cX_n + (Y_n - c)X_n$ and applying the sum
part (whose proof only used $Y_n' := (Y_n - c)X_n \to 0$ in
probability, with constant $0$): $Y_nX_n \Rightarrow cX$.

\emph{Application.} By the strong law of large numbers
(\cref{thm:b3:probability:slln}), $\hat p_n \to p$ a.s., so
by continuity $\hat\sigma_n = \sqrt{\hat p_n(1 - \hat p_n)}
\to \sigma = \sqrt{p(1 - p)} > 0$ a.s., hence
$\frac{\sigma}{\hat\sigma_n} \to 1$ in probability. Slutsky's
product rule upgrades $\frac{S_n - np}{\sigma\sqrt n}
\Rightarrow \mathcal N(0,1)$ to $\frac{S_n -
np}{\hat\sigma_n\sqrt n} = \frac{\sigma}{\hat\sigma_n}\cdot
\frac{S_n - np}{\sigma\sqrt n} \Rightarrow \mathcal N(0,1)$:
the \emph{usable} confidence interval $\hat p_n \pm
1.96\,\hat\sigma_n/\sqrt n$, built from the data alone,
keeps its asymptotic $95\%$ level.
\end{solution}

\begin{solution}{exo:b3:clt:9}
(a) Splitting the expectation over the two values of
$\varepsilon$ (independence): for Borel $B$, $\P(Y \in B) =
\frac12\P(X \in B) + \frac12\P(-X \in B) = \P(X \in B)$,
since $-X \sim X$ ($\mathcal N(0,1)$ is symmetric): $Y \sim
\mathcal N(0,1)$. And $\operatorname{Cov}(X, Y) =
\E[\varepsilon X^2] = \E[\varepsilon]\,\E[X^2] = 0 \cdot 1 =
0$.
(b) $\abs Y = \abs X$, so $\P(\abs X \leq 1,\ \abs Y \geq 2)
= 0$ while $\P(\abs X \leq 1)\,\P(\abs Y \geq 2) > 0$: not
independent. If $(X, Y)$ were a Gaussian vector, $X + Y =
(1 + \varepsilon)X$ would be a real Gaussian variable
(\cref{def:b3:clt:gaussianvector} with $t = (1,1)$); but
$\P(X + Y = 0) = \P(\varepsilon = -1) = \frac12$, whereas a
Gaussian variable has an atom only if it is a.s.\ constant
--- and $X + Y$ equals $2X \neq 0$ a.s.\ on
$\{\varepsilon = 1\}$. Contradiction: $(X, Y)$ is not
Gaussian.
(c) Each marginal is Gaussian and the covariance vanishes,
yet independence fails --- because the \emph{pair} is not
jointly Gaussian. \cref{thm:b3:clt:gaussianvector}(2) cannot
be weakened to ``Gaussian marginals''.
\end{solution}

\begin{solution}{exo:b3:clt:10}
(a) \cref{exo:b3:fouriertransform:1} computes
$\widehat{\eu^{-\abs\cdot}}(\xi) = \frac{2}{1 + \xi^2}$;
both sides being integrable, Fourier inversion
(\cref{thm:b3:fouriertransform:inversion}) turns this
around:
\[
\int_\R\eu^{\iu\xi x}\,\frac{\dd x}{\pi(1 + x^2)} =
\eu^{-\abs\xi},
\]
which is exactly $\varphi_X(\xi)$ for a Cauchy variable $X$.
(b) By independence, $\varphi_{S_n}(\xi) =
\bigl(\eu^{-\abs\xi}\bigr)^n = \eu^{-n\abs\xi}$, so
$\varphi_{S_n/n}(\xi) = \varphi_{S_n}(\xi/n) =
\eu^{-\abs\xi}$: the empirical mean $\frac{S_n}n$ is again
standard Cauchy for every $n$ (injectivity). The average
never concentrates: its fluctuations at time $10^6$ are
those of a single observation.
(c) $\E\abs{X_1} = \frac2\pi\int_0^\infty\frac{x\,\dd x}{1 +
x^2} = +\infty$: the Cauchy law is not integrable, so the
strong law of large numbers
(\cref{thm:b3:probability:slln}) does not apply, and the CLT
(which needs a finite variance) even less. Here their
conclusions genuinely fail, not merely their proofs.
Consistency check: $\varphi(\xi) = \eu^{-\abs\xi}$ is not
differentiable at $0$, as \cref{prop:b3:clt:cfbasics}(c)
read contrapositively predicts for a non-integrable
variable.
\end{solution}

\begin{solution}{pb:b3:clt:1}
\textbf{1.} The family $(X_1, \dots, X_n, N_1, \dots, N_n)$
is independent: the two blocks are independent of each
other by construction and each block is i.i.d. $W_i$ is a
measurable function of the variables $(X_j)_{j<i}$ and
$(N_j)_{j>i}$ only, all distinct from $X_i$ and $N_i$: by
the coalition principle
(\cref{thm:b3:probability:independence}), $W_i$ is
independent of the pair $(X_i, N_i)$. The decompositions
$H_i = W_i + \frac{X_i}{\sqrt n}$ and $H_{i-1} = W_i +
\frac{N_i}{\sqrt n}$ are immediate from the definitions:
passing from $H_i$ to $H_{i-1}$ swaps the single summand
$X_i$ for $N_i$.

\textbf{2.} Taylor--Lagrange at order $3$: there is $c$
between $w$ and $w + h$ with $f(w + h) = f(w) + f'(w)h +
\frac12f''(w)h^2 + \frac16f'''(c)h^3$, and $\abs{f'''(c)}
\leq M_3$ gives the bound.

\textbf{3.} Subtracting the two expansions at the common
base point $w = W_i$:
\[
f(H_i) - f(H_{i-1}) = f'(W_i)\,\frac{X_i - N_i}{\sqrt n} +
\frac{f''(W_i)}{2}\,\frac{X_i^2 - N_i^2}{n} + R_i,
\qquad
\abs{R_i} \leq \frac{M_3}{6}\cdot
\frac{\abs{X_i}^3 + \abs{N_i}^3}{n^{3/2}} .
\]
Take expectations. By question 1, $f'(W_i)$ and $f''(W_i)$
are independent of $(X_i, N_i)$, so the mixed expectations
factor:
\begin{align*}
\E\Bigl[f'(W_i)\,\frac{X_i - N_i}{\sqrt n}\Bigr] &=
\E\bigl[f'(W_i)\bigr]\,\frac{\E X_i - \E N_i}{\sqrt n} = 0,
\\
\E\Bigl[f''(W_i)\,\frac{X_i^2 - N_i^2}{n}\Bigr] &=
\E\bigl[f''(W_i)\bigr]\,\frac{1 - 1}{n} = 0 :
\end{align*}
the first two moments of $X_i$ and $N_i$ \emph{match}, and
only the remainder survives:
\[
\bigl|\E f(H_i) - \E f(H_{i-1})\bigr| \leq \E\abs{R_i} \leq
\frac{M_3}{6}\cdot\frac{\beta + \gamma}{n^{3/2}} .
\]
The Gaussian third moment: $\gamma = \E\abs{N_1}^3 =
2\int_0^\infty x^3\,\frac{\eu^{-x^2/2}}{\sqrt{2\pi}}\,\dd x
= \frac{2}{\sqrt{2\pi}}\int_0^\infty 2u\,\eu^{-u}\dd u =
\frac{4}{\sqrt{2\pi}} = \frac{2\sqrt2}{\sqrt\pi}$
(substitution $u = x^2/2$, then $\Gamma(2) = 1$).

\textbf{4.} Telescoping $\E f(T_n) - \E f(G_n) =
\sum_{i=1}^n\bigl(\E f(H_i) - \E f(H_{i-1})\bigr)$ and
applying question 3 to each of the $n$ terms:
\[
\bigl|\E f(T_n) - \E f(G_n)\bigr| \leq
n \cdot \frac{M_3(\beta + \gamma)}{6\,n^{3/2}} =
\frac{M_3\,(\beta + \gamma)}{6\,\sqrt n} .
\]

\textbf{5.} $G_n$ is $\mathcal N(0,1)$ \emph{exactly} for
every $n$ (a normalized sum of independent standard
Gaussians, \cref{exo:b3:clt:3}), so $\E f(G_n) = \E f(N)$
and question 4 reads $\abs{\E f(T_n) - \E f(N)} \leq
\frac{M_3(\beta+\gamma)}{6\sqrt n} \to 0$ for $f \in
\mathcal C^3_b$. \emph{Upgrade.} Let $f$ be bounded
continuous, $M = \sup\abs f$, $\varepsilon > 0$. Choose $A
\geq 1$ with $\frac1{A^2} \leq \varepsilon$: Chebyshev with
$\V(T_n) = 1$ gives $\P(\abs{T_n} > A) \leq \varepsilon$
for all $n$, and likewise $\P(\abs N > A) \leq \varepsilon$.
Let $\chi$ be $\mathcal C^\infty$ with
$\mathbf 1_{\intcc{-A}A} \leq \chi \leq
\mathbf 1_{\intcc{-A-1}{A+1}}$ (a smooth plateau, built by
mollifying $\mathbf 1_{\intcc{-A-\frac12}{A+\frac12}}$,
\cref{thm:b3:lp:regularization}); $g = f\chi$ is continuous
with compact support, hence uniformly continuous, so its
mollification $g_\eta = g * \rho_\eta$ is $\mathcal
C^\infty$ with bounded derivatives of all orders and
$\norm{g - g_\eta}_\infty \leq \varepsilon$ for $\eta$
small enough. For $T = T_n$ or $N$, since $f = g$ on
$\intcc{-A}A$ and $\abs{f - g} \leq 2M$ everywhere:
\[
\bigl|\E f(T) - \E g_\eta(T)\bigr| \leq
\E\abs{(f - g)(T)} + \norm{g - g_\eta}_\infty
\leq 2M\,\P(\abs T > A) + \varepsilon \leq
(2M + 1)\,\varepsilon .
\]
Combining with $\E g_\eta(T_n) \to \E g_\eta(N)$ (question
4 applies: $g_\eta \in \mathcal C^3_b$):
\[
\limsup_n\;\bigl|\E f(T_n) - \E f(N)\bigr| \leq
2(2M + 1)\,\varepsilon ,
\]
and $\varepsilon$ was arbitrary: $\E f(T_n) \to \E f(N)$ for
every bounded continuous $f$, i.e.\ $T_n \Rightarrow
\mathcal N(0,1)$.

\textbf{6.} Identical distribution entered only through one
sentence: ``$X_i$ and $N_i$ have the same first two
moments''. So let $X_1, \dots, X_n$ be independent,
centered, with variances $\sigma_i^2$ and finite third
moments, $s_n^2 = \sum_i\sigma_i^2 > 0$, and take $N_i \sim
\mathcal N(0, \sigma_i^2)$ independent of everything.
Define the hybrids with normalization $s_n$: $H_i =
\frac1{s_n}(\sum_{j\leq i}X_j + \sum_{j>i}N_j)$. In the
$i$-th swap, $\E X_i = \E N_i = 0$ and $\E X_i^2 = \E N_i^2
= \sigma_i^2$ again kill the $f'$ and $f''$ terms, and the
remainder gives (using $\E\abs{N_i}^3 = \sigma_i^3\gamma$ by
scaling):
\[
\bigl|\E f(H_i) - \E f(H_{i-1})\bigr| \leq
\frac{M_3}{6\,s_n^3}\bigl(\E\abs{X_i}^3 +
\sigma_i^3\gamma\bigr) .
\]
Telescoping:
\[
\Bigl|\E f\Bigl(\frac{X_1 + \dots + X_n}{s_n}\Bigr) -
\E f(N)\Bigr| \leq \frac{M_3}{6\,s_n^3}\sum_{i=1}^n
\Bigl(\E\abs{X_i}^3 + \V(X_i)^{3/2}\,\gamma\Bigr) .
\]
Since $\sigma_i^3 = (\E X_i^2)^{3/2} \leq \E\abs{X_i}^3$
(the power--mean inequality, i.e.\ Jensen for $t \mapsto
t^{3/2}$ applied to $X_i^2$), the right side is at most
$\frac{M_3(1 + \gamma)}{6}\cdot
\frac{\sum_i\E\abs{X_i}^3}{s_n^3}$: under \emph{Lyapunov's
condition} $\frac1{s_n^3}\sum_i\E\abs{X_i}^3 \to 0$, the
normalized sums converge in law to $\mathcal N(0,1)$ ---
the CLT without identical distribution.

\textbf{7.} Let $\rho \in \mathcal C^\infty_c(\intoo01)$
with $\int\rho = 1$ and set $\psi(x) = \int_x^1\rho(s)\dd
s$: $\psi$ is $\mathcal C^\infty$, nonincreasing, $\psi = 1$
on $\R_-$, $\psi = 0$ on $\intco1\infty$; let $K =
\norm{\psi'''}_\infty$. For $t \in \R$ and $\delta > 0$
define $\psi_\delta(x) = \psi\bigl(\frac{x - t}\delta\bigr)$
and $\tilde\psi_\delta(x) = \psi\bigl(\frac{x - t}\delta +
1\bigr)$: these are $\mathcal C^3_b$ with third derivative
bounded by $K/\delta^3$, and
\[
\mathbf 1_{\intoc{-\infty}{t-\delta}} \leq
\tilde\psi_\delta \leq \mathbf 1_{\intoc{-\infty}t} \leq
\psi_\delta \leq \mathbf 1_{\intoc{-\infty}{t+\delta}} .
\]
Upper bound: by question 4 applied to $\psi_\delta$ (with
$M_3 = K/\delta^3$),
\[
\P(T_n \leq t) \leq \E\psi_\delta(T_n) \leq
\E\psi_\delta(N) + \frac{K(\beta +
\gamma)}{6\,\delta^3\sqrt n}
\leq \Phi(t + \delta) + \frac{K(\beta +
\gamma)}{6\,\delta^3\sqrt n}
\leq \Phi(t) + \frac{\delta}{\sqrt{2\pi}} +
\frac{K(\beta + \gamma)}{6\,\delta^3\sqrt n},
\]
because $\Phi$ is Lipschitz with constant
$\frac1{\sqrt{2\pi}}$ (its density is bounded by
$\frac1{\sqrt{2\pi}}$). The symmetric lower bound via
$\tilde\psi_\delta$ gives the two-term estimate
\[
\sup_{t\in\R}\,\bigl|\P(T_n \leq t) - \Phi(t)\bigr| \leq
\frac{K(\beta + \gamma)}{6}\cdot\frac{1}{\delta^3\sqrt n} +
\frac{\delta}{\sqrt{2\pi}}
\qquad(\delta > 0\ \text{arbitrary}).
\]
The two terms balance when $\delta^{-3}n^{-1/2} \asymp
\delta$, i.e.\ $\delta = n^{-1/8}$: both are then
$O(n^{-1/8})$, an explicit uniform rate valid for every
$n$. (The optimal Berry--Esseen rate $C\beta/\sqrt n$
requires the Fourier smoothing inequality; swapping trades
sharpness for complete elementarity.)

\textbf{8.} For $X_i = 2B_i - 1$ (fair signs): centered,
variance $1$, and $\abs{X_i} = 1$ so $\beta = 1$. Question 7
then bounds $\sup_t\abs{\P(\frac{S_n}{\sqrt n} \leq t) -
\Phi(t)}$ explicitly and uniformly for \emph{every} finite
$n$ --- a global, non-asymptotic statement about the
distribution function. The local estimate of
\cref{pb:b3:product:1}, question 7, gives instead the exact
asymptotics of an individual atom, $\P(S_{2n} = 2k) \sim
\frac{\eu^{-k^2/n}}{\sqrt{\pi n}}$: it resolves
probabilities of size $n^{-1/2}$, far below question 7's
$n^{-1/8}$ resolution, but it is pointwise, asymptotic
(no explicit error at fixed $n$) and tied to this
particular lattice law. Local precision versus global
uniformity: the two methods are complementary, and summing
the local estimate over $k \in \intint{a\sqrt n}{b\sqrt n}$
recovers de Moivre--Laplace on intervals --- with a sharper
rate, but only for this law.

\textbf{9.} The swapping argument used nothing about the law
of the $X_i$ beyond $\E X_i = 0$, $\E X_i^2 = 1$ and the
finiteness of $\E\abs{X_i}^3$: had we replaced the Gaussians
$N_i$ by any other i.i.d.\ family with the same first two
moments and finite third moment, the same telescoping would
bound $\abs{\E f(\text{sum}_X) - \E f(\text{sum}_Y)}$ by
$O(n^{-1/2})$ for every smooth $f$. Smooth statistics of
large independent sums are therefore \emph{universal}: up to
a quantified error, they depend on the summands' law only
through two numbers. This is the invariance principle:
prove a limit theorem for the most computable law (the
Gaussian, where everything is exact), then transfer it to
all laws by swapping. The same scheme --- with sums replaced
by more elaborate functionals --- drives Wigner's semicircle
law for random matrices, the universality of roots of random
polynomials, and much of modern probability; the central
limit theorem is its first and simplest instance.
\end{solution}
