\section*{Chapter \ref{ch:b3:probability} --- Probability:
Foundations and the \omterm{def:b3:probability:space}{Law} of Large Numbers}

\begin{solution}{exo:b3:probability:1}
(a) For $u \in \intoo01$: $\P(F(X) \leq u) = \P(X \leq
F^{-1}(u)) = F(F^{-1}(u)) = u$ (\omterm{def:b3:topology:continuity}{continuity} and strict
monotonicity make $F$ a bijection onto $\intoo01$ with
$\{F(X) \leq u\} = \{X \leq F^{-1}(u)\}$): $F(X)$ is uniform.
Conversely $\P(G(U) \leq t) = \P(U \leq F(t)) = F(t)$: to
simulate a \omterm{def:b3:probability:space}{law}, apply the inverse distribution function to a
uniform sample.

(b) $Y = X^2$, $X$ uniform on $\intcc{-1}1$: for $t \in
\intcc01$, $F_Y(t) = \P(-\sqrt t \leq X \leq \sqrt t) = \sqrt
t$: \omterm{ex:b3:lebesgue:gamma}{density} $\frac1{2\sqrt t}\mathbf 1_{\intoo01}$. And
$\P\bigl(-\frac1\lambda\ln U \leq t\bigr) = \P(U \geq
\eu^{-\lambda t}) = 1 - \eu^{-\lambda t}$: the exponential
$\mathcal E(\lambda)$ --- inversion in action.
\end{solution}

\begin{solution}{exo:b3:probability:2}
(a) Poisson: $\E X = \sum_{k\geq0}k\,\eu^{-\lambda}
\frac{\lambda^k}{k!} = \lambda$, $\E[X(X-1)] = \lambda^2$, so
$\V = \lambda^2 + \lambda - \lambda^2 = \lambda$. Geometric
($\P(X = k) = p(1-p)^{k-1}$): $\E X = \frac1p$, $\V =
\frac{1-p}{p^2}$ (differentiate the geometric series twice).

(b) $G(t) = \P(T > t)$ is nonincreasing with $G(0^+)\dots$
$G \colon \intco0\infty \to \intoc01$; memorylessness reads
$G(t + s) = G(t)G(s)$. Then $G(n t) = G(t)^n$ and $G(t/n) =
G(t)^{1/n}$: $G(q) = G(1)^q$ for rational $q \geq 0$; writing
$G(1) = \eu^{-\lambda}$ ($\in \intoo01$: $G(1) = 1$ would
force $G \equiv 1$, impossible for a finite \omterm{def:b3:probability:space}{random variable};
$G(1) = 0$ is excluded by hypothesis) and squeezing an
arbitrary $t$ between rationals (monotonicity): $G(t) =
\eu^{-\lambda t}$ --- the exponential \omterm{def:b3:probability:space}{law}. The converse is a
computation.
\end{solution}

\begin{solution}{exo:b3:probability:3}
(a) $\sigma(f_i(X_i)) = f_i(X_i)^{-1}(\mathcal B) \subseteq
X_i^{-1}(\mathcal B) = \sigma(X_i)$ ($f_i$ Borel), and
sub-$\sigma$-algebras of \omterm{def:b3:probability:independence}{independent} $\sigma$-algebras are
\omterm{def:b3:probability:independence}{independent} (the defining identity holds a fortiori).

(b) $\sigma(A_i) = \{\varnothing, A_i, A_i^c, \Omega\} =
\sigma(A_i^c) = \sigma(\mathbf 1_{A_i})$: all three
statements assert \omterm{def:b3:probability:independence}{independence} of the same
$\sigma$-algebras. (That factorization over the $A_i$
propagates to complements is the $\lambda$-system argument
inside \cref{def:b3:probability:independence}'s equivalence
--- or direct inclusion-exclusion.)

(c) $\P(A) = \P(B) = \P(C) = \frac12$; $A\cap B = A\cap C =
B\cap C$ on pairs: each intersection is ``both heads'' or
analogous, of probability $\frac14$: pairwise \omterm{def:b3:probability:independence}{independent}.
But $\P(A\cap B\cap C) = \P(\text{HH}) = \frac14 \neq
\frac18$: not \omterm{def:b3:probability:independence}{independent} --- $C$ is determined by $A$ and
$B$.
\end{solution}

\begin{solution}{exo:b3:probability:4}
(a) Let the text $T$ have length $L$ and $q = a^{-L}$ ($a$
the alphabet size). The events $E_k = \{$positions $kL+1,
\dots, (k+1)L$ spell $T\}$ are \omterm{def:b3:probability:independence}{independent} (disjoint blocks
of i.i.d.\ letters), each of probability $q > 0$: $\sum\P(E_k)
= \infty$, and Borel--Cantelli (2) gives infinitely many
occurrences a.s.

(b) Upper: $\P\bigl(R_n \geq (1+\varepsilon)\log_2n\bigr)
\leq 2^{-(1+\varepsilon)\log_2n} = n^{-(1+\varepsilon)}$,
summable: by Borel--Cantelli (1), a.s.\ only finitely many
such $n$. Lower: pack disjoint blocks --- the $j$-th of
length $\ell_j = \lceil\log_2s_j\rceil$ starting at $s_j =
\sum_{i<j}\ell_i$; the events ``block $j$ is all ones'' are
\omterm{def:b3:probability:independence}{independent} with probability $2^{-\ell_j} \asymp
\frac1{s_j} \asymp \frac1{j\log_2 j}$, whose sum diverges:
Borel--Cantelli (2) gives infinitely many all-ones blocks,
i.e.\ $R_{s_j} \geq \log_2 s_j$ infinitely often. Together:
the maximal run length in the first $n$ digits is
$(1 + o(1))\log_2n$ a.s.
\end{solution}

\begin{solution}{exo:b3:probability:5}
(a) $R = \bigl(\limsup\abs{X_n}^{1/n}\bigr)^{-1}$ is
unchanged if finitely many $X_n$ are modified: for every $N$,
$R$ is $\sigma(X_N, X_{N+1}, \dots)$-measurable, i.e.\
\omterm{def:b3:lebesgue:measurable}{tail-measurable}. Then each event $\{R \leq c\}$ has
probability $0$ or $1$ (\cref{thm:b3:probability:zeroone}),
so the distribution function of $R$ takes only the values
$0, 1$: it jumps at a single point $c_0 \in
\intcc0{+\infty}$, and $R = c_0$ a.s.

(b) Convergence of $\sum X_n$ and of $\frac{S_n}n$ are
insensitive to changing finitely many terms (for the second:
the modified terms contribute $O(1/n) \to 0$): tail events;
\omterm{thm:b3:probability:zeroone}{zero--one law}.

(c) $\{X_1 > 0\}$ depends on $X_1$: for i.i.d.\ signs
($\P(X_1 = \pm1) = \frac12$), its probability is $\frac12
\notin \{0,1\}$ --- no contradiction, it is not a tail
event.
\end{solution}

\begin{solution}{exo:b3:probability:6}
Work on $(\intcc01, \lambda)$.
(a) The typewriter $\mathbf 1_{I_n}$
(\cref{exo:b3:lp:3}): $\norm{X_n}_p^p = \lambda(I_n) \to 0$
(all $p < \infty$), hence also in probability; at every
$\omega$ the values $0$ and $1$ both recur: no pointwise
convergence anywhere.
(b) $X_n = n\mathbf 1_{\intoo0{1/n}} \to 0$ off $0$, but
$\norm{X_n}_p \geq n^{1 - 1/p} \geq 1$.
(c) $X_n = \sqrt n\,\mathbf 1_{\intoo0{1/n}}$: $\E\abs{X_n} =
n^{-1/2} \to 0$, $\E X_n^2 = 1$.
(d) From any subsequence extract (convergence in
probability) a further subsequence converging a.s.\
(\cref{prop:b3:probability:modes}(c)); dominated convergence
gives $L^1$ convergence along it, with the \emph{same} limit
$X$. Thus every subsequence of the numerical sequence
$\E\abs{X_n - X}$ has a subsubsequence tending to $0$: the
whole sequence tends to $0$.
\end{solution}

\begin{solution}{exo:b3:probability:7}
(a) $\hat p_n = \frac{S_n}n$ with $S_n$ binomial: $\V(\hat
p_n) = \frac{p(1-p)}n \leq \frac1{4n}$, and Chebyshev
(\cref{prop:b3:probability:markov}) gives the bound.
(b) Solve $\frac1{4n(0.03)^2} \leq 0.05$: $n \geq
\frac{1}{4\cdot0.0009\cdot0.05} \approx 5556$. The central
limit theorem will justify $n \approx 1070$ for the same
guarantee: Chebyshev pays for its generality with a factor
$\approx 5$.
\end{solution}

\begin{solution}{exo:b3:probability:8}
(a) If $S_n \sim \mathcal B(n, x)$, the transfer theorem
gives $\E\bigl[f(\frac{S_n}n)\bigr] =
\sum_k\binom nkx^k(1-x)^{n-k}f(\frac kn) = B_nf(x)$.

(b)--(c) Let $\omega = \omega_f$ be the modulus of \omterm{def:b3:topology:continuity}{continuity}
($\abs{f(u) - f(v)} \leq \omega(\abs{u - v})$, and $\omega(c
\delta) \leq (1 + c)\,\omega(\delta)$ by chaining steps).
Then, for any $\delta > 0$,
\[
\abs{f(u) - f(x)} \leq \Bigl(1 + \frac{(u -
x)^2}{\delta^2}\Bigr)\omega(\delta)
\]
(if $\abs{u - x} \leq \delta$, clear; otherwise
$\omega(\abs{u-x}) \leq (1 + \frac{\abs{u-x}}\delta)
\omega(\delta) \leq (1 + \frac{(u-x)^2}{\delta^2})
\omega(\delta)$). Take \omterm{def:b3:probability:space}{expectations} at $u = \frac{S_n}n$:
\[
\abs{B_nf(x) - f(x)} \leq
\Bigl(1 + \frac{\V(S_n/n)}{\delta^2}\Bigr)\omega(\delta)
\leq \Bigl(1 + \frac{1}{4n\delta^2}\Bigr)\omega(\delta) ;
\]
with $\delta = n^{-1/2}$: $\norm{B_nf - f}_\infty \leq
\frac54\,\omega\bigl(n^{-1/2}\bigr) \leq
\frac32\,\omega\bigl(n^{-1/2}\bigr) \to 0$ (uniform
\omterm{def:b3:topology:continuity}{continuity} on the \omterm{def:b3:topology:compact}{compact}): a probabilistic Weierstrass
theorem, with an explicit and uniform rate.
\end{solution}

\begin{solution}{exo:b3:probability:9}
(a) After $k - 1$ types are collected, each draw is new with
probability $p_k = \frac{n-k+1}n$: $\tau_k$ is geometric
$(p_k)$, and the $\tau_k$ are \omterm{def:b3:probability:independence}{independent} (the draws are).
Sums: $\E T_n = \sum_k\frac n{n-k+1} = nH_n \sim n\ln n$;
$\V(T_n) = \sum\frac{1 - p_k}{p_k^2} \leq
n^2\sum_{j=1}^n\frac1{j^2} \leq \frac{\pi^2}6n^2$.

(b) Chebyshev: $\P\bigl(\abs{T_n - nH_n} \geq \varepsilon
n\ln n\bigr) \leq \frac{\pi^2n^2/6}{\varepsilon^2n^2\ln^2n}
\to 0$, and $\frac{nH_n}{n\ln n} \to 1$: $\frac{T_n}{n\ln n}
\to 1$ in probability.

(c) Union bound: $T_n > t$ means some type is unseen after
$\lceil t\rceil$ draws, so $\P(T_n > t) \leq n(1 -
\frac1n)^{t} \leq n\,\eu^{-t/n}$; at $t = \beta n\ln n$:
$\leq n^{1 - \beta}$. For $\beta > 1$, $\sum_m
2^{m(1-\beta)} < \infty$: Borel--Cantelli gives, along $n =
2^m$, a.s.\ $T_n \leq \beta n\ln n$ eventually --- in
particular for every $\beta > 2$ as stated (any $\beta > 1$
works along the subsequence).
\end{solution}

\begin{solution}{exo:b3:probability:10}
(a) The even-indexed digits $(b_{2k})_k$ are i.i.d.\ fair
bits (a subfamily of the \omterm{def:b3:probability:independence}{independent} digit family), so $U =
\sum_kb_{2k}2^{-k}$ gives every dyadic interval its correct
probability (as in
\cref{thm:b3:probability:existence}): uniform; likewise $V$;
and $(U, V)$ depend on disjoint digit blocks: \omterm{def:b3:probability:independence}{independent}
(factorization on dyadic rectangles, then Dynkin).

(b) $\Phi(\omega) = (U(\omega), V(\omega))$ is \omterm{def:b3:lebesgue:measurable}{measurable}
with $\Phi_*\lambda = \lambda\otimes\lambda = \lambda_2$
(agreement on dyadic rectangles + uniqueness). Interleaving
digits defines an inverse defined off the (null) set of
dyadic rationals in either factor: a measure-preserving
bijection between \omterm{def:b3:measure:measure}{full-measure} subsets of $\intcc01$ and
$\intcc01^2$. Contrast with Peano
(\cref{pb:b3:topology:1}): \omterm{def:b3:topology:continuity}{continuity} forced surjectivity
without injectivity; dropping \omterm{def:b3:topology:continuity}{continuity} for mere
\omterm{def:b3:lebesgue:measurable}{measurability} buys a measure-isomorphism --- dimension is
invisible to \omterm{def:b3:measure:measure}{measure} theory, visible to \omterm{def:b3:topology:topology}{topology}.
\end{solution}

\begin{solution}{exo:b3:probability:11}
(a) \omterm{def:b3:topology:continuity}{Continuity} of the distribution makes ties null events
(as in the chapter's order-statistics arguments), and the
$n!$ relative orderings of $(X_1, \dots, X_n)$ are
exchangeable, hence equally likely. $R_n = 1$ means the
maximum sits in the last position: probability
$\frac{(n-1)!}{n!} = \frac1n$.

(b) Fix $n$ and condition on the relative ordering of $X_1,
\dots, X_{n-1}$: inserting $X_n$ into the $n$ possible rank
slots is uniform and \omterm{def:b3:probability:independence}{independent} of that ordering
(exchangeability of the $n$-tuple). Hence $R_n$ (the event
``$X_n$ takes the top slot'') is \omterm{def:b3:probability:independence}{independent} of the whole
record history $(R_1, \dots, R_{n-1})$, which is a function
of the relative ordering of the first $n - 1$ variables.
Induction gives full \omterm{def:b3:probability:independence}{independence} with $\P(R_n = 1) =
\frac1n$.

(c) $\sum\P(R_n = 1) = \sum\frac1n = \infty$ with
\omterm{def:b3:probability:independence}{independence}: the second Borel--Cantelli half gives records
infinitely often a.s.\ (records never stop --- but they
thin out logarithmically: $\E[\#\text{records} \leq n] =
H_n \approx \ln n$). Consecutive records: $\P(R_n = R_{n+1}
= 1) = \frac1{n(n+1)}$ (\omterm{def:b3:probability:independence}{independence}), and
\[
\sum_n\frac1{n(n+1)} = \sum_n\Bigl(\frac1n -
\frac1{n+1}\Bigr) = 1 < \infty :
\]
the first Borel--Cantelli half applies --- only finitely
many consecutive-record pairs occur, a.s.
\end{solution}

\begin{solution}{exo:b3:probability:12}
(a) A run of length $\ell$ starting at position $i \leq n$
has probability $2^{-\ell}$; union bound: $\P(L_n \geq \ell)
\leq n2^{-\ell}$. With $\ell_n = (1 +
\varepsilon)\log_2n$: $\P(L_n \geq \ell_n) \leq
n^{-\varepsilon}$. Along $n = 2^k$:
$\sum_k2^{-k\varepsilon} < \infty$, so a.s.\ $L_{2^k} <
(1+\varepsilon)k$ eventually (Borel--Cantelli); for general
$n$ pick $2^{k-1} < n \leq 2^k$ and use monotonicity of
$L_n$ plus $\log_22^{k-1} \leq \log_2n$: $L_n \leq L_{2^k}
< (1 + \varepsilon)k \leq (1 + \varepsilon)\frac{k}{k-1}
\log_2n$, and the extra factor is absorbed by enlarging
$\varepsilon$ slightly.

(b) With $\ell = \lceil(1 - \varepsilon)\log_2n\rceil$ and
$m = \lfloor n/\ell\rfloor$ disjoint blocks: the blocks are
\omterm{def:b3:probability:independence}{independent}, each all-heads with probability $2^{-\ell}
\geq n^{-(1-\varepsilon)}/2$, so
\[
\P(L_n < \ell) \leq \bigl(1 - 2^{-\ell}\bigr)^{m}
\leq \exp\bigl(-m2^{-\ell}\bigr)
\leq \exp\Bigl(-c\,\frac{n^{\varepsilon}}{\log_2n}\Bigr)
\]
for a constant $c > 0$ and $n$ large. These probabilities
are summable along $n = 2^k$ (indeed along all $n$):
Borel--Cantelli gives a.s.\ $L_n \geq (1 -
\varepsilon)\log_2n$ eventually (monotonicity fills in
between the $2^k$ as in (a), harmlessly).

(c) Both bounds along a sequence $\varepsilon = \frac1j$,
intersecting countably many \omterm{def:b3:measure:measure}{full-measure} events:
$\frac{L_n}{\log_2n} \to 1$ a.s. For $n = 10^6$: $\log_2n
\approx 19.9$ --- a run of $\approx 20$ heads is not a
suspicious anomaly but a mathematical certainty, and its
absence is evidence of a human faking ``randomness'' (humans
rarely dare write more than $5$ or $6$ heads in a row).
\end{solution}

\begin{solution}{pb:b3:probability:1}
\textbf{1.} $X_n^{\pm}$ are Borel functions of $X_n$: they
remain pairwise \omterm{def:b3:probability:independence}{independent}
(\cref{exo:b3:probability:3}(a)) and identically
distributed, \omterm{def:b3:lebesgue:l1}{integrable}, with $\E X_1 = \E X_1^+ - \E
X_1^-$. If the theorem holds for nonnegative variables,
apply it to both halves and subtract:
$\frac{S_n}n = \frac{S_n^+}n - \frac{S_n^-}n \to \E X_1^+ -
\E X_1^- = m$ a.s.

\textbf{2.} $\P(X_n \neq Y_n) = \P(X_n > n) = \P(X_1 > n)$
(identical \omterm{def:b3:probability:space}{laws}), and $\sum_n\P(X_1 > n) \leq \sum_n\P(X_1
\geq n) \leq \E X_1 < \infty$
(\cref{exo:b3:product:3}(a)). Borel--Cantelli (1): a.s.\
$X_n = Y_n$ for all large $n$, so $S_n - S_n^*$ is
eventually constant in $n$: $\frac{S_n - S_n^*}n \to 0$
a.s., and the two normalized sums share their asymptotic
behavior.

\textbf{3.} $X_1\mathbf 1_{X_1 \leq n} \nearrow X_1$: MCT
gives $\E Y_n \to m$; Ces\`aro means of a convergent
sequence converge to the same limit: $\frac{\E S_n^*}n =
\frac1n\sum_{k\leq n}\E Y_k \to m$. Hence it suffices to
prove $\frac{S^*_n - \E S^*_n}{n} \to 0$ a.s.

\textbf{4.} $\V(Y_n) \leq \E Y_n^2 = \E[X_1^2\mathbf
1_{X_1\leq n}]$. By Tonelli for series,
\[
\sum_n\frac{\E[X_1^2\mathbf 1_{X_1\leq n}]}{n^2}
= \E\Bigl[X_1^2\!\!\sum_{n \geq \max(X_1, 1)}\!\frac1{n^2}
\Bigr]
\leq \E\Bigl[X_1^2\cdot\frac{4}{\max(X_1,1)}\Bigr]
\leq 4\,\E[X_1] < \infty,
\]
using $\sum_{n\geq x}n^{-2} \leq \frac4x$ for $x \geq 1$
(for $x \geq 2$: $\leq \frac1{x-1} \leq \frac2x$; for $1
\leq x < 2$: $\leq \frac{\pi^2}6 \leq \frac4x$ since
$\frac4x > 2$), and $X_1^2/\max(X_1, 1) \leq X_1$ in both
cases $X_1 \gtrless 1$.

\textbf{5.} Pairwise \omterm{def:b3:probability:independence}{independence} gives $\E[(Y_i - \E
Y_i)(Y_j - \E Y_j)] = 0$ for $i \neq j$ (the product
formula for two variables), so variances add: $\V(S^*_k) =
\sum_{n\leq k}\V(Y_n)$. Chebyshev on each $k_j$ and
summing:
\[
\sum_j\P\Bigl(\abs{S^*_{k_j} - \E S^*_{k_j}} \geq
\varepsilon k_j\Bigr)
\leq \frac1{\varepsilon^2}\sum_j\frac1{k_j^2}\sum_{n\leq
k_j}\V(Y_n)
= \frac1{\varepsilon^2}\sum_n\V(Y_n)\!\!\sum_{j : k_j\geq
n}\!\frac1{k_j^2}
\]
(Tonelli for the nonnegative double series).

\textbf{6.} $k_j = \lfloor\alpha^j\rfloor \geq
\frac{\alpha^j}2$ (valid once $\alpha^j \geq 1$, i.e.\ all
$j \geq 0$: $\lfloor x\rfloor \geq \frac x2$ for $x \geq
1$). Hence
\[
\sum_{j : k_j \geq n}\frac1{k_j^2}
\leq 4\sum_{j : \alpha^j \geq n}\alpha^{-2j}
\leq \frac{4}{1 - \alpha^{-2}}\cdot\frac1{n^2}
= \frac{C_\alpha}{n^2},
\]
(geometric series from the first $j$ with $\alpha^j \geq
n$). Combining with questions 4--5, the double sum is
finite; Borel--Cantelli (1), applied for each rational
$\varepsilon$ and intersected, gives
$\frac{S^*_{k_j} - \E S^*_{k_j}}{k_j} \to 0$ a.s., and with
question 3: $\frac{S^*_{k_j}}{k_j} \to m$ a.s.

\textbf{7.} $Y_n \geq 0$ makes $n \mapsto S^*_n$
nondecreasing: for $k_j \leq n \leq k_{j+1}$,
\[
\frac{S^*_{k_j}}{k_{j+1}} \leq \frac{S^*_n}{n} \leq
\frac{S^*_{k_{j+1}}}{k_j},
\]
which is the displayed sandwich after inserting
$\frac{k_j}{k_{j+1}}$ and $\frac{k_{j+1}}{k_j}$. Since
$\frac{k_{j+1}}{k_j} \to \alpha$, question 6 gives a.s.
\[
\frac m\alpha \leq \liminf_n\frac{S^*_n}n \leq
\limsup_n\frac{S^*_n}n \leq \alpha m .
\]

\textbf{8.} Apply question 7 for $\alpha = 1 + \frac1p$, $p
\in \N^*$: countably many a.s.\ events; on their
intersection, letting $p \to \infty$: $\lim\frac{S^*_n}n =
m$ a.s. With questions 1--3, $\frac{S_n}n \to \E X_1$
a.s.: the strong \omterm{def:b3:probability:space}{law} of large numbers, under pairwise
\omterm{def:b3:probability:independence}{independence}.

\textbf{9.} Independence-type hypotheses appeared three
times: (i) additivity of variances (question 5) ---
pairwise suffices; (ii) identical distribution, in the
truncation sums (question 2) and the mean computation
(question 3) --- no \omterm{def:b3:probability:independence}{independence} at all; (iii)
Borel--Cantelli (1) (questions 2 and 6) --- valid without
any \omterm{def:b3:probability:independence}{independence}. Full mutual \omterm{def:b3:probability:independence}{independence} was never
invoked: Etemadi's observation.

\textbf{10.} Fix a base $b$ and a digit $r$. The base-$b$
digits $(d_k)$ of a uniform $\omega$ are i.i.d.\ uniform on
$\{0, \dots, b-1\}$ (each digit vector value occupies an
interval of length $b^{-m}$: the argument of
\cref{thm:b3:probability:existence} verbatim). The strong
\omterm{def:b3:probability:space}{law} applied to the i.i.d.\ bounded variables $\mathbf
1_{d_k = r}$ gives: a.s., the frequency of the digit $r$
tends to $\frac1b$. Intersecting over the countably many
pairs $(b, r)$: almost every number is
\emph{simply normal in every base}. An explicit non-normal
number: $x = 0.100100100\ldots_2$ (frequency of ones
$\frac13 \neq \frac12$). The contrast is humbling: almost
all numbers are normal, yet for $\sqrt2$, $\eu$, or $\pi$
normality remains unproved --- \omterm{def:b3:measure:measure}{measure} theory counts
without exhibiting.

\textbf{11.} By \cref{exo:b3:probability:10} iterated, a
single uniform variable yields a sequence of i.i.d.\
uniform \emph{vectors} $U_k$ on $\intcc01^d$ (split the
digit set of each $U_n$ of
\cref{thm:b3:probability:existence} into $d$ subfamilies).
For $g \in L^1(\intcc01^d)$, the variables $g(U_k)$ are
i.i.d.\ \omterm{def:b3:lebesgue:l1}{integrable} with mean $\int g\,\dd\lambda_d$
(transfer): the strong \omterm{def:b3:probability:space}{law} gives
\[
\frac1n\sum_{k=1}^ng(U_k)
\xrightarrow[n\to\infty]{\text{a.s.}}
\int_{\intcc01^d}g\,\dd\lambda_d :
\]
\omterm{ex:b3:probability:sllnapps}{Monte Carlo} integration converges almost surely, in every
dimension --- the error size is the business of the central
limit theorem (\cref{ch:b3:clt}).

\textbf{12.} Let $A_k = \{\abs{S_k} \geq \varepsilon\} \cap
\bigcap_{j<k}\{\abs{S_j} < \varepsilon\}$: the $A_k$ are
disjoint with union $A = \{\max_{k\leq n}\abs{S_k} \geq
\varepsilon\}$. Then
\[
\E S_n^2 \geq \sum_{k=1}^n\E\bigl[S_n^2\mathbf 1_{A_k}\bigr]
= \sum_{k=1}^n\E\Bigl[\bigl(S_k^2 + 2S_k(S_n - S_k) + (S_n
- S_k)^2\bigr)\mathbf 1_{A_k}\Bigr]
\geq \sum_{k=1}^n\E\bigl[S_k^2\mathbf 1_{A_k}\bigr],
\]
because the cross term vanishes: $S_k\mathbf 1_{A_k}$ is a
Borel function of the coalition $(Z_1, \dots, Z_k)$, which
is \omterm{def:b3:probability:independence}{independent} of $S_n - S_k$, a function of $(Z_{k+1},
\dots, Z_n)$ (\cref{thm:b3:probability:independence}), so
$\E[S_k\mathbf 1_{A_k}(S_n - S_k)] = \E[S_k\mathbf
1_{A_k}]\,\E[S_n - S_k] = 0$. On $A_k$, $S_k^2 \geq
\varepsilon^2$, whence $\E S_n^2 \geq
\varepsilon^2\sum_k\P(A_k) = \varepsilon^2\P(A)$; and $\E
S_n^2 = \sum_{k\leq n}\V(Z_k)$ (variances add). The
decisive step is the factorization: $S_k\mathbf 1_{A_k}$ is
a \emph{nonlinear} function of the whole first block, and
its \omterm{def:b3:probability:independence}{independence} from the second block is coalition
\omterm{def:b3:probability:independence}{independence} --- pairwise \omterm{def:b3:probability:independence}{independence} of the $Z_i$ only
decorrelates pairs and would not justify it.

\textbf{13.} Fix $N$ and apply question 12 to $Z_{N+1},
\dots, Z_{N+m}$:
\[
\P\Bigl(\max_{N < k \leq N+m}\abs{S_k - S_N} >
\varepsilon\Bigr) \leq
\frac1{\varepsilon^2}\sum_{j=N+1}^{N+m}\V(Z_j) \leq
\frac{r_N}{\varepsilon^2},
\qquad r_N = \sum_{j>N}\V(Z_j) .
\]
The events increase with $m$; \omterm{def:b3:topology:continuity}{continuity} from below gives
$\P(\sup_{k>N}\abs{S_k - S_N} > \varepsilon) \leq
r_N/\varepsilon^2$, and $r_N \to 0$ by hypothesis. Hence
for each $p \in \N^*$, $\P\bigl(\bigcap_N\{\sup_{k>N}
\abs{S_k - S_N} > \frac1p\}\bigr) \leq \inf_Np^2r_N = 0$:
almost surely, for every $p$ there is $N$ with
$\sup_{k>N}\abs{S_k - S_N} \leq \frac1p$ (intersect the
countably many a.s.\ events over $p$), so that $\abs{S_k -
S_l} \leq \frac2p$ for all $k, l > N$: the partial sums
are a.s.\ Cauchy, hence a.s.\ convergent.

\textbf{14.} The variables $Z_n = x_n\varepsilon_n$ are
\omterm{def:b3:probability:independence}{independent} (Borel functions of \omterm{def:b3:probability:independence}{independent} variables,
\cref{exo:b3:probability:3}(a)), centered, with $\V(Z_n) =
x_n^2$: question 13 applies when $\sum_nx_n^2 < \infty$ and
gives a.s.\ convergence. In general, for each $N$ the
convergence of $\sum_nx_n\varepsilon_n$ is unaffected by
the values of $\varepsilon_1, \dots, \varepsilon_N$: the
convergence event lies in the tail $\sigma$-algebra of the
\omterm{def:b3:probability:independence}{independent} sequence $(\varepsilon_n)$, so Kolmogorov's
\omterm{thm:b3:probability:zeroone}{zero--one law} (\cref{thm:b3:probability:zeroone}) forces
its probability to be $0$ or $1$.

\textbf{15.} (a) Splitting at the level $\theta\E Z$ and
using Cauchy--Schwarz on the upper piece,
\[
\E Z = \E\bigl[Z\mathbf 1_{Z \leq \theta\E Z}\bigr] +
\E\bigl[Z\mathbf 1_{Z > \theta\E Z}\bigr]
\leq \theta\,\E Z + \sqrt{\E Z^2}\,
\sqrt{\P(Z > \theta\E Z)} ,
\]
so $(1 - \theta)\E Z \leq \sqrt{\E Z^2\,\P(Z > \theta\E
Z)}$; square. (b) Expand $T_n^4 =
\sum_{i,j,k,l}x_ix_jx_kx_l\,
\E[\varepsilon_i\varepsilon_j\varepsilon_k\varepsilon_l]$:
the \omterm{def:b3:probability:space}{expectation} is $1$ when the indices pair off (all four
equal, or two distinct pairs, the latter in $3$
arrangements) and $0$ otherwise (an unpaired sign has zero
mean and factors out by \omterm{def:b3:probability:independence}{independence}). Hence
\[
\E T_n^4 = \sum_kx_k^4 + 3\sum_{i\neq j}x_i^2x_j^2 =
3s_n^4 - 2\sum_kx_k^4 \leq 3s_n^4 .
\]
(c) Paley--Zygmund with $Z = T_n^2$, $\E Z = s_n^2$,
$\theta = \frac14$:
\[
\P\Bigl(\abs{T_n} > \frac{s_n}2\Bigr) = \P\Bigl(T_n^2 >
\frac{s_n^2}4\Bigr) \geq \Bigl(\frac34\Bigr)^2
\frac{s_n^4}{3s_n^4} = \frac3{16} .
\]
If the series converged with positive probability, it
would converge a.s.\ (question 14), so $\sup_n\abs{T_n} <
\infty$ a.s., and some $M$ would satisfy
$\P(\sup_n\abs{T_n} > M) < \frac3{16}$; but as soon as $s_n
> 2M$, $\P(\abs{T_n} > M) \geq \P(\abs{T_n} > \frac{s_n}2)
\geq \frac3{16}$: contradiction. So divergence is almost
sure, and with question 14 the dichotomy is \omterm{def:b3:complete:complete}{complete}.

\textbf{16.} Here $x_n = n^{-s}$ and $\sum_nn^{-2s} <
\infty$ exactly when $s > \frac12$: by questions 14--15,
$\sum_n\frac{\varepsilon_n}{n^s}$ converges a.s.\ if and
only if $s > \frac12$ (for $s \leq \frac12$, a.s.\
divergence). For $\frac12 < s \leq 1$ the convergence is
never absolute. The comparison is instructive: perfectly
alternating signs cancel at strength $n^{-s}$ for every $s
> 0$, while typical random signs cancel only at
square-root strength --- the random walk of question 21
grows like $\sqrt n$, and Abel summation converts exactly
that growth into convergence of $\sum\varepsilon_nn^{-s}$
for $s > \frac12$.

\textbf{17.} (a) $\cosh\lambda =
\sum_k\frac{\lambda^{2k}}{(2k)!}$ and $\eu^{\lambda^2/2} =
\sum_k\frac{\lambda^{2k}}{2^kk!}$; and $(2k)! \geq 2^kk!$
holds termwise, because $\frac{(2k)!}{k!} =
\prod_{i=1}^k(k + i) \geq \prod_{i=1}^k(2i) = 2^kk!$ (each
factor satisfies $k + i \geq 2i$ for $i \leq k$), so that
in fact $(2k)! \geq 2^k(k!)^2 \geq 2^kk!$. (b) Note $a \leq 0 \leq b$ ($Z$ is
centered), and by convexity of $z \mapsto \eu^{\lambda z}$,
for $z \in \intcc ab$:
\[
\eu^{\lambda z} \leq \frac{b - z}{b - a}\,\eu^{\lambda a} +
\frac{z - a}{b - a}\,\eu^{\lambda b},
\qquad\text{so}\qquad
\E\,\eu^{\lambda Z} \leq \frac{b\,\eu^{\lambda a} -
a\,\eu^{\lambda b}}{b - a}
= (1 - p)\eu^{-pt} + p\,\eu^{(1-p)t} = \eu^{\varphi(t)}
\]
with $p = \frac{-a}{b-a} \in \intcc01$, $t = \lambda(b -
a)$, $\varphi(t) = -pt + \log(1 - p + p\eu^t)$. Then
$\varphi(0) = 0$, $\varphi'(t) = -p + \frac{p\eu^t}{1 - p +
p\eu^t}$ vanishes at $0$, and $\varphi''(t) = \rho(1 -
\rho) \leq \frac14$ for $\rho = \frac{p\eu^t}{1 - p +
p\eu^t} \in \intcc01$: Taylor at order $2$ gives
$\varphi(t) \leq \frac{t^2}8 = \frac{\lambda^2(b-a)^2}8$.

\textbf{18.} For $\lambda > 0$, Markov applied to the
positive variable $\eu^{\lambda(S_n - \E S_n)}$
(\cref{prop:b3:probability:markov}) and the product
formula for \omterm{def:b3:probability:independence}{independent} variables give
\[
\P(S_n - \E S_n \geq t) \leq \eu^{-\lambda
t}\prod_{i=1}^n\E\,\eu^{\lambda(X_i - \E X_i)}
\leq \exp\Bigl(-\lambda t +
\frac{\lambda^2}8\sum_i(b_i - a_i)^2\Bigr),
\]
by question 17(b) applied to each centered $X_i - \E X_i
\in \intcc{a_i - \E X_i}{b_i - \E X_i}$ (same width).
Minimizing the exponent at $\lambda = \frac{4t}{D}$, $D =
\sum_i(b_i - a_i)^2$, yields $-\frac{2t^2}D$. The lower
tail follows by applying the result to $(-X_i)$.

\textbf{19.} Take $t = n\varepsilon$ and $D = n(b - a)^2$:
\[
\P\Bigl(\Bigl|\frac{S_n}n - m\Bigr| \geq \varepsilon\Bigr)
\leq 2\exp\Bigl(\frac{-2n^2\varepsilon^2}{n(b-a)^2}\Bigr)
= 2\exp\Bigl(\frac{-2n\varepsilon^2}{(b-a)^2}\Bigr),
\]
which is summable in $n$ (a geometric-type series):
Borel--Cantelli (\cref{thm:b3:probability:borelcantelli})
gives that a.s.\ $\abs{\frac{S_n}n - m} < \varepsilon$
eventually; intersecting over $\varepsilon = \frac1p$
yields $\frac{S_n}n \to m$ a.s. Comparison: Etemadi asks
only $X_1 \in L^1$ and pairwise \omterm{def:b3:probability:independence}{independence}, and delivers
no rate; Hoeffding asks boundedness and full \omterm{def:b3:probability:independence}{independence},
and delivers an explicit exponential guarantee at every
finite $n$ --- the two theorems answer different questions
about the same limit.

\textbf{20.} The $g(U_k)$ are i.i.d.\ with values in
$\intcc01$ and mean $\int g\,\dd\lambda_d$ (transfer), so
question 18 with $b_i - a_i = 1$, $t = n\varepsilon$ gives
the two-sided bound $2\eu^{-2n\varepsilon^2} \leq \delta$
as soon as $\eu^{2n\varepsilon^2} \geq \frac2\delta$, i.e.\
$n \geq \frac{\log(2/\delta)}{2\varepsilon^2}$. For
$\varepsilon = \delta = 10^{-2}$:
\[
n \geq \frac{\log 200}{2\cdot10^{-4}} =
\frac{5.2983\ldots}{0.0002} \approx 26\,492 :
\]
about $26\,500$ samples guarantee a $1\%$ accuracy with
$99\%$ confidence --- in every dimension $d$, for every
\omterm{def:b3:lebesgue:measurable}{measurable} integrand with values in $\intcc01$. Question
11's strong \omterm{def:b3:probability:space}{law} promised convergence with no finite-$n$
guarantee; a deterministic grid with $k$ points per axis
costs $k^d$ evaluations, exponential in $d$. Concentration
is what makes \omterm{ex:b3:probability:sllnapps}{Monte Carlo} a \emph{method} rather than a
hope.

\textbf{21.} \omterm{def:b3:probability:independence}{Independence} and the product formula:
$\E\,\eu^{\lambda S_n} = (\E\,\eu^{\lambda\varepsilon_1})^n
= (\cosh\lambda)^n \leq \eu^{n\lambda^2/2}$ by question
17(a). Markov on $\eu^{\lambda S_n}$:
\[
\P(S_n \geq x) \leq \eu^{-\lambda x + n\lambda^2/2}
= \eu^{-x^2/(2n)}
\qquad\text{at the optimum } \lambda = \frac xn,
\]
and the symmetric bound for $-S_n$ (same \omterm{def:b3:probability:space}{law}) doubles the
constant for $\abs{S_n}$.

\textbf{22.} Fix $\eta > 0$ and set $x_n = (1 +
\eta)\sqrt{2n\log n}$ for $n \geq 2$:
\[
\P(\abs{S_n} \geq x_n) \leq 2\exp\bigl(-(1 +
\eta)^2\log n\bigr) = \frac{2}{n^{(1+\eta)^2}},
\]
summable since $(1 + \eta)^2 > 1$. Borel--Cantelli: a.s.\
$\abs{S_n} < (1 + \eta)\sqrt{2n\log n}$ for all large $n$,
so $\limsup_n\frac{\abs{S_n}}{\sqrt{2n\log n}} \leq 1 +
\eta$ a.s.; intersecting the a.s.\ events for $\eta =
\frac1p$, $p \in \N^*$, gives the claim. The walk of size
$n$ has typical amplitude $\sqrt n$ (its variance), and
even its worst excursions exceed that scale by at most
$\sqrt{2\log n}$.

\textbf{23.} With $n_j = 2^j$ and $x = (1 +
\eta)\sqrt{2n_j\log\log n_j}$ (defined for $j \geq 2$),
question 21 gives
\[
\P\bigl(S_{n_j} \geq x\bigr) \leq \exp\bigl(-(1 +
\eta)^2\log\log n_j\bigr) = (j\log 2)^{-(1+\eta)^2},
\]
summable in $j$ since $(1 + \eta)^2 > 1$: Borel--Cantelli
and $\eta = \frac1p$ give $\limsup_jS_{n_j}/\sqrt{2n_j
\log\log n_j} \leq 1$ a.s. What is missing for the full
upper half is the bridge between checkpoints: one must
show $\max_{n_j\leq n\leq n_{j+1}}S_n$ exceeds
$(1+\eta)\sqrt{2n_j\log\log n_j}$ only finitely often,
which demands a maximal inequality with \emph{Gaussian}
tails (L\'evy's reflection inequality or Ottaviani's
inequality, not proved here). Question 12 is quantitatively
too weak: it bounds the probability by
\[
\frac{n_j}{(1+\eta)^2\,2n_j\log\log n_j}
= \frac{1}{2(1+\eta)^2\log(j\log2)},
\]
which tends to $0$ but is \emph{not summable} in $j$:
Borel--Cantelli cannot conclude. The lower half of the \omterm{def:b3:probability:space}{law}
of the iterated logarithm applies the second
Borel--Cantelli lemma to the \omterm{def:b3:probability:independence}{independent} increments
$S_{n_{j+1}} - S_{n_j}$, using matching lower bounds for
Gaussian-type tails. Both refinements are genuine Year 3
probability, one course further along; what this problem
delivers unaided is the exact iterated-logarithm scale
along geometric times.

\textbf{24.} Each $\hat p_i$ is an average of $n$ i.i.d.\
indicator variables with values in $\intcc01$ and mean
$\P(A_i)$: Hoeffding gives $\P(\abs{\hat p_i - \P(A_i)} >
\varepsilon) \leq 2\eu^{-2n\varepsilon^2}$. The union bound
multiplies by $N$. Solving $2N\eu^{-2n\varepsilon^2} \leq
\delta$: $n \geq \frac{\ln(2N/\delta)}{2\varepsilon^2}$.
Numerically: $\ln\frac{2\cdot10^6}{0.05} =
\ln(4\cdot10^7) \approx 17.5$, so $n \geq
\frac{17.5}{2\cdot10^{-4}} \approx 87\,600$: estimating
\emph{one} probability to $\pm1\%$ takes about $18\,500$
samples ($\ln(2/\delta)/2\varepsilon^2$), and a
\emph{million} probabilities only $\approx 4.7$ times more
--- uniformity costs $\ln N$, not $N$: the observation that
makes empirical risk minimization, and with it machine
learning, statistically possible.

\textbf{25.} The variables $X_n = \frac{\varepsilon_n}
{n^\alpha}$ are \omterm{def:b3:probability:independence}{independent}, centered, bounded, with
$\sum_n\V(X_n) = \sum_nn^{-2\alpha}$. If $\alpha >
\frac12$: the variance series converges, and the
one-series theorem (Part VI) gives a.s.\ convergence of
$\sum X_n$. If $\alpha \leq \frac12$: the variance series
diverges, and the converse half (Part VI's
Paley--Zygmund argument, applicable since the summands are
bounded by $1$) gives a.s.\ divergence. Absolute
convergence asks $\sum n^{-\alpha} < \infty$: $\alpha > 1$.
On $\intoc{\frac12}1$, the series converges a.s.\ although
$\sum\abs{X_n} = \infty$ surely: the signs conspire to
cancel, with probability one --- convergence by
cancellation, invisible to any absolute test, and (by the
\omterm{thm:b3:probability:zeroone}{zero--one law}) with a deterministic verdict all the same.
\end{solution}
