\section*{Chapter \ref{ch:b1:counting} --- Counting}

\begin{solution}{exo:b1:counting:1}
Independent stages and the product rule: $26^2 \times 10^3 \times 26^2
= 26^4 \times 1000 = 456\,976\,000$ plates. With the four letters
pairwise distinct, the letter stages form a $4$-arrangement of the
alphabet: $26 \times 25 \times 24 \times 23 = 358\,800$ ways, so
$358\,800 \times 1000 = 358\,800\,000$ plates.
\end{solution}

\begin{solution}{exo:b1:counting:2}
\textsc{orange} has $6$ distinct letters: $6! = 720$ anagrams.
\textsc{banana} has $6$ letters with repetitions ($3$ a's, $2$ n's,
$1$ b): each anagram is determined by the positions of the a's
($\binom 63$ choices), then of the n's among the remaining $3$ spots
($\binom 32$), the b taking the last spot:
$\binom{6}{3}\binom{3}{2} = 20 \times 3 = 60$ anagrams
(equivalently $6!/(3!\,2!\,1!) = 60$).
\end{solution}

\begin{solution}{exo:b1:counting:3}
Total: $\binom{12}{4} = 495$. Exactly $2$ women: choose them
($\binom 72 = 21$) and $2$ men ($\binom 52 = 10$): $210$ committees.
At least one man: complement of ``no man'',
$\binom{12}{4} - \binom{7}{4} = 495 - 35 = 460$.
\end{solution}

\begin{solution}{exo:b1:counting:4}
\emph{Birthdays:} the boxes are the $12$ months; $13$ people into $12$
boxes force two in the same box
(\cref{cor:b1:counting:pigeonhole}).

\emph{Consecutive integers:} the boxes are the $n$ pairs $\{1,2\},
\{3,4\}, \dots, \{2n-1, 2n\}$, which \omterm{thm:b1:logic:partition}{partition} $\intint{1}{2n}$.
Choosing $n + 1$ integers puts two in the same pair, and the two
elements of a pair are consecutive.
\end{solution}

\begin{solution}{exo:b1:counting:5}
For $1 \leq k \leq n$,
\[
k \binom nk = k\,\frac{n!}{k!\,(n-k)!}
= n\,\frac{(n-1)!}{(k-1)!\,(n-k)!} = n \binom{n-1}{k-1}.
\]
Summing and reindexing with $j = k - 1$:
\[
\sum_{k=0}^{n} k \binom nk = n \sum_{j=0}^{n-1} \binom{n-1}{j}
= n\, 2^{n-1}
\]
by \cref{prop:b1:counting:identities}. (Alternative: differentiate
$(1+x)^n = \sum_k \binom nk x^k$ and set $x = 1$.)
\end{solution}

\begin{solution}{exo:b1:counting:6}
A strictly increasing \omterm{def:b1:logic:map}{map} $f \colon \intint{1}{k} \to \intint{1}{n}$
is determined by its image, a $k$-subset of $\intint{1}{n}$ (list the
subset in increasing order); conversely every $k$-subset gives exactly
one such \omterm{def:b1:logic:map}{map}. Hence $\binom nk$ strictly increasing \omterm{def:b1:logic:map}{maps}.

If $f$ is merely increasing, set $g(i) = f(i) + i - 1$. Then $g$ is
strictly increasing (between consecutive arguments, $f$ gains $\geq 0$
and $i - 1$ gains $1$) with values in $\intint{1}{n + k - 1}$; and
$f(i) = g(i) - i + 1$ recovers $f$ from any strictly increasing $g$
into $\intint{1}{n+k-1}$. This is a bijection, so there are
$\binom{n + k - 1}{k}$ increasing \omterm{def:b1:logic:map}{maps}.
\end{solution}

\begin{solution}{exo:b1:counting:7}
Split a \omterm{def:b1:logic:sets}{set} $E$ with $m + n$ elements into blocks $M$ ($m$ elements)
and $N$ ($n$ elements). A $k$-subset of $E$ contains some $j$ elements
of $M$ ($0 \leq j \leq k$) and $k - j$ of $N$; for fixed $j$ there are
$\binom mj \binom{n}{k-j}$ such subsets, and the cases $j = 0, \dots,
k$ \omterm{thm:b1:logic:partition}{partition} the $k$-subsets. The sum rule gives Vandermonde's
identity\index{Vandermonde's identity}.

With $m = n = k$: $\binom{2n}{n} = \sum_{j=0}^{n} \binom nj
\binom{n}{n-j} = \sum_{j=0}^{n} \binom nj^2$, using $\binom{n}{n-j} =
\binom nj$.
\end{solution}

\begin{solution}{exo:b1:counting:8}
Let $A_d$ be the multiples of $d$ in $\intint{1}{1000}$, so
$\abs{A_d} = \lfloor 1000/d \rfloor$. Inclusion--exclusion
(\cref{thm:b1:counting:inclexcl}) with $A_2, A_3, A_5$, noting
$A_2 \cap A_3 = A_6$ etc.:
\[
500 + 333 + 200 - 166 - 100 - 66 + 33 = 734 .
\]
So $734$ integers are divisible by $2$, $3$ or $5$.
\end{solution}

\begin{solution}{exo:b1:counting:9}
Onto $2$ elements: all $2^4 = 16$ \omterm{def:b1:logic:map}{maps} except the $2$ constant \omterm{def:b1:logic:map}{maps}:
$14$ surjections.

Onto $3$ elements: by inclusion--exclusion on the missed values, the
number of \omterm{def:b1:logic:map}{maps} from a $4$-set to a $3$-set missing at least one value
is $\binom 31 2^4 - \binom 32 1^4 = 48 - 3 = 45$; total \omterm{def:b1:logic:map}{maps} $3^4 =
81$; surjections: $81 - 45 = 36$. (Check: a surjection from $4$ onto
$3$ elements doubles up exactly one value: choose the doubled value
($3$), the pair mapping to it ($\binom 42 = 6$), and a bijection for
the rest ($2$): $3 \times 6 \times 2 = 36$.)
\end{solution}

\begin{solution}{exo:b1:counting:10}
A solution of $x_1 + \dots + x_n = k$ in $\N^n$ encodes as a row of
$k$ stars and $n - 1$ bars: write $x_1$ stars, a bar, $x_2$ stars, a
bar, \dots, ending with $x_n$ stars. This is a bijection onto the
words of length $k + n - 1$ using $k$ stars and $n - 1$ bars, and
those words are determined by the positions of the stars:
$\binom{n + k - 1}{k}$. Selections with repetition correspond to
solutions of the equation ($x_i$ = number of copies of object $i$), so
the count is the same.
\end{solution}

\begin{solution}{exo:b1:counting:11}
With $A_i = \{\sigma : \sigma(i) = i\}$, a \omterm{def:b1:counting:objects}{permutation} in
$\bigcap_{i \in I} A_i$ fixes every $i \in I$ and permutes the other
$n - \abs I$ points freely: $\abs{\bigcap_{i \in I} A_i} =
(n - \abs I)!$. Inclusion--exclusion:
\[
\Bigl|\bigcup_i A_i\Bigr|
= \sum_{k=1}^{n} (-1)^{k+1} \binom nk (n-k)!
= \sum_{k=1}^{n} (-1)^{k+1} \frac{n!}{k!} ,
\]
since there are $\binom nk$ subsets $I$ of size $k$. Hence
\[
D_n = n! - \Bigl|\bigcup_i A_i\Bigr|
= n!\Bigl(1 - \sum_{k=1}^{n} \frac{(-1)^{k+1}}{k!}\Bigr)
= n! \sum_{k=0}^{n} \frac{(-1)^k}{k!} .
\]
For the second identity: classify the \omterm{def:b1:counting:objects}{permutations} $\sigma$ of
$\intint{1}{n}$ by their fixed-point \omterm{def:b1:logic:sets}{set} $F(\sigma)$. For a fixed
$k$-subset $F$, the \omterm{def:b1:counting:objects}{permutations} with $F(\sigma) = F$ are exactly the
derangements of the complement: $D_{n-k}$ of them. Summing over the
$\binom nk$ choices of $F$ for each $k$:
$n! = \sum_{k=0}^{n} \binom nk D_{n-k}$.
\end{solution}

\begin{solution}{exo:b1:counting:12}
\emph{First identity.} Count the pairs $(A, a)$ where $A \subseteq E$
($\abs E = n$) and $a \in A$. By size of $A$: $\sum_k \binom nk k$
pairs. By choosing the marked element first: $n$ choices for $a$, then
any subset of the remaining $n - 1$ elements to complete $A$:
$n\,2^{n-1}$ pairs.

\emph{Second identity.} Count the triples $(A, a, b)$ with $a, b \in
A$ (possibly $a = b$). By size: $\sum_k k^2 \binom nk$. Directly:
either $a = b$ ($n\,2^{n-1}$ triples, previous count) or $a \neq b$
($n(n-1)$ ordered choices, then any subset of the other $n - 2$
elements: $n(n-1)\,2^{n-2}$). Total
\[
n\,2^{n-1} + n(n-1)\,2^{n-2} = n\,2^{n-2}\,(2 + n - 1)
= n(n+1)\,2^{n-2} .
\]
\end{solution}

\begin{solution}{pb:b1:counting:1}
\textbf{1.} $D_1 = 0$ (the only \omterm{def:b1:counting:objects}{permutation} fixes $1$), $D_2 = 1$
(the swap), $D_3 = 2$ (in one-line notation: $231$ and $312$). For
$n = 4$, group by $\sigma(1)$: with $\sigma(1) = 2$ the derangements
are $2143$, $2341$, $2413$; with $\sigma(1) = 3$: $3142$, $3412$,
$3421$; with $\sigma(1) = 4$: $4123$, $4312$, $4321$. Three in each
group: $D_4 = 9$.

\textbf{2.} A \omterm{def:b1:counting:objects}{permutation} with exactly $k$ fixed points is
determined by the choice of its fixed-point \omterm{def:b1:logic:sets}{set} $F$ ($\binom nk$
ways) together with its restriction to the complement, which must
be a \omterm{def:b1:counting:objects}{permutation} of $n - k$ points with \emph{no} fixed point
($D_{n-k}$ ways). The two choices are independent and the
correspondence is \omterm{def:b1:logic:inj}{bijective}: $P_k(n) = \binom nk D_{n-k}$.

\textbf{3.} $P_0(4) = D_4 = 9$; $P_1(4) = \binom41 D_3 = 4 \times 2
= 8$; $P_2(4) = \binom42 D_2 = 6$; $P_3(4) = \binom43 D_1 = 0$
(three fixed points force a fourth); $P_4(4) = 1$. Sum: $9 + 8 + 6
+ 0 + 1 = 24 = 4!$. No match ($9$ cases) beats exactly one match
($8$ cases) --- narrowly.

\textbf{4.} Count the pairs $(\sigma, i)$ with $\sigma(i) = i$. For
fixed $i$, the \omterm{def:b1:counting:objects}{permutations} fixing $i$ are the \omterm{def:b1:counting:objects}{permutations} of the
other $n - 1$ points: $(n-1)!$ of them. Hence the number of pairs
is $n \cdot (n-1)! = n!$, and this number is also
$\sum_\sigma \abs{\mathrm{Fix}(\sigma)}$. Dividing by the number
$n!$ of \omterm{def:b1:counting:objects}{permutations}: the average number of fixed points is
exactly $1$, for every $n \geq 1$.

\textbf{5.} Let $\sigma$ be a derangement of $\intint1{n+1}$ and
$j = \sigma(n+1) \in \intint1n$: $n$ possible values. \emph{Case
$\sigma(j) = n+1$:} the points $j$ and $n+1$ swap, and $\sigma$
restricted to the remaining $n - 1$ points is an arbitrary
derangement of them: $D_{n-1}$ possibilities. \emph{Case $\sigma(j)
\neq n+1$:} let $i_0 = \sigma^{-1}(n+1)$; here $i_0 \neq j$ and
$i_0 \leq n$. Define $\tau$ on $\intint1n$ by $\tau(i) = \sigma(i)$
for $i \neq i_0$ and $\tau(i_0) = j$. Then $\tau$ is a \omterm{def:b1:counting:objects}{permutation}
of $\intint1n$ (the value $n+1$ has been replaced by the missing
value $j$), and it is a derangement: $\tau(i_0) = j \neq i_0$, and
$\tau(i) = \sigma(i) \neq i$ elsewhere. Conversely, from a
derangement $\tau$ of $\intint1n$ and the value $j$, one recovers
$\sigma$ by setting $\sigma(n+1) = j$, $\sigma(\tau^{-1}(j)) = n+1$
and $\sigma = \tau$ elsewhere: a bijection, giving $D_n$
possibilities. Summing over $j$: $D_{n+1} = n(D_n + D_{n-1})$.
Numerically: $D_5 = 4(9 + 2) = 44$, $D_6 = 5(44 + 9) = 265$.

\textbf{6.} From question 5, $D_{n+1} = nD_n + nD_{n-1}$, so
\[
u_{n+1} = D_{n+1} - (n+1)D_n = nD_n + nD_{n-1} - (n+1)D_n
= -(D_n - nD_{n-1}) = -u_n .
\]
Since $u_1 = D_1 - 1 \cdot D_0 = -1$, induction gives $u_n =
(-1)^n$, i.e.\ $D_n = nD_{n-1} + (-1)^n$ for $n \geq 1$.

\textbf{7.} Induction on $n$. Base: $D_0 = 1 = 0!\,s_0$. Step:
assuming $D_{n-1} = (n-1)!\,s_{n-1}$,
\[
D_n = nD_{n-1} + (-1)^n = n!\,s_{n-1} + (-1)^n
= n!\Bigl(s_{n-1} + \frac{(-1)^n}{n!}\Bigr) = n!\,s_n ,
\]
which is the formula. No inclusion--exclusion was used: only the
combinatorial recurrence of question 5.

\textbf{8.} Trinomial revision, by factorials:
\[
\binom nk \binom kj
= \frac{n!}{k!\,(n-k)!} \cdot \frac{k!}{j!\,(k-j)!}
= \frac{n!}{j!\,(n-j)!} \cdot \frac{(n-j)!}{(k-j)!\,(n-k)!}
= \binom nj \binom{n-j}{k-j} .
\]
Now substitute $a_k = \sum_j \binom kj b_j$ and exchange the two
\omterm{def:b1:counting:card}{finite} sums:
\[
\sum_{k=0}^{n} (-1)^{n-k} \binom nk a_k
= \sum_{j=0}^{n} b_j \binom nj
\sum_{k=j}^{n} (-1)^{n-k} \binom{n-j}{k-j}
= \sum_{j=0}^{n} b_j \binom nj
\sum_{i=0}^{n-j} (-1)^{(n-j)-i} \binom{n-j}{i} .
\]
The inner sum is the expansion of $(1 + (-1))^{n-j} = 0^{n-j}$
(binomial theorem, \cref{thm:b1:counting:binomial}): it vanishes
for $j < n$ and equals $1$ for $j = n$. Only $j = n$ survives, and
the right-hand side is $b_n$, as claimed.

\textbf{9.} By the symmetry $\binom nk = \binom n{n-k}$, the
identity of \cref{exo:b1:counting:11} rewrites as $n! =
\sum_{k=0}^n \binom nk D_k$. Apply question 8 with $a_n = n!$ and
$b_k = D_k$:
\[
D_n = \sum_{k=0}^{n} (-1)^{n-k} \binom nk k!
= \sum_{k=0}^{n} (-1)^{n-k} \frac{n!}{(n-k)!}
= n! \sum_{j=0}^{n} \frac{(-1)^j}{j!} ,
\]
reindexing by $j = n - k$: the formula a third time.

\textbf{10.} $D_n = n!\,s_n$ (question 7), so
\[
\Bigl| D_n - \frac{n!}{\eu} \Bigr|
= n!\,\abs{s_n - \eu^{-1}} < \frac{n!}{(n+1)!} = \frac1{n+1} .
\]

\textbf{11.} For $n \geq 1$, $\frac1{n+1} \leq \frac12$, and the
inequality of question 10 is strict: $D_n$ lies at distance
$< \frac12$ from $n!/\eu$, so it is the unique nearest integer. For
$n = 0$ the bound only gives distance $< 1$, and indeed the claim
fails there: $0!/\eu \approx 0.368$ has nearest integer $0$, while
$D_0 = 1$.

\textbf{12.} $\eu^{-1} - s_n = \sum_{k \geq n+1} (-1)^k/k!$ is an
alternating series with strictly decreasing terms, so its sign is
the sign of its first term $(-1)^{n+1}/(n+1)!$. Hence $s_n -
\eu^{-1}$ has the sign of $(-1)^n$: for $n$ even, $s_n > \eu^{-1}$
and $D_n = n!\,s_n > n!/\eu$; for $n$ odd, $D_n < n!/\eu$.

\textbf{13.} $D_7 = 6(265 + 44) = 6 \times 309 = 1854$; $D_8 =
7(1854 + 265) = 7 \times 2119 = 14\,833$; $D_9 = 8(14\,833 + 1854)
= 8 \times 16\,687 = 133\,496$; $D_{10} = 9(133\,496 + 14\,833) =
9 \times 148\,329 = 1\,334\,961$. Check: $10!/\eu = 3\,628\,800 /
2.718281828 \approx 1\,334\,960.92$, whose nearest integer is
$1\,334\,961$ --- and $D_{10} > 10!/\eu$, as question 12 predicts
for even $n$.

\textbf{14.} $\abs{p_n - \eu^{-1}} = \abs{s_n - \eu^{-1}} <
\frac1{(n+1)!}$. For $n = 6$: $p_6 = 265/720 = 0.36806$ (five
decimals), against $\eu^{-1} = 0.36788$; the gap is below $1/7! =
1/5040 < 2 \times 10^{-4}$. The bound $1/(n+1)!$ collapses so fast
that the probability is fixed to many decimal places already for a
dozen letters: the answer ``about $36.8\%$'' is, for every
practical purpose, independent of $n$ --- the famous surprise of
the problem.

\textbf{15.} By question 2 and $D_m = m!\,s_m$:
\[
\frac{P_k(n)}{n!} = \frac{\binom nk D_{n-k}}{n!}
= \frac{D_{n-k}}{k!\,(n-k)!} = \frac{s_{n-k}}{k!}
\;\longrightarrow\; \frac{\eu^{-1}}{k!}
\]
as $n \to \infty$ with $k$ fixed, since $s_{n-k} \to \eu^{-1}$.
The limit values $\eu^{-1}/k!$ ($k \in \N$) are the weights of the
Poisson distribution of parameter $1$.

\textbf{16.} Count the triples $(\sigma, i, j)$ with $i \neq j$,
$\sigma(i) = i$, $\sigma(j) = j$. Choosing the ordered pair first:
$n(n-1)$ ways; the \omterm{def:b1:counting:objects}{permutations} fixing both $i$ and $j$ are the
\omterm{def:b1:counting:objects}{permutations} of the remaining $n - 2$ points: $(n-2)!$ of them.
Total: $n(n-1)(n-2)! = n!$. Summing instead over $\sigma$ first
counts, for each $\sigma$, the ordered pairs of distinct fixed
points: $\abs{\mathrm{Fix}(\sigma)}(\abs{\mathrm{Fix}(\sigma)}-1)$.
Hence the stated identity; dividing by $n!$, the average of
$\abs{\mathrm{Fix}}(\abs{\mathrm{Fix}} - 1)$ is $1$, so the average
of $\abs{\mathrm{Fix}}^2$ is $1 + 1 = 2$ and the variance is $2 -
1^2 = 1$.

\textbf{17.} The proportions $1 - p_n$: for $n = 4$, $1 - \frac
9{24} = \frac{15}{24} = 0.6250$; for $n = 5$, $1 - \frac{44}{120} =
\frac{76}{120} = 0.6333$; for $n = 6$, $1 - \frac{265}{720} =
\frac{455}{720} = 0.6319$. All within a percent of $1 - \eu^{-1}
\approx 0.6321$, oscillating around it.

\textbf{18.} Directly:
\[
s_{n+2} - s_n = \frac{(-1)^{n+1}}{(n+1)!} +
\frac{(-1)^{n+2}}{(n+2)!}
= (-1)^{n+1}\Bigl(\frac1{(n+1)!} - \frac1{(n+2)!}\Bigr),
\]
and the parenthesis is $> 0$. For $n$ even the difference is
negative: $s_{n+2} < s_n$, so $p_0 > p_2 > p_4 > \dots$; for $n$
odd it is positive: $p_1 < p_3 < p_5 < \dots$ Combined with
question 12 (evens above $\eu^{-1}$, odds below) and question 14
(distance to $\eu^{-1}$ tends to $0$): the two staircases squeeze
$\eu^{-1}$ between them.

\textbf{19.} One complete drawing is a uniform random \omterm{def:b1:counting:objects}{permutation},
valid when it is a derangement: probability $p_n \approx \eu^{-1}$.
By the quoted fact, the average number of drawings until success is
$1/p_n$, and question 14 gives $1/p_n \approx \eu$ up to an error
that is negligible already for small $n$. So a Secret Santa with
restarts costs on average about $\eu \approx 2.72$ complete
drawings --- whether the office has $6$ people or $600$.

\textbf{20.} Fix $j \geq 2$ and run question 5's classification on
the value $\sigma(1) = j$. If $\sigma(j) = 1$: the remaining
$n - 2$ points carry an arbitrary derangement, $D_{n-2}$ ways. If
$\sigma(j) \neq 1$: redirect the \omterm{def:b1:logic:map}{preimage} $i_0 = \sigma^{-1}(1)$ to
$j$ exactly as in question 5; this is a bijection with the
derangements of the $n - 1$ points $\{2, \dots, n\}$: $D_{n-1}$
ways. Total $D_{n-1} + D_{n-2}$, the same for every $j$. Summing
over the $n - 1$ values of $j$: $D_n = (n-1)(D_{n-1} + D_{n-2})$,
which displays the factor $n - 1$: $(n-1) \mid D_n$.

\textbf{21.} Claim: $D_n$ is odd iff $n$ is even. Induction using
$D_n = nD_{n-1} + (-1)^n$, i.e.\ $D_n \equiv nD_{n-1} + 1 \pmod 2$.
Base: $D_1 = 0$ is even, $n = 1$ odd: claim holds. If $n$ is even,
$nD_{n-1}$ is even and $D_n \equiv 1$: odd, as claimed. If $n$ is
odd, then $n - 1$ is even, so $D_{n-1}$ is odd by hypothesis, and
$D_n \equiv D_{n-1} + 1 \equiv 0$: even. The induction closes.

\textbf{22.} Reducing $D_n = nD_{n-1} + (-1)^n$ modulo $n$ kills
the first term: $D_n \equiv (-1)^n \pmod n$. For $n = 10$:
$(-1)^{10} = 1$, and indeed $D_{10} = 1\,334\,961$ ends in the
digit $1$.

\textbf{23.} For $n \geq 3$, $D_{n-1} \geq 1$ and division of
question 6's recurrence by $D_{n-1}$ gives $D_n/D_{n-1} = n +
(-1)^n/D_{n-1}$, with $\abs{(-1)^n/D_{n-1}} \leq 1$ and rapidly
tending to $0$. Consistency: if $D_n \approx n!/\eu$, then
$D_n/D_{n-1} \approx n!/(n-1)! = n$ --- the factor $\eu$ cancels
in the ratio, and the recurrence confirms it to accuracy
$1/D_{n-1}$.

\textbf{24.} (i) The product and sum rules underlie every count:
questions 2 and 5 \omterm{thm:b1:logic:partition}{partition} \omterm{def:b1:logic:sets}{sets} of \omterm{def:b1:counting:objects}{permutations} into independent
stages. (ii) Double counting gave the mean (question 4) and the
variance (question 16) of the number of fixed points with no
formula for $D_n$ at all. (iii) The binomial theorem evaluated the
alternating inner sum $(1-1)^{n-j}$ that makes binomial inversion
work (question 8). (iv) The alternating-series bound converted the
exact but opaque sum $n!\,s_n$ into the transparent \omterm{def:b1:logic:statement}{statement}
``nearest integer to $n!/\eu$'' (questions 10--14).

\textbf{25.} Inclusion--exclusion
(\cref{ex:b1:counting:derangement} and \cref{exo:b1:counting:11})
is the conceptual proof: it explains the alternating sum as
overcounting corrections, and it generalizes verbatim to counting
elements avoiding any family of ``bad'' \omterm{def:b1:logic:sets}{sets}. The recurrence route
(questions 5--7) computes fastest --- linear time, exact integer
arithmetic, no factorials --- and is the source of the arithmetic
facts of Part V. Binomial inversion (questions 8--9) places the
formula inside a general transform that will reappear wherever
two triangular systems of identities face each other. And the
appearance of $\eu$ is best explained by the formula itself: the
proportion of derangements is the partial sum $s_n$ of the series
for $\eu^{-1}$, so Montmort's envelopes were, three decades before
Euler's notation, already computing the number $\eu$.
\end{solution}
