\section*{Chapter \ref{ch:b1:functions} --- Standard Functions}

\begin{solution}{exo:b1:functions:1}
$\arcsin\frac{\sqrt 3}{2} = \frac\pi3$;
$\;\arccos\bigl(-\frac12\bigr) = \frac{2\pi}{3}$;
$\;\arctan(-1) = -\frac\pi4$.

$\arcsin\bigl(\sin\frac{5\pi}{6}\bigr)$: $\sin\frac{5\pi}{6} =
\frac12$, and the angle of $\intcc{-\frac\pi2}{\frac\pi2}$ with sine
$\frac12$ is $\frac\pi6$ (not $\frac{5\pi}{6}$).

$\arccos\bigl(\cos\frac{7\pi}{4}\bigr)$: $\cos\frac{7\pi}{4} =
\frac{\sqrt 2}{2}$, and the angle of $\intcc{0}{\pi}$ with that cosine
is $\frac\pi4$.
\end{solution}

\begin{solution}{exo:b1:functions:2}
$f(x) = \arcsin(2x - 1)$ requires $-1 \leq 2x - 1 \leq 1$, i.e.\ $x
\in \intcc{0}{1}$. On $\intoo{0}{1}$, the chain rule and
\cref{prop:b1:functions:arcderiv} give
\[
f'(x) = \frac{2}{\sqrt{1 - (2x-1)^2}} = \frac{2}{\sqrt{4x - 4x^2}}
= \frac{1}{\sqrt{x(1 - x)}} .
\]
$g(x) = \arctan\sqrt x$ is defined for $x \geq 0$, and for $x > 0$:
\[
g'(x) = \frac{1}{1 + x} \cdot \frac{1}{2\sqrt x}
= \frac{1}{2\sqrt x\,(1 + x)} .
\]
\end{solution}

\begin{solution}{exo:b1:functions:3}
$\cosh x + \sinh x = \frac{\eu^x + \eu^{-x}}{2} + \frac{\eu^x -
\eu^{-x}}{2} = \eu^x$. Hence for $n \in \Z$:
\[
(\cosh x + \sinh x)^n = (\eu^x)^n = \eu^{nx}
= \cosh nx + \sinh nx .
\]
\end{solution}

\begin{solution}{exo:b1:functions:4}
With $y = \arcsin x$: $\cos 2y = 1 - 2\sin^2 y = 1 - 2x^2$, so
$\cos(2\arcsin x) = 1 - 2x^2$.

With $y = \arctan x$: $\sin 2y = 2 \sin y \cos y = 2 \tan y \cos^2 y =
\frac{2\tan y}{1 + \tan^2 y}$, so $\sin(2\arctan x) =
\dfrac{2x}{1 + x^2}$.
\end{solution}

\begin{solution}{exo:b1:functions:5}
From slowest to fastest:
\[
(\ln x)^{1000} \;\ll\; x^{100} \;\ll\; x^{\ln x} \;\ll\;
\eu^{\sqrt x} \;\ll\; \eu^{x/100} .
\]
Justifications: $(\ln x)^{1000}/x^{100} \to 0$ by
\cref{prop:b1:functions:powerrules} (logs lose to powers).
$x^{100} = \eu^{100\ln x}$ and $x^{\ln x} = \eu^{(\ln x)^2}$: since
$(\ln x)^2 - 100\ln x \to +\infty$, the second wins. $x^{\ln x} =
\eu^{(\ln x)^2} \ll \eu^{\sqrt x}$ because $(\ln x)^2/\sqrt x \to 0$
(logs lose to the power $x^{1/4}$, squared). Finally $\sqrt x - x/100
\to -\infty$, so $\eu^{\sqrt x} \ll \eu^{x/100}$.
\end{solution}

\begin{solution}{exo:b1:functions:6}
Domain: $x \neq \pm 1$, three intervals. On each,
\[
f'(x) = \frac{\bigl(\frac{2x}{1-x^2}\bigr)'}{1 +
\frac{4x^2}{(1-x^2)^2}}
= \frac{\frac{2(1-x^2) + 4x^2}{(1-x^2)^2}}
{\frac{(1-x^2)^2 + 4x^2}{(1-x^2)^2}}
= \frac{2(1 + x^2)}{(1 + x^2)^2} = \frac{2}{1 + x^2}
= (2\arctan x)' .
\]
So $f(x) - 2\arctan x$ is constant on each interval. Values: at $x =
0$, $f(0) = 0$: the constant is $0$ on $\intoo{-1}{1}$. As $x \to
+\infty$, $\frac{2x}{1-x^2} \to 0^-$ so $f \to 0$, while $2\arctan x
\to \pi$: the constant is $-\pi$ on $\intoo{1}{+\infty}$. By oddness,
it is $+\pi$ on $\intoo{-\infty}{-1}$. Summary: $f = 2\arctan x$ on
$\intoo{-1}{1}$, $f = 2\arctan x - \pi$ for $x > 1$, $f = 2\arctan x +
\pi$ for $x < -1$. (This is the double-angle formula for the tangent,
read through $\arctan$.)
\end{solution}

\begin{solution}{exo:b1:functions:7}
$\cosh x = 2$: with $u = \eu^x > 0$, $u + u^{-1} = 4$, so $u^2 - 4u +
1 = 0$, $u = 2 \pm \sqrt 3$: $x = \ln(2 + \sqrt 3)$ or $x = \ln(2 -
\sqrt 3) = -\ln(2 + \sqrt 3)$ (the two solutions are opposite, as
$\cosh$ is even; both are valid). Equivalently $x = \pm
\operatorname{arcosh} 2$.

$5\cosh x - 4\sinh x = 3$: substituting the exponential definitions,
$\frac{5(u + u^{-1}) - 4(u - u^{-1})}{2} = 3$, i.e.\ $u + 9u^{-1} =
6$, i.e.\ $u^2 - 6u + 9 = (u - 3)^2 = 0$: $u = 3$, one solution $x =
\ln 3$.
\end{solution}

\begin{solution}{exo:b1:functions:8}
Let $\alpha = \arctan\frac12 + \arctan\frac13$. Addition formula:
\[
\tan\alpha = \frac{\frac12 + \frac13}{1 - \frac12\cdot\frac13}
= \frac{5/6}{5/6} = 1 .
\]
Both \omterm{def:b1:functions:arc}{arctangents} lie in $\intoo{0}{\frac\pi4}$ (their arguments are in
$\intoo{0}{1}$), so $\alpha \in \intoo{0}{\frac\pi2}$; the only angle
there with tangent $1$ is $\frac\pi4$.

Machin: let $\beta = \arctan\frac15$. Double angle twice:
\[
\tan 2\beta = \frac{2/5}{1 - 1/25} = \frac{5}{12},
\qquad
\tan 4\beta = \frac{2 \cdot 5/12}{1 - 25/144} = \frac{120}{119}.
\]
Then
\[
\tan\Bigl(4\beta - \frac\pi4\Bigr)
= \frac{\frac{120}{119} - 1}{1 + \frac{120}{119}}
= \frac{1/119}{239/119} = \frac{1}{239}.
\]
Location: $\beta < \arctan 1 = \frac\pi4$ and in fact $\tan 4\beta =
\frac{120}{119}$ close to $1$ with $4\beta \in \intoo{0}{\frac\pi2}$
(since $\beta < \frac\pi8$, as $\tan\frac\pi8 = \sqrt 2 - 1 > \frac15$);
so $4\beta - \frac\pi4 \in \intoo{-\frac\pi4}{\frac\pi4}$, where
$\arctan$ inverts $\tan$: $4\beta - \frac\pi4 =
\arctan\frac{1}{239}$, which is Machin's formula.
\end{solution}

\begin{solution}{exo:b1:functions:9}
Let $f(x) = x - \sin x$: $f(0) = 0$ and $f'(x) = 1 - \cos x \geq 0$,
so $f \geq 0$ on $\R_+$: $\sin x \leq x$.

Let $g(x) = \sin x - x + \frac{x^3}{6}$: $g(0) = 0$, $g'(x) = \cos x -
1 + \frac{x^2}{2}$, $g'(0) = 0$, $g''(x) = -\sin x + x = f(x) \geq 0$
on $\R_+$. So $g'$ is increasing with $g'(0) = 0$, hence $g' \geq 0$,
hence $g$ is increasing with $g(0) = 0$: $g \geq 0$, i.e.\ $x -
\frac{x^3}{6} \leq \sin x$ on $\R_+$.

Consequently $0 \leq \frac{x - \sin x}{x^3} \leq \frac 16$ for $x >
0$: the ratio is trapped near $\frac16$'s scale, and
\cref{ch:b1:taylor} shows its limit is exactly $\frac 16$.
\end{solution}

\begin{solution}{exo:b1:functions:10}
Write $y = \arcsin x$, so $2x\sqrt{1 - x^2} = 2\sin y\cos y = \sin
2y$ and $h(x) = \arcsin(\sin 2y) - 2y$.

For $\abs x \leq \frac{\sqrt 2}{2}$: $2y \in
\intcc{-\frac\pi2}{\frac\pi2}$, so $\arcsin(\sin 2y) = 2y$ and $h =
0$.

For $x \in \intoc{\frac{\sqrt 2}{2}}{1}$: $2y \in
\intoc{\frac\pi2}{\pi}$, and the angle of
$\intcc{-\frac\pi2}{\frac\pi2}$ with sine $\sin 2y$ is $\pi - 2y$:
so $h(x) = \pi - 4\arcsin x$. (Check by derivative: there
$h'(x) = \frac{2(1 - 2x^2)}{\abs{1 - 2x^2}\sqrt{1 - x^2}} -
\frac{2}{\sqrt{1-x^2}} = \frac{-4}{\sqrt{1 - x^2}}$, the derivative
of $-4\arcsin x$; and at $x = 1$, $h(1) = \arcsin 0 - 2\cdot\frac\pi2
= -\pi = \pi - 4\cdot\frac\pi2$.)

For $x \in \intco{-1}{-\frac{\sqrt 2}{2}}$, by oddness of $h$:
$h(x) = -\pi - 4\arcsin x$.
\end{solution}

\begin{solution}{exo:b1:functions:11}
$g = \arctan \circ \sinh$ is a composition of odd, strictly
increasing functions, so it is odd and strictly increasing; as $x \to
+\infty$, $\sinh x \to +\infty$ so $g(x) \to \frac\pi2$, and $g$ is a
bijection onto $\intoo{-\frac\pi2}{\frac\pi2}$ (continuity plus the
intermediate value theorem). Chain rule with
\cref{prop:b1:functions:hyprules}~(1):
\[
g'(x) = \frac{\cosh x}{1 + \sinh^2 x} = \frac{\cosh x}{\cosh^2 x}
= \frac{1}{\cosh x} .
\]
By construction $\tan g(x) = \sinh x$. Then, since $g(x) \in
\intoo{-\frac\pi2}{\frac\pi2}$ has positive cosine,
\begin{align*}
\cos g(x) &= \frac{1}{\sqrt{1 + \tan^2 g(x)}}
= \frac{1}{\sqrt{1 + \sinh^2 x}} = \frac{1}{\cosh x},\\
\sin g(x) &= \tan g(x) \cos g(x) = \frac{\sinh x}{\cosh x} = \tanh x .
\end{align*}
\end{solution}

\begin{solution}{exo:b1:functions:12}
Set $u = \arctan 2$ and $v = \arctan 3$; both lie in
$\intoo{\frac\pi4}{\frac\pi2}$ (their arguments exceed $1$), so $u
+ v \in \intoo{\frac\pi2}{\pi}$. The addition formula gives
\[
\tan(u + v) = \frac{2 + 3}{1 - 6} = -1 ,
\]
and the unique angle of $\intoo{\frac\pi2}{\pi}$ with tangent $-1$
is $\frac{3\pi}4$: so $\arctan 2 + \arctan 3 = \frac{3\pi}4$.
(Blindly applying $\arctan$ to the tangent would give
$-\frac\pi4$, off by $\pi$ --- the localization is what saves the
computation, and the negative denominator $1 - 6$ is precisely the
signal that the sum has left
$\intoo{-\frac\pi2}{\frac\pi2}$.) Adding $\arctan 1 = \frac\pi4$:
\[
\arctan 1 + \arctan 2 + \arctan 3 = \frac\pi4 + \frac{3\pi}4 = \pi .
\]
\end{solution}

\begin{solution}{pb:b1:functions:1}
\textbf{1.} $f$ is a quotient of differentiable functions with
nonvanishing denominator on $\intoo0{+\infty}$, and
\[
f'(t) = \frac{\frac1t \cdot t - \ln t}{t^2} = \frac{1 - \ln
t}{t^2} ,
\]
positive for $t < \eu$, zero at $t = \eu$, negative for $t > \eu$:
$f$ increases strictly on $\intoc0\eu$, decreases strictly on
$\intco\eu{+\infty}$, with maximum $f(\eu) = \frac1\eu$.

\textbf{2.} As $t \to 0^+$: $\ln t \to -\infty$ and $\frac1t \to
+\infty$, so $f(t) \to -\infty$. As $t \to +\infty$: $f(t) \to 0$
by the \omterm{prop:b1:functions:powerrules}{growth comparison} of \cref{prop:b1:functions:powerrules}
($\beta = \alpha = 1$). Sign: that of $\ln t$, so $f < 0$ on
$\intoo01$, $f(1) = 0$, $f > 0$ on $\intoo1{+\infty}$. The graph
climbs from $-\infty$, crosses zero at $1$, peaks at $(\eu,
\frac1\eu)$, then decays to $0^+$.

\textbf{3.} $f(4) = \frac{\ln 4}4 = \frac{2\ln 2}4 = \frac{\ln 2}2
= f(2)$.

\textbf{4.} By the variation table and the intermediate value
theorem (used at High School level; formalized in
\cref{ch:b1:continuity}). For $c < 0$: solutions exist only where
$f < 0$, i.e.\ in $\intoo01$ where $f$ is a strictly increasing
bijection onto $\intoo{-\infty}0$: exactly one. For $c = 0$: only
$t = 1$. For $0 < c < \frac1\eu$: on $\intoo1\eu$, $f$ increases
from $0$ to $\frac1\eu$: one solution; on $\intoo\eu{+\infty}$,
$f$ decreases from $\frac1\eu$ to $0$: one more; total two. For
$c = \frac1\eu$: only the maximum point $t = \eu$. For $c >
\frac1\eu$: none.

\textbf{5.} For $x \neq \eu$, strict maximality gives $f(x) <
f(\eu)$, i.e.\ $\frac{\ln x}x < \frac1\eu$, i.e.\ $\eu\ln x < x$,
i.e.\ $\ln(x^\eu) < \ln(\eu^x)$: $x^\eu < \eu^x$. With $x = \pi$:
$\eu^\pi > \pi^\eu$ (numerically $23.14 > 22.46$).

\textbf{6.} For $x, y > 0$, both sides are positive, so
\[
x^y = y^x \iff y\ln x = x\ln y \iff \frac{\ln x}x = \frac{\ln y}y
\iff f(x) = f(y) ,
\]
dividing by $xy > 0$.

\textbf{7.} Let $f(x) = f(y) = c$ with $x \neq y$. If $c \leq 0$ or
$c = \frac1\eu$, question 4 says the equation $f = c$ has a single
solution: impossible. So $0 < c < \frac1\eu$, and again by
question 4 the two solutions are one point of $\intoo1\eu$ and one
of $\intoo\eu{+\infty}$: both coordinates exceed $1$ and they
straddle $\eu$.

\textbf{8.} Substituting $y = tx$ in $y\ln x = x\ln y$:
\[
tx\ln x = x(\ln t + \ln x)
\iff (t - 1)\ln x = \ln t
\iff \ln x = \frac{\ln t}{t - 1} ,
\]
so $x = t^{1/(t-1)}$ and $y = tx = t^{1 + \frac1{t-1}} =
t^{t/(t-1)}$. Conversely, for these values, $\ln y = \frac{t\ln
t}{t-1} = t\ln x$ and $y = tx$, so $\frac{\ln y}y = \frac{t\ln
x}{tx} = \frac{\ln x}x$: a solution. Every nontrivial solution has
some ratio $t = y/x \in \intoo0{+\infty} \setminus \{1\}$, so the
parametrization is complete.

\textbf{9.} $t = 2$: $x = 2^{1/1} = 2$, $y = 2^{2/1} = 4$. And
\[
x(1/t) = (1/t)^{\frac1{\frac1t - 1}}
= (t^{-1})^{\frac{t}{1 - t}}
= t^{\frac{t}{t - 1}} = y(t) ,
\]
then $y(1/t) = \frac1t\,x(1/t) = \frac{y(t)}t = x(t)$: the
parameter change $t \mapsto 1/t$ swaps the coordinates, as the
symmetry of the equation demands.

\textbf{10.} As $t \to 1$: $\frac{\ln t}{t-1} \to 1$ (it is the
difference quotient of $\ln$ at $1$), so $x \to \eu^1 = \eu$ and
$y = tx \to \eu$. As $t \to +\infty$: $\ln x = \frac{\ln t}{t-1}
\to 0$ so $x \to 1$, while $\ln y = \frac{t\ln t}{t-1} \to
+\infty$ so $y \to +\infty$. As $t \to 0^+$: $\ln x = \frac{\ln
t}{t-1} \to \frac{-\infty}{-1} = +\infty$ so $x \to +\infty$,
while $\ln y = \frac{t\ln t}{t-1} \to \frac{0}{-1} = 0$ so $y \to
1$ (using $t\ln t \to 0$, \cref{ex:b1:functions:xx}). The branch
therefore runs from the asymptote $y = 1$ (far right), up through
$(\eu, \eu)$ on the diagonal, and off along the asymptote $x = 1$
(far top) --- symmetric about the diagonal by question 9.

\textbf{11.} $\ln x(t) = g(t) = \frac{\ln t}{t-1}$, and
\[
g'(t) = \frac{\frac{t-1}t - \ln t}{(t-1)^2}
= \frac{h(t)}{(t-1)^2},
\qquad h(t) = 1 - \frac1t - \ln t .
\]
$h(1) = 0$ and $h'(t) = \frac1{t^2} - \frac1t = \frac{1 -
t}{t^2} < 0$ for $t > 1$: $h < 0$ on $\intoo1{+\infty}$, so $g' <
0$ and $x = \eu^g$ is strictly decreasing there (from $\eu$ to
$1$, by question 10).

\textbf{12.} By question 11, $t \mapsto x(t)$ is a strictly
decreasing bijection from $\intoo1{+\infty}$ onto $\intoo1\eu$;
write $t(x)$ for its inverse (also strictly decreasing) and set
$\varphi(x) = y(t(x))$. Note $h < 0$ also on $\intoo01$ (there
$h' > 0$ and $h(1) = 0$), so $x(\cdot)$ is decreasing on
$\intoo01$ as well; hence $y(t) = x(1/t)$ (question 9) is
\emph{increasing} in $t$ on $\intoo1{+\infty}$, from $\eu$ to
$+\infty$. Composing: $\varphi = y \circ t(\cdot)$ is strictly
decreasing from $\intoo1\eu$ onto $\intoo\eu{+\infty}$, and
$x^{\varphi(x)} = \varphi(x)^x$ by question 8. Finally, for a
common value $c = f(x) \in \intoo0{\frac1\eu}$, the pair $\{x,
\varphi(x)\}$ is \emph{the} two-element solution \omterm{def:b1:logic:sets}{set} of $f = c$
(question 4); extending $\varphi$ to $\intoo\eu{+\infty}$ as the
inverse \omterm{def:b1:logic:map}{map}, $\varphi(\varphi(x))$ returns to the other (i.e.\
original) element: $\varphi \circ \varphi = \mathrm{id}$.

\textbf{13.} If $(x, y)$ is a nontrivial integer solution with $x
< y$, question 7 puts $x \in \intoo1\eu$: the only integer there
is $x = 2$. Then $f(y) = f(2) = \frac{\ln2}2$ with $y > \eu$; by
question 4 the equation $f = \frac{\ln2}2$ has exactly one
solution beyond $\eu$, and question 3 exhibits it: $y = 4$. Hence
$(2, 4)$, and its swap, are the only ones.

\textbf{14.} $t = 1 + \frac1n = \frac{n+1}n$ gives $\frac1{t - 1}
= n$ and $\frac t{t-1} = n + 1$, so
\[
x_n = \Bigl(\frac{n+1}n\Bigr)^{n},
\qquad
y_n = \Bigl(\frac{n+1}n\Bigr)^{n+1},
\]
manifestly rational, distinct ($y_n = t\,x_n \neq x_n$), and $n =
1$ gives $x_1 = 2$, $y_1 = 4$.

\textbf{15.} $t = y/x$ is rational and $> 1$; write $t - 1 =
\frac rs$ in lowest terms, so $t = \frac{s + r}s$ with $\gcd(s +
r, s) = \gcd(r, s) = 1$. Question 8 gives $x = t^{s/r}$, hence
\[
x^r = t^s = \frac{(s+r)^s}{s^s} ,
\]
a fraction in lowest terms (no prime divides both $s + r$ and
$s$). Writing $x = \frac pq$ in lowest terms, $x^r = \frac{p^r}
{q^r}$ is also in lowest terms, and by uniqueness of the reduced
representation: $p^r = (s+r)^s$ and $q^r = s^s$. Compare the
exponent of any prime $\ell$ in $q^r = s^s$: $r \cdot
v_\ell(q) = s \cdot v_\ell(s)$, so $r$ divides $s\,v_\ell(s)$;
since $\gcd(r, s) = 1$, $r$ divides $v_\ell(s)$ for every prime
$\ell$ (unique factorization, admitted; proved in
\cref{ch:b1:arith}), so $s = b^r$ for an integer $b$. The same
argument on $p^r = (s+r)^s$ gives $s + r = a^r$.

\textbf{16.} Suppose $a^r - b^r = r$ with integers $a > b \geq 1$
and $r \geq 2$. Then $a \geq b + 1$, so by the binomial theorem
\[
a^r - b^r \geq (b+1)^r - b^r
= \sum_{k=0}^{r-1}\binom rk b^k
\geq \binom r0 + \binom r{r-1} b^{r-1}
\geq 1 + r > r ,
\]
a contradiction. From question 15, $a^r - b^r = (s + r) - s = r$
forces $r = 1$: $t = 1 + \frac1s$, and the solution is $(x_s,
y_s)$ of question 14. Together with the swapped pairs, the
classification is complete.

\textbf{17.} $f\bigl(\tfrac94\bigr) = \frac{\ln 2.25}{2.25} =
\frac{0.81093}{2.25} = 0.3604$ and $f\bigl(\tfrac{27}8\bigr) =
\frac{\ln 3.375}{3.375} = \frac{1.21640}{3.375} = 0.3604$ (four
decimals): equal, as the construction promises --- so
$\bigl(\tfrac94\bigr)^{27/8} = \bigl(\tfrac{27}8\bigr)^{9/4}$.

\textbf{18.} The parameters $t_n = 1 + \frac1n$ decrease strictly
to $1$. Since $x(\cdot)$ is strictly decreasing on
$\intoo1{+\infty}$ (question 11), $x_n = x(t_n)$ is strictly
\emph{increasing}; since $y(\cdot)$ is strictly increasing there
(question 12), $y_n = y(t_n)$ is strictly \emph{decreasing}. By
question 7, $x_n \in \intoo1\eu$ and $y_n \in
\intoo\eu{+\infty}$: so $x_n < \eu < y_n$ for every $n$. Finally
$t_n \to 1$ and question 10 give $x_n \to \eu$ and $y_n \to \eu$.
The classical monotone convergence of $\bigl(1 +
\frac1n\bigr)^n$ falls out of the geometry of the branch, with no
new inequality.

\textbf{19.} If $\eu \leq a < b$: $f$ strictly decreasing on
$\intco\eu{+\infty}$ gives $f(a) > f(b)$, i.e.\ $b\ln a > a\ln
b$, i.e.\ $a^b > b^a$. If $1 < a < b \leq \eu$: $f$ strictly
increasing gives $f(a) < f(b)$, hence $a^b < b^a$. Mixed case:
$2 < \eu < 3$ and $2^3 = 8 < 9 = 3^2$; but $2 < \eu < 5$ and
$2^5 = 32 > 25 = 5^2$: both outcomes occur, decided by which side
of the branch the point $(a, b)$ falls.

\textbf{20.} The branch meets the diagonal at $\eu$ and $\varphi$
extends continuously there with $\varphi(\eu) = \eu$.
Differentiating $\varphi(\varphi(x)) = x$ at $x = \eu$ by the
chain rule: $\varphi'(\varphi(\eu))\,\varphi'(\eu) =
\varphi'(\eu)^2 = 1$, so $\varphi'(\eu) = \pm1$; $\varphi$ is
decreasing, so $\varphi'(\eu) = -1$. The branch crosses the
diagonal with slope $-1$: perpendicularly.

\textbf{21.} $y = 3x$ is the case $t = 3$: $x = 3^{1/2} = \sqrt3$
and $y = 3^{3/2} = 3\sqrt3$. Check: $f(\sqrt3) =
\frac{0.54931}{1.73205} = 0.3171$ and $f(3\sqrt3) =
\frac{\ln 5.19615}{5.19615} = \frac{1.64792}{5.19615} = 0.3171$:
equal to four decimals, so $(\sqrt3)^{3\sqrt3} =
(3\sqrt3)^{\sqrt3}$.

\textbf{22.} $n^{1/n} = \eu^{f(n)}$, and $\exp$ is increasing, so
the ordering is that of $f(n)$: $f(3) = \frac{\ln3}3 \approx
0.3662$ beats $f(2) = f(4) = \frac{\ln2}2 \approx 0.3466$ (question
3) and $f(5) \approx 0.3219$. The largest is $\sqrt[3]3$, and the
equal pair is $\sqrt2 = \sqrt[4]4$ --- the integer pair $(2, 4)$
in yet another disguise.

\textbf{23.} The solution \omterm{def:b1:logic:sets}{set} is the union of the diagonal $\{(x,
x) : x > 0\}$ and a single branch, symmetric about the diagonal,
strictly decreasing from the asymptote $x = 1$ (as $y \to
+\infty$) to the asymptote $y = 1$ (as $x \to +\infty$), crossing
the diagonal exactly once, at $(\eu, \eu)$, with slope $-1$
there. On the branch sit the integer points $(2, 4)$ and $(4, 2)$
--- the only ones --- and the rational points $(x_n, y_n)$,
which march monotonically along the branch toward $(\eu, \eu)$
without ever reaching it ($\eu$ is irrational; the branch's
rational points accumulate at an irrational corner).

\textbf{24.} (i) \omterm{prop:b1:functions:powerrules}{Growth comparisons} gave the limits of $f$ at
$+\infty$ (question 2) and $t\ln t \to 0$ (question 10),
shaping both the variation table and the asymptotes. (ii) The
intermediate value theorem, through the bijection \omterm{def:b1:logic:statement}{statements},
converted the variation table into exact solution counts
(question 4) and into the existence of the inverse \omterm{def:b1:logic:map}{map} $t(x)$
(question 12). (iii) Unique factorization powered the two
lowest-terms identifications of question 15, the arithmetic heart
of the rational-point classification.

\textbf{25.} One derivative computation --- the sign of $1 - \ln
t$ --- generated everything: the solution count for each level
$c$, the shape and asymptotes of the branch, the extremal
inequality $\eu^x \geq x^\eu$, the classification of integer and
rational solutions, and the monotone convergence of $\bigl(1 +
\frac1n\bigr)^n$. This is the economy of thinking with variation
tables: a one-dimensional study resolves a two-variable equation
because the equation factors through a single function.
\cref{ch:b1:seq} will redo the convergence of $(x_n)$ with the
general theory of monotone sequences; \cref{ch:b1:derivative}
founds variation tables rigorously on the mean value theorem; and
\cref{ch:b1:taylor} supplies what the table cannot --- the
\emph{speed} of the convergence $x_n \to \eu$ (it is of order
$1/n$).
\end{solution}
