\section*{Chapter \ref{ch:b1:structures} --- Algebraic Structures}

\begin{solution}{exo:b1:structures:1}
\emph{Stability:} $x * y = 1 \iff x + y - xy = 1 \iff (1-x)(1-y) = 0$,
impossible for $x, y \neq 1$. Indeed the key identity is
\[
1 - x * y = (1 - x)(1 - y):
\]
the \omterm{def:b1:logic:map}{map} $\varphi(x) = 1 - x$ sends $(E, *)$ to $(\R^*, \times)$ with
$\varphi(x * y) = \varphi(x)\varphi(y)$ --- a \omterm{def:b1:logic:inj}{bijective} \omterm{def:b1:structures:morphism}{morphism}. All
axioms now transport: associativity and commutativity follow from
those of $\times$; the identity is $\varphi^{-1}(1) = 0$ (check: $x *
0 = x$); the inverse of $x$ is $\varphi^{-1}\bigl((1-x)^{-1}\bigr) =
1 - \frac{1}{1-x} = \frac{x}{x - 1}$ (which is $\neq 1$). So $(E, *)$
is an \omterm{def:b1:structures:group}{abelian group}.
\end{solution}

\begin{solution}{exo:b1:structures:2}
\begin{enumerate}
  \item Yes: product of positives is positive, identity $1$, inverse
        $\frac 1x$, associativity inherited from $\R^*$.
  \item No: not stable ($1 + 1 = 2 \notin \{-1,0,1\}$).
  \item Yes: the standard example.
  \item No: not stable (odd $+$ odd $=$ even), and no identity ($0$ is
        even).
\end{enumerate}
\end{solution}

\begin{solution}{exo:b1:structures:3}
Write $\mathrm{id}$, the transpositions $\tau_{12}, \tau_{13},
\tau_{23}$ (swapping the two named points), and the cycles $c =
(1\,2\,3)$ (i.e.\ $1 \mapsto 2 \mapsto 3 \mapsto 1$) and $c^2 =
(1\,3\,2)$. The table of $\sigma\rho$ (row $\sigma$, column $\rho$,
apply $\rho$ first):
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{c|cccccc}
$\sigma\backslash\rho$ & $\mathrm{id}$ & $c$ & $c^2$ & $\tau_{12}$ &
$\tau_{13}$ & $\tau_{23}$ \\
\hline
$\mathrm{id}$ & $\mathrm{id}$ & $c$ & $c^2$ & $\tau_{12}$ &
$\tau_{13}$ & $\tau_{23}$ \\
$c$ & $c$ & $c^2$ & $\mathrm{id}$ & $\tau_{13}$ & $\tau_{23}$ &
$\tau_{12}$ \\
$c^2$ & $c^2$ & $\mathrm{id}$ & $c$ & $\tau_{23}$ & $\tau_{12}$ &
$\tau_{13}$ \\
$\tau_{12}$ & $\tau_{12}$ & $\tau_{23}$ & $\tau_{13}$ & $\mathrm{id}$
& $c^2$ & $c$ \\
$\tau_{13}$ & $\tau_{13}$ & $\tau_{12}$ & $\tau_{23}$ & $c$ &
$\mathrm{id}$ & $c^2$ \\
$\tau_{23}$ & $\tau_{23}$ & $\tau_{13}$ & $\tau_{12}$ & $c^2$ & $c$ &
$\mathrm{id}$ \\
\end{tabular}
\end{center}
Non-commuting pair: $\tau_{12}\tau_{13} = c^2$ while
$\tau_{13}\tau_{12} = c$. (To check one entry:
$\tau_{12}\tau_{13}$ sends $1 \xmapsto{\tau_{13}} 3
\xmapsto{\tau_{12}} 3$, $3 \mapsto 1 \mapsto 2$, $2 \mapsto 2 \mapsto
1$: that is $1 \mapsto 3 \mapsto 2 \mapsto 1$, the cycle $c^2 =
(1\,3\,2)$.)
\end{solution}

\begin{solution}{exo:b1:structures:4}
$H$: $1 \in H$; for $z, w \in H$, $\abs{zw^{-1}} = \abs z / \abs w =
1$: the criterion applies. $\R_+^*$: same, $\abs{xy^{-1}}$ replaced by
positivity. Union: $\iu \in H$ and $2 \in \R_+^*$, but $2\iu$ has
\omterm{def:b1:complex:field}{modulus} $2 \neq 1$ and is not a positive real: $2\iu \notin H \cup
\R_+^*$, so the union is not stable --- not a \omterm{def:b1:structures:subgroup}{subgroup} (as predicted
by \cref{exo:b1:structures:6}, neither \omterm{def:b1:structures:subgroup}{subgroup} contains the other).
\end{solution}

\begin{solution}{exo:b1:structures:5}
\omterm{def:b1:structures:morphism}{Morphism}: $\eu^{\iu(\theta + \varphi)} =
\eu^{\iu\theta}\eu^{\iu\varphi}$ (\cref{thm:b1:complex:funceq}).
\omterm{def:b1:structures:morphism}{Kernel}: $\eu^{\iu\theta} = 1 \iff \theta \in 2\pi\Z$, so $\ker f =
2\pi\Z \neq \{0\}$: not \omterm{def:b1:logic:inj}{injective}. Image: every unit complex number is
$\eu^{\iu\theta}$ for some $\theta$ (polar form), so
$\operatorname{im} f = \mathbb{U}$, the unit circle. The restriction
of $f$ to a half-open interval of length $2\pi$, such as
$\intco{0}{2\pi}$ or $\intoc{-\pi}{\pi}$, is \omterm{def:b1:logic:inj}{injective} (two angles
with the same image differ by a multiple of $2\pi$, and only one
representative of each class fits in the interval); no interval of
greater length works, since it contains two points at distance
$2\pi$.
\end{solution}

\begin{solution}{exo:b1:structures:6}
Intersection: $e \in H \cap K$, and $x, y \in H \cap K$ gives
$xy^{-1}$ in both $H$ and $K$. Union: if $H \subseteq K$ the union is
$K$, a \omterm{def:b1:structures:subgroup}{subgroup} (and symmetrically). Conversely, suppose neither
inclusion holds: pick $h \in H \setminus K$ and $k \in K \setminus
H$, and suppose $H \cup K$ were a \omterm{def:b1:structures:subgroup}{subgroup}; then $hk \in H \cup K$.
If $hk \in H$, then $k = h^{-1}(hk) \in H$: contradiction. If $hk \in
K$, then $h = (hk)k^{-1} \in K$: contradiction. So $H \cup K$ is not
a \omterm{def:b1:structures:subgroup}{subgroup}.
\end{solution}

\begin{solution}{exo:b1:structures:7}
Note first that $x^2 = e$ means $x^{-1} = x$ for every $x$. Then for
$x, y \in G$:
\[
xy = (xy)^{-1} = y^{-1} x^{-1} = yx ,
\]
using \cref{prop:b1:structures:rules}~(2). So $G$ is \omterm{def:b1:structures:group}{abelian}.
\end{solution}

\begin{solution}{exo:b1:structures:8}
Units of $\Z/18\Z$: classes \omterm{cor:b1:arith:bezout}{coprime} to $18 = 2 \times 3^2$:
$\overline 1, \overline 5, \overline 7, \overline{11}, \overline{13},
\overline{17}$. Inverse of $\overline 5$: $5 \times 11 = 55 = 3\times
18 + 1$, so $\overline 5^{-1} = \overline{11}$.

$\overline 5 x = \overline 7$: multiply by $\overline{11}$: $x =
\overline{77} = \overline 5$ (since $77 = 4\times 18 + 5$). Unique
solution.

$\overline 6 x = \overline 3$: the equation $6x \equiv 3 \pmod{18}$
means $18 \mid 6x - 3$. But $6x - 3 = 3(2x - 1)$ is odd, while $18$
is even: an even number cannot divide an odd one. No solution.

$\overline 6 x = \overline{12}$: $6x \equiv 12 \pmod{18} \iff x
\equiv 2 \pmod 3$: solutions $x \in \{\overline 2, \overline 5,
\overline 8, \overline{11}, \overline{14}, \overline{17}\}$ --- six
of them.
\end{solution}

\begin{solution}{exo:b1:structures:9}
$\Z[\sqrt 2]$ contains $0$ and $1$, and is stable under subtraction
and product:
\[
(a + b\sqrt 2)(c + d\sqrt 2) = (ac + 2bd) + (ad + bc)\sqrt 2 ,
\]
so it is a subring of $\R$ (commutativity, associativity,
distributivity are inherited). Unit: $(1 + \sqrt 2)(-1 + \sqrt 2) =
2 - 1 = 1$, so $1 + \sqrt 2$ is invertible with inverse $\sqrt 2 - 1
\in \Z[\sqrt 2]$. Its powers $(1 + \sqrt 2)^n$ are strictly
increasing (the base is $> 1$), hence pairwise distinct, and each is
a unit ($\bigl((1+\sqrt2)^n\bigr)^{-1} = (\sqrt 2 - 1)^n$): the \omterm{def:b1:structures:group}{group}
of units is infinite.
\end{solution}

\begin{solution}{exo:b1:structures:10}
$x + x = (x + x)^2 = x^2 + x^2 + x^2 + x^2 = 4x^2 = 4x$ --- so $2x =
4x$, giving $2x = 0$, i.e.\ $x + x = 0$ (each element is its own
additive inverse). Then
\[
x + y = (x+y)^2 = x^2 + xy + yx + y^2 = x + xy + yx + y ,
\]
so $xy + yx = 0$, i.e.\ $xy = -yx = yx$ (using $-z = z$). Hence $A$ is
commutative.

Example: on $\mathcal{P}(E)$, define $A + B = (A \cup B) \setminus (A
\cap B)$ (symmetric difference) and $A \times B = A \cap B$. One
checks: $(\mathcal{P}(E), +)$ is an \omterm{def:b1:structures:group}{abelian group} with identity
$\emptyset$ and each \omterm{def:b1:logic:sets}{set} its own inverse; $\cap$ is associative,
commutative, with identity $E$; distributivity $A \cap (B + C) = (A
\cap B) + (A \cap C)$ holds (an element lies in the left side iff it
is in $A$ and in exactly one of $B, C$). And $A \cap A = A$: every
element is idempotent, as required.
\end{solution}

\begin{solution}{exo:b1:structures:11}
Write the hypothesis for $n$ and $n+1$:
\[
(xy)^{n+1} = x^{n+1} y^{n+1}
\quad\text{and}\quad
(xy)^{n+1} = (xy)(xy)^n = xy\,x^n y^n .
\]
Equating: $x^{n+1} y^{n+1} = x\,y\,x^n\,y^n$; cancel $x$ on the left
and $y^n$ on the right: $x^n y = y x^n$. The same computation one
degree higher ($n+1$ and $n+2$) gives $x^{n+1} y = y x^{n+1}$. Then
\[
y\,x^{n+1} = x^{n+1} y = x\,(x^n y) = x\,y\,x^n ,
\]
and cancelling $x^n$ on the right of $y x \cdot x^n = x y \cdot x^n$:
$yx = xy$. So $G$ is \omterm{def:b1:structures:group}{abelian}.
\end{solution}

\begin{solution}{exo:b1:structures:12}
\begin{enumerate}
  \item Let $f \colon \Z \to \Z$ be additive and $a = f(1)$. By
        induction $f(k) = ka$ for $k \in \N$, and $f(-k) = -f(k) =
        -ka$: so $f$ is the multiplication by $a$. Conversely
        every \omterm{def:b1:logic:map}{map} $k \mapsto ak$ is a \omterm{def:b1:structures:morphism}{morphism}: the \omterm{def:b1:structures:morphism}{morphisms}
        $(\Z,+) \to (\Z,+)$ are exactly the multiplications by a
        fixed integer.
  \item Let $f \colon \Q \to \Z$ be a \omterm{def:b1:structures:morphism}{morphism}, $x \in \Q$ and
        $n \in \N^*$. Then
        \[
        f(x) = f\Bigl(\underbrace{\tfrac xn + \dots +
        \tfrac xn}_{n}\Bigr) = n\,f\Bigl(\frac xn\Bigr) ,
        \]
        so the integer $f(x)$ is divisible by every $n \geq 1$.
        The only such integer is $0$: $f \equiv 0$.
\end{enumerate}
\end{solution}

\begin{solution}{pb:b1:structures:1}
\textbf{1.} A \omterm{def:b1:counting:objects}{permutation} is a bijection of $\intint1n$, i.e.\ an
$n$-arrangement of $n$ objects: there are $n!$ of them
(\cref{thm:b1:counting:counts}). With $\sigma = [2,3,1]$, $\tau =
[1,3,2]$: $\sigma\tau$ sends $1 \mapsto 1 \mapsto 2$, $2 \mapsto 3
\mapsto 1$, $3 \mapsto 2 \mapsto 3$: $\sigma\tau = [2,1,3]$; and
$\tau\sigma$ sends $1 \mapsto 2 \mapsto 3$, $2 \mapsto 3 \mapsto
2$, $3 \mapsto 1 \mapsto 1$: $\tau\sigma = [3,2,1] \neq
\sigma\tau$.

\textbf{2.} Let $\gamma, \gamma'$ have disjoint supports $S, S'$.
For $x \in S$: $\gamma'(x) = x$ and $\gamma(x) \in S$, so
$\gamma\gamma'(x) = \gamma(x) = \gamma'\gamma(x)$. Symmetrically
for $x \in S'$; and both sides fix every $x \notin S \cup S'$. So
$\gamma\gamma' = \gamma'\gamma$.

\textbf{3.} For $i \in \intint1n$, the values $i, \sigma(i),
\sigma^2(i), \dots$ live in a \omterm{def:b1:counting:card}{finite set}, so $\sigma^a(i) =
\sigma^b(i)$ for some $a < b$; injectivity gives $\sigma^{b-a}(i)
= i$: the sequence returns to $i$. Call \emph{orbit} of $i$ the
set $\{i, \sigma(i), \dots, \sigma^{k-1}(i)\}$ with $k \geq 1$
minimal such that $\sigma^k(i) = i$. Two orbits meeting in one
point coincide (both are the forward $\sigma$-images of that
point), so the orbits \omterm{thm:b1:logic:partition}{partition} $\intint1n$; $\sigma$ acts on each
orbit of size $k \geq 2$ as the $k$-cycle $(i\ \sigma(i)\ \cdots\
\sigma^{k-1}(i))$ and fixes the singletons. The product of these
disjoint cycles agrees with $\sigma$ everywhere. Uniqueness: in
any decomposition into disjoint cycles, the cycle through $i$ must
be $(i\ \sigma(i)\ \cdots)$ --- the cycles are forced to be the
orbits with their induced action.

\textbf{4.} Following the orbits: $1 \to 4 \to 2 \to 1$, $3 \to 5
\to 3$, $6 \to 7 \to 8 \to 6$:
\[
\sigma = (1\ 4\ 2)(3\ 5)(6\ 7\ 8) .
\]
If $\sigma = \gamma_1\cdots\gamma_r$ with disjoint cycles of
lengths $k_1, \dots, k_r$, commutation (question 2) gives
$\sigma^m = \gamma_1^m\cdots\gamma_r^m$, and since the supports
are disjoint, $\sigma^m = \mathrm{id}$ iff each $\gamma_i^m =
\mathrm{id}$ iff $k_i \mid m$ for all $i$ (a $k$-cycle has \omterm{def:b1:structures:order}{order}
$k$: $\gamma^m$ sends $a_1$ to $a_{1 + (m \bmod k)}$). The least
such $m$ is $\operatorname{lcm}(k_1, \dots, k_r)$. Here:
$\operatorname{lcm}(3, 2, 3) = 6$.

\textbf{5.} Apply the right-hand side to each point, rightmost
factor first. $a_1 \mapsto a_2$ by $(a_1\ a_2)$, then every later
factor fixes $a_2$: net $a_1 \mapsto a_2$. For $2 \leq i < k$:
$a_i$ is untouched until $(a_1\ a_i)$ sends it to $a_1$, and the
very next factor $(a_1\ a_{i+1})$ sends $a_1$ to $a_{i+1}$, after
which nothing moves it: net $a_i \mapsto a_{i+1}$. Finally $a_k$
is fixed by all factors but the leftmost, which sends it to $a_1$.
This is exactly the cycle. Since every \omterm{def:b1:counting:objects}{permutation} is a product of
cycles (question 3), it is a product of transpositions. For
question 4's $\sigma$:
\[
\sigma = (1\ 2)(1\ 4)\;(3\ 5)\;(6\ 8)(6\ 7),
\]
five transpositions.

\textbf{6.} Induction on $b - a$. For $b = a + 1$ the identity is
trivial ($1 = 2\cdot1 - 1$ factor). For $b > a + 1$, verify
directly that $(a\ b) = (a\ \ a{+}1)\,(a{+}1\ \ b)\,(a\ \ a{+}1)$:
the right side sends $a \mapsto a{+}1 \mapsto b \mapsto b$, $b
\mapsto b \mapsto a{+}1 \mapsto a$, $a{+}1 \mapsto a \mapsto a
\mapsto a{+}1$, and fixes the rest. By induction $(a{+}1\ \ b)$ is
a palindromic product of $2(b - a - 1) - 1$ adjacent
transpositions, so $(a\ b)$ is one of $2(b - a) - 1$: an odd
number.

\textbf{7.} $N(\mathrm{id}) = 0$, $\varepsilon = +1$. For $(i\ \
i{+}1)$, the only inverted pair is $(i, i+1)$: $N = 1$,
$\varepsilon = -1$. For $[2, 3, 1]$: the inverted pairs are $(1,
3)$ (values $2 > 1$) and $(2, 3)$ (values $3 > 1$): $N = 2$,
$\varepsilon = +1$.

\textbf{8.} The value lists of $\sigma$ and $\sigma\tau$ differ
only by the swap of positions $i$ and $i + 1$. For a pair of
positions not involving $i, i+1$, nothing changes. For $k < i$,
the two pairs $(k, i)$ and $(k, i+1)$ exchange their inversion
statuses (the same two values are compared with $\sigma(k)$, in
the other \omterm{def:b1:structures:order}{order} of positions): their total contribution is
unchanged; likewise for $k > i + 1$. The one remaining pair $(i,
i+1)$ flips its status. Hence $N(\sigma\tau) = N(\sigma) \pm 1$.

\textbf{9.} Let $\tau = (a\ b)$ be any transposition: by
question 6 it is a product of an odd number of adjacent
transpositions, so right-multiplying by $\tau$ changes $N$ by an
odd total (question 8, applied repeatedly): $\varepsilon(\sigma
\tau) = -\varepsilon(\sigma)$. Now if $\sigma = \tau_1\cdots
\tau_p$ (transpositions), build it up from the identity by $p$
right-multiplications: $\varepsilon(\sigma) =
(-1)^p\varepsilon(\mathrm{id}) = (-1)^p$. Since
$\varepsilon(\sigma)$ is defined by inversions --- independently
of any factorization --- the parity of $p$ is an invariant of
$\sigma$. \omterm{def:b1:structures:morphism}{Morphism}: writing $\sigma$ with $p$ and $\sigma'$ with
$q$ transpositions, $\sigma\sigma'$ uses $p + q$ of them:
$\varepsilon(\sigma\sigma') = (-1)^{p+q} =
\varepsilon(\sigma)\varepsilon(\sigma')$.

\textbf{10.} A $k$-cycle is a product of $k - 1$ transpositions
(question 5): $\varepsilon = (-1)^{k-1}$. For general $\sigma$
with orbits of sizes $k_1, \dots, k_r$ ($k_i \geq 2$) plus $f$
fixed points, $c(\sigma) = r + f$ and $n = k_1 + \dots + k_r + f$,
so
\[
\varepsilon(\sigma) = \prod_{i=1}^r (-1)^{k_i - 1}
= (-1)^{\sum_i k_i - r} = (-1)^{n - f - r} = (-1)^{n -
c(\sigma)} .
\]

\textbf{11.} $\mathfrak A_n = \ker\varepsilon$ is a \omterm{def:b1:structures:subgroup}{subgroup} as
the \omterm{def:b1:structures:morphism}{kernel} of a \omterm{def:b1:structures:morphism}{morphism} (\cref{def:b1:structures:morphism}). Fix
a transposition $\tau_0$ (exists for $n \geq 2$). The \omterm{def:b1:logic:map}{map} $\sigma
\mapsto \sigma\tau_0$ is a bijection of $\mathfrak S_n$ (its own
inverse) exchanging $\mathfrak A_n$ with the \omterm{def:b1:logic:sets}{set} of odd
\omterm{def:b1:counting:objects}{permutations} (question 9). The two \omterm{def:b1:logic:sets}{sets} \omterm{thm:b1:logic:partition}{partition} $\mathfrak S_n$
and have equal size: $\abs{\mathfrak A_n} = \frac{n!}2$.

\textbf{12.} \emph{Inversions} of $[4, 1, 5, 2, 3, 7, 8, 6]$: from
value $4$: over $1, 2, 3$: three; from $5$: over $2, 3$: two;
from $7$: over $6$: one; from $8$: over $6$: one. $N = 7$,
$\varepsilon = -1$. \emph{Cycle type}: $c = 3$ orbits, $n = 8$:
$\varepsilon = (-1)^{8-3} = -1$. \emph{Transposition count}: five
transpositions in question 5: $(-1)^5 = -1$. All three agree.

\textbf{13.} $(a\ b)(a\ c)$ (rightmost first): $a \mapsto c
\mapsto c$; $c \mapsto a \mapsto b$; $b \mapsto b \mapsto a$:
the $3$-cycle $(a\ c\ b)$. And $(a\ c\ b)(a\ c\ d)$: $a \mapsto c
\mapsto b$; $b \mapsto b \mapsto a$; $c \mapsto d \mapsto d$; $d
\mapsto a \mapsto c$: that is $(a\ b)(c\ d)$, as claimed.

\textbf{14.} Let $\sigma \in \mathfrak A_n$: by question 9,
$\sigma = \tau_1\cdots\tau_{2m}$ with an even number of
transpositions. Group them in consecutive pairs
$\tau_{2i-1}\tau_{2i}$: if the two are equal, the pair is the
identity and disappears; if they share exactly one point, the
first identity of question 13 writes the pair as one $3$-cycle;
if they are disjoint, the second identity writes it as two
$3$-cycles. Hence $\sigma$ is a product of $3$-cycles (or the
identity, an empty product --- and for $n \geq 3$ also $(1\ 2\
3)^3$).

\textbf{15.} $(1\ 2)(3\ 4) = (1\ 3\ 2)(1\ 3\ 4)$ (question 13
with $a{=}1, b{=}2, c{=}3, d{=}4$). For the $5$-cycle: by
question 5, $(1\ 2\ 3\ 4\ 5) = (1\ 5)(1\ 4)(1\ 3)(1\ 2)$, and
pairing: $(1\ 5)(1\ 4) = (1\ 4\ 5)$, $(1\ 3)(1\ 2) = (1\ 2\ 3)$:
\[
(1\ 2\ 3\ 4\ 5) = (1\ 4\ 5)(1\ 2\ 3) .
\]
(Check on $3$: $(1\ 2\ 3)$ sends $3 \to 1$, then $(1\ 4\ 5)$
sends $1 \to 4$: net $3 \to 4$, correct.)

\textbf{16.} Apply both sides to an arbitrary point. For $i =
\sigma(a_j)$: the left side gives $\sigma\bigl((a_1\ \dots\
a_k)(a_j)\bigr) = \sigma(a_{j+1})$ (indices mod $k$), which is
what the right side does to $\sigma(a_j)$. For $i$ not of this
form: $\sigma^{-1}(i)$ is outside the support, so the left side
fixes $i$, and so does the right side. Equal everywhere.

\textbf{17.} Sliding the tile of cell $c'$ into the empty cell
$c$ swaps the contents of cells $c$ and $c'$ (tile $9$, the
blank, moves to $c'$). If tile $\sigma(i)$ sat in cell $i$, the
new position is $\sigma' = \sigma \circ (c\ c')$: same contents
except that cells $c, c'$ read each other's former contents. By
question 9, $\varepsilon(\sigma') = -\varepsilon(\sigma)$.

\textbf{18.} A move sends the empty cell to an adjacent cell: its
row or its column changes by exactly $1$, so the taxicab distance
$d$ to cell $9$ changes by $\pm1$, and $(-1)^d$ flips. Since each
move flips both $\varepsilon(\sigma)$ and $(-1)^{d(\sigma)}$,
their product $I(\sigma)$ is unchanged by every move: an
invariant.

\textbf{19.} The solved position has $\varepsilon = +1$, $d = 0$:
$I = +1$. The target (tiles $7, 8$ swapped, blank home) is the
transposition of the contents of cells $7$ and $8$: $\varepsilon
= -1$, $d = 0$: $I = -1$. Since $I$ is invariant and the two
values differ, no sequence of moves joins them.

\textbf{20.} A position with the blank at home is a \omterm{def:b1:counting:objects}{permutation}
of the $8$ tiles among cells $1, \dots, 8$, i.e.\ an element of
$\mathfrak S_8$; it has $d = 0$, so $I = \varepsilon(\sigma)$.
Reachable forces $I = +1$, i.e.\ $\sigma \in \mathfrak A_8$; the
admitted converse says all of $\mathfrak A_8$ is reached. Count:
$\abs{\mathfrak A_8} = \frac{8!}2 = 20\,160$ (question 11).

\textbf{21.} By question 20 the reachable blank-at-home
arrangements form exactly $\mathfrak A_8$ --- in particular a
\emph{\omterm{def:b1:structures:subgroup}{subgroup}} of $\mathfrak S_8$: composing two solvable
scrambles, or inverting one, stays solvable, which is far from
obvious by pure puzzle reasoning.

\textbf{22.} (a) A $3$-cycle of tiles with blank home:
$\varepsilon = +1$ (question 10), $d = 0$, so $I = +1$: reachable
(by the admitted converse) --- one can cycle three tiles.
(b) Tile $5$ and the blank exchanged: the position is the
transposition of the contents of cells $5$ and $9$, so
$\varepsilon = -1$; the blank sits at the center, at taxicab
distance $d = 2$ from home, so $(-1)^d = +1$ and $I = -1$:
unreachable. One cannot simply ``park the blank in the middle''
leaving the tiles otherwise sorted.

\textbf{23.} Let $f \colon \mathfrak S_n \to \{\pm1\}$ be a
\omterm{def:b1:structures:morphism}{morphism}. For any two transpositions $\tau, \tau'$, question 16
provides $\sigma$ with $\sigma\tau\sigma^{-1} = \tau'$ (\omterm{def:b1:logic:map}{map} the
two moved points onto the two others; $n \geq 3$ guarantees room
to do so, though even $n = 2$ is trivial here). Then $f(\tau') =
f(\sigma)f(\tau)f(\sigma)^{-1} = f(\tau)$ since $\{\pm1\}$ is
\omterm{def:b1:structures:group}{abelian}: $f$ is constant on transpositions. If that constant is
$+1$, then $f = 1$ on all products of transpositions, i.e.\
everywhere (question 5). If it is $-1$, then $f(\sigma) =
(-1)^p = \varepsilon(\sigma)$ on a product of $p$ transpositions.
So $f \in \{1, \varepsilon\}$.

\textbf{24.} (i) The \omterm{def:b1:structures:morphism}{morphism} property of $\varepsilon$ and the
\omterm{def:b1:structures:morphism}{kernel} machinery gave $\mathfrak A_n$ its \omterm{def:b1:structures:subgroup}{subgroup} structure and
its size, and \cref{prop:b1:structures:kernel}-style reasoning
runs through questions 11 and 21. (ii) Counting:
$\abs{\mathfrak S_n} = n!$, the halving argument of question 11,
and the count $20\,160$ of question 20 are
\cref{ch:b1:counting} at work. (iii) Questions 8--9 resolve a
genuine well-definedness problem --- ``the parity of the number
of transpositions'' presupposes that this parity does not depend
on the factorization, exactly as $\Z/n\Z$'s operations required
representative-independence in \cref{def:b1:structures:zn}.

\textbf{25.} The signature is a single $\{\pm1\}$-valued
computation, proved once to be well defined, and it does three
jobs at once: internally, it cuts $\mathfrak S_n$ in half and
isolates $\mathfrak A_n$ with its $3$-cycle generators;
externally, it decides in one line a question (``can these two
tiles be swapped?'') that naive search could never settle, since
no finite list of failed move sequences proves impossibility; and
structurally, it is the alternating-sign engine inside the
formula $\det A = \sum_\sigma \varepsilon(\sigma)\,
a_{1\sigma(1)}\cdots a_{n\sigma(n)}$ of \cref{ch:b1:det}. The
pattern --- find a quantity conserved by every elementary move,
compute it at the start and at the target --- is the
mathematician's standard weapon against ``is it possible?''
questions, and it will return whenever a \omterm{def:b1:structures:group}{group} acts on a \omterm{def:b1:logic:sets}{set} of
states.
\end{solution}
