\section*{Chapter \ref{ch:b1:reals} --- Real Numbers}

\begin{solution}{exo:b1:reals:1}
$A$: every element is $\leq 1$ and $1 \in A$: $\sup A = \max A = 1$.
Lower bounds: $0$ bounds below; for $\varepsilon > 0$, Archimedes
provides $n$ with $\frac 1n < \varepsilon$, so no positive number
bounds $A$ below: $\inf A = 0$, not attained (no min).

$B$: terms $0, -\frac12, \frac23, -\frac34, \frac45, \dots$ The even
terms $\frac{n}{n+1}$ ($n$ even) increase to $1$ without reaching it;
the odd terms $-\frac{n}{n+1}$ decrease to $-1$. So $\sup B = 1$ and
$\inf B = -1$, neither attained: no max, no min. (Bounds: $\abs{b}
< 1$ for all $b \in B$; and $\frac{n}{n+1} = 1 - \frac{1}{n+1} > 1 -
\varepsilon$ for $n$ large, similarly below.)

$C = \intoo{-\sqrt 3}{\sqrt 3}$: $\sup C = \sqrt 3$, $\inf C =
-\sqrt 3$, neither attained.
\end{solution}

\begin{solution}{exo:b1:reals:2}
Write $x = \lfloor x \rfloor + u$, $y = \lfloor y \rfloor + v$ with
$u, v \in \intco{0}{1}$. Then $x + y = \lfloor x \rfloor + \lfloor y
\rfloor + (u + v)$ with $u + v \in \intco{0}{2}$. If $u + v < 1$,
$\lfloor x + y\rfloor = \lfloor x\rfloor + \lfloor y \rfloor$; if
$1 \leq u + v < 2$, $\lfloor x+y \rfloor = \lfloor x \rfloor +
\lfloor y \rfloor + 1$. Both cases occur: $(x, y) = (0.2,\, 0.3)$
gives the left equality, $(0.7,\, 0.8)$ the right one.
\end{solution}

\begin{solution}{exo:b1:reals:3}
Let $k = \lfloor x \rfloor$, so $k \leq x < k + 1$. Multiplying by
$n$: $nk \leq nx < nk + n$, and taking \omterm{thm:b1:reals:floor}{floors} (an increasing
operation on integers' side): $nk \leq \lfloor nx \rfloor \leq nk + n
- 1$. Dividing by $n$: $k \leq \frac{\lfloor nx \rfloor}{n} < k + 1$,
so the outer \omterm{thm:b1:reals:floor}{floor} is $k$.
\end{solution}

\begin{solution}{exo:b1:reals:4}
Every element of $A$ is in $B$, so $\sup B$ bounds $A$ above:
therefore $\sup A \leq \sup B$ ($\sup A$ is the \emph{least} \omterm{def:b1:reals:bounds}{upper
bound}). Symmetrically $\inf B \leq \inf A$. Finally $\inf A \leq
\sup A$ because $A$ is nonempty: any $a \in A$ sits between them.
\end{solution}

\begin{solution}{exo:b1:reals:5}
Let $s = \sup A$, $t = \sup B$. Every $a + b \leq s + t$: \omterm{def:b1:reals:bounds}{upper
bound}. For $\varepsilon > 0$, choose $a > s - \frac\varepsilon2$ and
$b > t - \frac\varepsilon2$ (\cref{prop:b1:reals:epsilon}): then $a +
b > s + t - \varepsilon$. By the $\varepsilon$-characterization,
$\sup(A+B) = s + t$.

For $-A$: $m$ bounds $-A$ above $\iff$ $-m$ bounds $A$ below; the
least \omterm{def:b1:reals:bounds}{upper bound} of $-A$ therefore corresponds to the greatest lower
bound of $A$: $\sup(-A) = -\inf A$.
\end{solution}

\begin{solution}{exo:b1:reals:6}
For every $x$: $f(x) + g(x) \leq \sup f + \sup g$; taking the sup of
the left side gives the inequality. Strict example: $E = \{0, 1\}$,
$f = \mathbf{1}_{\{0\}}$ (value $1$ at $0$, else $0$), $g =
\mathbf{1}_{\{1\}}$: $\sup(f + g) = 1 < 2 = \sup f + \sup g$.

No contradiction with \cref{exo:b1:reals:5}: there, $a \in A$ and $b
\in B$ vary \emph{independently}; here the same $x$ feeds both $f$
and $g$ --- the \omterm{def:b1:logic:sets}{set} $\{f(x) + g(x) : x \in E\}$ is smaller than the
set $\{f(x) + g(y) : x, y \in E\}$.
\end{solution}

\begin{solution}{exo:b1:reals:7}
$\sqrt 6$ is irrational: $6 = 2 \times 3$ is not a perfect square,
and $v_2(6q^2) = 1 + 2v_2(q)$ odd prevents $6q^2 = r^2$ (as in
\cref{exo:b1:arith:7}). Now suppose $x = \sqrt 2 + \sqrt 3 \in \Q$.
Then $x^2 = 5 + 2\sqrt 6 \in \Q$, so $\sqrt 6 = \frac{x^2 - 5}{2} \in
\Q$: contradiction. Hence $\sqrt 2 + \sqrt 3 \notin \Q$.
\end{solution}

\begin{solution}{exo:b1:reals:8}
Let $x < y$. Pick $n \in \N$ with $2^n > \frac{1}{y - x}$ (Archimedes:
$2^n \geq n + 1$ by an easy induction, so some power of $2$ exceeds
any real). Then, as in the proof of \cref{thm:b1:reals:density} with
$2^n$ in place of $n$: $m = \lfloor 2^n x \rfloor + 1$ satisfies $x <
\frac{m}{2^n} < y$. So $D$ is dense.
\end{solution}

\begin{solution}{exo:b1:reals:9}
\begin{enumerate}
  \item Suppose $\alpha > 0$. First, $\alpha \in G$. Suppose not: by
        the $\varepsilon$-characterization of the \omterm{def:b1:reals:bounds}{infimum} with
        $\varepsilon = \alpha$, there is $g \in G$ with $\alpha < g <
        2\alpha$ (strict on the left since $\alpha \notin G$); then,
        with $\varepsilon = g - \alpha$, there is $h \in G$ with
        $\alpha < h < g$. Now $g - h \in G$ and $0 < g - h < g -
        \alpha < \alpha$: an element of $G \cap \intoo{0}{+\infty}$
        below its \omterm{def:b1:reals:bounds}{infimum}, absurd. So $\alpha \in G$, and $\alpha\Z
        \subseteq G$ ($G$ is a \omterm{def:b1:structures:group}{group}). Conversely, for $x \in G$, let
        $k = \lfloor x/\alpha \rfloor$: then $x - k\alpha \in G$ and
        $0 \leq x - k\alpha < \alpha$, and the definition of $\alpha$
        forces $x - k\alpha = 0$. Hence $G = \alpha\Z$.
  \item Suppose $\alpha = 0$, and let $x < y$. There is $g \in G$
        with $0 < g < y - x$. The multiple $kg$ with $k = \lfloor x/g
        \rfloor + 1$ satisfies $x < kg \leq x + g < y$, and $kg \in
        G$: density.
\end{enumerate}
$G = \Z + \sqrt 2\,\Z$ is a \omterm{def:b1:structures:subgroup}{subgroup} of $(\R, +)$. It is not of the
form $\alpha\Z$: otherwise $1 = p\alpha$ and $\sqrt 2 = q\alpha$
($p, q \in \Z$) would give $\sqrt 2 = \frac qp \in \Q$,
contradiction. By the dichotomy, $G$ is dense in $\R$.
\end{solution}

\begin{solution}{exo:b1:reals:10}
Let $s = \sup A > 0$, $t = \sup B > 0$. For $a \in A$, $b \in B$: $ab
\leq st$ (multiplying inequalities between \emph{positive} numbers).
For $0 < \varepsilon < \min(s, t)$: choose $a > s - \varepsilon$ and
$b > t - \varepsilon$; then
\[
ab > (s - \varepsilon)(t - \varepsilon)
= st - \varepsilon(s + t) + \varepsilon^2
> st - \varepsilon (s + t),
\]
and $\varepsilon(s+t)$ can be made arbitrarily small: by the
$\varepsilon$-characterization (in the form: no number $< st$ bounds
$AB$ above), $\sup AB = st$.

Positivity is essential: $A = B = \{-1, 0\}$ gives $AB = \{0, 1\}$,
$\sup AB = 1$, while $\sup A \cdot \sup B = 0 \times 0 = 0$.
\end{solution}

\begin{solution}{exo:b1:reals:11}
Write $s = \sup A$, $i = \inf A$. For $a, a' \in A$: $a \leq s$
and $a' \geq i$ give $a - a' \leq s - i$; by symmetry $\abs{a -
a'} \leq s - i$, so $s - i$ bounds the \omterm{def:b1:logic:sets}{set} of gaps above. For
$\varepsilon > 0$, choose $a > s - \frac\varepsilon2$ and $a' < i
+ \frac\varepsilon2$ (\cref{prop:b1:reals:epsilon} and its mirror
for the \omterm{def:b1:reals:bounds}{infimum}): then $\abs{a - a'} \geq a - a' > s - i -
\varepsilon$. By the $\varepsilon$-characterization,
$\operatorname{diam} A = s - i$.

Every $a \in A$ satisfies $i \leq a \leq s$, so $A \subseteq
\intcc{i}{s}$, a closed \omterm{prop:b1:reals:intervals}{interval} of length $\operatorname{diam}
A$. If a closed \omterm{prop:b1:reals:intervals}{interval} $\intcc{u}{v}$ contains $A$, then $v$ is
an \omterm{def:b1:reals:bounds}{upper bound} and $u$ a lower bound of $A$, so $u \leq i$ and $v
\geq s$: $\intcc{i}{s} \subseteq \intcc{u}{v}$. Hence
$\intcc{i}{s}$ is the smallest one.
\end{solution}

\begin{solution}{exo:b1:reals:12}
$0 \in E$, so $E \neq \emptyset$. If $x > \max(1, y)$ then $x^2 >
x > y$, so $E$ is bounded above by $\max(1, y)$: $s = \sup E$
exists (\cref{thm:b1:reals:sup}), and $s \geq \min(1, y) > 0$
because $\min(1, y) \in E$: indeed if $y \geq 1$ then $1^2 = 1
\leq y$, and if $y < 1$ then $y^2 < y$.

\emph{$s^2 < y$ is impossible.} Choose $0 < h < 1$ with $h <
\frac{y - s^2}{2s + 1}$. Then
\[
(s + h)^2 = s^2 + 2sh + h^2 \leq s^2 + (2s + 1)h < y ,
\]
so $s + h \in E$, contradicting that $s$ bounds $E$ above.

\emph{$s^2 > y$ is impossible.} Choose $0 < h < s$ with $h <
\frac{s^2 - y}{2s}$. Then $(s - h)^2 = s^2 - 2sh + h^2 > s^2 -
2sh > y$; every $x \in E$ satisfies $x^2 \leq y < (s - h)^2$,
hence $x < s - h$ (both are $\geq 0$): $s - h$ is an \omterm{def:b1:reals:bounds}{upper bound}
of $E$ smaller than $s$, contradicting leastness.

Therefore $s^2 = y$. Uniqueness: if $0 < s < s'$ then $s^2 <
s'^2$, so two distinct positive roots cannot both square to $y$.
Monotonicity: if $0 < y < y'$, then $\sqrt y \neq \sqrt{y'}$, and
$\sqrt y > \sqrt{y'}$ would give $y > y'$ by squaring: so $\sqrt
y < \sqrt{y'}$.
\end{solution}

\begin{solution}{pb:b1:reals:1}
\textbf{1.} For $\frac 18$: $10 = 8 \cdot 1 + 2$, $20 = 8 \cdot 2
+ 4$, $40 = 8 \cdot 5 + 0$; digits $1, 2, 5$, remainder $0$, then
only zeros: $\frac 18 = 0.125$. For $\frac 17$: $10 = 7 \cdot 1 +
3$, $30 = 7 \cdot 4 + 2$, $20 = 7 \cdot 2 + 6$, $60 = 7 \cdot 8 +
4$, $40 = 7 \cdot 5 + 5$, $50 = 7 \cdot 7 + 1$: digits $1, 4, 2,
8, 5, 7$, remainders $3, 2, 6, 4, 5, 1$. The remainder has
returned to $r = 1$, so the six steps repeat verbatim forever:
$\frac 17 = 0.(142857)$, remainders cycling through $1, 3, 2, 6,
4, 5$.

\textbf{2.} $\frac 13$ in base $2$ ($r_0 = 1$): $2 = 3 \cdot 0 +
2$, $4 = 3 \cdot 1 + 1$, and $r = 1$ recurs: $\frac 13 =
(0.\overline{01})_2$. $\frac{5}{16}$ in base $2$: $10 = 16 \cdot
0 + 10$, $20 = 16 \cdot 1 + 4$, $8 = 16 \cdot 0 + 8$, $16 = 16
\cdot 1 + 0$: $\frac{5}{16} = (0.0101)_2$, terminating. $\frac 12$
in base $3$: $3 = 2 \cdot 1 + 1$, and $r = 1$ recurs at once:
$\frac 12 = (0.\overline{1})_3$.

\textbf{3.} The algorithm's remainder after $N$ steps is $r_N =
b^N p \bmod q$ (proved formally in question 14; here it is the
observation that each step multiplies the remainder by $b$ and
reduces mod $q$). All later digits are $0$ if and only if some
$r_N = 0$, i.e.\ $q \mid b^N p$; since $\gcd(p, q) = 1$, Gauss's
lemma gives $q \mid b^N$. If $q \mid b^N$, every \omterm{def:b1:arith:prime}{prime} factor of
$q$ \omterm{def:b1:arith:divides}{divides} $b^N$, hence \omterm{def:b1:arith:divides}{divides} $b$ (primality). Conversely, if
every \omterm{def:b1:arith:prime}{prime} of $q = p_1^{a_1} \cdots p_r^{a_r}$ \omterm{def:b1:arith:divides}{divides} $b$, then
with $A = \max_i a_i$ each $p_i^{a_i}$ \omterm{def:b1:arith:divides}{divides} $b^A$, and the
$p_i^{a_i}$ are pairwise \omterm{cor:b1:arith:bezout}{coprime}, so $q \mid b^A$. For $q = 20 =
2^2 \cdot 5$: both \omterm{def:b1:arith:prime}{primes} divide $10$ ($\frac{1}{20} = 0.05$),
but $2 \nmid 3$, so $\frac{1}{20}$ repeats forever in base $3$.

\textbf{4.} $1^2 = 1 < 2 < 4 = 2^2$ gives $s_0 = 1$. Then $1.4^2
= 1.96 < 2 < 2.25 = 1.5^2$: $\lfloor 10\sqrt 2 \rfloor = 14$,
$s_1 = 1.4$. Next $1.41^2 = 1.9881 < 2 < 2.0164 = 1.42^2$: $s_2 =
1.41$; $1.414^2 = 1.999396 < 2 < 2.002225 = 1.415^2$: $s_3 =
1.414$; $1.4142^2 = 1.99996164 < 2 < 2.00024449 = 1.4143^2$: $s_4
= 1.4142$. In each case the displayed inequalities say exactly
$s_n \leq \sqrt 2 < s_n + 10^{-n}$, which is the definition of
the \omterm{thm:b1:reals:floor}{floor} of $10^n \sqrt 2$.

\textbf{5.} $A_0 = \lfloor x \rfloor = 0$ since $0 \leq x < 1$.
From $A_{n-1} \leq b^{n-1} x < A_{n-1} + 1$, multiply by $b$:
\[
b\,A_{n-1} \leq b^n x < b\,A_{n-1} + b .
\]
The integer $b\,A_{n-1}$ is $\leq b^n x$, so $b\,A_{n-1} \leq
A_n$; and $b^n x < b\,A_{n-1} + b$ with $b\,A_{n-1} + b$ an
integer forces $A_n \leq b\,A_{n-1} + b - 1$. Hence $0 \leq d_n =
A_n - b\,A_{n-1} \leq b - 1$: a digit.

\textbf{6.} Telescoping: $d_k b^{-k} = A_k b^{-k} - A_{k-1}
b^{-(k-1)}$, so
\[
\sum_{k=1}^n d_k b^{-k} = A_n b^{-n} - A_0 = s_n .
\]
Dividing $A_n \leq b^n x < A_n + 1$ by $b^n$ gives $s_n \leq x <
s_n + b^{-n}$.

\textbf{7.} Induction: $b^0 = 1 \geq 1$, and $b^{n+1} = b \cdot
b^n \geq 2(n + 1) \geq n + 2$. Monotonicity: $s_n - s_{n-1} = d_n
b^{-n} \geq 0$. Each $s_n \leq x$ (question 6): $x$ is an \omterm{def:b1:reals:bounds}{upper
bound} of $\{s_n\}$. For $\varepsilon > 0$, the \omterm{thm:b1:reals:archimedes}{Archimedean
property} provides $n$ with $n + 1 > \frac1\varepsilon$, hence
$b^{-n} < \varepsilon$, and then $s_n > x - b^{-n} > x -
\varepsilon$ by question 6. By \cref{prop:b1:reals:epsilon}, $x =
\sup_n s_n$.

\textbf{8.} Suppose $d_k = b - 1$ for all $k > N$. For $n > N$,
the finite geometric sum gives
\[
s_n = s_N + (b - 1)\sum_{k=N+1}^{n} b^{-k}
= s_N + b^{-N} - b^{-n} .
\]
So $x \geq s_n = s_N + b^{-N} - b^{-n}$ for every $n$; letting
the last term shrink below any $\varepsilon$ (question 7), $x
\geq s_N + b^{-N}$. But question 6 at rank $N$ says $x < s_N +
b^{-N}$: contradiction. The string $(d_n)$ is proper.

\textbf{9.} Bounded: $t_n \leq (b-1)\sum_{k=1}^n b^{-k} = 1 -
b^{-n} < 1$, and $(t_n)$ is nondecreasing, so $y = \sup t_n$
exists with $0 \leq y \leq 1$. Fix $n$. For the two-sided
estimate: $t_n \leq y$ is clear. By properness pick $m > n$ with
$e_m \leq b - 2$. For $p \geq m$:
\[
t_p - t_n = \sum_{k=n+1}^{p} e_k b^{-k}
\leq (b^{-n} - b^{-p}) - b^{-m} < b^{-n} - b^{-m},
\]
the middle sum losing at least $b^{-m}$ against the all-$(b-1)$
maximum; for $p \leq m$, $t_p \leq t_m \leq t_n + b^{-n} -
b^{-m}$ as well (monotonicity plus the case $p = m$). Hence every
$t_p \leq t_n + b^{-n} - b^{-m}$, so $y \leq t_n + b^{-n} -
b^{-m} < t_n + b^{-n}$. (With $n = 0$: $y < 1$, so $y \in
\intco{0}{1}$.) Now $b^n t_n = \sum_{k \leq n} e_k b^{n-k}$ is an
integer, and $b^n t_n \leq b^n y < b^n t_n + 1$: so $\lfloor b^n
y \rfloor = b^n t_n$. Finally the digits of $y$: $d_n(y) = b^n
t_n - b \cdot b^{n-1} t_{n-1} = b^n(t_n - t_{n-1}) = e_n$.

\textbf{10.} Question 9 says: (value of string) has digits (the
string); questions 5--8 say: (digits of $x$) form a proper string
whose truncations have \omterm{def:b1:reals:bounds}{supremum} $x$ (question 7). So the two \omterm{def:b1:logic:map}{maps}
compose to the identity in both orders: they are mutually inverse
bijections between $\intco{0}{1}$ and the proper strings. If two
proper strings had equal value, applying the digit \omterm{def:b1:logic:map}{map} would make
them equal: uniqueness. This is the \emph{$b$-adic expansion
theorem}.

\textbf{11.} Let $e_n = b - 1$ for $n > M$, $M \geq 0$ minimal.
As in question 8, $t_n = t_M + b^{-M} - b^{-n}$ for $n \geq M$,
so the value is $\sup t_n = t_M + b^{-M}$. If $M = 0$ the value
is $0 + 1 = 1$: in base $10$, $0.999\dots = 1$ exactly --- not
approximately. If $M \geq 1$, minimality gives $e_M \leq b - 2$,
and the value is
\[
t_M + b^{-M} = \frac{b^M t_M + 1}{b^M} \in \intoo{0}{1},
\]
a $b$-adic fraction, whose \emph{proper} expansion is $e_1 \dots
e_{M-1}\,(e_M + 1)\,000\dots$ (a terminating string is proper,
and its value is the same number). Conversely a real with two
representations must have one improper (properness pins down the
representation, question 10), hence be of this form. And each
$m/b^N \in \intoo{0}{1}$, written with last nonzero digit $d_N$,
does have the improper twin $d_1 \dots d_{N-1}(d_N -
1)(b-1)(b-1)\dots$: exactly the $b$-adic fractions carry two
names, all other reals one.

\textbf{12.} Say the strings agree up to $m - 1$, with common
truncation $P = s_{m-1}$, and $d_m < e_m$. By question 9 (strict
upper estimate at rank $m$), $x < P + d_m b^{-m} + b^{-m} = P +
(d_m + 1)b^{-m} \leq P + e_m b^{-m} \leq y$, the last step
because $P + e_m b^{-m}$ is $y$'s truncation $t_m \leq y$. So
$d_m < e_m \implies x < y$; exchanging roles, $e_m < d_m \implies
y < x$; and since the strings differ at $m$, one of the two
holds. Both directions follow.

\textbf{13.} Let $z = bx - A_1 \in \intco{0}{1}$ (indeed $A_1
\leq bx < A_1 + 1$). For $n \geq 0$: $b^n z = b^{n+1} x - b^n
A_1$ with $b^n A_1 \in \Z$, so by $\lfloor u - K \rfloor =
\lfloor u \rfloor - K$ ($K$ integer),
\[
A_n(z) = A_{n+1}(x) - b^n A_1(x) .
\]
Hence $d_n(z) = A_n(z) - b\,A_{n-1}(z) = A_{n+1} - b^n A_1 -
b\,A_n + b^n A_1 = d_{n+1}(x)$. So the fractional part of $bx$
carries the shifted digits; iterating $m$ times, the fractional
part of $b^m x$ has digits $(d_{n+m})_{n \geq 1}$.

\textbf{14.} Euclidean division: $b^n p = q\,Q_n + r_n$ with $0
\leq r_n < q$. Divide by $q$: $b^n x = Q_n + \frac{r_n}{q}$ with
$0 \leq \frac{r_n}{q} < 1$, so $Q_n = \lfloor b^n x \rfloor =
A_n$, giving $A_n = \frac{b^n p - r_n}{q}$. For the recurrence:
$b^n p = b(q\,A_{n-1} + r_{n-1}) = q\,(b\,A_{n-1}) +
b\,r_{n-1}$, so $b^n p$ and $b\,r_{n-1}$ differ by a multiple of
$q$: $r_n = (b\,r_{n-1}) \bmod q$.

\textbf{15.} Divide $b\,r_{n-1}$ by $q$: $b\,r_{n-1} = q\,c +
r_n$ with $c = \lfloor b\,r_{n-1}/q \rfloor$. Substituting into
the display of question 14: $b^n p = q(b\,A_{n-1} + c) + r_n$,
and uniqueness of Euclidean division identifies $A_n = b\,A_{n-1}
+ c$, that is $d_n = c = \lfloor b\,r_{n-1}/q \rfloor$. Digit
$n$ depends only on $r_{n-1}$ --- the long division loop of Part
I, now certified.

\textbf{16.} The $q + 1$ remainders $r_0, \dots, r_q$ take values
in the $q$-element \omterm{def:b1:logic:sets}{set} $\intint{0}{q-1}$: by the pigeonhole
principle (\cref{cor:b1:counting:pigeonhole}) two coincide, say
$r_N = r_{N+T}$ with $0 \leq N < N + T \leq q$. Since $r_n$
determines $r_{n+1}$ (question 14), induction gives $r_{n+T} =
r_n$ for all $n \geq N$; since $r_{n-1}$ determines $d_n$
(question 15), $d_{n+T} = d_n$ for all $n \geq N + 1$. Every
rational's expansion is eventually periodic, with preperiod $\leq
q$ and period $\leq q$.

\textbf{17.} The digits of the fractional part of $b^T y$ are
$(d_{n+T}) = (d_n)$ (shift lemma, then pure periodicity): the
same proper string as $y$. By question 10 the values are equal:
$b^T y - \lfloor b^T y \rfloor = y$, so $(b^T - 1)\,y = \lfloor
b^T y \rfloor = A_T \in \N$ and
\[
y = \frac{A_T}{b^T - 1} ,
\]
rational with denominator dividing $b^T - 1$; the numerator $A_T$
is the integer whose base-$b$ digits are $d_1 \dots d_T$. Check:
$0.(142857) = \frac{142857}{999999}$, and $142857 \times 7 =
999999$, so this is $\frac 17$.

\textbf{18.} If $d_{n+T} = d_n$ for $n > N$, the fractional part
$z$ of $b^N x$ has digits $(d_{N+n})_{n\geq1}$ (shift lemma),
which are purely periodic; by question 17, $z \in \Q$. Then $b^N
x = A_N + z$ gives $x = (A_N + z)/b^N \in \Q$. With question 16:
$x$ rational $\iff$ expansion eventually periodic. The right side
mentions the base, the left does not: periodicity in one base is
equivalent to rationality, hence to periodicity in every base.

\textbf{19.} For $x = \frac 1q$, $r_n = b^n \bmod q$. If
$\gcd(b, q) = 1$, then $r_T = r_0 = 1$ if and only if $b^T \equiv
1 \pmod q$; such $T$ exists (pigeonhole gives $b^i \equiv b^j$,
$i < j$, and $b$ is invertible modulo $q$, so $b^{j-i} \equiv
1$), and the least one --- the multiplicative order --- makes the
remainders, hence the digits, purely periodic of period $T$.
No smaller period is possible: a period $T'$ would give $(b^{T'}
- 1)\frac1q \in \N$ (question 17), i.e.\ $q \mid b^{T'} - 1$.
For $q = 7$, $b = 10$: $10 \equiv 3$, $10^2 \equiv 2$, $10^3
\equiv 6$, $10^4 \equiv 4$, $10^5 \equiv 5$, $10^6 \equiv 1
\pmod 7$: order $6$, and indeed $\frac 17$ has period six.

\textbf{20.} The string has infinitely many $0$s ($x^*$'s digits
are mostly zero), so it is proper, and $x^*$ is well defined
(question 9). Suppose the digits eventually periodic with period
$T$ beyond $N$. Infinitely many digits equal $1$ (one per
triangular number), so some $1$ sits at a position $j > N$; then
periodicity puts a $1$ at every position $j + kT$: from $j$
onward, gaps between consecutive $1$s are at most $T$. But the
$1$s sit exactly at the triangular numbers, whose consecutive
gaps $\frac{(j+1)(j+2)}{2} - \frac{j(j+1)}{2} = j + 1$ exceed $T$
eventually: contradiction. Not eventually periodic, so by
question 18, $x^* \notin \Q$ --- irrationality read off the
rhythm of the digits alone.

\textbf{21.} Given $x < y$, question 7 provides $n$ with $b^{-n}
< y - x$; set $m = \lfloor b^n x \rfloor + 1$. Then $b^n x < m
\leq b^n x + 1 < b^n y$, so $x < \frac{m}{b^n} < y$: density,
for every base at once ($b = 2$ recovers
\cref{exo:b1:reals:8}). For $\frac pq \in \intoo{0}{1}$ in base
$b = q$: the first digit is $\lfloor q \cdot \frac pq \rfloor =
p$ and the fractional part of $q \cdot \frac pq = p$ is $0$: all
later digits vanish, a terminating expansion $\frac pq =
(0.p)_q$. Terminating depends on the base; periodicity ---
rationality --- does not (question 18).

\textbf{22.} Each $e_k \in \{1, 2\}$ is a digit of base $10$, and
the string never ends in all $9$s: proper. Its value $y$ lies in
$\intco{0}{1}$ and has digits exactly $(e_k)$ (question 9). Fix
$k$: the $k$-th digit of $y$ is $e_k$, chosen $\neq$ the $k$-th
digit of $x_k$, so the proper strings of $y$ and $x_k$ differ, so
$y \neq x_k$ (question 10: the coding is \omterm{def:b1:logic:inj}{injective}). Thus $y$ is
in no list: no \omterm{def:b1:logic:map}{map} $\N^* \to \intco{0}{1}$ is \omterm{def:b1:logic:inj}{surjective}. The
reals, unlike the rationals, cannot be enumerated ---
uncountability, whose theory \cref{ch:b1:topology} develops.

\textbf{23.} If the expansions agree up to $n$, then $x$ and $y$
have the same truncation $s_n$, and question 6 puts both in
$\intco{s_n}{s_n + b^{-n}}$, an \omterm{prop:b1:reals:intervals}{interval} of length $b^{-n}$:
$\abs{x - y} < b^{-n}$. Converse: $x = 0.1$ and $y = 0.0999$
(terminating, hence proper) satisfy $\abs{x - y} = 10^{-4} <
10^{-3}$, yet their expansions differ at the very first digit.
The culprits are the $b$-adic fractions of question 11: near
them, a tiny move flips every displayed digit ($0.0999 \to
0.1000$), because they are precisely the reals where the
improper twin lurks.

\textbf{24.} $p = 1$, $q = 10$, $b = 2$, $r_0 = 1$: $2 = 10 \cdot
0 + 2$, $4 = 10 \cdot 0 + 4$, $8 = 10 \cdot 0 + 8$, $16 = 10
\cdot 1 + 6$, $12 = 10 \cdot 1 + 2$ --- and $r_5 = 2 = r_1$: the
remainders cycle $(2, 4, 8, 6)$ from index $1$. Digits: $d_1 =
0$, then the repeating block $d_2 d_3 d_4 d_5 = 0, 0, 1, 1$:
\[
\tfrac{1}{10} = (0.0\overline{0011})_2 ,
\]
preperiod $1$, period $4$. By question 3, a terminating base-$2$
expansion would need every \omterm{def:b1:arith:prime}{prime} of $10$ to divide $2$; the \omterm{def:b1:arith:prime}{prime}
$5$ refuses. So $0.1$ is \emph{not} representable by any finite
binary string --- a computer storing finitely many bits keeps
only a truncation, and the accumulated truncation errors are why
floating-point $0.1 + 0.2$ differs from $0.3$ in the last bits.

\textbf{25.} (i) Completeness produced the values: $x = \sup s_n$
and $y = \sup t_n$ (questions 7 and 9) --- over $\Q$ alone, the
proper string of $\sqrt 2$ would name nothing. (ii) The
\omterm{thm:b1:reals:archimedes}{Archimedean property} made $b^{-n}$ eventually smaller than any
$\varepsilon$, forcing the truncations to close in on their
\omterm{def:b1:reals:bounds}{supremum} (questions 7, 21). (iii) The uniqueness clause of the
\omterm{thm:b1:reals:floor}{floor} identified $Q_n = A_n$ in question 14 and legitimized every
digit extraction $\lfloor u - K \rfloor = \lfloor u \rfloor - K$
(question 13). (iv) The pigeonhole principle, applied to finitely
many remainders, is the sole engine of periodicity (question 16).
Moral: proper strings code $\intco{0}{1}$ faithfully and turn
rationality into a visible rhythm; but addition of digit strings
requires carries propagating from infinitely far right, so no
finite-stage rule computes even the first digit of a sum ---
whereas the \omterm{def:b1:reals:bounds}{supremum} interface of \cref{thm:b1:reals:sup} handles
all of analysis with one axiom. Digits are a magnificent
\emph{picture} of $\R$; the \omterm{def:b1:reals:bounds}{supremum} is its \emph{engine}.
\end{solution}
