\section*{Chapter \ref{ch:b1:derivative} --- Differentiation}

\begin{solution}{exo:b1:derivative:1}
$x^x = \eu^{x\ln x}$ on $\intoo{0}{+\infty}$: \omterm{def:b1:derivative:def}{derivative} $(\ln x +
1)\,x^x$.

$\ln(x + \sqrt{x^2+1})$ on $\R$ (the argument is always $> 0$):
\omterm{def:b1:derivative:def}{derivative} $\frac{1}{\sqrt{x^2+1}}$ (computed in
\cref{prop:b1:functions:invhyp} --- it is
$\operatorname{arsinh}$).

$\arctan\frac1x$ on $\R^*$: \omterm{def:b1:derivative:def}{derivative} $\frac{-1/x^2}{1 + 1/x^2} =
\frac{-1}{1 + x^2}$ (consistent with
\cref{prop:b1:functions:arcidentities}~(2): the function is
$\pm\frac\pi2 - \arctan x$ on each half-line).

$\sqrt{1 + \eu^{2x}}$ on $\R$: \omterm{def:b1:derivative:def}{derivative}
$\frac{\eu^{2x}}{\sqrt{1 + \eu^{2x}}}$.
\end{solution}

\begin{solution}{exo:b1:derivative:2}
At $0$: $\bigl|\frac{f(h) - 0}{h}\bigr| = \abs{h \sin\frac1h} \leq
\abs h \to 0$, so $f'(0) = 0$. For $x \neq 0$, the usual rules give
$f'(x) = 2x\sin\frac1x - \cos\frac1x$. Along $x_n = \frac{1}{2\pi
n}$: $f'(x_n) = 0 - 1 \to -1$; along $y_n = \frac{1}{(2n+1)\pi}$:
$f'(y_n) = 0 + 1 \to 1$. Two sequences tending to $0$ with different
limits of $f'$: no limit (\cref{thm:b1:continuity:seqchar}), so $f'$
is not \omterm{def:b1:continuity:continuous}{continuous} at $0$ and $f$ is \omterm{def:b1:derivative:def}{differentiable} without being
$C^1$.
\end{solution}

\begin{solution}{exo:b1:derivative:3}
$\ln(1+x) < x$ for $x > 0$: concavity tangent inequality at $0$
(strict away from the contact point since $\ln$ is strictly concave;
or apply the mean value theorem: $\ln(1+x) = \frac{x}{1+c}$ for some
$c \in \intoo{0}{x}$, and $\frac{x}{1+c} < x$). The same mean value
identity gives the lower bound: $\frac{x}{1+c} > \frac{x}{1+x}$.

Consequence: with $x/n$ in place of $x$,
\[
\frac{x/n}{1 + x/n} < \ln\Bigl(1 + \frac xn\Bigr) < \frac xn
\quad\implies\quad
\frac{x}{1 + x/n} < n \ln\Bigl(1 + \frac xn\Bigr) < x .
\]
The left member tends to $x$: by the squeeze, $n\ln(1 + \frac xn)
\to x$, and by \omterm{def:b1:continuity:continuous}{continuity} of $\exp$, $\bigl(1 + \frac xn\bigr)^n =
\eu^{n\ln(1 + x/n)} \to \eu^x$.
\end{solution}

\begin{solution}{exo:b1:derivative:4}
Let $x_1 < x_2 < \dots < x_k$ be distinct roots of $P$. On each
$\intcc{x_i}{x_{i+1}}$, Rolle (\cref{thm:b1:derivative:rolle})
produces $c_i \in \intoo{x_i}{x_{i+1}}$ with $P'(c_i) = 0$: that is
$k - 1$ roots of $P'$, distinct because the \omterm{def:b1:topology:open}{open} \omterm{prop:b1:reals:intervals}{intervals} are
disjoint --- and interlaced by construction.

If $P$ (degree $n$) has all roots real, write them with
multiplicities $m_1 + \dots + m_k = n$. Each root of \omterm{def:b1:poly:derivative}{multiplicity}
$m_i \geq 2$ is a root of $P'$ of \omterm{def:b1:poly:derivative}{multiplicity} $m_i - 1$
(\cref{prop:b1:poly:multiplicity}), contributing $\sum (m_i - 1) = n
- k$; Rolle contributes $k - 1$ more, all distinct from these. Total
$\geq n - 1 = \deg P'$: all roots of $P'$ are real.
\end{solution}

\begin{solution}{exo:b1:derivative:5}
Fix $\varepsilon > 0$ and $A$ with $\abs{f'(t) - \ell} \leq
\varepsilon$ for $t \geq A$. For $x > A$, the mean value theorem on
$\intcc{A}{x}$ gives $c \in \intoo{A}{x}$ with
\[
f(x) = f(A) + f'(c)(x - A),
\qquad\text{so}\qquad
\Bigl|\frac{f(x)}{x} - \ell\Bigr|
\leq \frac{\abs{f(A)} + \abs\ell A}{x} + \abs{f'(c) - \ell}
\cdot\frac{x - A}{x} \leq \frac{C_A}{x} + \varepsilon .
\]
For $x$ large, $\frac{C_A}{x} \leq \varepsilon$: hence
$\frac{f(x)}{x} \to \ell$.

Yes: $f(x+1) - f(x) = f'(c_x)$ with $c_x \in \intoo{x}{x+1}$ (mean
value theorem on $\intcc{x}{x+1}$), and $c_x \to +\infty$, so
$f(x+1) - f(x) \to \ell$.
\end{solution}

\begin{solution}{exo:b1:derivative:6}
Induction on $n$. For $n = 1$: Rolle. If the claim holds for $n - 1$:
$f$ vanishes at $n+1$ points, so by Rolle applied on the $n$ gaps,
$f'$ vanishes at $n$ distinct points; the induction hypothesis
applied to $f'$ ($n-1$ times \omterm{def:b1:derivative:def}{differentiable}, $n$ zeros) makes
$(f')^{(n-1)} = f^{(n)}$ vanish somewhere.

Application: if $P$ of degree $\leq n$ vanishes at $n+1$ points, then
$P^{(n)}$, a constant equal to $n!$ times the leading coefficient,
vanishes: the leading coefficient is $0$, and one concludes by
downward induction (or directly: all coefficients vanish).
\end{solution}

\begin{solution}{exo:b1:derivative:7}
By the mean value theorem on $\intcc{a}{x_0}$ and on
$\intcc{x_0}{b}$:
\[
f'(c_1) = \frac{f(x_0) - f(a)}{x_0 - a} = \frac{f(x_0)}{x_0 - a} > 0,
\qquad
f'(c_2) = \frac{f(b) - f(x_0)}{b - x_0} = \frac{-f(x_0)}{b - x_0} < 0,
\]
with $c_1 < x_0 < c_2$. Then the mean value theorem applied to $f'$
on $\intcc{c_1}{c_2}$ gives $c$ with
\[
f''(c) = \frac{f'(c_2) - f'(c_1)}{c_2 - c_1} < 0 . \qedhere
\]
\end{solution}

\begin{solution}{exo:b1:derivative:8}
$f'(x) = \frac{1 - \ln x}{x^2}$: $f$ increases on $\intoc{0}{\eu}$,
decreases on $\intco{\eu}{+\infty}$, with maximum $f(\eu) =
\frac1\eu$; limits $-\infty$ at $0^+$ and $0$ at $+\infty$ (\omterm{prop:b1:functions:powerrules}{growth
comparison}).

For $\eu \leq a < b$: $f$ strictly decreasing there gives
$\frac{\ln a}{a} > \frac{\ln b}{b}$, i.e.\ $b \ln a > a \ln b$,
i.e.\ $a^b > b^a$.

With $a = \eu < b = \pi$: $\eu^\pi > \pi^\eu$.

Small pairs: $(2, 3)$: $2^3 = 8 < 9 = 3^2$ --- reversed! The reason:
$2 < \eu$, and on $\intoo{0}{\eu}$ the function $f$ is
\emph{increasing}, so the comparison flips when both numbers sit
below $\eu$, and is unpredictable across $\eu$ ($f(2) = f(4)$
explains the tie $2^4 = 4^2 = 16$).
\end{solution}

\begin{solution}{exo:b1:derivative:9}
\emph{Jensen for $\ln$, by induction on $n$.} Claim: for positive
$x_i$ and weights $\lambda_i > 0$ with $\sum \lambda_i = 1$:
$\ln\bigl(\sum \lambda_i x_i\bigr) \geq \sum \lambda_i \ln x_i$. For
$n = 2$ this is concavity. Step: with $\Lambda = \lambda_1 + \dots +
\lambda_{n-1} = 1 - \lambda_n$ and $y = \sum_{i<n}
\frac{\lambda_i}{\Lambda} x_i$,
\[
\ln\Bigl(\sum_{i \leq n} \lambda_i x_i\Bigr)
= \ln\bigl(\Lambda y + \lambda_n x_n\bigr)
\geq \Lambda \ln y + \lambda_n \ln x_n
\geq \Lambda \sum_{i<n} \frac{\lambda_i}{\Lambda}\ln x_i
+ \lambda_n \ln x_n,
\]
using concavity ($n = 2$) then the induction hypothesis.

With $\lambda_i = \frac 1n$ and $x_i = a_i$: $\ln\frac{\sum a_i}{n}
\geq \frac 1n \sum \ln a_i = \ln\sqrt[n]{a_1\cdots a_n}$;
exponentiate. Equality: $\ln$ is \emph{strictly} concave ($\ln'' < 0$),
so equality at each step forces the averaged points to coincide ---
i.e.\ all $a_i$ equal; and if all are equal, equality is clear.
\end{solution}

\begin{solution}{exo:b1:derivative:10}
Let $g(x) = f(x) - vx$: \omterm{def:b1:derivative:def}{differentiable}, with $g'(a) = f'(a) - v < 0$
and $g'(b) = f'(b) - v > 0$. By the extreme value theorem, $g$
attains its minimum on $\intcc{a}{b}$ at some $c$. It is not at $a$:
since $g'(a) < 0$, points just right of $a$ have $g < g(a)$. It is
not at $b$: since $g'(b) > 0$, points just left of $b$ have $g <
g(b)$. So $c$ is \omterm{def:b1:topology:closure}{interior}, and \cref{prop:b1:derivative:fermat}
gives $g'(c) = 0$, i.e.\ $f'(c) = v$.
\end{solution}

\begin{solution}{exo:b1:derivative:11}
\emph{Uniqueness:} two fixed points $\ell \neq \ell'$ would give
$\abs{\ell - \ell'} = \abs{f(\ell) - f(\ell')} \leq k\abs{\ell -
\ell'} < \abs{\ell - \ell'}$, absurd.

\emph{Existence:} $g(x) = f(x) - x$ satisfies, by the mean value
inequality, $f(x) \leq f(0) + k\abs x$; so for $x \geq
\frac{\abs{f(0)}}{1 - k}$, $g(x) \leq f(0) + kx - x \leq 0$, and
symmetrically $g(-x) \geq 0$ for $x$ large. The intermediate value
theorem gives a zero $\ell$ of $g$: a fixed point.

\emph{Convergence:} the mean value inequality again:
\[
\abs{u_{n+1} - \ell} = \abs{f(u_n) - f(\ell)} \leq k\abs{u_n - \ell},
\]
so by induction $\abs{u_n - \ell} \leq k^n \abs{u_0 - \ell} \to 0$.
\end{solution}

\begin{solution}{exo:b1:derivative:12}
\begin{enumerate}
  \item If $g(b) = g(a)$, Rolle would give an \omterm{def:b1:topology:closure}{interior} zero of
        $g'$: excluded. Set $\lambda = \frac{f(b) - f(a)}{g(b) -
        g(a)}$ and $h = f - \lambda g$: $h$ is \omterm{def:b1:continuity:continuous}{continuous} on
        $\intcc{a}{b}$, \omterm{def:b1:derivative:def}{differentiable} inside, and $h(b) - h(a)
        = f(b) - f(a) - \lambda(g(b) - g(a)) = 0$. Rolle
        provides $c$ with $h'(c) = 0$, i.e.\ $f'(c) =
        \lambda\,g'(c)$; divide by $g'(c) \neq 0$.
  \item For $x > a$ close to $a$, part (1) on $\intcc{a}{x}$
        (where $g' \neq 0$) gives $g(x) \neq 0$ and $c_x \in
        \intoo{a}{x}$ with
        \[
        \frac{f(x)}{g(x)} = \frac{f(x) - f(a)}{g(x) - g(a)}
        = \frac{f'(c_x)}{g'(c_x)} .
        \]
        As $x \to a^+$, $c_x \to a^+$ (squeeze), so the right
        side tends to $\ell$: $\frac{f}{g} \to \ell$.
  \item $\frac{f(x)}{g(x)} = x\sin\frac1x \to 0$, while
        $\frac{f'(x)}{g'(x)} = 2x\sin\frac1x - \cos\frac1x$ has
        no limit at $0$ (\cref{exo:b1:derivative:2}):
        l'Hospital's rule transfers information only from
        $\frac{f'}{g'}$ to $\frac fg$, never back.
\end{enumerate}
\end{solution}

\begin{solution}{pb:b1:derivative:1}
\textbf{1.} $\bigl|\frac ab - \frac pq\bigr| = \frac{\abs{aq -
bp}}{bq}$, and $aq - bp$ is a nonzero integer when the fractions
differ: the distance is $\geq \frac{1}{bq}$. So no rational
other than $x$ itself enters the punctured \omterm{prop:b1:reals:intervals}{interval} of radius
$\frac{1}{bq}$ around $x = \frac ab$.

\textbf{2.} If $\bigl|\sqrt2 - \frac pq\bigr| \geq 1 \geq
\frac{1}{4q^2}$, done. Otherwise $\frac pq \in \intoo{\sqrt2 -
1}{\sqrt2 + 1}$, so $0 < \sqrt2 + \frac pq < 2\sqrt2 + 1 < 4$.
Since $\sqrt 2 \notin \Q$, $p^2 - 2q^2$ is a nonzero integer, and
\[
\Bigl|\sqrt2 - \frac pq\Bigr|
= \frac{\abs{2q^2 - p^2}}{q^2\,\bigl(\sqrt2 + \frac pq\bigr)}
\geq \frac{1}{4q^2} .
\]

\textbf{3.} $(p + 2q)^2 - 2(p + q)^2 = -(p^2 - 2q^2)$: the value
$\pm1$ propagates. From $(1,1)$:
\[
(3,2),\ (7,5),\ (17,12),\ (41,29),\ (99,70),
\]
with $p^2 - 2q^2$ alternating $-1, +1, \dots$
For these, $\frac pq \geq 1$, so $\sqrt2 + \frac pq > 2$ and
\[
\Bigl|\sqrt2 - \frac pq\Bigr| =
\frac{1}{q^2(\sqrt2 + p/q)} < \frac{1}{2q^2} ,
\]
with $q \to \infty$: infinitely many order-$2$ approximations.
With question 2, the exponent $2$ is exact for $\sqrt 2$.

\textbf{4.} The $N + 1$ numbers $kx - \lfloor kx\rfloor$ ($0
\leq k \leq N$) lie in the $N$ boxes $\intco{\frac
jN}{\frac{j+1}{N}}$: two share a box
(\cref{cor:b1:counting:pigeonhole}), say for $i < j$. With $q =
j - i \leq N$ and $p = \lfloor jx\rfloor - \lfloor ix\rfloor$:
$\abs{qx - p} < \frac1N$, so $\bigl|x - \frac pq\bigr| <
\frac{1}{Nq} \leq \frac{1}{q^2}$. Letting $N \to \infty$: since
$x$ is irrational, each fixed fraction has positive distance to
$x$, while $\frac{1}{Nq} \leq \frac 1N \to 0$ forces new
fractions to appear: infinitely many distinct $\frac pq$ with
$\bigl|x - \frac pq\bigr| < \frac{1}{q^2}$.

\textbf{5.} Start from any nonzero integer $P_0$ with $P_0(x) =
0$. If $P_0$ has a rational root $\frac ab$, the factor theorem
(\cref{thm:b1:poly:factor}) writes $P_0 = \bigl(X - \frac
ab\bigr)Q$ with $Q \in \Q[X]$;
since $x \neq \frac ab$ ($x$ irrational), $Q(x) = 0$, and
clearing denominators gives a nonzero \emph{integer} \omterm{def:b1:poly:def}{polynomial}
of smaller degree vanishing at $x$. The degree drops at each
step, so the process stops: we reach $P \in \Z[X]$, $P(x) = 0$,
with no rational root, of some degree $d$. If $d \leq 1$, $P =
uX + v$ would make $x = -\frac vu$ rational: so $d \geq 2$.

\textbf{6.} $q^d\,P\bigl(\frac pq\bigr) = a_d p^d + a_{d-1}
p^{d-1} q + \dots + a_0 q^d$ is an integer, and it is nonzero
because $P$ has no rational root: $\bigl|P\bigl(\frac
pq\bigr)\bigr| \geq q^{-d}$.

\textbf{7.} Note $M > 0$: $P'$ is a nonzero \omterm{def:b1:poly:def}{polynomial} ($d \geq
2$), so it cannot vanish identically on
$\intcc{x-1}{x+1}$. If $\bigl|x - \frac pq\bigr| > 1$, then it
exceeds $\frac{C}{q^d}$ trivially. Otherwise $\frac pq \in
\intcc{x-1}{x+1}$ and the mean value theorem
(\cref{thm:b1:derivative:mvt}) gives $c$ between $x$ and
$\frac pq$ with
\[
\Bigl|P\Bigl(\frac pq\Bigr)\Bigr|
= \Bigl|P\Bigl(\frac pq\Bigr) - P(x)\Bigr|
= \abs{P'(c)}\,\Bigl|x - \frac pq\Bigr|
\leq M\,\Bigl|x - \frac pq\Bigr| ,
\]
so with question 6: $\bigl|x - \frac pq\bigr| \geq
\frac{1}{Mq^d} \geq \frac{C}{q^d}$.

\textbf{8.} Suppose $x = \frac ab$ is Liouville. Pick $n$ with
$2^{n-1} > b$ and the corresponding $\frac pq$, $q \geq 2$:
\[
0 < \Bigl|x - \frac pq\Bigr| < \frac{1}{q^n}
= \frac{1}{q^{n-1}\,q} \leq \frac{1}{2^{n-1} q} <
\frac{1}{bq} ,
\]
contradicting question 1. So Liouville numbers are irrational.

\textbf{9.} If a Liouville $x$ were algebraic: it is irrational
(question 8), so questions 5--7 provide $d \geq 2$ and $C > 0$
with $\bigl|x - \frac pq\bigr| \geq \frac{C}{q^d}$ always. For
each $n$, the Liouville approximant gives $\frac{C}{q^d} <
q^{-n}$, i.e.\ $C < q^{d-n} \leq 2^{d-n}$ (as $q \geq 2$). For
$n$ large, $2^{d-n} < C$: contradiction. Liouville numbers are
transcendental.

\textbf{10.} Ones at positions $1, 2, 6, 24$; all other digits
among the first $25$ vanish:
\[
L = 0.1100010000\,0000000000\,00010\dots
\]

\textbf{11.} For $m > k$, the positions $n!$ with $n > k$ are
distinct integers $\geq (k+1)!$, so the finite geometric sum
gives
\[
t_m - t_k = \sum_{n=k+1}^{m} 10^{-n!}
\leq \sum_{j = (k+1)!}^{m!} 10^{-j}
< 10^{-(k+1)!}\,\frac{1}{1 - \frac1{10}}
= \frac{10}{9}\,10^{-(k+1)!} ;
\]
taking the \omterm{def:b1:reals:bounds}{supremum} over $m$: $L - t_k \leq
\frac{10}{9}10^{-(k+1)!} < 2\cdot10^{-(k+1)!}$. Lower bound: $L
\geq t_{k+1} = t_k + 10^{-(k+1)!}$.

\textbf{12.} $p_k = 10^{k!}\,t_k \in \N$, $q_k = 10^{k!}$, and
$(k+1)! = (k+1)\,k!$ gives $10^{-(k+1)!} = q_k^{-(k+1)}$:
question 11 reads
\[
0 < L - \frac{p_k}{q_k} < \frac{2}{q_k^{\,k+1}} .
\]
Given $n$: for $k \geq n$, $2\,q_k^{-(k+1)} \leq q_k^{-n}$
(indeed $q_k^{\,k+1-n} \geq q_k \geq 10 > 2$), and $q_k \geq 2$:
the definition of question 8 is met. $L$ is a Liouville number.

\textbf{13.} By question 9, $L$ is transcendental --- the first
number in history proved transcendental (Liouville, 1844).
Digit cross-check: the string has infinitely many ones with
consecutive gaps $(k+1)! - k! = k\cdot k! \to \infty$, so it is
not eventually periodic, and $L \notin \Q$ by the periodicity
criterion of \cref{pb:b1:reals:1} (question 18) --- consistent.

\textbf{14.} With digits $d_n \in \intint{1}{9}$ at the
factorial positions: the tail bound of question 11 scales by at
most $9$: $0 < L' - t'_k \leq 9\cdot\frac{10}{9}\,10^{-(k+1)!} =
10\,q_k^{-(k+1)}$ (positivity because the digit at position
$(k+1)!$ is nonzero). For $k \geq n$: $10\,q_k^{-(k+1)} \leq
q_k^{-n}$ since $q_k^{\,k+1-n} \geq 10$: again a Liouville
number, hence transcendental. These values are pairwise
distinct for distinct digit choices (the strings are proper ---
zeros abound --- and proper strings determine their value,
\cref{pb:b1:reals:1}, question 10). Given any list $k \mapsto
x_k$ of them, choose the $k$-th factorial digit in
$\intint{1}{9}$ different from that of $x_k$: a number of the
same shape missing from the list. Uncountably many explicit
transcendentals.

\textbf{15.} First a lemma: \emph{if $\bigl|x - \frac pq\bigr|
\geq \frac{C}{q^s}$ for all $\frac pq \neq x$, then $x$ is not
approximable to any order $\mu > s$.} Indeed infinitely many
$\frac pq \neq x$ with $\bigl|x - \frac pq\bigr| <
\frac{c}{q^\mu}$ would force $\frac{C}{q^s} < \frac{c}{q^\mu}$,
i.e.\ $q^{\mu - s} < \frac cC$: the $q$ are bounded, and
boundedly many fractions lie within distance $1$ of $x$ ---
finitely many candidates, not infinitely many. Now the
hierarchy: rationals are approximable to order $1$ ($\frac
pq$ with $p = \lfloor qx\rfloor + 1$ gives error $\leq \frac1q <
\frac2q$) and to no order $\mu > 1$ (question 1 gives the
hypothesis of the lemma with $s = 1$, $C = \frac1b$); $\sqrt2$:
order $2$ (question 3) and no more (question 2 and the lemma);
every irrational: at least $2$ (question 4); algebraic of degree
$d$: at most $d$ (question 7 and the lemma); Liouville numbers:
every order (question 12's display, with $c = 2$).

\textbf{16.} With $r = \frac ab$: $\frac{p_k}{q_k} + \frac ab =
\frac{b p_k + a q_k}{b q_k} =: \frac{P_k}{Q_k}$, $Q_k = b q_k
\geq 2$, and
\[
\Bigl|(L + r) - \frac{P_k}{Q_k}\Bigr| = L - \frac{p_k}{q_k}
< 2\,q_k^{-(k+1)} = 2\,b^{\,k+1} Q_k^{-(k+1)} .
\]
Given $n$: for large $k$, $Q_k^{\,k+1-n} \geq Q_k = b\,10^{k!}
\geq 2\,b^{\,k+1}$ (the factorial crushes the power), so the
error is $< Q_k^{-n}$: $L + r$ is Liouville. Since $\Q$ is \omterm{def:b1:topology:dense}{dense}
and each $L + r$ is transcendental, \omterm{pb:b1:logic:1}{transcendental numbers} are
\omterm{def:b1:topology:dense}{dense} in $\R$.

\textbf{17.} For $h \geq 1$ there are finitely many $P \in
\Z[X]$ with $\deg P + \sum_i \abs{a_i} \leq h$ (degree $\leq h$
and each coefficient in $\intint{-h}{h}$: at most
$(2h+1)^{h+1}$). Every nonzero integer \omterm{def:b1:poly:def}{polynomial} has such a
height, and has at most $\deg P$ real roots: the \omterm{pb:b1:logic:1}{algebraic
numbers} form a countable union (over $h$) of \omterm{def:b1:counting:card}{finite sets}, hence
can be listed as a single sequence. If the transcendentals could
also be listed, interleaving the two lists would list $\R$,
contradicting \cref{pb:b1:reals:1} (question 22). So
\omterm{pb:b1:logic:1}{transcendental numbers} form an uncountable \omterm{def:b1:logic:sets}{set}. Comparison:
Cantor proves \emph{most} reals are transcendental yet exhibits
none; Liouville exhibits one, with effective constants (Part V)
--- existence by abundance versus existence by construction.

\textbf{18.} From question 2, the lemma hypothesis holds with $s
= 2$, $C = \frac14$. If $\sqrt2$ were Liouville, then for $n =
3$: $\frac{1}{4q^2} < q^{-3}$ forces $q < 4$, so $q \in \{2,
3\}$; only finitely many $\frac pq$ with these $q$ lie within
$1$ of $\sqrt2$, each at some positive distance
$\geq \varepsilon_0$ ($\sqrt2$ irrational); choosing $n$ with
$2^{-n} < \varepsilon_0$ leaves no admissible $\frac pq$ at all:
contradiction. The same argument with $\frac{C}{q^d}$ shows no
\omterm{pb:b1:logic:1}{algebraic number} is Liouville --- question 9 in effective
clothing.

\textbf{19.} $99^2 - 2\cdot70^2 = 9801 - 9800 = 1$. Hence
\[
\sqrt2 - \frac{99}{70} =
\frac{2 - (99/70)^2}{\sqrt2 + 99/70}
= \frac{-1}{4900\,\bigl(\sqrt2 + \tfrac{99}{70}\bigr)} ,
\qquad
\Bigl|\sqrt2 - \frac{99}{70}\Bigr|
= \frac{1}{4900 \times 2.8284\dots} \approx 7.2\cdot10^{-5} :
\]
$\frac{99}{70} = 1.414285\dots$ against $\sqrt2 =
1.414213\dots$ --- five correct digits.

\textbf{20.} Rational-root test for $P = X^3 - 2$: candidates
$\pm1, \pm2, \pm\frac12$, none a root. So $d = 3$ and Part II
applies to $x = 2^{1/3} = 1.2599\dots$ On $\intcc{x - 1}{x + 1}
\subseteq \intcc{0.25}{2.26}$: $\abs{P'(t)} = 3t^2 \leq 3\,(1 +
2^{1/3})^2 < 3\times(2.26)^2 = 15.32 < 16$, so $M < 16$ and $C
\geq \frac{1}{16}$:
\[
\Bigl|2^{1/3} - \frac pq\Bigr| \geq \frac{1}{16\,q^3}
\qquad\text{for all rationals.}
\]

\textbf{21.} If $\bigl|2^{1/3} - \frac pq\bigr| < 10^{-6}$, then
$\frac{1}{16 q^3} < 10^{-6}$, i.e.\ $q^3 > \frac{10^6}{16} =
62\,500$; since $39^3 = 59\,319 < 62\,500 \leq 64\,000 = 40^3$:
$q \geq 40$.

\textbf{22.} Run Part III in base $2$: $B = \sup_k \sum_{n\leq
k} 2^{-n!}$, $q_k = 2^{k!}$, and the geometric tail (ratio
$\frac12$) gives $0 < B - \frac{p_k}{q_k} \leq 2\cdot2^{-(k+1)!}
= 2\,q_k^{-(k+1)} \leq q_k^{-n}$ for $k \geq n$. So $B$ is
Liouville, hence transcendental: nothing in the argument is
decimal.

\textbf{23.} With ones at positions $3^k$: $q_k = 10^{3^k}$ and
the tail bound gives $0 < x^\dagger - \frac{p_k}{q_k} <
2\cdot10^{-3^{k+1}} = 2\,q_k^{-3}$ (as $3^{k+1} = 3\cdot3^k$):
infinitely many approximations of order $3$. By the lemma of
question 15: order $3 > 1$ rules out rationality, and order $3 >
2$ rules out being a quadratic irrational (whose Liouville
inequality has $s = d = 2$). But an \omterm{pb:b1:logic:1}{algebraic number} of degree
$\geq 3$ is only repelled at order $d \geq 3$: Liouville's
method cannot separate $x^\dagger$ from the cubics. The gap is
closed by Roth's theorem --- every algebraic irrational has
approximation order exactly $2$ --- a twentieth-century result
far beyond this volume; granted it, $x^\dagger$ too is
transcendental.

\textbf{24.} There are at most $(2H+1)^{d+1}$ tuples $(a_0,
\dots, a_d)$ with entries in $\intint{-H}{H}$, and each nonzero
\omterm{def:b1:poly:def}{polynomial} among them has at most $d$ real roots: at most
$d\,(2H+1)^{d+1}$ \omterm{pb:b1:logic:1}{algebraic numbers} arise --- the finiteness
that let question 17 enumerate them all.

\textbf{25.} (i) The single analytic ingredient is the mean
value theorem, in question 7, converting the vanishing $P(x) =
0$ into the Lipschitz repulsion $\abs{P(p/q)} \leq M\abs{x -
p/q}$. (ii) The tension: integrality pushes $\abs{P(p/q)}$ up to
$q^{-d}$, smoothness pulls it down to $M\abs{x - p/q}$ --- a
rational too close to $x$ would be crushed between the two.
(iii) Liouville constructs one transcendental with effective
constants; Cantor shows almost all reals are transcendental
without naming one: construction versus \omterm{def:b1:counting:card}{cardinality}. (iv) The
weekend problem of \cref{ch:b1:integration} proves the
irrationality of $\pi$ by the same squeeze --- an integral that
would be a positive integer yet is trapped in $\intoo{0}{1}$ ---
with integration replacing differentiation as the analytic
half.
\end{solution}
