\section*{Chapter \ref{ch:b1:taylor} --- Taylor Formulas and Asymptotic
Expansions}

\begin{solution}{exo:b1:taylor:1}
\[
\eu^{2x} = 1 + 2x + 2x^2 + \frac{4x^3}{3} + o(x^3);
\qquad
\ln(1 - x) = -x - \frac{x^2}{2} - \frac{x^3}{3} - \frac{x^4}{4} +
o(x^4);
\]
\[
\sqrt{1+x} = 1 + \frac x2 - \frac{x^2}{8} + \frac{x^3}{16} + o(x^3);
\qquad
\frac{1}{1 + x^2} = 1 - x^2 + x^4 - x^6 + o(x^6),
\]
the last by substituting $-x^2$ into the geometric expansion.
\end{solution}

\begin{solution}{exo:b1:taylor:2}
$\eu^x - 1 - x = \frac{x^2}{2} + o(x^2)$: limit $\dfrac12$.

$\cos x = 1 - \frac{x^2}{2} + \frac{x^4}{24} + o(x^4)$ and
$\sqrt{1 - x^2} = 1 - \frac{x^2}{2} - \frac{x^4}{8} + o(x^4)$:
difference $\frac{x^4}{24} + \frac{x^4}{8} = \frac{x^4}{6} + o(x^4)$:
limit $\dfrac16$.

$\ln(1+x) - \sin x = \bigl(x - \frac{x^2}{2}\bigr) - x + o(x^2) =
-\frac{x^2}{2} + o(x^2)$: limit $-\dfrac12$.
\end{solution}

\begin{solution}{exo:b1:taylor:3}
$\frac{1}{1+t^2} = 1 - t^2 + t^4 + o(t^5)$; integrating from $0$ to
$x$ (\cref{met:b1:taylor:compute}~(5)):
\[
\arctan x = x - \frac{x^3}{3} + \frac{x^5}{5} + o(x^5)\ \ (\text{even
}o(x^6)\text{, by oddness}).
\]
$(1 - t^2)^{-1/2} = 1 + \frac{t^2}{2} + \frac38 t^4 + o(t^5)$ (binomial
expansion with $\alpha = -\frac12$, $x = -t^2$:
$\binom{-1/2}{2} = \frac{(-\frac12)(-\frac32)}{2} = \frac38$);
integrating:
\[
\arcsin x = x + \frac{x^3}{6} + \frac{3x^5}{40} + o(x^5) .
\]
\end{solution}

\begin{solution}{exo:b1:taylor:4}
Taylor--Lagrange (\cref{thm:b1:taylor:lagrange}) for $\exp$ at $a =
0$, $x = 1$: the $(n+1)$-st derivative is $\eu^t \leq \eu < 3$ on
$\intcc{0}{1}$, so
\[
\Bigl|\eu - \sum_{k=0}^{n} \frac{1}{k!}\Bigr| \leq
\frac{3}{(n+1)!} .
\]
For $6$ exact decimals, want $\frac{3}{(n+1)!} < 5\times 10^{-7}$,
i.e.\ $(n+1)! > 6\times 10^{6}$: since $10! = 3\,628\,800$ and $11! =
39\,916\,800$, $n + 1 = 11$, i.e.\ $n = 10$ suffices.
\end{solution}

\begin{solution}{exo:b1:taylor:5}
$n\ln\bigl(1 + \frac1n\bigr) = n\Bigl(\frac1n - \frac{1}{2n^2} +
\frac{1}{3n^3} + o\bigl(\frac{1}{n^3}\bigr)\Bigr) = 1 - \frac{1}{2n}
+ \frac{1}{3n^2} + o\bigl(\frac{1}{n^2}\bigr)$. Exponentiating, with
$u = -\frac{1}{2n} + \frac{1}{3n^2}$ and $\eu^u = 1 + u + \frac{u^2}2
+ o(u^2)$:
\[
\Bigl(1 + \frac1n\Bigr)^n = \eu\cdot \eu^{u}
= \eu\Bigl(1 - \frac{1}{2n} + \frac{1}{3n^2} + \frac{1}{8n^2}
+ o\Bigl(\frac{1}{n^2}\Bigr)\Bigr)
= \eu\Bigl(1 - \frac{1}{2n} + \frac{11}{24n^2} +
o\Bigl(\frac{1}{n^2}\Bigr)\Bigr).
\]
Limit $\eu$; the error is $\sim \dfrac{\eu}{2n}$: slow (one digit per
tenfold increase of $n$).
\end{solution}

\begin{solution}{exo:b1:taylor:6}
$f(x) = x^2 - x^4 = x^2(1 + o(1))$: first term $x^2$, $p = 2$ even,
coefficient $> 0$: local minimum at $0$ (not global: $f(2) = -12$).

$g(x) = x^3 + x^5$: first term $x^3$, odd $p$: no extremum; $g$
crosses its (horizontal) tangent: inflection at $0$.

$h = \exp$ at $a = 1$: $h(x) = \eu + \eu(x-1) + \frac{\eu}{2}(x-1)^2
+ o((x-1)^2)$; the difference with the tangent is
$\frac{\eu}{2}(x-1)^2 + o(\cdot) > 0$ locally. Globally: $\eu^x -
\eu x \geq 0$ for all $x$ by convexity
(\cref{thm:b1:derivative:convexchar}~(3)): the graph lies above every
tangent, with equality only at the contact point.
\end{solution}

\begin{solution}{exo:b1:taylor:7}
For $x \to +\infty$:
\[
f(x) = x\Bigl(1 + \frac1x\Bigr)^{1/3}
= x\Bigl(1 + \frac{1}{3x} - \frac{1}{9x^2} +
o\Bigl(\frac{1}{x^2}\Bigr)\Bigr)
= x + \frac13 - \frac{1}{9x} + o\Bigl(\frac1x\Bigr):
\]
asymptote $y = x + \frac13$, curve below it near $+\infty$. As $x
\to -\infty$, the same computation is valid (the cube root is
defined for all reals, and $\frac1x \to 0$): same asymptote $y = x +
\frac13$, but now $-\frac{1}{9x} > 0$: curve \emph{above} the line.
\end{solution}

\begin{solution}{exo:b1:taylor:8}
$u_n = \sqrt n\bigl(\sqrt{1 + \tfrac1n} - 1\bigr) = \sqrt
n\bigl(\frac{1}{2n} + o(\frac1n)\bigr) \sim \dfrac{1}{2\sqrt n}$.

$v_n = \ln\bigl(1 + \frac1n\bigr) \sim \dfrac 1n$.

$w_n$: with $h = \frac1n \to 0$, $\sin h - \tan h = \bigl(h -
\frac{h^3}{6}\bigr) - \bigl(h + \frac{h^3}{3}\bigr) + o(h^3) =
-\frac{h^3}{2} + o(h^3)$, so $w_n \sim -\dfrac{1}{2n^3}$.
\end{solution}

\begin{solution}{exo:b1:taylor:9}
Taylor--Lagrange at order $1$ around $x$, on both sides:
\[
f(x + h) = f(x) + h f'(x) + R_+,\quad
f(x - h) = f(x) - h f'(x) + R_-,
\qquad \abs{R_\pm} \leq \frac{h^2}{2} \sup \abs{f''} .
\]
Subtracting: $f(x+h) - f(x-h) = 2h f'(x) + (R_+ - R_-)$, so
\[
\abs{f'(x)} \leq \frac{\abs{f(x+h) - f(x-h)}}{2h} +
\frac{h}{2}\sup_{\intcc{x-h}{x+h}}\abs{f''} .
\]
With global bounds: $\abs{f'(x)} \leq \frac{M_0}{h} + \frac{M_2
h}{2}$ for every $h > 0$. The right side is minimized at $h =
\sqrt{2M_0/M_2}$ (derivative zero), with value $\sqrt{2M_0M_2}$
--- hence $\abs{f'} \leq \sqrt{2M_0M_2}$. (If $M_2 = 0$, let $h \to
\infty$: $f' = 0$, consistent.)
\end{solution}

\begin{solution}{exo:b1:taylor:10}
On $\intoo{0}{\pi}$: $0 < \sin u < u$, so $(u_n)$ is strictly
decreasing, positive, hence convergent; the limit is a fixed point
of $\sin$ in $\intcc{0}{\pi}$, and $\sin \ell = \ell$ forces $\ell =
0$ (since $\sin x < x$ for $x > 0$).
\begin{enumerate}
  \item Using $\sin u = u - \frac{u^3}{6} + o(u^3)$ as $u \to 0$:
        \[
        v_{n+1} - v_n = \frac{1}{\sin^2 u_n} - \frac{1}{u_n^2}
        = \frac{1}{u_n^2}\Bigl(\Bigl(1 - \frac{u_n^2}{6} +
        o(u_n^2)\Bigr)^{\!-2} - 1\Bigr)
        = \frac{1}{u_n^2}\Bigl(\frac{u_n^2}{3} + o(u_n^2)\Bigr)
        \longrightarrow \frac13 .
        \]
  \item By \cref{exo:b1:seq:10}~(3) (Ces\`aro for differences),
        $\frac{v_n}{n} \to \frac13$, i.e.\ $v_n \sim \frac n3$, i.e.\
        $u_n^2 \sim \frac 3n$: since $u_n > 0$,
        \[
        u_n \sim \sqrt{\frac{3}{n}} .
        \]
\end{enumerate}
\end{solution}

\begin{solution}{exo:b1:taylor:11}
Common denominators. First limit:
\[
\frac{1}{x^2} - \frac{1}{\sin^2 x}
= \frac{\sin^2 x - x^2}{x^2\sin^2 x},
\qquad
\sin^2 x = \Bigl(x - \frac{x^3}{6} + o(x^4)\Bigr)^{\!2}
= x^2 - \frac{x^4}{3} + o(x^5) ,
\]
so the numerator is $-\frac{x^4}{3} + o(x^4)$ while the
denominator is $\sim x^4$: the limit is $-\dfrac13$.

Second: $\dfrac1x - \dfrac{1}{\ln(1+x)} = \dfrac{\ln(1+x) -
x}{x\ln(1+x)}$; the numerator is $-\frac{x^2}{2} + o(x^2)$, the
denominator $x\bigl(x + o(x)\bigr) \sim x^2$: the limit is
$-\dfrac12$.
\end{solution}

\begin{solution}{exo:b1:taylor:12}
On $I_k = \intoo{k\pi - \frac\pi2}{k\pi + \frac\pi2}$, the
function $g(x) = \tan x - x$ has derivative $\tan^2 x \geq 0$,
vanishing only at the single point $k\pi$: $g$ is strictly
increasing on $I_k$ (\cref{cor:b1:derivative:monotone}~(2)),
with limits $-\infty$ and $+\infty$ at the ends: exactly one
zero $x_k$. For $k \geq 1$, $g(k\pi) = -k\pi < 0$, so $x_k \in
\intoo{k\pi}{k\pi + \frac\pi2}$: write $x_k = k\pi + \frac\pi2 -
\varepsilon_k$ with $\varepsilon_k \in \intoo{0}{\frac\pi2}$.
Then
\[
x_k = \tan x_k = \tan\Bigl(\frac\pi2 - \varepsilon_k\Bigr)
= \frac{1}{\tan\varepsilon_k}
\quad\Longrightarrow\quad
\varepsilon_k = \arctan\frac{1}{x_k} ,
\]
using $\tan\varepsilon_k = \frac{1}{x_k}$ and $\varepsilon_k \in
\intoo{0}{\frac\pi2}$. Since $x_k \geq k\pi \to \infty$:
$\varepsilon_k \to 0$, and
\[
\varepsilon_k = \arctan\frac{1}{x_k} = \frac{1}{x_k} +
O\Bigl(\frac{1}{x_k^3}\Bigr)
= \frac{1}{k\pi + O(1)} + O\Bigl(\frac{1}{k^3}\Bigr)
= \frac{1}{k\pi} + O\Bigl(\frac{1}{k^2}\Bigr) ,
\]
whence $x_k = k\pi + \frac\pi2 - \frac{1}{k\pi} +
o\bigl(\frac1k\bigr)$.
\end{solution}

\begin{solution}{pb:b1:taylor:1}
\textbf{1.} $S_{2n+1} - S_{2n-1} = a_{2n} - a_{2n+1} \geq 0$ and
$S_{2n+2} - S_{2n} = a_{2n+2} - a_{2n+1} \leq 0$, while $S_{2n}
- S_{2n+1} = a_{2n+1} \to 0$: the sequences $(S_{2n+1})$,
$(S_{2n})$ are adjacent, converging to a common $S$
(\cref{thm:b1:seq:adjacent}) with $S_{2n+1} \leq S \leq S_{2n}$.
For even $n$: $S_{n+1} \leq S \leq S_n$ gives $-a_{n+1} \leq S -
S_n \leq 0$; for odd $n$: $0 \leq S - S_n \leq a_{n+1}$. In both
cases $\abs{S - S_n} \leq a_{n+1}$ and $S - S_n$ has the sign of
$(-1)^{n+1}$, the first omitted term. Strict decrease makes
every displayed inequality strict, in particular $0 < \abs{S -
S_n} < a_{n+1}$.

\textbf{2.} $a_k = \frac{1}{k!}$ decreases strictly to $0$:
question 1 applies. Taylor--Lagrange
(\cref{thm:b1:taylor:lagrange}) for $\exp$ between $-1$ and $0$:
$\abs{\eu^{-1} - T_n} \leq \frac{1}{(n+1)!}$ (the derivative
$\eu^t$ is $\leq 1$ there), so $T_n \to \eu^{-1}$, and the limit
$S$ of question 1 \emph{is} $\eu^{-1}$, with the strict bounds
$0 < \abs{\eu^{-1} - T_n} < \frac{1}{(n+1)!}$.

\textbf{3.} Integrating the identity over $\intcc{0}{1}$: the
left side is $\arctan 1 = \frac\pi4$ (fundamental theorem), the
$k$-th term gives $\frac{(-1)^k}{2k+1}$, and
\[
\rho_n = (-1)^{n+1}\int_0^1 \frac{t^{2n+2}}{1+t^2}\dd t,
\qquad
\abs{\rho_n} \leq \int_0^1 t^{2n+2}\dd t = \frac{1}{2n+3} .
\]

\textbf{4.} The error on $\pi$ is $4\abs{\rho_n} \leq
\frac{4}{2n+3}$: below $10^{-6}$ requires $2n + 3 > 4\cdot10^6$,
about two million terms. Meanwhile $4S_4 = 4\bigl(1 - \frac13 +
\frac15 - \frac17 + \frac19\bigr) = 4 \times 0.834921 =
3.339683$, almost $0.2$ away from $\pi$: five terms, not even
one digit.

\textbf{5.} Integrating over $\intcc{0}{1}$: $\ln 2 =
\sum_{k=1}^{n} \frac{(-1)^{k-1}}{k} + (-1)^n R_n$ with $R_n =
\int_0^1 \frac{t^n}{1+t}\dd t$; from $\frac12 \leq \frac{1}{1+t}
\leq 1$: $\frac{1}{2(n+1)} \leq R_n \leq \frac{1}{n+1}$. The
error is trapped between two multiples of $\frac1n$: six
decimals cost about a million terms.

\textbf{6.} Integrating from $0$ to $x$: $\frac12\ln\frac{1 +
x}{1 - x} = \sum_{k=0}^{n} \frac{x^{2k+1}}{2k+1} + \int_0^x
\frac{t^{2n+2}}{1-t^2}\dd t$. At $x = \frac13$: $\frac{1 +
1/3}{1 - 1/3} = 2$, and on $\intcc{0}{\frac13}$, $\frac{1}{1 -
t^2} \leq \frac98$:
\[
\ln 2 = 2\sum_{k=0}^{n} \frac{(1/3)^{2k+1}}{2k+1} +
\tilde\rho_n,
\qquad
0 < \tilde\rho_n \leq \frac{9}{4}\cdot
\frac{(1/3)^{2n+3}}{2n+3} :
\]
each extra term divides the error by about $9$.

\textbf{7.} Induction: for $n = 1$: $1 - \frac12 = \frac12 = H_2
- H_1$. Step:
\[
H_{2n+2} - H_{n+1} = (H_{2n} - H_n) + \frac{1}{2n+1} +
\frac{1}{2n+2} - \frac{1}{n+1}
= (H_{2n} - H_n) + \frac{1}{2n+1} - \frac{1}{2n+2},
\]
which is exactly the increment of the alternating sum. And
$H_{2n} - H_n = \sum_{k=1}^{n}\frac{1}{n+k}$ is the Riemann sum
of \cref{ex:b1:integration:riemannexample}, converging to $\ln
2$: the even partial sums of Way 1 \emph{are} the Riemann sums
of Way 3.

\textbf{8.} $\sum_{k=1}^{6}\frac{(-1)^{k-1}}{k} = 0.61667$,
error $0.0765$; Way 2 at $n = 5$ gives $0.6931471$ with error
$\leq \frac94\cdot\frac{(1/3)^{13}}{13} = 1.1\cdot10^{-7}$. The
reason: Way 1 evaluates the logarithm series at the boundary
point $x = 1$, where the terms decay like $\frac1k$; Way 2
evaluates at $x = \frac13$, deep inside, where each term carries
a fresh factor $\frac19$.

\textbf{9.} Ten decimals: want $\tilde\rho_n \leq
5\cdot10^{-11}$. At $n = 10$: $\frac94 \cdot
\frac{(1/3)^{23}}{23} = \frac94\cdot\frac{1.06\cdot10^{-11}}{23}
\approx 1.0\cdot10^{-12} < 5\cdot10^{-11}$: eleven terms
suffice.

\textbf{10.} $(5+\iu)^2 = 24 + 10\iu$, then $(5+\iu)^4 = (24 +
10\iu)^2 = 476 + 480\iu$; and $2(1+\iu)(239+\iu) = 2(238 +
240\iu) = 476 + 480\iu$: equal. Arguments: $\arg(5 + \iu) =
\arctan\frac15$, so the left side has argument
$4\arctan\frac15 \approx 0.79 \in \intoo{0}{\pi}$; the right
side has argument $\frac\pi4 + \arctan\frac{1}{239} \in
\intoo{0}{\pi}$. Two equal complex numbers with arguments in the
same interval of length $< 2\pi$:
\[
4\arctan\frac15 = \frac\pi4 + \arctan\frac{1}{239} ,
\]
which is Machin's formula.

\textbf{11.} Integrate $\frac{1}{1+t^2} = \sum_{k=0}^n (-1)^k
t^{2k} + \frac{(-1)^{n+1}t^{2n+2}}{1+t^2}$ from $0$ to $x$:
\[
\arctan x = \sum_{k=0}^{n}\frac{(-1)^k x^{2k+1}}{2k+1} +
r_n(x),
\qquad
\abs{r_n(x)} \leq \int_0^x t^{2n+2}\dd t =
\frac{x^{2n+3}}{2n+3} .
\]

\textbf{12.} Errors: $16\,\frac{(1/5)^{11}}{11} = 3.0\cdot
10^{-8}$ and $4\,\frac{(1/239)^5}{5} \approx 1\cdot10^{-12}$:
total $< 5\cdot10^{-8}$. The displayed sum evaluates to
$3.14159268\dots$, hence $\pi = 3.1415926\dots$ certified to
$5\cdot10^{-8}$: seven decimals from six terms (five at
$\frac15$, two at $\frac1{239}$ counting generously).

\textbf{13.} Leibniz: error $\sim \frac1n$, so each new digit
multiplies the workload by ten. Dalzell
(\cref{pb:b1:integration:1}, question 22): error $4^{1-5m}$,
about three digits per step, each step a heavier polynomial.
Machin: error ratio $\frac{1}{25}$ per term, about $1.4$ digits
per term, each term one division. The moral: the remainder of a
geometric-type expansion scales like $x^{2n}$, so making $x$
small buys digits at a \emph{fixed} cost per term --- Machin's
complex identity is precisely a machine for shrinking $x$.

\textbf{14.} $a_k = \frac{1}{(2k)!}$ decreases strictly to $0$;
by question 1 and \cref{prop:b1:taylor:standard} (Lagrange
bound as in question 2), $\sum_{k \leq n}\frac{(-1)^k}{(2k)!}
\to \cos 1$ with the \emph{strict} bracket $0 < \bigl|\cos 1 -
S'_n\bigr| < \frac{1}{(2n+2)!}$. Suppose $\cos 1 = \frac pq$ and
take $2n \geq q$: then $(2n)!\,S'_n = \sum_{k\leq n} (-1)^k
\frac{(2n)!}{(2k)!} \in \Z$ and $(2n)!\,\frac pq \in \Z$, while
\[
0 < \Bigl|(2n)!\cos 1 - (2n)!S'_n\Bigr| <
\frac{(2n)!}{(2n+2)!} = \frac{1}{(2n+1)(2n+2)} < 1 :
\]
a nonzero integer of absolute value $< 1$. Contradiction: $\cos
1 \notin \Q$.

\textbf{15.} Identically with $a_k = \frac{1}{(2k+1)!}$,
multiplying by $(2n+1)!$ with $2n + 1 \geq q$: $\sin 1 \notin
\Q$. Yet $\cos^2 1 + \sin^2 1 = 1 \in \Q$: products and sums of
irrationals may be rational --- irrationality passes through no
algebraic operation for free.

\textbf{16.} Tail bound: for $m > n$,
\[
\sum_{k=n+1}^{m} \frac{1}{(2k)!}
\leq \frac{1}{(2n+2)!}\Bigl(1 + \frac12 + \frac14 +
\dots\Bigr) \leq \frac{2}{(2n+2)!} ,
\]
since each successive ratio is $\frac{1}{(2k+1)(2k+2)} \leq
\frac12$; the tail is positive (its first term is). So $0 <
\cosh 1 - \sum_{k\leq n}\frac{1}{(2k)!} < \frac{2}{(2n+2)!}$,
and multiplying by $(2n)!$ with $2n \geq q$ traps a nonzero
integer in $\intoo{0}{1}$ again: $\cosh 1 \notin \Q$.

\textbf{17.} $\cos\frac1m = \sum_k
\frac{(-1)^k}{m^{2k}(2k)!}$: the terms decrease strictly to
zero, and $m^{2n}(2n)!\cdot\frac{1}{m^{2k}(2k)!} =
m^{2(n-k)}\frac{(2n)!}{(2k)!} \in \Z$ for $k \leq n$. If
$\cos\frac1m = \frac pq$, multiply the strict bracket by
$q\,m^{2n}(2n)!$: the error is bounded by
$\frac{q}{m^2(2n+1)(2n+2)} < 1$ for large $n$: contradiction.
For $\frac ab$ with $a \geq 2$: clearing denominators multiplies
the tail by $b^{2n}(2n)!$, but the first omitted term is
$\frac{a^{2n+2}}{b^{2n+2}(2n+2)!}$, and the product
$\frac{a^{2n+2}}{b^2(2n+1)(2n+2)}$ \emph{explodes}: the
numerator $a^{2n+2}$ no longer clears, and the trap jams.
(The result is still true --- via Niven-style machinery, not
this one.)

\textbf{18.} At $t = x$ every term of $g$ vanishes except $f(x)
- f(x) = 0$: $g(x) = 0$; $A$ is chosen so $g(a) = 0$.
Differentiating, the sum telescopes:
\[
g'(t) = -\frac{f^{(n+1)}(t)}{n!}(x - t)^n +
A\,\frac{(x-t)^n}{n!}
= \frac{(x-t)^n}{n!}\bigl(A - f^{(n+1)}(t)\bigr) .
\]
Rolle on the segment from $a$ to $x$ gives $c$ strictly between
with $g'(c) = 0$; since $(x - c)^n \neq 0$: $A =
f^{(n+1)}(c)$. Unfolding $g(a) = 0$ yields the Taylor equality
with remainder $\frac{f^{(n+1)}(c)}{(n+1)!}(x-a)^{n+1}$.

\textbf{19.} For $x > 0$, the remainder is $\frac{\eu^{c}}
{(n+1)!}x^{n+1} > 0$: the exponential exceeds each of its
Taylor polynomials, strictly, at every order. For $x < 0$ the
sign of the remainder is that of $x^{n+1}$: $\eu^x$ is
\emph{above} the polynomial for $n$ odd, \emph{below} it for
$n$ even --- alternating sides, as the graphs of $1 + x$ and $1
+ x + \frac{x^2}{2}$ against $\eu^x$ already show.

\textbf{20.} True error: $\sin 0.5 - 0.4791667 =
2.59\cdot10^{-4}$, against the bound $\frac{0.5^5}{120} =
2.60\cdot10^{-4}$: nearly attained (the next term dominates the
tail). Young: for limits and local analysis, where no constant
is needed. Lagrange: for certified decimals. Alternating: when
applicable, same bound \emph{plus} the direction of the error
--- the best of the three, but the rarest.

\textbf{21.} Continuity at $0$: with $u = \frac{1}{x^2} \to
+\infty$, $f(x) = \eu^{-u} \to 0 = f(0)$. Derivative at $0$:
$\bigl|\frac{f(h)}{h}\bigr| = \sqrt u\,\eu^{-u} \to 0$
(\cref{prop:b1:functions:powerrules}): $f'(0) = 0$. For $x \neq
0$, $f'(x) = \frac{2}{x^3}\eu^{-1/x^2}$: the form
$P_1\bigl(\frac1x\bigr)\eu^{-1/x^2}$ with $P_1(X) = 2X^3$;
inductively, differentiating $P_k(\frac1x)\eu^{-1/x^2}$ gives
$P_{k+1}(X) = 2X^3 P_k(X) - X^2 P_k'(X)$, a polynomial. Then
\[
\frac{f^{(k)}(h) - 0}{h} = v\,P_k(v)\,\eu^{-v^2}
\Big|_{v = 1/h} \longrightarrow 0
\]
(polynomial against $\eu^{-v^2}$, growth comparison at
$\pm\infty$): by induction $f^{(k)}(0) = 0$ for all $k$. All
Taylor polynomials of $f$ at $0$ vanish, yet $f > 0$ off $0$:
Taylor--Young is exact at every order and blind beyond the
germ. An expansion is local information only.

\textbf{22.} By question 2, $0 < \abs{\eu^{-1} - T_n} <
\frac{1}{(n+1)!}$, strictly. If $\eu^{-1} = \frac pq$, take $n
\geq q$ and multiply by $n!$: $n!\,T_n \in \Z$ and $n!\frac pq
\in \Z$, so a nonzero integer has absolute value $<
\frac{n!}{(n+1)!} = \frac{1}{n+1} < 1$: contradiction. Hence
$\eu^{-1} \notin \Q$, and $\eu = \frac{1}{\eu^{-1}}$ is
irrational too. The three proofs: adjacent sequences squeezing
$q!\,\eu$ (\cref{exo:b1:seq:9}); the integral recurrence $A_n =
\eu - nA_{n-1}$ (\cref{pb:b1:integration:1}); the alternating
bracket (here). One trap, three certificates of smallness.

\textbf{23.} Group the partial sums in pairs: $\sum_{k=1}^{2n}
(-1)^k b_k = E_n - O_n$ with $E_n = \sum_{j=1}^{n}
\frac{1}{4j^2}$, bounded (by the telescoping bound of
\cref{ex:b1:seq:basel}, $E_n \leq \frac12$), and $O_n =
\sum_{j=1}^{n}\frac{1}{2j-1} \geq \frac12 H_n \to +\infty$
(\cref{exo:b1:seq:5}): the partial sums tend to $-\infty$.
Monotonicity was used in question 1 exactly where $S_{2n+1} -
S_{2n-1} = a_{2n} - a_{2n+1}$ needed a sign: without decrease,
the even and odd subsequences need not be monotone, and
adjacency collapses.

\textbf{24.} $(2+\iu)(3+\iu) = 5 + 5\iu = 5(1+\iu)$; taking
arguments (all in $\intoo{0}{\frac\pi2}$): $\arctan\frac12 +
\arctan\frac13 = \frac\pi4$. Series cost for six decimals:
error $\leq 4\bigl(\frac{(1/2)^{2n+3}}{2n+3} +
\frac{(1/3)^{2n+3}}{2n+3}\bigr)$, which at $n = 10$ is
$\approx 2\cdot10^{-8} < 5\cdot10^{-7}$: eleven terms. Ranking:
better than Leibniz by an exponential margin, behind Machin
(whose dominant point $\frac15$ is smaller than $\frac12$):
roughly $0.6$ digits per term against Machin's $1.4$.

\textbf{25.} (i) For decreasing $a_k \to 0$, the alternating
partial sums converge with $\abs{S - S_n} \leq a_{n+1}$ and the
error carries the sign of the first omitted term. (ii) A finite
identity with explicit remainder can be \emph{evaluated and
bounded at a chosen point}, while a limit statement only
promises eventual closeness --- certification needs the former.
(iii) Extracted: $\pi$ to $5\cdot10^{-8}$ by Machin, $\ln 2$ to
ten decimals by the $\frac13$-series, irrationality of $\cos 1$,
$\sin 1$, $\cosh 1$, $\cos\frac1m$ and $\eu^{-1}$, the
Taylor--Lagrange equality, and the flat-function warning. (iv)
\cref{ch:b1:series} upgrades question 1 into the alternating
series test, separates absolute from conditional convergence,
and its weekend problem stages the rearrangement drama for
which the alternating harmonic series of Way 1 is the star
witness.
\end{solution}
