\section*{Chapter \ref{ch:b1:findim} --- Finite Dimension}

\begin{solution}{exo:b1:findim:1}
\begin{enumerate}
  \item $z = -x - y$: $F = \{(x, y, -x-y)\} =
        \operatorname{Vect}\bigl((1,0,-1), (0,1,-1)\bigr)$; the two
        vectors are \omterm{def:b1:vspaces:free}{free} (\omterm{prop:b1:vspaces:coordinates}{coordinates}): $\dim F = 2$.
  \item $G = \{(x, x, 2t, t)\} =
        \operatorname{Vect}\bigl((1,1,0,0), (0,0,2,1)\bigr)$: \omterm{def:b1:vspaces:free}{free},
        $\dim G = 2$.
  \item $P(1) = P'(1) = 0$ means $(X-1)^2 \mid P$
        (\cref{prop:b1:poly:multiplicity}): $P = (X-1)^2(aX + b)$.
        \omterm{def:b1:vspaces:free}{Basis} $\bigl((X-1)^2, X(X-1)^2\bigr)$, dimension $2$.
\end{enumerate}
\end{solution}

\begin{solution}{exo:b1:findim:2}
$(3,5,7) = (1,1,1) + 2(1,2,3)$ and $(0,1,2) = (1,2,3) - (1,1,1)$:
both are combinations of the first two, which are \omterm{def:b1:vspaces:free}{free} (not
proportional). \omterm{def:b1:findim:rank}{Rank} $2$; \omterm{def:b1:vspaces:free}{basis} of the \omterm{def:b1:vspaces:span}{span}:
$\bigl((1,1,1), (1,2,3)\bigr)$.
\end{solution}

\begin{solution}{exo:b1:findim:3}
Try appending $e_1 = (1,0,0,0)$ and $e_3 = (0,0,1,0)$. The family
$\bigl((1,1,0,0), (0,0,1,1), e_1, e_3\bigr)$ is \omterm{def:b1:vspaces:free}{free}: a null
combination $\alpha(1,1,0,0) + \beta(0,0,1,1) + \gamma e_1 + \delta
e_3 = 0$ reads $(\alpha + \gamma, \alpha, \beta + \delta, \beta) =
0$, so $\alpha = \beta = 0$, then $\gamma = \delta = 0$. Four \omterm{def:b1:vspaces:free}{free}
vectors in dimension $4$: a \omterm{def:b1:vspaces:free}{basis}
(\cref{prop:b1:findim:twoofthree}).
\end{solution}

\begin{solution}{exo:b1:findim:4}
Degrees $0, 1, 2, 3$ pairwise distinct: \omterm{def:b1:vspaces:free}{free}
(\cref{prop:b1:vspaces:freecriteria}), four vectors in dimension
$4$: \omterm{def:b1:vspaces:free}{basis}. For $X^3$: expand downward,
\[
X(X-1)(X-2) = X^3 - 3X^2 + 2X,
\qquad
X(X-1) = X^2 - X,
\]
so $X^3 - X(X-1)(X-2) = 3X^2 - 2X$; and $3X^2 - 2X = 3(X^2 - X) + X
= 3\,X(X-1) + X$. Hence
\[
X^3 = X(X-1)(X-2) + 3\,X(X-1) + 1\cdot X + 0\cdot 1 :
\]
\omterm{prop:b1:vspaces:coordinates}{coordinates} $(0, 1, 3, 1)$ on $\bigl(1, X, X(X-1),
X(X-1)(X-2)\bigr)$. (These are Stirling numbers in disguise.)
\end{solution}

\begin{solution}{exo:b1:findim:5}
Grassmann: $\dim(F \cap G) = 4 + 5 - \dim(F + G)$, and $F + G$ is a
\omterm{def:b1:vspaces:subspace}{subspace} of $E$ containing $G$: $5 \leq \dim(F+G) \leq 7$. Hence
$\dim(F \cap G) \in \{2, 3, 4\}$. Realizations in $\R^7$ with
canonical \omterm{def:b1:vspaces:free}{basis} $(e_1, \dots, e_7)$, taking $G =
\operatorname{Vect}(e_1, \dots, e_5)$:
\begin{itemize}
  \item $F = \operatorname{Vect}(e_1, e_2, e_6, e_7)$: $F + G = \R^7$,
        intersection $\operatorname{Vect}(e_1, e_2)$, dimension $2$;
  \item $F = \operatorname{Vect}(e_1, e_2, e_3, e_6)$: intersection
        of dimension $3$;
  \item $F = \operatorname{Vect}(e_1, e_2, e_3, e_4) \subseteq G$:
        dimension $4$.
\end{itemize}
\end{solution}

\begin{solution}{exo:b1:findim:6}
$H$ is a \omterm{def:b1:vspaces:subspace}{subspace} (\cref{exo:b1:vspaces:1}~(4) verbatim). By the
factor theorem (\cref{thm:b1:poly:factor}), $P \in H \iff P =
(X-1)Q$ with $\deg Q \leq n - 1$: the \omterm{def:b1:logic:map}{map} $Q \mapsto (X-1)Q$ is a
linear bijection from $\R_{n-1}[X]$ onto $H$, so a \omterm{def:b1:vspaces:free}{basis} of $H$ is
\[
\bigl((X-1),\ (X-1)X,\ (X-1)X^2,\ \dots,\ (X-1)X^{n-1}\bigr),
\qquad \dim H = n .
\]
A \omterm{def:b1:vspaces:sum}{supplementary} line: $\operatorname{Vect}(1)$ (constants). Indeed
$H \cap \operatorname{Vect}(1) = \{0\}$ (a nonzero constant does not
vanish at $1$) and dimensions add up to $n + 1$: by Grassmann, $H
\oplus \operatorname{Vect}(1) = \R_n[X]$.
\end{solution}

\begin{solution}{exo:b1:findim:7}
Deletion: removing $u_p$, the \omterm{def:b1:vspaces:span}{span} can only shrink; and it shrinks
by at most one dimension, since adding $u_p$ back to a \omterm{def:b1:vspaces:free}{basis} of the
smaller \omterm{def:b1:vspaces:span}{span} gives a generating family of the larger with at most
one extra vector. Symmetrically, appending a vector $v$: the new
\omterm{def:b1:vspaces:span}{span} contains the old with at most one extra generator, so its
dimension is $r$ (if $v$ was already in the \omterm{def:b1:vspaces:span}{span}) or $r + 1$
(otherwise, by \cref{prop:b1:vspaces:freecriteria}~(2) a \omterm{def:b1:vspaces:free}{basis}
extends). Both \omterm{def:b1:logic:statement}{statements} together give the ``\omterm{def:b1:findim:rank}{rank} is
$1$-Lipschitz'' conclusion.
\end{solution}

\begin{solution}{exo:b1:findim:8}
Along a strictly increasing chain, dimensions strictly increase
($F_i \subseteq F_{i+1}$, $F_i \neq F_{i+1}$ and
\cref{thm:b1:findim:subspaces}: equality of dimensions would force
equality of spaces). So $\dim F_1 < \dim F_2 < \dots < \dim F_k$ is
a strictly increasing sequence of integers in $\intint{0}{n}$: at
most $n + 1$ values, $k \leq n + 1$. Maximal chain in $\R^n$:
\[
\{0\} \subsetneq \operatorname{Vect}(e_1) \subsetneq
\operatorname{Vect}(e_1, e_2) \subsetneq \dots \subsetneq \R^n ,
\]
of length exactly $n + 1$.
\end{solution}

\begin{solution}{exo:b1:findim:9}
$F + G$ strictly contains $F$ (since $G \not\subseteq F$), so
$\dim(F + G) = n$ and Grassmann gives $\dim(F \cap G) = (n-1) +
(n-1) - n = n - 2$.

General claim, by induction on $k$: the intersection $I_k$ of $k$
hyperplanes has $\dim I_k \geq n - k$. True for $k = 1$. Step: $I_{k+1}
= I_k \cap H_{k+1}$, and Grassmann inside $E$:
\[
\dim(I_k \cap H_{k+1}) = \dim I_k + (n - 1) - \dim(I_k + H_{k+1})
\geq \dim I_k + (n-1) - n \geq n - k - 1 . \qedhere
\]
\end{solution}

\begin{solution}{exo:b1:findim:10}
\begin{enumerate}
  \item The condition $u_{n+2} - 3u_{n+1} + 2u_n = 0$ is linear
        and satisfied by the zero sequence: $E$ is a \omterm{def:b1:vspaces:subspace}{subspace}. By
        induction, $u_0$ and $u_1$ determine every $u_n$, and the
        dependence is linear (each step is a linear combination of
        the two previous values); conversely every pair $(a, b)$
        arises from exactly one sequence of $E$ (define $u_n$ by
        the recurrence). As in \cref{ex:b1:findim:dims}, $E$ is
        parametrized \omterm{def:b1:logic:inj}{bijectively} and linearly by $(u_0, u_1) \in
        \R^2$: $\dim E = 2$.
  \item Constants: $3\cdot1 - 2\cdot1 = 1$. Geometric: $3\cdot
        2^{n+1} - 2\cdot 2^n = (6 - 2)2^n = 2^{n+2}$. Both lie in
        $E$. Freeness: $a\cdot 1 + b\cdot 2^n = 0$ for all $n$
        gives, at $n = 0$ and $n = 1$: $a + b = 0$, $a + 2b = 0$,
        so $a = b = 0$. Two \omterm{def:b1:vspaces:free}{free} vectors in dimension $2$: a \omterm{def:b1:vspaces:free}{basis}
        (\cref{prop:b1:findim:twoofthree}).
  \item Solve $a + b = 0$, $a + 2b = 1$: $b = 1$, $a = -1$, so
        $u_n = 2^n - 1$ (the Mersenne sequence).
\end{enumerate}
\end{solution}

\begin{solution}{exo:b1:findim:11}
\begin{enumerate}
  \item Grassmann: $\dim(F \cap G) = \dim F + \dim G - \dim(F +
        G) \geq \dim F + \dim G - n > 0$, so $F \cap G \neq
        \{0\}$. With $n = 100$: $51 + 50 - 100 = 1 > 0$, the
        intersection contains a line.
  \item If $F \cap H = \{0\}$, the sum is direct and $\dim F +
        (n - 1) = \dim(F \oplus H) \leq n$, so $\dim F \leq 1$.
        (Conversely a line not contained in $H$ does satisfy
        this: hyperplanes miss almost nothing.)
\end{enumerate}
\end{solution}

\begin{solution}{exo:b1:findim:12}
Downward induction on $d = \dim F = \dim G$, from $d = n$ to $d =
0$. If $d = n$: $F = G = E$ and $S = \{0\}$ works. Suppose the
\omterm{def:b1:logic:statement}{statement} holds for \omterm{def:b1:vspaces:subspace}{subspace} pairs of dimension $d + 1 \leq n$,
and let $\dim F = \dim G = d < n$.

If $F = G$: take for $S$ any \omterm{def:b1:vspaces:sum}{supplementary} of $F$
(\cref{thm:b1:findim:subspaces}). If $F \neq G$: both are proper,
so by \cref{exo:b1:vspaces:12} there exists $x \notin F \cup G$.
The sums $F' = F \oplus \operatorname{Vect}(x)$ and $G' = G \oplus
\operatorname{Vect}(x)$ are direct ($x \notin F$, $x \notin G$)
and have dimension $d + 1$; by the induction hypothesis they
admit a common \omterm{def:b1:vspaces:sum}{supplementary} $S'$: $E = F' \oplus S' = G' \oplus
S'$. Set $S = \operatorname{Vect}(x) \oplus S'$ --- direct,
because $S' \cap \operatorname{Vect}(x) \subseteq S' \cap F' =
\{0\}$.

Then $F + S = F + \operatorname{Vect}(x) + S' = F' + S' = E$, and
$F \cap S = \{0\}$: if $f = \lambda x + s'$ with $f \in F$, $s'
\in S'$, then $s' = f - \lambda x \in F' \cap S' = \{0\}$, so $f
= \lambda x$, forcing $\lambda = 0$ ($x \notin F$) and $f = 0$.
Hence $E = F \oplus S$, and symmetrically $E = G \oplus S$.
\end{solution}

\begin{solution}{pb:b1:findim:1}
\textbf{1.} $\Q$ contains $0 \neq 1$, is stable under addition,
multiplication and opposites, and every nonzero rational has a
rational inverse: a \omterm{def:b1:structures:field}{field} (\cref{def:b1:structures:field}). The
definitions of \omterm{def:b1:vspaces:span}{span}, freeness, \omterm{def:b1:vspaces:free}{basis} and the proofs of Chapters
18--19 use only vector addition, distributivity and scalar
arithmetic in a \omterm{def:b1:structures:field}{field} --- never an absolute value, an order or a
limit. So $\R$, with its own addition and the multiplication
$\Q \times \R \to \R$, is a $\Q$-vector space, and all the
theorems apply. Division by a scalar occurs once in the proof of
\cref{thm:b1:findim:exchange}: to ``solve for $g_k$'' one \omterm{def:b1:arith:divides}{divides}
by its nonzero coefficient.

\textbf{2.} If $a + b\sqrt2 = 0$ with $a, b \in \Q$ not both
zero: $b \neq 0$ would give $\sqrt2 = -a/b \in \Q$, contradicting
the irrationality of $\sqrt2$ (\cref{exo:b1:arith:7} with $p =
2$); so $b = 0$, then $a = 0$: \omterm{def:b1:vspaces:free}{free} over $\Q$. Over $\R$ the
relation $\sqrt2\cdot 1 + (-1)\cdot\sqrt2 = 0$ is nontrivial:
linked. Hence $\dim_\Q \Q(\sqrt2) = 2$, with unique \omterm{prop:b1:vspaces:coordinates}{coordinates}
$(a, b)$.

\textbf{3.} $(a + b\sqrt2)(c + d\sqrt2) = (ac + 2bd) + (ad +
bc)\sqrt2 \in \Q(\sqrt2)$. Then
\[
\sigma(uv) = (ac + 2bd) - (ad + bc)\sqrt2
= (a - b\sqrt2)(c - d\sqrt2) = \sigma(u)\sigma(v),
\]
so $N(uv) = uv\,\sigma(uv) = u\sigma(u)\,v\sigma(v) = N(u)N(v)$.
If $N(u) = a^2 - 2b^2 = 0$ with $u \neq 0$: $b \neq 0$ would give
$(a/b)^2 = 2$, a rational square root of $2$; so $b = 0$, then $a
= 0$ and $u = 0$: contradiction. Hence $N(u) \neq 0$ for $u \neq
0$.

\textbf{4.} $u \cdot \dfrac{\sigma(u)}{N(u)} = \dfrac{N(u)}{N(u)}
= 1$, and $\sigma(u)/N(u) \in \Q(\sqrt2)$: every nonzero element
is invertible inside $\Q(\sqrt2)$, which is therefore a subfield
of $\R$. Examples: $N(3 + 2\sqrt2) = 9 - 8 = 1$, so
\[
\frac1{3 + 2\sqrt2} = 3 - 2\sqrt2 ;
\qquad
N(1 + \sqrt2) = -1, \quad
\frac1{1 + \sqrt2} = \sqrt2 - 1 .
\]

\textbf{5.} If $\sqrt6 = p/q$ then $6q^2 = p^2$; the exponent of
$2$ is $1 + 2v_2(q)$, odd, on the left, and $2v_2(p)$, even, on
the right: impossible. Now suppose $\sqrt3 = a + b\sqrt2$ with
$a, b \in \Q$. Squaring: $3 = a^2 + 2b^2 + 2ab\sqrt2$. If $ab
\neq 0$, then $\sqrt2 = (3 - a^2 - 2b^2)/(2ab) \in \Q$:
impossible. If $b = 0$: $\sqrt3 = a \in \Q$, contradicting
\cref{exo:b1:arith:7} ($p = 3$). If $a = 0$: $\sqrt3 = b\sqrt2$,
and multiplying by $\sqrt2$: $\sqrt6 = 2b \in \Q$: impossible.
So $\sqrt3 \notin \Q(\sqrt2)$.

\textbf{6.} The products of \omterm{def:b1:vspaces:free}{basis} vectors are
\[
\sqrt2\,\sqrt3 = \sqrt6,\quad
\sqrt2\,\sqrt6 = 2\sqrt3,\quad
\sqrt3\,\sqrt6 = 3\sqrt2,\quad
(\sqrt2)^2 = 2,\ (\sqrt3)^2 = 3,\ (\sqrt6)^2 = 6,
\]
all in $M$. A product of two elements of $M$ expands by
bilinearity into rational combinations of these: $M$ is stable
under multiplication.

\textbf{7.} Let $x + y\sqrt3 = 0$ with $x, y \in K = \Q(\sqrt2)$.
If $y \neq 0$, then $\sqrt3 = -x/y \in K$ (question 4: $K$ is a
\omterm{def:b1:structures:field}{field}), contradicting question 5. So $y = 0$, then $x = 0$: $(1,
\sqrt3)$ is \omterm{def:b1:vspaces:free}{free} over $K$. It generates: $K + K\sqrt3 =
\{(a + b\sqrt2) + (c + d\sqrt2)\sqrt3\} = \operatorname{Vect}_\Q
(1, \sqrt2, \sqrt3, \sqrt6) = M$. Hence $\dim_K M = 2$.

\textbf{8.} A relation $a + b\sqrt2 + c\sqrt3 + d\sqrt6 = 0$
regroups as $(a + b\sqrt2) + (c + d\sqrt2)\sqrt3 = 0$ with
coefficients in $K$; by question 7 both vanish, and by question 2
$a = b = 0$ and $c = d = 0$. So the family is \omterm{def:b1:vspaces:free}{free} and $\dim_\Q M
= 4$.

\textbf{9.} Every $x \in M$ writes $x = \sum_j y_j f_j$ with $y_j
\in K$ (\omterm{def:b1:vspaces:free}{basis} of $M$ over $K$), and each $y_j = \sum_i
\lambda_{ij} e_i$ with $\lambda_{ij} \in \Q$ (\omterm{def:b1:vspaces:free}{basis} of $K$ over
$\Q$); substituting, $x = \sum_{i,j} \lambda_{ij}\, e_i f_j$: the
products generate $M$ over $\Q$.

\textbf{10.} Suppose $\sum_{i,j} \lambda_{ij}\, e_i f_j = 0$ with
$\lambda_{ij} \in \Q$. Regroup: $\sum_j \bigl(\sum_i \lambda_{ij}
e_i\bigr) f_j = 0$, and the inner sums belong to $K$. Freeness of
$(f_j)$ over $K$ gives $\sum_i \lambda_{ij} e_i = 0$ for each
$j$; freeness of $(e_i)$ over $\Q$ then gives $\lambda_{ij} = 0$
for all $i, j$.

\textbf{11.} By questions 9 and 10, $(e_i f_j)_{i \leq m,\, j
\leq n}$ is a \omterm{def:b1:vspaces:free}{basis} of $M$ over $\Q$ with $mn$ elements:
\[
\dim_\Q M = m\,n = \dim_\Q K \cdot \dim_K M .
\]
Check: $\dim_\Q K = 2$ and $\dim_K M = 2$ (questions 2 and 7)
give $\dim_\Q M = 4$, which is question 8.

\textbf{12.} The vectors $u a_1, \dots, u a_d$ belong to $A$
(stability). They are \omterm{def:b1:vspaces:free}{free}: if $\sum_i \lambda_i\, u a_i = 0$,
then $u \sum_i \lambda_i a_i = 0$ in $\R$, and $u \neq 0$ forces
$\sum_i \lambda_i a_i = 0$, hence $\lambda_i = 0$ (the $a_i$ form
a \omterm{def:b1:vspaces:free}{basis}). So $(u a_1, \dots, u a_d)$ is a \omterm{def:b1:vspaces:free}{free family} of $d$
vectors in $A$, $\dim_\Q A = d$: a \omterm{def:b1:vspaces:free}{basis}
(\cref{prop:b1:findim:twoofthree}). In particular $1 \in A$
decomposes as $1 = \sum_i \mu_i\, u a_i = u\,v$ with $v = \sum_i
\mu_i a_i \in A$: the inverse of $u$ lies in $A$. This recovers
question 4 ($A = \Q(\sqrt2)$) and proves at one stroke that $M$
is a subfield of $\R$ (with question 6).

\textbf{13.} The $d + 1$ vectors $1, x, x^2, \dots, x^{d}$ all
lie in $A$ (stability under products); a \omterm{def:b1:vspaces:free}{free family} of $A$ has
at most $d$ vectors (\cref{prop:b1:findim:twoofthree}), so they
are linked: there are rationals $\lambda_0, \dots, \lambda_d$,
not all zero, with $\sum_k \lambda_k x^k = 0$. The \omterm{def:b1:poly:def}{polynomial} $P
= \sum_k \lambda_k X^k$ is nonzero, has rational coefficients,
degree $\leq d$, and $P(x) = 0$.

\textbf{14.} $s^2 = 2 + 2\sqrt6 + 3 = 5 + 2\sqrt6$: \omterm{prop:b1:vspaces:coordinates}{coordinates}
$(5, 0, 0, 2)$. Then
\[
s^3 = s\,s^2 = (\sqrt2 + \sqrt3)(5 + 2\sqrt6)
= 5\sqrt2 + 2\sqrt{12} + 5\sqrt3 + 2\sqrt{18}
= 11\sqrt2 + 9\sqrt3 ,
\]
\omterm{prop:b1:vspaces:coordinates}{coordinates} $(0, 11, 9, 0)$ (using $\sqrt{12} = 2\sqrt3$,
$\sqrt{18} = 3\sqrt2$). Finally $s^4 = (s^2)^2 = 25 + 20\sqrt6 +
24 = 49 + 20\sqrt6$: \omterm{prop:b1:vspaces:coordinates}{coordinates} $(49, 0, 0, 20)$.

\textbf{15.} Suppose $a + bs + cs^2 + ds^3 = 0$. Reading the four
\omterm{prop:b1:vspaces:coordinates}{coordinates} on $(1, \sqrt2, \sqrt3, \sqrt6)$:
\[
a + 5c = 0,\qquad b + 11d = 0,\qquad b + 9d = 0,\qquad 2c = 0 .
\]
So $c = 0$, then $a = 0$; subtracting the middle equations, $2d =
0$, then $b = 0$: $(1, s, s^2, s^3)$ is \omterm{def:b1:vspaces:free}{free}. Four \omterm{def:b1:vspaces:free}{free} vectors
in $M$ of dimension $4$: a \omterm{def:b1:vspaces:free}{basis}, so
$\operatorname{Vect}_\Q(1, s, s^2, s^3) = M$. From question 14,
$s^3 - 9s = 2\sqrt2$ and $11s - s^3 = 2\sqrt3$:
\[
\sqrt2 = \frac{s^3 - 9s}{2},
\qquad
\sqrt3 = \frac{11s - s^3}{2} .
\]

\textbf{16.} $s^4 - 10s^2 + 1 = (49 + 20\sqrt6) - 10(5 +
2\sqrt6) + 1 = 0$. A \omterm{def:b1:poly:def}{monic polynomial} of degree $\leq 3$
vanishing at $s$ would produce a nontrivial null combination of
$(1, s, s^2, s^3)$, contradicting question 15: $X^4 - 10X^2 + 1$
has least degree. Its roots: $X^2 = 5 \pm 2\sqrt6 = (\sqrt3 \pm
\sqrt2)^2$, so the four real roots are $\pm(\sqrt3 + \sqrt2)$ and
$\pm(\sqrt3 - \sqrt2)$, i.e.\ $\pm\sqrt2 \pm \sqrt3$.

\textbf{17.} From $s^4 - 10s^2 + 1 = 0$: $s\,(10s - s^3) = 1$, so
\[
\frac1s = 10s - s^3 = 10(\sqrt2 + \sqrt3) - (11\sqrt2 + 9\sqrt3)
= \sqrt3 - \sqrt2 ,
\]
consistent with $(\sqrt3 + \sqrt2)(\sqrt3 - \sqrt2) = 1$. If $a +
b\sqrt2 + c\sqrt3 + d\sqrt6 = r \in \Q$, then $(a - r) + b\sqrt2
+ c\sqrt3 + d\sqrt6 = 0$, and freeness (question 8) forces $b = c
= d = 0$ (and $a = r$). For $\sqrt2 + \sqrt3 + \sqrt6$ the
\omterm{prop:b1:vspaces:coordinates}{coordinates} $(0, 1, 1, 1)$ are not of this form: irrational ---
a \omterm{def:b1:logic:statement}{statement} stronger than \cref{exo:b1:reals:7}, obtained without
any squaring tricks.

\textbf{18.} A rational root $p/q$ (lowest terms) of $X^3 - 2$
satisfies $p^3 = 2q^3$, and the criterion of
\cref{exo:b1:poly:5} gives $p \mid 2$, $q \mid 1$: candidates
$\pm1, \pm2$, whose cubes are $\pm1, \pm8 \neq 2$. No rational
root; in particular $t = 2^{1/3} \notin \Q$, so $(1, t)$ is \omterm{def:b1:vspaces:free}{free}
over $\Q$ (as in question 2).

\textbf{19.} Suppose $(1, t, t^2)$ linked: some nonzero $P \in
\Q[X]$ with $\deg P \leq 2$ has $P(t) = 0$; choose such a $P$ of
\emph{least} degree $d \geq 1$. By question 18, $d \neq 1$, so $d
= 2$. Euclidean division (\cref{thm:b1:poly:division}): $X^3 - 2
= PQ + R$ with $Q, R \in \Q[X]$, $\deg R < 2$. Evaluating at $t$:
$0 = P(t)Q(t) + R(t) = R(t)$, so $R$ vanishes at $t$; minimality
of $d$ forces $R = 0$. Then $X^3 - 2 = PQ$ with $\deg Q = 1$: the
rational root of $Q$ is a rational root of $X^3 - 2$,
contradicting question 18. Hence $(1, t, t^2)$ is \omterm{def:b1:vspaces:free}{free} and
$\dim_\Q A = 3$. Stability: $t^3 = 2$ reduces every product of
$1, t, t^2$ to a combination of them ($t\cdot t^2 = 2$, $t^2\cdot
t^2 = 2t$); $A$ contains $1$: by question 12, $A$ is a subfield
of $\R$.

\textbf{20.} Suppose $t \in M$. Then $t^2 \in M$ (stability), so
$A \subseteq M$, and $M$ is a \omterm{def:b1:vspaces:def}{vector space} over the \omterm{def:b1:structures:field}{field} $A$:
the axioms are those of $\R$-arithmetic, restricted. It is
\omterm{def:b1:findim:def}{finite-dimensional} over $A$ (a finite $\Q$-generating family
generates a fortiori over $A \supseteq \Q$). The \omterm{pb:b1:findim:1}{tower law} for
$\Q \subseteq A \subseteq M$ gives
\[
4 = \dim_\Q M = \dim_\Q A \cdot \dim_A M = 3\,\dim_A M ,
\]
impossible: $3$ does not divide $4$. So $2^{1/3} \notin M$: no
rational combination of $1, \sqrt2, \sqrt3, \sqrt6$ equals
$2^{1/3}$.

\textbf{21.} $\Q(\sqrt2) \subseteq M$, so $t \notin \Q(\sqrt2)$.
If $t + \sqrt2 = r \in \Q$, then $t = r - \sqrt2 \in \Q(\sqrt2)$:
contradiction --- $2^{1/3} + \sqrt2$ is irrational. If $t = a +
b\sqrt3$, then $t \in \operatorname{Vect}_\Q(1, \sqrt3) \subseteq
M$: contradiction again.

\textbf{22.} $B$ is a \omterm{def:b1:vspaces:def}{vector space} over the \omterm{def:b1:structures:field}{field} $A$
(restriction of scalars), \omterm{def:b1:findim:def}{finite-dimensional} since $\dim_\Q B$ is
finite. The \omterm{pb:b1:findim:1}{tower law} for $\Q \subseteq A \subseteq B$ gives
$\dim_\Q B = \dim_\Q A \cdot \dim_A B$: the left factor \omterm{def:b1:arith:divides}{divides}.
Subfields of $M$ therefore have $\Q$-dimension $1$, $2$ or $4$:
dimension $1$ is $\Q$ itself, dimension $4$ is $M$.

\textbf{23.} Let $F \subseteq M$ be a subfield with $\dim_\Q F =
2$, and $w \in F \setminus \Q$: $(1, w)$ is \omterm{def:b1:vspaces:free}{free}, hence a \omterm{def:b1:vspaces:free}{basis}
of $F$. By question 13 ($d = 2$), $w^2 = \alpha + \beta w$ for
rationals $\alpha, \beta$. Setting $v = w - \beta/2 \in F
\setminus \Q$:
\[
v^2 = w^2 - \beta w + \frac{\beta^2}4 = \alpha + \frac{\beta^2}4
\;\in\; \Q,
\]
and $F = \operatorname{Vect}(1, v)$. Now write $v = a + b\sqrt2 +
c\sqrt3 + d\sqrt6$ and expand:
\[
v^2 = (a^2 + 2b^2 + 3c^2 + 6d^2)
+ (2ab + 6cd)\sqrt2 + (2ac + 4bd)\sqrt3 + (2ad + 2bc)\sqrt6 .
\]
The three irrational \omterm{prop:b1:vspaces:coordinates}{coordinates} vanish: $ab + 3cd = ac + 2bd =
ad + bc = 0$. If $a \neq 0$: $b = -3cd/a$, and the second
condition becomes $c(a^2 - 6d^2) = 0$; since $a^2 = 6d^2$ with $d
\neq 0$ would make $\sqrt6 = \abs{a/d}$ rational, either $c = 0$
or $d = 0$, and in both cases the remaining conditions force $b =
c = d = 0$, i.e.\ $v \in \Q$: excluded. So $a = 0$, and the
conditions read $3cd = 2bd = bc = 0$: at most one of $b, c, d$ is
nonzero. Hence $v$ is a rational multiple of $\sqrt2$, $\sqrt3$
or $\sqrt6$, and
\[
F \in \bigl\{\Q(\sqrt2),\ \Q(\sqrt3),\ \Q(\sqrt6)\bigr\} ,
\]
each of which is indeed a $2$-dimensional subfield (stability as
in question 6, inverses by question 12): exactly three quadratic
subfields.

\textbf{24.} $(1 + \sqrt2 + \sqrt3)(1 + \sqrt2 - \sqrt3) = (1 +
\sqrt2)^2 - 3 = 2\sqrt2$, so
\[
\frac1{1 + \sqrt2 + \sqrt3}
= \frac{1 + \sqrt2 - \sqrt3}{2\sqrt2}
= \frac{\sqrt2 + 2 - \sqrt6}{4}
= \frac12 + \frac14\sqrt2 + 0\cdot\sqrt3 - \frac14\sqrt6 .
\]
(Check: $(1 + \sqrt2 + \sqrt3)(2 + \sqrt2 - \sqrt6) = 4$ after
expansion.)

\textbf{25.} (i) $\Q$-dimension attaches an \emph{integer} to
each subfield of $\R$, and the \omterm{pb:b1:findim:1}{tower law} makes these integers
multiply along inclusions: dimension behaves like an arithmetic
invariant, and \omterm{def:b1:arith:divides}{divisibility} constraints become impossibility
proofs. (ii) \omterm{def:b1:findim:def}{Finite dimension} forces every element to satisfy a
nonzero rational \omterm{def:b1:poly:def}{polynomial} equation of degree at most the
dimension: finiteness means algebraicity. (iii) The
two-out-of-three rule turned ``multiplication by $u$ \omterm{def:b1:logic:map}{maps} a \omterm{def:b1:vspaces:free}{basis}
to a \omterm{def:b1:vspaces:free}{free family} of maximal size'' into surjectivity, producing
$u^{-1}$ with no formula: inverses came from counting. (iv) The
\omterm{pb:b1:findim:1}{tower law} forbids a $3$-dimensional \omterm{def:b1:structures:field}{field} inside a
$4$-dimensional one, which is why $2^{1/3}$ cannot be reached
from $\sqrt2$ and $\sqrt3$. The theorem of Part III is
\emph{Dedekind's \omterm{pb:b1:findim:1}{tower law}}, the opening move of Galois theory,
developed in the Year~3 volume.
\end{solution}
