\chapter{Problems and Charts}\label{ch:g5:data}

The last chapter of the year puts everything to work: multi-step
problems with money, quantities and durations, and the charts that
present information at a glance. The star idea is ``per one'' ---
finding the price of a single item --- which will grow into
proportionality in \cref{ch:g6:propdata}.

\section{Multi-step problems}

\begin{method}[Taming a long problem]\label{met:g5:data:steps}
\begin{enumerate}
  \item Read to the end; say what the final question is;
  \item plan backwards: to answer it, what do I need first?
  \item solve the steps in order, writing one operation and one
        short sentence per step;
  \item check the final answer against common sense and against an
        estimate.
\end{enumerate}
\end{method}

\begin{example}\label{ex:g5:data:multistep}
``A school orders $4$ boxes of $25$ books at $6$ each, and pays
$180$ for delivery on top. What is the total cost?''
\begin{enumerate}
  \item Books: $4 \times 25 = 100$ books.
  \item Book cost: $100 \times 6 = 600$.
  \item Total: $600 + 180 = 780$.
\end{enumerate}
Estimate check: about a hundred books at $6$ is about $600$, plus
about $200$: about $800$ --- the answer $780$ is plausible.
\end{example}

\begin{example}[Per one]\label{ex:g5:data:perone}
``$5$ identical mugs cost $12.50$; what do $8$ mugs cost?''
Find the price of \emph{one} mug first:
\[
12.50 \div 5 = 2.50,
\qquad\text{then}\qquad
8 \times 2.50 = 20 .
\]
Through one, everything becomes reachable --- this two-step pattern
solves a whole family of problems (recipes, speeds, prices), and is
the seed of \cref{ch:g6:propdata}.
\end{example}

\begin{example}[Change and budget]\label{ex:g5:data:budget}
Zoe has $30$. She buys a book at $12.40$ and two pens at $2.30$
each. Can she also afford a $13$ board game?
\begin{enumerate}
  \item Pens: $2 \times 2.30 = 4.60$.
  \item Spent so far: $12.40 + 4.60 = 17$.
  \item Remaining: $30 - 17 = 13$ --- exactly enough for the game,
        with nothing left.
\end{enumerate}
\end{example}

\section{Reading charts}

\begin{method}[Reading any chart]\label{met:g5:data:read}
\begin{enumerate}
  \item Read the title: what is being counted?
  \item Read the axes or the key: what does one graduation, one bar,
        one symbol stand for?
  \item Only then answer questions --- pointing at the chart with a
        ruler helps read heights exactly.
\end{enumerate}
\end{method}

\begin{example}\label{ex:g5:data:chart}
Rainfall in a town, month by month:

\begin{omfigure}
\begin{tikzpicture}
\begin{axis}[
  width=8.8cm, height=5.0cm,
  ybar, bar width=13pt,
  symbolic x coords={Jan,Feb,Mar,Apr,May,Jun},
  xtick=data,
  ymin=0, ymax=90,
  ytick={20,40,60,80},
  ylabel={rain (mm)},
  tick label style={font=\small},
  label style={font=\small}]
  \addplot[fill=omDef!30, draw=omDef] coordinates
    {(Jan,60) (Feb,45) (Mar,55) (Apr,70) (May,80) (Jun,30)};
\end{axis}
\end{tikzpicture}

{\small A bar chart of rainfall. One graduation $= 20$ mm.}
\end{omfigure}

Readings: the wettest month is May ($80$ mm); June got half of
April's rain ($30$ against $70$? no --- half of $70$ is $35$, so
\emph{not quite} half: reading precisely matters!); total for the
six months: $60 + 45 + 55 + 70 + 80 + 30 = 340$ mm.
\end{example}

\begin{example}[A changing quantity]\label{ex:g5:data:line}
A plant's height, measured every week:

\begin{omfigure}
\begin{tikzpicture}
\begin{axis}[omaxis,
  width=8.4cm, height=5.0cm,
  xmin=0, xmax=6.5, ymin=0, ymax=24,
  xlabel={week}, ylabel={height (cm)},
  xtick={1,...,6}, ytick={5,10,15,20}]
  \addplot[omDef, thick, mark=*, mark size=1.6pt] coordinates
    {(1,3) (2,6) (3,10) (4,15) (5,19) (6,21)};
\end{axis}
\end{tikzpicture}

{\small Joining the measurement points shows the \emph{growth}: the
steeper the segment, the faster the plant grew that week.}
\end{omfigure}

The plant grew fastest between weeks $3$ and $4$ ($+5$ cm) and
slowed down at the end ($+2$ cm in week $6$).
\end{example}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:g5:data:1}
Solve in steps: ``A club rents a bus for $240$ and buys $32$ museum
tickets at $7$ each. What is the total cost of the trip?''
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g5:data:2}
Solve with ``per one'': ``$3$ kg of apples cost $7.50$. What do
$5$ kg cost?''
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g5:data:3}
Solve: ``Six friends share equally the $87$ cost of a picnic. How
much does each pay?''
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g5:data:4}
Sam has $25$. He buys a T-shirt at $13.90$ and a cap at $8.50$. How
much money has he left?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g5:data:5}
Using the rainfall chart of \cref{ex:g5:data:chart}: which months
got more than $50$ mm? How much more rain fell in May than in
June?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g5:data:6}
Using the plant chart of \cref{ex:g5:data:line}: how tall was the
plant in week $2$? Between which weeks did it grow by exactly
$4$ cm?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g5:data:7}
Draw a bar chart for the number of goals a team scored:
September $4$, October $7$, November $3$, December $6$. Choose the
graduation, and give the total.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g5:data:8}
A pack of $8$ yogurts costs $3.20$; sold alone, one yogurt costs
$0.50$. Compute the ``per one'' price in the pack, and how much one
saves per yogurt by buying the pack.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g5:data:9}
``A cinema sold $148$ tickets at $9.50$ on Saturday and $95$ tickets
at the same price on Sunday. How much more money did Saturday bring
than Sunday?'' Solve it two ways: computing both days' takings, or
computing with the \emph{difference} in tickets --- and check that
the answers agree.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g5:data:10}
A recipe for $4$ people needs $300$ g of rice. Aline cooks for
$10$ people. How much rice does she need? (Per one person first ---
decimals allowed.)
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g5:data:11}
Invent a three-step problem using money whose final answer is
$5.50$, write its full solution in the style of
\cref{met:g5:data:steps}, and test it on a classmate.
\end{exercise}
