\chapter{Standard Functions}\label{ch:b1:functions}

Analysis is only as useful as the stock of functions one masters. To
the collection inherited from secondary school --- powers, exponential,
logarithm, trigonometric functions --- this chapter adds their inverse
functions ($\arcsin$, $\arccos$, $\arctan$) and the hyperbolic family.
Derivatives are used freely at High School level; the
theory behind inverse functions is completed in
\cref{ch:b1:continuity,ch:b1:derivative}.

\section{Exponential, logarithm, powers}

\begin{proposition}[Recap and characterization]\label{prop:b1:functions:expln}
$\exp \colon \R \to \intoo{0}{+\infty}$ and $\ln \colon
\intoo{0}{+\infty} \to \R$ are reciprocal bijections, strictly
increasing, with
\[
\exp(x + y) = \exp x \exp y, \qquad \ln(xy) = \ln x + \ln y,
\qquad \exp' = \exp, \qquad \ln'(x) = \frac 1x .
\]
Moreover $\exp$ is the \emph{only} differentiable function $f \colon
\R \to \R$ with $f' = f$ and $f(0) = 1$.
\end{proposition}

\begin{proof}
The recap is High School material. For uniqueness, let $f' = f$,
$f(0) = 1$, and set $g(x) = f(x)\,\eu^{-x}$. Then $g' = f'\eu^{-x} -
f\eu^{-x} = 0$, so $g$ is constant equal to $g(0) = 1$: $f = \exp$.
\end{proof}

\begin{definition}[General powers]\label{def:b1:functions:powers}
For $x > 0$ and $\alpha \in \R$:
$\;x^\alpha = \eu^{\alpha \ln x}$\index{power function}. For $a > 0$,
$a \neq 1$, the \emph{logarithm to base $a$} is $\log_a x =
\frac{\ln x}{\ln a}$, the inverse of $x \mapsto a^x$.
\end{definition}

\begin{example}[Solving exponential equations]\label{ex:b1:functions:expequations}
Solve $2^x = 5^{\,x-1}$ in $\R$. Both sides are positive, so take
logarithms --- a reversible step:
\[
x\ln2 = (x - 1)\ln5
\iff x(\ln2 - \ln5) = -\ln5
\iff x = \frac{\ln5}{\ln5 - \ln2} = \frac{\ln 5}{\ln\frac52}
\approx 1.756 .
\]
Next, solve $x^{\sqrt2} = 3$ for $x > 0$: raise to the power
$\frac1{\sqrt2}$ (i.e.\ apply the reciprocal bijection):
$x = 3^{1/\sqrt2} = \eu^{(\ln 3)/\sqrt2} \approx 2.175$. The
insight: every equation mixing powers unwinds through $\ln$ and
$\exp$, because the definition $x^\alpha = \eu^{\alpha\ln x}$
reduces all power manipulations to arithmetic of exponents ---
but only on the domain $x > 0$ where that definition lives.
\end{example}

\begin{example}[Doubling times]\label{ex:b1:functions:doubling}
A quantity grows by $3\%$ per step: after $n$ steps it is
multiplied by $(1.03)^n$. When does it double? Solve $(1.03)^n
\geq 2$:
\[
n\ln(1.03) \geq \ln2
\iff n \geq \frac{\ln 2}{\ln 1.03}
= \frac{0.6931}{0.02956} \approx 23.4 ,
\]
so the first doubling occurs at step $24$. (The financiers'
``rule of $72$'', which estimates the doubling time as
$72$ divided by the rate in percent, is this computation with
the approximation $\ln(1 + x) \approx x$, quantified in
\cref{ch:b1:taylor}.) Exponential
processes are best reasoned about through their logarithms: on
that scale, growth is linear and questions become divisions.
\end{example}

\begin{example}[How long is $2^{2026}$?]\label{ex:b1:functions:digits}
The number of decimal digits of an integer $N \geq 1$ is
$\floor{\log_{10} N} + 1$ (indeed $N$ has $d$ digits exactly when
$10^{d-1} \leq N < 10^d$, i.e.\ $d - 1 \leq \log_{10}N < d$). For
$N = 2^{2026}$:
\[
\log_{10} 2^{2026} = 2026\,\log_{10}2
= 2026 \times 0.301030 = 609.887 ,
\]
so $2^{2026}$ has $610$ digits. The fractional part carries a
bonus: $10^{0.887} \approx 7.7$, so the number \emph{begins} with
$7$. One multiplication answered a question about a number no one
will ever write out --- logarithms compress multiplicative size
into additive size, which is their whole historical point
(\cref{ex:b1:structures:expmorphism} makes that sentence precise).
\end{example}

\begin{proposition}[Power rules and growth comparison]\label{prop:b1:functions:powerrules}
For $x, y > 0$ and $\alpha, \beta \in \R$:
\[
x^{\alpha+\beta} = x^\alpha x^\beta, \quad
(x^\alpha)^\beta = x^{\alpha\beta}, \quad
(xy)^\alpha = x^\alpha y^\alpha, \quad
(x^\alpha)' = \alpha\, x^{\alpha - 1}.
\]
Growth scale\index{growth comparison} as $x \to +\infty$, for every
$\alpha > 0$ and $\beta > 0$:
\[
\frac{(\ln x)^\beta}{x^\alpha} \longrightarrow 0,
\qquad
\frac{x^\alpha}{\eu^{\beta x}} \longrightarrow 0 .
\]
\end{proposition}

\begin{proof}
The identities transcribe those of $\exp$ and $\ln$ through the
definition. In detail for the second (the least obvious): $x^\alpha
> 0$ and $\ln(x^\alpha) = \alpha\ln x$ (apply $\ln$ to the
definition), so
\[
(x^\alpha)^\beta = \eu^{\beta\ln(x^\alpha)}
= \eu^{\beta\alpha\ln x} = x^{\alpha\beta} ;
\]
the first and third are the same two-line unwindings via
$\exp(u + v) = \exp u\exp v$ and $\ln(xy) = \ln x + \ln y$. The
derivative is the chain rule: $(\eu^{\alpha\ln x})' =
\frac{\alpha}{x} \eu^{\alpha\ln x}$.

Comparisons: from $\frac{\ln t}{t} \to 0$ (proved in the High
School volume), substitute $t = x^{\alpha/\beta}$:
$\frac{\ln x}{x^{\alpha/\beta}} \to 0$, then raise to the power
$\beta$. For the second, substitute $x = \ln u$ in the first.
\end{proof}

\begin{example}[Two limits every reader should own]\label{ex:b1:functions:xx}
What are $\lim_{x \to 0^+} x^x$ and $\lim_{x \to +\infty}
x^{1/x}$? Both powers are defined through the exponential, so
settle the exponent first:
\[
x^x = \eu^{x\ln x}, \qquad
x\ln x = -\frac{\ln(1/x)}{1/x} \longrightarrow 0
\quad (x \to 0^+),
\qquad\text{so } x^x \longrightarrow \eu^0 = 1 ;
\]
and $x^{1/x} = \eu^{(\ln x)/x} \to \eu^0 = 1$ as $x \to +\infty$,
directly by the \omterm{prop:b1:functions:powerrules}{growth comparison}. The insight: an indeterminate
power (of shape $0^0$ or $\infty^0$) is \emph{always} handled by
rewriting $u^v = \eu^{v\ln u}$ and analyzing the product $v \ln u$
--- never by guessing from base and exponent separately. The
function $x \mapsto \frac{\ln x}{x}$ that decided both limits is
studied exhaustively in the weekend problem of this chapter.
\end{example}

\begin{omfigure}
\begin{tikzpicture}
\begin{axis}[omaxis, width=9.5cm, height=6cm,
    xmin=0, xmax=5.2, ymin=-1.8, ymax=8.5,
    xtick={1,2,3,4,5}, ytick={2,4,6,8}]
  \addplot[omMeth, thick, domain=0.02:5, samples=200] {ln(x)}
    node[pos=0.97, below, font=\small] {$\ln x$};
  \addplot[omDef, thick, domain=0:5, samples=100] {sqrt(x)}
    node[pos=0.9, above, font=\small] {$\sqrt x$};
  \addplot[omProp, thick, domain=0:5] {x}
    node[pos=0.72, left, font=\small] {$x$};
  \addplot[omThm, thick, domain=0:2.14, samples=120] {exp(x)}
    node[pos=0.85, left, font=\small] {$\eu^x$};
\end{axis}
\end{tikzpicture}

{\small The growth scale of
\cref{prop:b1:functions:powerrules}: near the right edge, $\ln x$
has barely passed $1.6$ while $\eu^x$ has left the frame. Every
ratio (log over power, power over exponential) tends to $0$ ---
the picture only hints at what the substitutions in the proof make
exact.}
\end{omfigure}

\begin{remark}[Common pitfalls with powers and logarithms]\label{rem:b1:functions:pitfalls}
\begin{enumerate}
  \item \emph{Domains.} $x^\alpha$ for irrational $\alpha$ needs
        $x > 0$; $(-8)^{1/3}$ is best avoided in favor of ``the
        real cube root of $-8$'', because the rule
        $(x^\alpha)^\beta = x^{\alpha\beta}$ silently fails on
        negatives: $\bigl((-8)^2\bigr)^{1/6} = 2 \neq -2$.
  \item \emph{$\sqrt{x^2} = \abs x$}, not $x$: forgetting the
        absolute value is the classic source of lost negative
        solutions.
  \item \emph{$\ln(xy) = \ln x + \ln y$ requires $x, y > 0$.} For
        negative arguments, $\ln(xy)$ can be defined while the
        right-hand side is not.
  \item \emph{Which function grows?} In $x^\alpha$ the
        \emph{base} varies; in $a^x$ the \emph{exponent} does.
        Their growths are wildly different
        (\cref{prop:b1:functions:powerrules}), and hybrid
        expressions like $x^{\ln x}$ or $x^{1/x}$ must be
        rewritten as $\eu^{(\cdot)}$ before any reasoning ---
        as in \cref{ex:b1:functions:xx}.
\end{enumerate}
\end{remark}

\section{Inverse trigonometric functions}

\begin{definition}[$\arcsin$, $\arccos$, $\arctan$]\label{def:b1:functions:arc}
The restrictions
\[
\sin \colon \intcc{-\tfrac\pi2}{\tfrac\pi2} \to \intcc{-1}{1},
\qquad
\cos \colon \intcc{0}{\pi} \to \intcc{-1}{1},
\qquad
\tan \colon \intoo{-\tfrac\pi2}{\tfrac\pi2} \to \R
\]
are strictly monotonic bijections. Their inverse \omterm{def:b1:logic:map}{maps} are written
$\arcsin$\index{arcsine}, $\arccos$\index{arccosine} and
$\arctan$\index{arctangent}. Thus, for instance, $y = \arcsin x$
($x \in \intcc{-1}{1}$) is \emph{the} angle in
$\intcc{-\frac\pi2}{\frac\pi2}$ whose sine is $x$.
\end{definition}

\begin{omfigure}
\begin{tikzpicture}
\begin{axis}[omaxis, width=5.6cm, height=5.6cm,
    xmin=-1.25, xmax=1.25, ymin=-1.8, ymax=1.8,
    xtick={-1,1}, ytick={-1.5708,1.5708},
    yticklabels={$-\frac{\pi}{2}$,$\frac{\pi}{2}$}]
  \addplot[omDef, very thick, domain=-1:1, samples=200]
    {rad(asin(x))};
\end{axis}
\begin{scope}[xshift=5.3cm]
\begin{axis}[omaxis, width=5.6cm, height=5.6cm,
    xmin=-1.25, xmax=1.25, ymin=-0.4, ymax=3.4,
    xtick={-1,1}, ytick={1.5708,3.1416},
    yticklabels={$\frac{\pi}{2}$,$\pi$}]
  \addplot[omThm, very thick, domain=-1:1, samples=200]
    {rad(acos(x))};
\end{axis}
\end{scope}
\begin{scope}[xshift=10.6cm]
\begin{axis}[omaxis, width=5.6cm, height=5.6cm,
    xmin=-4, xmax=4, ymin=-1.9, ymax=1.9,
    xtick={-2,2}, ytick={-1.5708,1.5708},
    yticklabels={$-\frac{\pi}{2}$,$\frac{\pi}{2}$}]
  \addplot[omMeth, very thick, domain=-4:4, samples=200]
    {rad(atan(x))};
  \addplot[dashed, gray, domain=-4:4] {1.5708};
  \addplot[dashed, gray, domain=-4:4] {-1.5708};
\end{axis}
\end{scope}
\end{tikzpicture}

{\small From left to right: $\arcsin$, $\arccos$, $\arctan$. Each
inherits its graph from the restricted direct function by reflection
in the line $y = x$.}
\end{omfigure}

\begin{proposition}[Derivatives]\label{prop:b1:functions:arcderiv}
On the interior of their domains:
\[
\arcsin' x = \frac{1}{\sqrt{1 - x^2}},
\qquad
\arccos' x = \frac{-1}{\sqrt{1 - x^2}},
\qquad
\arctan' x = \frac{1}{1 + x^2}.
\]
\end{proposition}

\begin{proof}
Anticipating the inverse-function rule proved in
\cref{ch:b1:derivative}: if $f$ is a differentiable bijection and
$f' \neq 0$, then $(f^{-1})'(x) = \frac{1}{f'(f^{-1}(x))}$. For
$\arcsin$: with $y = \arcsin x$,
\[
\arcsin' x = \frac{1}{\cos y}
= \frac{1}{\sqrt{1 - \sin^2 y}} = \frac{1}{\sqrt{1 - x^2}},
\]
where $\cos y = +\sqrt{1 - \sin^2 y}$ because $y \in
\intoo{-\frac\pi2}{\frac\pi2}$ forces $\cos y > 0$. Similarly
$\arccos' x = \frac{1}{-\sin y}$ with $\sin y > 0$ on
$\intoo{0}{\pi}$, and $\arctan' x = \frac{1}{1 + \tan^2 y} =
\frac{1}{1 + x^2}$ using $\tan' = 1 + \tan^2$.
\end{proof}

\begin{example}[Recognizing a hidden constant]\label{ex:b1:functions:hiddenconst}
Study $g(x) = \arctan\dfrac{1 - x}{1 + x}$ on $\intoo{-1}
{+\infty}$. Chain rule and a short computation:
\[
g'(x) = \frac{1}{1 + \bigl(\frac{1-x}{1+x}\bigr)^2}\cdot
\frac{-(1+x) - (1-x)}{(1+x)^2}
= \frac{-2}{(1+x)^2 + (1-x)^2}
= \frac{-1}{1 + x^2} ,
\]
since $(1+x)^2 + (1-x)^2 = 2 + 2x^2$. So $g' = -\arctan'$: the
function $g + \arctan$ has zero derivative on the \emph{interval}
$\intoo{-1}{+\infty}$, hence is constant there; its value at $x =
0$ is $\arctan 1 + \arctan 0 = \frac\pi4$. Conclusion:
\[
\arctan\frac{1 - x}{1 + x} = \frac\pi4 - \arctan x
\qquad (x > -1) .
\]
On $\intoo{-\infty}{-1}$ the same derivative computation holds but
the constant is different ($-\frac{3\pi}4$: evaluate the limit as
$x \to -\infty$). The insight: ``zero derivative implies
constant'' is an interval-by-interval \omterm{def:b1:logic:statement}{statement} --- exactly the
subtlety exploited in \cref{exo:b1:functions:6} and
\cref{exo:b1:functions:10}.
\end{example}

\begin{proposition}[Standard identities]\label{prop:b1:functions:arcidentities}
\begin{enumerate}
  \item For $x \in \intcc{-1}{1}$: $\arcsin x + \arccos x =
        \dfrac{\pi}{2}$.
  \item For $x > 0$: $\arctan x + \arctan\dfrac 1x = \dfrac{\pi}{2}$
        (and $-\frac\pi2$ for $x < 0$).
  \item $\sin(\arccos x) = \cos(\arcsin x) = \sqrt{1 - x^2}$;
        $\;\tan(\arcsin x) = \dfrac{x}{\sqrt{1 - x^2}}$ for
        $\abs x < 1$.
\end{enumerate}
\end{proposition}

\begin{proof}
(1) The derivative of $x \mapsto \arcsin x + \arccos x$ is zero on
$\intoo{-1}{1}$ (\cref{prop:b1:functions:arcderiv}), so the function
is constant there, equal to its value $\frac\pi2$ at $0$; the endpoint
values $\pm 1$ are checked directly ($\frac\pi2 + 0$ and $-\frac\pi2 +
\pi$).

(2) Same method on $\intoo{0}{+\infty}$: the derivative is
$\frac{1}{1+x^2} + \frac{-1/x^2}{1 + 1/x^2} = \frac{1}{1+x^2} -
\frac{1}{x^2 + 1} = 0$, and at $x = 1$ the sum is $2\arctan 1 =
\frac\pi2$. For $x < 0$, use oddness of $\arctan$.

(3) With $y = \arccos x \in \intcc{0}{\pi}$: $\sin y \geq 0$, so
$\sin y = \sqrt{1 - \cos^2 y} = \sqrt{1 - x^2}$; likewise for
$\cos(\arcsin x)$, and the tangent formula is the quotient.
\end{proof}

\begin{remark}
$\arcsin(\sin\theta) = \theta$ holds \emph{only} for $\theta \in
\intcc{-\frac\pi2}{\frac\pi2}$: for instance
$\arcsin(\sin\pi) = 0 \neq \pi$. The composition in the other
direction, $\sin(\arcsin x) = x$, is valid on all of
$\intcc{-1}{1}$.
\end{remark}

\begin{example}[Wrapping back to the principal interval]
\label{ex:b1:functions:wrap}
Compute
\[
\arctan\Bigl(\tan\frac{3\pi}4\Bigr)
\qquad\text{and}\qquad
\arctan\Bigl(\tan\frac{17\pi}5\Bigr).
\]
The recipe: replace the
angle by the unique angle of $\intoo{-\frac\pi2}{\frac\pi2}$ with
the same tangent, i.e.\ subtract the right multiple of $\pi$
(the period of $\tan$). First: $\frac{3\pi}4 - \pi =
-\frac\pi4$, so the answer is $-\frac\pi4$. Second: $\frac{17\pi}
5 - 3\pi = \frac{2\pi}5 \in \intoo{-\frac\pi2}{\frac\pi2}$, so
the answer is $\frac{2\pi}5$. The computation is a Euclidean
division of the angle by $\pi$ in disguise --- and the analogous
recipes for $\arcsin$ (reflect into
$\intcc{-\frac\pi2}{\frac\pi2}$, period $2\pi$) and $\arccos$
(reflect into $\intcc0\pi$) drive \cref{exo:b1:functions:1} and
the piecewise answer of \cref{exo:b1:functions:10}.
\end{example}

\begin{example}[Adding arctangents safely]\label{ex:b1:functions:arctansum}
Let us prove
\[
\arctan\frac12 + \arctan\frac15 + \arctan\frac18 = \frac\pi4 .
\]
Two ingredients: the tangent addition formula, and --- the step
beginners forget --- a \emph{localization} of the sum. First,
$\tan(u + v) = \frac{\tan u + \tan v}{1 - \tan u\tan v}$ with
$u = \arctan\frac12$, $v = \arctan\frac15$ gives
\[
\tan(u + v) = \frac{\frac12 + \frac15}{1 - \frac1{10}}
= \frac{7/10}{9/10} = \frac79,
\qquad\text{then}\qquad
\tan\Bigl(u + v + \arctan\frac18\Bigr)
= \frac{\frac79 + \frac18}{1 - \frac7{72}}
= \frac{65/72}{65/72} = 1 .
\]
Second, each of the three angles lies in $\intoo0{\frac\pi4}$
(their arguments are less than $1$), so the sum lies in
$\intoo0{\frac{3\pi}4}$; the only angle there with tangent $1$ is
$\frac\pi4$. Without the localization the conclusion would be
``$\frac\pi4$ up to a multiple of $\pi$'' --- half a proof. The
same two-step discipline runs \cref{exo:b1:functions:8} and
\cref{exo:b1:functions:12}.
\end{example}

\begin{example}[An arctangent equation]\label{ex:b1:functions:arctaneq}
Solve $\arctan x + \arctan 2x = \dfrac\pi4$. First localize: the
left side has the sign of $x$ (both terms do), so any solution has
$x > 0$, and then the sum lies in $\intoo0\pi$. Take tangents
(\omterm{def:b1:logic:inj}{injective} on no interval of length $\pi$, but combined with the
localization this will suffice): the addition formula gives
\[
\tan\bigl(\arctan x + \arctan 2x\bigr)
= \frac{3x}{1 - 2x^2} = 1
\iff 2x^2 + 3x - 1 = 0
\iff x = \frac{-3 \pm \sqrt{17}}4 .
\]
The negative root is discarded by the localization; for the
positive candidate $x_0 = \frac{\sqrt{17} - 3}4 \approx 0.28$,
the sum lies in $\intoo0\pi$ with tangent $1$, and the only such
angle is $\frac\pi4$ (in $\intoo{\frac\pi2}\pi$ the tangent is
negative): $x_0$ is the unique solution. Note the shape of the
argument: taking tangents may create solutions, never lose them,
so one solves the polynomial equation \emph{then} filters by
localization --- the same forward-check discipline as squaring
an equation.
\end{example}

\section{Hyperbolic functions}

\begin{definition}[$\cosh$, $\sinh$, $\tanh$]\label{def:b1:functions:hyperbolic}
For $x \in \R$:
\[
\cosh x = \frac{\eu^x + \eu^{-x}}{2},
\qquad
\sinh x = \frac{\eu^x - \eu^{-x}}{2},
\qquad
\tanh x = \frac{\sinh x}{\cosh x}
\]
(\emph{hyperbolic cosine, sine, tangent}\index{hyperbolic
functions}). $\cosh$ is even, $\sinh$ and $\tanh$ are odd.
\end{definition}

\begin{proposition}[Basic properties]\label{prop:b1:functions:hyprules}
For all $x, y \in \R$:
\begin{enumerate}
  \item $\cosh^2 x - \sinh^2 x = 1$;
  \item $\cosh' = \sinh$, $\;\sinh' = \cosh$, $\;\tanh' = 1 - \tanh^2
        = \dfrac{1}{\cosh^2}$;
  \item $\cosh(x+y) = \cosh x\cosh y + \sinh x \sinh y$ and
        $\sinh(x+y) = \sinh x \cosh y + \cosh x \sinh y$;
  \item $\sinh$ is a strictly increasing bijection of $\R$ onto $\R$;
        $\cosh$ restricted to $\R_+$ is a strictly increasing
        bijection onto $\intco{1}{+\infty}$; $\tanh$ is a strictly
        increasing bijection of $\R$ onto $\intoo{-1}{1}$.
\end{enumerate}
\end{proposition}

\begin{proof}
(1) $\bigl(\frac{\eu^x + \eu^{-x}}{2}\bigr)^2 - \bigl(\frac{\eu^x -
\eu^{-x}}{2}\bigr)^2 = \frac{(\eu^{2x} + 2 + \eu^{-2x}) - (\eu^{2x} -
2 + \eu^{-2x})}{4} = 1$.

(2) Differentiate the defining formulas; for $\tanh$, the quotient
rule gives $\frac{\cosh^2 - \sinh^2}{\cosh^2}$, which is both
$1 - \tanh^2$ and $\frac{1}{\cosh^2}$ by (1).

(3) Expand the right-hand sides using the definitions. In detail
for the first formula:
\[
\cosh x\cosh y + \sinh x\sinh y
= \frac{(\eu^x + \eu^{-x})(\eu^y + \eu^{-y})
+ (\eu^x - \eu^{-x})(\eu^y - \eu^{-y})}{4} ;
\]
of the eight products, the four ``mixed'' ones
($\eu^{x-y}$, $\eu^{y-x}$) cancel in pairs, while
$\eu^{x+y}$ and $\eu^{-(x+y)}$ each appear twice:
the total is $\frac{2\eu^{x+y} + 2\eu^{-(x+y)}}{4} = \cosh(x+y)$.
The sine formula is identical with the mixed terms surviving
instead.

(4) $\sinh' = \cosh \geq 1 > 0$, so $\sinh$ is strictly increasing,
with limits $\pm\infty$ (dominant term $\pm\frac{\eu^{\abs
x}}{2}$); the bijection \omterm{def:b1:logic:statement}{statement} then follows from the intermediate
value theorem (used at High School level; systematic treatment in
\cref{ch:b1:continuity}). On $\R_+$, $\cosh' = \sinh \geq 0$
vanishing only at $0$: strictly increasing from $\cosh 0 = 1$ to
$+\infty$. And $\tanh' > 0$ with limits $\pm 1$ at $\pm\infty$: in
detail,
\[
\tanh x = \frac{\eu^x - \eu^{-x}}{\eu^x + \eu^{-x}}
= \frac{1 - \eu^{-2x}}{1 + \eu^{-2x}}
\xrightarrow[x \to +\infty]{} 1 ,
\]
after dividing numerator and denominator by $\eu^x$, and oddness
gives the limit $-1$ at $-\infty$; the strictly increasing
function $\tanh$ therefore \omterm{def:b1:logic:map}{maps} $\R$ onto $\intoo{-1}1$.
\end{proof}

\begin{omfigure}
\begin{tikzpicture}
\begin{axis}[omaxis, width=6.4cm, height=6cm,
    xmin=-2.6, xmax=2.6, ymin=-3.6, ymax=3.9,
    xtick={-2,-1,1,2}, ytick={-3,-2,-1,1,2,3}]
  \addplot[omDef, very thick, domain=-2.05:2.05, samples=120]
    {cosh(x)} node[pos=0.1, right, font=\small] {$\cosh$};
  \addplot[omThm, very thick, domain=-2.05:2.05, samples=120]
    {sinh(x)} node[pos=0.93, above left, font=\small] {$\sinh$};
  \addplot[dashed, gray, domain=0:2.5] {exp(x)/2};
\end{axis}
\begin{scope}[xshift=7.4cm]
\begin{axis}[omaxis, width=6.4cm, height=6cm,
    xmin=-3.2, xmax=3.2, ymin=-1.5, ymax=1.5,
    xtick={-2,-1,1,2}, ytick={-1,1}]
  \addplot[omMeth, very thick, domain=-3:3, samples=150]
    {tanh(x)} node[pos=0.9, above left, font=\small] {$\tanh$};
  \addplot[dashed, gray, domain=-3:3] {1};
  \addplot[dashed, gray, domain=-3:3] {-1};
\end{axis}
\end{scope}
\end{tikzpicture}

{\small Left: $\cosh$ and $\sinh$, asymptotically glued to
$\frac{\eu^x}{2}$ (dashed). Right: $\tanh$, increasing from $-1$ to
$1$.}
\end{omfigure}

\begin{remark}[Why ``hyperbolic''?]
The point $(\cosh t, \sinh t)$ runs along the branch $x > 0$ of the
hyperbola $x^2 - y^2 = 1$ (by
\cref{prop:b1:functions:hyprules}~(1)), exactly as $(\cos t, \sin
t)$ runs along the circle $x^2 + y^2 = 1$. Every circular identity has
a hyperbolic sibling, with sign changes governed by
$\sinh^2 \leftrightarrow -\sin^2$.
\end{remark}

\begin{example}[The addition formula for $\tanh$]\label{ex:b1:functions:tanhadd}
Dividing the two addition formulas of
\cref{prop:b1:functions:hyprules}~(3) by $\cosh x\cosh y$:
\[
\tanh(x + y)
= \frac{\sinh x\cosh y + \cosh x\sinh y}
{\cosh x\cosh y + \sinh x\sinh y}
= \frac{\tanh x + \tanh y}{1 + \tanh x\,\tanh y} ,
\]
the hyperbolic sibling of the tangent addition formula --- with a
$+$ where trigonometry has a $-$. A dividend: since $\abs{\tanh}
< 1$, the right-hand side is a ``velocity addition'' rule that
never leaves $\intoo{-1}1$: if $u, v \in \intoo{-1}1$ then
$\frac{u + v}{1 + uv} \in \intoo{-1}1$ as well (write $u = \tanh
a$, $v = \tanh b$, possible by bijectivity, and read the formula
backwards). Checking that algebraically, without \omterm{def:b1:functions:hyperbolic}{hyperbolic
functions}, is a slightly painful exercise; parametrizing by
$\tanh$ makes it one line --- the same strategy that circular
functions provide for the unit circle.
\end{example}

\begin{proposition}[Inverse hyperbolic functions]\label{prop:b1:functions:invhyp}
The inverses granted by \cref{prop:b1:functions:hyprules}~(4) have
closed forms:
\[
\operatorname{arsinh} x = \ln\bigl(x + \sqrt{x^2 + 1}\bigr)
\ (x \in \R),
\qquad
\operatorname{arcosh} x = \ln\bigl(x + \sqrt{x^2 - 1}\bigr)
\ (x \geq 1),
\]
\[
\operatorname{artanh} x = \frac 12 \ln\frac{1 + x}{1 - x}
\ (\abs x < 1),
\]
with derivatives $\frac{1}{\sqrt{x^2+1}}$, $\frac{1}{\sqrt{x^2-1}}$
($x > 1$) and $\frac{1}{1 - x^2}$ respectively.
\end{proposition}

\begin{proof}
For $\operatorname{arsinh}$: solve $x = \sinh y = \frac{\eu^y -
\eu^{-y}}{2}$. Setting $u = \eu^y > 0$: $u^2 - 2xu - 1 = 0$, so $u = x
+ \sqrt{x^2 + 1}$ (the root $x - \sqrt{x^2+1}$ is negative), and $y =
\ln(x + \sqrt{x^2+1})$. The other two are identical computations
($u^2 - 2xu + 1 = 0$ for $\operatorname{arcosh}$, keeping the root
$\geq 1$; a two-line rearrangement for $\operatorname{artanh}$).
Derivatives: differentiate the logarithmic expressions, e.g.
\[
\bigl(\ln(x + \sqrt{x^2+1})\bigr)'
= \frac{1 + \frac{x}{\sqrt{x^2+1}}}{x + \sqrt{x^2+1}}
= \frac{1}{\sqrt{x^2 + 1}} . \qedhere
\]
\end{proof}

\begin{example}[Closed forms at work]\label{ex:b1:functions:arcoshtwo}
The solution of $\cosh t = 2$ with $t \geq 0$ is, by the closed
form, $t = \operatorname{arcosh} 2 = \ln(2 + \sqrt3) \approx
1.317$; the other solution is $-t$, by evenness --- and indeed
$\ln(2 - \sqrt3) = \ln\frac1{2 + \sqrt3} = -\ln(2 + \sqrt3)$: the
two roots $u = \eu^t$ of the quadratic $u^2 - 4u + 1 = 0$ are
reciprocals, as their product $1$ (Vieta) demands. This tiny
computation displays the general pattern: hyperbolic equations
convert to quadratics in $\eu^t$, and the symmetry $t \mapsto -t$
appears as the symmetry $u \mapsto 1/u$ of the quadratic ---
worth remembering when solving \cref{exo:b1:functions:7}.
\end{example}

\begin{method}[Choosing the right primitive form]\label{met:b1:functions:primitive}
The three derivative patterns to memorize for integration
(\cref{ch:b1:integration}):
\[
\int \frac{\dd x}{1 + x^2} = \arctan x + C,
\qquad
\int \frac{\dd x}{\sqrt{1 - x^2}} = \arcsin x + C,
\qquad
\int \frac{\dd x}{\sqrt{x^2 + 1}} = \operatorname{arsinh} x + C,
\]
and for a general quadratic, reduce to these by completing the square
and rescaling.
\end{method}

\begin{remark}[Where this chapter is used]\label{rem:b1:functions:whereused}
This chapter is the working vocabulary of all the analysis to come.
The growth scale of \cref{prop:b1:functions:powerrules} decides
convergence questions throughout
\cref{ch:b1:seq,ch:b1:series}; the derivative formulas of
\cref{prop:b1:functions:arcderiv} and
\cref{prop:b1:functions:invhyp} are the primitives most often
needed in \cref{ch:b1:integration}, via
\cref{met:b1:functions:primitive}; \omterm{def:b1:functions:hyperbolic}{hyperbolic functions}
parametrize the solutions of the equation $y'' = y$ in
\cref{ch:b1:diffeq} exactly as circular functions parametrize
those of $y'' = -y$. The characterization of $\exp$ by $f' = f$,
$f(0) = 1$ (\cref{prop:b1:functions:expln}) is the one-dimensional
seed of the theory of linear differential equations, and the
catenary curve $y = \cosh x$ returns among the plane curves of
\cref{ch:b1:curves}.
\end{remark}

\begin{remark}[Interlude: the inverse-function program]\label{rem:b1:functions:interlude}
This chapter ran one program four times: restrict a function
until it becomes a strictly monotonic bijection, name the inverse,
transport every formula through it. What was \emph{used} at each
step --- that a continuous strictly monotonic function on an
interval is a bijection onto an interval, and that its inverse is
continuous, then differentiable away from critical points --- was
borrowed on High School credit. The debt is repaid inside this
volume: \cref{ch:b1:continuity} proves the bijection \omterm{def:b1:logic:statement}{statement}
(the monotone inverse theorem, resting on the intermediate value
theorem), and \cref{ch:b1:derivative} proves the derivative rule
$(f^{-1})' = 1/(f' \circ f^{-1})$ that silently produced every
formula of \cref{prop:b1:functions:arcderiv} and
\cref{prop:b1:functions:invhyp}. Reading those chapters with
$\arcsin$ and $\operatorname{arsinh}$ in mind --- as the worked
examples the theory was built to justify --- is the intended way
around the apparent circularity.
\end{remark}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:b1:functions:1}
Compute without calculator: $\arcsin\frac{\sqrt 3}{2}$;
$\;\arccos\bigl(-\frac 12\bigr)$; $\;\arctan(-1)$;
$\;\arcsin\bigl(\sin\frac{5\pi}{6}\bigr)$;
$\;\arccos\bigl(\cos\frac{7\pi}{4}\bigr)$.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b1:functions:2}
Give the domain of definition of $f(x) = \arcsin(2x - 1)$ and compute
$f'$ where defined. Same questions for $g(x) = \arctan\sqrt{x}$.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b1:functions:3}
Prove that for all $x \in \R$: $\cosh x + \sinh x = \eu^x$ and
$(\cosh x + \sinh x)^n = \cosh nx + \sinh nx$ for every $n \in \Z$
(a hyperbolic de Moivre formula).
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b1:functions:4}
Simplify $\cos(2\arcsin x)$ and $\sin(2\arctan x)$ into algebraic
expressions of $x$.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b1:functions:5}
Order, for large $x$, the functions
$x^{100}$, $\;\eu^{\sqrt x}$, $\;(\ln x)^{1000}$, $\;\eu^{x/100}$,
$\;x^{\ln x}$, from slowest to fastest growth, with justifications
based on \cref{prop:b1:functions:powerrules}.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:functions:6}
Study the function $f(x) = \arctan\dfrac{2x}{1 - x^2}$ on its domain:
compute $f'$, compare with $(2\arctan x)'$, and express $f(x)$ in
terms of $\arctan x$ on each of the three intervals of the domain.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:functions:7}
Solve in $\R$: $\;\cosh x = 2$; then $5\cosh x - 4 \sinh x = 3$
(express the solutions with logarithms). \emph{Hint for the second:
write everything with $u = \eu^x$.}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:functions:8}
Prove the identity $\arctan\frac{1}{2} + \arctan\frac{1}{3} =
\frac{\pi}{4}$, then Machin's formula\index{Machin's formula}
\[
4\arctan\frac 15 - \arctan\frac{1}{239} = \frac{\pi}{4}.
\]
\emph{Hint: compute the tangent of both sides using the addition
formula, and control in which interval each side lies.}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:functions:9}
Prove that for all $x \geq 0$: $\;x - \dfrac{x^3}{6} \leq \sin x \leq
x$ \emph{(study the successive derivatives of the differences)}, and
deduce $\lim_{x \to 0^+} \frac{x - \sin x}{x^3} = \frac 16$ is
plausible --- the limit itself is established in
\cref{ch:b1:taylor}.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b1:functions:10}
For $x \in \intcc{-1}{1}$, set
$h(x) = \arcsin\bigl(2x\sqrt{1 - x^2}\,\bigr) - 2\arcsin x$. One might
expect $h = 0$ from the identity $\sin 2y = 2\sin y\cos y$ --- but
$h$ is \emph{not} identically zero. Compute $h'$ on the open intervals
where it exists, evaluate $h$ at well-chosen points, and give the full
piecewise-constant description of $h$ on $\intcc{-1}{1}$.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b1:functions:11}
(Gudermannian) Let $g(x) = \arctan(\sinh x)$ for $x \in \R$. Prove
that $g$ is an odd, strictly increasing bijection from $\R$ onto
$\intoo{-\frac\pi2}{\frac\pi2}$, that $g'(x) = \frac{1}{\cosh x}$,
and that $\tan g(x) = \sinh x$, $\;\sin g(x) = \tanh x$,
$\;\cos g(x) = \frac{1}{\cosh x}$: the function $g$ links circular
and hyperbolic trigonometry without complex numbers.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:functions:12}
Prove that
\[
\arctan 1 + \arctan 2 + \arctan 3 = \pi .
\]
\emph{Hint: compute $\arctan 2 + \arctan 3$ first, localizing the
sum as in \cref{ex:b1:functions:arctansum}; beware that the
addition formula's denominator is negative here.}
\end{exercise}

\section{Problem: The equation $x^y = y^x$}

\begin{problem}\label{pb:b1:functions:1}
Which pairs of positive numbers satisfy $x^y = y^x$? Everyone knows
one accidental-looking example, $2^4 = 4^2 = 16$; this problem shows
that nothing about it is accidental. The whole equation is governed
by the variations of the single function
\[
f(t) = \frac{\ln t}{t} \qquad (t > 0),
\]
whose study yields: the complete solution \omterm{def:b1:logic:sets}{set} (a diagonal plus one
curved branch through $(\eu, \eu)$), a rational parametrization of
the branch, the fact that $(2, 4)$ is its \emph{only} integer point
and the classification of all its rational points --- plus, as
dividends, the comparison of $\eu^\pi$ with $\pi^\eu$ and the
monotone convergence of $\bigl(1 + \frac1n\bigr)^n$ to $\eu$.
\index{power function}

\medskip
\noindent\textbf{Part I --- The function $f(t) = \ln t / t$.}
\begin{enumerate}
  \item Justify that $f$ is differentiable on $\intoo0{+\infty}$,
        compute $f'$, and draw up the variation table: $f$
        increases strictly on $\intoc0\eu$, decreases strictly on
        $\intco\eu{+\infty}$, with maximum $f(\eu) = \frac1\eu$.
  \item Determine the limits of $f$ at $0^+$ and $+\infty$
        (\cref{prop:b1:functions:powerrules}), the sign of $f$
        (negative on $\intoo01$, zero at $1$, positive beyond),
        and sketch the graph.
  \item Check by direct computation that $f(2) = f(4)$. (Keep this
        equality in mind: the entire problem grows out of it.)
  \item Let $c \in \R$. Discuss, according to the value of $c$, the
        number of solutions of $f(t) = c$: exactly one for $c \leq
        0$; exactly two (one in $\intoo1\eu$, one in
        $\intoo\eu{+\infty}$) for $0 < c < \frac1\eu$; exactly one
        for $c = \frac1\eu$; none for $c > \frac1\eu$.
  \item Deduce that $\eu^x > x^\eu$ for every $x > 0$ with $x \neq
        \eu$, and in particular settle which of $\eu^\pi$ and
        $\pi^\eu$ is larger.
\end{enumerate}

\medskip
\noindent\textbf{Part II --- The equation and its curve.} In this
part $x, y > 0$.
\begin{enumerate}[resume]
  \item Show that $x^y = y^x \iff f(x) = f(y)$.
  \item Deduce the structure of the solution \omterm{def:b1:logic:sets}{set}: all diagonal
        pairs $(x, x)$; and the \emph{nontrivial} pairs ($x \neq
        y$), which satisfy: both coordinates are $> 1$, and one
        lies in $\intoo1\eu$ while the other lies in
        $\intoo\eu{+\infty}$.
  \item Parametrize the nontrivial pairs: writing $y = tx$ with $t
        > 0$, $t \neq 1$, show that $x^y = y^x$ forces
        \[
        x(t) = t^{\frac1{t-1}},
        \qquad
        y(t) = t\,x(t) = t^{\frac{t}{t-1}},
        \]
        and that conversely every such pair is a solution.
  \item Verify that $t = 2$ gives $(2, 4)$, and prove the symmetry
        $x(1/t) = y(t)$, $y(1/t) = x(t)$: inverting the parameter
        swaps the two coordinates.
  \item Determine the limits of $x(t)$ and $y(t)$ as $t \to 1$
        (both tend to $\eu$), as $t \to +\infty$ ($x \to 1$, $y
        \to +\infty$) and as $t \to 0^+$ ($x \to +\infty$, $y \to
        1$). Describe the resulting branch: a curve asymptotic to
        the lines $x = 1$ and $y = 1$, crossing the diagonal at
        $(\eu, \eu)$.
  \item Show that $t \mapsto x(t)$ is strictly decreasing on
        $\intoo1{+\infty}$. \emph{(Study $h(t) = 1 - \frac1t - \ln
        t$, whose sign controls the derivative of $\ln x(t) =
        \frac{\ln t}{t - 1}$.)}
  \item Conclude that the nontrivial solutions define a strictly
        decreasing bijection $\varphi \colon \intoo1\eu \to
        \intoo\eu{+\infty}$ with $x^{\varphi(x)} = \varphi(x)^x$,
        and that $\varphi$ is an involution of the branch:
        $\varphi(\varphi(x)) = x$ wherever both sides are defined.
\end{enumerate}

\medskip
\noindent\textbf{Part III --- Integer and rational points.}
\begin{enumerate}[resume]
  \item Prove that $(2, 4)$ and $(4, 2)$ are the only nontrivial
        \emph{integer} solutions of $x^y = y^x$.
  \item For $n \in \N^*$, apply the parametrization with $t = 1 +
        \frac1n$ and show that
        \[
        x_n = \Bigl(1 + \frac1n\Bigr)^{n},
        \qquad
        y_n = \Bigl(1 + \frac1n\Bigr)^{n+1}
        \]
        is a nontrivial \emph{rational} solution for every $n$,
        with $(x_1, y_1) = (2, 4)$.
  \item Conversely, let $(x, y)$ be a nontrivial rational solution
        with $y > x$, and write $t = y/x = 1 + \frac rs$ in lowest
        terms ($r, s \in \N^*$). Using $x = t^{s/r}$ and admitting
        the uniqueness of prime factorization (familiar from
        school; proved in \cref{ch:b1:arith}), show that $x$
        rational forces both $s$ and $s + r$ to be $r$-th powers
        of integers.
  \item Show with the binomial theorem that $a^r - b^r = r$ has no
        integer solutions $a > b \geq 1$ when $r \geq 2$, and
        conclude: the rational solutions of $x^y = y^x$ are
        exactly the pairs $(x_n, y_n)$ of question 14 (and their
        swaps).
  \item Verify the case $n = 2$ numerically: compute $f(9/4)$ and
        $f(27/8)$ to four decimal places and check they agree.
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- Dividends.}
\begin{enumerate}[resume]
  \item Use questions 7 and 11 to prove, with no further
        computation, that the sequence $x_n = \bigl(1 +
        \frac1n\bigr)^n$ is strictly increasing with $x_n < \eu$,
        that $y_n = \bigl(1 + \frac1n\bigr)^{n+1}$ is strictly
        decreasing with $y_n > \eu$, and that both converge to
        $\eu$. (The parameter $t_n = 1 + \frac1n$ decreases to
        $1$.)
  \item Establish the general comparison rule for $1 < a < b$:
        if $\eu \leq a$ then $a^b > b^a$; if $b \leq \eu$ then
        $a^b < b^a$; and show by the two examples $(2, 3)$ and
        $(2, 5)$ that in the mixed case $a < \eu < b$ both
        outcomes really occur.
  \item Suppose the involution $\varphi$ of question 12 is
        differentiable at $\eu$ (it is). Differentiate the
        identity $\varphi(\varphi(x)) = x$ at $x = \eu$ and deduce
        $\varphi'(\eu) = -1$: the branch crosses the diagonal at
        right angles to it.
  \item Find the unique solution pair with $y = 3x$, in closed
        form, and check it numerically to four decimals via $f$.
  \item Among the numbers $\sqrt2$, $\sqrt[3]3$, $\sqrt[4]4$,
        $\sqrt[5]5$, determine the largest and identify the two
        that are equal. \emph{(Compare $n^{1/n} =
        \eu^{f(n)}$.)}
\end{enumerate}

\medskip
\noindent\textbf{Part V --- Synthesis.}
\begin{enumerate}[resume]
  \item Describe the complete solution \omterm{def:b1:logic:sets}{set} of $x^y = y^x$ in the
        quarter-plane $x, y > 0$ --- diagonal plus branch, their
        intersection $(\eu, \eu)$, the asymptotes, the integer
        point $(2, 4)$, the rational points accumulating at
        $(\eu, \eu)$ --- in a form you could sketch from memory.
  \item Where exactly did the problem use: (i) the \omterm{prop:b1:functions:powerrules}{growth
        comparisons} of \cref{prop:b1:functions:powerrules}; (ii)
        the intermediate value theorem (through the bijection
        \omterm{def:b1:logic:statement}{statements}); (iii) the admitted uniqueness of prime
        factorization? One sentence each.
  \item Moral, in a short paragraph: a single variation table
        solved an equation in two unknowns, classified its
        rational points, and proved the monotone convergence of
        $\bigl(1 + \frac1n\bigr)^n$ --- comment on this economy,
        and name where each thread is industrialized later in the
        volume (\cref{ch:b1:seq} for the sequence,
        \cref{ch:b1:derivative} for variation tables,
        \cref{ch:b1:taylor} for the precision the table lacks).
\end{enumerate}
\end{problem}
