\chapter{Topology of the Real Line}\label{ch:b1:topology}

Limits keep referring to the same geometric vocabulary: points
``close to'' a set, sets ``without boundary leaks'', intervals in
which sequences cannot escape. This chapter fixes that vocabulary ---
open and closed sets, interior and closure, density --- on the real
line, and proves the compactness of segments in its sequential form.
The same notions, in normed vector spaces, are second-year material;
on $\R$ they are within reach and immediately useful for
\cref{ch:b1:continuity}.

\section{Open sets, closed sets}

\begin{definition}[Neighborhood, open set]\label{def:b1:topology:open}
A set $V \subseteq \R$ is a \emph{neighborhood}\index{neighborhood}
of $x \in \R$ when it contains an interval $\intoo{x - r}{x + r}$ for
some $r > 0$. A set $U \subseteq \R$ is \emph{open}\index{open set}
when it is a neighborhood of each of its points:
\[
\forall x \in U,\ \exists r > 0, \quad \intoo{x - r}{x + r}
\subseteq U .
\]
\end{definition}

\begin{example}\label{ex:b1:topology:open}
Open intervals are open: for $x \in \intoo{a}{b}$, take $r = \min(x -
a,\, b - x) > 0$. Half-lines $\intoo{a}{+\infty}$ are open; $\R$ and
$\emptyset$ are open (the latter vacuously). $\intcc{0}{1}$ is
\emph{not} open: no interval around $0$ stays inside.
\end{example}

\begin{proposition}[Stability of open sets]\label{prop:b1:topology:openstable}
Any union of open sets is open; a \emph{finite} intersection of open
sets is open. Infinite intersections may fail:
$\bigcap_{n \geq 1} \intoo{-\frac 1n}{\frac 1n} = \{0\}$, not open.
\end{proposition}

\begin{proof}
Union: if $x \in \bigcup_i U_i$, then $x \in U_{i_0}$ for some $i_0$,
and the interval provided by $U_{i_0}$ sits inside the union. Finite
intersection: if $x \in U_1 \cap \dots \cap U_k$, take $r = \min(r_1,
\dots, r_k) > 0$ of the radii provided by each $U_j$. For the
counterexample: any interval around $0$ contains some $\frac{1}{n}$
(Archimedes), hence leaves the intersection.
\end{proof}

\begin{example}[Certifying openness with explicit radii]\label{ex:b1:topology:radii}
Is $U = \{x \in \R : x^2 > 2\}$ open? Yes, and the certificate
can be written down: $U = \intoo{-\infty}{-\sqrt2} \cup
\intoo{\sqrt2}{+\infty}$, a union of two open half-lines, open
by \cref{prop:b1:topology:openstable}. Alternatively, argue
point by point: for $x \in U$ with $x > \sqrt 2$, take $r = x -
\sqrt2 > 0$: every $y \in \intoo{x - r}{x + r}$ satisfies $y >
\sqrt 2$, hence $y^2 > 2$; symmetrically on the left. Both
styles matter --- the structural one (build from known open
sets by unions and finite intersections) scales better, the
$\varepsilon$-style one works when no structure is visible; and
\cref{ch:b1:continuity} will add a third, the most powerful:
$U$ is the preimage of the open $\intoo{2}{+\infty}$ under the
continuous $x \mapsto x^2$.
\end{example}

\begin{definition}[Closed set]\label{def:b1:topology:closed}
A set $F \subseteq \R$ is \emph{closed}\index{closed set} when its
complement $\R \setminus F$ is open. By de Morgan and
\cref{prop:b1:topology:openstable}: any intersection of closed sets
is closed, finite unions of closed sets are closed.
\end{definition}

\begin{theorem}[Sequential characterization of closed sets]\label{thm:b1:topology:seqclosed}
$F$ is closed if and only if: for every sequence $(u_n)$ of points of
$F$ converging to some $\ell \in \R$, the limit $\ell$ belongs to
$F$. (``Closed'' $=$ ``stable under limits''.)
\end{theorem}

\begin{proof}
($\Rightarrow$) Let $F$ be closed, $u_n \in F$, $u_n \to \ell$, and
suppose $\ell \notin F$. The complement is open: some
$\intoo{\ell - r}{\ell + r}$ avoids $F$. But convergence puts $u_n$
in that interval for large $n$: contradiction with $u_n \in F$.

($\Leftarrow$) Suppose $F$ is not closed: the complement is not open,
so some $x \notin F$ has no interval $\intoo{x - r}{x + r}$ inside
the complement; taking $r = \frac{1}{n+1}$, choose $u_n \in F$ with
$\abs{u_n - x} < \frac{1}{n+1}$. Then $u_n \in F$, $u_n \to x \notin
F$: the sequential property fails.
\end{proof}

\begin{example}[The sequential test, both ways]\label{ex:b1:topology:seqtest}
\emph{Closed:} $F = \Z \cup \bigl\{n + \frac1n : n \geq 2\bigr\}$.
Let $u_k \in F$ with $u_k \to \ell$. The window $\intcc{\ell -
1}{\ell + 1}$ contains only finitely many points of $F$ (finitely
many integers, finitely many $n + \frac1n$), and beyond some rank
all $u_k$ lie in it: the sequence then takes finitely many
values, and, converging, is eventually constant (as in
\cref{exo:b1:topology:3}): $\ell \in F$. Closed --- even though
$F$ contains pairs of points at distance $\frac1n$, arbitrarily
close.

\emph{Not closed:} $G = \bigl\{\frac1m + \frac1n : m, n \in
\N^*\bigr\}$. The sequence $\frac1n + \frac1n \in G$ tends to
$0$, and $0 \notin G$ (a sum of two positive terms): the
sequential test fails, $G$ is not closed. Interestingly, each
$\frac1m$ \emph{does} belong to $\overline G \cap G$: indeed
$\frac1m = \frac{1}{m+1} + \frac{1}{m(m+1)} \in G$. The closing
insight: to prove closedness, control \emph{all} convergent
sequences at once (usually via a local finiteness or a
closed-formula argument); to disprove it, one well-chosen escaping
sequence suffices --- the asymmetry makes the negative direction
the easy one, and the counterexamples of this chapter all have
this one-line shape.
\end{example}

\begin{example}\label{ex:b1:topology:closed}
Segments $\intcc{a}{b}$, half-lines $\intco{a}{+\infty}$, finite
sets, $\Z$ (a convergent sequence of integers is eventually constant)
are closed. $\intoc{0}{1}$ is neither open (fails at $1$) nor closed
($\frac 1n \to 0 \notin$ the set): most sets are neither. $\R$ and
$\emptyset$ are both open and closed --- and they are the only such
subsets of $\R$ (\cref{exo:b1:topology:9}).
\end{example}

\begin{example}[An open set assembled from infinitely many pieces]\label{ex:b1:topology:rminusz}
$\R \setminus \Z = \bigcup_{n \in \Z} \intoo{n}{n+1}$: an
infinite union of open intervals, open by
\cref{prop:b1:topology:openstable} --- so $\Z$ is closed with no
sequential argument needed. Note the division of labor in the
stability rules: \emph{unions} of open sets may be arbitrary
(each point only needs its own certificate, supplied by the one
set containing it), while \emph{intersections} must stay finite
(certificates must be intersected, and infinitely many radii can
shrink to nothing). \Cref{exo:b1:topology:10} will show this
example is the general shape: every open subset of $\R$ is a
countable disjoint union of open intervals.
\end{example}

\section{Interior, closure, density}

\begin{definition}[Interior, closure, boundary]\label{def:b1:topology:closure}
Let $A \subseteq \R$.
\begin{itemize}
  \item A point $x$ is \emph{interior} to $A$ when $A$ is a
        neighborhood of $x$; the \emph{interior}
        $\mathring{A}$\index{interior} is the set of interior points.
  \item A point $x$ is \emph{adherent} to $A$ when every neighborhood
        of $x$ meets $A$; the \emph{closure}
        $\overline{A}$\index{closure} is the set of adherent points.
  \item The \emph{boundary} is $\partial A = \overline A \setminus
        \mathring A$.\index{boundary}
\end{itemize}
Then $\mathring A \subseteq A \subseteq \overline A$.
\end{definition}

\begin{proposition}[Main properties]\label{prop:b1:topology:closureprops}
\begin{enumerate}
  \item $\mathring A$ is the largest open set contained in $A$; $A$
        is open iff $A = \mathring A$.
  \item $\overline A$ is the smallest closed set containing $A$; $A$
        is closed iff $A = \overline A$.
  \item (Sequential characterization of adherence) $x \in \overline
        A$ if and only if $x$ is the limit of a sequence of points of
        $A$.
  \item Complementation exchanges the notions: $\R \setminus
        \overline A = \bigl(\R \setminus A\bigr)^{\!\circ}$.
\end{enumerate}
\end{proposition}

\begin{proof}
(4) $x \notin \overline A$ $\iff$ some neighborhood of $x$ avoids $A$
$\iff$ some interval around $x$ lies in $\R \setminus A$ $\iff$ $x$
is interior to $\R \setminus A$.

(1) $\mathring A$ is open: if $x \in \mathring A$, some
$\intoo{x-r}{x+r} \subseteq A$; every point $y$ of that interval has
a smaller interval around it inside it, hence inside $A$: the whole
interval is in $\mathring A$. Any open $U \subseteq A$ consists of
interior points of $A$, so $U \subseteq \mathring A$: largest. The
characterization of openness follows.

(2) In detail. By (4), $\R \setminus \overline A$ is the interior
of $\R \setminus A$, an open set by (1): so $\overline A$ is
closed, and it contains $A$. Minimality: let $F \supseteq A$ be
closed. Then $\R \setminus F$ is open and contained in $\R
\setminus A$, so by the maximality in (1),
\[
\R \setminus F \subseteq \bigl(\R \setminus A\bigr)^{\!\circ}
= \R \setminus \overline A ,
\]
and taking complements again: $\overline A \subseteq F$. Thus
$\overline A$ is the smallest closed superset. Characterization:
if $A = \overline A$ then $A$ is closed (just shown); if $A$ is
closed, it is itself a closed superset of $A$, so minimality
forces $\overline A \subseteq A$, and equality.

(3) If $u_n \in A$, $u_n \to x$: every neighborhood of $x$ contains
some $u_n \in A$, so $x \in \overline A$. Conversely, if $x \in
\overline A$: each interval $\intoo{x - \frac{1}{n+1}}{x +
\frac{1}{n+1}}$ meets $A$ at some $u_n$, and $u_n \to x$.
\end{proof}

\begin{example}\label{ex:b1:topology:closureexamples}
$\overline{\intoo{0}{1}} = \intcc{0}{1}$;
$\mathring{\intcc{0}{1}} = \intoo{0}{1}$; $\partial\intoo{0}{1} =
\{0, 1\}$. For $A = \{\frac 1n : n \in \N^*\}$: $\overline A = A
\cup \{0\}$, $\mathring A = \emptyset$, $\partial A = A \cup \{0\}$.
For $\Q$: by density (\cref{thm:b1:reals:density}) every real is
adherent to $\Q$, so $\overline{\Q} = \R$ while $\mathring{\Q} =
\emptyset$ (every interval contains irrationals): the boundary of
$\Q$ is all of $\R$.
\end{example}

\begin{example}[A full anatomy]\label{ex:b1:topology:anatomy}
Let $A = \intoc{0}{1} \,\cup\, \bigl(\Q \cap \intoo{2}{3}\bigr)
\,\cup\, \{4\}$. We compute the three sets of
\cref{def:b1:topology:closure}, piece by piece.

\emph{Interior.} A point of $\intoo{0}{1}$ has a whole interval
inside $A$: interior. The point $1$: every interval around it
leaks right of $1$, where $A$ has nothing until $2$: not
interior. No point of $\Q \cap \intoo{2}{3}$ is interior (every
interval contains irrationals, \cref{thm:b1:reals:density});
neither is the isolated $4$. So $\mathring A = \intoo{0}{1}$.

\emph{Closure.} Limits of points of $A$: all of $\intcc{0}{1}$
($0 = \lim \frac1n$ with $\frac 1n \in A$); all of
$\intcc{2}{3}$ (every real there is a limit of rationals of the
interval, density again); and $4$. Nothing else: a point
outside $\intcc{0}{1} \cup \intcc{2}{3} \cup \{4\}$ has positive
distance to that closed set. So $\overline A = \intcc{0}{1} \cup
\intcc{2}{3} \cup \{4\}$.

\emph{Boundary.} $\partial A = \overline A \setminus \mathring A
= \{0, 1\} \cup \intcc{2}{3} \cup \{4\}$.

The closing insight: the three operations act \emph{locally} ---
each piece of $A$ contributes according to its own nature (a
solid interval keeps its inside, a dense-but-porous piece turns
entirely into boundary, an isolated point is pure boundary), and
a two-line drawing of $A$ predicts every answer before any proof
is written.
\end{example}

\begin{remark}[Common pitfalls in point-set reasoning]
(i) \emph{``Not open'' does not mean ``closed''}: most sets are
neither ($\intoc{0}{1}$), and two sets are both
($\emptyset$, $\R$) --- open and closed are not opposites but
duals through complementation. (ii) \emph{Interior and closure
do not commute}: for $A = \Q$,
\[
\overline{\mathring A} = \overline\emptyset = \emptyset
\qquad\text{while}\qquad
\bigl(\,\overline A\,\bigr)^{\!\circ} = \mathring \R = \R :
\]
the two iterated operators differ as much as sets can. (iii)
\emph{Infinite unions of closed sets can fail to be closed}:
$\bigcup_{n\geq1} \intcc{\frac1n}{1} = \intoc{0}{1}$ --- the
mirror of the intersection counterexample of
\cref{prop:b1:topology:openstable}. (iv) \emph{Dense does not
mean big}: $\Q$ is dense, countable, with empty interior, and
its complement is dense too; density says ``arbitrarily close to
everything'', not ``almost everything'' --- the weekend problem's
Cantor set (\cref{pb:b1:topology:1}) makes the opposite point,
a topologically small set that is uncountably big.
\end{remark}

\begin{example}[A closure computed exactly]\label{ex:b1:topology:closurecomputed}
Let $G = \bigl\{\frac1m + \frac1n : m, n \in \N^*\bigr\}$ (from
\cref{ex:b1:topology:seqtest}). Claim:
\[
\overline G = G \,\cup\, \Bigl\{\frac1m : m \in \N^*\Bigr\}
\,\cup\, \{0\} .
\]
($\supseteq$) $\frac1m = \lim_n \bigl(\frac1m + \frac1n\bigr)$
and $0 = \lim_n \frac2n$: adherent by the sequential
characterization. ($\subseteq$) Let $x = \lim_k \bigl(
\frac{1}{m_k} + \frac{1}{n_k}\bigr)$; order each pair so that
$m_k \leq n_k$. If $(m_k)$ is unbounded, a subsequence has $m_k
\to \infty$, hence $n_k \to \infty$ too and $x = 0$. Otherwise
$(m_k)$ takes finitely many values, one of them, say $m$,
infinitely often; along that subsequence $\frac{1}{n_k} \to x -
\frac1m$: if $(n_k)$ is bounded it takes some value $n$
infinitely often and $x = \frac1m + \frac1n \in G$; if not, $x =
\frac1m$. Every case lands in the announced set. The closing
insight: computing a closure is a compactness-style case
analysis on indices --- bounded index means finitely many
values (pigeonhole), unbounded index means a limit escapes ---
and the answer displays the typical two-layer structure of
limit points: the set, its first-generation limits, and their
limit $0$.
\end{example}

\begin{definition}[Density, topological form]\label{def:b1:topology:dense}
$A$ is \emph{dense}\index{density} in $\R$ when $\overline A = \R$
--- equivalently, every nonempty open interval meets $A$;
equivalently (by \cref{prop:b1:topology:closureprops}~(3)), every
real is a limit of elements of $A$. Examples: $\Q$, $\R \setminus
\Q$, the dyadics (\cref{exo:b1:reals:8}), dense subgroups
(\cref{exo:b1:reals:9}).
\end{definition}

\begin{example}[Density is relative]\label{ex:b1:topology:relativedensity}
``Dense'' as defined here means dense \emph{in $\R$}; a set can
instead be dense in a part of the line only. The dyadics of
$\intcc{0}{1}$, i.e.\ $D \cap \intcc{0}{1}$
(\cref{exo:b1:reals:8}), meet every open interval included in
$\intcc{0}{1}$ but of course miss $\intoo{2}{3}$ entirely: they
are dense \emph{in} $\intcc{0}{1}$, meaning $\overline{D \cap
\intcc{0}{1}} = \intcc{0}{1}$. The general phrase ``$A$ is dense
in $B$'' abbreviates $B \subseteq \overline A$ --- always name
the ambient set, since the weekend problem's endpoints are dense
\emph{in the Cantor set} while being nowhere dense \emph{in
$\R$}: the same set, two truthful and opposite-sounding
descriptions.
\end{example}

\begin{example}[Handling density]\label{ex:b1:topology:densitytricks}
Three quick moves that recur constantly. \emph{Enlarging}: if
$A$ is dense and $A \subseteq B$, then $B$ is dense (every
interval already meets $A$). \emph{Transporting}: if $A$ is
dense, so is $\lambda A + \mu$ for $\lambda \neq 0$ --- an
interval $I$ meets $\lambda A + \mu$ iff the interval
$\frac{I - \mu}{\lambda}$ meets $A$; thus the odd multiples of
$10^{-9}$, say, are dense. \emph{Intersecting fails}: two dense
sets can miss each other entirely ($\Q$ and $\R \setminus \Q$):
density survives unions and affine maps, never intersections.
\end{example}

\section{Compactness of segments}

\begin{theorem}[Segments are sequentially compact]\label{thm:b1:topology:compact}
Let $a \leq b$. Every sequence of points of $\intcc{a}{b}$ has a
subsequence converging to a point \emph{of}
$\intcc{a}{b}$.\index{compactness}

More generally, the subsets of $\R$ with this property (every
sequence has a subsequence converging in the set) are exactly the
\emph{closed and bounded} sets.
\end{theorem}

\begin{proof}
A sequence in $\intcc{a}{b}$ is bounded, so Bolzano--Weierstrass
(\cref{thm:b1:seq:bw}) extracts a convergent subsequence; its limit
stays in $\intcc{a}{b}$ because segments are closed
(\cref{thm:b1:topology:seqclosed}).

General case. \emph{(Closed bounded $\Rightarrow$ compact)}: let
$F$ be closed and bounded, and $(u_n)$ a sequence in $F$.
Boundedness of $F$ bounds the sequence, so Bolzano--Weierstrass
extracts $u_{\varphi(n)} \to \ell$; and $\ell \in F$ because $F$
is closed and the subsequence is a convergent sequence of points
of $F$ (\cref{thm:b1:topology:seqclosed}): the two hypotheses
are consumed one each, boundedness for existence of the limit,
closedness for its membership. \emph{(Compact $\Rightarrow$ closed and
bounded)}: if $F$ is unbounded, pick $u_n \in F$ with $\abs{u_n} \geq
n$; every subsequence is unbounded, hence divergent
(\cref{prop:b1:seq:first}): no convergent subsequence at all. If $F$
is not closed, take $u_n \in F$ with $u_n \to \ell \notin F$
(\cref{thm:b1:topology:seqclosed}): every subsequence converges to
$\ell \notin F$, so no subsequence converges \emph{in} $F$.
\end{proof}

\begin{example}[Nested compact sets]\label{ex:b1:topology:nested}
A first workout for the theorem. Let $K_0 \supseteq K_1 \supseteq
K_2 \supseteq \dots$ be nonempty compact (closed bounded) subsets
of $\R$. Then $\bigcap_n K_n \neq \emptyset$. Indeed, pick $x_n
\in K_n$ for each $n$: the sequence lives in the compact $K_0$,
so a subsequence $x_{\varphi(n)}$ converges to some $x$
(\cref{thm:b1:topology:compact}). For every fixed $m$, the terms
$x_{\varphi(n)}$ with $\varphi(n) \geq m$ all lie in the closed
set $K_m$, so the limit $x$ lies in $K_m$
(\cref{thm:b1:topology:seqclosed}); as $m$ was arbitrary, $x \in
\bigcap_n K_n$. A useful companion: if an open set $U$ contains
$\bigcap_n K_n$, then $U \supseteq K_n$ for some $n$ --- apply
the same argument to points $x_n \in K_n \setminus U$; the limit
$x$ would lie in $\bigcap K_n \subseteq U$, yet $U$ open forces
$x_{\varphi(n)} \in U$ eventually, a contradiction. Both
statements fail without compactness: $\bigcap_n \intoo{0}{\frac
1n} = \emptyset$ and $\bigcap_n \intco{n}{+\infty} = \emptyset$.
The closing insight: compactness converts an infinite chain of
non-emptiness assertions into a single limit point --- it is the
tool that survives passage to the infinite intersection, and the
weekend problem (\cref{pb:b1:topology:1}) will lean on it twice.
\end{example}

\begin{omfigure}
\begin{tikzpicture}[scale=9.5]
  \node[left, font=\small] at (-0.02,0) {$C_0$};
  \fill[omDef] (0,-0.011) rectangle (1,0.011);
  \node[left, font=\small] at (-0.02,-0.09) {$C_1$};
  \fill[omDef] (0,-0.101) rectangle (0.3333,-0.079);
  \fill[omDef] (0.6667,-0.101) rectangle (1,-0.079);
  \node[left, font=\small] at (-0.02,-0.18) {$C_2$};
  \fill[omDef] (0,-0.191) rectangle (0.1111,-0.169);
  \fill[omDef] (0.2222,-0.191) rectangle (0.3333,-0.169);
  \fill[omDef] (0.6667,-0.191) rectangle (0.7778,-0.169);
  \fill[omDef] (0.8889,-0.191) rectangle (1,-0.169);
  \node[left, font=\small] at (-0.02,-0.27) {$C_3$};
  \fill[omDef] (0,-0.281) rectangle (0.0370,-0.259);
  \fill[omDef] (0.0741,-0.281) rectangle (0.1111,-0.259);
  \fill[omDef] (0.2222,-0.281) rectangle (0.2593,-0.259);
  \fill[omDef] (0.2963,-0.281) rectangle (0.3333,-0.259);
  \fill[omDef] (0.6667,-0.281) rectangle (0.7037,-0.259);
  \fill[omDef] (0.7407,-0.281) rectangle (0.7778,-0.259);
  \fill[omDef] (0.8889,-0.281) rectangle (0.9259,-0.259);
  \fill[omDef] (0.9630,-0.281) rectangle (1,-0.259);
  \node[below, font=\small] at (0,-0.30) {$0$};
  \node[below, font=\small] at (0.3333,-0.30) {$\tfrac13$};
  \node[below, font=\small] at (0.6667,-0.30) {$\tfrac23$};
  \node[below, font=\small] at (1,-0.30) {$1$};
  \draw[omThm, thick, -Stealth] (0.5,-0.042) -- (0.5,-0.062);
  \draw[omThm, thick, -Stealth] (0.5,-0.132) -- (0.5,-0.152);
  \draw[omThm, thick, -Stealth] (0.5,-0.222) -- (0.5,-0.242);
\end{tikzpicture}

{\small The first four stages of the middle-thirds construction:
each segment of $C_n$ loses its open middle third, leaving the
$2^{n+1}$ segments of $C_{n+1}$, of total length
$(\frac23)^{n+1}$. The Cantor set $C = \bigcap_n C_n$ --- the
subject of the weekend problem \cref{pb:b1:topology:1} --- is
the nonempty compact residue guaranteed by the nested-compacts
argument above: length zero, yet uncountably many points
survive.}
\end{omfigure}

\begin{remark}
This is the engine behind the extreme value theorem
(\cref{ch:b1:continuity}) and Heine's theorem on uniform continuity.
The name ``compact'' will acquire its general (covering) definition
in the second year; on $\R$, sequential compactness is all we need,
and ``compact $=$ closed $+$ bounded'' is the statement to remember.
\end{remark}

\begin{example}[Boundaries under unions]\label{ex:b1:topology:boundaryunion}
Always $\partial(A \cup B) \subseteq \partial A \cup \partial
B$: a point of $\partial(A \cup B)$ has every neighborhood
meeting $A \cup B$ (hence $A$ or $B$, infinitely often one of
them) and meeting the complement of $A \cup B$, which lies in
both complements --- a short check then places the point in
$\partial A$ or $\partial B$. The inclusion can be spectacularly
strict: with $A = \Q$ and $B = \R\setminus\Q$,
\[
\partial(A \cup B) = \partial \R = \emptyset ,
\qquad
\partial A \cup \partial B = \R \cup \R = \R :
\]
two ragged sets can glue into a seamless one, their boundaries
annihilating each other. The closing insight: interiors and
closures behave \emph{monotonically} under unions and
intersections, but boundaries do not --- treat $\partial$ as a
derived quantity ($\overline A \setminus \mathring A$), never as
an operator with algebra of its own.
\end{example}

\begin{remark}[Perspectives inside this volume]
The vocabulary built here is consumed twice more in this book.
In \cref{ch:b1:continuity}, every theorem is a topology
statement in disguise: the intermediate value theorem says
continuous maps preserve the interval property, the extreme
value theorem says they preserve compactness --- and the proofs
call \cref{thm:b1:topology:seqclosed,thm:b1:topology:compact}
by name. In \cref{ch:b1:multivar}, the same definitions are
re-read in $\R^2$ with disks in place of intervals: open sets,
closures and compactness transfer word for word, and the
two-variable extreme value theorem again rides on
Bolzano--Weierstrass (extract on each coordinate). The one
notion that does \emph{not} generalize painlessly is the
interval itself --- in the plane, connectedness replaces
convexity, a story that begins with
\cref{exo:b1:topology:9}'s ``only $\emptyset$ and $\R$ are open
and closed''.
\end{remark}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:b1:topology:1}
For each set, say whether it is open, closed, both, or neither (with
justification): $\intoo{0}{1} \cup \intoo{2}{3}$; $\;\intco{0}{1}$;
$\;\{0\} \cup \intcc{1}{2}$; $\;\R \setminus \Z$; $\;\Q \cap
\intoo{0}{1}$.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b1:topology:2}
Determine $\mathring A$, $\overline A$ and $\partial A$ for:
$A = \intoc{0}{1} \cup \{2\}$; $\;A = \R \setminus \Q$;
$\;A = \bigl\{\frac{(-1)^n n}{n+1} : n \in \N\bigr\}$.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b1:topology:3}
Prove that a finite set is closed, first via complements, then via
the sequential characterization.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b1:topology:4}
Prove that for any $A, B \subseteq \R$: $\overline{A \cup B} =
\overline A \cup \overline B$. Show by example that
$\overline{A \cap B}$ can differ from $\overline A \cap \overline
B$.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:topology:5}
Let $u_n \to \ell$ in $\R$. Prove that the set $\{u_n : n \in \N\}
\cup \{\ell\}$ is closed (hence compact if we add that it is
bounded --- which it is).
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:topology:6}
Let $U$ be open and $A$ arbitrary. Prove that $U + A = \{u + a\}$ is
open. Deduce that the sum of an open set and any set is open, and
contrast: exhibit two \emph{closed} sets whose sum is not closed.
\emph{(Try $\Z$ and $\sqrt 2\,\Z$, with \cref{exo:b1:reals:9}.)}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:topology:7}
A point $x \in A$ is \emph{isolated} in $A$ when some neighborhood of
$x$ meets $A$ only at $x$. Prove that every point of $\Z$ is isolated
in $\Z$, that $A = \{\frac1n\}$ has all its points isolated yet
$\overline A \neq A$, and that a set whose points are all isolated
has empty interior.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:topology:8}
Prove that the closure of a bounded set is bounded, and that
$\sup \overline A = \sup A$ for $A$ nonempty bounded above. Deduce
that $\sup A \in \overline A$: the supremum is always adherent.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b1:topology:9}
Prove that the only subsets of $\R$ that are both open and closed are
$\emptyset$ and $\R$. \emph{Hint: suppose $A$ is open, closed, with
$A \neq \emptyset$ and $\R \setminus A \neq \emptyset$; pick $a \in
A$, $b \notin A$, say $a < b$, and consider $s = \sup\,(A \cap
\intcc{a}{b})$; decide whether $s$ can belong to $A$ or to its
complement.}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b1:topology:10}
(Structure of open sets) Let $U \subseteq \R$ be open, nonempty. For
$x \in U$, let $I_x$ be the union of all open intervals containing
$x$ and contained in $U$. Prove that $I_x$ is an open interval, that
two sets $I_x$, $I_y$ are equal or disjoint, and that $U$ is a union
of \emph{countably many} pairwise disjoint open intervals \emph{(pick
a rational in each)}.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:topology:11}
A point $x \in \R$ is an \emph{accumulation point} of $A$ when
every neighborhood of $x$ meets $A \setminus \{x\}$; their set is
the \emph{derived set} $A'$. Prove that $\overline A = A \cup A'$,
and that $A$ is closed if and only if $A' \subseteq A$. Determine
$A'$ for $A = \{\frac 1n : n \in \N^*\}$, for $A = \Z$, and for
$A = \Q$.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b1:topology:12}
For $A \subseteq \R$ nonempty, define $d_A(x) = \inf\{\abs{x - a}
: a \in A\}$. Prove:
\begin{enumerate}
  \item $\abs{d_A(x) - d_A(y)} \leq \abs{x - y}$ for all $x, y$
        ($d_A$ is $1$-Lipschitz);
  \item $d_A(x) = 0$ if and only if $x \in \overline A$; in
        particular, if $F$ is closed and $x \notin F$, then
        $d_F(x) > 0$;
  \item for every $\varepsilon > 0$, the set $V_\varepsilon =
        \{x : d_F(x) < \varepsilon\}$ is open, contains $F$, and
        $\bigcap_{\varepsilon > 0} V_\varepsilon = F$ for $F$
        closed: every closed set is a countable intersection of
        open sets.
\end{enumerate}
\end{exercise}

\section{Problem: The Cantor set, small and enormous at once}

\begin{problem}[{Weekend problem --- the Cantor middle-thirds
set: length zero, uncountable, perfect, and $C + C =
\intcc{0}{2}$}]
\label{pb:b1:topology:1}
Remove from $\intcc{0}{1}$ its open middle third, then the middle
third of each remaining segment, and repeat forever: what
survives is the \emph{Cantor set} $C$, the fundamental
counterexample factory of analysis. This problem constructs it,
reads it through the base-$3$ machinery of \cref{pb:b1:reals:1},
and establishes its paradoxical portrait: total length zero, yet
uncountable; empty interior, yet no isolated point; totally
disconnected, yet $C + C$ fills the whole segment
$\intcc{0}{2}$.\index{Cantor set} Formally: $C_0 = \intcc{0}{1}$,
and $C_{n+1}$ is obtained from $C_n$ by deleting the open middle
third of each segment of $C_n$; finally $C = \bigcap_{n \geq 0}
C_n$. Throughout, a \emph{ternary code} of $x \in \intcc{0}{1}$
is any digit string $(d_k)_{k\geq1}$ with $d_k \in \{0, 1, 2\}$
whose value $\sup_n \sum_{k=1}^n d_k 3^{-k}$ equals $x$ ---
improper codes (eventually $2$) are allowed; by
\cref{pb:b1:reals:1} (questions 9--11), every $x \in
\intcc{0}{1}$ has one or two codes, two exactly when $x = m/3^N
\in \intoo{0}{1}$.

\medskip
\noindent\textbf{Part I --- The construction.}
\begin{enumerate}
  \item Describe $C_1$ and $C_2$ explicitly as unions of
        segments, and prove by induction: $C_n$ is a disjoint
        union of $2^n$ closed segments, each of length $3^{-n}$.
  \item Show that $C$ is closed, bounded --- hence compact
        (\cref{thm:b1:topology:compact}) --- nonempty, and that
        every endpoint of every segment of every $C_n$ belongs
        to $C$.
  \item The total length of $C_n$ is $\bigl(\frac23\bigr)^n$.
        Deduce that for every $\varepsilon > 0$, the set $C$ can
        be covered by finitely many segments of total length
        $\leq \varepsilon$: the Cantor set has \emph{length
        zero}.
  \item Show that an interval contained in $C$ has length $\leq
        3^{-n}$ for every $n$, hence is a singleton or empty:
        $\mathring C = \emptyset$. Being closed with empty
        interior, $C$ is \emph{nowhere dense}.
\end{enumerate}

\medskip
\noindent\textbf{Part II --- The ternary code.}
\begin{enumerate}[resume]
  \item Prove the self-similarity recursion
        \[
        C_{n+1} = \tfrac13 C_n \,\cup\,
        \bigl(\tfrac23 + \tfrac13 C_n\bigr),
        \qquad\text{hence}\qquad
        C = \tfrac13 C \,\cup\, \bigl(\tfrac23 + \tfrac13
        C\bigr),
        \]
        the two pieces being disjoint: $C$ is two copies of
        itself at scale $\frac13$.
  \item Prove by induction on $n$: $x \in C_n$ if and only if
        $x$ has a ternary code whose first $n$ digits lie in
        $\{0, 2\}$. Deduce, using the fact that $x$ has at most
        two codes: $x \in C$ if and only if $x$ has a code with
        \emph{no digit equal to $1$} (a \emph{$1$-free} code).
  \item Codes in action: give $1$-free codes for $0$, $1$,
        $\frac13$, $\frac23$; show $\frac14 = (0.\overline{02})_3$
        and $\frac34 = (0.\overline{20})_3$, so both belong to
        $C$; and check that $\frac14$ is \emph{not} an endpoint
        of any $C_n$ (endpoints have the form $m/3^n$).
  \item Show that each $x \in C$ has \emph{exactly one} $1$-free
        code \emph{(when $x$ has two codes, prove that exactly
        one of the pair contains the digit $1$)}. Conclude: the
        value map is a bijection from $\{0,2\}$-strings onto
        $C$.
  \item (Diagonal) Let $k \mapsto x_k$ be any map $\N^* \to C$.
        Build a $\{0, 2\}$-string differing at index $k$ from
        the code of $x_k$, and conclude that $C$ is uncountable
        --- while, by contrast, question 3 says it is
        metrically negligible.
\end{enumerate}

\medskip
\noindent\textbf{Part III --- Topological portrait.}
\begin{enumerate}[resume]
  \item Assemble the record so far: $C$ is compact, uncountable,
        of length zero, nowhere dense. Which single containment
        $C \subseteq C_n$ carries each property?
  \item ($C$ is perfect) Let $x \in C$ with $1$-free code
        $(d_k)$. Flipping the digit $d_n$ ($0 \leftrightarrow
        2$) produces $x_n \in C$ with $\abs{x_n - x} = 2 \cdot
        3^{-n}$. Conclude that $C$ has no isolated point: every
        point of $C$ is a limit of \emph{other} points of $C$.
  \item Show that the endpoints of question 2 form a countable
        dense subset of $C$ \emph{(truncate the code after $n$
        digits and continue with $0$s; countability as in
        \cref{exo:b1:topology:10})}. Conclude: the typical point
        of $C$ --- like $\frac14$ --- is \emph{not} an endpoint:
        endpoints are a countable skeleton inside an uncountable
        body.
  \item (Totally disconnected) Let $x < y$ in $C$. Choose $n$
        with $3^{-n} < y - x$ and produce a point $z \in
        \intoo{x}{y}$ with $z \notin C$. Conclude that the only
        nonempty intervals contained in $C$ are singletons.
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- Arithmetic of $C$.}
\begin{enumerate}[resume]
  \item Show $1 - C = C$ \emph{(what does $x \mapsto 1 - x$ do
        to a $1$-free code? recall $1 = (0.\overline{2})_3$)}.
  \item (Addition of codes) Show that if $x$, $x'$ have codes
        $(a_k)$, $(b_k)$, then $x + x' = \lim_n\,(t_n + t'_n)$
        where $t_n, t'_n$ are the partial sums. Deduce: every $y
        \in \intcc{0}{1}$ is the \emph{midpoint} of two points
        of $C$ --- given a code $(e_k)$ of $y$, choose digits
        $a_k, b_k \in \{0, 2\}$ with $\frac{a_k + b_k}{2} =
        e_k$.
  \item Conclude $C + C = \intcc{0}{2}$ and, with question 14,
        $C - C = \intcc{-1}{1}$. Concrete instance: write $1$ as
        a sum of the two non-endpoints found in question 7.
  \item Reflect: a set of length zero whose difference set fills
        $\intcc{-1}{1}$. Why is there no contradiction between
        ``$C$ is metrically negligible'' and ``$C + C$ has full
        length''? (One sentence; think about what length
        controls and what it does not.)
\end{enumerate}

\medskip
\noindent\textbf{Part V --- Members, rational and irrational.}
\begin{enumerate}[resume]
  \item Combine question 8 with the periodicity criterion of
        \cref{pb:b1:reals:1}: a point of $C$ is rational if and
        only if its $1$-free code is eventually periodic. Run
        the base-$3$ long division to check $\frac1{13} =
        (0.\overline{002})_3 \in C$.
  \item Produce an explicitly \emph{irrational} member of $C$:
        the value of the code with $d_k = 2$ at the triangular
        positions $k = \frac{j(j+1)}{2}$ and $d_k = 0$
        elsewhere. Justify irrationality by the growing-gaps
        argument of \cref{pb:b1:reals:1} (question 20).
  \item (Onto a full segment) Consider $h$ mapping the point of
        $C$ with $1$-free code $(d_k)$ to the value of the
        \emph{binary} string $\bigl(\frac{d_k}2\bigr)$, i.e.\
        $h(x) = \sup_n \sum_{k=1}^n \frac{d_k}{2}\,2^{-k}$.
        Show that $h$ maps $C$ \emph{onto} $\intcc{0}{1}$. So
        the negligible $C$ surjects onto a segment of full
        length --- a second proof that $C$ is uncountable.
  \item (Self-similar length) Suppose some notion of length $L$
        were defined for $C$ and its shrunken copies, respecting
        scaling ($L(\lambda A) = \lambda L(A)$), translation
        invariance, and additivity over the disjoint decomposition
        of question 5. Show that then $L(C) = \frac23\,L(C)$,
        forcing $L(C) = 0$: self-similarity alone already
        sentences $C$ to length zero.
\end{enumerate}

\medskip
\noindent\textbf{Part VI --- A fat cousin, and the moral.}
\begin{enumerate}[resume]
  \item (Fat Cantor set) Repeat the construction, but at stage
        $n$ ($n = 0, 1, 2, \dots$) remove from each of the $2^n$
        current segments a central open interval of length
        $4^{-(n+1)}$ only. Show that the segment lengths $l_n$
        obey $l_{n+1} = \frac{l_n - 4^{-(n+1)}}{2}$, $l_n =
        \frac{2^n + 1}{2\cdot 4^n} > 0$, that the resulting
        $K = \bigcap K_n$ is compact with empty interior, and
        that the total removed length is $\sum_{n\geq0} 2^n
        4^{-(n+1)} = \frac12$. Admitting the (intuitive, Year 3)
        additivity of length for finite unions of intervals, and
        using both statements of \cref{ex:b1:topology:nested},
        show that any finite family of open intervals covering
        $K$ has total length $\geq \frac12$: $K$ is nowhere
        dense but \emph{not} negligible. Smallness has several
        inequivalent meanings.
  \item (Distances) Show that for a nonempty closed $F \subseteq
        \R$ and $x \in \R$, the infimum $d(x, F)$ is
        \emph{attained} \emph{(minimizing sequence plus
        Bolzano--Weierstrass)}. Then compute
        \[
        \max_{y \in \intcc{0}{1}} d(y, C) = \frac16 ,
        \]
        attained exactly at the center $y = \frac12$ \emph{(a
        point of a gap created at stage $n$ is within
        $\frac{3^{-n}}{2}$ of the gap's endpoints, which lie in
        $C$)}.
  \item (Every point a subsequential limit) Using questions 12
        and 2, produce a single sequence in $C$ whose set of
        subsequential limits is \emph{all} of $C$. (Compare:
        for a convergent sequence that set is one point ---
        $C$ realizes the opposite extreme among compacts.)
  \item Synthesis, one sentence each: (i) which theorems of this
        chapter did the construction actually consume (stability
        of closed sets, compactness, sequential
        characterizations)? (ii) list the four paradoxical
        pairings of the portrait (length zero/uncountable,
        closed/empty interior, perfect/totally disconnected,
        negligible/$C+C$ full); (iii) where does $C$ resurface
        later (the devil's staircase built on $h$ in the theory
        of continuity, and the measure theory of the Year 3
        volume, where $C$ separates ``countable'' from
        ``negligible'')?
\end{enumerate}
\end{problem}
