\chapter{Taylor Formulas and Asymptotic Expansions}\label{ch:b1:taylor}

Near a point, a smooth function is as good as a polynomial --- with a
controllable error. The Taylor formulas make this exact in three
flavors (integral remainder, Lagrange remainder, Young remainder),
and the resulting \emph{asymptotic expansions}, manipulated
algebraically, become the sharpest tool of elementary analysis:
limits, equivalents, local behavior, asymptotes.

\section{Comparison notation}

\begin{definition}[Landau notation]\label{def:b1:taylor:landau}
Let $f, g$ be defined near $x_0$ ($x_0 \in \R$ or $\pm\infty$). One
writes, as $x \to x_0$:\index{Landau notation}
\begin{itemize}
  \item $f = o(g)$ (``little-o'') when $f = \varepsilon g$ with
        $\varepsilon(x) \to 0$;
  \item $f = O(g)$ (``big-O'') when $f = u g$ with $u$ bounded near
        $x_0$;
  \item $f \sim g$ (``equivalent''\index{equivalent functions}) when
        $f = (1 + \varepsilon) g$ with $\varepsilon \to 0$ ---
        equivalently $f - g = o(g)$.
\end{itemize}
The same notation applies to sequences ($n \to \infty$).
\end{definition}

\begin{proposition}[Rules]\label{prop:b1:taylor:landaurules}
As $x \to x_0$:
\begin{enumerate}
  \item $\sim$ is an equivalence relation; $f \sim g$ implies that
        $f$ and $g$ share limits, signs (near $x_0$), and zeros'
        absence;
  \item equivalents \emph{multiply and divide}: $f_1 \sim g_1$, $f_2
        \sim g_2$ imply $f_1 f_2 \sim g_1 g_2$ and $\frac{f_1}{f_2}
        \sim \frac{g_1}{g_2}$;
  \item equivalents do \emph{not} add: $x + 1 \sim x$ and $-x \sim
        -x + 2$ at $+\infty$, yet the sums $1$ and $2$ are not
        equivalent. To add, return to expansions with explicit
        $o(\cdot)$ terms;
  \item $o(g) + o(g) = o(g)$, $\;u \cdot o(g) = o(ug)$, $\;o(o(g)) =
        o(g)$, and $f \sim g \iff f = g + o(g)$.
\end{enumerate}
\end{proposition}

\begin{proof}
Each is a short manipulation of the definitions; for instance
$f_1 f_2 = (1+\varepsilon_1)(1+\varepsilon_2) g_1 g_2$ and $(1 +
\varepsilon_1)(1+\varepsilon_2) \to 1$. Two items of (4) deserve
their line. $u\cdot o(g) = o(ug)$: if $f = \varepsilon g$ with
$\varepsilon \to 0$, then $uf = \varepsilon\,(ug)$ with the same
$\varepsilon$. $o(o(g)) = o(g)$: if $f = \varepsilon_1 h$ and
$h = \varepsilon_2 g$ with both $\varepsilon_i \to 0$, then $f =
(\varepsilon_1\varepsilon_2) g$ and the product of the two
infinitesimals is one. The equivalence $f \sim g \iff f = g +
o(g)$ is the definition read twice: $f - g = \varepsilon g$.
The counterexample in (3) is the proof of (3).
\end{proof}

\begin{example}[The comparison scale]\label{ex:b1:taylor:scale}
As $x \to +\infty$, the standard scale reads, in increasing
order of strength:
\[
1 \;=\; o(\ln x), \quad
\ln x = o(x^{0.01}), \quad
x^{0.01} = o(\sqrt x), \quad
\sqrt x = o(x^{10}), \quad
x^{10} = o(\eu^{x}), \quad
\eu^x = o(\eu^{2x}) ,
\]
every step being an instance of the growth comparisons of
\cref{prop:b1:functions:powerrules} (powers beat logarithms,
exponentials beat powers, and within one family the exponent
decides). Two habits worth forming: first, an $O(\cdot)$ landing
in a smaller class upgrades silently ($O(\ln x)$ is also
$o(x^{0.01})$); second, at $x \to 0^+$ the whole ladder
reverses through the substitution $x \mapsto \frac1x$ ---
$\ln x = o(x^{-0.01})$ there, so ``$x^\alpha \ln x \to 0$''
holds for every $\alpha > 0$. Keeping the scale straight is
half of every asymptotic argument in
\cref{ch:b1:series}.
\end{example}

\begin{example}[Uniqueness of expansions, and a parity dividend]\label{ex:b1:taylor:uniqueness}
If a function admits two expansions at $0$ to the same order,
\[
a_0 + a_1 x + \dots + a_n x^n + o(x^n)
= b_0 + b_1 x + \dots + b_n x^n + o(x^n),
\]
then $a_k = b_k$ for every $k$: subtracting and setting $c_k =
a_k - b_k$, evaluate the identity $c_0 + c_1 x + \dots + c_n x^n
= o(x^n)$ at $x \to 0$ to get $c_0 = 0$; divide by $x$ and
repeat --- each division is legitimate because the remaining
expression is again $o(x^{n-k})$. Coefficients are therefore
intrinsic, and one may compute them by \emph{any} route (Taylor
derivatives, algebra on known expansions, integration): all
routes must agree. Dividend: an \emph{even} function has only
even powers in its expansion --- replace $x$ by $-x$ and invoke
uniqueness; likewise odd functions have odd powers. This is why
$\cos$ carries $o(x^{2p+1})$ rather than $o(x^{2p})$ in the
table below: the absent odd term is free information, one order
of precision for nothing.
\end{example}

\section{The three Taylor formulas}

\begin{theorem}[Taylor with integral remainder]\label{thm:b1:taylor:integral}
Let $f$ be of class $C^{n+1}$ on an interval containing $a$ and $x$.
Then\index{Taylor formula}
\[
f(x) = \sum_{k=0}^{n} \frac{f^{(k)}(a)}{k!}\,(x - a)^k
+ \int_a^x \frac{(x - t)^n}{n!}\, f^{(n+1)}(t)\, \dd t .
\]
\end{theorem}

\begin{proof}
Induction on $n$. For $n = 0$: $f(x) = f(a) + \int_a^x f'(t)\dd t$ is
the fundamental theorem (\cref{thm:b1:integration:ftc}). Step:
integrate the remainder by parts,
\[
\int_a^x \frac{(x-t)^n}{n!} f^{(n+1)}(t)\,\dd t
= \Bigl[-\frac{(x-t)^{n+1}}{(n+1)!} f^{(n+1)}(t)\Bigr]_a^x
+ \int_a^x \frac{(x-t)^{n+1}}{(n+1)!} f^{(n+2)}(t)\,\dd t ,
\]
the bracket contributing the term $\frac{f^{(n+1)}(a)}{(n+1)!}(x -
a)^{n+1}$.
\end{proof}

\begin{example}[An exact expansion with its remainder]\label{ex:b1:taylor:lnexact}
For $\ln(1 + x)$ the integral remainder can be made completely
explicit without differentiating anything $n$ times: integrate
the finite geometric identity $\frac{1}{1+t} =
\sum_{k=0}^{n-1}(-t)^k + \frac{(-t)^n}{1+t}$ from $0$ to $x$:
\[
\ln(1 + x) = \sum_{k=1}^{n} \frac{(-1)^{k-1}x^k}{k}
+ (-1)^n \int_0^x \frac{t^n}{1 + t}\,\dd t ,
\]
and for $0 \leq x \leq 1$ the remainder is bounded by $\int_0^x
t^n\,\dd t = \frac{x^{n+1}}{n+1}$. This is stronger than
Taylor--Young in two ways: it is an \emph{identity} valid for a
fixed $x$ (not only $x \to 0$), and the error bound is numerical.
The weekend problem (\cref{pb:b1:taylor:1}) lives on such exact
forms; Taylor--Young below is the lighter tool for limits, where
only the shape of the error matters.
\end{example}

\begin{theorem}[Taylor--Lagrange inequality]\label{thm:b1:taylor:lagrange}
Let $f$ be $C^{n+1}$ with $\abs{f^{(n+1)}} \leq M$ between $a$ and
$x$. Then
\[
\Bigl| f(x) - \sum_{k=0}^{n} \frac{f^{(k)}(a)}{k!}(x-a)^k \Bigr|
\leq M\, \frac{\abs{x - a}^{n+1}}{(n+1)!} .
\]
\end{theorem}

\begin{proof}
Bound the integral remainder: $\bigl|\int_a^x \frac{(x-t)^n}{n!}
f^{(n+1)}(t)\,\dd t\bigr| \leq M \bigl|\int_a^x
\frac{\abs{x-t}^n}{n!}\dd t\bigr| = M\frac{\abs{x-a}^{n+1}}{(n+1)!}$.
\end{proof}

\begin{example}[Certified numerics]\label{ex:b1:taylor:certified}
What is $\sqrt{1.02}$? Apply Taylor--Lagrange to $f(t) =
\sqrt{1 + t}$ at $a = 0$, order $2$, $x = 0.02$:
\[
\sqrt{1.02} \approx 1 + \frac{0.02}{2} - \frac{0.02^2}{8}
= 1.00995 ,
\qquad
\abs{f'''(t)} = \frac{3}{8}(1+t)^{-5/2} \leq \frac 38 ,
\]
so the error is at most $\frac38 \cdot \frac{0.02^3}{6} =
5\cdot10^{-7}$: $\sqrt{1.02} = 1.00995$ with six certified
decimals (true value $1.0099504938\dots$ --- the bound is nearly
sharp). The closing insight: Taylor--Young says only \emph{how
fast} the error vanishes; Taylor--Lagrange converts the same
polynomial into a \emph{certificate}, a number plus a proven
error bar. Whenever a decimal claim is made in this book, a
Lagrange-type bound is standing behind it; the weekend problem
(\cref{pb:b1:taylor:1}) industrializes the idea.
\end{example}

\begin{theorem}[Taylor--Young]\label{thm:b1:taylor:young}
Let $f$ be $n$ times differentiable at $a$. Then, as $x \to a$:
\[
f(x) = \sum_{k=0}^{n} \frac{f^{(k)}(a)}{k!}\,(x-a)^k
+ o\bigl((x-a)^n\bigr) .
\]
\end{theorem}

\begin{proof}
Induction on $n$. For $n = 1$ this is the definition of the
derivative (\cref{def:b1:derivative:def}). Assume the statement at
order $n - 1$, and let $f$ be $n$ times differentiable at $a$. Apply
the induction hypothesis to $f'$ (which is $n-1$ times differentiable
at $a$):
\[
f'(t) = \sum_{k=0}^{n-1} \frac{f^{(k+1)}(a)}{k!}(t-a)^k + r(t),
\qquad r(t) = o\bigl((t-a)^{n-1}\bigr).
\]
Let $g(x) = f(x) - \sum_{k=0}^{n} \frac{f^{(k)}(a)}{k!}(x-a)^k$; then
$g' = r$ and $g(a) = 0$. Given $\varepsilon > 0$, choose $\delta$
with $\abs{r(t)} \leq \varepsilon\abs{t - a}^{n-1}$ for $\abs{t-a}
\leq \delta$; the mean value inequality
(\cref{thm:b1:derivative:mvt}) applied on the segment from $a$ to
$x$ (where $\abs{g'} \leq \varepsilon\abs{x-a}^{n-1}$) yields
$\abs{g(x)} \leq \varepsilon\abs{x - a}^n$: exactly $g(x) =
o((x-a)^n)$.
\end{proof}

\begin{remark}[Three formulas, three prices, three products]
The hypotheses grade exactly with the conclusions.
Taylor--Young asks the least ($n$ derivatives \emph{at the
point} only) and yields the least: a qualitative $o((x-a)^n)$,
perfect for limits, useless for certified digits. The Lagrange
inequality asks for $C^{n+1}$ \emph{on the interval} and a bound
$M$ there, and returns a numerical error bar. The integral form
asks the same regularity and returns the most: the error as an
explicit object one can transform (integrate by parts, bound
piecewise, change variables) --- it is the form that powered the
irrationality machine of \cref{pb:b1:integration:1}. Choosing
the weakest formula that supports the goal is not pedantry: the
flat function of \cref{pb:b1:taylor:1} satisfies Taylor--Young
at every order while every stronger conclusion about it is
false away from $0$.
\end{remark}

\begin{proposition}[Standard expansions at $0$]
As $x \to 0$, for every fixed order $n$:
\index{Taylor expansion!standard}
\begin{align*}
\eu^x &= 1 + x + \frac{x^2}{2!} + \dots + \frac{x^n}{n!} + o(x^n),\\
\cos x &= 1 - \frac{x^2}{2!} + \frac{x^4}{4!} - \dots
+ \frac{(-1)^p x^{2p}}{(2p)!} + o(x^{2p+1}),\\
\sin x &= x - \frac{x^3}{3!} + \dots + \frac{(-1)^p
x^{2p+1}}{(2p+1)!} + o(x^{2p+2}),\\
\frac{1}{1 - x} &= 1 + x + x^2 + \dots + x^n + o(x^n),\\
\ln(1 + x) &= x - \frac{x^2}{2} + \frac{x^3}{3} - \dots +
\frac{(-1)^{n-1} x^n}{n} + o(x^n),\\
(1 + x)^\alpha &= 1 + \alpha x + \frac{\alpha(\alpha-1)}{2!}x^2 +
\dots + \binom{\alpha}{n} x^n + o(x^n),
\end{align*}
where $\binom{\alpha}{n} = \frac{\alpha(\alpha - 1)\cdots(\alpha - n
+ 1)}{n!}$ for real $\alpha$. ($\cosh$ and $\sinh$: same as $\cos$,
$\sin$ without the alternating signs.)
\label{prop:b1:taylor:standard}
\end{proposition}

\begin{proof}
Each function is smooth near $0$ with derivatives easy to evaluate:
$(\eu^x)^{(k)} = \eu^x$; the derivatives of $\sin$ and $\cos$ cycle
with period $4$; $\bigl((1+x)^\alpha\bigr)^{(k)} = \alpha(\alpha -
1)\cdots(\alpha - k + 1)(1+x)^{\alpha - k}$;
$\bigl(\ln(1+x)\bigr)^{(k)} = \frac{(-1)^{k-1}(k-1)!}{(1+x)^k}$.
Apply Taylor--Young at $a = 0$. (The geometric one is exact:
$\frac{1}{1-x} - \sum_0^n x^k = \frac{x^{n+1}}{1 - x} = o(x^n)$.)
\end{proof}

\begin{method}[Computing with expansions]\label{met:b1:taylor:compute}
\begin{enumerate}
  \item \emph{Fix the target order} $n$ first, and truncate every
        intermediate result there --- carrying higher terms is
        wasted work, dropping lower ones is an error.
  \item \emph{Sums, products}: expand each factor to order $n$ and
        multiply, discarding beyond $x^n$.
  \item \emph{Composition} $f(u(x))$ with $u(x) \to 0$: substitute
        the expansion of $u$ into that of $f$, order by order.
  \item \emph{Quotients}: write $\frac{1}{1 + v}$ with $v \to 0$ and
        use the geometric expansion.
  \item \emph{Integrate} an expansion term by term (differentiating
        requires more care --- justification: integral of $o(t^n)$
        from $0$ to $x$ is $o(x^{n+1})$, by direct bounding).
\end{enumerate}
\end{method}

\begin{example}[Composition, with the bookkeeping shown]\label{ex:b1:taylor:esin}
Expand $\eu^{\sin x}$ to order $3$. Inner expansion: $u = \sin x
= x - \frac{x^3}{6} + o(x^3)$, which indeed tends to $0$. Outer:
$\eu^u = 1 + u + \frac{u^2}{2} + \frac{u^3}{6} + o(u^3)$, and
$o(u^3) = o(x^3)$ since $u \sim x$. Powers of $u$, truncated at
$x^3$:
\[
u^2 = x^2 + o(x^3), \qquad u^3 = x^3 + o(x^3)
\]
(the cross term $2x\cdot(-\frac{x^3}{6})$ is already $x^4$).
Assemble:
\[
\eu^{\sin x} = 1 + \Bigl(x - \frac{x^3}{6}\Bigr) + \frac{x^2}{2}
+ \frac{x^3}{6} + o(x^3)
= 1 + x + \frac{x^2}{2} + o(x^3) :
\]
the two $x^3$ contributions cancel exactly. The closing insight:
$\eu^{\sin x}$ and $\eu^x$ agree to order $3$ --- not
because $\sin x \approx x$ crudely, but because the first
disagreement of the exponents ($-\frac{x^3}{6}$) enters
multiplied by $\eu^0 = 1$ and is then met by the cubic term of
the outer exponential; order-by-order bookkeeping detects such
conspiracies, eyeballing never does. (The next term is
$-\frac{x^4}{8}$: the truce ends at order $4$.)
\end{example}

\begin{example}\label{ex:b1:taylor:tan}
Expansion of $\tan$ at order $5$. Write $\tan x =
\sin x \cdot \frac{1}{\cos x}$:
\[
\frac{1}{\cos x} = \frac{1}{1 - \bigl(\frac{x^2}{2} - \frac{x^4}{24}
+ o(x^5)\bigr)}
= 1 + \Bigl(\frac{x^2}{2} - \frac{x^4}{24}\Bigr)
+ \Bigl(\frac{x^2}{2}\Bigr)^{\!2} + o(x^5)
= 1 + \frac{x^2}{2} + \frac{5x^4}{24} + o(x^5),
\]
then
\[
\tan x = \Bigl(x - \frac{x^3}{6} + \frac{x^5}{120}\Bigr)
\Bigl(1 + \frac{x^2}{2} + \frac{5x^4}{24}\Bigr) + o(x^5)
= x + \frac{x^3}{3} + \frac{2x^5}{15} + o(x^5) .
\]
\end{example}

\begin{omfigure}
\begin{tikzpicture}[x=1.15cm, y=1.05cm]
  \draw[-Stealth] (-3.5,0) -- (3.6,0) node[below, font=\small]
    {$x$};
  \draw[-Stealth] (0,-2.15) -- (0,2.2);
  \draw (3.1416,-0.05) -- (3.1416,0.05)
    node[below=3pt, font=\small] {$\pi$};
  \draw (-3.1416,-0.05) -- (-3.1416,0.05)
    node[below=3pt, font=\small] {$-\pi$};
  \draw (-0.05,1) -- (0.05,1) node[above left, font=\small,
    inner sep=1pt] {$1$};
  \draw (-0.05,-1) -- (0.05,-1) node[below right, font=\small,
    inner sep=1pt] {$-1$};
  \draw[omThm, very thick, smooth]
    plot[domain=-3.3:3.3, samples=120] (\x, {sin(\x r)});
  \draw[omDef, thick]
    plot[domain=-2.0:2.0] (\x, \x)
    node[right, font=\small] {$T_1$};
  \draw[omDef, thick, dashed, smooth]
    plot[domain=-2.7:2.7, samples=80] (\x, {\x - \x^3/6})
    node[below right, font=\small, inner sep=1pt] {$T_3$};
  \draw[omDef, thick, dotted, smooth]
    plot[domain=-3.25:3.25, samples=90]
    (\x, {\x - \x^3/6 + \x^5/120})
    node[above right, font=\small, inner sep=1pt] {$T_5$};
\end{tikzpicture}

{\small The sine (dark) against its Taylor polynomials at $0$:
$T_1 = x$, $T_3 = x - \frac{x^3}{6}$ (dashed), $T_5 = x -
\frac{x^3}{6} + \frac{x^5}{120}$ (dotted). Each new pair of terms
hugs the curve over a visibly wider window, but every polynomial
eventually tears away: a Taylor expansion is a \emph{local}
contract, sharpened at $0$ and silent far away. The
Taylor--Lagrange bound $\frac{\abs{x}^{n+1}}{(n+1)!}$ quantifies
the window; the flat function of \cref{pb:b1:taylor:1} shows the
contract can even be empty beyond the point itself.}
\end{omfigure}

\section{Applications}

\begin{example}[Limits]\label{ex:b1:taylor:limits}
\[
\lim_{x \to 0} \frac{x - \sin x}{x^3}:
\qquad
x - \sin x = \frac{x^3}{6} + o(x^3) \sim \frac{x^3}{6},
\qquad\text{so the limit is } \frac16
\]
--- settling the question raised in \cref{exo:b1:functions:9}.
Likewise $\displaystyle\lim_{x\to0}\Bigl(\frac{\sin
x}{x}\Bigr)^{1/x^2}$: the logarithm is
\[
\frac{1}{x^2}\ln\Bigl(1 - \frac{x^2}{6} + o(x^2)\Bigr)
= \frac{1}{x^2}\Bigl(-\frac{x^2}{6} + o(x^2)\Bigr)
\longrightarrow -\frac16,
\qquad\text{limit } \eu^{-1/6}.
\]
\end{example}

\begin{remark}[Common pitfalls with expansions]
(i) \emph{Never add or subtract equivalents}: from $\tan x \sim
x$ and $\sin x \sim x$ one may \emph{not} conclude $\tan x -
\sin x \sim 0$ (meaningless) --- the honest route is
expansions:
\[
\tan x - \sin x
= \Bigl(x + \frac{x^3}{3}\Bigr) - \Bigl(x - \frac{x^3}{6}\Bigr)
+ o(x^3) = \frac{x^3}{2} + o(x^3) \sim \frac{x^3}{2} .
\]
(ii) \emph{Expand past the massacre}: in the same computation,
order $1$ sees only $x - x = 0$; whenever leading terms cancel,
raise the order until a nonzero coefficient survives, and only
then convert back to an equivalent. (iii) \emph{Equivalents do
not pass through exponentials}: $n^2 + n \sim n^2$, yet
$\eu^{n^2+n} = \eu^{n}\,\eu^{n^2}$ is \emph{not} equivalent to
$\eu^{n^2}$ --- exponentiate only \emph{expansions of the
exponent} whose error tends to $0$, never equivalents of the
exponent. (Logarithms are safer: if $u_n \sim v_n \to \ell \neq
1$, $\ell > 0$, then $\ln u_n \sim \ln v_n$.) (iv) \emph{The
$o(\cdot)$ calculus is one-directional}: $o(x^2) + o(x^2) =
o(x^2)$, $5\,o(x^2) = o(x^2)$, $x\cdot o(x^2) = o(x^3)$ --- but
an $o(x^2)$ is not a specific function, so never cancel two of
them against each other: $o(x^2) - o(x^2)$ is $o(x^2)$, not
$0$.
\end{remark}

\begin{proposition}[Local behavior]\label{prop:b1:taylor:local}
Suppose $f(x) = f(a) + c\,(x - a)^p + o\bigl((x-a)^p\bigr)$ with $c
\neq 0$ (first nonzero term after the constant; $p \geq 2$ at a
critical point).
\begin{itemize}
  \item If $p$ is even: $f$ has a local minimum at $a$ if $c > 0$, a
        local maximum if $c < 0$.
  \item If $p$ is odd: no extremum ($f - f(a)$ changes sign);
        if moreover the expansion starts after a linear term $f'(a)(x
        - a)$, the graph crosses its tangent: an \emph{inflection}.
\end{itemize}
\end{proposition}

\begin{proof}
Near $a$, $f(x) - f(a) = (x-a)^p\bigl(c + o(1)\bigr)$ has the sign
of $c\,(x-a)^p$: constant sign for $p$ even, changing for $p$ odd.
\end{proof}

\begin{example}[Exponents must be expanded to $o(1)$]\label{ex:b1:taylor:expdiscipline}
Find an equivalent of $u_n = \bigl(1 + \frac1n\bigr)^{n^2}$.
Expand the \emph{exponent} until its error tends to $0$:
\[
n^2 \ln\Bigl(1 + \frac1n\Bigr)
= n^2\Bigl(\frac1n - \frac{1}{2n^2} +
O\Bigl(\frac{1}{n^3}\Bigr)\Bigr)
= n - \frac12 + O\Bigl(\frac1n\Bigr),
\]
so $u_n = \eu^{\,n - 1/2}\,\eu^{O(1/n)}$ with $\eu^{O(1/n)} \to
1$:
\[
u_n \;\sim\; \eu^{\,n - \frac12} .
\]
Note what would have gone wrong with less care: stopping the
exponent at $n^2\cdot\frac1n = n + O(1)$ leaves an
$\eu^{O(1)}$ factor --- bounded but not tending to $1$ --- and
no equivalent can be asserted. The rule of
the pitfalls above, in positive form: an equivalent of
$\eu^{a_n}$ requires the expansion of $a_n$ \emph{up to a term
tending to zero}, every coefficient before that being kept
exactly.
\end{example}

\begin{example}[Classifying a flat critical point]\label{ex:b1:taylor:flatmin}
Study $f(x) = \cos x + \frac{x^2}{2}$ near $0$. Both $f'(0) =
0$ and $f''(0) = -\cos 0 + 1 = 0$: the second-derivative test is
mute. Expand instead:
\[
f(x) = \Bigl(1 - \frac{x^2}{2} + \frac{x^4}{24} +
o(x^4)\Bigr) + \frac{x^2}{2}
= 1 + \frac{x^4}{24} + o(x^4) :
\]
first nonzero term $c\,x^p$ with $p = 4$ even and $c =
\frac{1}{24} > 0$: a local minimum, of unusual flatness (the
graph leaves its minimum value like $x^4$, not $x^2$). The
closing insight: the expansion sees in one line what iterated
differentiation obscures --- and \cref{prop:b1:taylor:local} is
the systematic dictionary from ``first surviving term'' to
``local shape''.
\end{example}

\begin{example}[Expansions at infinity]\label{ex:b1:taylor:atinfinity}
Two computations where the variable runs to $+\infty$ and the
substitution $h = \frac1x \to 0^+$ imports the whole toolbox.
First,
\[
x - x^2\ln\Bigl(1 + \frac1x\Bigr)
= x - x^2\Bigl(\frac1x - \frac{1}{2x^2} +
O\Bigl(\frac{1}{x^3}\Bigr)\Bigr)
= \frac12 + O\Bigl(\frac1x\Bigr) \longrightarrow \frac12 .
\]
Second, the arctangent at infinity: from $\arctan x +
\arctan\frac1x = \frac\pi2$ for $x > 0$
(\cref{prop:b1:functions:arcidentities}) and the expansion of
$\arctan$ at $0$ (\cref{exo:b1:taylor:3}),
\[
\arctan x = \frac\pi2 - \arctan\frac1x
= \frac\pi2 - \frac1x + \frac{1}{3x^3} +
o\Bigl(\frac{1}{x^3}\Bigr) :
\]
the graph approaches its asymptote $y = \frac\pi2$ from below at
speed $\frac1x$. The closing insight: there is no separate
theory of expansions at infinity --- one reciprocal substitution
reduces them to expansions at $0$, provided every intermediate
$O$ and $o$ is carried along honestly.
\end{example}

\begin{example}[Asymptote by expansion]\label{ex:b1:taylor:asymptote}
As $x \to +\infty$,
\[
\sqrt{x^2 + x} = x\sqrt{1 + \tfrac1x}
= x\Bigl(1 + \frac{1}{2x} - \frac{1}{8x^2} + o\bigl(\tfrac{1}{x^2}\bigr)\Bigr)
= x + \frac12 - \frac{1}{8x} + o\bigl(\tfrac 1x\bigr):
\]
the line $y = x + \frac12$ is an asymptote, approached \emph{from
below} (the next term $-\frac{1}{8x}$ is negative).
\end{example}

\begin{remark}[Where expansions work next]
Asymptotic expansions are the standing language of the rest of
the book: in \cref{ch:b1:series} they decide convergence
(equivalents feed the comparison tests, and the study of $\sum
\frac{1}{n^\alpha}$ is an expansion in disguise); in the Year 2
volume they become \emph{power series}, where the Taylor
polynomial acquires infinitely many terms and a radius of
convergence; and every linearization in physics --- the pendulum,
first-order perturbation --- is a Taylor--Young statement with
the $o(\cdot)$ silently dropped. The one warning worth engraving:
an expansion describes a function only \emph{near one point} ---
see the flat function of the weekend problem, whose expansion at
$0$ is identically zero without the function being so.
\end{remark}

\begin{remark}[Perspectives inside this volume]
Expansions are the working language of the remaining analysis
and of the geometry to come. \Cref{ch:b1:series} converts them
into convergence verdicts: an equivalent of the general term is
an expansion truncated at its first term, and the finer tests
(alternating with error control) consume the second term too.
\Cref{ch:b1:curves} reads local geometry from expansions of the
two coordinate functions: whether a parametrized curve crosses,
kisses, or cusps at a point is decided by which powers of $t$
survive in $x(t)$ and $y(t)$ --- the plane version of
\cref{prop:b1:taylor:local}. And \cref{ch:b1:multivar} stops at
order one on purpose: the tangent plane is a two-variable
Taylor--Young statement, with the full second-order theory
(Hessians, saddle points) deferred to the Year 2 volume. The
common thread: every ``local'' question in this book is answered
by writing down the first surviving term of an expansion.
\end{remark}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:b1:taylor:1}
Give the expansions at $0$: $\eu^{2x}$ to order $3$;
$\;\ln(1 - x)$ to order $4$; $\;\sqrt{1 + x}$ to order $3$;
$\;\dfrac{1}{1 + x^2}$ to order $6$.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b1:taylor:2}
Compute the limits:
\[
\lim_{x\to 0} \frac{\eu^x - 1 - x}{x^2},
\qquad
\lim_{x\to 0} \frac{\cos x - \sqrt{1 - x^2}}{x^4},
\qquad
\lim_{x\to 0} \frac{\ln(1+x) - \sin x}{x^2}.
\]
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b1:taylor:3}
Expand $\arctan x$ at $0$ to order $5$ by integrating the expansion
of $\frac{1}{1 + x^2}$, and $\arcsin x$ to order $5$ by integrating
that of $(1 - x^2)^{-1/2}$.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b1:taylor:4}
Using Taylor--Lagrange for $\exp$ on $\intcc{0}{1}$, prove that
\[
\Bigl| \eu - \sum_{k=0}^{n} \frac{1}{k!} \Bigr| \leq
\frac{3}{(n+1)!},
\]
and determine an $n$ guaranteeing $6$ exact decimals of $\eu$.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:taylor:5}
Expand to order $2$ in $\frac1n$ and deduce the limit and the
speed of convergence:
\[
\Bigl(1 + \frac 1n\Bigr)^{\!n}
= \eu\Bigl(1 - \frac{1}{2n} + \frac{11}{24n^2} +
o\Bigl(\frac{1}{n^2}\Bigr)\Bigr).
\]
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:taylor:6}
Study the local behavior at $0$ of $f(x) = x^2 - x^4$ and of $g(x) =
x^3 + x^5$; and find the position of the graph of $h(x) = \eu^x$
relative to its tangent at $a = 1$, locally then globally.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:taylor:7}
Determine the asymptotes at $\pm\infty$ of $f(x) =
\sqrt[3]{x^3 + x^2}$ and the position of the curve relative to
them.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:taylor:8}
Find the equivalent, as $n \to \infty$, of
\[
u_n = \sqrt{n+1} - \sqrt n, \qquad
v_n = \ln(n+1) - \ln n, \qquad
w_n = \sin\frac{1}{n} - \tan\frac{1}{n},
\]
each as a power of $n$ times a constant.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b1:taylor:9}
Let $f$ be $C^2$ on $\R$. Prove that for every $x$ and $h > 0$:
\[
\abs{f'(x)} \leq \frac{\abs{f(x+h) - f(x-h)}}{2h} +
\frac{h}{2}\sup_{\intcc{x-h}{x+h}}\abs{f''} ,
\]
and deduce the Landau--Kolmogorov-type inequality: if $\abs f \leq
M_0$ and $\abs{f''} \leq M_2$ on $\R$, then $\abs{f'} \leq
\sqrt{2 M_0 M_2}$ everywhere. \emph{(Optimize over $h$.)}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b1:taylor:10}
The sequence $u_0 \in \intoo{0}{\pi}$, $u_{n+1} = \sin u_n$
decreases to $0$ (justify briefly). To find its speed, consider $v_n
= \frac{1}{u_n^2}$:
\begin{enumerate}
  \item using the expansion of $\sin$, prove $v_{n+1} - v_n \to
        \frac13$;
  \item with Ces\`aro (\cref{exo:b1:seq:10}), deduce $\frac{v_n}{n}
        \to \frac13$, then the equivalent $u_n \sim
        \sqrt{\dfrac{3}{n}}$.
\end{enumerate}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:taylor:11}
(Differences of infinities) Compute
\[
\lim_{x \to 0} \Bigl(\frac{1}{x^2} - \frac{1}{\sin^2 x}\Bigr)
\qquad\text{and}\qquad
\lim_{x \to 0^+} \Bigl(\frac 1x - \frac{1}{\ln(1 + x)}\Bigr)
\]
by reducing to a common denominator and expanding numerator and
denominator separately.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b1:taylor:12}
(Asymptotics of implicit roots) Show that for every $k \in
\N^*$ the equation $\tan x = x$ has exactly one solution $x_k$
in $\intoo{k\pi - \frac\pi2}{k\pi + \frac\pi2}$, that $x_k = k\pi
+ \frac\pi2 - \varepsilon_k$ with $\varepsilon_k =
\arctan\frac{1}{x_k}$, and deduce the expansion
\[
x_k = k\pi + \frac\pi2 - \frac{1}{k\pi} + o\Bigl(\frac
1k\Bigr) \qquad (k \to \infty).
\]
\end{exercise}

\section{Problem: Alternating sums, certified digits, and the
irrationality of $\cos 1$}

\begin{problem}[{Weekend problem --- the alternating estimate
$\abs{S - S_n} \leq a_{n+1}$: $\ln 2$ and $\pi$ with proven
decimals, Machin's formula, and $\cos 1 \notin \Q$}]
\label{pb:b1:taylor:1}
An alternating sum with decreasing terms is the friendliest
object in numerical analysis: its error is bounded by the first
omitted term, \emph{with known sign}. This problem proves that
principle with the adjacent-sequences theorem, then spends it
three ways: certified decimals for $\ln 2$ (three competing
routes) and for $\pi$ (Leibniz, then Machin's 1706 formula, still
the idea behind record computations for centuries), the
irrationality of $\cos 1$, $\sin 1$ and $\cosh 1$, and, as a
counterweight, the Taylor--Lagrange \emph{equality} and the flat
function whose Taylor expansion lies.\index{alternating series
estimate}\index{Machin's formula}\index{irrationality!of cos 1@of
$\cos 1$} Throughout, ``series'' language is informal: every sum
here is a \emph{sequence of partial sums}, as in
\cref{ex:b1:seq:e}; the theory proper opens in
\cref{ch:b1:series}.

\medskip
\noindent\textbf{Part I --- The alternating estimate.} Let
$(a_k)_{k \geq 0}$ decrease to $0$ and $S_n = \sum_{k=0}^{n}
(-1)^k a_k$.
\begin{enumerate}
  \item Show that $(S_{2n+1})$ is nondecreasing, $(S_{2n})$
        nonincreasing, and that they are adjacent
        (\cref{thm:b1:seq:adjacent}): both converge to a common
        $S$ with, for every $n$,
        \[
        S_{2n+1} \leq S \leq S_{2n},
        \qquad
        \abs{S - S_n} \leq a_{n+1} ,
        \]
        the error having the sign of the first omitted term.
        Show moreover that if the decrease is \emph{strict}, all
        these inequalities are strict.
  \item First dividend: for $x = 1$ in the exponential series,
        compare with \cref{thm:b1:taylor:lagrange} at $a = 0$:
        show that $T_n = \sum_{k=0}^{n} \frac{(-1)^k}{k!}$
        converges to $\eu^{-1}$ with $\abs{\eu^{-1} - T_n} <
        \frac{1}{(n+1)!}$.
  \item (Leibniz, 1674) From the \emph{exact} finite identity
        \[
        \frac{1}{1 + t^2} = \sum_{k=0}^{n} (-1)^k t^{2k}
        + \frac{(-1)^{n+1} t^{2n+2}}{1 + t^2} ,
        \]
        integrated over $\intcc{0}{1}$, prove
        \[
        \frac\pi4 = \sum_{k=0}^{n} \frac{(-1)^k}{2k+1} + \rho_n,
        \qquad \abs{\rho_n} \leq \frac{1}{2n+3} .
        \]
  \item Slowness: how many terms of Leibniz guarantee six exact
        decimals of $\pi$? (About two million.) Evaluate $4 S_4
        = 4\bigl(1 - \frac13 + \frac15 - \frac17 +
        \frac19\bigr)$ and its distance to $\pi$, to feel the
        pain.
\end{enumerate}

\medskip
\noindent\textbf{Part II --- $\ln 2$ three ways.}
\begin{enumerate}[resume]
  \item (Way 1: alternating harmonic) From $\frac{1}{1+t} =
        \sum_{k=0}^{n-1}(-1)^k t^k + \frac{(-1)^n t^n}{1+t}$
        integrated over $\intcc{0}{1}$:
        \[
        \ln 2 = \sum_{k=1}^{n} \frac{(-1)^{k-1}}{k} + (-1)^n
        R_n, \qquad \frac{1}{2(n+1)} \leq R_n \leq
        \frac{1}{n+1} :
        \]
        the error is of exact order $\frac 1n$ --- a million
        terms for six decimals.
  \item (Way 2: the fast series) Integrate $\frac{1}{1 - t^2} =
        \sum_{k=0}^{n} t^{2k} + \frac{t^{2n+2}}{1-t^2}$ from $0$
        to $x \in \intoo{0}{1}$ and evaluate at $x = \frac13$
        (note $\frac{1 + 1/3}{1 - 1/3} = 2$):
        \[
        \ln 2 = 2\sum_{k=0}^{n} \frac{(1/3)^{2k+1}}{2k+1} +
        \tilde\rho_n,
        \qquad
        0 < \tilde\rho_n \leq \frac{9}{4}\cdot
        \frac{(1/3)^{2n+3}}{2n+3} :
        \]
        geometric convergence, roughly one digit per term.
  \item (Way 3: Riemann sums and a hidden identity) Prove by
        induction the identity
        \[
        \sum_{k=1}^{2n} \frac{(-1)^{k-1}}{k} = H_{2n} - H_n =
        \sum_{k=1}^{n} \frac{1}{n + k} ,
        \]
        and recover $\ln 2$ as the Riemann-sum limit of
        \cref{ex:b1:integration:riemannexample}: Ways 1 and 3
        are secretly the same number seen twice.
  \item Shootout at six terms: compare $\sum_{k=1}^{6}
        \frac{(-1)^{k-1}}{k} = 0.6167$ with Way 2 at $n = 5$,
        which already gives $\ln 2 = 0.693147$ with error $\leq
        1.1\cdot10^{-7}$. Explain the structural reason
        (evaluation point deep inside the interval of
        convergence versus on its boundary).
  \item How many terms of Way 2 certify \emph{ten} decimals of
        $\ln 2$? Show $n = 10$ suffices.
\end{enumerate}

\medskip
\noindent\textbf{Part III --- Machin's formula.}
\begin{enumerate}[resume]
  \item Compute $(5 + \iu)^4$ and verify the complex identity
        \[
        (5 + \iu)^4 = 2\,(1 + \iu)\,(239 + \iu) .
        \]
        Taking arguments (\cref{ch:b1:complex} conventions),
        deduce Machin's formula
        \[
        \frac\pi4 = 4\arctan\frac15 - \arctan\frac{1}{239} .
        \]
        (Check that no argument leaves
        $\intoo{-\frac\pi2}{\frac\pi2}$.)
  \item As in question 3, establish for $0 < x < 1$:
        \[
        \arctan x = \sum_{k=0}^{n} \frac{(-1)^k
        x^{2k+1}}{2k+1} + r_n(x), \qquad \abs{r_n(x)} \leq
        \frac{x^{2n+3}}{2n+3} .
        \]
  \item Certify $\pi$ to seven decimals with six terms: bound
        the total error of
        \[
        \pi \approx 16\sum_{k=0}^{4}
        \frac{(-1)^k (1/5)^{2k+1}}{2k+1}
        - 4\Bigl(\frac{1}{239} - \frac{1}{3\cdot239^3}\Bigr)
        \]
        by $16\,\frac{(1/5)^{11}}{11} +
        4\,\frac{(1/239)^5}{5} < 5\cdot10^{-8}$, and give the
        resulting value $3.1415926\dots$
  \item Compare the three routes to $\pi$ now available ---
        Leibniz (question 4), the Dalzell integrals of
        \cref{pb:b1:integration:1} (error $4^{1-5m}$), Machin
        (error $\approx 16\cdot 5^{-(2n+3)}$) --- in digits per
        term, and explain why shrinking the evaluation point
        beats everything.
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- The integer trap, alternating
edition.}
\begin{enumerate}[resume]
  \item Suppose $\cos 1 = \frac pq$. Multiply the strict
        alternating bracketing of $\sum_{k}
        \frac{(-1)^k}{(2k)!}$ (question 1) by $(2n)!$ with $2n
        \geq q$, and derive a contradiction: $\cos 1$ is
        irrational.
  \item Adapt to $\sin 1 = \sum_k \frac{(-1)^k}{(2k+1)!}$
        (multiply by $(2n+1)!$): $\sin 1 \notin \Q$. Both
        irrational, yet $\cos^2 1 + \sin^2 1 = 1$: irrationality
        is not stable under algebra.
  \item The non-alternating cousin: $\cosh 1 = \sum_k
        \frac{1}{(2k)!}$ (in the partial-sum sense, with the
        two-sided tail bound $0 < \cosh 1 - \sum_{k \leq n}
        \frac{1}{(2k)!} < \frac{2}{(2n+2)!}$, to be proved).
        Conclude $\cosh 1 \notin \Q$ by the same trap.
  \item Push to $\cos\frac 1m$ for every integer $m \geq 1$:
        multiply by $m^{2n}(2n)!$ and conclude $\cos\frac1m
        \notin \Q$. Where does the same attempt break for
        $\cos\frac ab$ with $b > 1$ general? (Identify the
        denominator that no longer clears.)
\end{enumerate}

\medskip
\noindent\textbf{Part V --- Sharper and darker: the equality
form, and a function that fools Taylor.}
\begin{enumerate}[resume]
  \item (Taylor--Lagrange, equality form) Let $f$ be $n + 1$
        times differentiable between $a$ and $x$. Define $g(t) =
        f(x) - \sum_{k=0}^{n} \frac{f^{(k)}(t)}{k!}(x - t)^k -
        A\,\frac{(x-t)^{n+1}}{(n+1)!}$ with the constant $A$
        chosen so that $g(a) = 0$. Compute $g'$ (the sum
        telescopes), apply Rolle on $\intcc{a}{x}$, and conclude
        that there exists $c$ strictly between $a$ and $x$ with
        \[
        f(x) = \sum_{k=0}^{n} \frac{f^{(k)}(a)}{k!}(x-a)^k
        + \frac{f^{(n+1)}(c)}{(n+1)!}\,(x-a)^{n+1} .
        \]
  \item Dividend of the equality: for $x > 0$ show
        \[
        \eu^x > 1 + x + \frac{x^2}{2!} + \dots +
        \frac{x^n}{n!}
        \]
        (strictly, for every $n$), and locate where the
        inequality reverses for $x < 0$ according to the parity
        of $n$.
  \item Benchmark the remainders on $\sin(0.5)$ at order $3$:
        Young gives only $o(x^3)$ (no number); Lagrange gives
        $\abs{\sin 0.5 - (0.5 - \frac{0.5^3}{6})} \leq
        \frac{0.5^5}{120} = 2.61\cdot10^{-4}$; the alternating
        estimate gives the same bound \emph{plus} the sign
        information $\sin 0.5 > 0.5 - \frac{0.5^3}{6}$. Compare
        with the true error $2.59\cdot10^{-4}$: the bound is
        nearly attained. Which tool would you reach for, and
        when?
  \item (The flat function) Let $f(x) = \eu^{-1/x^2}$ for $x
        \neq 0$, $f(0) = 0$. Show $f$ is continuous at $0$, that
        $f'(0) = 0$, and more generally --- by proving that
        every derivative has the form $f^{(k)}(x) =
        P_k\bigl(\frac1x\bigr)\eu^{-1/x^2}$ for a polynomial
        $P_k$ (induction) --- that $f^{(k)}(0) = 0$ for all $k$
        \emph{(growth comparison
        \cref{prop:b1:functions:powerrules})}. Conclude: all
        Taylor polynomials of $f$ at $0$ vanish, yet $f(x) > 0$
        for $x \neq 0$: Taylor--Young holds at every order, and
        says nothing about $f$ away from $0$. Expansions
        describe \emph{germs}, not functions.
\end{enumerate}

\medskip
\noindent\textbf{Part VI --- Synthesis.}
\begin{enumerate}[resume]
  \item Run the trap once more, on $\eu^{-1} = \sum_k
        \frac{(-1)^k}{k!}$: multiply the strict alternating
        bracketing by $n!$ and conclude $\eu^{-1} \notin \Q$,
        hence $\eu \notin \Q$ --- the third proof of this fact
        in the volume. List the three (adjacent sequences,
        \cref{exo:b1:seq:9}; integrals,
        \cref{pb:b1:integration:1}; alternating sums, here) and
        what each needed.
  \item The fine print: monotonicity is not decorative. Let
        $b_k = \frac1k$ for odd $k$ and $b_k = \frac{1}{k^2}$
        for even $k$: the $b_k$ are positive and tend to $0$,
        yet the partial sums of $\sum (-1)^k b_k$ diverge to
        $-\infty$. Prove it \emph{(split the partial sum into
        the even part, bounded via
        \cref{ex:b1:seq:basel}, and the odd part, which
        dominates half the harmonic series,
        \cref{exo:b1:seq:5})}, and say exactly which step of
        question 1 used monotonicity.
  \item Verify Euler's simpler identity $\arctan\frac12 +
        \arctan\frac13 = \frac\pi4$ via $(2 + \iu)(3 + \iu) =
        5(1 + \iu)$, estimate the terms needed for six decimals
        of $\pi$ by this route ($n = 10$ suffices), and place it
        between Leibniz and Machin in the ranking of question
        13.
  \item Synthesis, one sentence each: (i) state the alternating
        estimate and its two outputs (bound and sign); (ii) why
        exact finite identities with explicit remainders beat
        limit statements for certified numerics; (iii) inventory
        of the problem ($\pi$ to $10^{-7}$ by hand, $\ln 2$ to
        ten decimals, four irrationality proofs, one equality
        theorem, one warning example); (iv) which of these
        threads \cref{ch:b1:series} will pick up (the alternating
        series test, absolute versus conditional convergence,
        and the rearrangement drama of its weekend problem).
\end{enumerate}
\end{problem}
