\chapter{Numerical Series}\label{ch:b1:series}

Summing infinitely many numbers means taking the limit of the partial
sums --- nothing more, nothing less. This chapter sets up the
definitions and the convergence tests usable in first year:
comparison and equivalents for positive terms, the ratio test, the
integral comparison giving the Riemann series, absolute convergence,
and the alternating series theorem. The finer theory (products of
series, summation by packets, series of functions) belongs to the
second year.

\section{Generalities}

\begin{definition}\label{def:b1:series:def}
Given a sequence $(u_n)$, the \emph{series}\index{series} $\sum u_n$
is the sequence of \emph{partial sums} $S_N = \sum_{n=0}^{N} u_n$.
The series \emph{converges} when $(S_N)$ converges; the limit is the
\emph{sum} $\sum_{n=0}^{\infty} u_n$, and $R_N = \sum_{n > N} u_n =
S - S_N$ is the \emph{remainder}, which tends to $0$.
\end{definition}

\begin{example}[Geometric series]\label{ex:b1:series:geometric}
For $q \in \C$: $\;S_N = \sum_{n=0}^{N} q^n = \frac{1 -
q^{N+1}}{1-q}$ ($q \neq 1$). The series converges iff $\abs q < 1$
(\cref{exo:b1:seq:3}), with\index{geometric series}
\[
\sum_{n=0}^{\infty} q^n = \frac{1}{1 - q} .
\]
\end{example}

\begin{example}[Periodic decimals are geometric series]\label{ex:b1:series:periodicdecimal}
What number is $0.363636\dots$? Its very writing is a series:
\[
0.\overline{36} = \sum_{k=1}^{\infty} \frac{36}{100^k}
= 36\cdot\frac{1/100}{1 - 1/100} = \frac{36}{99} =
\frac{4}{11} ,
\]
by the geometric sum with $q = \frac{1}{100}$. In general a
block $B$ of $p$ digits repeating forever is worth
$\frac{B}{10^p - 1}$ --- the mechanism behind the periodicity
criterion of \cref{pb:b1:reals:1}, which this chapter's language
finally states in one line: a decimal expansion is a convergent
series, eventually periodic exactly when its sum is rational.
The digit machinery of Chapter 10, built there with bare
suprema, was series theory traveling incognito.
\end{example}

\begin{proposition}[First facts]\label{prop:b1:series:first}
\begin{enumerate}
  \item If $\sum u_n$ converges, then $u_n \to 0$. (The converse is
        \emph{false}: the harmonic series.)
  \item Linearity: convergent series add and scale, with the expected
        sums.
  \item (Telescoping\index{telescoping series}) $\sum (v_{n+1} -
        v_n)$ converges iff $(v_n)$ converges, with sum $\lim v_n -
        v_0$.
  \item Changing finitely many terms does not affect convergence
        (only the sum).
\end{enumerate}
\end{proposition}

\begin{proof}
(1) $u_N = S_N - S_{N-1} \to S - S = 0$. The harmonic series has
$u_n = \frac1n \to 0$ yet diverges (\cref{exo:b1:seq:5}). (2)
Operations on limits. (3) $S_N = v_{N+1} - v_0$. (4) The partial
sums change by an eventually constant amount.
\end{proof}

\begin{example}[Planning digits with the geometric remainder]\label{ex:b1:series:georemainder}
For $\abs q < 1$ the remainder of the geometric series is
explicit:
\[
R_N = \sum_{n = N+1}^{\infty} q^n = \frac{q^{N+1}}{1 - q} .
\]
This converts accuracy goals into term counts before any
computation. To evaluate $\sum_{n\geq0} \bigl(\frac13\bigr)^n =
\frac32$ within $10^{-10}$: need $\frac{(1/3)^{N+1}}{2/3} \leq
10^{-10}$, i.e.\ $3^{N} \geq \frac{3}{2}\cdot 10^{10}$, i.e.\ $N
\geq 22$ (as $3^{22} \approx 3.1\cdot10^{10}$): twenty-three
terms, known in advance. Every geometric-rate estimate of the
weekend problems (the $\frac13$-series for $\ln 2$, Machin's
arctangents in \cref{pb:b1:taylor:1}) is this two-line budget
in professional dress.
\end{example}

\begin{example}[A longer telescope]\label{ex:b1:series:telescope3}
Compute $\sum_{n\geq1} \frac{1}{n(n+1)(n+2)}$. Partial
fractions (\cref{ch:b1:fractions}):
\[
\frac{1}{n(n+1)(n+2)}
= \frac{1/2}{n} - \frac{1}{n+1} + \frac{1/2}{n+2}
= \frac12\Bigl(\frac{1}{n(n+1)} - \frac{1}{(n+1)(n+2)}\Bigr),
\]
where the second form --- a difference of consecutive values of
$w_n = \frac{1}{n(n+1)}$ --- is the telescoping one. Hence
\[
\sum_{n=1}^{N} \frac{1}{n(n+1)(n+2)}
= \frac12\Bigl(w_1 - w_{N+1}\Bigr)
= \frac12\Bigl(\frac12 - \frac{1}{(N+1)(N+2)}\Bigr)
\longrightarrow \frac14 .
\]
The closing insight: three-term partial fractions rarely
telescope as written; regroup them into a difference $w_n -
w_{n+1}$ first --- the reward is not just convergence but the
exact sum, which no comparison test ever delivers.
\end{example}

\section{Series with nonnegative terms}

\begin{theorem}[Bounded partial sums]\label{thm:b1:series:positive}
If $u_n \geq 0$ for all $n$, the partial sums increase, so: $\sum
u_n$ converges $\iff$ its partial sums are bounded above. Hence the
\emph{comparison test}: if $0 \leq u_n \leq v_n$ for all (large)
$n$,
\[
\sum v_n \text{ converges} \implies \sum u_n \text{ converges},
\qquad
\sum u_n \text{ diverges} \implies \sum v_n \text{ diverges}.
\]
And the \emph{equivalents test}: if $u_n \sim v_n$ with $v_n \geq
0$, the two series have the same nature.
\end{theorem}

\begin{proof}
Monotone limit theorem (\cref{thm:b1:seq:monotone}) for the first
point; comparison of partial sums for the second. Equivalents: for
large $n$, $\frac12 v_n \leq u_n \leq 2 v_n$ (definition of $\sim$
with $\varepsilon = \frac12$), and comparison applies both ways.
\end{proof}

\begin{example}[An equivalent that proves divergence]\label{ex:b1:series:equivdiverge}
Nature of $\sum_{n\geq1} n\sin\dfrac{1}{n^2}$? Since
$\frac{1}{n^2} \to 0$ and $\sin h \sim h$ at $0$:
\[
n\sin\frac{1}{n^2} \;\sim\; n\cdot\frac{1}{n^2} = \frac1n ,
\]
and the equivalents test transfers the divergence of the
harmonic series: divergent --- even though the terms tend to
$0$. One expansion, one scale, one verdict; the same two-step
pattern (equivalent, then Riemann or geometric lookup) decides
all four series of \cref{exo:b1:series:3}.
\end{example}

\begin{example}[The equivalents test in one line]\label{ex:b1:series:equivtest}
Nature of $\sum_{n \geq 1} \frac{\sqrt{n+1} - \sqrt n}{n}$?
Conjugate the numerator:
\[
\frac{\sqrt{n+1} - \sqrt n}{n}
= \frac{1}{n\,(\sqrt{n+1} + \sqrt n)}
\sim \frac{1}{2\,n^{3/2}} ,
\]
a convergent Riemann scale ($\alpha = \frac32 > 1$): the series
converges. The whole decision took one equivalent and one lookup
--- provided the terms are nonnegative, which they are. The
closing insight: for positive series, the entire convergence
theory is a \emph{dictionary of scales} ($n^{-\alpha}$, $q^n$,
$\frac{1}{n(\ln n)^\alpha}$) plus the license to replace a term
by an equivalent; the analytic work is in the asymptotics
(\cref{ch:b1:taylor}), never in the summation.
\end{example}

\begin{theorem}[Integral comparison; Riemann series]\label{thm:b1:series:riemann}
Let $f$ be continuous, nonnegative and \emph{decreasing} on
$\intco{1}{+\infty}$. Then
\[
\int_1^{N+1} f(t)\,\dd t \;\leq\; \sum_{n=1}^{N} f(n) \;\leq\; f(1)
+ \int_1^{N} f(t)\,\dd t ,
\]
so $\sum f(n)$ converges iff $\bigl(\int_1^x f\bigr)$ is bounded. In
particular, for $\alpha \in \R$:\index{Riemann series}
\[
\sum_{n \geq 1} \frac{1}{n^\alpha} \text{ converges}
\iff \alpha > 1,
\]
and $\sum_{n=1}^{N} \frac1n = \ln N + O(1)$.
\end{theorem}

\begin{proof}
For $n \leq t \leq n+1$, monotonicity gives $f(n+1) \leq f(t)
\leq f(n)$; integrating over $\intcc{n}{n+1}$ (a segment of
length $1$):
\[
f(n+1) \;\leq\; \int_n^{n+1} f(t)\,\dd t \;\leq\; f(n) .
\]
Summing the right inequalities for $n = 1, \dots, N-1$ gives
$\int_1^{N} f \leq \sum_{n=1}^{N-1} f(n)$, hence the upper
framing after adding $f(N) \leq f(1)$; summing the left ones for
$n = 1, \dots, N$ gives $\sum_{n=2}^{N+1} f(n) \leq
\int_1^{N+1} f$, which after reindexing is the lower framing.
Convergence: the partial sums and the integrals $\int_1^x f$
bound each other within the constant $f(1)$, and both are
nondecreasing, so one is bounded iff the other is
(\cref{thm:b1:series:positive}). For $f(t)
= t^{-\alpha}$ ($\alpha \neq 1$): $\int_1^x t^{-\alpha}\dd t =
\frac{x^{1-\alpha} - 1}{1 - \alpha}$, bounded iff $\alpha > 1$; for
$\alpha = 1$ the integral is $\ln x \to \infty$, and the framing
gives $\ln(N+1) \leq H_N \leq 1 + \ln N$. For $\alpha \leq 0$ the
terms do not tend to $0$.
\end{proof}

\begin{example}[The harmonic stack]\label{ex:b1:series:harmonicstack}
How many terms must the harmonic series accumulate to pass $20$?
The framing $\ln(N+1) \leq H_N \leq 1 + \ln N$ answers with no
summation at all: $H_N \geq 20$ requires $1 + \ln N \geq 20$,
i.e.\ $N \geq \eu^{19} \approx 1.8\cdot10^{8}$, and is guaranteed
once $\ln(N + 1) \geq 20$, i.e.\ $N \approx \eu^{20} \approx
4.9\cdot10^{8}$. (The weekend problem sharpens this to $N \approx
\eu^{20 - \gamma} \approx 2.7\cdot10^{8}$ via Euler's constant.)
The closing insight: the integral comparison does not merely
decide convergence --- it \emph{locates} partial sums with
logarithmic precision, turning a hopeless computation
(hundreds of millions of terms) into a two-line estimate.
\end{example}

\begin{theorem}[Ratio test (d'Alembert)]\label{thm:b1:series:ratio}
Let $u_n > 0$ with $\frac{u_{n+1}}{u_n} \to \ell$.\index{ratio test}
\begin{itemize}
  \item If $\ell < 1$: $\sum u_n$ converges;
  \item if $\ell > 1$: $u_n \to +\infty$, divergence;
  \item if $\ell = 1$: no conclusion ($\sum \frac1n$ diverges, $\sum
        \frac{1}{n^2}$ converges).
\end{itemize}
\end{theorem}

\begin{proof}
If $\ell < 1$, fix $q \in \intoo{\ell}{1}$: beyond some $N$,
$u_{n+1} \leq q\,u_n$, so $u_n \leq u_N q^{\,n-N}$ by induction:
comparison with a geometric series. If $\ell > 1$: beyond some $N$
the sequence $(u_n)$ is increasing, so it cannot tend to $0$
(its limit, if any, is $\geq u_N > 0$); by
\cref{prop:b1:series:first}~(1), divergence --- and in fact $u_n
\geq u_N q^{n-N}$ with $q > 1$ gives $u_n \to \infty$.
\end{proof}

\begin{example}\label{ex:b1:series:ratio}
$\sum \frac{x^n}{n!}$ converges for every $x > 0$: ratio
$\frac{x}{n+1} \to 0$. Its sum is $\eu^x$: by Taylor--Lagrange
(\cref{thm:b1:taylor:lagrange}) on $\intcc{0}{x}$,
\[
\Bigl| \eu^x - \sum_{k=0}^{n} \frac{x^k}{k!} \Bigr|
\leq \eu^{x}\, \frac{x^{n+1}}{(n+1)!} \xrightarrow[n\to\infty]{} 0 ,
\]
the bound tending to $0$ because the factorial dominates
(\cref{exo:b1:integration:9}~(1) used the same fact). The same
argument sums the $\sin$, $\cos$, $\sinh$, $\cosh$ series on all of
$\R$.
\end{example}

\begin{example}[The ratio test is sufficient, not necessary]\label{ex:b1:series:rationotnecessary}
Let $u_n = 2^{-n}$ for even $n$ and $u_n = 2^{-n-2}$ for odd
$n$. The consecutive ratios oscillate between $\frac{1}{8}$ and
$\frac12\cdot4 = 2$, so $\frac{u_{n+1}}{u_n}$ has no limit and
d'Alembert is mute --- yet $u_n \leq 2^{-n}$ and the comparison
test settles convergence instantly. The test's hypothesis (the
ratio \emph{converges}) is a real restriction: it suits terms
with one dominant multiplicative structure (factorials, powers),
and fails on anything that breathes. When ratios misbehave,
step back to comparison against a geometric envelope --- which
is all the ratio test ever was, as its proof shows.
\end{example}

\begin{example}[Ratio test on factorial battles]\label{ex:b1:series:ratiobinom}
Nature of $\sum_{n\geq0} \dfrac{(n!)^2}{(2n)!}$ (reciprocals of
the central binomial coefficients, up to the factor $n + 1$)?
The ratio collapses the factorials:
\[
\frac{u_{n+1}}{u_n}
= \frac{((n+1)!)^2}{(n!)^2}\cdot\frac{(2n)!}{(2n+2)!}
= \frac{(n+1)^2}{(2n+1)(2n+2)}
\longrightarrow \frac14 < 1 :
\]
convergent, with room to spare --- the terms decay essentially
like $4^{-n}$, consistent with $\binom{2n}{n} \geq
\frac{4^n}{2n+1}$ from \cref{pb:b1:integration:1}. The closing
insight: quotients of factorials are exactly what the ratio test
digests --- every factorial cancels into a rational function of
$n$, whose limit is read off the leading terms.
\end{example}

\section{Absolute convergence; alternating series}

\begin{theorem}[Absolute convergence]\label{thm:b1:series:absolute}
If $\sum \abs{u_n}$ converges (\emph{absolute
convergence}\index{absolute convergence}), then $\sum u_n$ converges,
and $\bigl|\sum u_n\bigr| \leq \sum \abs{u_n}$. This holds for real
or complex terms.
\end{theorem}

\begin{proof}
The partial sums satisfy, for $M > N$ (Cauchy criterion,
\cref{thm:b1:seq:complete}):
\[
\abs{S_M - S_N} = \Bigl| \sum_{n=N+1}^{M} u_n \Bigr|
\leq \sum_{n=N+1}^{M} \abs{u_n},
\]
which is small for large $N$ since the partial sums of $\sum\abs{u_n}$
form a Cauchy sequence. So $(S_N)$ is Cauchy, hence convergent. The
inequality passes to the limit from the finite triangle inequality.
\end{proof}

\begin{example}[Absolute convergence, real and complex]\label{ex:b1:series:absexamples}
$\sum_{n\geq1} \frac{\sin n}{n^2}$: the terms change sign
erratically (indeed $(\sin n)$ is dense in $\intcc{-1}{1}$,
\cref{exo:b1:seq:12}), and no alternating structure is in
sight. Absolute convergence rescues everything at once:
$\bigl|\frac{\sin n}{n^2}\bigr| \leq \frac{1}{n^2}$, a
convergent scale, so the series converges. The same shield works
over $\C$: $\sum_{n\geq1}\frac{\eu^{\iu n}}{n^2}$ converges
because $\bigl|\frac{\eu^{\iu n}}{n^2}\bigr| = \frac{1}{n^2}$
--- sign patterns, even two-dimensional ones, are irrelevant
once the moduli are summable. The closing insight: absolute
convergence is the only tool of this chapter that never asks how
the signs are organized; try it first
(\cref{met:b1:series:decide}), and reserve the delicate tests
for the series that fail it.
\end{example}

\begin{theorem}[Alternating series test]\label{thm:b1:series:alternating}
Let $(a_n)$ be decreasing with $a_n \to 0$. Then the alternating
series $\sum (-1)^n a_n$ converges; its sum lies between any two
consecutive partial sums, and\index{alternating series}
\[
\abs{R_N} = \Bigl| \sum_{n > N} (-1)^n a_n \Bigr| \leq a_{N+1} .
\]
\end{theorem}

\begin{proof}
The even and odd partial sums are adjacent: $S_{2p+2} - S_{2p} =
a_{2p+2} - a_{2p+1} \leq 0$ (decreasing), $S_{2p+1} - S_{2p-1} =
a_{2p} - a_{2p+1} \geq 0$ (increasing), and $S_{2p} - S_{2p+1} =
a_{2p+1} \to 0$. By \cref{thm:b1:seq:adjacent} they share a limit
$S$, which the two subsequences criterion
(\cref{prop:b1:seq:subsequences}) makes the limit of $(S_N)$;
moreover $S$ is trapped between consecutive partial sums, and
$\abs{S - S_N}$ is at most the gap to the next one, $a_{N+1}$.
\end{proof}

\begin{example}[Alternating harmonic series]\label{ex:b1:series:ln2}
$\sum_{n \geq 1} \frac{(-1)^{n-1}}{n}$ converges (alternating test)
but not absolutely (harmonic series). Its sum is $\ln 2$: from the
finite geometric identity $\frac{1}{1+t} = \sum_{k=0}^{n-1} (-t)^k +
\frac{(-t)^n}{1+t}$, integrate over $\intcc{0}{1}$:
\[
\ln 2 = \sum_{k=1}^{n} \frac{(-1)^{k-1}}{k}
+ (-1)^n \int_0^1 \frac{t^n}{1+t}\,\dd t ,
\qquad
0 \leq \int_0^1 \frac{t^n}{1+t}\,\dd t \leq \frac{1}{n+1} \to 0 .
\]
The convergence is painfully slow ($R_N \approx \frac{1}{N}$) ---
alternating series converge by cancellation, not by smallness.
\end{example}

\begin{omfigure}
\begin{tikzpicture}[x=0.78cm, y=7.2cm]
  \draw[-Stealth] (0.3,0.42) -- (11,0.42)
    node[below, font=\small] {$N$};
  \draw (1,0.415) -- (1,0.425) node[below=1pt, font=\small]
    {$1$};
  \draw (5,0.415) -- (5,0.425) node[below=1pt, font=\small]
    {$5$};
  \draw (10,0.415) -- (10,0.425) node[below=1pt, font=\small]
    {$10$};
  \draw[gray, dashed] (0.3,0.6931) -- (10.7,0.6931)
    node[right, font=\small, text=black] {$\ln 2$};
  \node[left, font=\small] at (0.3,1) {$1$};
  \node[left, font=\small] at (0.3,0.5) {$0.5$};
  \draw[omDef, thick]
    plot coordinates {(1,1) (2,0.5) (3,0.8333) (4,0.5833)
    (5,0.7833) (6,0.6167) (7,0.7595) (8,0.6345)
    (9,0.7456) (10,0.6456)};
  \fill[omDef] (1,1) circle (1.5pt);
  \fill[omDef] (2,0.5) circle (1.5pt);
  \fill[omDef] (3,0.8333) circle (1.5pt);
  \fill[omDef] (4,0.5833) circle (1.5pt);
  \fill[omDef] (5,0.7833) circle (1.5pt);
  \fill[omDef] (6,0.6167) circle (1.5pt);
  \fill[omDef] (7,0.7595) circle (1.5pt);
  \fill[omDef] (8,0.6345) circle (1.5pt);
  \fill[omDef] (9,0.7456) circle (1.5pt);
  \fill[omDef] (10,0.6456) circle (1.5pt);
\end{tikzpicture}

{\small The partial sums $S_N$ of the alternating harmonic
series $1 - \frac12 + \frac13 - \cdots$ hop over their limit
$\ln 2$ at every step: odd sums from above, even sums from
below, each hop of size $\frac{1}{N+1}$. The bracketing is the
proof of \cref{thm:b1:series:alternating} made visible --- and
the slow closing of the pincers ($\abs{S_N - \ln 2} \approx
\frac{1}{2N}$, weekend problem \cref{pb:b1:series:1}) is why
nobody computes $\ln 2$ this way.}
\end{omfigure}

\begin{remark}[Common pitfalls with series]
(i) \emph{The equivalents test needs a sign}: let $v_n =
\frac{(-1)^n}{\sqrt n}$ and $u_n = v_n + \frac1n$. Then
$\frac{u_n}{v_n} = 1 + \frac{(-1)^n}{\sqrt n} \to 1$, so $u_n
\sim v_n$; yet $\sum v_n$ converges (alternating test) while
$\sum u_n = \sum v_n + \sum \frac1n$ diverges. Equivalence
controls the \emph{size} of terms, and for signed series size
is not destiny --- the test is stated, and true, for
(eventually) nonnegative terms only. (ii) \emph{$u_n \to 0$
proves nothing}: the harmonic series is the eternal
counterexample; the converse direction
(\cref{prop:b1:series:first}~(1)) is only a quick divergence
test. (iii) \emph{Ratio limit $1$ is silence, not
convergence}: both $\sum\frac1n$ and $\sum\frac{1}{n^2}$ have
ratio $\to 1$; switch to Riemann scales or integral comparison.
(iv) \emph{Alternating needs decreasing}: $\sum
\frac{(-1)^n}{n + (-1)^n}$ looks alternating and is handled
only by expansion (\cref{exo:b1:series:5}); the weekend problem
of \cref{ch:b1:taylor} (question 23 there) shows the test can
fail outright without monotonicity. (v) \emph{Grouping and
reordering are not free}: inserting parentheses is harmless for
convergent series but can create convergence from divergence
($1 - 1 + 1 - \cdots$ grouped in pairs), and reordering can
change the sum itself --- the drama staged in this chapter's
weekend problem (\cref{pb:b1:series:1}).
\end{remark}

\begin{method}[Deciding the nature of a series]\label{met:b1:series:decide}
\begin{enumerate}
  \item Does $u_n \to 0$? If not, divergence, stop.
  \item Nonnegative terms: seek an equivalent of $u_n$ (expansions,
        \cref{ch:b1:taylor}!), compare with Riemann or geometric
        scales; factorials and powers call for the ratio test;
        decreasing $f(n)$ calls for integral comparison.
  \item Signs vary: try absolute convergence first; if it fails, the
        alternating test (check \emph{decreasing} carefully); beyond
        that, second-year tools.
\end{enumerate}
\end{method}

\begin{example}[Odd denominators, half the telescope]\label{ex:b1:series:oddtelescope}
Compute $\sum_{n \geq 1} \dfrac{1}{4n^2 - 1}$. Partial
fractions: $\frac{1}{(2n-1)(2n+1)} =
\frac12\bigl(\frac{1}{2n-1} - \frac{1}{2n+1}\bigr)$, so
\[
\sum_{n=1}^{N} \frac{1}{4n^2 - 1}
= \frac12\Bigl(1 - \frac{1}{2N+1}\Bigr)
\longrightarrow \frac12 .
\]
Compare with $\sum \frac{1}{n(n+1)} = 1$
(\cref{exo:b1:series:1}): same telescoping skeleton, but the
consecutive terms here are two apart in the odd numbers, and
the factor $\frac12$ records the step. The closing insight:
telescoping is a change of viewpoint, not a trick --- whenever
the general term is a difference $w_n - w_{n+1}$ of a sequence
with a limit, the sum is $w_1 - \lim w$, exactly
\cref{prop:b1:series:first}~(3).
\end{example}

\begin{remark}[The analysis pipeline, in retrospect]
This chapter is where the volume's analysis converges, and each
test names its ancestor. Bounded partial sums is the monotone
limit theorem (\cref{ch:b1:seq}), itself the completeness axiom
of \cref{ch:b1:reals}; absolute convergence is the Cauchy
criterion; the integral test is \cref{ch:b1:integration}'s
framing of areas; equivalents of general terms are
\cref{ch:b1:taylor}'s expansions; and the alternating theorem is
the adjacent-sequences lemma in its Sunday clothes. Read
backwards, the pipeline explains what each chapter was
\emph{for} --- and the weekend problems threaded through it
($b$-adic digits, Ces\`aro--Stolz, the irrationality machines,
Euler's constant) are the same few ideas meeting at higher and
higher altitude. The linear algebra that follows changes
subject, not standards: the habit of exact statements with
certified error survives the move from limits to dimensions.
\end{remark}

\begin{remark}[Where series go next]
This chapter closes the analysis of the volume and opens three
doors. In the Year 2 volume, series acquire a variable ($\sum
a_n x^n$: power series, with their radius of convergence) and
then a function-valued theory (Fourier series); the dichotomy
absolute-versus-conditional convergence, dramatized in the
weekend problem below, becomes the cornerstone of both. In
probability (Year 3 volume), expectations of discrete random
variables \emph{are} series, and absolute convergence is what
makes them well defined. And the Riemann series $\sum n^{-s}$,
pushed to complex $s$, becomes the zeta function --- the single
most studied series in mathematics.
\end{remark}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:b1:series:1}
Nature (and sum, when telescoping) of:
\[
\sum_{n\geq1} \frac{1}{n(n+1)},
\qquad
\sum_{n\geq2} \ln\Bigl(1 - \frac{1}{n^2}\Bigr),
\qquad
\sum_{n\geq0} \frac{3^n + 4^n}{5^n} .
\]
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b1:series:2}
Nature of: $\;\sum \dfrac{n^2}{2^n}$; $\;\sum \dfrac{n!}{n^n}$;
$\;\sum \dfrac{2^n\,n!}{n^n}$; $\;\sum \dfrac{3^n\,n!}{n^n}$.
\emph{(Ratio test; recall $\bigl(1 + \frac1n\bigr)^n \to \eu$.)}
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b1:series:3}
Nature of: $\;\sum \sin\dfrac{1}{n^2}$; $\;\sum
\Bigl(1 - \cos\dfrac1n\Bigr)$; $\;\sum \dfrac{1}{\sqrt{n(n+1)}}$;
$\;\sum \dfrac{\ln n}{n^2}$ \emph{(compare with $n^{-3/2}$)}.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b1:series:4}
Prove that $\sum_{n\geq1} \frac{1}{n^2}$ converges with sum $\leq 2$,
using $\frac{1}{n^2} \leq \frac{1}{n(n-1)}$ for $n \geq 2$ and a
telescoping bound.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:series:5}
Nature of $\;\sum \dfrac{(-1)^n}{\sqrt n}$, of $\;\sum
\dfrac{(-1)^n}{n + (-1)^n}$ \emph{(expand: the alternating test does
not apply directly --- why?)}, and of $\;\sum
\sin\bigl(\pi\sqrt{n^2+1}\,\bigr)$ \emph{(reduce modulo $\pi$:
$\sqrt{n^2+1} = n + \frac{1}{2n} + O(n^{-3})$)}.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:series:6}
For which $\alpha > 0$ does $\sum_{n \geq 2}
\dfrac{1}{n (\ln n)^{\alpha}}$ converge? \emph{(Integral comparison;
substitute $u = \ln t$.)}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:series:7}
Let $u_n = \dfrac{1}{n} - \ln\Bigl(1 + \dfrac1n\Bigr)$. Prove that
$0 \leq u_n \leq \dfrac{1}{2n^2}$, that $\sum u_n$ converges, and
deduce the existence of Euler's constant\index{Euler's constant}:
\[
\gamma = \lim_{N \to \infty} \Bigl( \sum_{n=1}^{N} \frac 1n - \ln N
\Bigr) .
\]
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:series:8}
Compute the sums
\[
\sum_{n=1}^{\infty} \frac{1}{n(n+2)}
\qquad\text{and}\qquad
\sum_{n=0}^{\infty} \frac{n}{2^n} .
\]
\emph{(For the first: partial fractions. For the second: compute
$\sum_{n=1}^{N} n x^{n-1}$ in closed form and let $N \to \infty$ at
$x = \frac12$.)}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b1:series:9}
(Cauchy condensation) Let $(u_n)$ be nonnegative and decreasing.
Prove that
\[
\sum_{n \geq 1} u_n \text{ converges}
\iff
\sum_{k \geq 0} 2^k\, u_{2^k} \text{ converges},
\]
by comparing packets of terms between consecutive powers of $2$.
Recover from it the Riemann criterion and
\cref{exo:b1:series:6}.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b1:series:10}
Using the integral identity of \cref{ex:b1:series:ln2} adapted to
$\frac{1}{1+t^2}$, prove Leibniz's formula
\[
\frac{\pi}{4} = \sum_{n=0}^{\infty} \frac{(-1)^n}{2n+1}
= 1 - \frac13 + \frac15 - \frac17 + \cdots
\]
with the error bound $\abs{R_N} \leq \frac{1}{2N+3}$.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b1:series:11}
Nature of $\displaystyle\sum_{n \geq 1} \frac{1}{n^{1 + 1/n}}$.
\emph{(Compute the limit of $n^{1/n}$ and find an equivalent of
the general term: the Riemann test needs a \emph{fixed}
exponent.)}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b1:series:12}
Let $(u_n)$ be nonnegative and \emph{decreasing} with $\sum u_n$
convergent. Prove that $n\,u_n \to 0$ \emph{(bound $n\,u_{2n}$
by a slice $\sum_{k=n+1}^{2n} u_k$ and use the Cauchy
criterion)}. Show that the converse fails, and that the
monotonicity hypothesis cannot be removed.
\end{exercise}

\section{Problem: Euler's constant and the series that changes
its sum}

\begin{problem}[{Weekend problem --- $H_n = \ln n + \gamma +
\frac{1}{2n} + O(n^{-2})$, and rearranging $1 - \frac12 +
\frac13 - \dots$ to $\frac{\ln 2}{2}$}]
\label{pb:b1:series:1}
Two stories share the harmonic series. First, the exact
bookkeeping of its divergence: $H_n - \ln n$ converges to Euler's
constant $\gamma$ (\cref{exo:b1:series:7}), and this problem
sharpens the statement into a two-sided law $\frac{1}{2(n+1)}
\leq H_n - \ln n - \gamma \leq \frac{1}{2n}$, certifying $\gamma
= 0.5772\dots$ by hand. Second, the scandal of conditional
convergence: the alternating harmonic series sums to $\ln 2$
(\cref{ex:b1:series:ln2}), yet \emph{the same terms, in a
different order}, sum to $\frac{\ln 2}{2}$ --- or to $\ln 2 +
\frac12\ln\frac pq$ for any $p, q$, or to any real whatsoever
(Riemann). The two stories are one: the rearranged sums are
computed \emph{with} the $\gamma$-law.\index{Euler's
constant}\index{rearrangement of series}

\medskip
\noindent\textbf{Part I --- $\gamma$, bracketed.} Set $a_n = H_n
- \ln n$ and $b_n = H_n - \ln(n+1)$.
\begin{enumerate}
  \item Using $\frac{t}{1+t} \leq \ln(1 + t) \leq t$, show that
        $(a_n)$ decreases, $(b_n)$ increases, and that they are
        adjacent; their common limit is $\gamma$, with $b_n \leq
        \gamma \leq a_n$ for every $n$.
  \item Numerical first shot: from $H_{10} = 2.928968\dots$,
        bracket $\gamma$ between $b_{10} = 0.5311$ and $a_{10} =
        0.6264$. How large an $n$ would this crude bracket need
        for four decimals?
  \item Show the exact tail representation $a_n - \gamma =
        \sum_{k \geq n} w_k$ (limit of partial sums), where
        \[
        w_k = a_k - a_{k+1}
        = \ln\Bigl(1 + \frac1k\Bigr) - \frac{1}{k+1}
        = \int_k^{k+1} \frac{(k + 1 - t)}{t\,(k+1)}\,\dd t ,
        \]
        and deduce from the integral form the two-sided bound
        $\dfrac{1}{2(k+1)^2} \leq w_k \leq \dfrac{1}{2k(k+1)}$.
\end{enumerate}

\medskip
\noindent\textbf{Part II --- The $\frac{1}{2n}$ law.}
\begin{enumerate}[resume]
  \item Sum the bounds of question 3 (both sides telescope or
        compare to telescopes) and conclude the law:
        \[
        \frac{1}{2(n+1)} \;\leq\; H_n - \ln n - \gamma \;\leq\;
        \frac{1}{2n} \qquad (n \geq 1).
        \]
  \item Deduce $H_n = \ln n + \gamma + \frac{1}{2n} +
        O\bigl(\frac{1}{n^2}\bigr)$; precisely, show that
        $\gamma_n = H_n - \ln n - \frac{1}{2n}$ satisfies
        $-\frac{1}{2n(n+1)} \leq \gamma_n - \gamma \leq 0$.
  \item Certify four decimals with $n = 100$: given $H_{100} =
        5.1873775\dots$, compute $\gamma_{100} = 0.577207\dots$
        and conclude $\gamma = 0.5772 \pm 5\cdot10^{-5}$ (true
        value $0.5772156\dots$).
  \item Two dividends of the law, both needed later: as $m \to
        \infty$,
        \[
        H_{2m} - H_m = \ln 2 - \frac{1}{4m} +
        O\Bigl(\frac{1}{m^2}\Bigr),
        \qquad
        \sum_{j=1}^{m} \frac{1}{2j-1} = \frac{\ln m}{2} + \ln 2
        + \frac\gamma2 + o(1) ,
        \]
        the second via $\sum_{j \leq m} \frac{1}{2j-1} = H_{2m}
        - \frac12 H_m$, and likewise $\sum_{j=1}^{m}
        \frac{1}{2j} = \frac{\ln m}{2} + \frac\gamma2 + o(1)$.
\end{enumerate}

\medskip
\noindent\textbf{Part III --- The alternating harmonic series,
to second order.}
\begin{enumerate}[resume]
  \item Show (induction, or grouping) the identity
        $\sum_{k=1}^{2m} \frac{(-1)^{k-1}}{k} = H_{2m} - H_m$,
        and deduce both the sum $\ln 2$ (again) and the exact
        speed:
        \[
        \sum_{k=1}^{2m} \frac{(-1)^{k-1}}{k}
        = \ln 2 - \frac{1}{4m} + O\Bigl(\frac{1}{m^2}\Bigr) .
        \]
  \item Deduce the asymptotic error of the alternating harmonic
        series at \emph{any} index: $S - S_N \sim
        \frac{(-1)^N}{2N}$ --- twice smaller than the worst-case
        bound $a_{N+1} \approx \frac1N$ of
        \cref{thm:b1:series:alternating}.
  \item (Acceleration for free) Show that the averaged sums
        $\tilde S_N = \frac{S_N + S_{N+1}}{2}$ satisfy $\tilde
        S_N = \ln 2 + O\bigl(\frac{1}{N^2}\bigr)$. Check: $S_{10}
        = 0.64563$, $S_{11} = 0.73654$, $\tilde S_{10} =
        0.69109$, against $\ln 2 = 0.69315$: one average buys
        two decimal places.
  \item Explain in two sentences why no such trick can help a
        \emph{positive} divergent-tail phenomenon like question
        2's bracket: the alternating error oscillates (sign
        $(-1)^N$), so averaging cancels its leading term, while
        the $\gamma$-bracket error $\frac{1}{2n}$ has constant
        sign. (Averaging $a_n$ and $b_n$ \emph{does} help:
        relate $\frac{a_n + b_n}{2}$ to the midpoint estimate
        $H_n - \ln\bigl(n + \frac12\bigr)$ and show its error is
        $O\bigl(\frac{1}{n^2}\bigr)$.)
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- Rigidity and its failure.}
\begin{enumerate}[resume]
  \item Show that the positive part $\sum \frac{1}{2j-1}$ and
        the negative part $\sum \frac{1}{2j}$ of the alternating
        harmonic series both diverge --- the signature of
        \emph{conditional} convergence.
  \item Prove the general statement behind question 12: if
        $\sum u_n$ converges but $\sum \abs{u_n}$ diverges, then
        the series of positive parts $\sum u_n^+$ and of
        negative parts $\sum u_n^-$ both diverge \emph{(from
        $u_n^\pm = \frac{\abs{u_n} \pm u_n}{2}$: if one
        converged, so would the other, hence $\sum\abs{u_n}$)}.
        This inexhaustible reservoir of positive and negative
        mass is what Riemann's recipe will spend.
  \item (Rigidity) Prove: if $\sum u_n$ converges
        \emph{absolutely} and $\sigma \colon \N \to \N$ is a
        bijection, then $\sum u_{\sigma(n)}$ converges to the
        same sum \emph{(for $N$ large the first $M$ rearranged
        terms contain $u_0, \dots, u_N$; compare partial sums
        through the tail $\sum_{n > N}\abs{u_n}$)}.
  \item (Riemann's recipe) Let $t \in \R$. Describe the greedy
        rearrangement of the alternating harmonic series: take
        positive terms $1, \frac13, \frac15, \dots$ until the
        partial sum first exceeds $t$, then negative terms
        $-\frac12, -\frac14, \dots$ until it first drops below
        $t$, and repeat. Show that every term is used exactly
        once, that after the first crossing the partial sums
        stay within the last used term of $t$, and conclude that
        the rearranged series converges to $t$: \emph{any}
        prescribed sum is attainable.
\end{enumerate}

\medskip
\noindent\textbf{Part V --- The $(p, q)$ formula.} Fix integers
$p, q \geq 1$. Rearrange the alternating harmonic series in
blocks: $p$ positive terms (the next odd reciprocals), then $q$
negative terms (the next even reciprocals), and repeat.
\begin{enumerate}[resume]
  \item Check that this is a genuine rearrangement (every term
        exactly once), and that for $(p, q) = (1, 2)$ it reads
        \[
        1 - \frac12 - \frac14 + \frac13 - \frac16 - \frac18 +
        \frac15 - \frac1{10} - \frac1{12} + \dots
        \]
  \item (The exact halving) For $(p, q) = (1, 2)$, prove the
        block identity
        \[
        \frac{1}{2k-1} - \frac{1}{4k-2} - \frac{1}{4k}
        = \frac12\Bigl(\frac{1}{2k-1} - \frac{1}{2k}\Bigr),
        \]
        and deduce the \emph{exact} relation $T_{3K} = \frac12
        S_{2K}$ between the rearranged partial sums and the
        original ones: the halving of the sum is visible at
        every finite stage, not only in the limit.
  \item Show that the partial sum after $K$ complete blocks
        equals $\sum_{j=1}^{pK} \frac{1}{2j-1} -
        \sum_{j=1}^{qK} \frac{1}{2j}$, and compute its limit
        with question 7:
        \[
        \ln 2 + \frac12 \ln\frac pq .
        \]
  \item Control the partial sums \emph{inside} a block (the
        terms tend to $0$) and conclude that the $(p,
        q)$-rearranged series converges to $\ln 2 + \frac12\ln
        \frac pq$. In particular $(1, 2)$ gives $\frac{\ln
        2}{2}$: verify against the first nine terms, $T_9 =
        0.3083$, creeping toward $0.3466$.
  \item Sanity checks and range: $(1,1)$ recovers $\ln 2$;
        $(2,1)$ gives $\frac32\ln 2$; which sums are reachable
        by $(p, q)$-blocks, and how does this countable menu
        compare with Riemann's full carte (question 14)?
\end{enumerate}

\medskip
\noindent\textbf{Part VI --- Epilogue: $\gamma$ at work, and
synthesis.}
\begin{enumerate}[resume]
  \item Identify the sum of the convergent series
        $\sum_{k\geq1} \bigl(\frac1k - \ln\frac{k+1}{k}\bigr)$
        (\cref{exo:b1:series:7}): show it equals $\gamma$.
  \item Run Riemann's recipe (question 14) for the target $t =
        1$ and list the first twelve terms produced ($1, \frac13,
        -\frac12, \frac15, -\frac14, \frac17, \frac19, -\frac16,
        \frac1{11}, \frac1{13}, -\frac18, \frac1{15}$),
        computing the partial sum ($\approx 0.980$) --- watch
        the algorithm breathe around its target.
  \item Show that some rearrangement of the alternating harmonic
        series diverges to $+\infty$ \emph{(blocks of positive
        terms long enough to gain $1$ each time, using question
        12, separated by single negative terms)}.
  \item Sharpen \cref{ex:b1:series:harmonicstack} with the
        $\gamma$-law: show that the first index with $H_N \geq
        20$ satisfies $N = \eu^{\,20 - \gamma}\,(1 + o(1))
        \approx 2.7\cdot10^{8}$ --- Euler's constant is exactly
        the correction the crude framing was missing.
  \item Synthesis, one sentence each: (i) the $\gamma$-law and
        what each of its three pieces ($\ln n$, $\gamma$,
        $\frac{1}{2n}$) contributes; (ii) why conditional
        convergence makes the sum order-dependent while absolute
        convergence forbids it; (iii) how the $(p,q)$ formula
        was a \emph{computation} with the $\gamma$-law rather
        than an abstract existence claim; (iv) where these
        threads continue --- power series and products of series
        in the Year 2 volume, and the Year 3 volume's weekend
        problem on Stirling's formula, where the same
        sum-versus-integral bookkeeping runs at full power.
\end{enumerate}
\end{problem}
