\chapter{Preface}

This book is an open-source project with a simple ambition: a single,
coherent mathematics book covering the entire French curriculum, from
kindergarten to the end of the \emph{licence} (bachelor's degree), written in
English.

The style is that of the French \emph{classes préparatoires}: concise
courses built from definitions, examples, propositions and theorems, with
proofs whenever they are accessible at the given level, followed by carefully
chosen exercises. Results stated without proof are explicitly marked as
admitted. Full solutions to all exercises are collected at the end of the
book.

Each school year is a part of the book, and each part is split into
chapters. The current state of the project covers the \emph{Terminale} year
(grade 12), taking the union of the former \emph{Terminale scientifique}
program and the current \emph{Spécialité Mathématiques} and \emph{Maths
Expertes} programs.

\section*{How to read this book}

Statements are numbered within each chapter and color-coded: definitions in
blue, theorems in red, propositions and their relatives in orange, and
methods --- short recipes for standard tasks --- in green. Exercises are
graded from ($\star$) for direct applications to ($\star\star\star$) for
harder problems.

\section*{Contributing}

The source of this book lives in a public git repository. Contributions ---
new chapters, corrections, better proofs, additional exercises --- are
welcome; see \texttt{CONTRIBUTING.md} at the root of the repository.
