\chapter{The Fourier Transform}\label{ch:b3:fouriertransform}

Fourier series decompose periodic signals into discrete
harmonics; the Fourier \emph{transform} does the same for
signals on the whole line, with a continuum of frequencies. It
converts differentiation into multiplication, convolution into
products, and Gaussians into Gaussians --- the reasons it
solves differential equations, drives signal processing, and
will prove the central limit theorem in \cref{ch:b3:clt}. This
chapter develops the $L^1$ theory (Riemann--Lebesgue,
inversion, injectivity), the Schwartz class where the
transform is a perfect bijection, and the $L^2$ theory
(Plancherel: the transform is, up to a constant, a unitary
operator), with two showpiece applications: the heat equation,
solved end to end in the weekend problem, and the Poisson
summation formula. Convention:
\[
\hat f(\xi) = \int_\R f(x)\,\eu^{-\iu\xi x}\,\dd x .
\]

\section{The transform on \texorpdfstring{$L^1$}{L1}}

\begin{proposition}\label{prop:b3:fouriertransform:basic}
For $f \in L^1(\R)$: $\hat f$ is well defined, bounded
($\norm{\hat f}_\infty \leq \norm f_1$), continuous, and:
\begin{enumerate}
  \item $\widehat{\tau_af}(\xi) = \eu^{-\iu a\xi}\hat f(\xi)$
        and $\widehat{\eu^{\iu ax}f}(\xi) = \hat f(\xi - a)$;
  \item $\widehat{f(\cdot/\lambda)}(\xi) = \lambda\hat
        f(\lambda\xi)$ for $\lambda > 0$;
  \item if $xf \in L^1$, then $\hat f$ is $\mathcal C^1$ with
        $(\hat f)'(\xi) = \widehat{(-\iu x)f}(\xi)$;
  \item if $f \in \mathcal C^1$ with $f' \in L^1$ (and $f \to
        0$ at $\pm\infty$, automatic here), then
        $\widehat{f'}(\xi) = \iu\xi\hat f(\xi)$;
  \item $\widehat{f * g} = \hat f\,\hat g$ for $f, g \in L^1$.
\end{enumerate}
\end{proposition}

\begin{proof}
Boundedness: $\abs{\hat f} \leq \int\abs f$. Continuity: DCT
with dominator $\abs f$ (\cref{thm:b3:lebesgue:paramcont}).
(1), (2): substitutions
(\cref{thm:b3:product:linearchange}). (3): differentiation
under the integral, dominator $\abs{xf}$
(\cref{thm:b3:lebesgue:paramdiff}). (4): first, $f(x) = f(0)
+ \int_0^xf'$ has a limit at $\pm\infty$ ($f' \in L^1$), which
must be $0$ ($f \in L^1$); then integrate by parts on $[-A,
A]$ and let $A \to \infty$. (5): Fubini, legitimate since
$(x,y)\mapsto f(x - y)g(y)\eu^{-\iu\xi x}$ is absolutely
integrable (\cref{thm:b3:product:convolution}):
\[
\widehat{f*g}(\xi) = \iint f(x - y)g(y)\eu^{-\iu\xi(x - y)}
\eu^{-\iu\xi y}\dd x\,\dd y = \hat f(\xi)\,\hat g(\xi).
\]
\end{proof}

\begin{example}\label{ex:b3:fouriertransform:gaussian}
The Gaussian: for $a > 0$,\index{Gaussian, Fourier transform}
\[
\widehat{\eu^{-ax^2}}(\xi) =
\sqrt{\frac\pi a}\;\eu^{-\xi^2/4a} :
\]
by \cref{exo:b3:lebesgue:7} (the ODE $F' = -\frac\xi{2}F$
trick, rescaled), or by (3): $g = \widehat{\eu^{-ax^2}}$
satisfies $g'(\xi) = -\frac{\xi}{2a}g(\xi)$ (integrate by
parts), $g(0) = \sqrt{\pi/a}$. Gaussians are fixed points of
the transform up to scaling --- the deep reason they rule the
central limit theorem.
\end{example}

\begin{theorem}[Riemann--Lebesgue]
\label{thm:b3:fouriertransform:riemannlebesgue}
For $f \in L^1(\R)$: $\hat f(\xi) \to 0$ as $\abs\xi \to
\infty$. Thus $\widehat{\phantom f} \colon L^1 \to \mathcal
C_0(\R)$ (continuous functions vanishing at
infinity).\index{Riemann--Lebesgue lemma}
\end{theorem}

\begin{proof}
For an indicator of an interval, $\widehat{\mathbf 1_{\intcc
ab}}(\xi) = \frac{\eu^{-\iu a\xi} - \eu^{-\iu b\xi}}{\iu\xi}
\to 0$; hence for step functions. Step functions are dense in
$L^1$ (\cref{thm:b3:lp:density}(1) plus approximation of
finite-measure sets by finite unions of intervals,
\cref{exo:b3:measure:7}), and the transform is
$\norm\cdot_\infty$-$\norm\cdot_1$ continuous: for $\norm{f -
s}_1 < \varepsilon$, $\limsup_{\abs\xi\to\infty}\abs{\hat
f(\xi)} \leq \varepsilon$.
\end{proof}

\section{Inversion and injectivity}

\begin{lemma}[Multiplication formula]
\label{lem:b3:fouriertransform:multiplication}
For $f, g \in L^1(\R)$: $\displaystyle\int \hat f\,g =
\int f\,\hat g$.
\end{lemma}

\begin{proof}
Both sides equal $\iint f(x)g(\xi)\eu^{-\iu x\xi}\dd
x\,\dd\xi$ (Tonelli--Fubini: the double integral of the
absolute value is $\norm f_1\norm g_1$).
\end{proof}

\begin{theorem}[Inversion]\label{thm:b3:fouriertransform:inversion}
Let $f \in L^1(\R)$.\index{Fourier inversion theorem}
\begin{enumerate}
  \item (Gaussian summability) For every $x$,
        \[
        (f * g_\varepsilon)(x) =
        \frac1{2\pi}\int_\R \hat f(\xi)\,
        \eu^{-\varepsilon\xi^2}\,\eu^{\iu x\xi}\,\dd\xi,
        \qquad\text{where } g_\varepsilon(y) =
        \frac{1}{2\sqrt{\pi\varepsilon}}\,
        \eu^{-y^2/4\varepsilon},
        \]
        and $f * g_\varepsilon \to f$ in $L^1$ as $\varepsilon
        \to 0$.
  \item If moreover $\hat f \in L^1$, then for almost every
        $x$
        \[
        f(x) = \frac{1}{2\pi}\int_\R \hat
        f(\xi)\,\eu^{\iu x\xi}\,\dd\xi ,
        \]
        and $f$ has a continuous representative.
  \item (Injectivity) If $\hat f = 0$ then $f = 0$ a.e.
\end{enumerate}
\end{theorem}

\begin{proof}
(1) Fix $x$ and apply
\cref{lem:b3:fouriertransform:multiplication} to $f$ and
$g(\xi) = \frac1{2\pi}\eu^{-\varepsilon\xi^2}\eu^{\iu x\xi}$:
by \cref{ex:b3:fouriertransform:gaussian} (with the modulation
rule),
\[
\hat g(y) = \frac1{2\pi}\sqrt{\frac\pi\varepsilon}\,
\eu^{-(y - x)^2/4\varepsilon} = g_\varepsilon(x - y),
\]
so $\frac1{2\pi}\int\hat f(\xi)\eu^{-\varepsilon\xi^2}
\eu^{\iu x\xi}\dd\xi = \int f(y)g_\varepsilon(x - y)\dd y =
(f*g_\varepsilon)(x)$. The $g_\varepsilon$ are an
approximate identity: $g_\varepsilon \geq 0$,
$\int g_\varepsilon = 1$ (Gaussian integral), concentrating at
$0$; the proof of \cref{thm:b3:lp:regularization}(2) applies
verbatim (only $\int g_\varepsilon = 1$ and concentration were
used: for the tail, $\int_{\abs y > \delta}g_\varepsilon \to
0$): $\norm{f * g_\varepsilon - f}_1 \to 0$.

(2) If $\hat f \in L^1$: the right side of (1) converges, by
DCT (dominator $\abs{\hat f}$), to $\frac1{2\pi}\int\hat
f(\xi)\eu^{\iu x\xi}\dd\xi$ for \emph{every} $x$, and this
limit function is continuous (DCT again). On the other hand $f
* g_\varepsilon \to f$ in $L^1$, so along a subsequence a.e.\
(\cref{thm:b3:lp:complete}): the two limits agree a.e.

(3) $\hat f = 0$ makes the right side of (1) vanish: $f *
g_\varepsilon = 0$ for all $\varepsilon$, and $f *
g_\varepsilon \to f$ in $L^1$: $f = 0$ a.e.
\end{proof}

\section{The Schwartz class}

\begin{definition}\label{def:b3:fouriertransform:schwartz}
The \emph{Schwartz class}\index{Schwartz class} $\mathcal
S(\R)$ consists of the $\mathcal C^\infty$ functions $f$ with
$\sup_x\abs{x^m f^{(n)}(x)} < \infty$ for all $m, n \geq 0$
(all derivatives decay faster than any power). Examples:
$\eu^{-ax^2}$, $\mathcal C_c^\infty$. Clearly $\mathcal S
\subseteq L^p$ for every $p$ (bound by $C(1 + x^2)^{-1}$), and
$\mathcal S$ is stable under derivatives, multiplication by
polynomials, and products.
\end{definition}

\begin{theorem}\label{thm:b3:fouriertransform:schwartz}
The Fourier transform maps $\mathcal S(\R)$ bijectively onto
itself, with inverse $\check g(x) = \frac1{2\pi}\int
g(\xi)\eu^{\iu x\xi}\dd\xi$.
\end{theorem}

\begin{proof}
Let $f \in \mathcal S$. Iterating
\cref{prop:b3:fouriertransform:basic}(3), $(\hat f)^{(n)} =
\widehat{(-\iu x)^nf}$ (each $x^kf \in L^1$); iterating (4)
with $h = (-\iu x)^nf \in \mathcal S$ (all of whose
derivatives are integrable), $(\iu\xi)^m\hat h =
\widehat{h^{(m)}}$. Combining,
\[
\abs{\xi^m\,(\hat f)^{(n)}(\xi)}
= \bigl|\widehat{\,h^{(m)}}(\xi)\bigr|
\leq \bigl\|\bigl((-\iu x)^nf\bigr)^{(m)}\bigr\|_1 < \infty
\]
uniformly in $\xi$: $\hat f \in \mathcal S$. Since $\hat f \in L^1$, inversion
(\cref{thm:b3:fouriertransform:inversion}(2)) holds
everywhere (both sides continuous): $\check{\hat f} = f$, and
symmetrically $\widehat{\check g} = g$ (the check transform
is $g \mapsto \frac1{2\pi}\hat g(-\cdot)$, again preserving
$\mathcal S$): bijection.
\end{proof}

\section{Plancherel and \texorpdfstring{$L^2$}{L2}}

\begin{theorem}[Plancherel]\label{thm:b3:fouriertransform:plancherel}
For $f \in L^1 \cap L^2(\R)$:\index{Plancherel's theorem}
\[
\norm{\hat f}_2^2 = 2\pi\,\norm f_2^2 .
\]
Consequently $\widehat{\phantom f}$ extends uniquely to a
continuous linear map $\mathcal F \colon L^2(\R) \to L^2(\R)$
with $\norm{\mathcal Ff}_2 = \sqrt{2\pi}\norm f_2$; $\mathcal
F$ is bijective, with $\mathcal F^{-1} =
\frac1{2\pi}\,\mathcal F\circ\sigma$ where $\sigma f =
f(-\cdot)$, and it preserves inner products up to the factor
$2\pi$.
\end{theorem}

\begin{proof}
Let $f \in L^1\cap L^2$ and $h = f * \tilde f$ with
$\tilde f(x) = \overline{f(-x)}$. Then $h \in L^1$
(\cref{thm:b3:product:convolution}), $h$ is continuous and
bounded (\cref{exo:b3:lp:6}: $f, \tilde f \in L^2$), $h(0) =
\int f\bar f = \norm f_2^2$, and $\hat h = \hat
f\,\widehat{\tilde f} = \hat f\,\overline{\hat f} =
\abs{\hat f}^2 \geq 0$ (compute $\widehat{\tilde f} =
\overline{\hat f}$). Apply
\cref{thm:b3:fouriertransform:inversion}(1) to $h$ at $x = 0$:
\[
(h * g_\varepsilon)(0) = \frac1{2\pi}\int \hat
h(\xi)\,\eu^{-\varepsilon\xi^2}\dd\xi .
\]
As $\varepsilon \to 0$: the left side tends to $h(0)$ ($h$
continuous bounded: $(h*g_\varepsilon)(0) - h(0) = \int(h(-y)
- h(0))g_\varepsilon(y)\dd y \to 0$ by splitting small/large
$y$); the right side increases to $\frac1{2\pi}\int\hat h$ by
MCT ($\hat h \geq 0$). Hence $\frac1{2\pi}\int\abs{\hat f}^2 =
\norm f_2^2$, finite or not \emph{a priori} --- and finite,
proving both membership and the identity.

Extension: $L^1\cap L^2 \supseteq \mathcal C_c$ is dense in
$L^2$ (\cref{thm:b3:lp:density}); the transform is
$\sqrt{2\pi}$-isometric there, so extends uniquely to an
isometry-up-to-constant $\mathcal F$ on $L^2$
(\cref{thm:b3:complete:extension}). Inversion for $\mathcal S$
(\cref{thm:b3:fouriertransform:schwartz}) transfers by the
same density (both sides $L^2$-continuous): $\mathcal
F\bigl(\frac1{2\pi}\mathcal F(\sigma f)\bigr) = f$ on
$\mathcal S$, hence on $L^2$: bijectivity. Inner products:
polarization from the norm identity.
\end{proof}

\begin{theorem}[Poisson summation]
\label{thm:b3:fouriertransform:poisson}
Let $f \in \mathcal S(\R)$ (continuous $f$ with $\abs{f} +
\abs{\hat f} \leq C(1 + \abs\cdot)^{-2}$
suffices). Then\index{Poisson summation formula}
\[
\sum_{n\in\Z} f(n) \;=\; \sum_{k\in\Z}\hat f(2\pi k) .
\]
\end{theorem}

\begin{proof}
Let $F(x) = \sum_{n\in\Z}f(x + n)$: the series converges
normally on compacts (decay of $f$), so $F$ is continuous, and
it is $1$-periodic. Its Fourier coefficients (period $1$:
$c_k(F) = \int_0^1F(t)\eu^{-2\iu\pi kt}\dd t$):
\[
c_k(F) = \sum_n\int_0^1 f(t + n)\,\eu^{-2\iu\pi kt}\dd t
= \int_\R f(t)\,\eu^{-2\iu\pi kt}\dd t = \hat f(2\pi k)
\]
(normal convergence justifies the interchange; the phase is
$1$-periodic). The series $\sum_k\abs{c_k(F)}$ converges
(decay of $\hat f$), so the Fourier series of $F$ converges
normally; its sum is a continuous function with the same
Fourier coefficients as $F$, hence equals $F$ (injectivity on
the circle: the difference has zero coefficients, and
\cref{thm:b3:hilbert:fourier} gives zero in $L^2$, hence
everywhere by continuity). Evaluate at $x = 0$.
\end{proof}

\begin{example}[The theta identity]
\label{ex:b3:fouriertransform:theta}
Applying Poisson to $f(x) = \eu^{-\pi tx^2}$ ($t > 0$), whose
transform is $\hat f(\xi) = t^{-1/2}\eu^{-\xi^2/4\pi t}$
(\cref{ex:b3:fouriertransform:gaussian} with $a = \pi t$):
\[
\sum_{n\in\Z}\eu^{-\pi n^2t}
= \frac1{\sqrt t}\sum_{k\in\Z}\eu^{-\pi k^2/t} :
\]
the functional equation of Jacobi's theta function, key to the
functional equation of Riemann's $\zeta$ --- and a spectacular
numerical accelerator: for $t$ small, the left side converges
slowly, the right side blazingly fast.
\end{example}

\begin{method}\label{met:b3:fouriertransform:method}
Working ranges: $L^1$ --- transform defined pointwise,
inversion needs $\hat f \in L^1$; $\mathcal S$ --- everything
is legal, prove here first; $L^2$ --- transform defined by
density (not by the integral!), perfect symmetry, Parseval
bookkeeping. To compute a transform: reduce to the table
(indicator, exponential, Gaussian) by the rules of
\cref{prop:b3:fouriertransform:basic}; to prove an identity:
establish it on $\mathcal S$ (or $\mathcal C_c^\infty$) and
extend by density and continuity
(\cref{met:b3:lp:toolkit}); to solve a linear PDE or ODE with
constant coefficients: transform, divide, invert.
\end{method}

\begin{omfigure}
\begin{tikzpicture}[xscale=1.05, yscale=2.2]
  \draw[->] (-4.3,0) -- (4.3,0) node[right,font=\small] {$x$};
  \draw[->] (0,0) -- (0,1.25);
  \draw[omThm, thick, smooth]
    plot[domain=-4:4, samples=81]
    (\x, {exp(-\x*\x/0.4)/1.1209});
  \draw[omProp, thick, smooth]
    plot[domain=-4:4, samples=81]
    (\x, {exp(-\x*\x/1.6)/2.2418});
  \draw[omDef, thick, smooth]
    plot[domain=-4:4, samples=81]
    (\x, {exp(-\x*\x/6.4)/4.4836});
  \node[omThm, font=\small] at (1.35,0.75) {$t = 0.1$};
  \node[omProp, font=\small] at (2.4,0.36) {$t = 0.4$};
  \node[omDef, font=\small] at (3.5,0.17) {$t = 1.6$};
\end{tikzpicture}

{\small The heat kernel $g_t(x) =
\frac1{2\sqrt{\pi t}}\,\eu^{-x^2/4t}$ at three times: total
mass $1$ forever, height $\sim t^{-1/2}$, width $\sim
\sqrt t$. Convolving initial data with this spreading
Gaussian is the entire content of the weekend problem; in
frequency, the same picture reads
$\hat g_t(\xi) = \eu^{-t\xi^2}$ --- high frequencies die
first, and that asymmetry is the arrow of time.}
\end{omfigure}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:b3:fouriertransform:1}
Compute the Fourier transforms of: $\mathbf 1_{\intcc{-a}a}$;
$\eu^{-a\abs x}$ ($a > 0$); the tent function $\max(0, 1 -
\abs x)$; $\frac1{x^2 + a^2}$ \emph{(use inversion on the
second)}. Record the emerging table.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b3:fouriertransform:2}
Let $f \in L^1$. Express in terms of $\hat f$ the transforms
of: $f(x - a)$, $f(x)\cos(bx)$, $f(ax + b)$,
$\overline{f(-x)}$, $(f * f)(x)$. Verify each rule on the
Gaussian.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:fouriertransform:3}
(a) Show that $\mathbf 1_{\intcc{-1}1} * \mathbf
1_{\intcc{-1}1}$ has transform $\bigl(\frac{2\sin\xi}\xi
\bigr)^2$, and deduce
$\int_\R\bigl(\frac{\sin\xi}\xi\bigr)^2\dd\xi = \pi$ by
Plancherel --- or by inversion at $0$. Compare
\cref{pb:b3:lebesgue:1}.
(b) Compute $\int_\R\frac{\dd x}{(x^2+1)^2}$ via Plancherel
applied to $\eu^{-\abs x}$.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:fouriertransform:4}
(Heat kernel algebra) With $g_t(x) =
\frac1{2\sqrt{\pi t}}\eu^{-x^2/4t}$:
(a) verify $\hat g_t(\xi) = \eu^{-t\xi^2}$;
(b) deduce the semigroup law $g_t * g_s = g_{t+s}$ without any
integral computation;
(c) show $\norm{g_t}_1 = 1$ and $\norm{g_t}_2^2 =
(8\pi t)^{-1/2}$.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:fouriertransform:5}
Show that if $f \in L^1$ is even and real, $\hat f$ is even
and real; if $f$ is odd and real, $\hat f$ is odd and purely
imaginary. What does $\hat f(0)$ compute? Deduce that $f \geq
0$ forces $\norm{\hat f}_\infty = \hat f(0) = \int f$, and
interpret for probability densities (\cref{ch:b3:clt}: a
characteristic function has modulus $\leq 1$, attained at
$0$).
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b3:fouriertransform:6}
(Non-surjectivity) Show that $\widehat{\phantom f}\colon L^1
\to \mathcal C_0$ is injective and continuous, but \emph{not}
surjective, in three steps.
(i) Injectivity (\cref{thm:b3:fouriertransform:inversion}) and
continuity ($\norm{\hat f}_\infty \leq \norm f_1$), and
$\mathcal C_0$ is a Banach space (closed in
$\norm\cdot_\infty$).
(ii) If the map were surjective, it would be bijective, and
the open mapping theorem (\cref{thm:b3:banach:openmapping})
would give a constant $C$ with $\norm f_1 \leq C\norm{\hat
f}_\infty$ for all $f \in L^1$.
(iii) Contradict this with $f_n(x) = \frac{\sin
x}{x}\cdot\frac{\sin(x/n)}{x/n}$: its transform is (up to
constants) the convolution $\mathbf 1_{\intcc{-1}1} *
\mathbf 1_{\intcc{-1/n}{1/n}}$-type trapezoid --- show
$\norm{\hat f_n}_\infty \leq \pi$ uniformly, while
$\norm{f_n}_1 \geq c\ln n$ by counting the arches of
$\frac{\abs{\sin x}}x$ on $[1, n]$ (where the second factor is
bounded below), as in \cref{thm:b3:banach:fourierdiverge}.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:fouriertransform:7}
(Smoothness $\leftrightarrow$ decay dictionary) Prove: $f \in
L^1$ with $\hat f(\xi) = O(\abs\xi^{-k-1-\delta})$ for some
$\delta > 0$ implies $f$ has a $\mathcal C^k$ representative.
Conversely $f \in \mathcal C^k_c$ implies $\hat f(\xi) =
O(\abs\xi^{-k})$. Illustrate both directions on the tent
function.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b3:fouriertransform:8}
(Heisenberg's inequality) For $f \in \mathcal S(\R)$ real with
$\norm f_2 = 1$, prove
\[
\Bigl(\int x^2f(x)^2\dd x\Bigr)\cdot
\Bigl(\frac1{2\pi}\int \xi^2\abs{\hat f(\xi)}^2\dd\xi\Bigr)
\;\geq\; \frac14 ,
\]
with equality for Gaussians. \emph{(Write $1 = \int f^2 =
-\int x\,(f^2)'$ by parts, bound by Cauchy--Schwarz, and
convert $\norm{f'}_2$ by Plancherel.)} Interpretation: a
signal and its spectrum cannot both be concentrated.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:fouriertransform:9}
Justify \cref{ex:b3:fouriertransform:theta} in detail
(hypotheses of Poisson for the Gaussian), and use the identity
to evaluate $\sum_{n\in\Z}\eu^{-\pi n^2}$ to six decimals with
three terms. How many terms of the \emph{defining} series
would the same accuracy require at $t = 10^{-2}$, versus the
transformed series?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:fouriertransform:10}
(Band-limited functions) Let $f \in L^2(\R)$ with $\mathcal Ff$
supported in $\intcc{-\pi}\pi$. Show that $f$ has a
representative extending each of whose values is recoverable
from samples: prove the \emph{Shannon interpolation} at the
integers,
\[
f(x) = \sum_{n\in\Z} f(n)\,
\frac{\sin\bigl(\pi(x - n)\bigr)}{\pi(x - n)}
\quad\text{in } L^2,
\]
by expanding $\mathcal Ff$ in the Fourier basis of
$L^2(\intcc{-\pi}\pi)$ (\cref{thm:b3:hilbert:fourier}) and
transforming back term by term.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:fouriertransform:11}
(The transform as an operator of order four) On $\mathcal
S(\R)$, let $\mathcal F f = \hat f$.
(a) Using the inversion formula, show $(\mathcal F^2f)(x) =
2\pi\,f(-x)$, and deduce $\mathcal F^4 = (2\pi)^2\,
\mathrm{id}$.
(b) Deduce that every eigenvalue of $\mathcal F$ on
$\mathcal S$ belongs to $\{\pm\sqrt{2\pi},
\pm\iu\sqrt{2\pi}\}$, and exhibit an eigenfunction for
$+\sqrt{2\pi}$ \emph{(which function of this chapter is
proportional to its own transform?)}.
(c) Show that even functions satisfy $\mathcal F^2f = 2\pi
f$ and odd ones $\mathcal F^2f = -2\pi f$; produce an
eigenfunction for the eigenvalue $-\iu\sqrt{2\pi}$ from
$x\eu^{-x^2/2}$ by computing its transform (differentiate
the Gaussian's transform).
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:fouriertransform:12}
(Autocorrelation and Wiener's lemma) For $f \in L^2(\R)$
define $\tilde f(x) = \overline{f(-x)}$ and the
\emph{autocorrelation} $A_f = f * \tilde f$.
(a) Show that $A_f$ is a bounded continuous function with
$A_f(0) = \norm f_2^2 \geq \abs{A_f(x)}$ for all $x$
(\cref{exo:b3:lp:6} and Cauchy--Schwarz).
(b) Show, first for $f \in L^1\cap L^2$, that
$\widehat{A_f} = \abs{\hat f\,}^2 \geq 0$: the
autocorrelation has nonnegative transform --- spectra of
autocorrelations are power spectra.
(c) Deduce the identity $\int_\R\abs{\hat
f(\xi)}^2\eu^{\iu x\xi}\,\dd\xi = 2\pi A_f(x)$ (inversion;
justify its applicability when $\hat f \in L^2$ has
$\abs{\hat f}^2 \in L^1$), and evaluate it for $f = \mathbf
1_{\intcc{-1/2}{1/2}}$ at $x = 0$: recover
$\int_\R\bigl(\frac{\sin u}u\bigr)^2\dd u = \pi$.
\end{exercise}

\section{Problem: the heat equation on the line}

\begin{problem}[{Weekend problem --- $\partial_tu =
\partial^2_{xx}u$, solved end to end}]
\label{pb:b3:fouriertransform:1}
Heat spreads; the equation $\partial_tu = \partial_{xx}^2u$
says its density diffuses at a rate given by the local
curvature of the temperature profile. We solve the Cauchy
problem on $\R$ --- given $f$, find $u(t, x)$ for $t > 0$ with
$u(0, \cdot) = f$ --- prove the solution's remarkable
properties, and see why time cannot be reversed. Throughout,
$g_t(x) = \frac1{2\sqrt{\pi t}}\eu^{-x^2/4t}$ is the
\emph{heat kernel}\index{heat kernel} and $u(t, \cdot) = g_t *
f$.

\medskip
\noindent\textbf{Part I --- Deriving the kernel.} Work
formally first: suppose $u(t, \cdot) \in \mathcal S$ solves
the equation, and let $\hat u(t, \xi)$ be the transform in
$x$.
\begin{enumerate}
  \item Show (formally) $\partial_t\hat u = -\xi^2\hat u$,
        hence $\hat u(t,\xi) = \eu^{-t\xi^2}\hat f(\xi)$, and
        recognize $u(t) = g_t * f$
        (\cref{exo:b3:fouriertransform:4}). This motivates the
        \emph{definition} of $u$; everything is now proved
        directly, for $f \in \mathcal C_b(\R)$ (bounded
        continuous) or $f \in L^p$.
\end{enumerate}

\medskip
\noindent\textbf{Part II --- Verification.}
\begin{enumerate}[resume]
  \item Show that for $t > 0$, $u(t, x) = \int
        g_t(x-y)f(y)\dd y$ is well defined for $f \in \mathcal
        C_b$, and that $u$ is $\mathcal C^\infty$ in $(t, x)$
        on $\intoo0\infty\times\R$ \emph{(differentiate under
        the integral; dominate Gaussian derivatives locally
        uniformly in $(t,x)$)}.
  \item Verify $\partial_tg_t = \partial^2_{xx}g_t$ by direct
        computation, and deduce $\partial_tu =
        \partial^2_{xx}u$ for $t > 0$.
  \item (Initial condition) Show that for $f \in \mathcal
        C_b$, $u(t, x) \to f(x)$ as $t \to 0^+$, uniformly on
        compacts \emph{(approximate identity: split $\abs y
        \leq \delta$, $\abs y > \delta$)}; for $f \in L^p$
        ($p < \infty$), show $\norm{u(t) - f}_p \to 0$.
  \item (Instant smoothing) Conclude: even for merely bounded
        continuous $f$, the solution is $\mathcal C^\infty$
        for every $t > 0$ --- heat instantly erases
        roughness. Compute $u(t, \cdot)$ explicitly for $f =
        \mathbf 1_{\intoo0\infty}$ (an error function) and
        sketch its profile for three values of $t$.
\end{enumerate}

\medskip
\noindent\textbf{Part III --- Structural properties.}
\begin{enumerate}[resume]
  \item (Positivity and comparison) If $f \geq 0$ then $u > 0$
        for all $t > 0$ (strictly, unless $f = 0$ a.e.); if
        $f_1 \leq f_2$ then $u_1 \leq u_2$. A cold spot warms
        instantly: comment.
  \item (Conservation) For $f \in L^1$: $\int u(t, x)\dd x =
        \int f$ for all $t$ \emph{(Tonelli)} --- total heat is
        conserved.
  \item (Dissipation) For $f \in L^1\cap L^2$, show via
        Plancherel that $t \mapsto \norm{u(t)}_2$ is
        nonincreasing, strictly unless $f = 0$, and compute
        its limit as $t \to \infty$. Show moreover
        $\norm{u(t)}_\infty \leq \frac{\norm
        f_1}{2\sqrt{\pi t}} \to 0$: heat spreads and dies.
  \item (Uniqueness, $L^2$ class) Let $u$ be a solution with
        $u(t) \in L^2$ for all $t$, $u \in \mathcal
        C^1(\intoo0\infty, L^2)$ in the natural sense and
        $u(t) \to f$ in $L^2$ as $t\to0$; admitting that the
        transform converts it to $\partial_t\hat u =
        -\xi^2\hat u$ pointwise a.e.\ in $\xi$ for a.e.\ $t$
        \emph{(justified by testing against $\mathcal
        C_c^\infty$ in $\xi$ --- outline this)}, show
        $\hat u(t,\xi) = \eu^{-t\xi^2}\hat f(\xi)$, hence
        uniqueness in this class.
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- The arrow of time.}
\begin{enumerate}[resume]
  \item Show that the \emph{backward} problem is ill-posed:
        for the solution to exist at time $-s$ ($s > 0$) with
        data $f$ at time $0$ --- i.e.\ for $f = g_s * h$ to
        have a solution $h \in L^2$ --- it is necessary that
        $\eu^{s\xi^2}\hat f(\xi) \in L^2$: an extreme decay
        condition on $\hat f$. Exhibit an explicit smooth $f
        \in L^2$ for which no backward solution exists on any
        time interval: take the function with $\hat f(\xi) =
        \eu^{-\abs\xi}$ --- identify $f$
        (\cref{exo:b3:fouriertransform:1}) and show
        $\eu^{s\xi^2}\eu^{-\abs\xi} \notin L^2$ for every $s >
        0$.
  \item (Smoothing vs.\ information) Explain in a short
        paragraph, using questions 5, 9 and 10, why the heat
        semigroup $(f \mapsto g_t * f)_{t\geq0}$ is injective
        but not surjective on $L^2$, and why this expresses
        the irreversibility of diffusion.
\end{enumerate}

\medskip
\noindent\textbf{Part V --- Shannon's sampling theorem.} A
function $f \in L^2(\R)$ is \emph{band-limited} to $\Omega$
if $\hat f = 0$ a.e.\ outside $\intcc{-\Omega}\Omega$; write
$PW_\Omega$ (Paley--Wiener space) for these.
\begin{enumerate}[resume]
  \item Show that every $f \in PW_\Omega$ agrees a.e.\ with
        the $\mathcal C^\infty$ function
        $\frac1{2\pi}\int_{-\Omega}^{\Omega}\hat
        f(\xi)\eu^{\iu x\xi}\,\dd\xi$ (justify smoothness and
        the identification), with all derivatives bounded:
        band limitation is an extreme form of regularity.
        From now on $f$ denotes this representative.
  \item Expand $\hat f \in L^2(\intcc{-\Omega}\Omega)$ in the
        Fourier basis of that interval and identify the
        coefficients as \emph{samples} of $f$:
        \[
        \hat f(\xi) = \frac\pi\Omega\sum_{n\in\Z}
        f\Bigl(\frac{n\pi}\Omega\Bigr)\,
        \eu^{-\iu n\pi\xi/\Omega}
        \quad\text{in } L^2(\intcc{-\Omega}\Omega) .
        \]
  \item Deduce the \emph{sampling theorem}: for $f \in
        PW_\Omega$,
        \[
        f(x) = \sum_{n\in\Z}f\Bigl(\frac{n\pi}
        \Omega\Bigr)\,\operatorname{sinc}(\Omega x - n\pi),
        \qquad \operatorname{sinc}t = \frac{\sin t}t,
        \]
        with convergence in $L^2(\R)$ and uniformly on $\R$
        \emph{(inject the series of question 13 into the
        inversion formula and compute the elementary
        integral)}: a band-limited signal is entirely
        determined by its values on a grid of step
        $\pi/\Omega$ --- the Nyquist rate.
  \item Show that the functions $x \mapsto
        \operatorname{sinc}(\Omega x - n\pi)$, $n \in \Z$,
        form an orthogonal family in $L^2(\R)$ with constant
        norm $\sqrt{\pi/\Omega}$, and deduce the energy
        identity $\norm f_2^2 =
        \frac\pi\Omega\sum_n\abs{f(n\pi/\Omega)}^2$.
  \item (Aliasing) Exhibit a nonzero $g \in PW_{2\Omega}$
        vanishing at every sample point $\frac{n\pi}\Omega$
        \emph{(consider $g(x) =
        \sin(\Omega x)\operatorname{sinc}(\Omega x)$ and
        check its band)}: sampling below the Nyquist rate
        loses information --- two different signals can share
        all samples: the stroboscopic wagon-wheel effect,
        mathematized.
  \item (Degrees of freedom) Using questions 14--15, justify
        the engineering rule: a signal band-limited to
        $\Omega$ whose energy is essentially carried by a
        time window of length $T$ is described by
        approximately $\frac{\Omega T}\pi$ real samples ---
        make ``essentially'' precise through the energy
        identity and the tail $\sum_{\abs{n\pi/\Omega} >
        T/2}$.
  \item (Consistency checks) Verify the sampling theorem by
        hand on two members of $PW_\Omega$: (a) $f =
        \operatorname{sinc}(\Omega\,\cdot)$, whose samples
        are $\delta_{n0}$; (b) $f(x) = \cos(\omega x)
        \operatorname{sinc}(\varepsilon x)$-type narrowband
        signals --- more precisely, show that for $f \in
        PW_{\Omega'}$ with $\Omega' < \Omega$, the
        $\Omega$-rate series also reconstructs $f$
        (oversampling is harmless), by embedding $PW_
        {\Omega'} \subseteq PW_\Omega$.
\end{enumerate}

\medskip
\noindent\textbf{Part VI --- Uncertainty, twice more.}
Heisenberg's inequality (\cref{exo:b3:fouriertransform:8})
bounds \emph{how} concentrated $f$ and $\hat f$ can jointly
be; here are its all-or-nothing sibling and its exact
saturation.
\begin{enumerate}[resume]
  \item Let $f \in L^1$ with $\operatorname{supp}f \subseteq
        \intcc{-A}A$. Show that $\hat f$ is the sum of an
        everywhere-convergent power series:
        \[
        \hat f(\xi) = \sum_{k\geq0}\frac{(-\iu\xi)^k}{k!}
        \,m_k, \qquad m_k = \int_{-A}^{A}x^kf(x)\,\dd x,
        \quad \abs{m_k} \leq A^k\norm f_1
        \]
        \emph{(expand $\eu^{-\iu\xi x}$ and justify the
        interchange by normal convergence)}: the transform is
        \emph{real-analytic}, with infinite radius of
        convergence at every point.
  \item Deduce the support dichotomy: a real-analytic
        function vanishing on a nonempty open interval
        vanishes identically \emph{(the set where all
        derivatives vanish is open and closed --- work out
        the Taylor argument)}; conclude that no nonzero $f$
        has both $f$ and $\hat f$ compactly supported, and
        that $PW_\Omega$ contains no nonzero compactly
        supported function --- band-limited signals last
        forever, and time-limited signals leak into all
        frequencies.
  \item (Saturation of Heisenberg) On the Gaussian family $f
        = \eu^{-ax^2}$, compute both concentration factors
        and verify that the normalized product
        $\bigl(\int x^2\abs f^2\bigr)\bigl(\frac1{2\pi}\int
        \xi^2\abs{\hat f}^2\bigr)\big/\norm f_2^4$ equals
        $\frac14$ for \emph{every} $a$ --- the equality
        family of \cref{exo:b3:fouriertransform:8} in the
        flesh; explain by a scaling argument why the product
        must be constant along the family.
  \item Explain the physical reading (position/momentum
        densities of a quantum state; $\hbar$ in the
        normalization gives $\sigma_x\sigma_p \geq
        \frac\hbar2$), and connect across the chapter:
        instant smoothing (Part II), irreversibility (Part
        IV), sampling (Part V), Heisenberg and the support
        dichotomy are five expressions of one law --- the
        behavior of $\hat f$ at infinity legislates what $f$
        may do anywhere.
\end{enumerate}

\medskip
\noindent\textbf{Part VII --- The kernel's algebra, and one
solvable example.}
\begin{enumerate}[resume]
  \item (Semigroup) Prove the \emph{Chapman--Kolmogorov
        identity} $g_t * g_s = g_{t+s}$ for $t, s > 0$ (via
        the convolution theorem and injectivity of the
        transform on $L^1$), and deduce $u(t + s) = g_s *
        u(t)$: evolving for time $t + s$ is evolving for $t$,
        then for $s$. Sharpen the dissipation of question 8:
        writing $\norm{u(t)}_2^2 = \frac1{2\pi}\int
        \eu^{-2t\xi^2}\abs{\hat f(\xi)}^2\dd\xi$, show by
        Cauchy--Schwarz that
        \[
        t \longmapsto \ln\,\norm{u(t)}_2
        \quad\text{is convex on } \intoo0{+\infty} :
        \]
        the $L^2$ energy does not merely decrease, it
        decreases in a log-convex way.
  \item (Where the heat goes) Let $f \geq 0$, $f \in L^1$,
        with $\int x^2f(x)\dd x < \infty$. Show that for all
        $t > 0$
        \[
        \int_\R x\,u(t, x)\,\dd x = \int_\R x f(x)\,\dd x,
        \qquad
        \int_\R x^2u(t, x)\,\dd x
        = \int_\R x^2f(x)\,\dd x + 2t\int_\R f :
        \]
        the center of heat never moves, and the variance
        grows \emph{linearly} in time --- the diffusive
        scaling $x \sim \sqrt{2t}$, to be reread when
        Brownian motion appears in \cref{ch:b3:probability}.
        \emph{(Compute the first two moments of $g_t$ and use
        Tonelli on the convolution.)}
  \item (The Gaussian, solved end to end) For $f(x) =
        \eu^{-x^2}$, establish the closed form
        \[
        u(t, x) = \frac1{\sqrt{1 + 4t}}\,
        \exp\Bigl(-\frac{x^2}{1 + 4t}\Bigr),
        \]
        and verify on it, by hand: the equation $\partial_tu
        = \partial^2_{xx}u$; conservation $\int u(t) =
        \sqrt\pi$; the dissipation law $\norm{u(t)}_2 =
        (\pi/2)^{1/4}(1 + 4t)^{-1/4}$ (compare its $t^{-1/4}$
        decay with the $t^{-1/2}$ sup-norm decay of question
        8); and the exact variance growth of question 24. At
        $t = 6$: the peak has dropped to $\frac15$ of its
        initial height while the profile is five times wider
        --- same heat, spread out.
\end{enumerate}
\end{problem}
