\chapter{Compact Operators and the Spectral Theorem}
\label{ch:b3:spectral}

Diagonalization is finite-dimensional linear algebra's crown
jewel: a symmetric matrix has an orthonormal basis of
eigenvectors. In infinite dimension this fails for bounded
\omterm{def:b3:spectral:selfadjoint}{self-adjoint operators} in general --- multiplication by $x$ on
$L^2(\intcc01)$ has \emph{no} eigenvalues at all
(\cref{exo:b3:spectral:6}) --- but it survives, in \omterm{prop:b3:galois:perfect}{perfect}
form, for the operators that are \emph{almost finite
dimensional}: the \omterm{def:b3:spectral:compact}{compact} ones. The spectral theorem for
\omterm{def:b3:spectral:compact}{compact} \omterm{def:b3:spectral:selfadjoint}{self-adjoint operators} is the single most used theorem
of applied functional analysis: it diagonalizes integral
equations, drives the \omterm{thm:b3:spectral:fredholm}{Fredholm alternative}, and (weekend
problem) solves the vibrating string, producing the sine basis
of Fourier analysis from pure operator theory --- with Euler's
$\zeta(2) = \frac{\pi^2}6$ falling out of a trace formula as a
parting gift. Throughout, $H$ is a \omterm{def:b3:hilbert:inner}{Hilbert space} over $\C$ (or
$\R$; statements adapt), and operators are bounded.

\section{Compact operators}

\begin{definition}\label{def:b3:spectral:compact}
$T \in \mathcal L(E, F)$ ($E, F$ Banach) is
\emph{compact}\index{compact operator} if the image $T(B)$ of
the unit ball is relatively compact in $F$ --- equivalently,
every bounded sequence $(x_n)$ has a subsequence with
$(Tx_{n_k})$ convergent. Finite-rank operators are compact
(bounded sets in finite dimension); the identity of an
infinite-dimensional space never is (Riesz's theorem, Year 2).
\end{definition}

\begin{proposition}\label{prop:b3:spectral:ideal}
The \omterm{def:b3:spectral:compact}{compact operators} $\mathcal K(E, F)$ form a closed subspace
of $\mathcal L(E, F)$, and a two-sided \omterm{def:b3:rings:ideal}{ideal}: $S$ \omterm{def:b3:spectral:compact}{compact}
$\Rightarrow$ $AS$ and $SB$ \omterm{def:b3:spectral:compact}{compact} for bounded $A, B$.
Moreover, in a \omterm{def:b3:hilbert:inner}{Hilbert space}, every \omterm{def:b3:spectral:compact}{compact operator} is a
norm-limit of finite-rank operators.
\end{proposition}

\begin{proof}
Subspace: clear from the sequence characterization. \omterm{def:b3:rings:ideal}{Ideal}:
bounded maps send convergent sequences to convergent ones and
bounded to bounded. Closedness: let $T_n \to T$ with $T_n$
\omterm{def:b3:spectral:compact}{compact}, and $(x_k)$ bounded by $1$; a diagonal extraction
makes $(T_nx_{k_j})_j$ convergent for every $n$; then
$(Tx_{k_j})$ is Cauchy, since
\[
\norm{Tx_{k_j} - Tx_{k_l}} \leq 2\vertiii{T - T_n} +
\norm{T_nx_{k_j} - T_nx_{k_l}} ,
\]
choosing first $n$ then the indices. Approximation in \omterm{def:b3:hilbert:inner}{Hilbert
spaces}: let $T$ be \omterm{def:b3:spectral:compact}{compact}, $K = \overline{T(B)}$ \omterm{def:b3:spectral:compact}{compact};
given $\varepsilon$, cover $K$ by finitely many balls
$B(y_i, \varepsilon)$ and let $P$ be the orthogonal projection
onto $V = \operatorname{Vect}(y_1, \dots, y_m)$ (closed:
finite-dimensional). Then $PT$ has finite rank, and for
$\norm x \leq 1$: picking $y_i$ with $\norm{Tx - y_i} <
\varepsilon$,
\[
\norm{Tx - PTx} \leq \norm{Tx - y_i} + \norm{P(y_i - Tx)}
\leq 2\varepsilon
\]
($y_i = Py_i$; $\vertiii P \leq 1$): $\vertiii{T - PT} \leq
2\varepsilon$.
\end{proof}

\begin{example}\label{ex:b3:spectral:examples}
(a) Diagonal operators on $\ell^2$: $T(x_n) = (\lambda_nx_n)$
is \omterm{def:b3:spectral:compact}{compact} iff $\lambda_n \to 0$
(\cref{exo:b3:spectral:2}).
(b) Kernel operators on $\mathcal C(\intcc01)$: \omterm{def:b3:spectral:compact}{compact} by
Ascoli (\cref{exo:b3:complete:7}).
(c) \emph{Hilbert--Schmidt operators}\index{Hilbert--Schmidt
operator}: for $k \in L^2(\intcc01^2)$,
\[
(T_kf)(x) = \int_0^1k(x, y)\,f(y)\,\dd y
\]
defines a \omterm{def:b3:spectral:compact}{compact operator} on $L^2(\intcc01)$ with
$\vertiii{T_k} \leq \norm k_{L^2}$
(\cref{exo:b3:spectral:4}: truncating the basis expansion of
$k$ exhibits $T_k$ as a limit of finite-rank operators).
\end{example}

\section{Self-adjoint operators}

\begin{definition}\label{def:b3:spectral:selfadjoint}
$T \in \mathcal L(H)$ is \emph{self-adjoint}\index{self-adjoint
operator} if $T = T^*$ (\cref{exo:b3:hilbert:8}), i.e.\
$\langle Tx, y\rangle = \langle x, Ty\rangle$ for all $x, y$.
Then $\langle x, Tx\rangle \in \R$ for every $x$ (equal to its
conjugate).
\end{definition}

\begin{proposition}\label{prop:b3:spectral:sanorm}
For \omterm{def:b3:spectral:selfadjoint}{self-adjoint} $T$:\index{numerical radius}
\[
\vertiii T = \sup_{\norm x \leq 1}\ \abs{\langle x,
Tx\rangle} .
\]
Eigenvalues of $T$ are real, and eigenvectors for distinct
eigenvalues are orthogonal.
\end{proposition}

\begin{proof}
Let $M$ be the supremum; $M \leq \vertiii T$ by
Cauchy--Schwarz. Conversely, the polarization-type identity
\[
\langle x{+}y, T(x{+}y)\rangle - \langle x{-}y,
T(x{-}y)\rangle = 4\operatorname{Re}\langle y, Tx\rangle
\]
(expand; the cross terms $\langle y, Tx\rangle + \langle x,
Ty\rangle = 2\operatorname{Re}\langle y, Tx\rangle$ by
\omterm{def:b3:spectral:selfadjoint}{self-adjointness}) gives, with the parallelogram law,
\[
4\operatorname{Re}\langle y, Tx\rangle \leq
M\bigl(\norm{x{+}y}^2 + \norm{x{-}y}^2\bigr) = 2M\bigl(\norm
x^2 + \norm y^2\bigr).
\]
For $\norm x = 1$ with $Tx \neq 0$, take $y = Tx/\norm{Tx}$:
$4\norm{Tx} \leq 4M$. So $\vertiii T \leq M$. Eigenvalues:
$Tx = \lambda x$, $x \ne 0$ gives $\lambda\norm x^2 = \langle
x, Tx\rangle \in \R$. Orthogonality: $\lambda\langle x,
y\rangle = \langle Tx, y\rangle = \langle x, Ty\rangle =
\mu\langle x, y\rangle$ with $\lambda \neq \mu$ real.
\end{proof}

\section{The spectral theorem}

\begin{lemma}[Existence of an extreme eigenvalue]
\label{lem:b3:spectral:existence}
Let $T \neq 0$ be \omterm{def:b3:spectral:compact}{compact} and \omterm{def:b3:spectral:selfadjoint}{self-adjoint}. Then $\vertiii T$
or $-\vertiii T$ is an eigenvalue of $T$.
\end{lemma}

\begin{proof}
By \cref{prop:b3:spectral:sanorm}, pick unit vectors $x_n$
with $\langle x_n, Tx_n\rangle \to \mu$, where $\abs\mu =
\vertiii T > 0$ (pass to a subsequence to fix the sign). Then
\[
\norm{Tx_n - \mu x_n}^2
= \norm{Tx_n}^2 - 2\mu\langle x_n, Tx_n\rangle + \mu^2
\leq \vertiii T^2 - 2\mu\langle x_n, Tx_n\rangle + \mu^2
\longrightarrow 2\mu^2 - 2\mu\cdot\mu = 0 .
\]
By compactness, a subsequence $Tx_{n_k} \to y$; then $\mu
x_{n_k} = Tx_{n_k} - (Tx_{n_k} - \mu x_{n_k}) \to y$, so
$x_{n_k} \to x = y/\mu$, a unit vector, and \omterm{def:b3:topology:continuity}{continuity} gives
$Tx = \mu x$.
\end{proof}

\begin{theorem}[Spectral theorem for compact self-adjoint
operators]\label{thm:b3:spectral:spectral}
Let $T$ be a \omterm{def:b3:spectral:compact}{compact} \omterm{def:b3:spectral:selfadjoint}{self-adjoint operator} on a \omterm{def:b3:hilbert:inner}{Hilbert space}
$H$.\index{spectral theorem (compact self-adjoint)}
\begin{enumerate}
  \item $H$ admits an orthonormal system $(e_n)_{n \in N}$
        ($N$ finite or countable) of eigenvectors of $T$, with
        real nonzero eigenvalues $(\lambda_n)$, such that
        \[
        Tx = \sum_{n\in N}\lambda_n\,\langle e_n, x\rangle\,
        e_n \qquad (x \in H),
        \]
        and $H = \ker T \,\oplus^\perp\,
        \overline{\operatorname{Vect}}(e_n : n \in N)$.
  \item If $N$ is infinite, $\lambda_n \to 0$; for each
        $\delta > 0$ only finitely many $n$ have
        $\abs{\lambda_n} \geq \delta$, and each eigenspace
        $\ker(T - \lambda)$, $\lambda \neq 0$, is
        finite-dimensional.
  \item Completing $(e_n)$ by an orthonormal basis of $\ker T$
        yields, when $H$ is separable, an orthonormal basis of
        $H$ made of eigenvectors: $T$ is diagonalized.
\end{enumerate}
\end{theorem}

\begin{proof}
(2) first. If infinitely many orthonormal eigenvectors $x_k$
had $\abs{\lambda_{(k)}} \geq \delta$: $\norm{Tx_k - Tx_l}^2 =
\lambda_{(k)}^2 + \lambda_{(l)}^2 \geq 2\delta^2$
(orthogonality, Pythagoras): no convergent subsequence of
$(Tx_k)$, contradicting compactness of $T$ on the bounded
$(x_k)$. This bounds by a finite number, for each $\delta$,
the total multiplicity of eigenvalues outside
$\intoo{-\delta}\delta$; countability and $\lambda_n \to 0$
follow.

(1) Let $H_0$ be the closed span of \emph{all} eigenvectors
with nonzero eigenvalues, organized (by (2) and
Gram--Schmidt within each finite-dimensional eigenspace,
orthogonality across eigenspaces from
\cref{prop:b3:spectral:sanorm}) into an orthonormal system
$(e_n)$ with eigenvalues $\lambda_n \neq 0$. $T$ maps $H_0$
into $H_0$, and also $H_0^\perp$ into $H_0^\perp$: for $y
\perp H_0$ and $e$ an eigenvector, $\langle e, Ty\rangle =
\langle Te, y\rangle = \lambda\langle e, y\rangle = 0$. The
restriction $T' = T\restriction_{H_0^\perp}$ is \omterm{def:b3:spectral:compact}{compact}
\omterm{def:b3:spectral:selfadjoint}{self-adjoint} on the \omterm{def:b3:hilbert:inner}{Hilbert space} $H_0^\perp$; if $T' \neq
0$, \cref{lem:b3:spectral:existence} produces an eigenvector
of $T$ with nonzero eigenvalue inside $H_0^\perp$ ---
impossible, such vectors live in $H_0$. So $T' = 0$:
$H_0^\perp \subseteq \ker T$. Conversely $\ker T \perp$ every
$e_n$ ($\langle e_n, z\rangle = \frac1{\lambda_n}\langle
Te_n, z\rangle = \frac1{\lambda_n}\langle e_n, Tz\rangle =
0$): $\ker T \subseteq H_0^\perp$, whence $\ker T =
H_0^\perp$ and the orthogonal decomposition. The expansion:
for $x = z + \sum_nc_ne_n$ ($z \in \ker T$, $c_n = \langle
e_n, x\rangle$; \cref{thm:b3:hilbert:parseval}(1) on $H_0$),
\omterm{def:b3:topology:continuity}{continuity} of $T$ gives $Tx = \sum_nc_n\lambda_ne_n$.

(3) $\ker T$, a closed subspace of a separable space, is
separable: it has an orthonormal basis
(\cref{prop:b3:hilbert:gramschmidt}); the union is an
orthonormal basis of $H$ by the decomposition in (1).
\end{proof}

\begin{theorem}[Fredholm alternative]
\label{thm:b3:spectral:fredholm}
Let $T$ be \omterm{def:b3:spectral:compact}{compact} \omterm{def:b3:spectral:selfadjoint}{self-adjoint} and $\lambda \in \R\setminus
\{0\}$.\index{Fredholm alternative}
\begin{enumerate}
  \item If $\lambda$ is not an eigenvalue, then $T - \lambda I$
        is bijective with bounded inverse: for every $f$, the
        equation $Tx - \lambda x = f$ has exactly one
        solution, depending \omterm{def:b3:topology:continuity}{continuously} on $f$.
  \item If $\lambda$ is an eigenvalue, $Tx - \lambda x = f$ is
        \omterm{def:b3:groups:derived}{solvable} iff $f \perp \ker(T - \lambda I)$, and the
        solution is unique up to that (finite-dimensional)
        kernel.
\end{enumerate}
\end{theorem}

\begin{proof}
Decompose $x = z + \sum c_ne_n$ and $f = w + \sum d_ne_n$
along \cref{thm:b3:spectral:spectral} ($z, w \in \ker T$). The
equation reads
\[
-\lambda z = w, \qquad (\lambda_n - \lambda)\,c_n = d_n\
(n \in N).
\]
(1) $\lambda \notin \{\lambda_n\}\cup\{0\}$: by (2) of the
spectral theorem, $\inf_n\abs{\lambda_n - \lambda} = \delta >
0$ (eigenvalues accumulate only at $0 \neq \lambda$). Solve:
$z = -w/\lambda$, $c_n = d_n/(\lambda_n - \lambda)$, with
$\sum\abs{c_n}^2 \leq \delta^{-2}\sum\abs{d_n}^2$: a unique
solution with $\norm x \leq C\norm f$.
(2) $\lambda = \lambda_{n}$ for $n$ in a finite set $F$:
\omterm{def:b3:groups:derived}{solvability} of $(\lambda_n - \lambda)c_n = d_n$ for $n \in F$
requires $d_n = 0$, i.e.\ $f \perp e_n$ ($n \in F$), i.e.\ $f
\perp \ker(T - \lambda I)$; the $c_n$, $n \in F$, are then
free.
\end{proof}

\begin{example}\label{ex:b3:spectral:minxy}
On $L^2(\intcc01)$, let $Tf(x) = \int_0^1\min(x, y)f(y)\dd y$:
a \omterm{ex:b3:spectral:examples}{Hilbert--Schmidt operator} with real symmetric kernel:
\omterm{def:b3:spectral:compact}{compact} and \omterm{def:b3:spectral:selfadjoint}{self-adjoint}. Solving $Tf = \lambda f$: the
relation $\bigl(Tf\bigr)(x) = \int_0^xyf(y)\dd y +
x\int_x^1f(y)\dd y$ shows $u = Tf$ satisfies $u'' = -f$ (two
differentiations, legitimate for \omterm{def:b3:topology:continuity}{continuous} $f$, and $Tf$ is
\omterm{def:b3:topology:continuity}{continuous} for $f \in L^2$: dominated convergence), with $u(0)
= 0$ and $u'(1) = 0$. So eigenfunctions solve $\lambda u'' =
-u$, $u(0) = 0$, $u'(1) = 0$:
\[
u_n(x) = \sin\Bigl(\bigl(n + \tfrac12\bigr)\pi x\Bigr),
\qquad
\lambda_n = \frac{1}{\bigl(n + \frac12\bigr)^2\pi^2}
\quad (n \geq 0),
\]
and the spectral theorem asserts --- with no Fourier theory ---
that these sines form an orthonormal basis of $L^2(\intcc01)$
after normalization (the kernel of $T$ is $0$: $Tf = 0$
forces, by the two differentiations, $f = 0$ a.e.). The
weekend problem runs the same circle of ideas for the
vibrating string and extracts $\zeta(2)$ from the trace.
\end{example}

\begin{method}\label{met:b3:spectral:method}
Given an integral or differential equation: (1) recast it as
$(I - \lambda K)u = f$ or $Ku = \lambda u$ with $K$ an
integral operator; (2) verify $K$ \omterm{def:b3:spectral:compact}{compact} (Hilbert--Schmidt
kernel, or Ascoli) and, if possible, \omterm{def:b3:spectral:selfadjoint}{self-adjoint} (symmetric
real kernel); (3) diagonalize with the spectral theorem or
invoke the \omterm{thm:b3:spectral:fredholm}{Fredholm alternative} for \omterm{def:b3:groups:derived}{solvability}; (4) read off
existence, uniqueness, stability, and series formulas for
solutions in the eigenbasis. Differential operators are
unbounded, but their \emph{inverses} (Green operators) are
\omterm{def:b3:spectral:compact}{compact}: always invert first.
\end{method}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:b3:spectral:1}
(a) Show that a bounded operator with finite-dimensional range
is \omterm{def:b3:spectral:compact}{compact}.
(b) Show that the identity of a normed space is \omterm{def:b3:spectral:compact}{compact} iff
the dimension is finite (Riesz, Year 2). Deduce that a \omterm{def:b3:spectral:compact}{compact
operator} on an infinite-dimensional space is never invertible
with bounded inverse.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b3:spectral:2}
Let $T(x_1, x_2, \dots) = (\lambda_1x_1, \lambda_2x_2, \dots)$
on $\ell^2$, with $(\lambda_n)$ bounded.
(a) Show $\vertiii T = \sup\abs{\lambda_n}$.
(b) Show that $T$ is \omterm{def:b3:spectral:compact}{compact} iff $\lambda_n \to 0$.
\emph{(For $\Leftarrow$, truncate; for $\Rightarrow$, test on
$(e_n)$.)}
(c) When is $T$ \omterm{def:b3:spectral:selfadjoint}{self-adjoint}? Verify the spectral theorem by
inspection in that case.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:spectral:3}
Give the details of the \omterm{def:b3:rings:ideal}{ideal} property
(\cref{prop:b3:spectral:ideal}): if $S$ is \omterm{def:b3:spectral:compact}{compact} and $A, B$
bounded, then $ASB$ is \omterm{def:b3:spectral:compact}{compact}. Deduce that if $ST = TS = I$
for some bounded $S$, and $\dim H = \infty$, then $T$ is not
\omterm{def:b3:spectral:compact}{compact} --- and reconcile with \cref{exo:b3:spectral:1}(b).
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:spectral:4}
(Hilbert--Schmidt) Let $k \in L^2(\intcc01^2)$ and $(e_n)$ a
\omterm{def:b3:hilbert:onb}{Hilbert basis} of $L^2(\intcc01)$.
(a) Show that $\vertiii{T_k} \leq \norm k_{L^2}$
\emph{(Cauchy--Schwarz in the $y$-variable, then Tonelli)}.
(b) Expand $k(x,y) = \sum_{m,n}c_{mn}e_m(x)\overline{e_n(y)}$
in $L^2$ of the square (justify that the products form a
\omterm{def:b3:hilbert:onb}{Hilbert basis} there), and show that truncating the sum gives
finite-rank operators converging to $T_k$ in \omterm{def:b3:banach:operator}{operator norm}:
$T_k$ is \omterm{def:b3:spectral:compact}{compact}.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:spectral:5}
Let $T$ be \omterm{def:b3:spectral:selfadjoint}{self-adjoint} with $\langle x, Tx\rangle \geq 0$ for
all $x$ (\emph{positive} operator).
(a) Show that eigenvalues are $\geq 0$ and that $\vertiii T =
\sup_{\norm x\leq1}\langle x, Tx\rangle$.
(b) Prove the generalized Cauchy--Schwarz inequality
$\abs{\langle x, Ty\rangle}^2 \leq \langle x, Tx\rangle\langle
y, Ty\rangle$.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:spectral:6}
On $L^2(\intcc01)$, let $(Mf)(x) = x\,f(x)$.
(a) Show that $M$ is bounded, \omterm{def:b3:spectral:selfadjoint}{self-adjoint}, with $\vertiii M =
1$, but has \emph{no} eigenvalues.
(b) Show $M$ is not \omterm{def:b3:spectral:compact}{compact} \emph{(exhibit a bounded sequence
whose image has no convergent subsequence, e.g.\ normalized
indicators of shrinking intervals near $1$ --- or invoke the
spectral theorem)}.
(c) Where does \cref{lem:b3:spectral:existence}'s proof break
for $M$?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:spectral:7}
(Volterra) On $L^2(\intcc01)$, let $Vf(x) = \int_0^xf(y)\dd
y$.
(a) Show $V$ is \omterm{def:b3:spectral:compact}{compact} (Hilbert--Schmidt with kernel $\mathbf
1_{y < x}$) but not \omterm{def:b3:spectral:selfadjoint}{self-adjoint}; compute $V^*$.
(b) Show that $V$ has no nonzero eigenvalue. \emph{(From $Vf =
\lambda f$: $f$ has a \omterm{def:b3:topology:continuity}{continuous} representative, then is
$\mathcal C^1$, and solves $\lambda f' = f$, $f(0) = 0$.)}
(c) Conclude that compactness alone yields no eigenvectors:
\omterm{def:b3:spectral:selfadjoint}{self-adjointness} in \cref{thm:b3:spectral:spectral} is
essential.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b3:spectral:8}
(Courant--Fischer) Let $T$ be \omterm{def:b3:spectral:compact}{compact}, \omterm{def:b3:spectral:selfadjoint}{self-adjoint},
\emph{positive}, with eigenvalues $\mu_1 \geq \mu_2 \geq \dots
> 0$ (repeated by multiplicity, eigenvectors $e_1, e_2,
\dots$). Show:
\[
\mu_{k} = \max_{\substack{V \subseteq H \\ \dim V = k}}\
\min_{\substack{x \in V\\ \norm x = 1}}\ \langle x, Tx\rangle
= \min_{\substack{W \subseteq H\\ \operatorname{codim}W =
k-1}}\ \max_{\substack{x\in W\\ \norm x = 1}}\ \langle x,
Tx\rangle .
\]
\emph{(Test $V = \operatorname{Vect}(e_1,\dots,e_k)$; for the
upper bound intersect any $V$ with
$\operatorname{Vect}(e_k, e_{k+1}, \dots)$-type spaces:
dimension counting forces a nonzero intersection.)} Deduce
that eigenvalues depend monotonically on $T$ ($T \leq S
\Rightarrow \mu_k(T) \leq \mu_k(S)$).
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:spectral:9}
Using \cref{thm:b3:spectral:fredholm} for $Tf(x) =
\int_0^1\min(x,y)f(y)\dd y$ (\cref{ex:b3:spectral:minxy}):
for which $\lambda \in \R$ does the integral equation
\[
f(x) - \lambda\int_0^1\min(x,y)\,f(y)\,\dd y = g(x)
\]
have a unique solution $f \in L^2$ for every $g \in L^2$? What
happens at the exceptional values?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b3:spectral:10}
Let $S$ be the shift on $\ell^2$
(\cref{exo:b3:banach:1}).
(a) Show that $S$ has no eigenvalues, while every $\lambda$
with $\abs\lambda < 1$ is an eigenvalue of $S^*$ (find the
eigenvectors explicitly, geometric sequences).
(b) Neither $S$ nor $S^*$ is \omterm{def:b3:spectral:compact}{compact}: verify via
\cref{exo:b3:spectral:2}-style testing on $(e_n)$.
(c) Comment: for \omterm{def:b3:spectral:selfadjoint}{non-self-adjoint}, \omterm{def:b3:spectral:compact}{non-compact} operators, the
eigenvalue landscape can be anything from empty to a full
disc --- the notion that survives is the \emph{spectrum},
studied in a later course.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b3:spectral:11}
(Square roots) Let $T$ be \omterm{def:b3:spectral:compact}{compact}, \omterm{def:b3:spectral:selfadjoint}{self-adjoint}, positive
($\langle x, Tx\rangle \geq 0$) on a \omterm{def:b3:hilbert:inner}{Hilbert space} $H$, with
spectral decomposition $Tx = \sum_n\mu_n\langle e_n,
x\rangle e_n$ ($\mu_n > 0$).
(a) Define $Sx = \sum_n\sqrt{\mu_n}\,\langle e_n, x\rangle
e_n$; show $S$ is \omterm{def:b3:spectral:compact}{compact}, \omterm{def:b3:spectral:selfadjoint}{self-adjoint}, positive, with $S^2
= T$.
(b) Prove uniqueness: any \omterm{def:b3:spectral:compact}{compact} positive \omterm{def:b3:spectral:selfadjoint}{self-adjoint} $R$
with $R^2 = T$ preserves the eigenspaces of $T$
\emph{($RT = R^3 = TR$: $R$ commutes with $T$, so
$R(\ker(T - \mu)) \subseteq \ker(T - \mu)$)}, and on
$\ker(T - \mu)$, $R$ is a positive operator squaring to
$\mu\,\mathrm{id}$ on a finite-dimensional space:
diagonalize it there and conclude $R =
\sqrt\mu\,\mathrm{id}$ on each eigenspace, hence $R = S$.
(c) Compute $\sqrt G$ for the string operator $G$ of
\cref{pb:b3:spectral:1}: which kernel has eigenvalues
$\frac1{n\pi}$ on the sine basis? (Express $\sqrt G$ as the
$L^2$-limit of kernels; no closed form is required.)
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b3:spectral:12}
(Singular value decomposition) Let $T \in \mathcal L(H)$ be
\emph{\omterm{def:b3:spectral:compact}{compact}}, not necessarily \omterm{def:b3:spectral:selfadjoint}{self-adjoint}.
(a) Show that $T^*T$ is \omterm{def:b3:spectral:compact}{compact}, \omterm{def:b3:spectral:selfadjoint}{self-adjoint}, positive; let
$(e_n)$ be an orthonormal family of eigenvectors with
$T^*Te_n = s_n^2e_n$, $s_n > 0$ (the \emph{singular values}),
completed by $\ker(T^*T) = \ker T$ (prove this equality).
(b) Set $f_n = \frac{Te_n}{s_n}$; show $(f_n)$ is
orthonormal, and establish the \emph{SVD}:
\[
Tx = \sum_n s_n\,\langle e_n, x\rangle\,f_n
\qquad (x \in H),
\]
with convergence in $H$.
(c) Deduce: $\vertiii T = \max_ns_n$; $T$ is a norm-limit of
finite-rank operators (re-proving
\cref{prop:b3:spectral:ideal}'s converse for \omterm{def:b3:hilbert:inner}{Hilbert
spaces}); and for the Volterra operator $V$ of
\cref{exo:b3:spectral:7}, which has no eigenvalues, explain
why the SVD nevertheless exists and what its ingredients are
(identify $V^*V$ as a string-type kernel operator ---
computing its eigenvalues explicitly is
\cref{exo:b3:spectral:9}'s territory).
\end{exercise}

\section{Problem: the vibrating string and
\texorpdfstring{$\zeta(2)$}{zeta(2)}}

\begin{problem}[{Weekend problem --- Green's operator,
the sine basis, and a trace formula}]\label{pb:b3:spectral:1}
We solve the eigenvalue problem of the vibrating string with
fixed ends --- $-u'' = \nu u$, $u(0) = u(1) = 0$ --- by
operator theory, obtain the sine orthonormal basis without any
Fourier computation, and evaluate $\zeta(2)$ by comparing two
expressions for the \emph{trace} of the Green operator.
Define, on $L^2(\intcc01)$,
\[
(Gf)(x) = \int_0^1 g(x,y)\,f(y)\,\dd y,
\qquad
g(x, y) = \min(x,y)\,\bigl(1 - \max(x,y)\bigr).
\]

\medskip
\noindent\textbf{Part I --- The Green operator.}
\begin{enumerate}
  \item Show that $g$ is \omterm{def:b3:topology:continuity}{continuous}, symmetric, with $0 \leq g
        \leq \frac14$, and that $G$ is \omterm{def:b3:spectral:compact}{compact} and
        \omterm{def:b3:spectral:selfadjoint}{self-adjoint} (\cref{ex:b3:spectral:examples}(c)).
  \item For $f$ \omterm{def:b3:topology:continuity}{continuous}, show that $u = Gf$ is $\mathcal
        C^2$ with
        \[
        -u'' = f, \qquad u(0) = u(1) = 0
        \]
        \emph{(write $u(x) = (1-x)\int_0^xyf(y)\dd y +
        x\int_x^1(1-y)f(y)\dd y$ and differentiate twice)}.
        Conversely, if $u \in \mathcal C^2$ with $u(0) = u(1)
        = 0$, then $G(-u'') = u$: $G$ inverts the string
        operator.
  \item Show $\ker G = \{0\}$ \emph{(if $Gf = 0$ with $f \in
        L^2$: test against \omterm{def:b3:topology:continuity}{continuous} $\varphi$, transfer $G$
        by symmetry/Fubini onto $\varphi$, and use the
        fundamental lemma \cref{cor:b3:lp:fundlemma} --- or
        regularize)}, and that $G$ is a positive operator:
        $\langle f, Gf\rangle \geq 0$. \emph{(For \omterm{def:b3:topology:continuity}{continuous}
        $f$: $\langle f, Gf\rangle = \int_0^1 (u')^2$ with $u
        = Gf$, by parts; conclude by \omterm{ex:b3:lebesgue:gamma}{density}.)}
\end{enumerate}

\medskip
\noindent\textbf{Part II --- Diagonalization: the sine
basis.}
\begin{enumerate}[resume]
  \item Show that the eigenfunctions of $G$ with eigenvalue
        $\lambda \ne 0$ are, up to scalars, the solutions of
        $-\lambda u'' = u$, $u(0) = u(1) = 0$ \emph{(an
        eigenfunction has a \omterm{def:b3:topology:continuity}{continuous} representative --- $Gf$
        is \omterm{def:b3:topology:continuity}{continuous} for $f \in L^2$, why? --- hence is
        $\mathcal C^2$ by bootstrapping question 2)}.
  \item Solve the boundary value problem: the eigenvalues of
        $G$ are $\lambda_n = \frac1{n^2\pi^2}$ ($n \geq 1$),
        with normalized eigenfunctions $e_n(x) =
        \sqrt2\,\sin(n\pi x)$; check orthonormality by direct
        integration as a sanity test.
  \item Conclude from
        \cref{thm:b3:spectral:spectral} and question 3 that
        $\bigl(\sqrt2\sin(n\pi x)\bigr)_{n\geq1}$ is an
        orthonormal \emph{basis} of $L^2(\intcc01)$ --- no
        Stone--Weierstrass, no Fourier series needed. Expand
        $f(x) = x(1-x)$ in this basis and write \omterm{thm:b3:hilbert:parseval}{Parseval} for
        it.
\end{enumerate}

\medskip
\noindent\textbf{Part III --- The trace formula and
$\zeta(2)$.}
\begin{enumerate}[resume]
  \item Prove the two identities
        \[
        \langle e_n, Ge_n\rangle = \lambda_n
        \quad\text{and}\quad
        \sum_{n\geq1}\lambda_n = \int_0^1 g(x,x)\,\dd x .
        \]
        For the second (the \emph{trace formula}): expand
        $g(x, \cdot)$, for fixed $x$, in the basis $(e_n)$ ---
        show that the coefficients are $\lambda_ne_n(x)$, so
        that $g(x, \cdot) = \sum_n\lambda_ne_n(x)\,e_n$ in
        $L^2$. Here the sines are explicit: verify directly
        that $\sum_n\lambda_ne_n(x)e_n(y)$ converges
        \emph{uniformly} on the square (compare with $\sum
        \frac2{n^2\pi^2}$), so its sum is \omterm{def:b3:topology:continuity}{continuous} and, having
        the same $L^2$-expansions in $y$ for each $x$, equals
        $g(x,y)$ everywhere. Set $y = x$ and integrate term by
        term.
  \item Compute $\int_0^1g(x,x)\dd x = \int_0^1x(1-x)\dd x =
        \frac16$, and conclude
        \[
        \sum_{n\geq1}\frac{1}{n^2\pi^2} = \frac16,
        \qquad\text{i.e.}\qquad
        \boxed{\ \zeta(2) = \frac{\pi^2}{6}\ } :
        \]
        Euler's sum from an operator trace.
  \item Re-derive $\zeta(2)$ a third way: apply \omterm{thm:b3:hilbert:parseval}{Parseval} in
        the sine basis to the constant function $\mathbf 1$,
        compute $\sum_{n \text{ odd}}\frac1{n^2}$, and
        conclude. Then compare mechanisms: in what sense is
        the trace argument of questions 7--8 ``\omterm{thm:b3:hilbert:parseval}{Parseval}
        applied to the whole kernel at once''?
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- The string vibrates.}
\begin{enumerate}[resume]
  \item (Separation of variables, synthesized) For $f \in
        L^2$, define
        \[
        u(t, x) = \sum_{n\geq1}\;c_n\,
        \cos(n\pi t)\,\sqrt2\sin(n\pi x),
        \qquad c_n = \langle e_n, f\rangle .
        \]
        Show the series converges in $L^2(\intcc01)$ for each
        $t$, that $t\mapsto u(t, \cdot)$ is \omterm{def:b3:topology:continuity}{continuous} into
        $L^2$, and that for $f$ in the span of finitely many
        $e_n$ it solves the wave equation $\partial_t^2u =
        \partial_x^2u$ with $u(0) = f$, $\partial_tu(0) = 0$,
        fixed ends. The eigenvalues $n^2\pi^2$ are the squared
        frequencies: the string's harmonics --- explain the
        musical interpretation of
        \cref{thm:b3:spectral:spectral} in one paragraph.
\end{enumerate}

\medskip
\noindent\textbf{Part V --- Variational dividends: the power
method, Weyl stability, and a rigorous bound on $\pi$.} Let
$A$ be a \omterm{def:b3:spectral:compact}{compact} \omterm{def:b3:spectral:selfadjoint}{self-adjoint} \emph{positive} operator with
eigenvalues $\mu_1 \geq \mu_2 \geq \cdots > 0$ and
orthonormal eigenvectors $(u_n)$; $R(x) = \frac{\langle x,
Ax\rangle}{\norm x^2}$. The min--max formulas are
\cref{exo:b3:spectral:8}; here we spend them.
\begin{enumerate}[resume]
  \item (Power method) For $x \neq 0$ write $m_p =
        \sum_n\mu_n^p\abs{\langle u_n, x\rangle}^2$. Show
        $m_pm_{p+2} \geq m_{p+1}^2$ (Cauchy--Schwarz), deduce
        the chain
        \[
        R(x) \;\leq\; \frac{\langle Ax, Ax\rangle}{\langle x,
        Ax\rangle} \;\leq\; R(Ax) \;\leq\; \mu_1,
        \]
        and prove that if $\langle u_1, x\rangle \neq 0$,
        then $R(A^kx) \to \mu_1$: iterating the operator on
        any generic vector computes the top eigenvalue ---
        the power method of numerical analysis, certified.
  \item (Weyl stability) For \omterm{def:b3:spectral:compact}{compact} \omterm{def:b3:spectral:selfadjoint}{self-adjoint} positive
        $A, B$, deduce from \cref{exo:b3:spectral:8} that
        \[
        \abs{\mu_n(A) - \mu_n(B)} \;\leq\; \vertiii{A - B}
        \qquad\text{for every } n :
        \]
        the full spectrum is $1$-Lipschitz in the \omterm{def:b3:banach:operator}{operator
        norm} --- eigenvalues of big symmetric systems can be
        computed from approximations with guaranteed error.
  \item Apply the Rayleigh bound to $G$ with the test
        function $u(x) = x(1-x)$: solve $-w'' = u$, $w(0) =
        w(1) = 0$ to get $Gu = w = \frac{x(1-x)(1 + x -
        x^2)}{12}$, compute
        \[
        \norm u_2^2 = \frac1{30}, \qquad
        \langle u, Gu\rangle = \frac1{12}\Bigl(\frac1{30} +
        \frac1{140}\Bigr) = \frac{17}{5040},
        \qquad R(u) = \frac{17}{168},
        \]
        and conclude the \emph{rigorous} bound $\frac1{\pi^2}
        = \lambda_1 \geq \frac{17}{168}$, i.e.\ $\pi \leq
        \sqrt{168/17} < 3.1437$.
  \item One step of question 11's chain, on the same test
        function: using $\int_0^1(x - x^2)^4\dd x =
        \frac1{630}$, compute
        \[
        \norm{Gu}_2^2 = \frac1{144}\Bigl(\frac1{30} +
        \frac2{140} + \frac1{630}\Bigr) = \frac{31}{90720},
        \qquad
        \frac{\langle Gu, Gu\rangle}{\langle u, Gu\rangle} =
        \frac{31}{306},
        \]
        and conclude $\pi \leq \sqrt{306/31} < 3.1419$: two
        integrals, four correct digits. (Each further
        iteration roughly squares the accuracy: the
        eigenvector gap $\lambda_1/\lambda_2 = 4$ drives
        geometric convergence.)
\end{enumerate}

\medskip
\noindent\textbf{Part VI --- The trace of $G^2$, and
$\zeta(4)$.}
\begin{enumerate}[resume]
  \item Show that $\sum_n\lambda_n^2 =
        \iint_{\intcc01^2}g(x,y)^2\,\dd x\,\dd y$
        \emph{(expand $g$ on the product basis
        $(e_m(x)e_n(y))_{m,n}$ of $L^2(\intcc01^2)$ --- a
        \omterm{def:b3:hilbert:onb}{Hilbert basis}, cf.\ \cref{exo:b3:spectral:5} --- and
        apply \omterm{thm:b3:hilbert:parseval}{Parseval} on the square; question 7 identifies
        the coefficients)}.
  \item Compute the double integral:
        \[
        \iint g^2 = 2\int_0^1(1-x)^2\Bigl(\int_0^x
        y^2\,\dd y\Bigr)\dd x
        = \frac23\int_0^1x^3(1-x)^2\,\dd x = \frac1{90} .
        \]
  \item Conclude $\zeta(4) = \dfrac{\pi^4}{90}$; explain,
        without computation, how traces of higher powers
        $G^k$ produce $\zeta(2k) \in \pi^{2k}\,\Q$ for all
        $k \geq 1$, and why the odd values $\zeta(3),
        \zeta(5), \dots$ are structurally out of this
        machine's reach.
  \item ($\pi$ from below) From $\lambda_1^2 \leq
        \sum_n\lambda_n^2 = \frac1{90}$ deduce $\pi \geq
        90^{1/4} > 3.080$, and assemble with question 14 the
        two-sided verdict
        \[
        3.080 \;<\; \pi \;<\; 3.1419,
        \]
        obtained entirely from the vibrating string's
        arithmetic. Which side converges faster if one uses
        higher traces $(\operatorname{tr}G^{2k})^{-1/4k}$,
        and why?
\end{enumerate}

\medskip
\noindent\textbf{Part VII --- Forcing and resonance.} Fix
$\nu \in \R$ and consider the forced string $-u'' - \nu u =
f$, $u(0) = u(1) = 0$, with $f \in L^2$ and $c_n = \langle
e_n, f\rangle$.
\begin{enumerate}[resume]
  \item Suppose $\nu \notin \{n^2\pi^2 : n \geq 1\}$. Show
        that
        \[
        u = \sum_{n\geq1}\frac{c_n}{n^2\pi^2 - \nu}\,e_n
        \]
        converges in $L^2$ and uniformly on $\intcc01$
        \emph{(Cauchy--Schwarz between $(c_n)$ and the tails
        of $\sum n^{-4}$, with $\norm{e_n}_\infty =
        \sqrt2$)}, and that it satisfies $u = Gf + \nu Gu$
        --- the explicit-coordinates form of the \omterm{thm:b3:spectral:fredholm}{Fredholm
        alternative} (\cref{thm:b3:spectral:fredholm}), with
        uniqueness.
  \item Suppose $\nu = m^2\pi^2$. Show that $u = Gf + \nu
        Gu$ has a solution iff $c_m = 0$, unique up to
        adding multiples of $e_m$. Physical reading: pushing
        a swing exactly at its own frequency.
  \item For $\nu < \pi^2$, show that the solution operator
        $R_\nu\colon f \mapsto u$ is bounded on $L^2$ with
        norm $\frac1{\pi^2 - \nu}$, \omterm{def:b3:spectral:compact}{compact}, \omterm{def:b3:spectral:selfadjoint}{self-adjoint},
        and positive: the whole spectral analysis restarts,
        shifted by $\nu$.
  \item (Synthesis) Compile the dictionary of this problem:
        eigenvalue $\leftrightarrow$ squared frequency
        (harmonics); trace $\leftrightarrow$ $\zeta(2)$;
        Hilbert--Schmidt norm $\leftrightarrow$ $\zeta(4)$;
        \omterm{thm:b3:spectral:fredholm}{Fredholm alternative} $\leftrightarrow$ resonance;
        min--max $\leftrightarrow$ variational bounds ($\pi <
        3.1437$ from one polynomial). One integral operator,
        five chapters of analysis cashed in.
\end{enumerate}

\medskip
\noindent\textbf{Part VIII --- Three last echoes.}
\begin{enumerate}[resume]
  \item ($\zeta(6)$, for free) Question 6's \omterm{thm:b3:hilbert:parseval}{Parseval}
        identity gave $\sum_{n\text{ odd}}n^{-6} =
        \frac{\pi^6}{960}$. Split $\zeta(6)$ into odd and
        even $n$ and conclude
        \[
        \zeta(6) = \frac{\pi^6}{945},
        \]
        with no new integral: the machine of question 17
        (traces of $G^3$) would have produced the same value
        at the cost of an iterated kernel --- \omterm{thm:b3:hilbert:parseval}{Parseval} on one
        well-chosen function is the cheaper route here.
  \item (The ground state is positive) Let $A$ be a \omterm{def:b3:spectral:compact}{compact}
        \omterm{def:b3:spectral:selfadjoint}{self-adjoint} positive operator on $L^2(\intcc01)$
        given by a \omterm{def:b3:topology:continuity}{continuous} symmetric kernel $k > 0$ on
        $\intoo01^2$, with largest eigenvalue $\mu_1$. Show:
        (a) any maximizer of the Rayleigh quotient is a
        $\mu_1$-eigenfunction; (b) if $u$ is one, then
        $\langle\abs u, A\abs u\rangle \geq \langle u,
        Au\rangle$, with \emph{strict} inequality if $u$ takes
        both signs on sets of positive \omterm{def:b3:measure:measure}{measure} --- hence $u$
        has a.e.\ constant sign, and $u = \mu_1^{-1}Au$
        never vanishes on $\intoo01$; (c) $\mu_1$ is a
        \emph{simple} eigenvalue. Verify every claim on $G$:
        $e_1 = \sqrt2\sin(\pi x) > 0$, and each $e_n$, $n
        \geq 2$, being orthogonal to $e_1$, must change sign
        (it does: $n - 1$ \omterm{def:b3:topology:interior}{interior} zeros).
  \item (Distance to the spectrum, and the price of
        resonance) For $\nu \notin \{n^2\pi^2\}$, show that
        the solution operator $R_\nu$ of question 19 is
        bounded, \omterm{def:b3:spectral:selfadjoint}{self-adjoint}, \omterm{def:b3:spectral:compact}{compact}, with
        \[
        \vertiii{R_\nu} =
        \frac1{\min_{n\geq1}\,\abs{n^2\pi^2 - \nu}}
        = \frac1{\operatorname{dist}\bigl(\nu,
        \{n^2\pi^2\}\bigr)},
        \]
        the norm being attained on the nearest mode. Then
        quantify question 20's swing: forcing with $f = e_1$
        at $\nu = (1 - \varepsilon)\pi^2$ produces $u =
        \frac{1}{\varepsilon\pi^2}\,e_1$, an amplification by
        $\frac1\varepsilon$ of the static response $Ge_1 =
        \frac1{\pi^2}e_1$ --- at one percent below the
        fundamental ($\varepsilon = 10^{-2}$), the string
        answers a hundred times louder.
\end{enumerate}
\end{problem}
