\chapter{Areas and Volumes}\label{ch:g7:areas}

Grade 6 measured rectangles, \omterm{def:g6:shapes:triangles}{right triangles} and boxes
(\cref{ch:g6:measure}). With a pair of scissors --- cut a piece here,
glue it there --- the same ideas now give the \omterm{def:g6:measure:area}{areas} of \omterm{def:g7:central:parallelogram}{parallelograms},
of \emph{all} triangles and of the disk, and the \omterm{def:g6:measure:volume}{volumes} of \omterm{def:g7:areas:prism}{prisms} and
\omterm{def:g7:areas:prism}{cylinders}.

\section{Area of a parallelogram}

\begin{theorem}[Parallelogram]\label{thm:g7:areas:parallelogram}
A \omterm{def:g7:central:parallelogram}{parallelogram} with a side of length $b$ (its \emph{base}) and
distance $h$ between that side and the opposite one (its
\emph{height}) has \omterm{def:g6:measure:area}{area}
\[
A = b \times h .
\]
\end{theorem}

\begin{proof}[Proof by scissors]
Cut off the \omterm{def:g6:shapes:triangles}{right-angled} triangle sticking out on one side and glue it
on the other: the \omterm{def:g7:central:parallelogram}{parallelogram} becomes a $b \times h$ rectangle with
the same \omterm{def:g6:measure:area}{area}.
\end{proof}

\begin{omfigure}
\begin{tikzpicture}[scale=0.82]
  \fill[omDef!14] (0,0) -- (4,0) -- (5.2,1.8) -- (1.2,1.8) -- cycle;
  \draw[thick] (0,0) -- (4,0) -- (5.2,1.8) -- (1.2,1.8) -- cycle;
  \fill[omThm!25] (0,0) -- (1.2,1.8) -- (1.2,0) -- cycle;
  \draw[omThm, thick] (0,0) -- (1.2,1.8) -- (1.2,0) -- cycle;
  \draw[dashed] (1.2,0) -- (1.2,1.8);
  \node[below, font=\small] at (2.0,0) {$b$};
  \draw[Stealth-Stealth] (6.0,0) -- (6.0,1.8)
    node[midway, right, font=\small] {$h$};
  \begin{scope}[xshift=8.2cm]
  \fill[omDef!14] (0,0) rectangle (4,1.8);
  \draw[thick] (0,0) rectangle (4,1.8);
  \fill[omThm!25] (2.8,0) -- (4,1.8) -- (4,0) -- cycle;
  \draw[omThm, thick] (2.8,0) -- (4,1.8) -- (4,0) -- cycle;
  \node[below, font=\small] at (2.0,0) {$b$};
  \end{scope}
\end{tikzpicture}

{\small The scissors proof: sliding the red triangle from the left end
to the right end turns the \omterm{def:g7:central:parallelogram}{parallelogram} into a rectangle --- same
\omterm{def:g6:measure:area}{area}, $b \times h$.}
\end{omfigure}

\begin{remark}
The height is the \emph{\omterm{def:g6:lines:perp}{perpendicular}} distance between the two bases,
not the length of the slanted side! A very slanted \omterm{def:g7:central:parallelogram}{parallelogram} can
have long sides and a small \omterm{def:g6:measure:area}{area}.
\end{remark}

\section{Area of a triangle}

\begin{theorem}[Triangle]\label{thm:g7:areas:triangle}
A triangle with base $b$ and corresponding height $h$ (the
\omterm{def:g6:lines:perp}{perpendicular} distance from the opposite \omterm{def:g5:solids:def}{vertex} to the base line) has
\omterm{def:g6:measure:area}{area}
\[
A = \frac{b \times h}{2} .
\]
\end{theorem}

\begin{proof}
Two copies of the triangle, one turned by a half-turn around the
midpoint of one side, fit together into a \omterm{def:g7:central:parallelogram}{parallelogram} with base $b$
and height $h$ (that is \omterm{def:g7:central:def}{central symmetry} at work,
\cref{prop:g7:central:preserved}). The triangle is \omterm{def:g3:division:half}{half} of it:
$\frac{bh}{2}$.
\end{proof}

\begin{omfigure}
\begin{tikzpicture}[scale=0.9]
  \fill[omDef!14] (0,0) -- (4,0) -- (1.4,2.1) -- cycle;
  \draw[thick] (0,0) -- (4,0) -- (1.4,2.1) -- cycle;
  \fill[omThm!12] (4,0) -- (1.4,2.1) -- (5.4,2.1) -- cycle;
  \draw[thick, omThm, dashed] (4,0) -- (5.4,2.1) -- (1.4,2.1);
  \draw[dashed] (1.4,0) -- (1.4,2.1);
  \draw[thin] (1.4,0) ++(-0.2,0) -- ++(0,0.2) -- ++(0.2,0);
  \node[below, font=\small] at (2,0) {$b$};
  \node[right, font=\small] at (1.42,1.0) {$h$};
\end{tikzpicture}

{\small A triangle and its half-turned copy (dashed) make a
\omterm{def:g7:central:parallelogram}{parallelogram}: the triangle's \omterm{def:g6:measure:area}{area} is $\frac{b\times h}{2}$. Any of the
three sides can serve as the base --- with \emph{its own} height.}
\end{omfigure}

\begin{example}\label{ex:g7:areas:triangle}
A triangle has a base of $9$ cm and the corresponding height measures
$4$ cm:
\[
A = \frac{9 \times 4}{2} = \frac{36}{2} = 18 \text{ cm}^2 .
\]
For a \emph{right} triangle, the two legs are a base and its height:
we recover the formula of \cref{prop:g6:measure:formulas}.
\end{example}

\section{Area of a disk}

\begin{theorem}[Disk]\label{thm:g7:areas:disk}
A disk of \omterm{def:g3:shapes:shapes}{radius} $r$ has \omterm{def:g6:measure:area}{area}
\[
A = \pi r^2 .
\]
\end{theorem}

\begin{proof}[Idea of proof]
Cut the disk into many thin equal slices and lay them head to tail:
they almost form a \omterm{def:g7:central:parallelogram}{parallelogram} of height $r$ whose base is \omterm{def:g3:division:half}{half} the
circumference, $\pi r$ (\cref{prop:g6:measure:circle}). Its \omterm{def:g6:measure:area}{area} is
close to $\pi r \times r = \pi r^2$, and the approximation becomes
perfect as the slices get thinner. A complete proof needs integral
calculus, from the High School volume.
\end{proof}

\begin{omfigure}
\begin{tikzpicture}[scale=0.85]
  \foreach \a in {0,45,...,315}
    \fill[omDef!20, draw=omDef] (0,0) -- (\a:1.4)
      arc (\a:\a+45:1.4) -- cycle;
  \begin{scope}[xshift=4.3cm, yshift=-0.15cm]
  \foreach \i in {0,1,2,3} {
    \fill[omDef!20, draw=omDef]
      ({\i*1.1},0) -- ({\i*1.1+0.55},1.4) -- ({\i*1.1+1.1},0) -- cycle;
    \fill[omDef!20, draw=omDef]
      ({\i*1.1+0.55},1.4) -- ({\i*1.1+1.1},0) -- ({\i*1.1+1.65},1.4)
      -- cycle;
  }
  \draw[Stealth-Stealth] (5.2,0) -- (5.2,1.4)
    node[midway, right, font=\small] {$\approx r$};
  \node[below, font=\small] at (2.5,-0.1) {base $\approx \pi r$};
  \end{scope}
\end{tikzpicture}

{\small Slicing the disk and unrolling the slices: almost a
\omterm{def:g7:central:parallelogram}{parallelogram} of base $\pi r$ (\omterm{def:g3:division:half}{half} the border) and height $r$.}
\end{omfigure}

\begin{example}\label{ex:g7:areas:disk}
A round table top has \omterm{def:g3:shapes:shapes}{radius} $60$ cm. Its \omterm{def:g6:measure:area}{area} is
$\pi \times 60^2 = 3600\pi \approx 11\,310$ cm$^2$, a bit more than
$1.1$ m$^2$. Watch the \omterm{ex:g1:subtraction:difference}{difference}: the \emph{\omterm{def:g6:measure:perimeter}{perimeter}} $2\pi r$ uses
$r$, the \emph{\omterm{def:g6:measure:area}{area}} $\pi r^2$ uses $r^2$.
\end{example}

\section{Prisms and cylinders}

\begin{definition}[Prism, cylinder]\label{def:g7:areas:prism}
A \emph{prism}\index{prism} is a \omterm{def:g5:solids:def}{solid} with two identical \omterm{def:g6:lines:perp}{parallel}
polygonal \omterm{def:g5:solids:def}{faces} (the \emph{bases}) joined by rectangles; a
\emph{cylinder}\index{cylinder} is the same with disks as bases. The
\emph{height} is the distance between the two bases.
\end{definition}

\begin{theorem}[Volumes]\label{thm:g7:areas:volume}
For a \omterm{def:g7:areas:prism}{prism} or a \omterm{def:g7:areas:prism}{cylinder} with base \omterm{def:g6:measure:area}{area} $B$ and height $h$:
\[
V = B \times h .
\]
\end{theorem}
\admitted

\begin{example}\label{ex:g7:areas:volume}
A tent is a \omterm{def:g7:areas:prism}{prism} lying on its side: its bases are triangles of base
$2$ m and height $1.5$ m, and the tent is $3$ m long.
\begin{enumerate}
  \item Base \omterm{def:g6:measure:area}{area}: $B = \frac{2 \times 1.5}{2} = 1.5$ m$^2$.
  \item \omterm{def:g6:measure:volume}{Volume}: $V = B \times h = 1.5 \times 3 = 4.5$ m$^3$.
\end{enumerate}
A tin can of \omterm{def:g3:shapes:shapes}{radius} $4$ cm and height $11$ cm:
$V = \pi \times 4^2 \times 11 = 176\pi \approx 553$ cm$^3$, roughly
\omterm{def:g3:division:half}{half} a liter.
\end{example}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:g7:areas:1}
Compute the \omterm{def:g6:measure:area}{area} of a \omterm{def:g7:central:parallelogram}{parallelogram} with base $8$ cm and height
$5$ cm; of another with base $6.5$ m and height $4$ m.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g7:areas:2}
A \omterm{def:g7:central:parallelogram}{parallelogram} has sides $10$ cm and $6$ cm, and the height relative
to the $10$ cm base measures $4$ cm. Compute its \omterm{def:g6:measure:area}{area}. Why is the
answer not $60$ cm$^2$?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g7:areas:3}
Compute the \omterm{def:g6:measure:area}{area} of a triangle with base $12$ cm and height $7$ cm;
of a \omterm{def:g6:shapes:triangles}{right triangle} with legs $9$ and $6$; of a triangle with base
$5.5$ m and height $2$ m.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g7:areas:4}
Compute the \omterm{def:g6:measure:area}{area} of a disk of \omterm{def:g3:shapes:shapes}{radius} $5$ cm, then of a half-disk of
\omterm{def:g3:shapes:shapes}{radius} $10$ cm (exact values with $\pi$, then rounded to the unit).
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g7:areas:5}
Compute the \omterm{def:g6:measure:perimeter}{perimeter} \emph{and} the \omterm{def:g6:measure:area}{area} of a disk of \omterm{def:g3:shapes:shapes}{radius} $3$ cm.
Which of the two answers is in cm and which in cm$^2$?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g7:areas:6}
Compute the \omterm{def:g6:measure:volume}{volume} of a \omterm{def:g7:areas:prism}{prism} with base \omterm{def:g6:measure:area}{area} $24$ cm$^2$ and height
$10$ cm; of a \omterm{def:g7:areas:prism}{cylinder} of \omterm{def:g3:shapes:shapes}{radius} $3$ cm and height $8$ cm (exact,
then rounded).
\end{exercise}

\begin{exercise}[$\star$]\label{exo:g7:areas:7}
A triangle has \omterm{def:g6:measure:area}{area} $28$ cm$^2$ and base $8$ cm. Find the
corresponding height, writing the equation first.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g7:areas:8}
Draw any triangle and measure carefully the three base--height pairs.
Compute $\frac{b \times h}{2}$ for each pair: the three results should
agree (up to measuring error). Why?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g7:areas:9}
A trapezoid-shaped garden can be split, by one diagonal, into two
triangles sharing the same height $h = 20$ m, with bases $30$ m and
$18$ m. Compute its total \omterm{def:g6:measure:area}{area}. Can you guess a general formula for
the trapezoid?
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g7:areas:10}
A cylindrical vase of \omterm{def:g3:shapes:shapes}{radius} $6$ cm contains water to a height of
$15$ cm. All the water is poured into an empty \omterm{def:g5:solids:def}{box}
$18$ cm $\times$ $12$ cm at the base. What water height is reached in
the \omterm{def:g5:solids:def}{box}? (Same \omterm{def:g6:measure:volume}{volume}, different base.)
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:g7:areas:11}
A pizza of \omterm{def:g4:geometry:circle}{diameter} $30$ cm costs $9$ euros; one of \omterm{def:g4:geometry:circle}{diameter} $40$ cm
costs $14$ euros. Compute the two \omterm{def:g6:measure:area}{areas} and the price per $100$
cm$^2$. Which pizza is the better deal?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:g7:areas:12}
The two legs of a \omterm{def:g6:shapes:triangles}{right triangle} measure $6$ cm and $8$ cm, and its
hypotenuse measures $10$ cm. Compute its \omterm{def:g6:measure:area}{area} with the legs, then use
that \omterm{def:g6:measure:area}{area} to find the height relative to the \emph{hypotenuse}.
\end{exercise}

\section{Problem: The vanishing square}

\begin{problem}[{Weekend problem --- a famous ``proof'' that
$64 = 65$, unmasked by areas and slopes, with pizzas and rings for
dessert}]
\label{pb:g7:areas:1}
A magician cuts an $8 \times 8$ square into four pieces, slides
them around, and reassembles them into a $5 \times 13$ rectangle.
The audience counts: $8 \times 8 = 64$, but
$5 \times 13 = 65$ --- a square unit has appeared out of thin
air! This chapter's \omterm{def:g6:measure:area}{area} formulas are exactly the tools that
catch the trick. The problem then puts honest \omterm{def:g6:measure:area}{area} reasoning to
work: on trapezoids, on tilted stacks, on pizza economics --- and
finally returns to the magician for a second, \omterm{def:g3:numbers:evenodd}{even} stranger
performance.

\medskip
\noindent\textbf{Part I --- The trick, performed and unmasked.}
The four pieces of the $8 \times 8$ square are: two \omterm{def:g6:shapes:triangles}{right
triangles} with legs $8$ and $3$, and two right trapezoids with
\omterm{def:g6:lines:perp}{parallel} sides $5$ and $3$ and height $5$.
\begin{enumerate}
  \item Compute the \omterm{def:g6:measure:area}{area} of the original square and the \omterm{def:g6:measure:area}{area} of
        the claimed rectangle. State the scandal in one line.
  \item Compute the \omterm{def:g6:measure:area}{area} of each of the four pieces
        (\cref{thm:g7:areas:triangle}; for the trapezoids, cut
        or use \cref{exo:g7:areas:9}).
  \item Add the four \omterm{def:g6:measure:area}{areas}. Which of the two figures --- square
        or rectangle --- can the pieces really fill?
  \item In the rectangle arrangement, a triangle's hypotenuse
        (rising $3$ across $8$) is supposed to continue a
        trapezoid's slanted edge (rising $2$ across $5$) in one
        straight ``diagonal''. Test the alignment: are $3$ across
        $8$ and $2$ across $5$ the same steepness
        (\cref{thm:g7:prop:cross})?
  \item Explain where the sixty-fifth square hides: what shape is
        the gap along the false diagonal, what is its exact \omterm{def:g6:measure:area}{area},
        and why does the eye miss it? (How wide, on average, is a
        sliver of that \omterm{def:g6:measure:area}{area} stretched over the rectangle's
        diagonal, roughly $13$ units long?)
\end{enumerate}

\medskip
\noindent\textbf{Part II --- Honest \omterm{def:g6:measure:area}{area} reasoning.}
\begin{enumerate}[resume]
  \item Draw two \omterm{def:g6:lines:perp}{parallel lines} $4$ cm apart, a base of $6$ cm on
        one of them, and three different triangles on that base
        with apexes at various points of the other line. Compute
        the three \omterm{def:g6:measure:area}{areas}. Why are they all equal, however the apex
        slides along its line?
  \item Prove the trapezoid formula in general: a trapezoid with
        \omterm{def:g6:lines:perp}{parallel} sides $B$ and $b$ and height $h$, cut along a
        diagonal, gives two triangles of the same height $h$.
        Conclude:
        \[
        A = \frac{(B + b) \times h}{2} .
        \]
  \item Apply it to a trapezoid with $B = 9$ cm, $b = 5$ cm,
        $h = 4$ cm.
  \item Carpenters use the same formula in disguise: \omterm{def:g6:measure:area}{area} $=$
        (average of the two \omterm{def:g6:lines:perp}{parallel} sides) $\times$ height.
        Explain why this is the same rule, and recompute question
        8 that way.
  \item Tilt a neat stack of playing cards into a slanted stack.
        Explain why the \omterm{def:g6:measure:volume}{volume} has not changed, and what this
        says about the height in the \omterm{def:g7:areas:prism}{prism} formula
        $V = B \times h$ (\cref{thm:g7:areas:volume}): which
        height must one take for a slanted stack --- the slanted
        length or the vertical one? (Compare
        \cref{exo:g7:areas:2}, one dimension up.)
\end{enumerate}

\medskip
\noindent\textbf{Part III --- Rings, pizzas, and the magician's
return.}
\begin{enumerate}[resume]
  \item A circular pond of \omterm{def:g3:shapes:shapes}{radius} $3$ m sits at the center of a
        circular lawn of \omterm{def:g3:shapes:shapes}{radius} $5$ m. Compute the \omterm{def:g6:measure:area}{area} of the
        grass ring (exact with $\pi$, then rounded to the
        m$^2$).
  \item Find the \omterm{def:g3:shapes:shapes}{radius} of the single disk whose \omterm{def:g6:measure:area}{area} equals
        that ring's \omterm{def:g6:measure:area}{area} exactly. (The three radii you now have
        form a famous triple, whose story is told in
        \cref{ch:g8:pythagoras}.)
  \item Which is more pizza: one pizza of \omterm{def:g4:geometry:circle}{diameter} $40$ cm, or
        two pizzas of \omterm{def:g4:geometry:circle}{diameter} $28$ cm? Compute and compare.
  \item To \emph{double} a pizza's \omterm{def:g6:measure:area}{area}, by roughly what factor
        must its \omterm{def:g3:shapes:shapes}{radius} grow? Check the candidate $1.4$ (recall
        that \omterm{def:g6:measure:area}{areas} follow the \emph{square} of the scaling
        factor, \cref{pb:g6:propdata:1}), and say where the
        exact factor --- the number whose square is $2$ --- makes
        its official entrance
        (\cref{ex:g8:pythagoras:diagonal}).
  \item The magician returns: he cuts a $13 \times 13$ square
        with the numbers $5$, $8$, $13$ playing the roles that
        $3$, $5$, $8$ played before, and reassembles an
        $8 \times 21$ rectangle. Compute both \omterm{def:g6:measure:area}{areas}: what
        vanishes this time? The numbers $3, 5, 8, 13, 21$ are
        consecutive terms of the sequence of
        \cref{pb:g7:fractions:1} --- state the pattern of the
        magician's gains and losses, and the moral: why do \omterm{def:g6:measure:area}{areas}
        never lie?
\end{enumerate}
\end{problem}
