\section*{Chapter \ref{ch:g10:numbers} --- Numbers and Sets of Numbers}

\begin{solution}{exo:g10:numbers:1}
$\frac{15}{3} = 5 \in \N$. \quad $-7 \in \Z$. \quad
$\frac{22}{7} \in \Q$ (it is not an \omterm{def:g10:numbers:sets}{integer}: $22 = 7 \times 3 + 1$).
\quad $\sqrt 9 = 3 \in \N$. \quad $\sqrt{10} \in \R$ (\omterm{ex:g10:numbers:classify}{irrational}, since
$10$ is not the square of a rational --- admitted here, in the spirit of
\cref{thm:g10:numbers:sqrt2}). \quad $-2.4 = -\frac{24}{10} \in \Q$.
\quad $0 \in \N$.
\end{solution}

\begin{solution}{exo:g10:numbers:2}
(a) $\intco{-2}{5}$: filled dot at $-2$, hollow dot at $5$.
(b) $\intoo{3}{+\infty}$: hollow dot at $3$, shading to the right.
(c) $\intoc{-\infty}{-1}$: shading from the left up to a filled dot at
$-1$.
(d) ``distance from $x$ to $2$ at most $3$'' means $\abs{x - 2} \leq 3$,
i.e.\ $x \in \intcc{-1}{5}$: filled dots at $-1$ and $5$.
\end{solution}

\begin{solution}{exo:g10:numbers:3}
(a) The two \omterm{def:g10:numbers:interval}{intervals} overlap between $0$ and $2$:
$I \cap J = \intcc{0}{2}$ ($0 \in J$ and $2 \in I$, and both belong to the
other set as well), and $I \cup J = \intco{-3}{4}$.

(b) $I \cap J$ is the set of $x$ with $-2 \leq x$ and $x < 1$:
$\intco{-2}{1}$. The \omterm{def:g10:numbers:interunion}{union} covers everything: $I \cup J = \R$.
\end{solution}

\begin{solution}{exo:g10:numbers:4}
$\abs{-6} = 6$; \quad $\abs{4 - 9} = \abs{-5} = 5$; \quad
$\abs{-3 - 5} = \abs{-8} = 8$; \quad
$\sqrt 2 > 1$ so $\abs{\sqrt2 - 1} = \sqrt2 - 1$; \quad
$1 - \sqrt2 < 0$ so $\abs{1 - \sqrt2} = \sqrt2 - 1$ as well (a number and
its opposite have the same \omterm{def:g10:numbers:abs}{absolute value}).
\end{solution}

\begin{solution}{exo:g10:numbers:5}
$\abs{x} = 5$: distance $5$ from $0$, so $x = 5$ or $x = -5$.

$\abs{x - 1} = 3$: distance $3$ from $1$, so $x = 4$ or $x = -2$.

$\abs{x - 4} \leq 1$: distance at most $1$ from $4$, so
$x \in \intcc{3}{5}$.

$\abs{x + 2} < 3$: distance strictly less than $3$ from $-2$, so
$x \in \intoo{-5}{1}$.
\end{solution}

\begin{solution}{exo:g10:numbers:6}
Each \omterm{def:g10:numbers:interval}{interval} is described by its center $a$ (midpoint of the bounds) and
radius $r$ (half the length).

$\intcc{2}{8}$: $a = 5$, $r = 3$, so $\abs{x - 5} \leq 3$.

$\intoo{-5}{1}$: $a = -2$, $r = 3$, open bounds, so $\abs{x + 2} < 3$.

$\intcc{-7}{-3}$: $a = -5$, $r = 2$, so $\abs{x + 5} \leq 2$.
\end{solution}

\begin{solution}{exo:g10:numbers:7}
Let $x = 0.272727\dots$ Then $100x = 27.2727\dots$, and subtracting:
\[
100x - x = 27.2727\dots - 0.2727\dots = 27,
\]
so $99x = 27$ and $x = \frac{27}{99} = \frac{3}{11}$, a quotient of
\omterm{def:g10:numbers:sets}{integers}: $x$ is rational.
\end{solution}

\begin{solution}{exo:g10:numbers:8}
\emph{1. True:} adding \omterm{def:g10:numbers:sets}{integers} (positive or negative whole numbers)
always produces an \omterm{def:g10:numbers:sets}{integer}.

\emph{2. False:} $\frac{1}{2}$ is a quotient of the \omterm{def:g10:numbers:sets}{integers} $1$ and $2$
and is not an \omterm{def:g10:numbers:sets}{integer}.

\emph{3. True:} $\frac pq + \frac{p'}{q'} = \frac{pq' + p'q}{qq'}$ is
again a quotient of \omterm{def:g10:numbers:sets}{integers} (with nonzero denominator).

\emph{4. True:} suppose $r$ is rational, $t$ \omterm{ex:g10:numbers:classify}{irrational}, and $r + t = s$
were rational. Then $t = s - r$ would be a difference of two rationals,
hence rational (by 3, applied with $-r$) --- contradiction. So $r + t$ is
\omterm{ex:g10:numbers:classify}{irrational}.
\end{solution}

\begin{solution}{exo:g10:numbers:9}
Multiplying $1.414 \leq \sqrt2 \leq 1.415$ by $2 > 0$:
$2.828 \leq 2\sqrt2 \leq 2.830$.

Adding $3$: $4.414 \leq \sqrt2 + 3 \leq 4.415$.

Multiplying by $-1 < 0$ reverses the inequalities:
$-1.415 \leq -\sqrt2 \leq -1.414$.
\end{solution}

\begin{solution}{exo:g10:numbers:10}
First the auxiliary fact. Divide $p$ by $3$: the remainder is $0$, $1$ or
$2$, i.e.\ $p = 3k$, $p = 3k+1$ or $p = 3k+2$. Squaring:
\[
(3k)^2 = 3(3k^2), \quad
(3k+1)^2 = 3(3k^2 + 2k) + 1, \quad
(3k+2)^2 = 3(3k^2 + 4k + 1) + 1 .
\]
Only the first is a multiple of $3$: if $p^2$ is a multiple of $3$, then
$p$ is too.

Now suppose $\sqrt3 = \frac pq$ fully simplified. Squaring gives
$p^2 = 3q^2$, so $p^2$ is a multiple of $3$, so $p = 3k$. Then
$9k^2 = 3q^2$, so $q^2 = 3k^2$ and $q$ is also a multiple of $3$ --- but
then the fraction $\frac pq$ was not fully simplified: contradiction. So
$\sqrt3$ is \omterm{ex:g10:numbers:classify}{irrational}.
\end{solution}

\begin{solution}{pb:g10:numbers:1}
\textbf{1.} $-7 \in \Z$; \ $\frac{13}{4} \in \Q$; \
$\sqrt{16} = 4 \in \N$; \ $0.121212\ldots = \frac{12}{99} =
\frac{4}{33} \in \Q$ (the repeating-decimals weekend problem of the Middle School volume); \
$\sqrt8 = 2\sqrt2$ is \omterm{ex:g10:numbers:classify}{irrational}: smallest set $\R$; \ $\pi$:
$\R$.

\textbf{2.} $\frac pq + \frac rs = \frac{ps + qr}{qs}$ and
$\frac pq \times \frac rs = \frac{pr}{qs}$: quotients of
\omterm{def:g10:numbers:sets}{integers} with nonzero denominators --- rational.

\textbf{3.} (a) If $r$ is rational, $x$ \omterm{ex:g10:numbers:classify}{irrational} and
$r + x = q$ were rational, then $x = q - r$ would be a
difference of rationals, hence rational (question 2):
contradiction. (b) If $r \neq 0$ and $r x = q$ rational, then
$x = \frac qr$ is rational: contradiction.

\textbf{4.} $\sqrt2$ and $-\sqrt2$ are \omterm{ex:g10:numbers:classify}{irrational} with sum $0$;
$\sqrt2$ and $\sqrt2$ have product $2$. Rational results from
\omterm{ex:g10:numbers:classify}{irrational} ingredients: the \omterm{ex:g10:numbers:classify}{irrationals} form no stable kingdom.

\textbf{5.} $\sqrt8 = 2\sqrt2$, so
$\sqrt2 + \sqrt8 = 3\sqrt2$: \omterm{ex:g10:numbers:classify}{irrational} (question 3b, multiplier
$3$). And $\sqrt2 \times \sqrt8 = \sqrt{16} = 4 \in \N$: the
product of two \omterm{ex:g10:numbers:classify}{irrationals} lands in the smallest kingdom of
all.

\textbf{6.} $3.475$, then $3.471$ (or $3.4701$, $3.47001$,
\dots): appending decimal digits produces fresh rationals
between the two forever.

\textbf{7.} The multiples of $10^{-n}$ march along the line in
steps of $10^{-n} < g$. The first multiple strictly greater
than $a$ --- it exists, since the multiples eventually exceed
$a$ --- lies at most one step beyond $a$, hence before
$a + g = b$: strictly between $a$ and $b$. A grid point is a
decimal, hence a rational: every \omterm{def:g10:numbers:interval}{interval} of positive length
contains one.

\textbf{8.} Between $a$ and the rational $r_1$ found in
question 7 there is (question 7 again) a rational $r_2$; between
$a$ and $r_2$ a rational $r_3$; and so on: infinitely many, all
distinct, all in the original \omterm{def:g10:numbers:interval}{interval}.

\textbf{9.} $r + \frac{\sqrt2}{10^k}$ is the sum of a rational
and an \omterm{ex:g10:numbers:classify}{irrational} ($\frac{\sqrt2}{10^k}$ is \omterm{ex:g10:numbers:classify}{irrational} by
question 3b), hence \omterm{ex:g10:numbers:classify}{irrational}. Since
$\frac{\sqrt2}{10^k} < \frac{2}{10^k}$ shrinks below any bound,
for $k$ large enough $r + \frac{\sqrt2}{10^k}$ still lies
before $b$: an \omterm{ex:g10:numbers:classify}{irrational} inside the \omterm{def:g10:numbers:interval}{interval} --- and varying
$k$ gives infinitely many.

\textbf{10.} No smallest positive real: if $s > 0$ were it, the
\omterm{def:g10:numbers:interval}{interval} $(0, s)$ would still contain a rational (question 7),
positive and smaller than $s$. No number just after $3$: any
candidate $c > 3$ leaves the \omterm{def:g10:numbers:interval}{interval} $(3, c)$, which is not
empty --- the no-next-number weekend problem of the Middle School volume's game, now a theorem about
$\R$.

\textbf{11.} $\abs{x - 5} = 2$: $x = 3$ or $x = 7$.
$\abs{x - 5} < 2$: $x \in \intoo{3}{7}$.
$\abs{x + 1} \geq 3$: distance to $-1$ at least $3$:
$x \in \intoc{-\infty}{-4} \cup \intco{2}{+\infty}$.

\textbf{12.} $\abs{d - 80} \leq 0.05$, i.e.\
$d \in \intcc{79.95}{80.05}$. The piston at $79.97$ passes;
the one at $80.06$ fails by a hundredth of a millimeter.

\textbf{13.} $(3, -5)$: $\abs{-2} = 2 \leq 8$. $(-2, -7)$:
$\abs{-9} = 9 = 2 + 7$: equality. Same signs: $\abs{a + b}$ is
the sum of the distances, equal to $\abs a + \abs b$. Opposite
signs: the walk doubles back, $\abs{a + b}$ is the
\emph{difference} of the distances, strictly less than their
sum (unless one of them is $0$). Equality exactly when $a$ and
$b$ have the same sign or one is zero.

\textbf{14.} The solutions are the points equidistant from $2$
and $-4$: the midpoint, $x = -1$. It is the perpendicular
bisector of the two-mirrors weekend problem of the Middle School volume, collapsed to one
dimension: a single point.

\textbf{15.} $\sqrt{x^2} = \abs{x}$, not $x$: for $x = -3$,
$\sqrt{9} = 3 = \abs{-3} \neq -3$. Then $x^2 < 9$ reads
$\sqrt{x^2} < 3$, i.e.\ $\abs x < 3$:
$x \in \intoo{-3}{3}$.

\textbf{16.} The measurement asserts only that the length lies
in the \omterm{def:g10:numbers:interval}{interval} $\intcc{1.41420}{1.41422}$ --- and by questions
7 and 9 \emph{that \omterm{def:g10:numbers:interval}{interval} contains infinitely many rationals
and infinitely many \omterm{ex:g10:numbers:classify}{irrationals}}. The same holds for every
future tolerance, however small. No measurement can ever
separate the two families; only a proof about the exact length
--- like the one for the diagonal of the unit square
(the irrationality weekend problem of the Middle School volume) --- can.

\textbf{17.} They approach $\sqrt2$ with errors below
$10^{-1}, 10^{-2}, 10^{-3}, \dots$: rationals press against
$\sqrt2$ at every scale. The general fact: every \omterm{def:g10:numbers:sets}{real number} is
approached as closely as desired by rationals (its decimal
truncations) --- density once more, seen from the target's
side.

\textbf{18.} $a + b \in \intcc{2.4 + 1.1}{2.6 + 1.3} =
\intcc{3.5}{3.9}$: uncertainty $\pm 0.2$ (errors added).
$a \times b \in \intcc{2.4 \times 1.1}{2.6 \times 1.3} =
\intcc{2.64}{3.38}$: an uncertainty of about $\pm 0.37$ around
$3$ --- relative errors add for products, so precision degrades
faster.

\textbf{19.} Write the true values $a = 2.5 + e_1$,
$b = 1.2 + e_2$ with $\abs{e_1}, \abs{e_2} \leq 0.1$. Then the
error of the sum is $\abs{e_1 + e_2} \leq \abs{e_1} +
\abs{e_2} \leq 0.2$: the engineer's rule \emph{is} the triangle
inequality of question 13.

\textbf{20.} The real line has no gaps and no neighbours: after
any number, no ``next'' one (question 10). Its points split
into the rationals --- exactly the stopping or repeating
decimals (the repeating-decimals weekend problem of the Middle School volume) --- and the \omterm{ex:g10:numbers:classify}{irrationals},
and the two families are interwoven so tightly that every
\omterm{def:g10:numbers:interval}{interval} holds infinitely many of each (questions 8 and 9),
which is why no measurement, only proof, can tell them apart
(question 16). And the final surprise, left as a promise: the
\omterm{ex:g10:numbers:classify}{irrationals} outnumber the rationals --- not by counting one by
one, but in the precise sense of comparing infinities, a
theory built in the university volumes.
\end{solution}
