\section*{Chapter \ref{ch:g5:geometry} --- Geometric Constructions}

\begin{solution}{exo:g5:geometry:1}
Both constructions follow \cref{met:g5:geometry:basic}; the
perpendicular and the parallel through $A$ are themselves
perpendicular to each other.
\end{solution}

\begin{solution}{exo:g5:geometry:2}
Chain of perpendiculars as in \cref{ex:g5:geometry:chain}. If the
three built angles are exactly right and the lengths exact, the
fourth angle has no choice but to be right too --- a well-built
rectangle checks itself.
\end{solution}

\begin{solution}{exo:g5:geometry:3}
Base $[AB]$ of $7$ cm; arcs of radii $6$ cm (from $A$) and $3$ cm
(from $B$); the crossing is $C$.
\end{solution}

\begin{solution}{exo:g5:geometry:4}
Isosceles: base $5$ cm, both arcs with the same $7$ cm opening.
Equilateral: base $6$ cm, both arcs with the $6$ cm opening.
\end{solution}

\begin{solution}{exo:g5:geometry:5}
The set square guarantees the right angles; the compass (or the
ruler) carries the $5$ cm side to all four sides.
\end{solution}

\begin{solution}{exo:g5:geometry:6}
Draw a side $[AB]$ of $4$ cm, choose a direction for $[AD]$ (not
perpendicular, to avoid the square), $AD = 4$ cm; then arcs of
$4$ cm from $B$ and from $D$ cross at $C$. The compass confirms the
four equal sides.
\end{solution}

\begin{solution}{exo:g5:geometry:7}
The circle of center $A$, radius $4$ cm, and the circle of center
$B$, radius $3$ cm, cross in \emph{two} points (one above, one below
the line $(AB)$), because $4 + 3 > 5$ and the circles overlap.
\end{solution}

\begin{solution}{exo:g5:geometry:8}
One correct program: ``1. Draw a rectangle $ABCD$ with $AB = 6$ cm
(bottom) and $BC = 2$ cm. 2. Draw two arcs of radius $4$ cm centered
at $D$ and at $C$ (the top corners); name $S$ their crossing above
the rectangle. 3. Draw $[DS]$ and $[CS]$.'' (The point of the pencil
is the isosceles triangle $DSC$.)
\end{solution}

\begin{solution}{exo:g5:geometry:9}
The arcs of radii $3$ cm and $4$ cm, drawn from the two ends of the
$8$ cm base, never meet: even end to end, $3 + 4 = 7$ cm cannot
stretch across $8$ cm. Two sides together must always be longer than
the third --- otherwise the triangle cannot close.
\end{solution}

\begin{solution}{exo:g5:geometry:10}
Draw a segment of $6$ cm and its midpoint $O$; draw the
perpendicular at $O$; mark on it two points at $2$ cm from $O$ (one
on each side); join the four endpoints in order. The compass
confirms the four equal sides.
\end{solution}

\begin{solution}{exo:g5:geometry:11}
On the plan, the goat reaches the disk of center $A$ and radius
$4$ cm, cut off by the fence line at $3$ cm from $A$: her region is
the disk minus the sliver beyond the fence. Along the fence she can
graze on a segment (where the circle crosses the fence line);
beyond it, nothing.
\end{solution}
