\section*{Chapter \ref{ch:g5:solids} --- Cubes and Boxes}

\begin{solution}{exo:g5:solids:1}
For instance: a die (cube); a cereal box (box); a can (cylinder); a
football (ball); certain roofs or the monuments of Egypt
(pyramid).
\end{solution}

\begin{solution}{exo:g5:solids:2}
A box has $6$ faces, $12$ edges, $8$ vertices --- and
$6 + 8 = 12 + 2$. \checkmark
\end{solution}

\begin{solution}{exo:g5:solids:3}
The six faces come in three pairs of identical opposite faces
(top--bottom, front--back, left--right). On a cube, all six faces
are identical squares.
\end{solution}

\begin{solution}{exo:g5:solids:4}
Cross-shaped net: four $2$ cm squares in a row, one above and one
below the second square. Folding the ring of four makes the sides;
the two extra squares close the top and bottom.
\end{solution}

\begin{solution}{exo:g5:solids:5}
On the cross net (four squares in a row, one above, one below the
second), the opposite pairs are: squares $1$ and $3$ of the row;
squares $2$ and $4$ of the row; the top and bottom squares. One
correct labelling: row $= 1, 2, 6, 5$; top $= 3$; bottom $= 4$
--- then $1{+}6$, $2{+}5$, $3{+}4$ all make $7$.
\end{solution}

\begin{solution}{exo:g5:solids:6}
Box: $5 \times 2 = 10$ per layer, $3$ layers: $30$ cubes. Cube:
$3 \times 3 \times 3 = 27$. L-stack: $3 \times 2 \times 1 = 6$ plus
$1 \times 2 \times 1 = 2$: $8$ cubes.
\end{solution}

\begin{solution}{exo:g5:solids:7}
Layers of $12$: for instance $4 \times 3$ or $6 \times 2$ (or
$12 \times 1$). With $3$ layers, the volume is
$12 \times 3 = 36$ cubes in every case.
\end{solution}

\begin{solution}{exo:g5:solids:8}
$27 = 3 \times 3 \times 3$: a cube of side $3$.
$64 = 4 \times 4 \times 4$: a cube of side $4$.
\end{solution}

\begin{solution}{exo:g5:solids:9}
The box holds $10 \times 5 \times 4 = 200$ sugar cubes. At $6$ per
day: $200 = 6 \times 33 + 2$, so the box lasts $33$ full days (and
$2$ cubes remain for the morning of day $34$).
\end{solution}

\begin{solution}{exo:g5:solids:10}
Paint on $3$ faces: the $8$ corner cubes. On $2$ faces: the middles
of the $12$ edges: $12$ cubes. On $1$ face: the centers of the $6$
faces: $6$ cubes. On none: the single hidden center cube: $1$.
Check: $8 + 12 + 6 + 1 = 27$. \checkmark
\end{solution}
