\section*{Chapter \ref{ch:g12:deriv} --- Differentiation and Convexity}

\begin{solution}{exo:g12:deriv:1}
$f'(x) = 3x^2 - 5$.

Quotient rule:
$g'(x) = \dfrac{(x^2+1) - x\cdot 2x}{(x^2+1)^2} = \dfrac{1 - x^2}{(x^2+1)^2}$.

Chain rule with $u = 2x+1$: $h'(x) = 5 \cdot 2 \cdot (2x+1)^4 = 10(2x+1)^4$.

Chain rule with $u = x^2 + 1$:
$k'(x) = \dfrac{2x}{2\sqrt{x^2+1}} = \dfrac{x}{\sqrt{x^2+1}}$.
\end{solution}

\begin{solution}{exo:g12:deriv:2}
$f'(x) = 2x - 3$, so $f(2) = -1$ and $f'(2) = 1$: the \omterm{def:g12:deriv:tangent}{tangent} is
$y = -1 + (x - 2) = x - 3$. The gap is
\[
f(x) - (x - 3) = x^2 - 4x + 4 = (x-2)^2 \geq 0 ,
\]
so the curve lies above the \omterm{def:g12:deriv:tangent}{tangent} everywhere, touching it only at $x = 2$
(as expected: $f$ is \omterm{def:g12:deriv:convex}{convex}).
\end{solution}

\begin{solution}{exo:g12:deriv:3}
Both are difference quotients. With $f(x) = (1+x)^{100}$ at $0$:
$f'(x) = 100(1+x)^{99}$, so the limit is $f'(0) = 100$.
With $g(x) = \sqrt{x}$ at $4$: $g'(x) = \frac{1}{2\sqrt{x}}$, so the limit is
$g'(4) = \frac14$.
\end{solution}

\begin{solution}{exo:g12:deriv:4}
\omterm{def:g11:func:function}{Domain} $\R \setminus \{1\}$. Limits: at $\pm\infty$, $f(x) \sim x \to
\pm\infty$; at $1^\pm$, numerator $\to 4 > 0$ and denominator $\to 0^\pm$,
so $f \to \pm\infty$; the line $x = 1$ is a \omterm{def:g12:limcont:asymptote}{vertical asymptote}.

\[
f'(x) = \frac{2x(x-1) - (x^2+3)}{(x-1)^2} = \frac{x^2 - 2x - 3}{(x-1)^2}
= \frac{(x-3)(x+1)}{(x-1)^2}.
\]
Hence $f' > 0$ on $\intoo{-\infty}{-1}$ and $\intoo{3}{+\infty}$, $f' < 0$ on
$\intoo{-1}{1}$ and $\intoo{1}{3}$. The \omterm{def:g11:func:function}{function} increases to a \omterm{def:g12:deriv:extremum}{local maximum}
$f(-1) = -2$, decreases to $-\infty$, jumps to $+\infty$ after the \omterm{def:g12:limcont:asymptote}{asymptote},
decreases to a local \omterm{def:g10:functions:extrema}{minimum} $f(3) = 6$, then increases.
\end{solution}

\begin{solution}{exo:g12:deriv:5}
For $x \in \intoo{0}{15}$, the box has a square base of side $30 - 2x$ and
height $x$, so
\[
V(x) = x(30 - 2x)^2 .
\]
$V'(x) = (30-2x)^2 + x \cdot 2(30-2x)(-2) = (30-2x)(30 - 2x - 4x)
= (30-2x)(30-6x)$. On $\intoo{0}{15}$, $30 - 2x > 0$, so $V'$ has the sign of
$30 - 6x$: positive before $x = 5$, negative after. The volume is maximal for
$x = 5$ cm, giving $V(5) = 5 \times 20^2 = 2000\ \text{cm}^3$.
\end{solution}

\begin{solution}{exo:g12:deriv:6}
\emph{1.} $f'(x) = 4x^3 - 12x^2$ and $f''(x) = 12x^2 - 24x = 12x(x-2)$.
Thus $f'' > 0$ on $\intoo{-\infty}{0}$ and on $\intoo{2}{+\infty}$ (\omterm{def:g12:deriv:convex}{convex}),
$f'' < 0$ on $\intoo{0}{2}$ (\omterm{def:g12:deriv:convex}{concave}). $f''$ changes sign at $0$ and at $2$:
both are \omterm{def:g12:deriv:inflection}{inflection points}.

\emph{2.} At $a = 2$: $f(2) = 16 - 32 + 10 = -6$, $f'(2) = 32 - 48 = -16$,
\omterm{def:g12:deriv:tangent}{tangent} $y = -6 - 16(x-2) = -16x + 26$. The gap is
\[
g(x) = f(x) + 16x - 26 = x^4 - 4x^3 + 16x - 16 .
\]
Since $g(2) = g'(2) = g''(2) = 0$, $(x-2)^3$ divides $g$; division gives
$g(x) = (x-2)^3(x+2)$. Near $x = 2$, $x + 2 > 0$, so $g$ has the sign of
$(x-2)^3$: negative before $2$, positive after. The curve crosses the
\omterm{def:g12:deriv:tangent}{tangent}: $2$ is indeed an \omterm{def:g12:deriv:inflection}{inflection point}.
\end{solution}

\begin{solution}{exo:g12:deriv:7}
The \omterm{def:g11:func:function}{function} $f(x) = \sqrt{x}$ satisfies
$f''(x) = -\frac{1}{4}x^{-3/2} < 0$ on $\intoo{0}{+\infty}$: it is \omterm{def:g12:deriv:convex}{concave},
so its \omterm{def:g10:functions:graph}{graph} lies \emph{below} each \omterm{def:g12:deriv:tangent}{tangent}. The \omterm{def:g12:deriv:tangent}{tangent} at $a = 1$ is
\[
y = f(1) + f'(1)(x - 1) = 1 + \tfrac12 (x-1) = \frac{x+1}{2},
\]
whence $\sqrt{x} \leq \frac{x+1}{2}$ for all $x > 0$. Equality means the
curve touches the \omterm{def:g12:deriv:tangent}{tangent}, which happens only at the point of tangency
$x = 1$. (Equivalently: $\left(\sqrt x - 1\right)^2 \geq 0$.)
\end{solution}

\begin{solution}{exo:g12:deriv:8}
By \cref{thm:g12:deriv:convexity}, the \omterm{def:g10:functions:graph}{graph} of a \omterm{def:g12:deriv:derivative}{differentiable} \omterm{def:g12:deriv:convex}{convex}
\omterm{def:g11:func:function}{function} lies above each of its \omterm{def:g12:deriv:tangent}{tangents}. The \omterm{def:g12:deriv:tangent}{tangent} at $a$ is horizontal
($f'(a) = 0$), of \omterm{def:g10:algebra:equation}{equation} $y = f(a)$. Hence $f(x) \geq f(a)$ for all
$x \in \R$: $f(a)$ is the global \omterm{def:g10:functions:extrema}{minimum}.
\end{solution}

\begin{solution}{exo:g12:deriv:9}
\emph{Fixed perimeter.} A rectangle with sides $x$ and $\frac{p}{2} - x$
($0 < x < \frac p2$) has area $S(x) = x\left(\frac p2 - x\right)$. Then
$S'(x) = \frac p2 - 2x$ vanishes at $x = \frac p4$, positive before and
negative after: the \omterm{def:g10:functions:extrema}{maximum} is at $x = \frac{p}{4}$, where both sides equal
$\frac p4$ --- a square.

\emph{Fixed area.} Sides $x$ and $\frac Ax$ give perimeter
$P(x) = 2\left(x + \frac Ax\right)$, with
$P'(x) = 2\left(1 - \frac{A}{x^2}\right)$, negative for $x < \sqrt A$ and
positive after: \omterm{def:g10:functions:extrema}{minimum} at $x = \sqrt{A}$, a square again.

\emph{Relation.} The two statements are dual. Suppose a non-square rectangle
minimized the perimeter at fixed area $A$; the square of the same perimeter
would have area $> A$ by the first statement, and shrinking it homothetically
to area $A$ would strictly decrease its perimeter --- contradicting
minimality. Each statement thus implies the other.
\end{solution}

\begin{solution}{pb:g12:deriv:1}
\textbf{1.} $30x\left(3x^2 + 1\right)^4$; \quad
$\dfrac{x}{\sqrt{x^2 + 9}}$; \quad
$-\dfrac{2x}{\left(x^2 + 1\right)^2}$.

\textbf{2.} $y(4) = 5$, \omterm{def:g10:reffunc:affine}{slope} $\frac45$: \omterm{def:g12:deriv:tangent}{tangent}
$y = 5 + \frac45(x - 4)$, i.e.\ $y = \frac45 x + \frac95$.

\textbf{3.} $f'(x) = \frac{(x^2 + 1) - x \cdot 2x}{(x^2+1)^2}
= \frac{1 - x^2}{(x^2 + 1)^2}$: negative, positive, negative
across $-1$ and $1$: \omterm{def:g10:functions:extrema}{minimum} $f(-1) = -\frac12$, \omterm{def:g10:functions:extrema}{maximum}
$f(1) = \frac12$, with \omterm{def:g12:limcont:asymptote}{horizontal asymptote} $0$ on both sides.

\textbf{4.} $g'(x) = 3x^2 - 6x = 3x(x - 2)$: \omterm{def:g12:seq:monotonic}{increasing},
\omterm{def:g10:functions:variations}{decreasing}, \omterm{def:g12:seq:monotonic}{increasing} across $0$ and $2$; local max
$g(0) = 4$, local min $g(2) = 0$. $g''(x) = 6x - 6$: \omterm{def:g12:deriv:convex}{concave}
before $1$, \omterm{def:g12:deriv:convex}{convex} after: inflection at $(1, 2)$.

\textbf{5.} \omterm{def:g12:deriv:tangent}{Tangent} at $1$: \omterm{def:g10:reffunc:affine}{slope} $g'(1) = -3$:
$y = 2 - 3(x - 1) = -3x + 5$. Difference:
$x^3 - 3x^2 + 4 - (-3x + 5) = x^3 - 3x^2 + 3x - 1 =
(x - 1)^3$, which changes sign at $1$: the \omterm{def:g12:deriv:tangent}{tangent}
\emph{crosses} the curve --- the signature behavior at an
\omterm{def:g12:deriv:inflection}{inflection point}, where the curve switches sides.

\textbf{6.} $T(x) = \dfrac{\sqrt{x^2 + 900}}{5} +
\dfrac{\sqrt{(40 - x)^2 + 400}}{2}$. Entering before $0$ or
beyond $40$ lengthens \emph{both} legs: pointless.

\textbf{7.} $T'(x) = \dfrac{x}{5\sqrt{x^2 + 900}} -
\dfrac{40 - x}{2\sqrt{(40 - x)^2 + 400}}$.

\textbf{8.} In the running triangle, the side along the shore
is $x$ and the hypotenuse $\sqrt{x^2 + 900}$: the \omterm{def:g11:trigo:cossin}{sine} of the
angle with the perpendicular is precisely their quotient
(opposite over hypotenuse); likewise
$\sin\theta_2 = \frac{40 - x}{\sqrt{(40-x)^2 + 400}}$. Then
$T'(x) = 0$ reads
$\frac{\sin\theta_1}{5} = \frac{\sin\theta_2}{2}$ --- Snell's
law with speeds $5$ and $2$.

\textbf{9.} $T(24) \approx 20.5$ s; $T(40) = 10 + 10 = 20$ s;
$T(0) = 6 + \sqrt{2000}/2 \approx 28.4$ s. The straight line
loses even to ``run all the way, then swim straight'' --- and
both lose to the refracted path.

\textbf{10.} \omterm{thm:g12:limcont:ivt}{Dichotomy} on $T'$ (negative at $24$, positive at
$40$) \omterm{def:g12:seq:limit}{converges} to $x^* \approx 34$ m, with
$T(x^*) \approx 19.5$ s: about one second faster than the
straight line --- the difference between a rescue and a
tragedy.

\textbf{11.} Each leg-time is a \omterm{def:g12:deriv:convex}{convex} \omterm{def:g11:func:function}{function} of $x$
(a $\sqrt{\text{quadratic}}$ branch), so $T$ is \omterm{def:g12:deriv:convex}{convex}, and a
critical point of a \omterm{def:g12:deriv:convex}{convex} \omterm{def:g12:deriv:derivative}{differentiable} \omterm{def:g11:func:function}{function} is its
global \omterm{def:g10:functions:extrema}{minimum} (\cref{exo:g12:deriv:8}): $x^*$ is not merely
stationary, it is unbeatable.

\textbf{12.} Light in water is slower; by Fermat's principle
the fastest air-to-water path bends at the surface exactly by
Snell's law --- so rays from a submerged pencil reach the eye
bent, and the brain, extrapolating straight lines, sees the
pencil broken at the waterline.

\textbf{13.} $(x^2)'' = 2 > 0$: \omterm{def:g12:deriv:convex}{convex}. \omterm{def:g12:deriv:convex}{Convexity} at the
\omterm{prop:g10:coordgeom:midpoint}{midpoint}: $f\left(\frac{a+b}{2}\right) \leq
\frac{f(a) + f(b)}{2}$, which is the claim. By hand:
$\frac{a^2 + b^2}{2} - \left(\frac{a+b}{2}\right)^2 =
\frac{(a - b)^2}{4} \geq 0$.

\textbf{14.} Taking square roots (\omterm{def:g12:seq:monotonic}{increasing}) in question 13:
$\frac{a + b}{2} \leq \sqrt{\frac{a^2 + b^2}{2}}$. Full chain
on $2, 8$: harmonic $\frac{2 \cdot 16}{10} = 3.2$; geometric
$\sqrt{16} = 4$; arithmetic $5$; quadratic
$\sqrt{34} \approx 5.83$: each \omterm{def:g11:stat:mean}{mean} bows to the next.

\textbf{15.} \omterm{def:g12:deriv:tangent}{Tangent} to $\frac1x$ at $1$: value $1$, \omterm{def:g10:reffunc:affine}{slope}
$-1$: $y = 2 - x$. \omterm{def:g12:deriv:convex}{Convex} curves sit above their \omterm{def:g12:deriv:tangent}{tangents}
(\cref{thm:g12:deriv:convexity}): $\frac1x \geq 2 - x$ for all
$x > 0$, with equality exactly at the contact point $x = 1$.

\textbf{16.} At the inflection, the \emph{daily} case count
(the \omterm{def:g12:deriv:derivative}{derivative}) peaks: growth stops accelerating and begins
to decelerate --- the first mathematical signal that the wave
is turning, long before the counts themselves fall. Before it,
$f'' > 0$ (each day worse than the last); after it,
$f'' < 0$ (still growing, but more slowly).

\textbf{17.} $S'(r) = 4\pi r - \frac{2V}{r^2} = 0$ gives
$r^3 = \frac{V}{2\pi} = \frac{330}{2\pi}$:
$r \approx 3.74$ cm, and
$h = \frac{330}{\pi r^2} \approx 7.49$ cm.

\textbf{18.} $h = \frac{V}{\pi r^2} = \frac{2\pi r^3}{\pi r^2}
= 2r$ at the optimum: height equals diameter, whatever the
volume.

\textbf{19.} $S''(r) = 4\pi + \frac{4V}{r^3} > 0$: \omterm{def:g12:deriv:convex}{convex}, so
the critical radius is the global \omterm{def:g10:functions:extrema}{minimum}. Real cans stand
taller for grip, stacking, shelf presence and the thicker
top/bottom metal --- optimization always minimizes exactly
what you wrote, not what you meant.

\textbf{20.} Lifeguard: model $T(x)$, differentiate, Snell,
\omterm{def:g12:deriv:convex}{convexity}, one second saved. Can: model $S(r)$, differentiate,
$h = 2r$, \omterm{def:g12:deriv:convex}{convexity}, metal saved. Beam
(\cref{pb:g11:deriv:1}): model $S(w)$, differentiate,
$d = w\sqrt2$, table, stiffness gained. And by Fermat's
principle, light itself runs the pipeline at every surface it
meets --- nature was the first optimizer.
\end{solution}
