\section*{Chapter \ref{ch:g12:comb} --- Combinatorics and Counting}

\begin{solution}{exo:g12:comb:1}
Multiplication principle:
$26^2 \times 10^3 \times 26^2 = 26^4 \times 10^3 = 456\,976\,000$.

Without repeated characters, the four letters must be distinct
($26 \times 25 \times 24 \times 23$ ways, filling the letter positions in
order) and the three digits distinct ($10 \times 9 \times 8$):
\[
26 \times 25 \times 24 \times 23 \times 10 \times 9 \times 8
= 358\,800 \times 720 = 258\,336\,000 .
\]
\end{solution}

\begin{solution}{exo:g12:comb:2}
$\dbinom83 = \dfrac{8 \times 7 \times 6}{3!} = 56$;
$\dbinom{10}{8} = \dbinom{10}{2} = \dfrac{10 \times 9}{2} = 45$;
\[
\frac{\binom n2}{\binom{n+1}2}
= \frac{n(n-1)/2}{(n+1)n/2} = \frac{n-1}{n+1}.
\]
\end{solution}

\begin{solution}{exo:g12:comb:3}
Choose the committee: $\binom{30}{4}$ ways. Then choose president and
treasurer among the 4, in order: $4 \times 3 = 12$ ways. Total
\[
\binom{30}{4} \times 12 = 27\,405 \times 12 = 328\,860 .
\]
\end{solution}

\begin{solution}{exo:g12:comb:4}
\[
(x+2)^5 = x^5 + 10x^4 + 40x^3 + 80x^2 + 80x + 32 ,
\]
\[
(1-x)^6 = 1 - 6x + 15x^2 - 20x^3 + 15x^4 - 6x^5 + x^6 .
\]
In $(2x+3)^7$, the term in $x^3$ is
$\binom{7}{3}(2x)^3\,3^4 = 35 \times 8 \times 81\, x^3$: the coefficient is
$22\,680$.
\end{solution}

\begin{solution}{exo:g12:comb:5}
\emph{1.} $\dbinom{52}{5} = 2\,598\,960$.

\emph{2.} Exactly one ace: choose it ($4$ ways) and complete with $4$
non-aces: $4 \times \binom{48}{4} = 4 \times 194\,580 = 778\,320$.
At least one ace: complementary counting,
$\binom{52}{5} - \binom{48}{5} = 2\,598\,960 - 1\,712\,304 = 886\,656$.

\emph{3.} Choose the rank of the three-of-a-kind ($13$), its suits
($\binom43 = 4$), the rank of the pair ($12$ remaining), its suits
($\binom42 = 6$):
$13 \times 4 \times 12 \times 6 = 3744$.
\end{solution}

\begin{solution}{exo:g12:comb:6}
MATH has 4 distinct letters: $4! = 24$ anagrams.

BANANA has 6 letters: three A's, two N's, one B. Choose the positions of the
A's ($\binom63$), then of the N's among the rest ($\binom32$), the B takes
the last spot:
\[
\binom{6}{3}\binom{3}{2} = 20 \times 3 = 60 .
\]
(Equivalently $\frac{6!}{3!\,2!\,1!} = 60$.)
\end{solution}

\begin{solution}{exo:g12:comb:7}
\emph{Algebraically:}
\[
k\binom nk = \frac{k\,n!}{k!(n-k)!} = \frac{n!}{(k-1)!\,(n-k)!}
= n\,\frac{(n-1)!}{(k-1)!\bigl((n-1)-(k-1)\bigr)!} = n\binom{n-1}{k-1}.
\]
\emph{By double counting:} count pairs (committee of $k$, president in it).
Either choose the committee ($\binom nk$) then its president ($k$):
$k\binom nk$ pairs. Or choose the president first ($n$ options) then the
other $k-1$ members among the remaining $n-1$: $n\binom{n-1}{k-1}$ pairs.
\end{solution}

\begin{solution}{exo:g12:comb:8}
A path consists of exactly $m + n$ steps, of which $m$ are East and $n$ are
North; it is entirely determined by the set of instants (among the $m+n$) at
which one steps East. There are $\binom{m+n}{m}$ such choices.
\end{solution}

\begin{solution}{exo:g12:comb:9}
Split a set of $m + n$ people into a group $A$ of $m$ and a group $B$ of
$n$. A $k$-element subset contains some number $j$ of members of $A$
($0 \leq j \leq k$) and $k - j$ members of $B$; for fixed $j$ there are
$\binom mj \binom{n}{k-j}$ such subsets, and the addition principle over $j$
gives Vandermonde's identity.

With $m = n = k$:
\[
\binom{2n}{n} = \sum_{j=0}^n \binom nj \binom{n}{n-j}
= \sum_{j=0}^n \binom nj^{2},
\]
using the symmetry $\binom{n}{n-j} = \binom nj$.
\end{solution}

\begin{solution}{exo:g12:comb:10}
\emph{Via \cref{exo:g12:comb:7}:}
\[
\sum_{k=1}^n k\binom nk = \sum_{k=1}^n n\binom{n-1}{k-1}
= n\sum_{j=0}^{n-1}\binom{n-1}{j} = n\,2^{n-1},
\]
by \cref{cor:g12:comb:sums}. \emph{Via differentiation:} differentiating
$(1+x)^n = \sum_k \binom nk x^k$ gives
$n(1+x)^{n-1} = \sum_k k \binom nk x^{k-1}$; evaluate at $x = 1$.
\end{solution}

\begin{solution}{pb:g12:comb:1}
\textbf{1.} $26^2 \times 10^3 = 676\,000$ plates. BANANA: $6$
letters with A tripled and N doubled:
$\frac{6!}{3!\,2!} = 60$ anagrams.

\textbf{2.} $\binom{32}{5} = 201\,376$ hands;
$\binom42 \binom{28}{3} = 6 \times 3\,276 = 19\,656$ with
exactly two aces.

\textbf{3.} A path is a word with $4$ R's and $3$ U's: choose
the U positions: $\binom73 = 35$.

\textbf{4.} $(1+x)^4 = 1 + 4x + 6x^2 + 4x^3 + x^4$. At
$x = 1$: $\sum_k \binom nk = 2^n$; at $x = -1$:
$\sum_k (-1)^k \binom nk = 0$ --- row sums and alternating row
sums of \omterm{prop:g11:binom:pascal}{Pascal's triangle}.

\textbf{5.} Committees of $k$ people with a chair, from $n$:
choose the committee then its chair ($\binom nk \times k$), or
the chair then the other members
($n \times \binom{n-1}{k-1}$): equal. Summing over $k$: the
right side sums to $n \sum_j \binom{n-1}{j} = n\,2^{n-1}$.

\textbf{6.} A row of $10$ stars and $3$ bars encodes the order
(scoops of flavor 1 before the first bar, etc.); the row has
$13$ symbols and is determined by the bar positions:
$\binom{13}{3} = 286$ orders.

\textbf{7.} $12$ stars, $2$ bars: $\binom{14}{2} = 91$.

\textbf{8.} With $x', y', z' \geq 0$ and
$x' + y' + z' = 9$: $\binom{11}{2} = 55$.

\textbf{9.} A monomial $a^i b^j c^k$ with $i + j + k = 5$:
$\binom72 = 21$.

\textbf{10.} Formula: $3$ stars, $1$ bar: $\binom41 = 4$;
list: $(3,0)$, $(2,1)$, $(1,2)$, $(0,3)$: agreement.

\textbf{11.} ``Identical'' entered when an order was declared
to be nothing but the \emph{counts} per flavor --- the stars
carry no names. If the scoops are eaten in order, each of the
$10$ distinct positions picks a flavor freely:
$4^{10} = 1\,048\,576$ \omterm{def:g12:seq:sequence}{sequences} --- a different model and a
different world (\cref{met:g12:comb:model}: always ask
\emph{ordered? distinct? repetition allowed?}).

\textbf{12.} $D_1 = 0$; $D_2 = 1$ (swap); $D_3 = 2$ (the two
$3$-cycles); $D_4 = 9$.

\textbf{13.} Guest 1 gets hat $k \neq 1$: $n - 1$ choices. If
guest $k$ gets hat 1, the remaining $n - 2$ guests derange
their own hats: $D_{n-2}$ ways. If guest $k$ does \emph{not}
get hat 1, relabel hat 1 as guest $k$'s forbidden hat: the
$n - 1$ remaining guests derange: $D_{n-1}$ ways. Hence
$D_n = (n-1)(D_{n-1} + D_{n-2})$. Check:
$D_4 = 3(2 + 1) = 9$; and $D_5 = 4(9 + 2) = 44$.

\textbf{14.} Of the $3! = 6$ assignments, subtract those
fixing at least one hat: three fix a given hat ($2!$ each,
$3 \times 2 = 6$), overcounting the pairs ($3$ pairs, $1!$
each) which must return, and re-subtracting the identity
($1$): $D_3 = 6 - 6 + 3 - 1 = 2$, i.e.\
$3!\left(1 - 1 + \frac12 - \frac16\right) = 2$. In general
$D_n = n!\sum_{k=0}^{n} \frac{(-1)^k}{k!}$.

\textbf{15.} $\frac{D_5}{120} = \frac{44}{120} \approx
0.3667$, already close to $\frac1\eu \approx 0.3679$: the
alternating sum $1 - 1 + \frac{1}{2!} - \frac{1}{3!} + \dots$
marches to $\eu^{-1}$. A large party's hats derange about
$36.8\,\%$ of the time --- the lottery's and the secretary's
constant, third sighting.

\textbf{16.} $\P(\text{valid}) = \frac{D_{10}}{10!} \approx
0.368$. Each redraw succeeds with \omterm{def:g10:proba:distribution}{probability}
$\approx \frac1\eu$, so the expected number of draws is about
$\eu \approx 2.7$: budget three hat-passings.

\textbf{17.} Each handshake contributes $2$ to the total
degree count, so the sum of all guests' handshake-numbers is
even. A sum of \omterm{def:g10:numbers:sets}{integers} is even only if the number of odd
terms is even: odd-shakers come in even numbers. (At three
guests: possible handshake profiles never have exactly one or
three odd entries --- check the four possible \omterm{def:g10:functions:graph}{graphs}.)

\textbf{18.} $n = 1$: $1 = 1$. If
$1^3 + \dots + n^3 = \left(\frac{n(n+1)}{2}\right)^2$, adding
$(n+1)^3$:
\[
\frac{n^2(n+1)^2}{4} + (n+1)^3
= \frac{(n+1)^2\left(n^2 + 4n + 4\right)}{4}
= \left(\frac{(n+1)(n+2)}{2}\right)^{\!2} :
\]
heredity. For $n = 3$: $1 + 8 + 27 = 36 = 6^2$.

\textbf{19.} A path to $(n, n)$ makes $2n$ steps and crosses
the anti-diagonal $x + y = n$ at exactly one lattice point
$(j, n - j)$; the first half is a path with $j$ R's among $n$
steps ($\binom nj$ choices), the second half, read backwards,
likewise ($\binom nj$ again, by symmetry). Summing over the
crossing point: $\binom{2n}{n} = \sum_j \binom nj^2$ ---
Vandermonde's identity, drawn.

\textbf{20.} Multiply stages, add cases: plates and poker
hands. Encode: paths as RU-words, orders as stars and bars.
Count twice: committees-with-chair, handshakes, mid-cut paths.
Subtract and correct: the deranged hats, with $\frac1\eu$ as
the residue. Next stops: these counts under \omterm{def:g10:proba:distribution}{probability}'s
fractions, and the path-counting powers of the adjacency
matrices two chapters ahead.
\end{solution}
