\section*{Chapter \ref{ch:g12:complex} --- Complex Numbers}

\begin{solution}{exo:g12:complex:1}
$(3-2\iu)(1+4\iu) = 3 + 12\iu - 2\iu - 8\iu^2 = 11 + 10\iu$.

$\dfrac{2+\iu}{1-\iu} = \dfrac{(2+\iu)(1+\iu)}{(1-\iu)(1+\iu)}
= \dfrac{1 + 3\iu}{2} = \dfrac12 + \dfrac32\iu$.

$\iu^{2027}$: since $\iu^4 = 1$ and $2027 = 4 \times 506 + 3$,
$\iu^{2027} = \iu^3 = -\iu$.
\end{solution}

\begin{solution}{exo:g12:complex:2}
\emph{(a)} $\Delta = 16 - 52 = -36$:
$z = \dfrac{4 \pm 6\iu}{2} = 2 \pm 3\iu$. \omterm{def:g12:complex:conjugate}{Conjugate} pair; product
$= 4 + 9 = 13 = \frac ca$. \emph{(b)} $\Delta = 1 - 4 = -3$:
$z = \dfrac{-1 \pm \iu\sqrt3}{2}$ (the cube roots of unity $j$ and
$\conj j$); product $= \frac{1 + 3}{4} = 1$.
\end{solution}

\begin{solution}{exo:g12:complex:3}
$\abs{z_1} = \sqrt2$, \omterm{def:g12:complex:argument}{argument} $\frac{3\pi}{4}$ (second quadrant):
$z_1 = \sqrt2\,\eu^{3\iu\pi/4}$.
$\abs{z_2} = 2$, \omterm{def:g12:complex:argument}{argument} $-\frac\pi6$:
$z_2 = 2\,\eu^{-\iu\pi/6}$. Hence
\[
\frac{z_1}{z_2} = \frac{\sqrt2}{2}\,\eu^{\iu\left(\frac{3\pi}{4} + \frac{\pi}{6}\right)}
= \frac{\sqrt2}{2}\,\eu^{11\iu\pi/12}.
\]
Algebraically,
\[
\frac{z_1}{z_2} = \frac{-1+\iu}{\sqrt3 - \iu}
= \frac{(-1+\iu)(\sqrt3+\iu)}{4}
= \frac{-\sqrt3 - 1 + \iu(\sqrt3 - 1)}{4}.
\]
Identifying \omterm{def:g12:complex:def}{real parts}:
$\frac{\sqrt2}{2}\cos\frac{11\pi}{12} = \frac{-\sqrt3-1}{4}$, so
\[
\cos\frac{11\pi}{12} = -\frac{\sqrt6 + \sqrt2}{4}.
\]
\end{solution}

\begin{solution}{exo:g12:complex:4}
\emph{(a)} $\abs{z - 2} = \abs{z - (-\iu)}$ means $M$ is equidistant from
$A(2, 0)$ and $B(0, -1)$: the \emph{perpendicular bisector} of $[AB]$.

\emph{(b)} Distance $3$ from the point $(1,1)$: the \emph{circle} of center
$1 + \iu$ and radius $3$.

\emph{(c)} By \cref{prop:g12:complex:geometry}(4), $\frac{z-1}{z+1}$ purely
imaginary means the \omterm{def:g11:vect:vector}{vectors} from $A(1,0)$ to $M$ and from $B(-1,0)$ to $M$
are \omterm{def:g11:scal:orthogonal}{orthogonal}: $M$ sees the segment $[AB]$ under a right angle. The set is
the \emph{circle of diameter $[AB]$} (the unit circle) \emph{minus the
points $A$ and $B$} themselves.
\end{solution}

\begin{solution}{exo:g12:complex:5}
De Moivre: $\cos3\theta + \iu\sin3\theta = (\cos\theta + \iu\sin\theta)^3$.
\omterm{def:g10:algebra:expand}{Expanding} by the binomial theorem with $c = \cos\theta$, $s = \sin\theta$:
\[
(c + \iu s)^3 = c^3 + 3\iu c^2 s - 3 c s^2 - \iu s^3
= (c^3 - 3cs^2) + \iu(3c^2 s - s^3).
\]
Identifying parts and using $s^2 = 1 - c^2$, $c^2 = 1 - s^2$:
\[
\cos3\theta = 4\cos^3\theta - 3\cos\theta,
\qquad
\sin3\theta = 3\sin\theta - 4\sin^3\theta .
\]
\end{solution}

\begin{solution}{exo:g12:complex:6}
\[
\cos^3\theta = \left(\frac{\eu^{\iu\theta} + \eu^{-\iu\theta}}{2}\right)^{\!3}
= \frac{\eu^{3\iu\theta} + 3\eu^{\iu\theta} + 3\eu^{-\iu\theta} + \eu^{-3\iu\theta}}{8}
= \frac{\cos3\theta + 3\cos\theta}{4}.
\]
Hence
\[
\int_0^{\pi/2}\cos^3\theta\,\dd\theta
= \frac14\left[\frac{\sin3\theta}{3} + 3\sin\theta\right]_0^{\pi/2}
= \frac14\left(-\frac13 + 3\right) = \frac{2}{3}.
\]
(Consistent with $W_3 = \frac23$ in \cref{exo:g12:integ:9}, since
$\int_0^{\pi/2}\cos^3 = \int_0^{\pi/2}\sin^3$ by the symmetry
$\theta \mapsto \frac\pi2 - \theta$.)
\end{solution}

\begin{solution}{exo:g12:complex:7}
$c = a + \eu^{\pm\iu\pi/3}(b - a)$ with $b - a = 2 + \iu$ and
$\eu^{\pm\iu\pi/3} = \frac12 \pm \iu\frac{\sqrt3}{2}$:
\[
\eu^{\iu\pi/3}(2+\iu) = \left(\tfrac12 + \iu\tfrac{\sqrt3}{2}\right)(2+\iu)
= \left(1 - \tfrac{\sqrt3}{2}\right) + \iu\left(\tfrac12 + \sqrt3\right),
\]
so $c_1 = \left(2 - \frac{\sqrt3}{2}\right) +
\iu\left(\frac32 + \sqrt3\right)$, and similarly with the angle
$-\frac\pi3$:
$c_2 = \left(2 + \frac{\sqrt3}{2}\right) + \iu\left(\frac32 - \sqrt3\right)$.
\end{solution}

\begin{solution}{exo:g12:complex:8}
$-4 = 4\,\eu^{\iu\pi}$, so the solutions of $z^4 = -4$ are
\[
z_k = \sqrt2\,\eu^{\iu\left(\frac\pi4 + \frac{k\pi}{2}\right)},
\quad k = 0, 1, 2, 3,
\]
\emph{i.e.} $1 + \iu$, $-1 + \iu$, $-1 - \iu$, $1 - \iu$: the vertices of a
square of circumradius $\sqrt 2$. Grouping \omterm{def:g12:complex:conjugate}{conjugate} \omterm{def:g11:quad:discriminant}{roots},
\[
z^4 + 4 = \bigl(z - (1{+}\iu)\bigr)\bigl(z - (1{-}\iu)\bigr)
\bigl(z + (1{-}\iu)\bigr)\bigl(z + (1{+}\iu)\bigr)
= (z^2 - 2z + 2)(z^2 + 2z + 2).
\]
\end{solution}

\begin{solution}{exo:g12:complex:9}
With $q = \eu^{\iu\theta} \neq 1$:
\[
\sum_{k=0}^{n} \eu^{\iu k\theta} = \frac{\eu^{\iu(n+1)\theta} - 1}{\eu^{\iu\theta} - 1}
= \frac{\eu^{\iu(n+1)\theta/2}}{\eu^{\iu\theta/2}}
\cdot\frac{\eu^{\iu(n+1)\theta/2} - \eu^{-\iu(n+1)\theta/2}}{\eu^{\iu\theta/2} - \eu^{-\iu\theta/2}}
= \eu^{\iu n\theta/2}\,
\frac{\sin\frac{(n+1)\theta}{2}}{\sin\frac{\theta}{2}},
\]
using $\eu^{\iu\alpha} - \eu^{-\iu\alpha} = 2\iu\sin\alpha$ (the
\emph{half-angle \omterm{def:g10:algebra:expand}{factoring}}). Taking \omterm{def:g12:complex:def}{real parts}:
\[
S = \frac{\sin\frac{(n+1)\theta}{2}}{\sin\frac{\theta}{2}}\,
\cos\frac{n\theta}{2}.
\]
\end{solution}

\begin{solution}{exo:g12:complex:10}
\emph{1.} $\omega$ is a $5$-th \omterm{def:g11:quad:discriminant}{root} of unity different from $1$, so the sum
of all five \omterm{def:g11:quad:discriminant}{roots} vanishes (\cref{thm:g12:complex:rootsofunity}), which reads
$1 + \omega + \omega^2 + \omega^3 + \omega^4 = 0$.

\emph{2.} $u + v = \omega + \omega^2 + \omega^3 + \omega^4 = -1$ by point 1.
For the product, using $\omega^5 = 1$:
\[
uv = (\omega + \omega^4)(\omega^2 + \omega^3)
= \omega^3 + \omega^4 + \omega^6 + \omega^7
= \omega^3 + \omega^4 + \omega + \omega^2 = -1 .
\]
So $u, v$ are the \omterm{def:g11:quad:discriminant}{roots} of $X^2 + X - 1 = 0$:
$\left\{\frac{-1+\sqrt5}{2}, \frac{-1-\sqrt5}{2}\right\}$. Now
$u = \omega + \conj\omega = 2\cos\frac{2\pi}{5} > 0$ (since
$\frac{2\pi}{5} < \frac{\pi}{2}$), so $u = \frac{-1 + \sqrt5}{2}$.

\emph{3.} $\cos\frac{2\pi}{5} = \frac u2 = \frac{\sqrt5 - 1}{4}$.
\end{solution}

\begin{solution}{pb:g12:complex:1}
\textbf{1.} $(2 + \iu)(3 - \iu) = 6 - 2\iu + 3\iu + 1 =
7 + \iu$. \quad $\frac{1 + \iu}{1 - \iu} =
\frac{(1 + \iu)^2}{2} = \frac{2\iu}{2} = \iu$. \quad
$\abs{3 + 4\iu} = 5$.

\textbf{2.} $1 + \iu = \sqrt2\,\eu^{\iu\pi/4}$;
$-2 = 2\eu^{\iu\pi}$;
$1 - \iu\sqrt3 = 2\eu^{-\iu\pi/3}$.

\textbf{3.} $(1 + \iu)^8 = \left(\sqrt2\right)^8
\eu^{8\iu\pi/4} = 16\,\eu^{2\iu\pi} = 16$.

\textbf{4.} $z^2 = \eu^{\iu\pi/2}$:
$z = \pm\eu^{\iu\pi/4} = \pm\frac{\sqrt2}{2}(1 + \iu)$. And
$\Delta = 4 - 20 = -16$: $z = 1 \pm 2\iu$.

\textbf{5.} $z \mapsto \iu z$: rotation by a quarter turn
about the origin. $\abs{z - (1 + \iu)} = 2$: the circle of
center $1 + \iu$ and radius $2$.

\textbf{6.} $\left(\omega^k\right)^5 = \eu^{2\iu\pi k} = 1$,
and the five points $\eu^{2\iu k\pi/5}$ are distinct: all five
solutions of $z^5 = 1$. They sit on the unit circle at equal
angles of $72^\circ$: a regular pentagon with one vertex at
$1$.

\textbf{7.} $S = 1 + \omega + \dots + \omega^4$ satisfies
$(1 - \omega)S = 1 - \omega^5 = 0$, and $\omega \neq 1$:
$S = 0$. (The pentagon's balance point is its center.)

\textbf{8.} $\omega^4 = \overline\omega$ and
$\omega^3 = \overline{\omega^2}$: pairing \omterm{def:g12:complex:conjugate}{conjugates}, the \omterm{def:g12:complex:def}{real
part} of the sum reads
$1 + 2\cos\frac{2\pi}{5} + 2\cos\frac{4\pi}{5} = 0$.

\textbf{9.} $\cos\frac{4\pi}{5} = 2c^2 - 1$, so
$1 + 2c + 4c^2 - 2 = 0$, i.e.\ $4c^2 + 2c - 1 = 0$:
$c = \frac{-1 + \sqrt5}{4}$ (the positive \omterm{def:g11:quad:discriminant}{root}, since
$72^\circ$ is acute):
$\cos 72^\circ = \frac{\sqrt5 - 1}{4}$.

\textbf{10.} Numerically $c \approx 0.30902$: matches the
calculator. And $\frac{1}{2\varphi} = \frac{1}{1 + \sqrt5} =
\frac{\sqrt5 - 1}{4}$ (rationalize): $c = \frac{1}{2\varphi}$.
The pentagon is golden through and through: its diagonal cuts
its side in the ratio $\varphi$ --- the pentagram's endless
self-similarity.

\textbf{11.} Square roots are compass-constructible (the
half-circle machine of the Middle School volume's geometric
\omterm{def:g11:stat:mean}{mean}), so a number built from rationals by square roots alone
--- like $\cos 72^\circ$ --- is constructible; cube roots are
not, and $\cos 20^\circ$ needs one: the pentagon yields,
trisection resists. Gauss's criterion settles every regular
polygon --- his 17-gon at nineteen made him choose
mathematics.

\textbf{12.} $-4 = 4\eu^{\iu\pi}$: the fourth \omterm{def:g11:quad:discriminant}{roots} have
\omterm{def:g12:complex:modulus}{modulus} $\sqrt2$ and \omterm{def:g12:complex:argument}{arguments}
$\frac{\pi}{4} + k\frac{\pi}{2}$: the four numbers
$\pm 1 \pm \iu$. Pairing \omterm{def:g12:complex:conjugate}{conjugates}:
$(z - 1 - \iu)(z - 1 + \iu) = z^2 - 2z + 2$ and
$(z + 1 - \iu)(z + 1 + \iu) = z^2 + 2z + 2$: the
factorization of \cref{exo:g12:complex:8}.

\textbf{13.} $(\cos\theta + \iu\sin\theta)^3 = \cos^3\theta +
3\iu\cos^2\theta\sin\theta - 3\cos\theta\sin^2\theta -
\iu\sin^3\theta$; by de Moivre this equals
$\cos 3\theta + \iu \sin 3\theta$. \omterm{def:g12:complex:def}{Real parts}:
$\cos 3\theta = \cos^3\theta - 3\cos\theta(1 - \cos^2\theta)
= 4\cos^3\theta - 3\cos\theta$ --- and the \omterm{def:g12:complex:def}{imaginary parts}
hand over $\sin 3\theta$ for free.

\textbf{14.} It is the \omterm{def:g12:complex:def}{real part} of
$1 + \omega + \dots + \omega^4 = 0$: the sum equals $0$ ---
exactly question 8, seen through
\cref{exo:g12:complex:9}'s kernel.

\textbf{15.} $8\iu = 8\eu^{\iu\pi/2}$: \omterm{def:g11:quad:discriminant}{roots}
$2\eu^{\iu\pi/6} = \sqrt3 + \iu$;
$2\eu^{5\iu\pi/6} = -\sqrt3 + \iu$;
$2\eu^{3\iu\pi/2} = -2\iu$.

\textbf{16.} $1 + \iu = \sqrt2\,\eu^{\iu\pi/4}$: the map
rotates the plane by $45^\circ$ and scales it by $\sqrt2$. The
unit square becomes a tilted square of side $\sqrt2$ ---
doubled in area, spun an eighth of a turn.

\textbf{17.} $\abs{z_1 z_2}^2 = z_1 z_2 \overline{z_1 z_2} =
z_1\overline{z_1}\, z_2\overline{z_2} =
\abs{z_1}^2\abs{z_2}^2$: take square roots. Powers of
$1 + \iu$: $1 + \iu$, $2\iu$, $-2 + 2\iu$, $-4$: each a
$45^\circ$ turn further and $\sqrt2$ times longer --- an
outward spiral through moduli
$\sqrt2, 2, 2\sqrt2, 4, \dots$

\textbf{18.} $\mathrm{j}$ is a third \omterm{def:g11:quad:discriminant}{root} of unity
distinct from $1$, so (question 7's \omterm{def:g12:complex:argument}{argument} with $n = 3$)
$1 + \mathrm{j} + \mathrm{j}^2 = 0$. Criterion on the standard
triangle: $a + \mathrm{j}b + \mathrm{j}^2 c = 1 +
\mathrm{j}\cdot\mathrm{j} + \mathrm{j}^2 \cdot \mathrm{j}^2 =
1 + \mathrm{j}^2 + \mathrm{j}^4 = 1 + \mathrm{j}^2 +
\mathrm{j} = 0$: satisfied, as the equilateral triangle
$\left(1, \mathrm{j}, \mathrm{j}^2\right)$ demands.

\textbf{19.} Over $\C$, a real polynomial splits into linear
factors; its non-real \omterm{def:g11:quad:discriminant}{roots} come in \omterm{def:g12:complex:conjugate}{conjugate} pairs (\omterm{def:g12:complex:conjugate}{conjugate}
the \omterm{def:g10:algebra:equation}{equation}), and each pair multiplies into a real quadratic
$z^2 - 2\operatorname{Re}(z_0)\,z + \abs{z_0}^2$: hence the
real factorization into degree-$1$ and degree-$2$ pieces ---
question 12 performed it on $z^4 + 4$.

\textbf{20.} Points; rotation-and-scaling; regular polygons;
trigonometric identities by expansion; and a closed world
where every \omterm{def:g10:algebra:equation}{equation} of degree $n$ has its $n$ \omterm{def:g11:quad:discriminant}{roots}. Coda:
in question 10, $\pi$'s pentagon shook hands with the golden
ratio --- Euclid's two most famous guests, introduced at last
by a sum of five arrows equal to zero.
\end{solution}
