\section*{Chapter \ref{ch:g1:shapes} --- Shapes and Space}

\begin{solution}{exo:g1:shapes:1}
$3$ corners: a triangle. Round, no corner: a circle. $4$ equal
sides: a square.
\end{solution}

\begin{solution}{exo:g1:shapes:2}
Answers vary: a door and a book are rectangles; a plate or a clock
is a circle; a road-warning sign is a triangle; a tile can be a
square.
\end{solution}

\begin{solution}{exo:g1:shapes:3}
False: a square turned on its corner still has four equal sides and
four corners --- turning does not change the shape.
\end{solution}

\begin{solution}{exo:g1:shapes:4}
Grid drawings: the square is $3 \times 3$ squares, the rectangle
$5 \times 2$, and the triangle uses a slanted grid line as one
side.
\end{solution}

\begin{solution}{exo:g1:shapes:5}
Answers vary. Example: the ball is on the left, the car is between
the ball and the doll, the doll is on the right.
\end{solution}

\begin{solution}{exo:g1:shapes:6}
The cross is drawn inside the circle, the star outside, and the dot
right on the circle's line.
\end{solution}

\begin{solution}{exo:g1:shapes:7}
Following $\to\ \uparrow\ \to\ \uparrow\ \to$ moves $3$ squares
right and $2$ up. The way back is $\downarrow\ \downarrow$ then
$\leftarrow\ \leftarrow\ \leftarrow$ (or any path returning to the
start).
\end{solution}

\begin{solution}{exo:g1:shapes:8}
Two triangles and one square: $3 + 3 + 4 = 10$ corners. One circle
and two rectangles: $0 + 4 + 4 = 8$ corners.
\end{solution}

\begin{solution}{exo:g1:shapes:9}
$12$ matches make a square with $3$ matches on each side
($3 + 3 + 3 + 3 = 12$). With $10$ matches it cannot be done: $10$
does not split into $4$ equal whole parts (each side would need
$2$ matches and a half).
\end{solution}
