\section*{Chapter \ref{ch:g11:func} --- Functions and Variations}

\begin{solution}{exo:g11:func:1}
$f$: need $3 - x \geq 0$, \omterm{def:g11:func:function}{domain} $\intoc{-\infty}{3}$.

$g$: need $x^2 - 4 \neq 0$, i.e.\ $x \neq \pm 2$: \omterm{def:g11:func:function}{domain}
$\R \setminus \{-2, 2\}$.

$h$: need $x \geq 0$ and $x \neq 5$: \omterm{def:g11:func:function}{domain}
$\intco{0}{5} \cup \intoo{5}{+\infty}$.
\end{solution}

\begin{solution}{exo:g11:func:2}
$\sqrt{11} < \sqrt{13}$ since $\sqrt{}$ is \omterm{def:g11:func:monotone}{increasing} and $11 < 13$.

$(-2.1)^2 > (-2.05)^2$ since $x^2$ is \omterm{def:g10:functions:variations}{decreasing} on
$\intoc{-\infty}{0}$ and $-2.1 < -2.05$.

$\frac{1}{0.9} > \frac{1}{1.1}$ since $\frac1x$ is \omterm{def:g10:functions:variations}{decreasing} on
$\intoo{0}{+\infty}$ and $0.9 < 1.1$.

$(-1.01)^3 < (-0.99)^3$ since $x^3$ is \omterm{def:g11:func:monotone}{increasing} and $-1.01 < -0.99$.
\end{solution}

\begin{solution}{exo:g11:func:3}
$f(-x) = 3(-x)^4 - (-x)^2 = 3x^4 - x^2 = f(x)$: \omterm{def:g11:func:parity}{even}.

$g(-x) = -x^3 - x = -g(x)$: \omterm{def:g11:func:parity}{odd}.

$h(1) = 2$ and $h(-1) = 1 - 1 = 0$, so $h(-1)$ is neither $h(1)$ nor
$-h(1)$: neither \omterm{def:g11:func:parity}{even} nor \omterm{def:g11:func:parity}{odd}.

$k(-x) = \frac{-x}{x^2+1} = -k(x)$: \omterm{def:g11:func:parity}{odd}.
\end{solution}

\begin{solution}{exo:g11:func:4}
Read the \omterm{def:g10:functions:table}{variation table}. On $\intcc{-3}{1}$, $f$ climbs from $-2$ to $4$
and crosses the value $1$ exactly once; on $\intcc{1}{5}$ it goes down
from $4$ to $0$ and crosses $1$ once more: the \omterm{def:g10:algebra:equation}{equation} $f(x) = 1$ has
$2$ solutions. The \omterm{def:g10:functions:extrema}{maximum} of $f$ is $f(1) = 4 < 5$, so $f(x) = 5$ has no
solution.
\end{solution}

\begin{solution}{exo:g11:func:5}
Let $2 \leq u < v$. Then
\[
f(v) - f(u) = v^2 - 4v - u^2 + 4u = (v-u)(v+u) - 4(v-u)
= (v - u)(u + v - 4).
\]
Both factors are positive: $v - u > 0$, and $u + v > 2 + 2 = 4$ since
$u \geq 2$ and $v > 2$. Hence $f(v) > f(u)$: $f$ is strictly \omterm{def:g11:func:monotone}{increasing}
on $\intco{2}{+\infty}$.
\end{solution}

\begin{solution}{exo:g11:func:6}
$f = \sqrt{u}$ with $u(x) = 5 - x$ \omterm{def:g10:functions:variations}{decreasing} and nonnegative on
$\intoc{-\infty}{5}$: $f$ is \omterm{def:g10:functions:variations}{decreasing} there
(\cref{prop:g11:func:transforms}).

$g = \frac1u$ with $u(x) = x^2 + 1$ \omterm{def:g11:func:monotone}{increasing} and positive on
$\intco{0}{+\infty}$: $g$ is \omterm{def:g10:functions:variations}{decreasing} there.

$h = 3 - 2\sqrt x$: $\sqrt x$ is \omterm{def:g11:func:monotone}{increasing}, multiplying by $-2 < 0$
reverses, adding $3$ changes nothing: $h$ is \omterm{def:g10:functions:variations}{decreasing} on
$\intco{0}{+\infty}$.
\end{solution}

\begin{solution}{exo:g11:func:7}
\emph{1.} $f$ is in \omterm{thm:g11:quad:canonical}{canonical form}: \omterm{def:g10:functions:variations}{decreasing} on $\intoc{-\infty}{2}$,
\omterm{def:g11:func:monotone}{increasing} on $\intco{2}{+\infty}$, \omterm{def:g10:functions:extrema}{minimum} $f(2) = -3$.

\emph{2.} $g = \frac{1}{u}$ with $u(x) = (x-2)^2 + 1 = f(x) + 4$. The
shift by $4$ keeps the variations of $f$, and $u \geq 1 > 0$. Taking the
reciprocal reverses them: $g$ is \omterm{def:g11:func:monotone}{increasing} on $\intoc{-\infty}{2}$ and
\omterm{def:g10:functions:variations}{decreasing} on $\intco{2}{+\infty}$, with \omterm{def:g10:functions:extrema}{maximum}
$g(2) = \frac{1}{1} = 1$.
\end{solution}

\begin{solution}{exo:g11:func:8}
Any hill-shaped curve through the three points works. Then:
$f + 2$ passes through $(0,3)$, $(2,5)$, $(4,3)$, same variations
(vertical shift); $-f$ passes through $(0,-1)$, $(2,-3)$, $(4,-1)$ with
\emph{reversed} variations (the peak becomes a valley); $2f$ passes
through $(0,2)$, $(2,6)$, $(4,2)$, same variations but amplified
vertically.
\end{solution}

\begin{solution}{exo:g11:func:9}
$2 - \frac{1}{x+1} = \frac{2(x+1) - 1}{x+1} = \frac{2x+1}{x+1} = f(x)$.
On $\intoo{-1}{+\infty}$, $u(x) = x + 1$ is \omterm{def:g11:func:monotone}{increasing} and positive, so
$\frac1u$ is \omterm{def:g10:functions:variations}{decreasing}, $-\frac1u$ is \omterm{def:g11:func:monotone}{increasing}, and adding $2$ changes
nothing: $f$ is strictly \omterm{def:g11:func:monotone}{increasing} on $\intoo{-1}{+\infty}$.
\end{solution}

\begin{solution}{exo:g11:func:10}
\emph{1. True.} If $u < v$, then
$(f+g)(v) - (f+g)(u) = \bigl(f(v)-f(u)\bigr) + \bigl(g(v)-g(u)\bigr)$ is
a sum of two nonnegative numbers.

\emph{2. False.} $f(x) = g(x) = x$ are \omterm{def:g11:func:monotone}{increasing} on $\R$, but their
product $x^2$ is not \omterm{def:g11:func:monotone}{increasing} on $\R$ (it decreases on
$\intoc{-\infty}{0}$). (The statement becomes true if both \omterm{def:g11:func:function}{functions} are
also nonnegative.)

\emph{3. False.} Same counterexample: $f(x) = x$ is \omterm{def:g11:func:monotone}{increasing} on $\R$
but $f^2(x) = x^2$ is not.
\end{solution}

\begin{solution}{exo:g11:func:11}
\emph{1.} For $0 < u < v$:
\[
f(v) - f(u) = (v - u) + \frac1v - \frac1u
= (v - u) + \frac{u - v}{uv}
= (v - u)\left(1 - \frac{1}{uv}\right)
= (v - u)\,\frac{uv - 1}{uv}.
\]

\emph{2.} Always $v - u > 0$ and $uv > 0$. If $0 < u < v \leq 1$, then
$uv < 1$, so $uv - 1 < 0$ and $f(v) < f(u)$: $f$ is strictly \omterm{def:g10:functions:variations}{decreasing}
on $\intoc{0}{1}$. If $1 \leq u < v$, then $uv > 1$ and $f(v) > f(u)$:
$f$ is strictly \omterm{def:g11:func:monotone}{increasing} on $\intco{1}{+\infty}$.

\emph{3.} $f$ decreases before $1$ and increases after, so
$f(x) \geq f(1) = 2$ for all $x > 0$, with equality exactly at $x = 1$.
(This inequality reappears in many guises; compare
\cref{ex:g11:quad:canonical} and the tangent-line inequalities of
\cref{ch:g12:deriv}.)
\end{solution}

\begin{solution}{pb:g11:func:1}
\textbf{1.} $x^4 - 3x^2$: \omterm{def:g11:func:parity}{even} ($(-x)^4 - 3(-x)^2 = x^4 -
3x^2$). $x^3 - x$: \omterm{def:g11:func:parity}{odd}. $x^2 + x$: neither ($f(-1) = 0 \neq
f(1) = 2$ and $\neq -f(1)$). $\abs x$: \omterm{def:g11:func:parity}{even}. $\frac1x$: \omterm{def:g11:func:parity}{odd}.

\textbf{2.} \omterm{def:g11:func:parity}{Even}: the curve is symmetric across the $y$-axis
(mirror). \omterm{def:g11:func:parity}{Odd}: symmetric about the origin (half-turn).

\textbf{3.} If $f$ and $g$ are \omterm{def:g11:func:parity}{odd}:
$(fg)(-x) = f(-x)g(-x) = (-f(x))(-g(x)) = (fg)(x)$: \omterm{def:g11:func:parity}{even}.
Table: \omterm{def:g11:func:parity}{even} $\times$ \omterm{def:g11:func:parity}{even} $=$ \omterm{def:g11:func:parity}{even}; \omterm{def:g11:func:parity}{even} $\times$ \omterm{def:g11:func:parity}{odd} $=$ \omterm{def:g11:func:parity}{odd};
\omterm{def:g11:func:parity}{odd} $\times$ \omterm{def:g11:func:parity}{odd} $=$ \omterm{def:g11:func:parity}{even} --- the sign rule with parities.

\textbf{4.} \omterm{def:g11:func:parity}{Odd} at $0$: $f(0) = -f(0)$, so $f(0) = 0$. \omterm{def:g11:func:parity}{Even}:
$f(-3) = f(3) = 7$. Both \omterm{def:g11:func:parity}{even} and \omterm{def:g11:func:parity}{odd}:
$f(x) = f(-x) = -f(x)$ for all $x$, so $2f(x) = 0$: $f$ is the
zero \omterm{def:g11:func:function}{function}.

\textbf{5.} The right half ($x \geq 0$) determines an \omterm{def:g11:func:parity}{even
function} completely: the left half is its mirror. Neither \omterm{def:g11:func:parity}{even}
nor \omterm{def:g11:func:parity}{odd}: $x^2 + x$ (question 1) --- its two symmetries fail at
$x = 1$.

\textbf{6.} $E(x) = \frac{(x^3 + x^2 + 1) + (-x^3 + x^2 +
1)}{2} = x^2 + 1$ (\omterm{def:g11:func:parity}{even}), $O(x) = \frac{(x^3 + x^2 + 1) -
(-x^3 + x^2 + 1)}{2} = x^3$ (\omterm{def:g11:func:parity}{odd}); their sum is $f$.

\textbf{7.} $E(-x) = \frac{f(-x) + f(x)}{2} = E(x)$: \omterm{def:g11:func:parity}{even}.
$O(-x) = \frac{f(-x) - f(x)}{2} = -O(x)$: \omterm{def:g11:func:parity}{odd}. And
$E(x) + O(x) = f(x)$ by direct addition: every \omterm{def:g11:func:function}{function}
splits.

\textbf{8.} From $E_1 + O_1 = E_2 + O_2$:
$E_1 - E_2 = O_2 - O_1$. The left side is \omterm{def:g11:func:parity}{even}, the right side
\omterm{def:g11:func:parity}{odd} --- so this \omterm{def:g11:func:function}{function} is both, hence zero (question 4):
$E_1 = E_2$ and $O_1 = O_2$. One split only.

\textbf{9.} $E(x) = \frac12\left(\frac{1}{1 - x} + \frac{1}{1 +
x}\right) = \frac12 \cdot \frac{(1 + x) + (1 - x)}{1 - x^2}
= \frac{1}{1 - x^2}$, and
$O(x) = \frac12 \cdot \frac{(1 + x) - (1 - x)}{1 - x^2}
= \frac{x}{1 - x^2}$. Check: $E$ is \omterm{def:g11:func:parity}{even}, $O$ is \omterm{def:g11:func:parity}{odd}, and
their sum is $\frac{1 + x}{1 - x^2} =
\frac{1 + x}{(1 - x)(1 + x)} = \frac{1}{1 - x} = f(x)$.

\textbf{10.} Because each piece obeys a symmetry, half the
work (and half the data) suffices for each: solve on one side,
mirror to the other --- and many operations (integration,
series, signal analysis) treat symmetric pieces far more
simply than raw \omterm{def:g11:func:function}{functions}.

\textbf{11.} $(x-3)^2$: shifted $3$ right. $x^2 + 2$: shifted
$2$ up. $(x-3)^2 + 2$: both. $-x^2$: flipped upside down.
$2x^2$: stretched vertically by $2$ (narrower bowl).

\textbf{12.} For $x \geq 2$: $(x - 2) + (x + 2) = 2x$. For
$-2 \leq x \leq 2$: $(2 - x) + (x + 2) = 4$. For $x \leq -2$:
$-2x$. The curve descends with \omterm{def:g10:reffunc:affine}{slope} $-2$, runs flat at height
$4$ between $-2$ and $2$, then climbs with \omterm{def:g10:reffunc:affine}{slope} $2$: a valley
with a flat floor.

\textbf{13.} Vertical shift: $f(x) + 2$ stays \omterm{def:g11:func:parity}{even}
($f(-x) + 2 = f(x) + 2$). Vertical stretch: $2f(x)$ stays
\omterm{def:g11:func:parity}{even}. Horizontal shift: $(x - 3)^2$ is not \omterm{def:g11:func:parity}{even}
($f(-1) = 16 \neq 4 = f(1)$): sliding sideways destroys the
mirror.

\textbf{14.} On the flat floor the value is $4 \neq 6$. For
$x \geq 2$: $2x = 6$, $x = 3$. For $x \leq -2$: $-2x = 6$,
$x = -3$. Solutions: $\pm 3$.

\textbf{15.} $f(x) = a(x - 2)^2 - 3$ with $f(0) = 4a - 3 = 1$:
$a = 1$, so $f(x) = (x - 2)^2 - 3$.

\textbf{16.} $g(v) - g(u) = (v - u) + 4\,\frac{u - v}{uv}
= (v - u)\,\frac{uv - 4}{uv}$: negative for $u < v$ with
$uv < 4$, positive with $uv > 4$: \omterm{def:g10:functions:variations}{decreasing} on
$\intoc{0}{2}$, \omterm{def:g11:func:monotone}{increasing} on $\intco{2}{+\infty}$; \omterm{def:g10:functions:extrema}{minimum} at
$x = 2$, of value $g(2) = 4$.

\textbf{17.} Sides $x$ and $\frac{16}{x}$:
$P(x) = 2\left(x + \frac{16}{x}\right)$, minimized (question
16's machine with $4 \to 16$, turning point $\sqrt{16} = 4$)
at $x = 4$: the $4 \times 4$ \emph{square}, perimeter $16$ ---
Dido's answer, now proved for every real side length.

\textbf{18.} On $\intco{0}{+\infty}$: $x \mapsto x^2 + 1$
increases, and $\sqrt{\phantom{x}}$ increases, so the
composite increases. And $h(-x) = \sqrt{x^2 + 1} = h(x)$: $h$
is \omterm{def:g11:func:parity}{even}, so by mirror symmetry it decreases on
$\intoc{-\infty}{0}$. \omterm{def:g10:functions:extrema}{Minimum} $h(0) = 1$.

\textbf{19.} Squared distance to $(3, 0)$ of the point
$(x, \sqrt x)$:
$d^2(x) = (x - 3)^2 + x = x^2 - 5x + 9$, minimal at
$x = \frac52$: closest point
$\left(\frac52, \sqrt{\frac52}\right)$, distance
$\sqrt{\frac{11}{4}} = \frac{\sqrt{11}}{2}$. Minimizing $d^2$
is legitimate because squaring is strictly \omterm{def:g11:func:monotone}{increasing} on
positive numbers: $d$ and $d^2$ bottom out at the same spot.

\textbf{20.} Raw stock: the reference curves, known by heart.
Machines: shifts, flips and stretches assembling new \omterm{def:g10:functions:graph}{graphs}
from old. Quality control: parity and the even--odd split,
which halve every study. Measuring bench: \omterm{def:g10:algebra:expand}{factoring}
$f(v) - f(u)$, which reads variations and minima directly. The
next chapter's power tool: the \emph{derivative}, which reads
them at a glance.
\end{solution}
