\chapter{Sequences and Series}\label{ch:b2:series}

The theory of numerical series (Year 1 volume) matures here in three
directions: series with values in \omterm{def:b2:nvs:banach}{Banach spaces}, where absolute
convergence does the work; the finer tests for real series (\omterm{thm:b2:series:abel}{Abel
summation}); and \emph{\omterm{def:b2:series:summable}{summable} families} --- summation freed from
the order of the terms --- with the Fubini theorem for double sums
and the \omterm{thm:b2:series:fubini}{Cauchy product}. These tools carry all the function-series
chapters ahead.

\section{Series in normed spaces}

\begin{definition}\label{def:b2:series:def}
For a sequence $(u_n)$ in a normed space $E$, the series $\sum u_n$
converges when its partial sums do; it converges
\emph{absolutely}\index{absolute convergence!in a Banach space} when
$\sum \norm{u_n} < \infty$. In a \emph{Banach} space, absolute
convergence implies convergence (\cref{thm:b2:nvs:absoluteconvergence});
in a \omterm{def:b2:metric:complete}{non-complete} space this may fail
(\cref{exo:b2:series:9}).
\end{definition}

\begin{example}\label{ex:b2:series:neumann}
In $\mathcal{M}_n(K)$ (or $\mathcal{L}_c(E)$, $E$ Banach): for
$\vertiii A < 1$, the \emph{Neumann series} $\sum A^k$ converges
\omterm{def:b2:series:def}{absolutely} to $(I - A)^{-1}$ (proved in \cref{exo:b2:nvs:5});
$\sum \frac{A^k}{k!}$ converges \omterm{def:b2:series:def}{absolutely} to $\eu^A$ for every $A$
(\cref{ex:b2:nvs:matrixexp}). Operator-valued geometric and
exponential series behave like their scalar models --- the whole
point of the Banach framework.
\end{example}

\section{Abel summation}

\begin{theorem}[Abel's summation and test]\label{thm:b2:series:abel}
(Summation by parts) For scalars $a_n$ and vectors $b_n$, with $B_n
= \sum_{k=0}^{n} b_k$:\index{Abel summation}
\[
\sum_{n=0}^{N} a_n b_n
= a_N B_N - \sum_{n=0}^{N-1} (a_{n+1} - a_n) B_n .
\]
(Abel's test) If $(a_n)$ is a real sequence, decreasing to $0$, and
the partial sums $B_n$ are \emph{bounded} (in a \omterm{def:b2:nvs:banach}{Banach space}), then
$\sum a_n b_n$ converges.
\end{theorem}

\begin{proof}
The identity, step by step: with $B_{-1} = 0$, write $b_n = B_n -
B_{n-1}$ and split,
\[
\sum_{n=0}^{N} a_nb_n
= \sum_{n=0}^{N} a_nB_n - \sum_{n=0}^{N} a_nB_{n-1}
= \sum_{n=0}^{N} a_nB_n - \sum_{n=0}^{N-1} a_{n+1}B_{n} ,
\]
reindexing the second sum by $n \mapsto n + 1$ (the $B_{-1}$
term vanishes); collecting the common range $0 \leq n \leq N-1$
leaves $a_NB_N$ plus $\sum_{n\leq N-1}(a_n - a_{n+1})B_n$: the
stated formula. It is the discrete integration by parts, with
$(B_n)$ as antiderivative of $(b_n)$ and the difference $a_{n+1}
- a_n$ as derivative of $(a_n)$. For the test, with $\norm{B_n}
\leq M$: the
boundary term $a_N B_N \to 0$; the series $\sum (a_n -
a_{n+1})B_n$ converges \omterm{def:b2:series:def}{absolutely}, since
\[
\sum_n \norm{(a_{n+1} - a_n)B_n} \leq M \sum_n (a_n - a_{n+1})
= M a_0 < \infty
\]
(telescoping, $a_n \downarrow 0$). Both pieces of the identity
converge, hence so does $\sum a_n b_n$.
\end{proof}

\begin{example}\label{ex:b2:series:sinn}
$\sum \frac{\sin n}{n}$ converges: $a_n = \frac1n \downarrow 0$ and
$B_n = \sum_{k=1}^{n} \sin k$ is bounded --- indeed $B_n =
\Im\sum_{k \leq n} \eu^{\iu k} = \Im\,\frac{\eu^{\iu}(\eu^{\iu n} -
1)}{\eu^{\iu} - 1}$, of modulus $\leq \frac{2}{\abs{\eu^{\iu} - 1}}$.
It does \emph{not} converge \omterm{def:b2:series:def}{absolutely} ($\abs{\sin n} \geq \sin^2 n
= \frac{1 - \cos 2n}{2}$, and $\sum \frac{1 - \cos 2n}{2n}$ diverges
since $\sum \frac{\cos 2n}{n}$ converges by the same Abel test while
$\sum \frac{1}{2n}$ diverges). The alternating series test is the
special case $b_n = (-1)^n$.
\end{example}

\begin{example}[Abel on the circle of convergence]\label{ex:b2:series:abelboundary}
For which complex $z$ with $\abs z = 1$ does $\sum_{n \geq 1}
\frac{z^n}{n}$ converge? At $z = 1$ it is the harmonic series:
divergent. For $z \neq 1$ on the circle, Abel's test applies
with $a_n = \frac1n \downarrow 0$ and $b_n = z^n$, whose partial
sums are bounded independently of $N$:
\[
\Bigl|\sum_{n=1}^{N} z^n\Bigr|
= \Bigl|\frac{z(z^N - 1)}{z - 1}\Bigr|
\leq \frac{2}{\abs{z - 1}} .
\]
Convergent --- though never \omterm{def:b2:series:def}{absolutely} ($\sum\frac1n$). One
series, a circle of behaviours: divergence at a single point,
semi-convergence everywhere else. This is the standard boundary
behaviour of power series (\cref{ch:b2:powerseries}), met here
with bare hands; at $z = -1$ it recovers the alternating
harmonic series, and at $z = \eu^{\iu\theta}$ its real and
imaginary parts are the series $\sum\frac{\cos
n\theta}{n}$ and $\sum\frac{\sin n\theta}{n}$ of
\cref{exo:b2:series:4}.
\end{example}

\begin{example}[A booby-trapped alternating series]\label{ex:b2:series:trap}
Does $\sum_{n\geq2} \dfrac{(-1)^n}{\sqrt n + (-1)^n}$ converge?
The signs alternate and the terms tend to $0$ --- yet the
alternating test does \emph{not} apply: the moduli
$\frac{1}{\sqrt n + (-1)^n}$ are not decreasing (they jump up at
each odd $n$). Expand instead:
\[
\frac{(-1)^n}{\sqrt n + (-1)^n}
= \frac{(-1)^n}{\sqrt n}\cdot
\frac{1}{1 + \frac{(-1)^n}{\sqrt n}}
= \frac{(-1)^n}{\sqrt n} - \frac{1}{n}
+ O\Bigl(\frac{1}{n^{3/2}}\Bigr).
\]
The first piece converges (alternating test, honestly applied to
$\frac1{\sqrt n}\downarrow0$), the third converges \omterm{def:b2:series:def}{absolutely}
--- but the middle piece is the divergent harmonic series: the
sum \emph{diverges} to $-\infty$. The closing insight: when
monotonicity fails, expand until every piece is either
\omterm{def:b2:series:def}{absolutely} convergent or a clean test case; the hidden
$-\frac1n$ is invisible to sign-counting.
\end{example}

\begin{remark}[Common pitfalls]\label{rem:b2:series:pitfalls}
(i) ``Terms tend to $0$'' proves nothing: the harmonic series
diverges. (ii) The alternating test requires \emph{decreasing}
moduli --- \cref{ex:b2:series:trap} is the canonical
counterexample, and \cref{exo:b2:series:1}'s third series the
drill. (iii) Conditionally convergent series may not be
rearranged (\cref{ex:b2:series:rearrange}), and their \omterm{thm:b2:series:fubini}{Cauchy
products} may diverge: for $\sum\frac{(-1)^n}{\sqrt{n+1}}$
squared, the diagonal terms satisfy
\[
\abs{c_k} = \sum_{m=0}^{k}
\frac{1}{\sqrt{(m+1)(k-m+1)}}
\geq (k+1)\cdot\frac{2}{k+2} \longrightarrow 2 \neq 0
\]
(each factor is at most $\frac{k+2}2$ by AM--GM), so $\sum c_k$
diverges --- absolute convergence of at least one factor
(\cref{exo:b2:series:8}) is not a luxury. (iv) \omterm{def:b2:series:summable}{Summability} is
about \emph{absolute} bounds by definition: there is no such
thing as a conditionally \omterm{def:b2:series:summable}{summable family}.
\end{remark}

\section{Summable families}

\begin{definition}\label{def:b2:series:summable}
Let $I$ be a \omterm{def:b2:structures:countable}{countable} index set. A family $(u_i)_{i \in I}$ of
\emph{nonnegative reals} is \emph{summable}\index{summable family}
when the finite partial sums are bounded; its sum is
\[
\sum_{i \in I} u_i = \sup_{F \subseteq I \text{ finite}} \sum_{i
\in F} u_i \in \intcc{0}{+\infty} .
\]
A family of reals or complexes (or Banach vectors) is summable when
$(\norm{u_i})$ is; its sum is then defined by splitting into
positive/negative (or real/imaginary) parts --- equivalently, as the
common value of $\sum_{n} u_{\sigma(n)}$ over all enumerations
$\sigma$ of $I$ (see below).
\end{definition}

\begin{method}[Choosing a test]\label{met:b2:series:choosing}
Facing $\sum u_n$, in order: (1) if $u_n \not\to 0$, divergence,
stop. (2) If the terms have constant sign, compare: find an
equivalent (\cref{ch:b2:comparison}) and place it on the
Riemann--Bertrand map. (3) If signs alternate with
\emph{decreasing} moduli, the alternating test; if the moduli
are not monotone, expand the term until each piece is \omterm{def:b2:series:def}{absolutely}
convergent or a clean test case (\cref{ex:b2:series:trap}). (4)
If the sign pattern is oscillatory but structured ($\sin
n\theta$, $\eu^{\iu n\theta}$, matrix powers), Abel's test with
bounded partial sums. (5) Absolute convergence is always worth
checking first: it is stronger, order-proof, and unlocks \omterm{thm:b2:series:fubini}{Cauchy
products} and Fubini.
\end{method}

\begin{example}[Summability by diagonal counting]\label{ex:b2:series:diagonalcount}
For which $s > 0$ is the family
$\bigl((m + n)^{-s}\bigr)_{m, n \geq 1}$ \omterm{def:b2:series:summable}{summable}? Group the
finite partial sums by diagonals $m + n = k$: the diagonal $k$
carries $k - 1$ pairs, each contributing $k^{-s}$, so the
finite sums are exactly bounded by (and exhaust)
\[
\sum_{k \geq 2} \frac{k - 1}{k^{s}} ,
\]
a series with positive terms equivalent to $k^{1-s}$: \omterm{def:b2:series:summable}{summable}
iff $s - 1 > 1$, i.e.\ $s > 2$. The two-dimensional index eats
one full power: a plane of terms is ``one dimension more
divergent'' than a line --- the counting geometry of the index
set, not the size of individual terms, decides \omterm{def:b2:series:summable}{summability}.
(The same census shows $\bigl((m^2 + n^2)^{-1}\bigr)$ is not
\omterm{def:b2:series:summable}{summable}: on the diagonal $m + n = k$, each term is at least
$k^{-2}$, and $(k-1)\cdot k^{-2}$ sums like the harmonic
series.)
\end{example}

\begin{theorem}[Summability and order]\label{thm:b2:series:rearrangement}
\begin{enumerate}
  \item For nonnegative families, the sum is invariant under any
        enumeration: $\sum_{i} u_i = \sum_{n=0}^{\infty}
        u_{\sigma(n)}$ for every bijection $\sigma \colon \N \to I$.
  \item A real or complex series $\sum u_n$ is \emph{commutatively
        convergent} (every rearrangement converges, with the same
        sum) if and only if it is \omterm{def:b2:series:def}{absolutely} convergent.
\end{enumerate}
\end{theorem}

\begin{proof}
(1) Every partial sum $\sum_{n \leq N} u_{\sigma(n)}$ is a finite
partial sum of the family (so $\leq$ the sup); every finite $F$ is
contained in some $\{\sigma(0), \dots, \sigma(N)\}$ (so the sup
$\leq$ the series' limit). The two bounds match.

(2) If $\sum\abs{u_n} < \infty$: for any rearrangement $\sigma$ and
$\varepsilon > 0$, choose $N$ with $\sum_{n > N}\abs{u_n} \leq
\varepsilon$; beyond the rank where $\sigma$ has exhausted
$\intint{0}{N}$, the rearranged partial sums differ from the
original limit by at most $\varepsilon$: same sum. If $\sum
\abs{u_n} = \infty$ but $\sum u_n$ converges (real case; complex
follows coordinatewise): the positive and negative parts both
diverge, and one can rearrange to reach any prescribed limit
--- Riemann's theorem, carried out in \cref{exo:b2:series:5} ---
so commutative convergence fails.
\end{proof}

\begin{example}[A rearrangement caught red-handed]\label{ex:b2:series:rearrange}
The alternating harmonic series sums to
$\sum_{n\geq1}\frac{(-1)^{n-1}}{n} = \ln 2$ (Year 1
volume). Rearrange it as ``one positive, two negatives'':
\[
1 - \frac12 - \frac14 + \frac13 - \frac16 - \frac18 + \frac15 -
\cdots
\]
Grouping each block of three,
\[
\frac{1}{2k-1} - \frac{1}{4k-2} - \frac1{4k}
= \frac{1}{4k-2} - \frac{1}{4k}
= \frac12\Bigl(\frac{1}{2k-1} - \frac1{2k}\Bigr),
\]
so the rearranged series converges to $\frac12\ln 2$ --- half the
original sum, with exactly the same terms. \omterm{def:b2:series:def}{Non-absolutely}
convergent series remember the order of their terms; \omterm{def:b2:series:summable}{summable}
families are precisely the ones that do not.
\end{example}

\begin{example}[Grouping is safe, ungrouping is not]\label{ex:b2:series:grouping}
Grouping consecutive terms of a \emph{convergent} series never
changes the sum: the grouped partial sums form a subsequence of
the original ones. The converse operation is forbidden:
\[
(1 - 1) + (1 - 1) + (1 - 1) + \cdots = 0 + 0 + \cdots = 0,
\]
yet the ungrouped $1 - 1 + 1 - 1 + \cdots$ diverges (partial
sums oscillate between $1$ and $0$). Ungrouping is legitimate
only with a compensating hypothesis --- for instance, terms
tending to $0$ with bounded group lengths: then between two
grouped partial sums the original ones drift by at most a sum of
$o(1)$ terms of bounded number, and convergence transfers back.
That is exactly the clause under which the block computation of
\cref{ex:b2:series:rearrange} is a proof and not a sleight of
hand.
\end{example}

\begin{theorem}[Fubini for families; Cauchy products]\label{thm:b2:series:fubini}
Let $(u_{m,n})_{(m,n) \in \N^2}$ be a \omterm{def:b2:series:summable}{summable} double family (i.e.\
$\sup_F \sum_F \abs{u_{m,n}} < \infty$). Then\index{Fubini theorem
for series}
\[
\sum_{(m,n)} u_{m,n}
= \sum_{m=0}^{\infty}\Bigl(\sum_{n=0}^{\infty} u_{m,n}\Bigr)
= \sum_{n=0}^{\infty}\Bigl(\sum_{m=0}^{\infty} u_{m,n}\Bigr)
= \sum_{k=0}^{\infty}\Bigl(\sum_{m+n=k} u_{m,n}\Bigr),
\]
all inner series converging (\omterm{def:b2:series:def}{absolutely}). In particular, if $\sum
a_m$ and $\sum b_n$ converge \omterm{def:b2:series:def}{absolutely}, their \emph{Cauchy
product}\index{Cauchy product} converges \omterm{def:b2:series:def}{absolutely} with
\[
\Bigl(\sum_m a_m\Bigr)\Bigl(\sum_n b_n\Bigr)
= \sum_{k=0}^{\infty} c_k,
\qquad
c_k = \sum_{m=0}^{k} a_m b_{k-m} .
\]
\end{theorem}

\begin{proof}
\emph{Nonnegative case.} Each grouping (by rows, columns, or
diagonals) computes the same supremum: any finite set of pairs is
contained in a finite block of rows (bounding each grouped sum below
by finite partial sums and above by the total), and monotone
convergence of partial sums does the rest --- concretely, for rows:
$\sum_{m \leq M}\sum_{n \leq N} u_{m,n} \leq S$ gives, letting $N
\to \infty$ then $M \to \infty$, $\sum_m \sum_n u_{m,n} \leq S$;
conversely every finite $F$ sits in such a rectangle, so $S \leq
\sum_m\sum_n u_{m,n}$. Diagonals: same two bounds with triangles
instead of rectangles.

\emph{General case.} Split into positive and negative (real and
imaginary) parts, each a \omterm{def:b2:series:summable}{summable} nonnegative family; the four
groupings agree on each part, hence on the difference; absolute
convergence of the inner series comes from the nonnegative case
applied to $\abs{u_{m,n}}$.

\emph{\omterm{thm:b2:series:fubini}{Cauchy product}.} The family $u_{m,n} = a_m b_n$ is \omterm{def:b2:series:summable}{summable}:
finite partial sums of $\abs{a_mb_n}$ are bounded by
$\bigl(\sum\abs{a_m}\bigr)\bigl(\sum\abs{b_n}\bigr)$. Rows give
$\bigl(\sum a_m\bigr)\bigl(\sum b_n\bigr)$; diagonals give $\sum_k
c_k$.
\end{proof}

\begin{example}[The exponential identity, honestly]\label{ex:b2:series:exp}
For $a, b \in \C$ (or commuting matrices):
\[
\Bigl(\sum_m \frac{a^m}{m!}\Bigr)\Bigl(\sum_n
\frac{b^n}{n!}\Bigr)
= \sum_k \sum_{m+n=k} \frac{a^m b^n}{m!\,n!}
= \sum_k \frac{(a + b)^k}{k!},
\]
by the binomial theorem on each diagonal: $\eu^a \eu^b = \eu^{a+b}$
--- the functional equation of $\exp$ derived from the series alone.
(Commutation is used in the binomial step; for non-commuting
matrices the identity genuinely fails,
\cref{ch:b2:diffeq}.)
\end{example}

\begin{example}[Cauchy products as a computing device]\label{ex:b2:series:squaresum}
From the geometric series and \cref{exo:b2:series:2}'s $\sum_{n
\geq 1} nz^n = \frac{z}{(1-z)^2}$ ($\abs z < 1$), one more
\omterm{thm:b2:series:fubini}{Cauchy product} finishes the second moment. Multiply $\sum_m
mz^m$ by $\sum_n z^n$: the diagonal coefficient is $\sum_{m=0}^k
m = \frac{k(k+1)}2$, so
\[
\frac{z}{(1-z)^3} = \sum_{k\geq0}\frac{k(k+1)}{2}\,z^k ,
\]
and the identity $n^2 = 2\cdot\frac{n(n+1)}2 - n$ assembles
\[
\sum_{n\geq1} n^2z^n = \frac{2z}{(1-z)^3} - \frac{z}{(1-z)^2}
= \frac{z(1+z)}{(1-z)^3} .
\]
At $z = \frac12$: $\sum_{n\geq1}\frac{n^2}{2^n} =
\frac{\frac12\cdot\frac32}{\frac18} = 6$ --- a closed value with
no differentiation anywhere, just \omterm{def:b2:series:def}{absolutely} convergent series
multiplied like polynomials. The same telescoping of identities
computes every $\sum n^dz^n$, and probabilists will recognize
the second factorial moment of the geometric law
(\cref{ch:b2:genfun}).
\end{example}

\begin{example}[A double-sum evaluation]\label{ex:b2:series:doublesum}
For $s > 1$ real, let $\zeta(s) = \sum_{n\geq1} n^{-s}$. Counting
divisors by double summation --- the family $(m^{-s}n^{-s})$ over
$(m,n) \in (\N^*)^2$ is \omterm{def:b2:series:summable}{summable} (product of convergent positive
series) --- and grouping by the product $q = mn$:
\[
\zeta(s)^2 = \sum_{m,n} \frac{1}{(mn)^s}
= \sum_{q=1}^{\infty} \frac{d(q)}{q^s},
\]
where $d(q)$ is the number of divisors of $q$. \omterm{def:b2:series:summable}{Summable} families
turn combinatorics into analysis.
\end{example}

\begin{example}[A Fubini evaluation: $\sum_n (\zeta(n) - 1) = 1$]\label{ex:b2:series:zetasum}
For integer $n \geq 2$, $\zeta(n) - 1 = \sum_{k \geq 2} k^{-n}$.
The double family $(k^{-n})_{k, n \geq 2}$ is \omterm{def:b2:series:summable}{summable}: summing
the geometric columns first,
\[
\sum_{k\geq2}\sum_{n\geq2} \frac{1}{k^n}
= \sum_{k\geq2} \frac{1/k^2}{1 - 1/k}
= \sum_{k\geq2} \frac{1}{k(k-1)} = 1
\]
(telescoping), and all terms are positive, so
\cref{thm:b2:series:fubini} authorizes summing by rows instead:
\[
\sum_{n\geq2}\bigl(\zeta(n) - 1\bigr) = 1 .
\]
The infinitely many $\zeta$-values, each transcendental-looking,
have tails that add up to exactly $1$. Closing insight: when a
double sum has positive terms, compute it in whichever order
collapses --- here columns are geometric, rows are mysterious,
and Fubini transfers the collapse.
\end{example}

\begin{example}[The geometric series solves an equation]\label{ex:b2:series:neumannsolve}
In the \omterm{def:b2:nvs:banach}{Banach space} $\bigl(C(\intcc01),
\norm\cdot_\infty\bigr)$, solve $x - K(x) = y$ where $K(f)$ is
the constant function $\frac12\int_0^1 f$. The \omterm{thm:b2:nvs:continuouslinear}{operator norm} is
$\vertiii K \leq \frac12 < 1$, so the Neumann series applies
(\cref{ex:b2:series:neumann}): $x = \sum_{n\geq0} K^n(y)$.
Compute the iterates: $K(y) = \frac12\int_0^1 y$ (a constant),
and applying $K$ to a constant $c$ gives $\frac c2$, so
$K^n(y) = \frac{1}{2^{n-1}}\cdot\frac12\int_0^1 y$ for $n \geq
1$. Summing the geometric constants:
\[
x = y + \Bigl(\int_0^1 y\Bigr)
\sum_{n\geq1}\frac{1}{2^n} = y + \int_0^1 y .
\]
Check: $x - K(x) = y + \int y - \frac12\bigl(\int y + \int
y\bigr) = y$. An infinite series, a finite answer, and a
one-line verification --- the geometric series is an inversion
algorithm, not just a convergence statement.
\end{example}

\begin{example}[Telescoping by partial fractions]\label{ex:b2:series:telescope}
Exact summation is rare; telescoping is its main supplier.
Decompose
\[
\frac{1}{n(n+1)(n+2)}
= \frac{1}{2}\Bigl(\frac{1}{n(n+1)} -
\frac{1}{(n+1)(n+2)}\Bigr),
\]
(check by reduction to the common denominator), so the partial
sums collapse:
\[
\sum_{n=1}^{N}\frac{1}{n(n+1)(n+2)}
= \frac12\Bigl(\frac{1}{1\cdot2} -
\frac{1}{(N+1)(N+2)}\Bigr)
\longrightarrow \frac14 .
\]
The same pattern --- write the term as $c(u_n - u_{n+1})$ for
an explicit $(u_n)$ --- solved \cref{exo:b2:series:10}
(arctangents) and computes every $\sum\frac{1}{n(n+1)\cdots(n +
k)} = \frac{1}{k\cdot k!}$. When an exact sum exists at this
level, a telescope is usually hiding in the term.
\end{example}

\begin{remark}[Perspectives within this volume]\label{rem:b2:series:perspectives}
Three chapters ahead are direct clients. For
\cref{ch:b2:funcseq}: normal convergence of $\sum f_n$ is
absolute convergence of $\sum\norm{f_n}_\infty$ in the \omterm{def:b2:nvs:banach}{Banach
space} $\bigl(C, \norm\cdot_\infty\bigr)$ --- this chapter's
\cref{thm:b2:nvs:absoluteconvergence} in costume. For
\cref{ch:b2:powerseries}: inside the disk of convergence
everything is absolute and \omterm{def:b2:series:summable}{summable}, so \omterm{thm:b2:series:fubini}{Cauchy products} and
rearrangements run free (that is why power series multiply like
polynomials); on the boundary, Abel's test takes over
(\cref{ex:b2:series:abelboundary}). For \cref{ch:b2:genfun}:
probability generating functions are power series whose
manipulations --- products for sums of independent variables,
double sums for compound laws --- are all licensed by
\cref{thm:b2:series:fubini}. \omterm{def:b2:series:summable}{Summable} families are the legal
department of the analysis to come.
\end{remark}

\begin{remark}[Where this chapter is used]
Everything with an infinite sum passes through here: power series
(\cref{ch:b2:powerseries}) are \omterm{def:b2:series:summable}{summable} families in disguise,
Fourier coefficients get multiplied by \omterm{thm:b2:series:fubini}{Cauchy products} and
rearranged by Parseval (\cref{ch:b2:fourier}), and probability
generating functions (\cref{ch:b2:genfun}) are Fubini's theorem
applied to expectations. The Year 3 volume absorbs \omterm{def:b2:series:summable}{summable}
families into Lebesgue integration over the counting measure ---
where \cref{thm:b2:series:fubini} becomes a special case of the
Fubini--Tonelli theorem.
\end{remark}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:b2:series:1}
Nature of: $\sum \dfrac{\cos n}{n}$; $\;\sum \dfrac{(-1)^n}{\ln
n}$; $\;\sum \dfrac{(-1)^n}{n^{3/4} + \cos n}$ \emph{(expand as in
the Year 1 trap: the alternating test needs monotonicity)}.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b2:series:2}
Prove that for $\abs z < 1$: $\sum_{n\geq1} n z^{n} =
\dfrac{z}{(1-z)^2}$, via the \omterm{thm:b2:series:fubini}{Cauchy product} of $\sum z^n$ with
itself.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:series:3}
(Kronecker-type lemma) Let $\sum b_n$ be a convergent real series.
Prove, by \omterm{thm:b2:series:abel}{Abel summation}, that $\dfrac{1}{n}\sum_{k=1}^{n} k\,b_k
\to 0$.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:series:4}
Study the convergence of $\sum \dfrac{\sin(n\theta)}{n^\alpha}$
($\theta \in \R$, $\alpha > 0$) --- for which $(\theta, \alpha)$ is
it \omterm{def:b2:series:def}{absolutely} convergent, semi-convergent, divergent?
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b2:series:5}
(Riemann rearrangement) Let $\sum u_n$ be a convergent but not
\omterm{def:b2:series:def}{absolutely} convergent real series, and $\ell \in \R$. Prove that
some rearrangement of $\sum u_n$ converges to $\ell$.
\emph{(Show both subseries of positive and negative terms diverge;
then greedily alternate: take positive terms until exceeding
$\ell$, then negative until dropping below, and so on; the terms
tend to $0$, forcing convergence to $\ell$.)}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:series:6}
Prove that the family
$\Bigl(\dfrac{x^{m+n}}{m!\,n!}\Bigr)_{(m,n)\in\N^2}$ is \omterm{def:b2:series:summable}{summable}
for every $x \in \R$, and re-derive the identity $(\eu^x)^2 =
\eu^{2x}$ by grouping the double sum along the diagonals $m + n =
k$.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:series:7}
Prove that the family $\bigl(\frac{1}{m^2 n^2}\bigr)_{m,n \geq 1}$
is \omterm{def:b2:series:summable}{summable}, and that grouping by $\gcd$: with $q = \gcd(m,n)$,
\[
\zeta(2)^2 = \sum_{q\geq1} \frac{1}{q^4}
\sum_{\substack{a,b \geq 1\\ \gcd(a,b)=1}} \frac{1}{a^2b^2}
= \zeta(4) \cdot S,
\]
where $S = \sum_{\gcd(a,b)=1} \frac{1}{a^2b^2}$: deduce $S =
\zeta(2)^2/\zeta(4)$. \emph{(Every pair $(m,n)$ writes uniquely
$(qa, qb)$ with $\gcd(a,b) = 1$.)}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b2:series:8}
(Abel's theorem on products, light version) Suppose $\sum a_n$
converges \omterm{def:b2:series:def}{absolutely} and $\sum b_n$ converges. Prove that their
\omterm{thm:b2:series:fubini}{Cauchy product} $\sum c_n$ converges, with $\sum c_n = (\sum
a_n)(\sum b_n)$. \emph{(Write $C_N = \sum_{k\leq N} c_k = \sum_n
a_n B_{N-n}$ with $B$ the partial sums of $b$; split according to
$n \leq N/2$ or not, using boundedness of $(B_m)$ and the absolute
tail of $(a_n)$.)}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b2:series:9}
In the (\omterm{def:b2:metric:complete}{non-complete}) space $E$ of eventually-zero real sequences
with the sup \omterm{def:b2:nvs:norm}{norm}, exhibit an \omterm{def:b2:series:def}{absolutely} convergent series that does
not converge in $E$. \emph{(Try $u_n = 2^{-n} e_n$ with $(e_n)$ the
canonical sequences.)}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:series:10}
Verify the identity $\arctan(n+1) - \arctan(n) =
\arctan\dfrac{1}{n^2 + n + 1}$, and deduce the exact value of
\[
\sum_{n=1}^{\infty} \arctan\frac{1}{n^2 + n + 1} .
\]
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:series:11}
Determine the nature (with equivalents) of
\[
\sum_n \bigl(\sqrt{n+1} - \sqrt n\bigr)^{\alpha}
\ (\alpha > 0), \qquad
\sum_n \Bigl(1 - \cos\frac1n\Bigr), \qquad
\sum_n \Bigl(\eu - \Bigl(1 + \frac1n\Bigr)^{\!n}\Bigr).
\]
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b2:series:12}
Let $(a_n)$ be positive and \emph{decreasing} with $\sum a_n$
convergent. Prove that $n\,a_n \to 0$ \emph{(bound $n a_{2n}$ by a
tail)}. Show that the converse fails, and that monotonicity is
essential, with explicit counterexamples.
\end{exercise}

\section{Problem: Euler's $\zeta(2) = \pi^2/6$, by Cauchy's
Cotangent Sum}

Euler's most famous identity, $1 + \frac14 + \frac19 + \cdots =
\frac{\pi^2}6$, admits a completely elementary proof, due to
Cauchy: de Moivre's formula produces a polynomial whose roots are
the numbers $\cot^2\frac{k\pi}{2n+1}$, Vieta sums those roots
exactly, and the squeeze $\cot^2\theta < \frac{1}{\theta^2} <
1 + \cot^2\theta$ crushes the partial sums of $\sum\frac1{k^2}$
between two explicit rational bounds. We run the proof in full,
extract $\zeta(4) = \frac{\pi^4}{90}$ by the same method, then
map the entire frontier between convergence and divergence with
the Bertrand series --- and prove that the frontier carries no
slowest convergent series at all.

\begin{problem}[{Weekend problem --- $\zeta(2) = \pi^2/6$ and the
Bertrand panorama}]
\label{pb:b2:series:1}
Throughout, $n \geq 1$ and $\theta_k = \dfrac{k\pi}{2n+1}$ for $k
= 1, \dots, n$; note $0 < \theta_k < \frac\pi2$.

\medskip
\noindent\textbf{Part I --- The cotangent identity.}
\begin{enumerate}
  \item Prove de Moivre's formula $(\cos\theta +
        \iu\sin\theta)^m = \cos m\theta + \iu\sin m\theta$ ($m
        \in \N$), and deduce, for $m = 2n + 1$,
        \[
        \sin\bigl((2n{+}1)\theta\bigr) = \sum_{j=0}^{n}
        (-1)^j\binom{2n+1}{2j+1}\cos^{2(n-j)}\theta\,
        \sin^{2j+1}\theta .
        \]
  \item Deduce that for $\theta \in \intoo{0}{\frac\pi2}$,
        \[
        \sin\bigl((2n{+}1)\theta\bigr) =
        \sin^{2n+1}\theta\; P_n(\cot^2\theta),
        \qquad
        P_n(x) = \sum_{j=0}^{n}
        (-1)^j\binom{2n+1}{2j+1}x^{\,n-j},
        \]
        a polynomial of degree $n$ with leading coefficient $2n
        + 1$.
  \item Show that $x_k = \cot^2\theta_k$, $k = 1, \dots, n$, are
        $n$ \emph{distinct} roots of $P_n$ --- hence all of
        them.
  \item By Vieta, prove the exact identity
        \[
        \sum_{k=1}^{n} \cot^2\frac{k\pi}{2n+1}
        = \frac{n(2n-1)}{3}.
        \]
  \item Deduce also $\displaystyle\sum_{k=1}^{n}
        \frac{1}{\sin^2\theta_k} = \frac{2n(n+1)}{3}$.
  \item Prove the squeeze: $\cot^2\theta < \dfrac1{\theta^2} <
        \dfrac{1}{\sin^2\theta}$ for $\theta \in
        \intoo{0}{\frac\pi2}$ \emph{(from $\sin\theta < \theta <
        \tan\theta$)}.
\end{enumerate}

\medskip
\noindent\textbf{Part II --- The squeeze closes: Euler's
theorem.}
\begin{enumerate}[resume]
  \item Summing question 6 over $k = 1, \dots, n$ with $\theta =
        \theta_k$, establish
        \[
        \frac{n(2n-1)}{3} \;<\;
        \frac{(2n+1)^2}{\pi^2}\sum_{k=1}^{n}\frac1{k^2}
        \;<\; \frac{2n(n+1)}{3}.
        \]
  \item Conclude (\emph{Euler's theorem, by Cauchy's proof}):
        \[
        \zeta(2) = \sum_{k=1}^{\infty}\frac{1}{k^2} =
        \frac{\pi^2}{6}.
        \]
  \item Extract a rate from the sandwich: show
        \[
        \Bigl|\sum_{k=1}^{n}\frac1{k^2} - \frac{\pi^2}6\Bigr|
        = O\Bigl(\frac1n\Bigr),
        \]
        consistent with the exact tail $\sum_{k>n}k^{-2} =
        \frac1n - \frac{1}{2n^2} + O(n^{-3})$ of
        \cref{exo:b2:comparison:11}.
  \item Run the machine one floor higher: using the second Vieta
        function of $P_n$, show
        \[
        \sum_{k=1}^{n}\cot^4\theta_k =
        \Bigl(\frac{n(2n-1)}3\Bigr)^{\!2} -
        \frac{2n(2n-1)(2n-2)(2n-3)}{60}
        \;\sim\; \frac{8n^4}{45},
        \]
        and squeeze with $\cot^4 < \theta^{-4} < (1 +
        \cot^2)^2$ to obtain $\zeta(4) = \dfrac{\pi^4}{90}$.
\end{enumerate}

\medskip
\noindent\textbf{Part III --- Dividends.}
\begin{enumerate}[resume]
  \item Deduce from $\zeta(2) = \frac{\pi^2}6$:
        \[
        \sum_{k\geq0}\frac{1}{(2k+1)^2} = \frac{\pi^2}{8},
        \qquad
        \sum_{k\geq1}\frac{(-1)^{k-1}}{k^2} = \frac{\pi^2}{12}.
        \]
  \item Combine with \cref{exo:b2:series:7}: compute $S =
        \sum_{\gcd(a,b)=1}\frac{1}{a^2b^2} =
        \frac{\zeta(2)^2}{\zeta(4)} = \frac52$, and interpret
        $\frac{1}{\zeta(2)} = \frac{6}{\pi^2} \approx 0.608$ as
        the density of coprime pairs (state the heuristic
        honestly: the rigorous count is a Year 3 volume
        matter).
  \item (Certified acceleration) The tail formula of question 9
        gives $\sum_{k\leq n}k^{-2} + \frac1n - \frac1{2n^2} =
        \frac{\pi^2}6 + O(n^{-3})$. Compare the work needed for
        six digits of $\zeta(2)$: direct summation versus the
        corrected sum at $n = 100$ (where the error is
        $1.7\cdot10^{-7}$).
  \item Check question 4 by hand at $n = 1$ and $n = 2$ (the
        values $\cot^2\frac\pi3 = \frac13$ and
        $\cot^2\frac\pi5 + \cot^2\frac{2\pi}5 = 2$), using
        $\cos\frac\pi5 = \frac{1+\sqrt5}4$ or a numerical
        evaluation.
  \item Prove the companion identity
        \[
        \sum_{k=1}^{n}\tan^2\frac{k\pi}{2n+1} = n(2n+1)
        \]
        \emph{(the numbers $\tan^2\theta_k$ are the roots of the
        reversed polynomial $x^nP_n(1/x)$)}, and verify it at $n
        = 1$.
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- The Bertrand panorama.} For $\alpha,
\beta \in \R$, consider the \emph{Bertrand series}
\[
\sum_{n \geq 3} \frac{1}{n^{\alpha}(\ln n)^{\beta}} .
\]
\begin{enumerate}[resume]
  \item Show that for $\alpha > 1$ the series converges,
        whatever $\beta$ \emph{(compare with
        $n^{-(1+\alpha)/2}$)}.
  \item Show that for $\alpha < 1$ it diverges, whatever
        $\beta$.
  \item For $\alpha = 1$: using the series--integral comparison
        (\cref{thm:b2:comparison:seriesintegral}) with $f(t) =
        \frac{1}{t(\ln t)^\beta}$, prove convergence iff $\beta
        > 1$.
  \item Iterate the frontier: show $\sum\frac{1}{n\ln n\,\ln\ln
        n}$ diverges while $\sum\frac{1}{n\ln n\,(\ln\ln n)^2}$
        converges.
  \item Two traps: determine the nature of
        \[
        \sum_n \frac{1}{n^{1 + 1/\ln n}}
        \qquad\text{and}\qquad
        \sum_n \frac{1}{n^{1 + 1/\ln\ln n}}
        \]
        \emph{(compute $n^{1/\ln n}$ exactly; compare
        $n^{1/\ln\ln n}$ with every power of $\ln n$)}.
  \item (No slowest convergent series) Let $\sum a_n$ be any
        convergent series with $a_n > 0$, and $R_n = \sum_{k
        \geq n}a_k$ its tails. Prove that $\sum
        \frac{a_n}{\sqrt{R_n}}$ still \emph{converges}
        \emph{(compare with the telescoping $2(\sqrt{R_n} -
        \sqrt{R_{n+1}})$)}, although $\frac{a_n/\sqrt{R_n}}{a_n}
        \to \infty$: every convergent series is strictly
        dominated by another convergent series. The frontier of
        convergence is not a curve but a fog.
\end{enumerate}

\medskip
\noindent\textbf{Part V --- Cross-checks and synthesis.}
\begin{enumerate}[resume]
  \item (Cauchy condensation) Prove: for $(a_n)$ positive
        decreasing, $\sum a_n$ converges iff $\sum 2^k a_{2^k}$
        converges. Re-derive question 18's frontier from it.
  \item (The cost of slowness) For $\sum\frac1{n(\ln n)^2}$,
        bound the tail by an integral and show that summing up
        to $N = 10^6$ still leaves an error larger than
        $0.07$: convergence certified by theory can be useless
        for numerics --- contrast with question 13.
  \item Classify (with one-line justifications):
        $\sum\frac1{n\ln n}$, $\sum\frac1{n^{1.01}}$,
        $\sum\frac{(\ln n)^{100}}{n^{1.001}}$,
        $\sum\frac1{n(\ln n)(\ln\ln n)^{3}}$.
  \item (Synthesis) One sentence each: how de Moivre turned a
        trigonometric identity into a polynomial with computable
        root sums; where the squeeze needed \emph{exact}
        endpoint identities rather than equivalents; which tool
        of \cref{ch:b2:comparison} powered Part IV; and what
        question 21 says about the dream of a ``universal
        comparison test''. Name the two summits: Euler's
        $\zeta(2) = \frac{\pi^2}6$ (and its floor above,
        $\zeta(4) = \frac{\pi^4}{90}$), and the Bertrand
        classification. Note where $\zeta(2)$ will be proved
        again: by Parseval in \cref{ch:b2:fourier} --- one
        theorem, two civilizations.
\end{enumerate}
\end{problem}
