\chapter{Functions of a Real Variable}\label{ch:b2:realfun}

Before analysis moves to functions of functions (\cref{ch:b2:funcseq})
it pays to know the one-variable landscape in finer detail than
Year 1 required: how discontinuous a monotone function can be, how
regular a convex function must be, and what special properties
derivatives enjoy (Darboux). These structural results are short,
sharp, and beloved of examiners.

\section{Monotone functions}

\begin{theorem}[Regularity of monotone functions]\label{thm:b2:realfun:monotone}
Let $f \colon I \to \R$ be increasing on an interval.
\begin{enumerate}
  \item At every interior point $a$, the one-sided limits exist:
        \[
        f(a^-) = \sup_{x < a} f(x) \;\leq\; f(a) \;\leq\;
        f(a^+) = \inf_{x > a} f(x) ;
        \]
        every discontinuity is a \emph{jump}.
  \item The set of discontinuities of $f$ is at most
        countable.\index{monotone function!discontinuities}
\end{enumerate}
\end{theorem}

\begin{proof}
(1) The set $\{f(x) : x < a\}$ is nonempty, bounded above by $f(a)$:
its supremum $s$ satisfies $f(x) \to s$ as $x \to a^-$ (given
$\varepsilon$, some $f(x_0) > s - \varepsilon$, and monotonicity
traps $f(x) \in \intoc{s - \varepsilon}{s}$ for $x \in
\intoo{x_0}{a}$). Symmetrically on the right.

(2) To each discontinuity $a$ attach the nonempty open interval
$J_a = \intoo{f(a^-)}{f(a^+)}$ (a genuine jump). For $a < b$
discontinuities, $J_a$ and $J_b$ are disjoint: $f(a^+) \leq f(c)
\leq f(b^-)$ for any $c$ between. Each $J_a$ contains a rational;
distinct discontinuities get distinct rationals: an injection of the
discontinuity set into $\Q$, which is countable
(\cref{prop:b2:structures:countablestable}).
\end{proof}

\begin{example}
The bound is sharp: fix an enumeration $(r_n)$ of $\Q \cap
\intoo{0}{1}$ and set $f(x) = \sum_{n : r_n \leq x} 2^{-n}$
(a summable-family definition,
\cref{def:b2:series:summable}). Then $f$ is increasing on
$\intcc{0}{1}$ and discontinuous exactly at every rational of
$\intoo{0}{1}$ (jump $2^{-n}$ at $r_n$): a monotone function
\emph{can} be discontinuous on a dense countable set.
\end{example}

\begin{example}[The jumps cannot outweigh the rise]\label{ex:b2:realfun:jumpbudget}
For $f$ increasing on $\intcc{a}{b}$, the jumps have a budget:
if $a < c_1 < \dots < c_m < b$ are discontinuities with jumps
$s_i = f(c_i^+) - f(c_i^-) > 0$, then choosing interlacing
points $a < c_1 < t_1 < c_2 < \dots$ and using monotonicity on
each piece,
\[
\sum_{i=1}^{m} s_i \;\leq\; f(b) - f(a) :
\]
the total ascent bounds the total jumping. Consequence: for
each $k$, at most $k\,\bigl(f(b) - f(a)\bigr)$ discontinuities
have jump $\geq \frac1k$ --- a quantitative refinement of
\cref{thm:b2:realfun:monotone}~(2), since the discontinuity set
is the countable union over $k$ of these finite sets. On the
rational-jump function above, the budget is spent exactly:
the jumps $2^{-n}$ sum to $1 = f(1^+) - f(0^-)$ in the obvious
extended sense. Monotone functions may jump densely, but only
on a strict allowance.
\end{example}

\section{Convex functions}

\begin{lemma}[Slope inequality]\label{lem:b2:realfun:slopes}
Let $f$ be convex on $I$ and $x < y < z$ in $I$. Then
\[
\frac{f(y) - f(x)}{y - x}
\;\leq\; \frac{f(z) - f(x)}{z - x}
\;\leq\; \frac{f(z) - f(y)}{z - y} :
\]
slopes of chords increase in both endpoints.\index{convex
function!slope inequality}
\end{lemma}

\begin{proof}
Write $y = \frac{z - y}{z - x}\,x + \frac{y - x}{z - x}\,z$: a
convex combination, since the two coefficients are positive and
sum to $1$. Convexity gives
\[
f(y) \;\leq\; \frac{z-y}{z-x}\,f(x) + \frac{y-x}{z-x}\,f(z).
\]
For the left inequality, subtract $f(x)$ from both sides, using
$\frac{z-y}{z-x} - 1 = -\frac{y-x}{z-x}$:
\[
f(y) - f(x) \leq \frac{y - x}{z - x}\bigl(f(z) - f(x)\bigr),
\]
and divide by $y - x > 0$. For the right inequality, subtract
instead from $f(z)$:
\[
f(z) - f(y) \geq f(z) - \frac{z-y}{z-x}f(x) -
\frac{y-x}{z-x}f(z)
= \frac{z - y}{z - x}\bigl(f(z) - f(x)\bigr),
\]
and divide by $z - y > 0$. Both displayed steps are the same
barycentric identity read against a different endpoint.
\end{proof}

\begin{theorem}[Regularity of convex functions]\label{thm:b2:realfun:convexreg}
Let $f$ be convex on an interval $I$.
\begin{enumerate}
  \item At every interior point, $f$ has finite one-sided
        derivatives $f'_g \leq f'_d$; both are increasing functions
        of the point; in particular $f$ is continuous on the
        interior of $I$ (but possibly not at endpoints).
  \item $f$ lies above each \emph{support line}: for $a$ interior
        and any $m \in \intcc{f'_g(a)}{f'_d(a)}$,
        \[
        f(x) \geq f(a) + m(x - a) \qquad (x \in I).
        \]
  \item (Jensen, weighted) For $x_i \in I$ and weights $\lambda_i
        \geq 0$, $\sum\lambda_i = 1$:
        \[
        f\Bigl(\sum_i \lambda_i x_i\Bigr) \leq \sum_i \lambda_i
        f(x_i) .\index{Jensen's inequality}
        \]
\end{enumerate}
\end{theorem}

\begin{proof}
(1) Fix $a$ interior. By \cref{lem:b2:realfun:slopes}, the slope
$\tau(h) = \frac{f(a + h) - f(a)}{h}$ is an increasing function of
$h$ (on both sides, and $\tau(h_-) \leq \tau(h_+)$ for $h_- < 0 <
h_+$). Hence $\tau$ has a finite limit as $h \to 0^-$ (increasing,
bounded above by any right slope) --- this is $f'_g(a)$ --- and as
$h \to 0^+$ ($f'_d(a)$), with $f'_g(a) \leq f'_d(a)$. Finite
one-sided derivatives force continuity at $a$. Monotonicity in the
point: for $a < b$ interior, $f'_d(a) \leq \frac{f(b) - f(a)}{b -
a} \leq f'_g(b)$, again by the slope inequality.

(2) For $x > a$: $\frac{f(x) - f(a)}{x - a} \geq f'_d(a) \geq m$;
for $x < a$: $\frac{f(a) - f(x)}{a - x} \leq f'_g(a) \leq m$. Both
rearrange to the claim.

(3) Induction on the number of points exactly as in the Year 1
volume (the two-point case is the definition) --- or in one stroke:
apply (2) at $a = \sum\lambda_i x_i$ and average the support-line
inequalities at the points $x_i$ with weights $\lambda_i$: $\sum_i
\lambda_i f(x_i) \geq f(a) + m\sum_i\lambda_i(x_i - a) = f(a)$.
\end{proof}

\begin{omfigure}
\begin{tikzpicture}
\begin{axis}[omaxis, width=8.2cm, height=5.4cm,
    xmin=-2.4, xmax=2.5, ymin=-0.9, ymax=4.5,
    xtick={-1.5, 0.5, 2}, ytick=\empty]
  \addplot[omDef, very thick, domain=-2.1:2.15, samples=80]
    {x^2};
  \addplot[omThm, thick, domain=-1.5:2, samples=2]
    {0.5*x + 3};
  \addplot[omProp, thick, domain=-1.1:2.2, samples=2]
    {x - 0.25};
  \node[font=\small, omThm] at (axis cs:-1.05,3.15)
    {chord};
  \node[font=\small, omProp] at (axis cs:1.75,0.95)
    {support line};
  \node[font=\small, omDef] at (axis cs:-1.95,2.6) {$f$};
\end{axis}
\end{tikzpicture}

{\small Convexity in one picture: between $-1.5$ and $2$ the
graph of $f(x) = x^2$ stays below its chord (the definition) and
above the support line at $x = 0.5$
(\cref{thm:b2:realfun:convexreg}~(2)) --- every inequality of this
chapter's weekend problem is a rearrangement of these two
positions.}
\end{omfigure}

\begin{example}[Endpoint discontinuity]
On $\intcc{0}{1}$, the function $f(0) = 1$, $f(x) = 0$ for $x > 0$
is convex but discontinuous at the endpoint $0$: statement (1) is
sharp.
\end{example}

\begin{example}[Corners and the sheaf of support lines]\label{ex:b2:realfun:corner}
For $f(x) = \abs x$ at $a = 0$: the one-sided derivatives are
$f'_g(0) = -1$ and $f'_d(0) = +1$, and
\cref{thm:b2:realfun:convexreg}~(2) hands out a support line for
\emph{every} slope $m \in \intcc{-1}{1}$:
\[
\abs x \geq m\,x \qquad (x \in \R,\ -1 \leq m \leq 1),
\]
each an equality exactly on a half-line or at $0$. A convex
function is differentiable at $a$ precisely when the sheaf
collapses to a single line ($f'_g(a) = f'_d(a)$); corners carry
an interval of tangents. This sheaf is the finite-dimensional
germ of the \emph{subdifferential} of convex optimization ---
and the reason convex functions are so robust: even where the
derivative fails, the supporting geometry survives, which is all
that Jensen's proof used.
\end{example}

\begin{example}[Power mean inequality]\label{ex:b2:realfun:powermeans}
For $0 < p < q$ and positive $x_i$ with weights $\lambda_i$
summing to $1$, applying Jensen to the convex $t \mapsto t^{q/p}$
at the points $x_i^p$:
\[
\Bigl(\sum \lambda_i x_i^{p}\Bigr)^{1/p}
\leq \Bigl(\sum \lambda_i x_i^{q}\Bigr)^{1/q} :
\]
power means increase with the exponent --- containing AM--QM, and,
in the limit $p \to 0$ (\cref{exo:b2:realfun:6}), the AM--GM
inequality once more.
\end{example}

\begin{omfigure}
\begin{tikzpicture}
\begin{axis}[omaxis, width=8.6cm, height=5.6cm,
    xmin=-8.5, xmax=8.5, ymin=0.5, ymax=4.5,
    xtick={-4,-1,1,2,4}, ytick={1,2,4},
    xlabel={$p$}, ylabel={$M_p$}]
  \addplot[omDef, very thick, domain=-8:-0.4, samples=90]
    {((1/3)*(1^x + 2^x + 4^x))^(1/x)};
  \addplot[omDef, very thick, domain=0.4:8, samples=90]
    {((1/3)*(1^x + 2^x + 4^x))^(1/x)};
  \addplot[omThm, dashed, domain=-8.5:8.5, samples=2] {4};
  \addplot[omThm, dashed, domain=-8.5:8.5, samples=2] {1};
  \node[font=\small, omDef] at (axis cs:-5.4,1.55)
    {$M_p(1,2,4)$};
  \node[font=\small, omThm] at (axis cs:-6.4,3.75)
    {$\max = 4$};
  \node[font=\small, omThm] at (axis cs:6.3,1.25)
    {$\min = 1$};
  \node[font=\small, omProp] at (axis cs:0.2,1.83) {$2$};
\end{axis}
\end{tikzpicture}

{\small The power mean $M_p$ of the values $1, 2, 4$ (equal
weights), as a function of the exponent $p$: increasing from
$\min = 1$ (as $p \to -\infty$) to $\max = 4$ (as $p \to
+\infty$), through the harmonic ($p = -1$), geometric (the gap
at $p = 0$, value $2$), arithmetic ($p = 1$) and quadratic ($p
= 2$) means. The whole chain of classical mean inequalities is
one increasing curve --- proved in this chapter's weekend
problem, Part III.}
\end{omfigure}

\begin{example}[Maximal entropy]\label{ex:b2:realfun:entropy}
For a probability vector $(p_1, \dots, p_n)$ (positive,
summing to $1$), the entropy $H(p) = -\sum_i p_i\ln p_i$
satisfies
\[
H(p) \leq \ln n ,
\qquad\text{with equality iff } p_i = \frac1n \text{ for all }
i .
\]
Proof by Jensen (\cref{thm:b2:realfun:convexreg}~(3)) applied to
the \emph{concave} $\ln$ with weights $p_i$ at the points
$\frac{1}{p_i}$:
\[
H(p) = \sum_i p_i\ln\frac{1}{p_i}
\leq \ln\Bigl(\sum_i p_i\,\frac1{p_i}\Bigr) = \ln n ,
\]
equality forcing all points $\frac1{p_i}$ equal (strict
concavity), i.e.\ $p$ uniform. Equivalently, this is
\cref{exo:b2:realfun:7} with $q$ uniform. Uncertainty is
maximized by ignorance uniformly spread --- the variational
principle behind coding, statistical mechanics, and the entropy
appearances of \cref{ch:b2:randomvar}.
\end{example}

\begin{method}[Finding the convex function behind an inequality]\label{met:b2:realfun:spotting}
Most classical inequalities are Jensen in costume; to undress
one: (1) normalize so that a \emph{weighted average} appears
(weights positive, summing to $1$ --- divide by a total mass if
necessary); (2) look at what function is applied inside versus
outside the average: the claim ``$f(\text{average}) \leq$
average of $f$'' names the convex $f$; (3) certify convexity by
the second derivative, and handle equality via strictness; (4)
if no average is visible, take logarithms first --- products
and powers become averages, and the concavity of $\ln$ carries
AM--GM, Young and their relatives (this chapter's weekend
problem runs steps 1--4 on each of them). If even logarithms do
not reveal an average, try reading the inequality as
monotonicity of slopes (\cref{lem:b2:realfun:slopes}) ---
superadditivity statements like \cref{exo:b2:realfun:9} live
there.
\end{method}

\begin{remark}[Common pitfalls]\label{rem:b2:realfun:pitfalls}
(i) Convexity is not preserved by products: $x$ and $(x - 1)^2$
are convex on $\intcc{0}{2}$, but their product $x(x-1)^2$ has
second derivative $6x - 4$, negative on
$\intco{0}{\frac23}$ --- not convex; nor is convexity preserved
by composition without monotonicity
(\cref{exo:b2:realfun:10}). (ii) Jensen flips for concave
functions: half the classical inequalities are the concave
$\ln$-version; applying the convex form to $\ln$ is the
quickest way to prove AM--GM \emph{backwards}. (iii) Midpoint
convexity alone does not imply convexity --- continuity (or
mere boundedness) is needed (\cref{exo:b2:realfun:8}); the
pathological counterexamples live beyond this book's axioms. (iv)
A convex function on an \emph{open} interval is continuous,
even locally Lipschitz (\cref{exo:b2:realfun:12}); at
endpoints, nothing is free. (v) Derivatives obey Darboux but
need not be continuous (\cref{ex:b2:realfun:oscillation}):
``$f'$ has no jumps'' never means ``$f'$ is continuous''.
\end{remark}

\section{The Darboux property}

\begin{theorem}[Darboux]\label{thm:b2:realfun:darboux}
Let $f$ be differentiable on an interval $I$. Then $f'$ takes every
value between any two of its values --- even though $f'$ need not be
continuous.\index{Darboux's theorem}
\end{theorem}

\begin{proof}
Let $a < b$ in $I$ and $v$ strictly between $f'(a)$ and $f'(b)$,
say $f'(a) < v < f'(b)$. The function $g(x) = f(x) - vx$ is
differentiable with $g'(a) < 0 < g'(b)$: its minimum on
$\intcc{a}{b}$ (attained: continuity on a compact) is not at $a$
(just after $a$, $g$ decreases below $g(a)$) nor at $b$ (just
before $b$, $g$ is below $g(b)$): it is interior, and there $g'(c)
= 0$, i.e.\ $f'(c) = v$. (This was a Year 1 starred exercise; its
place in the theory is here.)
\end{proof}

\begin{example}[Which functions are derivatives?]\label{ex:b2:realfun:whichderivatives}
Darboux's theorem is a non-existence machine. The floor function
$\lfloor x\rfloor$ is not the derivative of any function on
$\R$: it takes the values $0$ and $1$ but skips $\frac12$ on
$\intcc{0}{1}$, which \cref{thm:b2:realfun:darboux} forbids for
derivatives. The same verdict hits every function with a jump
--- sign, Heaviside, all step functions --- however innocent
they look; their ``antiderivatives'' ($\abs x$ for sign, etc.)
exist only away from the jump and knot there with a corner.
Contrast: the wildly discontinuous $f'$ of
\cref{ex:b2:realfun:oscillation} \emph{is} a derivative --- its
discontinuity is an oscillation, which Darboux tolerates. The
boundary between the two behaviours is exactly the
no-jumps corollary below.
\end{example}

\begin{corollary}\label{cor:b2:realfun:nojumps}
A derivative has no jump discontinuities: if $f'(a^-)$ and
$f'(a^+)$ exist, they equal $f'(a)$. The discontinuities of a
derivative are always of oscillation type ($x^2\sin\frac1x$'s
derivative at $0$, Year 1 volume).
\end{corollary}

\begin{proof}
If $f'(a^+) = \lim_{x\to a^+} f'(x)$ exists and differs from
$f'(a)$, values strictly between them would be skipped by $f'$ on
a right neighborhood --- contradicting Darboux on intervals
$\intcc{a}{a + h}$. (Alternatively: the mean value theorem forces
$f'(a) = \lim_{h\to0^+} \frac{f(a+h)-f(a)}{h} = f'(a^+)$, the
difference quotient being an $f'$-value at an intermediate point.)
Same on the left.
\end{proof}

\begin{example}[The canonical oscillating derivative]\label{ex:b2:realfun:oscillation}
Let $f(x) = x^2\sin\frac1x$ for $x \neq 0$ and $f(0) = 0$. At
$0$: $\bigl|\frac{f(h) - f(0)}{h}\bigr| = \abs{h\sin\frac1h}
\leq \abs h \to 0$, so $f'(0) = 0$ exists. Away from $0$,
\[
f'(x) = 2x\sin\frac1x - \cos\frac1x ,
\]
whose first term tends to $0$ while $\cos\frac1x$ oscillates
through $\intcc{-1}{1}$ on every interval $\intoo{0}{\delta}$: the
limit $f'(0^+)$ does not exist. So $f'$ is defined everywhere but
discontinuous at $0$ --- and, exactly as
\cref{cor:b2:realfun:nojumps} predicts, the discontinuity is an
oscillation, not a jump: on each $\intoo{0}{\delta}$, $f'$ still
sweeps a full interval around $0$. Derivatives can be wild, but
only in the Darboux-compatible way.
\end{example}

\begin{remark}[Where this chapter is used]
Convexity is the engine of the inequality industry: this
chapter's weekend problem manufactures Young, H\"older, Minkowski
and the power-mean chain from it, which \cref{ch:b2:nvs}'s norm
theory and \cref{ch:b2:integration}'s integral estimates consume;
Jensen reappears in probability as the moment inequalities of
\cref{ch:b2:randomvar}. Monotone regularity returns in
\cref{ch:b2:integration} (monotone functions are integrable) and,
in the Year 3 volume, as the almost-everywhere differentiability
of monotone functions --- where ``countably many jumps'' becomes
the first step of Lebesgue's theory.
\end{remark}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:b2:realfun:1}
Determine the discontinuity sets and the jump sizes: $\lfloor x
\rfloor$; $\;x - \lfloor x\rfloor$; $\;\lfloor x \rfloor + \sqrt{x
- \lfloor x\rfloor}$; the function of the example following
\cref{thm:b2:realfun:monotone} restricted to dyadic rationals
$r_n$.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b2:realfun:2}
Prove that an increasing function $f \colon I \to \R$ with the
intermediate value property (its image of any subinterval is an
interval) is continuous.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b2:realfun:3}
Which of the following are convex on their domain? $x \mapsto
x\ln x$ ($x > 0$); $\;x \mapsto \ln(1 + \eu^x)$; $\;x \mapsto
\sqrt{1 + x^2}$; $\;x \mapsto x^3$.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:realfun:4}
Let $f$ be convex on $\R$ and bounded above. Prove that $f$ is
constant. \emph{(If $f(a) \neq f(b)$, the slope inequality
propagates the nonzero chord slope: beyond the point with the larger
value, $f$ grows at least linearly --- contradicting boundedness.
Treat both signs of the slope.)} Deduce that a convex function on
$\R$ with an asymptote at both ends is affine.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:realfun:5}
Let $f$ be differentiable on $I$ with $f'$ monotone. Prove that
$f'$ is continuous \emph{(combine
\cref{thm:b2:realfun:monotone} and
\cref{cor:b2:realfun:nojumps})}.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:realfun:6}
(Geometric mean as a limit) For positive $x_i$ and weights
$\lambda_i$ summing to $1$, prove
\[
\lim_{p \to 0^+} \Bigl(\sum_i \lambda_i x_i^p\Bigr)^{1/p}
= \prod_i x_i^{\lambda_i} ,
\]
via $x_i^p = \eu^{p\ln x_i} = 1 + p\ln x_i + O(p^2)$, and deduce
the weighted AM--GM inequality from
\cref{ex:b2:realfun:powermeans}.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:realfun:7}
(Entropy inequality) Using strict convexity of $t \mapsto t\ln t$,
prove that for positive $p_i, q_i$ with $\sum p_i = \sum q_i = 1$:
\[
\sum_i p_i \ln\frac{p_i}{q_i} \geq 0 ,
\]
with equality iff $p = q$. \emph{(Write the left side as $\sum q_i\,
\varphi\bigl(\frac{p_i}{q_i}\bigr)$ with $\varphi(t) = t\ln t$ and
apply Jensen with weights $q_i$.)}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b2:realfun:8}
(Midpoint convexity) $f \colon I \to \R$ is \emph{midpoint convex}
when $f\bigl(\frac{x+y}{2}\bigr) \leq \frac{f(x) + f(y)}{2}$
always. Prove that a \emph{continuous} midpoint convex function is
convex. \emph{(Establish the convexity inequality for dyadic
weights $\frac{k}{2^m}$ by induction on $m$, then pass to the limit
using density and continuity.)}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b2:realfun:9}
Let $f$ be convex on $\intco{0}{+\infty}$ with $f(0) \leq 0$. Prove
that $x \mapsto \frac{f(x)}{x}$ is increasing on
$\intoo{0}{+\infty}$, and deduce that for convex $f$ with $f(0) =
0$: $f(x + y) \geq f(x) + f(y)$ for $x, y \geq 0$
(superadditivity).
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b2:realfun:10}
Let $f$ be convex on $I$ and $g$ convex \emph{increasing} on an
interval containing $f(I)$. Prove that $g \circ f$ is convex, and
show by a counterexample that monotonicity of $g$ cannot be
dropped.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:realfun:11}
(Hermite--Hadamard) Let $f$ be convex and continuous on
$\intcc{a}{b}$. Prove
\[
f\Bigl(\frac{a+b}{2}\Bigr) \;\leq\; \frac{1}{b -
a}\int_a^b f(t)\,\dd t \;\leq\; \frac{f(a) + f(b)}{2} .
\]
\emph{(Left: integrate a support line at the midpoint. Right:
bound $f$ by the chord.)}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b2:realfun:12}
Prove that a convex function on an \emph{open} interval $I$ is
locally Lipschitz: for every segment $\intcc{a}{b} \subseteq I$
and margin $\delta > 0$ with $\intcc{a - \delta}{b + \delta}
\subseteq I$, the restriction of $f$ to $\intcc{a}{b}$ is
Lipschitz, with constant
$\max\Bigl(\bigl|\frac{f(a) - f(a - \delta)}{\delta}\bigr|,
\bigl|\frac{f(b + \delta) - f(b)}{\delta}\bigr|\Bigr)$
\emph{(trap every chord slope between these two by the slope
inequality)}.
\end{exercise}

\section{Problem: The Convexity Toolbox}

One definition --- the chord above the graph --- generates the
entire toolbox of classical inequalities. This weekend problem
builds it in logical order: convexity criteria and strict Jensen,
then Young, H\"older and Minkowski (the birth certificates of the
$p$-norms), the complete power-mean chain from the minimum to the
maximum, and two crown dividends --- Carleman's inequality, and
H\"older read as a duality. Everything is proved; nothing is
imported.

\begin{problem}[{Weekend problem --- Young, H\"older, Minkowski,
and the power-mean chain}]
\label{pb:b2:realfun:1}
Throughout, $p, q > 1$ are \emph{conjugate exponents}: $\frac1p +
\frac1q = 1$; vectors are $a = (a_1, \dots, a_n) \in \R^n$;
weights $\lambda_i > 0$ satisfy $\sum_i\lambda_i = 1$.

\medskip
\noindent\textbf{Part I --- Criteria and strict Jensen.}
\begin{enumerate}
  \item Let $f$ be differentiable on an interval $I$. Prove that
        $f$ is convex if and only if $f'$ is increasing
        \emph{(one direction by passing to the limit in the
        slope inequality \cref{lem:b2:realfun:slopes}; the
        other by the mean value theorem)}. Deduce the $C^2$
        criterion $f'' \geq 0$.
  \item Suppose $f''> 0$ on $I$. Prove that $f$ is
        \emph{strictly} convex (strict inequality for $x \neq y$
        and $\lambda \in \intoo01$), and that a strictly convex
        function satisfies Jensen's inequality
        (\cref{thm:b2:realfun:convexreg}~(3)) with equality
        \emph{only} when all the $x_i$ coincide.
  \item Certify the toolbox's raw materials: $-\ln$ is strictly
        convex on $\intoo{0}{+\infty}$; $t \mapsto t^r$ is
        strictly convex there for $r > 1$ and strictly concave
        for $0 < r < 1$; $\exp$ is strictly convex on $\R$.
  \item (Young's inequality) For $a, b \geq 0$, prove
        \[
        ab \;\leq\; \frac{a^p}{p} + \frac{b^q}{q},
        \]
        with equality if and only if $a^p = b^q$ \emph{(apply
        the concavity of $\ln$ to the two points $a^p, b^q$
        with weights $\frac1p, \frac1q$)}.
  \item Re-derive weighted AM--GM in one line from the
        concavity of $\ln$:
        \[
        \prod_i x_i^{\lambda_i} \leq \sum_i\lambda_ix_i \qquad
        (x_i > 0),
        \]
        with the equality case; compare with the limit route of
        \cref{exo:b2:realfun:6}.
\end{enumerate}

\medskip
\noindent\textbf{Part II --- H\"older and Minkowski.} Write
$\norm{a}_p = \bigl(\sum_i \abs{a_i}^p\bigr)^{1/p}$ and
$\norm{a}_\infty = \max_i\abs{a_i}$.
\begin{enumerate}[resume]
  \item (H\"older) Prove
        \[
        \sum_{i=1}^{n}\abs{a_ib_i} \;\leq\;
        \norm a_p\,\norm b_q ,
        \]
        with equality iff the vectors $(\abs{a_i}^p)$ and
        $(\abs{b_i}^q)$ are proportional \emph{(normalize
        $\norm a_p = \norm b_q = 1$ and apply Young termwise)}.
  \item Identify the special cases: $p = q = 2$
        (Cauchy--Schwarz), and the endpoint pair $(p, q) = (1,
        \infty)$: state and prove $\sum\abs{a_ib_i} \leq
        \norm a_1\norm b_\infty$.
  \item (Minkowski) For $p \geq 1$, prove
        \[
        \norm{a + b}_p \leq \norm a_p + \norm b_p
        \]
        \emph{(write $\abs{a_i + b_i}^p \leq \abs{a_i +
        b_i}^{p-1}(\abs{a_i} + \abs{b_i})$ and apply H\"older
        to each product)}. Conclude: $\norm\cdot_p$ is a norm
        on $\R^n$ for every $p \in \intco{1}{+\infty}$,
        completing the picture of \cref{ch:b2:nvs}.
  \item Integral versions: for $f, g$ continuous on
        $\intcc{a}{b}$, state and prove H\"older and Minkowski
        for $\norm f_p = \bigl(\int_a^b\abs
        f^p\bigr)^{1/p}$ \emph{(same proofs, with the strict
        positivity of the integral for the equality
        discussion)}.
  \item Prove the monotonicity $\norm a_q \leq \norm a_p$ for $1
        \leq p \leq q$, the limit $\norm a_p \to
        \norm a_\infty$ as $p \to \infty$, and the reverse
        comparison with the sharp constant:
        \[
        \norm a_p \leq n^{\frac1p - \frac1q}\,\norm a_q
        \]
        \emph{(H\"older against the constant vector)}. Identify
        the vectors achieving each equality.
  \item (Interpolation) For $1 \leq p < r < q$ and $\theta \in
        \intoo01$ with $\frac1r = \frac\theta p +
        \frac{1-\theta}q$, prove
        \[
        \norm a_r \leq \norm a_p^{\theta}\,
        \norm a_q^{1-\theta}
        \]
        \emph{(apply H\"older with exponents $\frac{p}{\theta
        r}$ and $\frac{q}{(1-\theta)r}$ to $\abs{a_i}^{\theta
        r}\abs{a_i}^{(1-\theta)r}$)}.
\end{enumerate}

\medskip
\noindent\textbf{Part III --- The power-mean chain, complete.}
For $p \neq 0$ set $M_p = \bigl(\sum_i\lambda_i
x_i^p\bigr)^{1/p}$ ($x_i > 0$), and $M_0 = \prod_i
x_i^{\lambda_i}$.
\begin{enumerate}[resume]
  \item Prove that $p \mapsto M_p$ is increasing on all of
        $\R^*$: treat $p < q < 0$ by the reciprocal identity
        $M_{-p}(x) = M_p(1/x)^{-1}$, and bridge through $0$ by
        showing $M_p \leq M_0 \leq M_q$ for $p < 0 < q$
        \emph{(apply the concavity of $\ln$ to $x_i^q$, and the
        reversed inequality for negative exponents)}.
  \item Prove the limits $M_p \to \max_i x_i$ as $p \to
        +\infty$ and $M_p \to \min_i x_i$ as $p \to -\infty$.
  \item Write out the chain $\min \leq \mathrm{HM} \leq
        \mathrm{GM} \leq \mathrm{AM} \leq \mathrm{QM} \leq
        \max$ for equal weights, and prove the classic
        consequence: for positive $a_1, \dots, a_n$,
        \[
        \Bigl(\sum_i a_i\Bigr)\Bigl(\sum_i\frac1{a_i}\Bigr)
        \geq n^2 .
        \]
  \item Relate means to norms: for equal weights $\lambda_i =
        \frac1n$, $M_p(x) = n^{-1/p}\norm x_p$. Reconcile the
        two monotonicities --- means \emph{increase} with $p$
        while norms \emph{decrease} (question 10) --- in one
        sentence about the factor $n^{-1/p}$.
  \item Determine the equality cases along the whole chain of
        question 14 (positive weights): equality anywhere
        forces all $x_i$ equal --- strict convexity pays off.
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- Dividends.}
\begin{enumerate}[resume]
  \item (Young with a knob) For $a, b \geq 0$ and $\varepsilon >
        0$, prove
        \[
        ab \leq \varepsilon\,\frac{a^p}{p} +
        \varepsilon^{-q/p}\,\frac{b^q}{q},
        \]
        and the workhorse case $ab \leq \varepsilon a^2 +
        \frac{b^2}{4\varepsilon}$: the absorption trick used
        throughout analysis.
  \item (Toward Carleman) Let $c_k = \frac{(k+1)^k}{k^{k-1}}$.
        Prove the telescoping identity $\prod_{k=1}^{n}c_k =
        (n+1)^n$, and deduce, by AM--GM applied to the numbers
        $c_ka_k$,
        \[
        (a_1a_2\cdots a_n)^{1/n} \leq
        \frac{1}{n(n+1)}\sum_{k=1}^{n} c_k a_k
        \qquad (a_k > 0).
        \]
  \item (Carleman's inequality) Sum over $n$, exchange the
        order of summation (positive summable families,
        \cref{thm:b2:series:fubini}), and use $\sum_{n \geq
        k}\frac{1}{n(n+1)} = \frac1k$ and $c_k/k = \bigl(1 +
        \frac1k\bigr)^k < \eu$ to conclude: for every
        convergent $\sum a_k$ with positive terms,
        \[
        \sum_{n=1}^{\infty}(a_1a_2\cdots a_n)^{1/n}
        \;\leq\; \eu\sum_{k=1}^{\infty}a_k .
        \]
  \item For $f$ continuous and positive on $\intcc{0}{1}$,
        prove
        \[
        \Bigl(\int_0^1 f\Bigr)\Bigl(\int_0^1\frac1f\Bigr)
        \geq 1,
        \]
        with equality iff $f$ is constant
        \emph{(Cauchy--Schwarz on $\sqrt f\cdot\frac1{\sqrt
        f}$)}.
  \item (Geometry of the balls) Using the equality case of
        Minkowski, show that for $1 < p < \infty$ the unit
        sphere of $\norm\cdot_p$ contains no segment (the norm
        is strictly convex in the sense of
        \cref{pb:b2:nvs:1}), whereas for $p = 1$ and $p =
        \infty$ it does: exhibit the flat pieces.
\end{enumerate}

\medskip
\noindent\textbf{Part V --- Duality and synthesis.}
\begin{enumerate}[resume]
  \item (H\"older as duality) Prove that for every $a \in
        \R^n$,
        \[
        \norm a_p = \max_{\norm b_q \leq 1}\ \sum_i a_ib_i ,
        \]
        exhibiting a maximizing $b$ explicitly. (The $p$-norm
        is the dual of the $q$-norm --- the finite-dimensional
        germ of $L^p$ duality.)
  \item (Moments) Let $X$ be a random variable taking finitely
        many positive values $x_i$ with probabilities
        $\lambda_i$. Restate question 12 as: $r \mapsto
        \E[X^r]^{1/r}$ is increasing --- the moment
        (Lyapunov) inequality, to be reused in
        \cref{ch:b2:randomvar}.
  \item Solve with named tools, in two lines each: (i) for
        positive $a, b, c$: $a^3 + b^3 + c^3 \geq
        \frac{(a+b+c)^3}{9}$; (ii) for positive $x_1, \dots,
        x_n$: $\bigl(\sum_i\sqrt{x_i}\bigr)^2 \leq
        n\sum_i x_i$.
  \item (Synthesis) Draw the genealogy in five sentences: chord
        definition to slope lemma; slopes to support lines to
        Jensen; $\ln$'s concavity to Young to H\"older to
        Minkowski to the $p$-norms; Jensen to the power-mean
        chain to moments; AM--GM to Carleman. Name the summits
        (H\"older--Minkowski; Carleman), and state where the
        toolbox is headed: the $L^p$ spaces of the Year 3
        volume, whose axioms are exactly questions 6 and 8.
\end{enumerate}
\end{problem}
