\chapter{Line Integrals and Multiple Integrals}\label{ch:b2:multint}

This chapter extends integration from intervals to curves and to
domains of the plane and of space. Line integrals integrate a
\emph{differential form} $P\,\dd x + Q\,\dd y$ along an oriented
arc; double and triple integrals integrate functions over
two- and three-dimensional domains. The two theories meet in the
\emph{Green--Riemann theorem}, the two-dimensional fundamental
theorem of calculus, and the main computational tool throughout is
the \emph{change of variables formula}, whose distortion factor is
the absolute Jacobian determinant.

\section{Line integrals}

\begin{definition}[Differential form; line integral]\label{def:b2:multint:lineint}
Let $U \subseteq \R^2$ be open. A \emph{differential form} of
degree $1$ and class $\mathcal{C}^0$ on $U$ is an expression
$\omega = P\,\dd x + Q\,\dd y$ with $P, Q \colon U \to \R$
continuous --- formally, a continuous map from $U$ into the dual
of $\R^2$, $\omega(M) = P(M)\,e_1^* + Q(M)\,e_2^*$. For a
$\mathcal{C}^1$ arc $\gamma \colon [a, b] \to U$, $\gamma(t) =
(x(t), y(t))$, the \emph{line integral} of $\omega$ along $\gamma$
is
\[
\int_\gamma \omega
= \int_a^b \Bigl(P(\gamma(t))\,x'(t)
+ Q(\gamma(t))\,y'(t)\Bigr)\,\dd t .
\]
The definitions extend verbatim to $\R^3$ (forms $P\,\dd x +
Q\,\dd y + R\,\dd z$) and to piecewise $\mathcal{C}^1$ arcs (sum
over the pieces).
\end{definition}

\begin{proposition}[Invariance and orientation]\label{prop:b2:multint:lineinv}
The line integral is unchanged under an increasing $\mathcal{C}^1$
change of parameter, and changes sign under a decreasing one. It
therefore only depends on the \emph{oriented} geometric arc.
\end{proposition}

\begin{proof}
If $\theta \colon [c, d] \to [a, b]$ is a change of parameter and
$\tilde\gamma = \gamma \circ \theta$, then by the chain rule and
the one-variable change of variables $t = \theta(u)$,
\[
\int_{\tilde\gamma}\omega
= \int_c^d \bigl(P(\gamma(\theta(u)))\,x'(\theta(u))
+ Q(\gamma(\theta(u)))\,y'(\theta(u))\bigr)\,\theta'(u)\,\dd u
= \pm\int_a^b \bigl(Px' + Qy'\bigr)(t)\,\dd t ,
\]
with sign $+$ if $\theta$ is increasing ($\theta(c) = a$) and $-$
if decreasing (the bounds swap).
\end{proof}

\begin{example}[Work of a force; circulation]\label{ex:b2:multint:work}
If $F = (P, Q)$ is a force field, $\int_\gamma P\dd x + Q\dd y =
\int_a^b \langle F(\gamma(t)), \gamma'(t)\rangle\,\dd t$ is the
\emph{work} of $F$ along $\gamma$. For $\omega = -y\,\dd x +
x\,\dd y$ along the counterclockwise unit circle $\gamma(t) =
(\cos t, \sin t)$:
\[
\int_\gamma \omega
= \int_0^{2\pi}\bigl((-\sin t)(-\sin t)
+ \cos t\cos t\bigr)\,\dd t = 2\pi ,
\]
twice the enclosed area --- a first hint of Green--Riemann.
\end{example}

\begin{example}[One integral, two parametrizations, one
sign trap]\label{ex:b2:multint:semicircle}
Compute $\int_\gamma x\,\dd y$ along the upper unit
semicircle from $(1, 0)$ to $(-1, 0)$. With $\gamma(t) =
(\cos t, \sin t)$, $t \in \intcc0\pi$:
\[
\int_0^\pi\cos t\cdot\cos t\,\dd t = \frac\pi2 .
\]
With the graph parametrization $x \mapsto (x, \sqrt{1 -
x^2})$, $x$ from $1$ to $-1$ (note the direction!):
\[
\int_1^{-1}x\cdot\frac{-x}{\sqrt{1 - x^2}}\,\dd x
= \int_{-1}^{1}\frac{x^2}{\sqrt{1 - x^2}}\,\dd x
= \frac\pi2
\]
($x = \sin u$ reduces it to a Wallis integral). Same value,
as \cref{prop:b2:multint:lineinv} guarantees --- but only
because both runs go from $(1,0)$ to $(-1,0)$; reversing the
travel flips the sign. Closing the path along the $x$-axis
(where $\dd y = 0$) adds nothing, and the total $\frac\pi2$
is the area of the half-disk: the first instance of the
boundary-area formulas of Green--Riemann below.
\end{example}

\begin{definition}[Exact and closed forms]\label{def:b2:multint:exact}
The form $\omega = P\,\dd x + Q\,\dd y$ of class $\mathcal{C}^0$
is \emph{exact} on $U$ if there is $f \in \mathcal{C}^1(U)$
(a \emph{potential}) with $\omega = \dd f$, i.e.\ $P = f_x$ and $Q
= f_y$. A $\mathcal{C}^1$ form is \emph{closed} if $P_y = Q_x$ on
$U$.
\end{definition}

\begin{theorem}[Fundamental theorem for line
integrals]\label{thm:b2:multint:ftc}
If $\omega = \dd f$ is exact and $\gamma$ is a piecewise
$\mathcal{C}^1$ arc in $U$ from $A$ to $B$, then
\[
\int_\gamma \omega = f(B) - f(A) .
\]
In particular the integral of an exact form along any closed arc
is zero, and every exact $\mathcal{C}^1$ form is closed.
\end{theorem}

\begin{proof}
$\frac{\dd}{\dd t}f(\gamma(t)) = f_x(\gamma(t))x'(t) +
f_y(\gamma(t))y'(t)$ by the chain rule
(\cref{ch:b2:diffcalc}), so the integrand in
\cref{def:b2:multint:lineint} is the derivative of $t \mapsto
f(\gamma(t))$, and the fundamental theorem of calculus gives the
result on each piece; the intermediate values telescope. Closedness
of exact $\mathcal{C}^1$ forms is Schwarz's theorem: $P_y =
f_{xy} = f_{yx} = Q_x$.
\end{proof}

\begin{example}[Reconstructing a potential]\label{ex:b2:multint:potential}
Let $\omega = y\,\eu^{xy}\,\dd x + (x\,\eu^{xy} + 2y)\,\dd y$
on $\R^2$. It is closed: both cross derivatives equal
$\eu^{xy}(1 + xy)$. To find a potential, integrate $P$ in
$x$ at fixed $y$:
\[
f(x, y) = \int y\,\eu^{xy}\,\dd x = \eu^{xy} + c(y),
\]
then adjust $c$ by matching $f_y$: $x\,\eu^{xy} + c'(y) =
x\,\eu^{xy} + 2y$ gives $c(y) = y^2$. So $f(x,y) = \eu^{xy} +
y^2$, and for any piecewise $\mathcal C^1$ arc from $(0,0)$
to $(1,1)$,
\[
\int_\gamma\omega = f(1,1) - f(0,0) = (\eu + 1) - 1 = \eu ,
\]
independently of the path --- the two-step recipe (integrate
in $x$, correct in $y$) is the practical converse of
\cref{thm:b2:multint:ftc} on domains where closed forms are
exact.
\end{example}

\begin{example}[Closed does not imply exact]\label{ex:b2:multint:angleform}
On $U = \R^2 \setminus \{0\}$, the \emph{angle form}
\[
\omega = \frac{-y\,\dd x + x\,\dd y}{x^2 + y^2}
\]
is closed (direct computation: both $P_y$ and $Q_x$ equal
$\frac{y^2 - x^2}{(x^2+y^2)^2}$), but its integral along the unit
circle is $2\pi \neq 0$ (same computation as
\cref{ex:b2:multint:work}, divided by $1$): $\omega$ is not exact
on $U$. Locally, $\omega = \dd\theta$ for a determination
$\theta$ of the polar angle; the failure is global --- the angle
cannot be defined continuously around the puncture. On domains
without holes the pathology disappears: on a \emph{star-shaped}
open set, every closed $\mathcal{C}^1$ form is exact
(Poincaré's lemma, \cref{exo:b2:multint:8}).
\end{example}

\section{Double integrals}

We take for granted the theory of the Riemann integral in one
variable (Year 1 volume, and \cref{ch:b2:integration}) and sketch
its two-variable version. A function $f$ continuous on a rectangle
$R = [a, b] \times [c, d]$ has a double integral $\iint_R f$,
defined by Riemann sums over grids exactly as in one variable, and
computed by iteration:

\begin{theorem}[Fubini on a rectangle]\label{thm:b2:multint:fubini}
For $f$ continuous on $R = [a,b] \times [c,d]$,
\[
\iint_R f
= \int_a^b \Bigl(\int_c^d f(x, y)\,\dd y\Bigr)\dd x
= \int_c^d \Bigl(\int_a^b f(x, y)\,\dd x\Bigr)\dd y .
\]
\end{theorem}

\begin{proof}
Set $F(x) = \int_c^d f(x, y)\,\dd y$. Uniform continuity of $f$ on
the compact $R$ makes $F$ continuous (dominated estimate:
$\abs{F(x) - F(x')} \leq (d - c)\sup_y\abs{f(x,y) - f(x',y)}$).
Now subdivide $[a,b]$ and $[c,d]$ into $n$ equal parts, giving a
grid of cells $R_{ij}$ of area $\Delta x\,\Delta y$. On each cell,
$\inf_{R_{ij}} f \cdot \Delta x \Delta y \leq
\int_{x_{i-1}}^{x_i}\int_{y_{j-1}}^{y_j} f(x,y)\,\dd y\,\dd x \leq
\sup_{R_{ij}} f \cdot \Delta x \Delta y$ by monotonicity of the
one-variable integral (applied twice). Summing over cells, the
iterated integral $\int_a^b F$ is squeezed between the lower and
upper Riemann sums of the grid; by uniform continuity both sums
converge to the common value defining $\iint_R f$ as $n \to
\infty$. The same argument applies with the roles of $x$ and $y$
exchanged, so both iterated integrals equal $\iint_R f$.
\end{proof}

\begin{remark}
Continuity on a compact rectangle is doing real work in
Fubini's proof: it provides the uniform continuity that
squeezes the Riemann sums. For wilder integrands the
statement genuinely fails --- there are functions whose two
iterated integrals exist and differ. The honest general
theorem, with integrability as the only hypothesis, is
Fubini's theorem for the Lebesgue integral, proved in the
Year 3 volume; everything in this chapter stays within the
continuous setting where the elementary proof above is
complete.
\end{remark}

\begin{definition}[Elementary domains]\label{def:b2:multint:domain}
A domain $D \subseteq \R^2$ is \emph{$y$-elementary} if
\[
D = \{(x, y) : a \leq x \leq b,\
\varphi_1(x) \leq y \leq \varphi_2(x)\}
\]
with $\varphi_1 \leq \varphi_2$ continuous on $[a,b]$
($x$-elementary: symmetrically). For $f$ continuous on a
$y$-elementary $D$,
\[
\iint_D f
= \int_a^b\Bigl(
\int_{\varphi_1(x)}^{\varphi_2(x)} f(x,y)\,\dd y\Bigr)\dd x ,
\]
and one checks (by extending $f$ by an approximation argument, or
by subdividing) that when $D$ is elementary in both directions the
two iterated integrals agree. Domains cut into finitely many
elementary pieces are handled by additivity, and $\operatorname
{Area}(D) = \iint_D 1$.
\end{definition}

\begin{example}\label{ex:b2:multint:triangle}
On the triangle $D = \{0 \leq x \leq 1,\ 0 \leq y \leq x\}$:
\[
\iint_D xy \,\dd x\,\dd y
= \int_0^1 x\Bigl(\int_0^x y\,\dd y\Bigr)\dd x
= \int_0^1 x\cdot\frac{x^2}{2}\,\dd x = \frac18 .
\]
Swapping the order ($x$ from $y$ to $1$): $\int_0^1
y\bigl(\int_y^1 x\,\dd x\bigr)\dd y = \int_0^1
y\,\frac{1 - y^2}{2}\,\dd y = \frac18$ --- same value, different
computation: choosing the order of integration well is half the
craft.
\end{example}

\begin{example}[When only one order works]\label{ex:b2:multint:swaporder}
Compute $I = \displaystyle\int_0^1\!\!\int_x^1
\eu^{y^2}\,\dd y\,\dd x$. As written, the inner integral
$\int\eu^{y^2}\dd y$ has no elementary antiderivative: the
computation is stuck. But the domain is the triangle $0 \leq
x \leq y \leq 1$, which is elementary in both directions;
swapping the order,
\[
I = \int_0^1\!\!\int_0^y \eu^{y^2}\,\dd x\,\dd y
= \int_0^1 y\,\eu^{y^2}\,\dd y
= \Bigl[\tfrac12\eu^{y^2}\Bigr]_0^1 = \frac{\eu - 1}{2} .
\]
The inner variable $x$ appeared nowhere in the integrand, so
integrating it first produced exactly the factor $y$ that
makes the outer integral immediate. Moral: Fubini is not just
a license to iterate --- it is a license to \emph{choose},
and the right order can turn an impossible integral into a
one-liner. Always sketch the domain and read off both
descriptions before starting.
\end{example}

\begin{theorem}[Change of variables]\label{thm:b2:multint:changevar}
Let $\Phi \colon U' \to U$ be a $\mathcal{C}^1$ diffeomorphism
between open sets of $\R^2$, let $K \subseteq U$ be a compact
domain cut into elementary pieces with $K' = \Phi^{-1}(K)$, and
let $f$ be continuous on $K$. Then
\[
\iint_K f(x, y)\,\dd x\,\dd y
= \iint_{K'} f\bigl(\Phi(u, v)\bigr)\,
\abs{\det J_\Phi(u, v)}\,\dd u\,\dd v .
\]
\end{theorem}
\admitted

\begin{remark}
The complete proof --- approximating $\Phi$ by its differential on
a fine grid and controlling the boundary cells --- is long though
not deep; it is done in full in the measure theory of Year 3, as a
consequence of the Lebesgue theory. The heuristic is the picture
already used for surface area: a small square of side $\dd u$ at
$(u, v)$ is mapped, to first order, onto the parallelogram spanned
by $\Phi_u\,\dd u$ and $\Phi_v\,\dd v$, whose area is
$\abs{\det J_\Phi}\,\dd u\,\dd v$
(\cref{lem:b2:surfaces:lagrange}).
\end{remark}

\begin{remark}[Method: choosing the change of variables]
Three reflexes cover most cases. \emph{Symmetry of the
integrand}: $x^2 + y^2$ calls for polar, a product structure
calls for keeping Cartesian axes. \emph{Shape of the
boundary}: boundaries $u(x,y) = c_1$, $v(x,y) = c_2$ beg for
the coordinates $(u, v)$ themselves, as in the hyperbolic
region example below --- the domain becomes a rectangle,
which is the whole victory. \emph{Linear structure}:
expressions in $x + y$ and $x - y$ invite the $45$-degree
rotation or a shear (\cref{ex:b2:multint:affine}). In all
cases, three boxes to tick before integrating: the map is a
bijection of the new domain onto the old one; its Jacobian
is computed \emph{in the direction actually used} (invert at
the end if easier); and the Jacobian enters with its
absolute value.
\end{remark}

\begin{example}[Polar coordinates]\label{ex:b2:multint:polar}
$\Phi(\rho, \alpha) = (\rho\cos\alpha,\ \rho\sin\alpha)$ has
\[
J_\Phi = \begin{pmatrix}
\cos\alpha & -\rho\sin\alpha\\
\sin\alpha & \rho\cos\alpha
\end{pmatrix},
\qquad \det J_\Phi = \rho ,
\]
so $\dd x\,\dd y = \rho\,\dd\rho\,\dd\alpha$. For the disk $D_R$
of radius $R$:
\[
\iint_{D_R} e^{-(x^2 + y^2)}\,\dd x\,\dd y
= \int_0^{2\pi}\!\!\int_0^R e^{-\rho^2}\rho\,\dd\rho\,\dd\alpha
= \pi\bigl(1 - e^{-R^2}\bigr)
\xrightarrow[R\to\infty]{} \pi .
\]
Comparing with the square $[-R, R]^2$ (which is squeezed between
the disks $D_R$ and $D_{R\sqrt2}$, all integrands positive) gives
$\bigl(\int_{-\infty}^\infty e^{-x^2}\dd x\bigr)^2 = \pi$:
\[
\boxed{\ \int_{-\infty}^{+\infty} e^{-x^2}\,\dd x = \sqrt{\pi}\ }
\]
--- the Gauss integral again, now by its most famous proof
(compare the one-variable derivation in
\cref{ch:b2:integration}).
\end{example}

\begin{example}[Affine changes of variables]\label{ex:b2:multint:affine}
For an affine map $\Phi(u, v) = M(u, v)^{\mathsf T} + C$ with
$M$ invertible, the Jacobian is the constant matrix $M$:
areas are multiplied by the constant factor $\abs{\det M}$
--- the promise made in \cref{ch:b2:affine} is now a theorem.
Two immediate uses. The ellipse $\frac{x^2}{a^2} +
\frac{y^2}{b^2} \leq 1$ is the image of the unit disk under
$(u, v) \mapsto (au, bv)$, so its area is $ab \cdot \pi$ ---
no computation. And for the integral of $f(x + y)$ over the
square $K = \intcc01^2$, the shear $\Phi(u, v) = (u - v, v)$
(determinant $1$) turns it into an integral of $f(u)$ over a
parallelogram, which Fubini slices at constant $u$: with $f =
\exp$,
\[
\iint_K \eu^{x+y}\,\dd x\,\dd y
= \Bigl(\int_0^1 \eu^x\,\dd x\Bigr)^2 = (\eu - 1)^2,
\]
as the product structure confirms. Choosing coordinates
adapted to the integrand --- not to the domain --- is the
other half of the craft.
\end{example}


\begin{example}[Coordinates adapted to a curvilinear
domain]\label{ex:b2:multint:hyperbolicregion}
Let $D$ be the region of the first quadrant bounded by the
hyperbolas $xy = 1$ and $xy = 3$ and the lines $y = x$ and
$y = 3x$. In the coordinates $u = xy$, $v = y/x$ the domain
becomes the square $\intcc13 \times \intcc13$; inverting,
\[
x = \sqrt{u/v}, \qquad y = \sqrt{uv},
\qquad
\det J = x_uy_v - x_vy_u = \frac{1}{2v}
\]
(a two-line computation with $x = u^{1/2}v^{-1/2}$, $y =
u^{1/2}v^{1/2}$). Hence
\[
\operatorname{Area}(D)
= \int_1^3\!\!\int_1^3\frac{\dd u\,\dd v}{2v}
= 2\cdot\frac{\ln 3}{2} = \ln 3 \approx 1.10 .
\]
Trying to slice $D$ in Cartesian coordinates means cutting it
into three pieces with hyperbolic and linear boundaries ---
feasible, joyless, and error-prone. The moral repeats
\cref{ex:b2:multint:affine} at full strength: read the
boundary equations, and let \emph{them} choose the
coordinates; the Jacobian then converts the curvilinear
mesh's cell area, exactly as $\rho$ did for polar
coordinates.
\end{example}

\begin{example}[Average values]\label{ex:b2:multint:average}
The \emph{mean value} of $f$ over a domain $D$ is
$\frac1{\operatorname{Area}(D)}\iint_Df$. Sample: the average
distance from the center for a uniformly chosen point of the
disk of radius $R$ is
\[
\frac{1}{\pi R^2}\int_0^{2\pi}\!\!\int_0^R
\rho\cdot\rho\,\dd\rho\,\dd\alpha
= \frac{2\pi R^3/3}{\pi R^2} = \frac{2R}3 ,
\]
not $R/2$: uniform area puts more mass at large radii (the
annulus at radius $\rho$ has weight proportional to $\rho$),
so the average sits beyond the halfway mark. Getting this
factor right is exactly the polar Jacobian at work, and the
same weighting explains the centroid $\bar z = 3R/8$ of the
half-ball computed later in the chapter rather than $R/2$.
\end{example}

\section{The Green--Riemann theorem}

\begin{theorem}[Green--Riemann]\label{thm:b2:multint:green}
Let $K \subseteq \R^2$ be a compact domain which is elementary in
both directions (or a finite union of such glued along segments),
with boundary $\partial K$ a piecewise $\mathcal{C}^1$ closed
curve oriented \emph{counterclockwise} (the domain stays on the
left). For $P, Q$ of class $\mathcal{C}^1$ on a neighbourhood of
$K$:
\[
\oint_{\partial K} P\,\dd x + Q\,\dd y
= \iint_K \Bigl(\frac{\partial Q}{\partial x}
- \frac{\partial P}{\partial y}\Bigr)\,\dd x\,\dd y .
\]
\end{theorem}

\begin{proof}
First, both sides are additive under cutting $K$ along a
segment into two pieces $K_1, K_2$: the double integrals add
by additivity of $\iint$; as for the boundary integrals, the
counterclockwise boundaries of $K_1$ and $K_2$ each traverse
the interior cut once, in \emph{opposite} directions, so in
the sum
\[
\oint_{\partial K_1} + \oint_{\partial K_2}
= \oint_{\partial K} + (\text{cut, both ways})
= \oint_{\partial K},
\]
the two passes along the cut cancel
(\cref{prop:b2:multint:lineinv}) and only the outer boundary
survives. Iterating finitely many cuts, it suffices to treat
an elementary domain. We prove $\oint P\,\dd x = -\iint_K P_y$ on a
$y$-elementary domain $D = \{a \leq x \leq b,\ \varphi_1(x) \leq y
\leq \varphi_2(x)\}$; the identity $\oint Q\,\dd y = \iint_K Q_x$
is symmetric ($x$-elementary), and the theorem is their sum.

Compute the double integral by Fubini and the one-variable
fundamental theorem:
\[
\iint_D \frac{\partial P}{\partial y}\,\dd x\,\dd y
= \int_a^b \bigl(P(x, \varphi_2(x)) - P(x, \varphi_1(x))\bigr)
\,\dd x .
\]
Now the boundary of $D$, counterclockwise, consists of: the lower
graph $y = \varphi_1(x)$ traversed left to right, the right
vertical segment $x = b$ (upward), the upper graph $y =
\varphi_2(x)$ traversed \emph{right to left}, the left vertical
segment $x = a$ (downward). Along the vertical segments $x$ is
constant, so they contribute $0$ to $\oint P\,\dd x$; the graphs,
parametrized by $x$, give
\[
\oint_{\partial D} P\,\dd x
= \int_a^b P(x, \varphi_1(x))\,\dd x
- \int_a^b P(x, \varphi_2(x))\,\dd x
= -\iint_D \frac{\partial P}{\partial y}\,\dd x\,\dd y .
\qedhere
\]
\end{proof}

\begin{corollary}[Area by the boundary]\label{cor:b2:multint:area}
Under the hypotheses of \cref{thm:b2:multint:green},
\[
\operatorname{Area}(K)
= \oint_{\partial K} x\,\dd y
= -\oint_{\partial K} y\,\dd x
= \frac12\oint_{\partial K} x\,\dd y - y\,\dd x .
\]
\end{corollary}

\begin{proof}
Apply Green--Riemann to $(P, Q) = (0, x)$, $(-y, 0)$ and
$\frac12(-y, x)$: each time $Q_x - P_y = 1$.
\end{proof}

\begin{remark}[Choosing among the three area formulas]
The three boundary formulas are equal, not interchangeable in
practice. Use $\oint x\,\dd y$ when the parametrization makes
$\dd y$ simple (graphs over the $y$-axis), $-\oint y\,\dd x$
symmetrically, and the symmetric half-sum when the
parametrization treats $x$ and $y$ evenly --- for the ellipse
it produced a constant integrand, with no trigonometric
linearization at all. On polygonal boundaries the half-sum
becomes the shoelace formula of \cref{exo:b2:multint:12},
the surveyors' algorithm. And when the boundary is traversed
clockwise by the given parametrization, all three formulas
return \emph{minus} the area: a negative result is not a
computational error but an orientation report --- flip the
sign, or the parametrization.
\end{remark}

\begin{example}[Area of the ellipse]\label{ex:b2:multint:ellipsearea}
For $x = a\cos t$, $y = b\sin t$, $t \in [0, 2\pi]$:
\[
\operatorname{Area}
= \frac12\int_0^{2\pi}\bigl(a\cos t \cdot b\cos t
- b\sin t\cdot(-a\sin t)\bigr)\,\dd t
= \frac{ab}{2}\int_0^{2\pi}\dd t = \pi ab .
\]
\end{example}

\begin{example}[Green--Riemann as a
cross-check]\label{ex:b2:multint:greencheck}
Take $P = -y^3$, $Q = x^3$ on the closed unit disk $D$.
Boundary side, with $\gamma(t) = (\cos t, \sin t)$:
\[
\oint_{\partial D}P\,\dd x + Q\,\dd y
= \int_0^{2\pi}\bigl(\sin^4 t + \cos^4 t\bigr)\dd t
= 2\pi\cdot\Bigl(\frac38 + \frac38\Bigr) = \frac{3\pi}2 ,
\]
by linearization ($\sin^4 + \cos^4 = \tfrac34 +
\tfrac14\cos4t$). Interior side:
\[
\iint_D(Q_x - P_y)\,\dd x\,\dd y
= \iint_D 3(x^2 + y^2)\,\dd x\,\dd y
= 3\int_0^{2\pi}\!\!\int_0^1\rho^3\,\dd\rho\,\dd\alpha
= \frac{3\pi}2 .
\]
Same number, two very different computations --- and that is
the practical use: whichever side of Green's identity is
easier becomes the computation, the other a verification. For
circulations of polynomial fields around closed curves, the
double integral is almost always the easy side.
\end{example}

\begin{remark}
Green--Riemann explains \cref{ex:b2:multint:angleform}: for a
closed form ($Q_x = P_y$), the integral around the boundary of any
domain contained in $U$ vanishes. The angle form fails to be exact
only because the puncture at the origin prevents the disk bounded
by the unit circle from lying inside $U$ --- line integrals of
closed forms detect the holes of the domain. (Pushed further,
this observation becomes de Rham cohomology.)
\end{remark}

\section{Triple integrals}

The theory extends to three variables with no new idea: Fubini
reduces $\iiint$ to three one-variable integrals (either by
\emph{slicing}: $\iiint_K f = \int\bigl(\iint_{K_z}
f\bigr)\dd z$ over the horizontal slices $K_z$, or by
\emph{stacking}: integrating in $z$ first along vertical sticks),
and the change of variables formula holds with the $3 \times 3$
Jacobian.

\begin{example}[Cylindrical and spherical coordinates]\label{ex:b2:multint:spherical}
\emph{Cylindrical} $(x, y, z) = (\rho\cos\alpha, \rho\sin\alpha,
z)$: $\dd x\,\dd y\,\dd z = \rho\,\dd\rho\,\dd\alpha\,\dd z$.
\emph{Spherical} $(x, y, z) = (r\cos\theta\cos\varphi,\
r\sin\theta\cos\varphi,\ r\sin\varphi)$ ($\theta$ longitude,
$\varphi \in [-\frac\pi2, \frac\pi2]$ latitude): expanding the $3
\times 3$ determinant along the last row,
\[
\det J = r^2\cos\varphi ,
\qquad
\dd x\,\dd y\,\dd z
= r^2\cos\varphi\;\dd r\,\dd\theta\,\dd\varphi .
\]
Volume of the ball of radius $R$:
\[
V = \int_0^R\!\!\int_0^{2\pi}\!\!\int_{-\pi/2}^{\pi/2}
r^2\cos\varphi\;\dd\varphi\,\dd\theta\,\dd r
= \frac{R^3}{3}\cdot 2\pi \cdot 2
= \boxed{\frac43\pi R^3} ,
\]
discharging at last the formula admitted in the volume chapters of
the earlier books.
\end{example}

\begin{example}[The tetrahedron, twice]
\label{ex:b2:multint:tetra}
The volume of $T = \{x, y, z \geq 0,\ x + y + z \leq 1\}$, by
stacking: for fixed $(x, y)$ in the triangle $x + y \leq 1$,
$z$ runs over $\intcc0{1 - x - y}$, so
\[
V = \int_0^1\!\!\int_0^{1-x}(1 - x - y)\,\dd y\,\dd x
= \int_0^1\frac{(1 - x)^2}{2}\,\dd x = \frac16 .
\]
By slicing: the section at height $z$ is the triangle $\{x, y
\geq 0,\ x + y \leq 1 - z\}$, of area $\frac{(1-z)^2}2$, and
$V = \int_0^1\frac{(1-z)^2}2\,\dd z = \frac16$ again --- the
two computations are the same integrals in a different
order, which is all Fubini claims. The value $\frac16 =
\frac13\cdot\frac12\cdot1$ is the cone formula
(\cref{ex:b2:multint:cone}) with triangular base, and the
$n$-dimensional version $1/n!$ is proved by exactly this
slicing in the weekend problem.
\end{example}

\begin{example}[Centroid of a half-ball]\label{ex:b2:multint:centroid}
For the upper half-ball $H$ of radius $R$ ($z \geq 0$), the
centroid height is $\bar z = \frac1{V}\iiint_H z$, with $V =
\frac23\pi R^3$. In spherical coordinates ($z =
r\sin\varphi$, $\varphi \in \intcc0{\pi/2}$):
\[
\iiint_H z
= \int_0^R r^3\,\dd r\int_0^{2\pi}\dd\theta
\int_0^{\pi/2}\sin\varphi\cos\varphi\,\dd\varphi
= \frac{R^4}4\cdot2\pi\cdot\frac12 = \frac{\pi R^4}4 ,
\]
so
\[
\bar z = \frac{\pi R^4/4}{2\pi R^3/3} = \frac{3R}8 :
\]
the balance point of a solid hemisphere sits three-eighths of
the radius above the flat face --- below the half-height
$R/2$, as it must, since the solid is fatter near the base.
Every centroid computation has this shape: one moment
integral, one volume, one ratio, and a plausibility check
against the geometry.
\end{example}

\begin{example}[Volume by slicing: the cone]\label{ex:b2:multint:cone}
A cone of base area $A$ and height $h$ (apex up, base at $z = 0$):
the slice at height $z$ is the base scaled by the factor $(1 -
z/h)$, of area $A(1 - z/h)^2$. Hence
\[
V = \int_0^h A\Bigl(1 - \frac zh\Bigr)^2\dd z = \frac{Ah}{3} :
\]
the one-third of the school formulas, valid for \emph{any} base
shape --- slicing turns it into the integral of a square.
\end{example}

\begin{example}[Integrability thresholds in the
plane]\label{ex:b2:multint:threshold}
For which $\alpha > 0$ does $\iint_{D}\rho^{-\alpha}\,\dd
x\,\dd y$ converge on the punctured unit disk $D$ (limit
over annuli $\varepsilon \leq \rho \leq 1$)? In polar
coordinates,
\[
\int_0^{2\pi}\!\!\int_\varepsilon^1\rho^{-\alpha}\,
\rho\,\dd\rho\,\dd\alpha
= 2\pi\int_\varepsilon^1\rho^{1-\alpha}\,\dd\rho ,
\]
which converges as $\varepsilon \to 0$ iff $1 - \alpha > -1$,
i.e.\ $\alpha < 2$: in dimension $2$ the critical singularity
exponent is the dimension itself, the extra $\rho$ from the
Jacobian softening the singularity by one power. (Likewise
$\alpha < 3$ for a point singularity in space, via $r^2$.)
Radial bookkeeping of this kind is how integrability is
decided at a glance in the Lebesgue framework of Year 3 ---
and it is the reason $\iiint 1/r$ converged effortlessly in
\cref{exo:b2:multint:7}.
\end{example}

\begin{remark}[Common pitfalls]
(i) \emph{Orientation}: a line integral changes sign with the
direction of travel, and Green--Riemann requires the boundary
counterclockwise (domain on the left); for a domain with a
hole, the inner boundary is traversed \emph{clockwise}. (ii)
\emph{The Jacobian enters with an absolute value}: change of
variables never produces negative area, and forgetting
$\abs{\det}$ typically flips signs exactly when the map
reverses orientation. (iii) \emph{The polar factor $\rho$}:
$\dd x\,\dd y = \rho\,\dd\rho\,\dd\alpha$, not
$\dd\rho\,\dd\alpha$ --- the most common error in the whole
chapter; dimensional analysis catches it, since
$\dd\rho\,\dd\alpha$ has the dimension of a length, not an
area. (iv) \emph{Improper double integrals}: limits over
growing disks and growing squares agree here because the
integrands are positive (squeezing); for sign-changing
integrands the limit may depend on the exhaustion, and no
claim is made without absolute convergence. (v)
\emph{Domains versus integrands}: a product integrand on a
non-product domain does \emph{not} factor the integral ---
factorization needs both, as in \cref{ex:b2:multint:polar}'s
square.
\end{remark}

\begin{remark}[Perspectives within this volume]
The Gauss integral computed here is quietly everywhere in
the probability chapters: the constant $\sqrt\pi$ inside
Stirling's formula (\cref{thm:b2:comparison:stirling}) is
this chapter's integral, and through Stirling it fixes the
$1/\sqrt{\pi n}$ asymptotics of the random walk's return
probabilities in \cref{ch:b2:proba}. The Wallis integrals of
the weekend problem reappear there too, driving the same
central-binomial estimates. In the other direction, the area
and volume elements of this chapter complete the geometry of
\cref{ch:b2:surfaces}, and Green's formula recomputes the
envelope areas of \cref{ch:b2:curves} (the astroid, in
\cref{exo:b2:multint:5}). One chapter, three services:
measure for geometry, constants for probability, and the
change-of-variables discipline used by both.
\end{remark}

\section{Exercises}

\begin{exercise}[$\star$]\label{exo:b2:multint:1}
Compute $\int_\gamma y^2\,\dd x + x\,\dd y$ along: (a) the segment
from $(0,0)$ to $(1,1)$; (b) the parabola arc $y = x^2$ from
$(0,0)$ to $(1,1)$. Is the form exact?
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b2:multint:2}
Show that $\omega = (2xy + y^3)\,\dd x + (x^2 + 3xy^2 + 1)\,\dd y$
is closed on $\R^2$, find a potential, and compute
$\int_\gamma\omega$ along any arc from $(0, 0)$ to $(1, 2)$.
\end{exercise}

\begin{exercise}[$\star$]\label{exo:b2:multint:3}
Compute $\iint_D (x + y)\,\dd x\,\dd y$ where $D$ is the domain
bounded by $y = x^2$ and $y = x$ ($0 \leq x \leq 1$), in both
orders of integration.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:multint:4}
Using polar coordinates, compute $\iint_D \frac{\dd x\,\dd y}{(1 +
x^2 + y^2)^2}$ over the whole plane (as a limit over disks), and
$\iint_{D'} xy\,\dd x\,\dd y$ over the quarter disk $D' = \{x, y
\geq 0,\ x^2 + y^2 \leq 1\}$.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:multint:5}
Compute the area enclosed by the astroid $x = \cos^3 t$, $y =
\sin^3 t$, $t \in [0, 2\pi]$, using
\cref{cor:b2:multint:area}. \emph{(Linearize $\sin^2 t\cos^2 t$.)}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:multint:6}
Compute the volume of the solid bounded below by the paraboloid $z
= x^2 + y^2$ and above by the plane $z = 1$, by both methods:
stacking (integrate $1 - x^2 - y^2$ over the unit disk, polar
coordinates) and slicing (horizontal slices are disks of radius
$\sqrt z$).
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:multint:7}
(Gravitational attraction of a ball --- Newton's theorem, special
case) Show that the volume of the spherical shell $a \leq r \leq
b$ is $\frac43\pi(b^3 - a^3)$ and compute $\iiint_{B}
\frac{\dd x\,\dd y\,\dd z}{r}$ over the ball $B$ of radius $R$
($r$ the distance to the origin). \emph{(Spherical
coordinates.)}
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b2:multint:8}
(Poincaré's lemma, star-shaped case) Let $U$ be star-shaped with
respect to $0$ (i.e.\ $M \in U \Rightarrow [0, M] \subseteq U$)
and $\omega = P\dd x + Q\dd y$ a closed $\mathcal{C}^1$ form on
$U$. Define
\[
f(x, y) = \int_0^1 \bigl(x\,P(tx, ty) + y\,Q(tx, ty)\bigr)\dd t .
\]
Using differentiation under the integral sign
(\cref{ch:b2:integration}) and $P_y = Q_x$, show that $f_x = P$
and $f_y = Q$: every closed form on a star-shaped open set is
exact.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b2:multint:9}
(Dirichlet integral by double integration) Justify and exploit
\[
\int_0^\infty\!\!\int_0^\infty e^{-xy}\sin x\;\dd y\,\dd x
\quad\text{vs}\quad
\int_0^\infty\!\!\int_0^\infty e^{-xy}\sin x\;\dd x\,\dd y
\]
on $[0, A] \times [0, \infty)$: show $\int_0^A \frac{\sin
x}{x}\dd x = \frac\pi2 - \int_0^\infty
e^{-Ay}\frac{y\sin A + \cos A}{1 + y^2}\dd y$ and recover
$\int_0^\infty \frac{\sin x}{x}\,\dd x = \frac\pi2$, comparing
with the parameter-integral proof of
\cref{ch:b2:integration}.
\end{exercise}

\begin{exercise}[$\star\star\star$]\label{exo:b2:multint:10}
(Isoperimetric inequality via Wirtinger) Let $\gamma$ be a simple
closed $\mathcal{C}^1$ curve of length $2\pi$, parametrized by arc
length on $[0, 2\pi]$, enclosing area $A$. Using
\cref{cor:b2:multint:area}, Parseval and the Wirtinger inequality
(exercises of \cref{ch:b2:fourier}), prove $A \leq \pi$, with
equality for the circle. \emph{(Normalize $\int_0^{2\pi} x(s)\dd s
= 0$; write $2A = \oint x\,\dd y - y\,\dd x$ and bound $2A \leq
\int (x^2 + y'^2)$ carefully via $2A = \int_0^{2\pi}(xy' - yx')\dd
s$ and $x^2 + y'^2 \geq 2xy'$.)}
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:multint:11}
(Moments of the ball) For the ball $B$ of radius $R$ in
$\R^3$, compute $\iiint_B z^2\,\dd x\,\dd y\,\dd z$ in
spherical coordinates, and deduce $\iiint_B (x^2 + y^2 +
z^2)\,\dd x\,\dd y\,\dd z$ by symmetry. Cross-check the
latter against the shell computation $\int_0^R r^2\cdot4\pi
r^2\,\dd r$.
\end{exercise}

\begin{exercise}[$\star\star$]\label{exo:b2:multint:12}
(Shoelace formula) Let $K$ be a polygon with vertices $(x_1,
y_1), \dots, (x_m, y_m)$ in counterclockwise order (indices
mod $m$). Deduce from \cref{cor:b2:multint:area} that
\[
\operatorname{Area}(K)
= \frac12\sum_{i=1}^m
\bigl(x_iy_{i+1} - x_{i+1}y_i\bigr) ,
\]
and check the formula on the triangle $(0,0)$, $(1,0)$,
$(0,1)$.
\end{exercise}

\section{Problem: the volume of the ball in dimension $n$}

\begin{problem}[{Weekend problem --- $V_n =
\pi^{n/2}/\Gamma(\frac n2 + 1)$, and the strangeness of high
dimensions}]\label{pb:b2:multint:1}
The disk has area $\pi$, the ball volume $\frac43\pi$ --- and
then? This problem computes the volume of the unit ball of
$\R^n$ for every $n$, twice (by a slicing recursion driven by
the Wallis integrals, then via the $\Gamma$ function and the
Gauss integral of \cref{ex:b2:multint:polar}), and then reads
off the geometry: the volumes peak in dimension five and rush
to zero, and almost all of a high-dimensional ball hides in a
thin shell near its boundary.\index{ball volume}\index{Wallis
integrals} For a continuous function on a ball of $\R^n$, the
integral is understood as the $n$-fold iterated integral
(slicing one coordinate at a time, as in the chapter for $n
\leq 3$); we write $B_n(R)$ for the closed ball of radius $R$
centered at $0$, $v_n(R)$ for its volume, and $V_n =
v_n(1)$, with $V_0 = 1$ by convention.

\medskip
\noindent\textbf{Part I --- The slicing recursion.}
\begin{enumerate}
  \item Substituting $x_i = Ru_i$ in each of the $n$ iterated
        integrals, show $v_n(R) = V_nR^n$.
  \item Slicing $B_n(1)$ along its last coordinate, show
        \[
        V_n = V_{n-1}\int_{-1}^{1}(1 -
        t^2)^{\frac{n-1}2}\,\dd t .
        \]
  \item With $t = \sin\theta$, identify the integral as a
        Wallis integral: $\int_{-1}^1(1 -
        t^2)^{\frac{n-1}2}\dd t = 2W_n$, where $W_n =
        \int_0^{\pi/2}\cos^n\theta\,\dd\theta =
        \int_0^{\pi/2}\sin^n\theta\,\dd\theta$.
  \item Prove the two Wallis identities (integrate by parts;
        then telescope $nW_nW_{n-1}$):
        \[
        W_n = \frac{n-1}nW_{n-2} \quad (n \geq 2),
        \qquad
        W_nW_{n-1} = \frac{\pi}{2n} \quad (n \geq 1).
        \]
\end{enumerate}

\medskip
\noindent\textbf{Part II --- The recursion solved.}
\begin{enumerate}[resume]
  \item Combine questions 2--4 into the two-step recursion
        \[
        V_n = \frac{2\pi}{n}\,V_{n-2}
        \qquad (n \geq 2).
        \]
  \item Deduce the closed forms, for $k \geq 0$:
        \[
        V_{2k} = \frac{\pi^k}{k!},
        \qquad
        V_{2k+1} = \frac{2^{k+1}\pi^k}{1\cdot3\cdot5\cdots
        (2k+1)} .
        \]
  \item Tabulate $V_1, \dots, V_7$ numerically. Using the
        ratio $V_n/V_{n-2} = 2\pi/n$ and the values of $2W_5$
        and $2W_6$, prove that the sequence $(V_n)$ increases
        up to its maximum $V_5 = \frac{8\pi^2}{15} \approx
        5.26$ and decreases from then on.
  \item Show that $V_n \to 0$ faster than any geometric
        sequence, and that $\sum_{n\geq1} V_n$ converges: all
        the unit balls together have finite total volume.
  \item Prove the generating identity
        \[
        \sum_{k\geq0} V_{2k}\,x^{2k} = \eu^{\pi x^2}
        \qquad (x \in \R),
        \]
        and deduce $\sum_{k\geq0}V_{2k} = \eu^\pi \approx
        23.14$.
\end{enumerate}

\medskip
\noindent\textbf{Part III --- Second route: $\Gamma$ and the
Gauss integral.}
\begin{enumerate}[resume]
  \item Show by Fubini (the integrand is a product) that
        \[
        I_n = \int_{\R^n}\eu^{-\norm
        x^2}\dd x
        = \Bigl(\int_{-\infty}^{+\infty}
        \eu^{-t^2}\dd t\Bigr)^{\!n} = \pi^{n/2},
        \]
        the $n$-dimensional Gauss integral, understood as a
        limit over cubes $\intcc{-R}{R}^n$.
  \item Recall $\Gamma(s) = \int_0^\infty t^{s-1}\eu^{-t}\dd
        t$ (\cref{def:b2:integration:gamma}). From
        $\Gamma(s+1) = s\,\Gamma(s)$
        (\cref{thm:b2:integration:gammaprops}) and
        $\Gamma(\tfrac12) = \sqrt\pi$ (substitute $t = u^2$
        and invoke the Gauss integral), compute
        \[
        \Gamma(k + 1) = k!,
        \qquad
        \Gamma\Bigl(k + \frac32\Bigr) =
        \frac{1\cdot3\cdots(2k+1)}{2^{k+1}}\,\sqrt\pi .
        \]
  \item Prove, by induction through the recursion of question
        5, the single formula
        \[
        V_n = \frac{\pi^{n/2}}{\Gamma\bigl(\frac n2 +
        1\bigr)} \qquad (n \geq 1),
        \]
        and check it reproduces both closed forms of question
        6.
  \item Show $\int_0^\infty \eu^{-r^2}r^{n-1}\dd r =
        \tfrac12\Gamma\bigl(\tfrac n2\bigr)$ and deduce the
        identity
        \[
        I_n = n\,V_n\int_0^\infty \eu^{-r^2}\,r^{n-1}\,\dd r.
        \]
        Interpret it: the Gaussian mass of $\R^n$ is
        collected along spherical shells whose
        ``$(n-1)$-dimensional area'' at radius $r$ is
        $nV_nr^{n-1}$ --- both sides are now proved
        independently, so the interpretation costs nothing.
  \item Set $s_{n-1} = nV_n$ (the area of the unit sphere
        $S^{n-1}$, consistently with $v_n(R) = \int_0^R
        s_{n-1}r^{n-1}\dd r$). Tabulate $s_0, \dots, s_3$ and
        check $s_1 = 2\pi$, $s_2 = 4\pi$, $s_3 = 2\pi^2$.
\end{enumerate}

\medskip
\noindent\textbf{Part IV --- High dimensions are strange.}
\begin{enumerate}[resume]
  \item From Stirling's formula
        (\cref{thm:b2:comparison:stirling}) applied to $k!$,
        show for even $n = 2k$:
        \[
        V_n \sim \frac{1}{\sqrt{\pi n}}
        \Bigl(\frac{2\pi\eu}{n}\Bigr)^{n/2}
        \qquad (n \to \infty, \ n \text{ even}),
        \]
        and explain why the same super-geometric decay bound
        extends to odd $n$ via the recursion.
  \item The unit ball sits in the cube $\intcc{-1}1^n$ of
        volume $2^n$. Compute the filling ratio $V_n/2^n$ for
        $n = 2, 3, 10$, and show it tends to $0$: in high
        dimension, essentially all of the cube lies in its
        corners.
  \item Show that the fraction of $v_n(1)$ lying within
        distance $\varepsilon$ of the boundary sphere is $1 -
        (1 - \varepsilon)^n \to 1$; numerically, what
        fraction of a $100$-dimensional ball lies in the
        outer shell of thickness $1\%$?
  \item Prove the Wallis asymptotic $W_n \sim
        \sqrt{\dfrac{\pi}{2n}}$ \emph{(monotonicity of
        $(W_n)$, the ratio $W_n/W_{n-2} \to 1$, and $W_nW_{n-1}
        = \frac\pi{2n}$)}, and the lower bound $W_n \geq
        \sqrt{\dfrac{\pi}{2(n+1)}}$ for all $n$.
  \item (Concentration on a slab) The fraction of the unit
        ball with first coordinate beyond $\delta$ is
        $\int_\delta^1(1 - x^2)^{\frac{n-1}2}\dd x \,\big/\,
        (2W_n)$. Using $1 - u \leq \eu^{-u}$ and the tail
        bound $\int_\delta^\infty \eu^{-a x^2}\dd x \leq
        \frac{\eu^{-a\delta^2}}{2a\delta}$, show this
        fraction is at most
        \[
        \frac{\eu^{-(n-1)\delta^2/2}}{(n-1)\,\delta}
        \Big/ \sqrt{\frac{2\pi}{n+1}}
        \]
        and conclude: for $\delta = s/\sqrt{n-1}$, all but a
        fraction $O(\eu^{-s^2/2}/s)$ of the ball lies in the
        slab $\abs{x_1} \leq s/\sqrt{n-1}$. A ball of high
        dimension is, statistically, a thin pancake in every
        direction at once.
  \item Assemble questions 16--19 into one paragraph: where
        the volume of $B_n(1)$ sits (near the boundary
        sphere, yet within $O(1/\sqrt n)$ slabs of every
        hyperplane through the center), and why these two
        statements do not contradict each other.
\end{enumerate}

\medskip
\noindent\textbf{Part V --- Other bodies, and synthesis.}
\begin{enumerate}[resume]
  \item (Simplex) Let $\Delta_n = \{x \in \R^n : x_i \geq 0,\
        \sum x_i \leq 1\}$. Show by slicing and induction that
        $\operatorname{vol}(\Delta_n) = \frac1{n!}$.
  \item (Cross-polytope) Deduce that $C_n = \{x :
        \sum\abs{x_i} \leq 1\}$ has volume $\frac{2^n}{n!}$,
        and verify the sandwich $C_n \subseteq B_n(1)
        \subseteq \intcc{-1}1^n$ at the level of volumes:
        $\frac{2^n}{n!} \leq V_n \leq 2^n$.
  \item Compute $V_4$ a third way: slice $\R^4 = \R^2 \times
        \R^2$, integrate the area of the $(z, w)$-disk over
        the $(x, y)$-disk in polar coordinates, and recover
        $V_4 = \frac{\pi^2}2$.
  \item (Monte Carlo in trouble) A point is drawn uniformly
        in the cube $\intcc{-1}1^{20}$. Show that the
        probability that it lands in the inscribed ball is
        $V_{20}/2^{20} \approx 2.5\cdot10^{-8}$, so about
        forty million draws are needed before the first hit
        is expected: estimating $V_n$ by rejection sampling
        collapses in high dimension (the curse of
        dimensionality).
  \item Synthesis. Two independent derivations met at $V_n =
        \pi^{n/2}/\Gamma(\frac n2 + 1)$: list which theorem
        of this chapter each one used (Fubini, change of
        variables, the polar Gauss integral), and which
        one-variable inputs (Wallis, $\Gamma$, Stirling).
        Where does the Year 3 volume redo this computation
        with Lebesgue's theory, and what does it add?
\end{enumerate}
\end{problem}
